### 2021-02-04 SPIE/article/main.tex
added final figures, some caption text

%\documentclass[]{spie}  %>>> use for US letter paper
\documentclass[a4paper]{spie}  %>>> use this instead for A4 paper
%\documentclass[nocompress]{spie}  %>>> to avoid compression of citations

\renewcommand{\baselinestretch}{1.0} % Change to 1.65 for double spacing

\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[dvipsnames]{xcolor}

\title{Quantifying the Impact \\ of Performance Improvements and Cost Reductions \\ from 20 years of Light-Emitting Diode Manufacturing}

\author[a,b]{Michael Weinold}
\author[b]{Sergey Kolesnikov}
\author[c]{Laura Diaz Anadon}
\affil[a]{ETH Zurich, Chair of Entrepreneurial Risks, Scheuchzerstrasse 7, 8092 Zurich, Switzerland}
\affil[b]{University of Cambridge, Centre for Environment, Energy and Natural Resource Governance, The David Attenborough Building, CB2 3QZ Cambridge, UK}

\authorinfo{Further author information: (Send correspondence to Michael Weinold)\\Michael Weinold: E-mail: michael.weinold@alumni.ethz.ch}

% Option to view page numbers
\pagestyle{empty} % change to \pagestyle{plain} for page numbers   
\setcounter{page}{301} % Set start page numbering at e.g. 301

\begin{document} 
\maketitle

\begin{abstract}
We collected historical data on device performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years. We used this information to identify and quantify the principal sources of performance improvements in LED manufacturing. We found that in order to quantify the impact of single technological changes, it is necessary to analyse performance improvements the device sub-efficiency level. We further developed a bottom-up manufacturing cost model with process step resolution that captures improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale to analyse cost reductions and their sources. It covers progress from early manufacturing in 2003 to today. We found that larger wafer sizes have been largely responsible for cost reductions. We found that XXX well known XXX.

\end{abstract}

% Include a list of keywords after the abstract 
\keywords{light-emitting diodes, innovation, efficiency, cost}

\section{INTRODUCTION}
\label{sec:intro}  % \label{} allows reference to this section

\clearpage
\section{METRICS}

Why did we choose the metrics we did?
What are the limitations of eg. Haitz's Law, Cost per Lamp, CRI, etc.

\begin{figure} [ht]
    \begin{center}
        \includegraphics[width=\textwidth]{haitz_law_white.pdf}
    \end{center}
    \caption{Historical increase in flux for white light-emitting diode chips and packages, inspired by \textit{"Haitz's Law"}. Note that the increase in flux per single chip is not increasing as starkly as the flux for multi-chip packages. Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.}
\end{figure}

Mention previous work by Tsao et al.
Mention limitations: Not done for 2003 (early devices) and 2020 (most recent devices)

The overall efficiency of a light-emitting diode is the product of the sub-efficiencies associated with an ensemble of different loss channels. Sub-efficiencies directly capture the effect of technology breakthroughs in device design and manufacturing process improvements. They thus provide the best metric to quantify technology breakthroughs.

\clearpage

\section{METHODS}
\label{sec:methods}

To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed to cover the entire manufacturing process of GaN-on-sapphire based phosphor converted low-to-mid power light-emitting diode packages of different chip architectures. We considered classical p-side-up lateral current spreading devices, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture. The model was constructed for the years 2003, 2012 and 2020. It was populated with equipment data from European and North American firms, selected for a virtual North American manufacturing location. Process specific step parameters were derived from scientific literature, company publications, archived product catalogs and patent literature. Details of the manufacturing process, as well as changes in the same were gathered from detailed patent analysis, augmented by interviews with experts from industry and academia. For 2012, data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland. The model includes both the wafer treatment process as well as the packaging process. It is important to note that the aim of the model is not to faithfully represent real world manufacturing conditions in Asia based plants, but rather to show the effect of single technological changes in the manufacturing process on total cost. This means that while the model does consider overhead costs and depreciation, it does so only to XXXX

Also to diagreggate contribution of single changes
To dis-aggregate the contribution of changes in manufacturing process steps to total manufacturing cost, we used an approach introduced by Kavlak et al. in a cost model with similar process step resolution for solar photovoltaics. It is based on the logarithmic derivative of the total differential of the cost function.

Equation here

To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture and types of down conversion phosphors used. We systematically identified the improvements related to each device sub-efficiency, gathered associated efficiency data from literature or computed the respective values from raw data.
For instance, in the case of spectral efficiency and the performance related to consumer experience metrics, we identified the prevalent red and yellow phosphors used during the period of interest. We then computed luminous efficacy of radiation as well as the color rendering index and color temperature from the spectral data. The development history of each phosphor was researched to determine if it had been developed specifically for use in light emitting devices, or if it was instead the result of a knowledge spillover.

figure: chip architectures
\begin{figure} [ht]
    \begin{center}
        \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
    \end{center}
    \caption{Cutaway side views of the evolution of chip architectures for classical chip designs (lateral current spreading). Note that dimensions are not to scale and smaller features are greatly exaggerated for visibility. Years correspond to earliest identified patent priority date. Dashed boxes indicate chip designs not brought to large scale production.}
    \label{fig:chip_arch}
\end{figure}

\begin{figure} [ht]
    \begin{center}
        \includegraphics[width=0.85\textwidth]{SPIE/article/droop_lumileds.pdf}
    \end{center}
    \caption{Luminous intensity of four different \textit{Lumileds} high-power light-emitting diodes normalized to the value at a test current of $A_{test}=350$mA. The black curves describe the real measured intensity, the orange curves describe the estimated ideal intensity. Droop is the difference between these curves at the test current. Sources for real curves: XXX}
    \label{fig:chip_arch}
\end{figure}

\section{PERFORMANCE IMPROVEMENTS}

\begin{figure} [ht]
    \begin{center}
        \includegraphics[width=0.85\textwidth]{SPIE/article/breakthroughs_efficiency.pdf}
    \end{center}
    \caption{Impact of technology breakthroughs and manufacturing process improvements on historical improvements in sub-efficiencies of phosphor-converted warm white light-emitting diodes with test currents of at least $I_\text{test}=350$mA. The overall lamp efficiency $\eta_L$ is displayed as the rightmost column. This figure takes as inputs the state-of-the-art sub-efficiencies discussed in section \ref{sec:methods}. Horizontal colored bars give state-of-the-art sub-efficiencies for five years: \textcolor{blue}{1997}, \textcolor{teal}{2002}, \textcolor{orange}{2010}, \textcolor{magenta}{2016} and \textcolor{red}{2020}. Colored annotation "N/A" indicates the sub-efficiency of the corresponding year cannot be computed for the following reasons: V$_\text{f}$E, Droop: depend on current, which was below 350mA at the time; CE, SE: warm white spectrum LEDs not available at the time. Physical limits are indicated by black horizontal bars. The possible range for the physical limit of V$_\text{f}$E exceeds 100\% and depends on electrical device parameters, which are discussed in \cite{david2016electrical}. The range is given by an upward pointing black arrow. Color of vertical bars indicates improvement through either technology spillovers or other improvements (technology breakthroughs or process learning). Spillovers are annotated and listed in an inset table. Efficiency acronyms: forward voltage efficiency (V$_\text{f}$E), internal quantum efficiency (IQE), light extraction efficiency (LEE), conversion efficiency (CE), spectral efficiency (SE), power conversion efficiency ($\eta_{PC}$). Spillover acronyms: patterned sapphire substrate (PSS), indium tin oxide (ITO). Sources: Data for sub efficiencies from figures XXX. Data for XXX.}
\end{figure}

\subsection{Physical Efficiency}

give two waterfall diagrams (2003,2020) and give one example
example: maybe droop with associated figure. is good example of process improvements

\subsection{Consumer Experience}

List phosphors, give a short overview but do not include figures
show cct spectrum with lines of YAG and 258+ phosphors

\section{MANUFACTURING COST}

show two waterfall diagrams (2003 only). mention work in progress

\acknowledgments % equivalent to \section*{ACKNOWLEDGMENTS}

The main author gratefully acknowledged funding by the \textit{Swiss Study Foundation} of Zurich, Switzerland. The authors further gratefully acknowledge project funding by the \textit{Alfred P. Sloan Foundation} of New York, USA. Cooperation by \textit{Stiftung Warentest} of Berlin, Germany in granting access to their publication archive is gratefully acknowledged.

% References
%\bibliography{report} % bibliography data in report.bib
%\bibliographystyle{spiebib} % makes bibtex use spiebib.bst

\end{document}

### 2021-02-04 SPIE/article/main.tex
added table

@@
 %\documentclass[]{spie}  %>>> use for US letter paper
-\documentclass[a4paper]{spie}  %>>> use this instead for A4 paper
+\documentclass[a4paper,nocompress]{spie}  %>>> use this instead for A4 paper
 %\documentclass[nocompress]{spie}  %>>> to avoid compression of citations
@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage{tabularx}
+\setlength{\extrarowheight}{2pt}
 
@@
 
-Why did we choose the metrics we did?
-What are the limitations of eg. Haitz's Law, Cost per Lamp, CRI, etc.
+However, the total flux per die is
 
@@
 They thus provide the best metric to quantify technology breakthroughs.
+
+        It describes the electrical and optical losses within the device, as well as the conversion losses in the phosphor layer.
+It further considers the mismatch between the device spectrum and an optimal spectrum, accounting for the wavelength dependent sensitivity of the human eye.
+
+    \begin{table}[h!]
+        \caption{List of the device sub-efficiencies used in our methodology.
+We follow the definitions used by previous authors, such as Tsao et al.
+\cite{tsao2010solid} and Pattison et al.
+\cite{pattison2017solid}.
+The historical development of the sub-efficiencies is displayed in figure \ref{fig:efficiency}.
+*Also called "lamp efficiency" or "cumulative efficiency" by authors, such as Tsao et al.
+\cite{tsao2010solid}}
+        \bigskip
+        \centering
+    	\begin{tabularx}{\textwidth}{|l|l|l|X|}
+    		\hline
+    			\textit{Symbol} & \textit{Sub-Efficiency} & \textit{Loss-Channel} & \textit{Definition} \\
+    		\hline
+    		    $\eta_{V_f}$ & Forward Voltage Efficiency* & Ohmic Resistance & $\eta_{V_f} = E_{h\nu} / V_f $ \\
+    		\hline
+    		    $\eta_{LE}$ & Light-Extraction Efficiency & Re-absorption and Reflection & $\eta_{LE}= P_{out} / P_{in} $ \\
+    		\hline
+    		    $\eta_{IQ}$ & Internal Quantum Efficiency & Non-radiative Recombinations & $\eta_{EQ} = \eta_{IQ} \times \eta_{LE}$ \\
+    		\hline
+    		    $\eta_{Droop}$ & (Electrical) Droop & Non-radiative Recombinations & $\eta_{Droop} = 1 - \eta_{IQE} / \eta_{IQE}(A \rightarrow 0) $ \\
+    		\hline
+    		    $\eta_C$ & Conversion Efficiency & Stokes Loss, Absorption, etc.
+& $\eta_{C} = E_{\textcolor{blue}{B}} / \sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i$ \\
+    		\hline
+    		    $\eta_{S}$ & Spectral Efficiency & Eye Sensitivity & $\eta_{S} = K / K_{max}(CRI,CCT)$ \\
+    		\hline
+    		    $\eta_L$ & Lamp Efficiency & N/A (Cumulative) & $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$ \\
+            \hline
+                \multicolumn{4}{|l|}{$\!\begin{aligned}
+                    E_{h\nu} &\dots \text{photon energy} \\
+                    V_f &\dots \text{forward voltage} \\
+                    A &\dots \text{electrical current} \\
+                    E_{B,\dots,G} &\dots \text{optical energy of monochromatic light (blue, red, orange, yellow, green)} \\
+                    K &\dots \text{luminous efficacy of radiation} \\
+                    CRI &\dots \text{color rendering index}, \ CCT \dots \text{color temperature} \\
+                \end{aligned}$} \\
+            \hline
+    	\end{tabularx}
+    \end{table}
 
@@
 Data for XXX.}
+    \label{fig:efficiency}
 \end{figure}
@@
 % References
-%\bibliography{report} % bibliography data in report.bib
-%\bibliographystyle{spiebib} % makes bibtex use spiebib.bst
+\bibliography{report} % bibliography data in report.bib
+\bibliographystyle{spiebib} % makes bibtex use spiebib.bst
 

### 2021-02-05 SPIE/article/main.tex
added sources to figures, added text to metrics section

@@
 
-However, the total flux per die is
+Data is readily available for metrics describing overall device performance.
+Luminous efficacy of devices as well as total flux per device can be extracted from datasheets, together with electrical device parameters.
+However,
+
+For instance, a highly cited and frequently updated metric is the total flux per package \cite{Liu2009}\cite{haitz2011solid}\cite{cho2017white}\cite{Fontoynont2018}.
+In combination with the cost per total flux, this visualization is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
+\cite{haitz1999case}.
+However, while these metrics are often used to showcase technological progress in light-emitting diode design and manufacturing, they have limited significance XXX.
+Firstly, it is not desirable in all applications to increase the total flux per device beyond a certain point.
+Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
+Secondly, even if one were to disregard these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if it was referring to single chips, instead of multi-chip paackages.
+Figure \ref{fig:haitz} shows an updated and expanded overview of best performing devices, both at the chip and package level, inspired by \textit{"Haitz's Law"}.
+It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
+While \textit{"Haitz's Law"} remains a popular visualization related to the progress in light-emitting diode design and manufacturing, care must be taken to consider its limitations.
 
@@
     \end{center}
-    \caption{Historical increase in flux for white light-emitting diode chips and packages, inspired by \textit{"Haitz's Law"}.
+    \caption{Historical increase in flux for highest performing white light-emitting diode \underline{chips and (multi-chip) packages}, inspired by \textit{"Haitz's Law"}.
 Note that the increase in flux per single chip is not increasing as starkly as the flux for multi-chip packages.
-Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.}
+Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
+Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, grey colored datapoints corresponding to the right ordinate.
+Sources: Data from Weinold et al.
+\cite{weinold2020technology}}
+    \label{fig:haitz}
 \end{figure}
 
-Mention previous work by Tsao et al.
-Mention limitations: Not done for 2003 (early devices) and 2020 (most recent devices)
+Overall efficiency is cumulative from many sub-efficiencies caused by different physical effects \cite{schubert2018light}
+In order to quantify the effect of single technological improvements on overall device performance, one must instead consider the metrics directly describing the various physical loss channels.
+Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
+The cumulative total lamp efficiency, defined as the product of all sub-efficiencies thus $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$ XXX
 
@@
 They thus provide the best metric to quantify technology breakthroughs.
+
+Flicker \cite{weinold2020long}
 
@@
     	\end{tabularx}
+    	\label{tab:eff}
     \end{table}
@@
 
-Equation here
+The following discussion follows an approach developed by Kavlak et al. \cite{kavlak2018evaluating}. For the detailed derivation, we refer to this publication.
 
@@
 Years correspond to earliest identified patent priority date.
-Dashed boxes indicate chip designs not brought to large scale production.}
+Dashed boxes indicate chip designs not brought to large scale production.
+Adapted from patents \cite{nagahama2013nitride}\cite{tanaka2010semiconductor}\cite{wierer2006photonic}}
     \label{fig:chip_arch}
@@
 Droop is the difference between these curves at the test current.
-Sources for real curves: XXX}
+Current-Intensity data extracted from device datasheets \cite{datasheet_lumileds_lux1}\cite{datasheet_lumileds_rebel}\cite{datasheet_lumileds_rebplus}\cite{lumi2019data}}
     \label{fig:chip_arch}
@@
 Spillover acronyms: patterned sapphire substrate (PSS), indium tin oxide (ITO).
-Sources: Data for sub efficiencies from figures XXX.
-Data for XXX.}
+Sources: Data for sub efficiencies from \cite{weinold2020technology}}
     \label{fig:efficiency}
@@
 
+\clearpage
 \acknowledgments % equivalent to \section*{ACKNOWLEDGMENTS}
@@
 Sloan Foundation} of New York, USA.
-Cooperation by \textit{Stiftung Warentest} of Berlin, Germany in granting access to their publication archive is gratefully acknowledged.
 

### 2021-02-05 SPIE/article/main.tex
added additional figure

@@
 \setlength{\extrarowheight}{2pt}
+\usepackage{chemformula}
 
@@
 The historical development of the sub-efficiencies is displayed in figure \ref{fig:efficiency}.
-*Also called "lamp efficiency" or "cumulative efficiency" by authors, such as Tsao et al.
+\\ *Also called "lamp efficiency" or "cumulative efficiency" by authors, such as Tsao et al.
 \cite{tsao2010solid}}
@@
 
+\begin{figure} [ht]
+    \begin{center}
+        \includegraphics[width=\textwidth]{SPIE/article/cct_spectrum.pdf}
+    \end{center}
+    \caption{Color temperature for temperature values between 1,000K and 12,000K.
+Vertical lines indicate the color temperature of blue light emitting diodes used in combinations with phosphors.
+Two data points are shown: The 1997 \textit{Nichia} \ch{YAG:Ce^{3+}} (YAG) phosphor at 5800K and the 2004 \textit{Lumileds} \ch{Sr_2Si_5N_8:Eu^{2+}} ("258") phosphor at below 4000K.
+Color data generated using the \textit{luxpy} package for \textit{Python}.
+CCT for phosphors computed using the \textit{color science} package for \textit{Python}, from spectral data in Bando et al.
+\cite{bando1998development} and Mueller-Mach et al.\cite{MuellerMach2005}.}
+    \label{fig:cct}
+\end{figure}
+
 \clearpage
@@
 
-figure: chip architectures
 \begin{figure} [ht]
@@
 Spillover acronyms: patterned sapphire substrate (PSS), indium tin oxide (ITO).
-Sources: Data for sub efficiencies from \cite{weinold2020technology}}
+Data for historical improvements of sub efficiencies from Weinold et al.
+\cite{weinold2020technology}}
     \label{fig:efficiency}
@@
 
-The main author gratefully acknowledged funding by the \textit{Swiss Study Foundation} of Zurich, Switzerland.
+The main author gratefully acknowledges funding by the \textit{Swiss Study Foundation} of Zurich, Switzerland.
 The authors further gratefully acknowledge project funding by the \textit{Alfred P.

### 2021-02-05 SPIE/article/main.tex
corrected citations, added equation to methods section

@@
 
-For instance, a highly cited and frequently updated metric is the total flux per package \cite{Liu2009}\cite{haitz2011solid}\cite{cho2017white}\cite{Fontoynont2018}.
+For instance, a highly cited and frequently updated metric is the total flux per package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
 In combination with the cost per total flux, this visualization is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
@@
 They thus provide the best metric to quantify technology breakthroughs.
-
-Flicker \cite{weinold2020long}
 
@@
 
+consumer experience important
+metrics relevant to consumer experience
+some are contested
+consumer experience metrics have been trated in my thesis \cite{weinold2020technology}.
+flicker is out of scope and has been treated in a separate publication
+Flicker \cite{weinold2020long}
+
 \begin{figure} [ht]
@@
 Color data generated using the \textit{luxpy} package for \textit{Python}.
-CCT for phosphors computed using the \textit{color science} package for \textit{Python}, from spectral data in Bando et al.
-\cite{bando1998development} and Mueller-Mach et al.\cite{MuellerMach2005}.}
+CCT for phosphors computed using the \textit{color science} package for \textit{Python}, from spectral data in publications  \cite{bando1998development,MuellerMach2005}.}
     \label{fig:cct}
@@
 \clearpage
-
 \section{METHODS}
@@
 
+    \begin{align}
+        C_i(\vec{r}) &= C_i^0 \prod_w g_{iw}(r_w) \\
+        \Delta C_x &= \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t \\
+        \Delta C_z (t_1,t_2) &\approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
+    \end{align}
+
+    where
+
 To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture and types of down conversion phosphors used.
@@
 Dashed boxes indicate chip designs not brought to large scale production.
-Adapted from patents \cite{nagahama2013nitride}\cite{tanaka2010semiconductor}\cite{wierer2006photonic}}
+Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
     \label{fig:chip_arch}
@@
 Droop is the difference between these curves at the test current.
-Current-Intensity data extracted from device datasheets \cite{datasheet_lumileds_lux1}\cite{datasheet_lumileds_rebel}\cite{datasheet_lumileds_rebplus}\cite{lumi2019data}}
+Current-Intensity data extracted from device datasheets \cite{datasheet_lumileds_lux1,datasheet_lumileds_rebel,datasheet_lumileds_rebplus,lumi2019data}}
     \label{fig:chip_arch}

### 2021-02-05 SPIE/article/main.tex
finished equations on dis-aggregation

@@
 
-Also to diagreggate contribution of single changes
-To dis-aggregate the contribution of changes in manufacturing process steps to total manufacturing cost, we used an approach introduced by Kavlak et al.
-in a cost model with similar process step resolution for solar photovoltaics.
+    To dis-aggregate the contribution of changes in single variables on total performance in both the device sub-efficiencies and the manufacturing cost, we used an approach introduced by Kavlak et al.
+\cite{kavlak2018evaluating}.
 It is based on the logarithmic derivative of the total differential of the cost function.
-
-The following discussion follows an approach developed by Kavlak et al. \cite{kavlak2018evaluating}. For the detailed derivation, we refer to this publication.
+We can write the cost function $C$ of the manufacturing process as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$, where $g_{iw}(r_w)$ gives the dependence of $i$th cost component on the $w$th variable.
+The contribution of changes in the cost model $\Delta C_z$ to total manufacturing cost $C$ can now be approximated.
+For the detailed derivation, we refer to the supplementary material of the original publication.
 
 	\begin{align}
-        C_i(\vec{r}) &= C_i^0 \prod_w g_{iw}(r_w) \\
+        C(\vec{r}) &= C(r_1,r_2, \dots) = \sum_i C_i = \sum_i C_i^0 \prod_w g_{iw}(r_w) \\
         \Delta C_x &= \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t \\
@@
     \end{align}
-
-    where
 

### 2021-02-06 SPIE/article/main.tex
added final figure

@@
 
+    \subsection{Manufacturing Cost Model}
 To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed to cover the entire manufacturing process of GaN-on-sapphire based phosphor converted low-to-mid power light-emitting diode packages of different chip architectures.
@@
 This means that while the model does consider overhead costs and depreciation, it does so only to XXXX
+
+\begin{figure} [ht]
+    \begin{center}
+        \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
+    \end{center}
+    \caption{Cutaway side views of the evolution of chip architectures for classical chip designs (lateral current spreading).
+Note that dimensions are not to scale and smaller features are greatly exaggerated for visibility.
+Years correspond to earliest identified patent priority date.
+Dashed boxes indicate chip designs not brought to large scale production.
+Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
+    \label{fig:chip_arch}
+\end{figure}
 
@@
 
+        \subsection{Literature Review and Performance Computation}
+
 To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture and types of down conversion phosphors used.
@@
 The development history of each phosphor was researched to determine if it had been developed specifically for use in light emitting devices, or if it was instead the result of a knowledge spillover.
-
-\begin{figure} [ht]
-    \begin{center}
-        \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
-    \end{center}
-    \caption{Cutaway side views of the evolution of chip architectures for classical chip designs (lateral current spreading).
-Note that dimensions are not to scale and smaller features are greatly exaggerated for visibility.
-Years correspond to earliest identified patent priority date.
-Dashed boxes indicate chip designs not brought to large scale production.
-Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
-    \label{fig:chip_arch}
-\end{figure}
 
@@
 \section{PERFORMANCE IMPROVEMENTS}
+
+\subsection{Physical Efficiency}
+
+give two waterfall diagrams (2003,2020) and give one example
+example: maybe droop with associated figure. is good example of process improvements
+
+\subsection{Consumer Experience}
+
+List phosphors, give a short overview but do not include figures
+show cct spectrum with lines of YAG and 258+ phosphors
+
+\section{MANUFACTURING COST REDUCTIONS}
+
+\clearpage
 
@@
 
-\subsection{Physical Efficiency}
-
-give two waterfall diagrams (2003,2020) and give one example
-example: maybe droop with associated figure. is good example of process improvements
-
-\subsection{Consumer Experience}
-
-List phosphors, give a short overview but do not include figures
-show cct spectrum with lines of YAG and 258+ phosphors
-
-\section{MANUFACTURING COST}
-
-show two waterfall diagrams (2003 only). mention work in progress
+\begin{figure} [ht]
+    \begin{center}
+        \includegraphics[width=0.85\textwidth]{SPIE/article/costmodel_calibration.pdf}
+    \end{center}
+    \caption{Modeled manufacturing cost for the specified architecture of light-emitting diode packages split by manufacturing step category, following the XXX in US DoE reports \cite{doe2016solid}.
+The top graph shows the data from \textit{LEDCOM} model and  round table discussions published by the United States Department of Energy.
+The bottom graph shows the results of the cost model developed in this thesis.
+Die per wafer calculated from de Vries \cite{deVries2005}}
+    \label{fig:cost}
+\end{figure}
 

### 2021-02-06 SPIE/article/main.tex
added text to metrics and methods sections

--- 2021-02-05 SPIE/article/main.tex
@@
 
-consumer experience important
-metrics relevant to consumer experience
-some are contested
-consumer experience metrics have been trated in my thesis \cite{weinold2020technology}.
-flicker is out of scope and has been treated in a separate publication
-Flicker \cite{weinold2020long}
+    Performance improvements in metrics related to consumer experience have also played a role in the adoption of light-emitting diode based luminaires \cite{cowan2011understanding}.
+Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
+For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources, shown in figure \ref{fig:cct}.
+A thorough treatment of consumer experience metrics is given by Weinold et al.
+\cite{weinold2020technology}.
+Flicker was not considered in this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
+We instead refer to a recent publication by Weinold \cite{weinold2020long}.
 
@@
     \caption{Color temperature for temperature values between 1,000K and 12,000K.
-Vertical lines indicate the color temperature of blue light emitting diodes used in combinations with phosphors.
-Two data points are shown: The 1997 \textit{Nichia} \ch{YAG:Ce^{3+}} (YAG) phosphor at 5800K and the 2004 \textit{Lumileds} \ch{Sr_2Si_5N_8:Eu^{2+}} ("258") phosphor at below 4000K.
+Vertical lines indicate the color temperature of blue light-emitting diodes used in combinations with phosphors.
+Arrows indicate flexibility through phosphor composition variation.
+Two data points are shown: The 1997 \textit{Nichia} \ch{YAG:Ce^{3+}} (YAG) phosphor at 5800K and the 2004 \textit{Lumileds} \ch{Sr_2Si_5N_8:Eu^{2+}} ("258") phosphor at below 4000K (tuneable).
 Color data generated using the \textit{luxpy} package for \textit{Python}.
@@
 
-To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed to cover the entire manufacturing process of GaN-on-sapphire based phosphor converted low-to-mid power light-emitting diode packages of different chip architectures.
-We considered classical p-side-up lateral current spreading devices, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture.
+    \subsection{Manufacturing Cost Model}
+
+        To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed.
+It covers the entire manufacturing process of GaN-on-sapphire-based phosphor-converted low-to-mid power light-emitting diode packages of different chip architectures.
+We considered the classical p-side-up lateral current spreading architecture, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture.
+An excerpt of the device architectures considered is shown in figure \ref{fig:chip_arch}.
 The model was constructed for the years 2003, 2012 and 2020.
@@
 Process specific step parameters were derived from scientific literature, company publications, archived product catalogs and patent literature.
-Details of the manufacturing process, as well as changes in the same were gathered from detailed patent analysis, augmented by interviews with experts from industry and academia.
-For 2012, data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland.
+Where necessary, this data was augmented using information gathered during semi-structured interviews with experts from industry and academia.
+Details of the manufacturing process, as well as changes in the same were gathered from detailed patent analysis, augmented by semi-structured interviews with experts from industry and academia.
+For 2012, data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}.
 The model includes both the wafer treatment process as well as the packaging process.
-It is important to note that the aim of the model is not to faithfully represent real world manufacturing conditions in Asia based plants, but rather to show the effect of single technological changes in the manufacturing process on total cost.
-This means that while the model does consider overhead costs and depreciation, it does so only to XXXX
-
-    To dis-aggregate the contribution of changes in single variables on total performance in both the device sub-efficiencies and the manufacturing cost, we used an approach introduced by Kavlak et al.
-\cite{kavlak2018evaluating}.
-It is based on the logarithmic derivative of the total differential of the cost function.
-We can write the cost function $C$ of the manufacturing process as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$, where $g_{iw}(r_w)$ gives the dependence of $i$th cost component on the $w$th variable.
-The contribution of changes in the cost model $\Delta C_z$ to total manufacturing cost $C$ can now be approximated.
-For the detailed derivation, we refer to the supplementary material of the original publication.
-
-	\begin{align}
-        C(\vec{r}) &= C(r_1,r_2, \dots) = \sum_i C_i = \sum_i C_i^0 \prod_w g_{iw}(r_w) \\
-        \Delta C_x &= \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t \\
-        \Delta C_z (t_1,t_2) &\approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
-    \end{align}
-
-To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture and types of down conversion phosphors used.
-We systematically identified the improvements related to each device sub-efficiency, gathered associated efficiency data from literature or computed the respective values from raw data.
-For instance, in the case of spectral efficiency and the performance related to consumer experience metrics, we identified the prevalent red and yellow phosphors used during the period of interest.
-We then computed luminous efficacy of radiation as well as the color rendering index and color temperature from the spectral data.
-The development history of each phosphor was researched to determine if it had been developed specifically for use in light emitting devices, or if it was instead the result of a knowledge spillover.
+While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach.
+The main aim of the model is not to faithfully represent real world manufacturing conditions in Asian manufacturing locations, but rather to show the effect of single technological changes in the manufacturing process on total cost.
+For instance, XXX
 
@@
 
+        To dis-aggregate the contribution of changes in single variables on total manufacturing cost, we used an approach introduced by Kavlak et al.
+\cite{kavlak2018evaluating}.
+It is based on the logarithmic derivative of the total differential of the cost function.
+We can write the cost function $C$ of the manufacturing process as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$, where $g_{iw}(r_w)$ gives the dependence of $i$th cost component on the $w$th variable.
+The contribution of changes in the cost model $\Delta C_z$ to total manufacturing cost $C$ can now be approximated.
+For the detailed derivation, we refer to the supplementary material of the original publication.
+
+    	\begin{align}
+            C(\vec{r}) &= C(r_1,r_2, \dots) = \sum_i C_i = \sum_i C_i^0 \prod_w g_{iw}(r_w) \\
+            \Delta C_x &= \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t \\
+            \Delta C_z (t_1,t_2) &\approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
+        \end{align}
+
+    \subsection{Literature Review and Performance Computation}
+
+        To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture and types of down conversion phosphors used.
+We systematically identified the improvements related to each device sub-efficiency, gathered associated efficiency data from literature or computed the respective values from raw data.
+For instance, to gather historical data on the improvement of losses associated with electrical droop, data was extracted from device datasheets, as shown in figure \ref{fig:droop}.
+This was augmented by industry forecasts from company publications \cite{osram2014osram}.
+Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gather a complete picture of the historical development of light-emitting diode efficiency.
+
         \begin{figure} [ht]
@@
 The black curves describe the real measured intensity, the orange curves describe the estimated ideal intensity.
-Droop is the difference between these curves at the test current.
+Droop $D$, as defined in table \ref{tab:eff}, is the difference between these curves at the test current.
 Current-Intensity data extracted from device datasheets \cite{datasheet_lumileds_lux1,datasheet_lumileds_rebel,datasheet_lumileds_rebplus,lumi2019data}}
-    \label{fig:chip_arch}
+            \label{fig:droop}
         \end{figure}
@@
 
-\subsection{Physical Efficiency}
-
-give two waterfall diagrams (2003,2020) and give one example
-example: maybe droop with associated figure. is good example of process improvements
-
-\subsection{Consumer Experience}
-
-List phosphors, give a short overview but do not include figures
-show cct spectrum with lines of YAG and 258+ phosphors
-
-\section{MANUFACTURING COST}
-
-show two waterfall diagrams (2003 only). mention work in progress
+\begin{figure} [ht]
+    \begin{center}
+        \includegraphics[width=0.85\textwidth]{SPIE/article/costmodel_calibration.pdf}
+    \end{center}
+    \caption{Modeled manufacturing cost for the specified architecture of light-emitting diode packages split by manufacturing step category, following the categories used in the US DoE reports \cite{doe2016solid}.
+The inset plot shows the number of die per wafer corresponding to wafers of different sizes, associated street width and exclusion zone.
+Calculated using approximations introduced by de Vries \cite{deVries2005}.}
+    \label{fig:cost}
+\end{figure}
 

### 2021-02-09 SPIE/article/main.tex
corrected error in Haitz law

@@
 \label{sec:intro}  % \label{} allows reference to this section
-
-\clearpage
-\section{METRICS}
-
-Data is readily available for metrics describing overall device performance.
-Luminous efficacy of devices as well as total flux per device can be extracted from datasheets, together with electrical device parameters.
-However,
-
-For instance, a highly cited and frequently updated metric is the total flux per package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
-In combination with the cost per total flux, this visualization is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
-\cite{haitz1999case}.
-However, while these metrics are often used to showcase technological progress in light-emitting diode design and manufacturing, they have limited significance XXX.
-Firstly, it is not desirable in all applications to increase the total flux per device beyond a certain point.
-Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
-Secondly, even if one were to disregard these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if it was referring to single chips, instead of multi-chip paackages.
-Figure \ref{fig:haitz} shows an updated and expanded overview of best performing devices, both at the chip and package level, inspired by \textit{"Haitz's Law"}.
-It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
-While \textit{"Haitz's Law"} remains a popular visualization related to the progress in light-emitting diode design and manufacturing, care must be taken to consider its limitations.
-
-\begin{figure} [ht]
-    \begin{center}
-        \includegraphics[width=\textwidth]{haitz_law_white.pdf}
-    \end{center}
-    \caption{Historical increase in flux for highest performing white light-emitting diode \underline{chips and (multi-chip) packages}, inspired by \textit{"Haitz's Law"}.
-Note that the increase in flux per single chip is not increasing as starkly as the flux for multi-chip packages.
-Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
-Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, grey colored datapoints corresponding to the right ordinate.
-Sources: Data from Weinold et al.
-\cite{weinold2020technology}}
-    \label{fig:haitz}
-\end{figure}
-
-Overall efficiency is cumulative from many sub-efficiencies caused by different physical effects \cite{schubert2018light}
-In order to quantify the effect of single technological improvements on overall device performance, one must instead consider the metrics directly describing the various physical loss channels.
-Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
-The cumulative total lamp efficiency, defined as the product of all sub-efficiencies thus $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$ XXX
-
-The overall efficiency of a light-emitting diode is the product of the sub-efficiencies associated with an ensemble of different loss channels.
-Sub-efficiencies directly capture the effect of technology breakthroughs in device design and manufacturing process improvements.
-They thus provide the best metric to quantify technology breakthroughs.
-
-        It describes the electrical and optical losses within the device, as well as the conversion losses in the phosphor layer.
-It further considers the mismatch between the device spectrum and an optimal spectrum, accounting for the wavelength dependent sensitivity of the human eye.
-
-    \begin{table}[h!]
-        \caption{List of the device sub-efficiencies used in our methodology.
-We follow the definitions used by previous authors, such as Tsao et al.
-\cite{tsao2010solid} and Pattison et al.
-\cite{pattison2017solid}.
-The historical development of the sub-efficiencies is displayed in figure \ref{fig:efficiency}.
-\\ *Also called "lamp efficiency" or "cumulative efficiency" by authors, such as Tsao et al.
-\cite{tsao2010solid}}
-        \bigskip
-        \centering
-    	\begin{tabularx}{\textwidth}{|l|l|l|X|}
-    		\hline
-    			\textit{Symbol} & \textit{Sub-Efficiency} & \textit{Loss-Channel} & \textit{Definition} \\
-    		\hline
-    		    $\eta_{V_f}$ & Forward Voltage Efficiency* & Ohmic Resistance & $\eta_{V_f} = E_{h\nu} / V_f $ \\
-    		\hline
-    		    $\eta_{LE}$ & Light-Extraction Efficiency & Re-absorption and Reflection & $\eta_{LE}= P_{out} / P_{in} $ \\
-    		\hline
-    		    $\eta_{IQ}$ & Internal Quantum Efficiency & Non-radiative Recombinations & $\eta_{EQ} = \eta_{IQ} \times \eta_{LE}$ \\
-    		\hline
-    		    $\eta_{Droop}$ & (Electrical) Droop & Non-radiative Recombinations & $\eta_{Droop} = 1 - \eta_{IQE} / \eta_{IQE}(A \rightarrow 0) $ \\
-    		\hline
-    		    $\eta_C$ & Conversion Efficiency & Stokes Loss, Absorption, etc.
-& $\eta_{C} = E_{\textcolor{blue}{B}} / \sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i$ \\
-    		\hline
-    		    $\eta_{S}$ & Spectral Efficiency & Eye Sensitivity & $\eta_{S} = K / K_{max}(CRI,CCT)$ \\
-    		\hline
-    		    $\eta_L$ & Lamp Efficiency & N/A (Cumulative) & $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$ \\
-            \hline
-                \multicolumn{4}{|l|}{$\!\begin{aligned}
-                    E_{h\nu} &\dots \text{photon energy} \\
-                    V_f &\dots \text{forward voltage} \\
-                    A &\dots \text{electrical current} \\
-                    E_{B,\dots,G} &\dots \text{optical energy of monochromatic light (blue, red, orange, yellow, green)} \\
-                    K &\dots \text{luminous efficacy of radiation} \\
-                    CRI &\dots \text{color rendering index}, \ CCT \dots \text{color temperature} \\
-                \end{aligned}$} \\
-            \hline
-    	\end{tabularx}
-    	\label{tab:eff}
-    \end{table}
-
-    Performance improvements in metrics related to consumer experience have also played a role in the adoption of light-emitting diode based luminaires \cite{cowan2011understanding}.
-Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
-For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources, shown in figure \ref{fig:cct}.
-A thorough treatment of consumer experience metrics is given by Weinold et al.
-\cite{weinold2020technology}.
-Flicker was not considered in this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
-We instead refer to a recent publication by Weinold \cite{weinold2020long}.
-
-\begin{figure} [ht]
-    \begin{center}
-        \includegraphics[width=\textwidth]{SPIE/article/cct_spectrum.pdf}
-    \end{center}
-    \caption{Color temperature for temperature values between 1,000K and 12,000K.
-Vertical lines indicate the color temperature of blue light-emitting diodes used in combinations with phosphors.
-Arrows indicate flexibility through phosphor composition variation.
-Two data points are shown: The 1997 \textit{Nichia} \ch{YAG:Ce^{3+}} (YAG) phosphor at 5800K and the 2004 \textit{Lumileds} \ch{Sr_2Si_5N_8:Eu^{2+}} ("258") phosphor at below 4000K (tuneable).
-Color data generated using the \textit{luxpy} package for \textit{Python}.
-CCT for phosphors computed using the \textit{color science} package for \textit{Python}, from spectral data in publications  \cite{bando1998development,MuellerMach2005}.}
-    \label{fig:cct}
-\end{figure}
 
@@
 
+\section{METRICS}
+
+    Once technological breakthroughs and spillovers have been identified, their effect must be quantified.
+However, in solid-state lighting device performance does not only have a purely physical and economic dimension.
+We must consider human factors in two dimensions: First, we must consider the wavelength dependent sensitivity of the human eyes.
+This allows us to us to extend the physical performance dimension to an optometric performance dimension.
+Second, we must consider the consumer experience dimension.
+Related to the spectrum of the light source, this includes the perceived temperature of white light or its ability to faithfully render colors.
+Related to the temporal modulation of the light source, this includes unwanted flickering.
+Only when considering all three dimensions can the effect of a technological breakthrough be fully quantified.
+
+    Metrics for light-emitting diodes must fulfill a number of requirements.
+First, the combined number of metrics must describe all dimensions of product performance introduced above.
+The development of a single compound metric, considering all dimensions, would be beyond the scope of this thesis.
+They must be widely accepted in industry.
+This ensures that they are reported in scientific literature and device datasheets.
+Metrics which have been proposed in literature but not reported by device manufacturers cannot be used to compare device performance.
+If neither is the case, they must at least allow for their computation from metrics which are available to us.
+They must further have the ability to directly describe the effect of breakthroughs on device performance.
+Otherwise, they must take as inputs metrics which directly describe the effect of breakthroughs on device performance.
+In this case, they must not require additional information which can not be fully quantified or is unavailable to us.
+
+    Data is readily available for metrics describing overall device performance.
+Luminous efficacy of devices as well as total flux per device can be extracted from datasheets, together with electrical device parameters.
+However,
+
+    For instance, a highly cited and frequently updated metric is the total luminous flux per light-emitting diode package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
+In combination with the cost per total flux, this visualization is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
+\cite{haitz1999case}.
+However, while these metrics are often used to showcase technological progress in light-emitting diode design and manufacturing, care must be taken to consider its limitations.
+
+    Firstly, the metric today retains only limited significance as a proxy for technological development in light-emitting diodes.
+This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
+Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
+Secondly, disregard these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
+Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
+
+    Figure \ref{fig:haitz} shows an updated and expanded overview of best performing devices, both at the chip and package level, inspired by \textit{"Haitz's Law"}.
+It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
+While \textit{"Haitz's Law"} remains a popular visualization related to the progress in light-emitting diode design and manufacturing, .
+
+    \begin{figure} [ht]
+        \begin{center}
+            \includegraphics[width=\textwidth]{haitz_law_white.pdf}
+        \end{center}
+        \caption{Historical increase in flux for highest performing white \underline{light-emitting diode chips and (multi-chip) packages}, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
+Note that flux per single chips has not increased as starkly as the flux for multi-chip packages.
+Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
+Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, grey colored datapoints corresponding to the right ordinate.
+Sources: Data from Weinold et al.
+\cite{weinold2020technology}}
+        \label{fig:haitz}
+    \end{figure}
+
+    Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency.
+This metric defined as the product of all device sub-efficiencies associated with an ensemble of different loss channels $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$.
+Sub-efficiencies directly capture the effect of technology breakthroughs in device design and manufacturing process improvements.
+They thus provide the best metric to quantify technology breakthroughs.
+Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
+
+    \begin{table}[h!]
+        \caption{List of the device sub-efficiencies used in our methodology.
+We follow the definitions used by previous authors, such as Tsao et al.
+\cite{tsao2010solid} and Pattison et al.
+\cite{pattison2017solid}.
+The historical development of the sub-efficiencies is displayed in figure \ref{fig:efficiency}.
+\\ *Also called "lamp efficiency" or "cumulative efficiency" by authors, such as Tsao et al.
+\cite{tsao2010solid}}
+        \bigskip
+        \centering
+    	\begin{tabularx}{\textwidth}{|l|l|l|X|}
+    		\hline
+    			\textit{Symbol} & \textit{Sub-Efficiency} & \textit{Loss-Channel} & \textit{Definition} \\
+    		\hline
+    		    $\eta_{V_f}$ & Forward Voltage Efficiency* & Ohmic Resistance & $\eta_{V_f} = E_{h\nu} / V_f $ \\
+    		\hline
+    		    $\eta_{LE}$ & Light-Extraction Efficiency & Re-absorption and Reflection & $\eta_{LE}= P_{out} / P_{in} $ \\
+    		\hline
+    		    $\eta_{IQ}$ & Internal Quantum Efficiency & Non-radiative Recombinations & $\eta_{EQ} = \eta_{IQ} \times \eta_{LE}$ \\
+    		\hline
+    		    $\eta_{Droop}$ & (Electrical) Droop & Non-radiative Recombinations & $\eta_{Droop} = 1 - \eta_{IQE} / \eta_{IQE}(A \rightarrow 0) $ \\
+    		\hline
+    		    $\eta_C$ & Conversion Efficiency & Stokes Loss, Absorption, etc.
+& $\eta_{C} = E_{\textcolor{blue}{B}} / \sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i$ \\
+    		\hline
+    		    $\eta_{S}$ & Spectral Efficiency & Eye Sensitivity & $\eta_{S} = K / K_{max}(CRI,CCT)$ \\
+    		\hline
+    		    $\eta_L$ & Lamp Efficiency & N/A (Cumulative) & $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$ \\
+            \hline
+                \multicolumn{4}{|l|}{$\!\begin{aligned}
+                    E_{h\nu} &\dots \text{photon energy} \\
+                    V_f &\dots \text{forward voltage} \\
+                    A &\dots \text{electrical current} \\
+                    E_{B,\dots,G} &\dots \text{optical energy of monochromatic light (blue, red, orange, yellow, green)} \\
+                    K &\dots \text{luminous efficacy of radiation} \\
+                    CRI &\dots \text{color rendering index}, \ CCT \dots \text{color temperature} \\
+                \end{aligned}$} \\
+            \hline
+    	\end{tabularx}
+    	\label{tab:eff}
+    \end{table}
+
+    Performance improvements in metrics related to consumer experience have also played a role in the adoption of light-emitting diode based luminaires \cite{cowan2011understanding}.
+Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
+For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources, shown in figure \ref{fig:cct}.
+A thorough treatment of consumer experience metrics is given by Weinold et al.
+\cite{weinold2020technology}.
+Flicker was not considered in this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
+We instead refer to a recent publication by Weinold \cite{weinold2020long}.
+
+\begin{figure} [ht]
+    \begin{center}
+        \includegraphics[width=\textwidth]{SPIE/article/cct_spectrum.pdf}
+    \end{center}
+    \caption{Color temperature for temperature values between 1,000K and 12,000K.
+Vertical lines indicate the color temperature of blue light-emitting diodes used in combinations with phosphors.
+Arrows indicate flexibility through phosphor composition variation.
+Two data points are shown: The 1997 \textit{Nichia} \ch{YAG:Ce^{3+}} (YAG) phosphor at 5800K and the 2004 \textit{Lumileds} \ch{Sr_2Si_5N_8:Eu^{2+}} ("258") phosphor at below 4000K (tuneable).
+Color data generated using the \textit{luxpy} package for \textit{Python}.
+CCT for phosphors computed using the \textit{color science} package for \textit{Python}, from spectral data in publications  \cite{bando1998development,MuellerMach2005}.}
+    \label{fig:cct}
+\end{figure}
+
 \section{PERFORMANCE IMPROVEMENTS}

### 2021-02-09 SPIE/article/main.tex
added text to sections

@@
 \section{INTRODUCTION}
-\label{sec:intro}  % \label{} allows reference to this section
+\label{sec:intro}
+
+mention: technology had important improvements in past 25 years
+mention anniversiary of technology
+mention adoption in EU and US
+mention commoditization
+lack of understanding of sources of improvements across 25 years
+mention works by tsao (and at least two others)
+
+We collected historical data on device performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
+We used this information to identify and quantify the principal sources of performance improvements in LED manufacturing.
+We found that in order to quantify the impact of single technological changes, it is necessary to analyse performance improvements the device sub-efficiency level.
+We further developed a bottom-up manufacturing cost model with process step resolution that captures improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale to analyse cost reductions and their sources.
+It covers progress from early manufacturing in 2003 to today.
+We found that larger wafer sizes have been largely responsible for cost reductions.
+We found that XXX well known XXX.
+We collected historical data on device performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
+We used this information to identify and quantify the principal sources of performance improvements in LED manufacturing.
+We found that in order to quantify the impact of single technological changes, it is necessary to analyse performance improvements the device sub-efficiency level.
+We further developed a bottom-up manufacturing cost model with process step resolution that captures improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale to analyse cost reductions and their sources.
+It covers progress from early manufacturing in 2003 to today.
+We found that larger wafer sizes have been largely responsible for cost reductions.
+We found that XXX well known XXX.
 
@@
 The main aim of the model is not to faithfully represent real world manufacturing conditions in Asian manufacturing locations, but rather to show the effect of single technological changes in the manufacturing process on total cost.
-For instance, XXX
 
@@
 
-        To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture and types of down conversion phosphors used.
-We systematically identified the improvements related to each device sub-efficiency, gathered associated efficiency data from literature or computed the respective values from raw data.
-For instance, to gather historical data on the improvement of losses associated with electrical droop, data was extracted from device datasheets, as shown in figure \ref{fig:droop}.
-This was augmented by industry forecasts from company publications \cite{osram2014osram}.
+        To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture, manufacturing processes and types of down conversion phosphors used.
+We systematically identified the architecture and manufacturing process improvements related to each physical loss channel, gathered associated efficiency data from literature or computed the respective values from raw data.
+
+        For instance, to gather historical data on the improvement of losses associated with electrical droop, device performance data was extracted from datasheets, as shown in figure \ref{fig:droop}.
+This data was cross-referenced with information on the different chip architectures and manufacturing processes used for each device.
+Data was then augmented by industry forecasts from company publications \cite{osram2014osram}.
 Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gather a complete picture of the historical development of light-emitting diode efficiency.
@@
 
-    Once technological breakthroughs and spillovers have been identified, their effect must be quantified.
-However, in solid-state lighting device performance does not only have a purely physical and economic dimension.
-We must consider human factors in two dimensions: First, we must consider the wavelength dependent sensitivity of the human eyes.
-This allows us to us to extend the physical performance dimension to an optometric performance dimension.
-Second, we must consider the consumer experience dimension.
-Related to the spectrum of the light source, this includes the perceived temperature of white light or its ability to faithfully render colors.
-Related to the temporal modulation of the light source, this includes unwanted flickering.
-Only when considering all three dimensions can the effect of a technological breakthrough be fully quantified.
+    Once technological breakthroughs and process improvements had been identified, their effect was quantified.
+The ensemble of metrics was required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
+They must further have the ability to directly describe the effect of single technological breakthroughs, device architectures and manufacturing process improvements on device performance.
 
-    Metrics for light-emitting diodes must fulfill a number of requirements.
-First, the combined number of metrics must describe all dimensions of product performance introduced above.
-The development of a single compound metric, considering all dimensions, would be beyond the scope of this thesis.
-They must be widely accepted in industry.
-This ensures that they are reported in scientific literature and device datasheets.
-Metrics which have been proposed in literature but not reported by device manufacturers cannot be used to compare device performance.
-If neither is the case, they must at least allow for their computation from metrics which are available to us.
-They must further have the ability to directly describe the effect of breakthroughs on device performance.
-Otherwise, they must take as inputs metrics which directly describe the effect of breakthroughs on device performance.
-In this case, they must not require additional information which can not be fully quantified or is unavailable to us.
-
-    Data is readily available for metrics describing overall device performance.
-Luminous efficacy of devices as well as total flux per device can be extracted from datasheets, together with electrical device parameters.
-However,
-
+    These requirements precluded the use of some established metrics.
 For instance, a highly cited and frequently updated metric is the total luminous flux per light-emitting diode package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
@@
 Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
-Secondly, disregard these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
+Secondly, disregarding these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
 Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
@@
 
+\clearpage
 \section{PERFORMANCE IMPROVEMENTS}
+
+    list performance improvements, mention physical limits
+    mention loss channels with process improvements vs channels with breakthroughs and spillovers
+    mention caveats in manufacturing cost decrease
+    mention most important factor in manufacturing cost decrease
 
@@
 \section{MANUFACTURING COST REDUCTIONS}
-
-\clearpage
 
@@
 
-\clearpage
 \acknowledgments % equivalent to \section*{ACKNOWLEDGMENTS}
@@
 
+\clearpage
 % References

### 2021-02-09 SPIE/article/main.tex
minor changes

@@
 \begin{abstract}
+
+    To understand the 
     We collected historical data on device performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
@@
 
-We collected historical data on device performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
-We used this information to identify and quantify the principal sources of performance improvements in LED manufacturing.
-We found that in order to quantify the impact of single technological changes, it is necessary to analyse performance improvements the device sub-efficiency level.
-We further developed a bottom-up manufacturing cost model with process step resolution that captures improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale to analyse cost reductions and their sources.
-It covers progress from early manufacturing in 2003 to today.
-We found that larger wafer sizes have been largely responsible for cost reductions.
-We found that XXX well known XXX.
-We collected historical data on device performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
-We used this information to identify and quantify the principal sources of performance improvements in LED manufacturing.
-We found that in order to quantify the impact of single technological changes, it is necessary to analyse performance improvements the device sub-efficiency level.
-We further developed a bottom-up manufacturing cost model with process step resolution that captures improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale to analyse cost reductions and their sources.
-It covers progress from early manufacturing in 2003 to today.
-We found that larger wafer sizes have been largely responsible for cost reductions.
-We found that XXX well known XXX.
+    Within only 25 years of the introduction of the first commercial white light-emitting diode, efficiency has increased by three orders of magnitude while manufacturing costs have fallen by two orders of magnitude.
+The total market penetration in general illumination has recently exceeded 50\% in the European Union and the United States \cite{eu2019impactass}\cite{stratunl2018}.
+Today, the solid-state lighting markets in the US and the EU together are valued at 58.7 Bn.\$(2020) \cite{gvr2020market_us}\cite{gvr2020market_eu}.
+Legislation and policy aimed at increasing the adoption of solid-state light sources has been implemented around the world \cite{2009-125-EC_regulation-2012_impact}.
+Estimates for the electrical energy saved annually from the adoption of energy efficiency lamps ranges from 131 TWh/year in 2020 for the EU to 442 TWh/year in 2020 for the US.
+
+    The early development history of light-emitting diodes is well documented \cite{Steigerwald2002}\cite{Nakamura2013}\cite{feezell2018invention}.
+Major works dealing with device efficiency improvements have documented overall progress \cite{azevedo2009transition}\cite{tsao2010solid}\cite{pattison2017solid}.
+Despite the excellent overview these publications provide, they did not systematically investigate the sources and impact of related technology improvements.
+Investigations regarding cost reductions are rarely reported in scientific literature.
+The focus is instead on the energy economics implications of decreased electricity consumption \cite{tsao2010world}\cite{Tsao2010SSL}.
 
@@
 Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
-For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources, shown in figure \ref{fig:cct}.
+For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources.
 A thorough treatment of consumer experience metrics is given by Weinold et al.
@@
 
-\begin{figure} [ht]
-    \begin{center}
-        \includegraphics[width=\textwidth]{SPIE/article/cct_spectrum.pdf}
-    \end{center}
-    \caption{Color temperature for temperature values between 1,000K and 12,000K.
-Vertical lines indicate the color temperature of blue light-emitting diodes used in combinations with phosphors.
-Arrows indicate flexibility through phosphor composition variation.
-Two data points are shown: The 1997 \textit{Nichia} \ch{YAG:Ce^{3+}} (YAG) phosphor at 5800K and the 2004 \textit{Lumileds} \ch{Sr_2Si_5N_8:Eu^{2+}} ("258") phosphor at below 4000K (tuneable).
-Color data generated using the \textit{luxpy} package for \textit{Python}.
-CCT for phosphors computed using the \textit{color science} package for \textit{Python}, from spectral data in publications  \cite{bando1998development,MuellerMach2005}.}
-    \label{fig:cct}
-\end{figure}
-
-\clearpage
 \section{PERFORMANCE IMPROVEMENTS}
 
-    list performance improvements, mention physical limits
-    mention loss channels with process improvements vs channels with breakthroughs and spillovers
-    mention caveats in manufacturing cost decrease
-    mention most important factor in manufacturing cost decrease
+     Overall efficiency in best performing devices improved from $\eta_L=5.8\%$ in 2002 to $\eta_L = 38.7\%$ in 2020.
+As previous authors have noted, no single loss channel dominates the overall the efficiency\cite{tsao2010solid}.
+We note, however, that those loss channels with a fixed physical limit become significantly more dominant in 2016 and 2020.
+For instance, Stokes loss, describing the energy dissipated upon conversion from short wavelength to long wavelength photons, becomes more dominant.
+Figure \ref{fig:efficiency} additionally shows the physical limits for the loss channels.
+Notably, sub-efficiencies for most current devices are only $\sim10\%$ below the physical limit.
+The exception is spectral efficiency, which at $\sim20\%$ below the physical limit shows larger potential for improvement.
 
-\subsection{Physical Efficiency}
+     Our comparison between the improvements in device sub-efficiencies between 2002 and 2020 shows that the aggregate efficiency improvement was not driven primarily by improvements in a single sub-efficiency.
+Instead, there has been consistent progress across the ensemble of loss channels corresponding to the device sub-efficiencies in the past 18 years: forward voltage efficiency ($70\%\rightarrow99.5\%$), internal quantum efficiency ($55\%\rightarrow90\%)$, electrical droop ($65\%\rightarrow90\%$), light-extraction efficiency ($60\%\rightarrow90\%$) and spectral efficiency ($74\% \rightarrow83\%$).
 
-give two waterfall diagrams (2003,2020) and give one example
-example: maybe droop with associated figure. is good example of process improvements
-
-\subsection{Consumer Experience}
-
-List phosphors, give a short overview but do not include figures
-show cct spectrum with lines of YAG and 258+ phosphors
+     We find that there are two distinct sources for improvements in device sub-efficiencies.
+In loss channels with physically well understood  XXXXX Our comparison between the improvements in device sub-efficiencies between 2002 and 2020 shows that the aggregate efficiency improvement was not driven primarily by improvements in a single sub-efficiency.
+Instead, there has been consistent progress across the ensemble of loss channels corresponding to the device sub-efficiencies in the past 18 years: forward voltage efficiency ($70\%\rightarrow99.5\%$), internal quantum efficiency ($55\%\rightarrow90\%)$, electrical droop ($65\%\rightarrow90\%$), light-extraction efficiency ($60\%\rightarrow90\%$) and spectral efficiency ($74\% \rightarrow83\%$).
 
@@
 
+    The cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, was estimated to have decreased from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
+
+    A preliminary analysis of cost factors suggests that an increase in wafer size in the manufacturing process is responsible for the largest part of the cost reduction.
+The number of die per wafer associated with this has been computed and is displayed in the inset plot of figure \ref{fig:cost}.
+The wafer diameter used in production has been steadily increasing since 2002.
+The model assumes diameters and associated number of die per wafer  $d(2002)=50$mm$\rightarrow851$ to $d(2020)=200$mm$\rightarrow26,838$.
+Note that the decrease in exclusion zone and cutting width also contribute to the increase in die per wafer.
+
+    As the number of die per wafer increases, the packaging steps carry a larger share of the total cost.
+This is because these steps are limited by the throughput of the associated equipment.
+The wafer processing steps are limited by the number of die per wafer and only to a lesser extend the throughput of equipment.
+
+\clearpage
 \begin{figure} [ht]
@@
     \end{center}
-    \caption{Modeled manufacturing cost for the specified architecture of light-emitting diode packages split by manufacturing step category, following the categories used in the US DoE reports \cite{doe2016solid}.
+    \caption{Modeled manufacturing cost for the specified architecture of light-emitting diode packages split by manufacturing step category, following the categories and color scheme used in the reports by the US DoE \cite{doe2016solid}.
 The inset plot shows the number of die per wafer corresponding to wafers of different sizes, associated street width and exclusion zone.

### 2021-02-11 SPIE/article/main.tex
revisions based on Sergey's comments

@@
 \author[b]{Sergey Kolesnikov}
-\author[c]{Laura Diaz Anadon}
-\affil[a]{ETH Zurich, Chair of Entrepreneurial Risks, Scheuchzerstrasse 7, 8092 Zurich, Switzerland}
-\affil[b]{University of Cambridge, Centre for Environment, Energy and Natural Resource Governance, The David Attenborough Building, CB2 3QZ Cambridge, UK}
+\author[b,c]{Laura Diaz Anadon}
+\affil[a]{Chair of Entrepreneurial Risks, ETH Zurich, Scheuchzerstrasse 7, CH-8092 Zurich, CH}
+\affil[b]{Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, CB3 9EP, UK}
+\affil[c]{Belfer Center for Science and International Affairs, Harvard Kennedy School, Harvard University, Cambridge, MA 02138, USA}
 
-\authorinfo{Further author information: (Send correspondence to Michael Weinold)\\Michael Weinold: E-mail: michael.weinold@alumni.ethz.ch}
+\authorinfo{Further author information: (Send correspondence to Michael Weinold) \\ MW: E-mail: mw799@cam.ac.uk \\ SK: E-mail: sk2063@cam.ac.uk \\ LDA: E-mail: lda24@cam.ac.uk}
 
@@
 
-    To understand the 
-    We collected historical data on device performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
-We used this information to identify and quantify the principal sources of performance improvements in LED manufacturing.
-We found that in order to quantify the impact of single technological changes, it is necessary to analyse performance improvements the device sub-efficiency level.
-We further developed a bottom-up manufacturing cost model with process step resolution that captures improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale to analyse cost reductions and their sources.
-It covers progress from early manufacturing in 2003 to today.
-We found that larger wafer sizes have been largely responsible for cost reductions.
-We found that XXX well known XXX.
+    With the aim of identifying and quantifying the principal sources of performance improvements and cost reductions in white light emitting diode manufacturing, we collect historical data on device cost and performance, technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
+We find that technological breakthroughs and process innovation contributed to performance improvements across the entire range of LED device sub-efficiencies, resulting in the overall increase in the lamp efficiency for the highest performing devices at a test current of 350mA from 5.8\% to 38.7\% between 2002 and 2020.
+We further develop a bottom-up manufacturing cost model with process-step resolution that captured improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale, to analyse progress in LED manufacturing cost structure between early manufacturing in 2003 and 2020.
+We estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location using state-of-the-art equipment has dropped from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
+We also find that the largest contribution to overall cost reductions has come from increase in wafer size, and that in 2020, chip packaging remains the largest contributor to cost at 60\%.
 
@@
 % Include a list of keywords after the abstract 
-\keywords{light-emitting diodes, innovation, efficiency, cost}
+\keywords{light-emitting diodes, LED manufacturing, LED efficiency, manufacturing cost, cost modeling, cost reduction, technological breakthroughs}
 
@@
 Details of the manufacturing process, as well as changes in the same were gathered from detailed patent analysis, augmented by semi-structured interviews with experts from industry and academia.
-For 2012, data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}.
-The model includes both the wafer treatment process as well as the packaging process.
+For 2012, additional data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}.
+Our model includes both the wafer treatment process as well as the packaging process.
 While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach.
-The main aim of the model is not to faithfully represent real world manufacturing conditions in Asian manufacturing locations, but rather to show the effect of single technological changes in the manufacturing process on total cost.
+The main aim of the model is not to faithfully represent real-world manufacturing such as those in present-day manufacturing locations in Asia, but rather to show the effect of technological change and manufacturing process improvements on total cost.
 
@@
 
-        To dis-aggregate the contribution of changes in single variables on total manufacturing cost, we used an approach introduced by Kavlak et al.
-\cite{kavlak2018evaluating}.
+        To dis-aggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak et al.
+in a cost model for photovoltaic modules \cite{kavlak2018evaluating}.
 It is based on the logarithmic derivative of the total differential of the cost function.
-We can write the cost function $C$ of the manufacturing process as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$, where $g_{iw}(r_w)$ gives the dependence of $i$th cost component on the $w$th variable.
-The contribution of changes in the cost model $\Delta C_z$ to total manufacturing cost $C$ can now be approximated.
+The contribution of changes in the cost model $\Delta C_z$ to total manufacturing cost $C$ can thus be approximated.
 For the detailed derivation, we refer to the supplementary material of the original publication.
 
-    	\begin{align}
-            C(\vec{r}) &= C(r_1,r_2, \dots) = \sum_i C_i = \sum_i C_i^0 \prod_w g_{iw}(r_w) \\
-            \Delta C_x &= \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t \\
-            \Delta C_z (t_1,t_2) &\approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
-        \end{align}
+    \subsection{Data Sources and Performance Calculations}+
 
-    \subsection{Literature Review and Performance Computation}
+        XXX mention patent analysis
+        XXX mention interviews
+        XXX mention company publications
+        XXX move any relevant sentences from cost model subsection here
 
-        To gain a detailed understanding of the sources and magnitude of efficiency improvements in devices, we gathered data on overall device performance and identified the associated device architecture, manufacturing processes and types of down conversion phosphors used.
+        To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in devices, we gathered data on overall device performance and identified the associated device architecture, manufacturing processes and types of down conversion phosphors used.
 We systematically identified the architecture and manufacturing process improvements related to each physical loss channel, gathered associated efficiency data from literature or computed the respective values from raw data.
-
-        For instance, to gather historical data on the improvement of losses associated with electrical droop, device performance data was extracted from datasheets, as shown in figure \ref{fig:droop}.
-This data was cross-referenced with information on the different chip architectures and manufacturing processes used for each device.
-Data was then augmented by industry forecasts from company publications \cite{osram2014osram}.
-Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gather a complete picture of the historical development of light-emitting diode efficiency.
-
-        \begin{figure} [ht]
-            \begin{center}
-                \includegraphics[width=0.85\textwidth]{SPIE/article/droop_lumileds.pdf}
-            \end{center}
-            \caption{Luminous intensity of four different \textit{Lumileds} high-power light-emitting diodes normalized to the value at a test current of $A_{test}=350$mA.
-The black curves describe the real measured intensity, the orange curves describe the estimated ideal intensity.
-Droop $D$, as defined in table \ref{tab:eff}, is the difference between these curves at the test current.
-Current-Intensity data extracted from device datasheets \cite{datasheet_lumileds_lux1,datasheet_lumileds_rebel,datasheet_lumileds_rebplus,lumi2019data}}
-            \label{fig:droop}
-        \end{figure}
-
-\section{METRICS}
-
-    Once technological breakthroughs and process improvements had been identified, their effect was quantified.
-The ensemble of metrics was required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
-They must further have the ability to directly describe the effect of single technological breakthroughs, device architectures and manufacturing process improvements on device performance.
-
-    These requirements precluded the use of some established metrics.
-For instance, a highly cited and frequently updated metric is the total luminous flux per light-emitting diode package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
-In combination with the cost per total flux, this visualization is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
-\cite{haitz1999case}.
-However, while these metrics are often used to showcase technological progress in light-emitting diode design and manufacturing, care must be taken to consider its limitations.
-
-    Firstly, the metric today retains only limited significance as a proxy for technological development in light-emitting diodes.
-This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
-Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
-Secondly, disregarding these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
-Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
-
-    Figure \ref{fig:haitz} shows an updated and expanded overview of best performing devices, both at the chip and package level, inspired by \textit{"Haitz's Law"}.
-It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
-While \textit{"Haitz's Law"} remains a popular visualization related to the progress in light-emitting diode design and manufacturing, .
-
-    \begin{figure} [ht]
-        \begin{center}
-            \includegraphics[width=\textwidth]{haitz_law_white.pdf}
-        \end{center}
-        \caption{Historical increase in flux for highest performing white \underline{light-emitting diode chips and (multi-chip) packages}, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
-Note that flux per single chips has not increased as starkly as the flux for multi-chip packages.
-Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
-Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, grey colored datapoints corresponding to the right ordinate.
-Sources: Data from Weinold et al.
-\cite{weinold2020technology}}
-        \label{fig:haitz}
-    \end{figure}
-
-    Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency.
-This metric defined as the product of all device sub-efficiencies associated with an ensemble of different loss channels $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$.
-Sub-efficiencies directly capture the effect of technology breakthroughs in device design and manufacturing process improvements.
-They thus provide the best metric to quantify technology breakthroughs.
-Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
 
@@
 \cite{pattison2017solid}.
-The historical development of the sub-efficiencies is displayed in figure \ref{fig:efficiency}.
-\\ *Also called "lamp efficiency" or "cumulative efficiency" by authors, such as Tsao et al.
-\cite{tsao2010solid}}
+The historical development of the sub-efficiencies is displayed in figure \ref{fig:efficiency}.}
         \bigskip
@@
 
+        For instance, to gather historical data on the reduction of losses associated with electrical droop, device performance data was calculated based on the information extracted from datasheets, as shown in figure \ref{fig:droop}.
+This data was cross-referenced with information on the different chip architectures and manufacturing processes used for each device.
+This data was then augmented by data provided in industry reports and company publications \cite{osram2014osram}.
+Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gather a complete picture of the historical development of light-emitting diode efficiency.
+
+        \begin{figure} [ht]
+            \begin{center}
+                \includegraphics[width=0.85\textwidth]{SPIE/article/droop_lumileds.pdf}
+            \end{center}
+            \caption{Luminous intensity of four different \textit{Lumileds} high-power light-emitting diodes normalized to the value at a test current of $A_{test}=350$mA.
+The black curves describe the real measured intensity, the orange curves describe the estimated ideal intensity.
+Droop $D$, as defined in table \ref{tab:eff}, is the difference between these curves at the test current.
+Current-Intensity data extracted from device datasheets \cite{datasheet_lumileds_lux1,datasheet_lumileds_rebel,datasheet_lumileds_rebplus,lumi2019data}}
+            \label{fig:droop}
+        \end{figure}
+
+\subsection{Choice of Metrics}
+
+    Once technological breakthroughs and process improvements had been identified, the next step was to quantify their effect.
+The ensemble of metrics is required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
+They must further have the ability to directly describe the effect of single technological breakthroughs, device architectures and manufacturing process improvements on device performance.
+
+    These requirements precluded the use of some established metrics.
+For instance, a highly cited and frequently updated metric is the total luminous flux per light-emitting diode package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
+In combination with the cost per total flux, it is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
+\cite{haitz1999case}.
+However, while it is often used to showcase technological progress in light-emitting diode design and manufacturing, care must be taken to consider its limitations.
+
+    Firstly, the metric today retains only limited significance as a proxy for technological development in light-emitting diodes.
+This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
+Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
+Secondly, disregarding these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
+Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
+
+    Figure \ref{fig:haitz} shows an updated and expanded overview of best performing devices, both at the chip and package level, to demonstrate the limitations of these metrics.
+It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
+
+    \begin{figure} [ht]
+        \begin{center}
+            \includegraphics[width=\textwidth]{haitz_law_white.pdf}
+        \end{center}
+        \caption{Historical increase in flux for highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
+Note that flux per single chips has not increased as starkly as the flux for multi-chip packages.
+Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
+Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, blue colored datapoints corresponding to the right ordinate.
+Sources: Data from Weinold et al.
+\cite{weinold2020technology}}
+        \label{fig:haitz}
+    \end{figure}
+
+    Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency.
+This metric is defined as the product of all device sub-efficiencies associated with an ensemble of different loss channels $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$.
+Sub-efficiencies directly capture the effect of particular technological breakthroughs in device design and manufacturing process improvements.
+Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
+
     Performance improvements in metrics related to consumer experience have also played a role in the adoption of light-emitting diode based luminaires \cite{cowan2011understanding}.
@@
 For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources.
-A thorough treatment of consumer experience metrics is given by Weinold et al.
-\cite{weinold2020technology}.
-Flicker was not considered in this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
+The work on analyzing consumer experience metrics is ongoing and will be presented in our future work.
+Notably, we exclude flicker from this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
 We instead refer to a recent publication by Weinold \cite{weinold2020long}.
 
-\section{PERFORMANCE IMPROVEMENTS}
+\section{RESULTS}
 
+\subsection{Performance Improvements}
+
+     Results for performance improvements are presented in Figure \ref{fig:efficiency}.
 Overall efficiency in best performing devices improved from $\eta_L=5.8\%$ in 2002 to $\eta_L = 38.7\%$ in 2020.
-As previous authors have noted, no single loss channel dominates the overall the efficiency\cite{tsao2010solid}.
+As previous studies have noted, no single loss channel dominates the overall the efficiency\cite{tsao2010solid}.
 We note, however, that those loss channels with a fixed physical limit become significantly more dominant in 2016 and 2020.
@@
 Figure \ref{fig:efficiency} additionally shows the physical limits for the loss channels.
-Notably, sub-efficiencies for most current devices are only $\sim10\%$ below the physical limit.
+Notably, sub-efficiencies for the most current devices are only $\sim10\%$ below the physical limit.
 The exception is spectral efficiency, which at $\sim20\%$ below the physical limit shows larger potential for improvement.
@@
 
-     We find that there are two distinct sources for improvements in device sub-efficiencies.
-In loss channels with physically well understood  XXXXX Our comparison between the improvements in device sub-efficiencies between 2002 and 2020 shows that the aggregate efficiency improvement was not driven primarily by improvements in a single sub-efficiency.
-Instead, there has been consistent progress across the ensemble of loss channels corresponding to the device sub-efficiencies in the past 18 years: forward voltage efficiency ($70\%\rightarrow99.5\%$), internal quantum efficiency ($55\%\rightarrow90\%)$, electrical droop ($65\%\rightarrow90\%$), light-extraction efficiency ($60\%\rightarrow90\%$) and spectral efficiency ($74\% \rightarrow83\%$).
+     XXXXXXXXXXXXXXXX
 
-\section{MANUFACTURING COST REDUCTIONS}
+\subsection{Manufacturing Cost Reductions}
 
-    The cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, was estimated to have decreased from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
+    Using our cost model, we estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, to have decreased from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
 
-    A preliminary analysis of cost factors suggests that an increase in wafer size in the manufacturing process is responsible for the largest part of the cost reduction.
-The number of die per wafer associated with this has been computed and is displayed in the inset plot of figure \ref{fig:cost}.
+    A preliminary analysis of components of the cost model suggests that an increase in wafer size in the manufacturing process is responsible for the largest part of the cost reduction.
+The number of die per wafer associated with this is displayed in the inset plot of figure \ref{fig:cost}.
 The wafer diameter used in production has been steadily increasing since 2002.
@@
 
+\subsection{Future Work}
+
 \clearpage
@@
     \end{center}
-    \caption{Impact of technology breakthroughs and manufacturing process improvements on historical improvements in sub-efficiencies of phosphor-converted warm white light-emitting diodes with test currents of at least $I_\text{test}=350$mA.
+    \caption{Impact of technology breakthroughs and manufacturing process improvements on historical developments in sub-efficiencies of phosphor-converted warm white light-emitting diodes with test currents of at least $I_\text{test}=350$mA.
 The overall lamp efficiency $\eta_L$ is displayed as the rightmost column.
@@
 Horizontal colored bars give state-of-the-art sub-efficiencies for five years: \textcolor{blue}{1997}, \textcolor{teal}{2002}, \textcolor{orange}{2010}, \textcolor{magenta}{2016} and \textcolor{red}{2020}.
-Colored annotation "N/A" indicates the sub-efficiency of the corresponding year cannot be computed for the following reasons: V$_\text{f}$E, Droop: depend on current, which was below 350mA at the time; CE, SE: warm white spectrum LEDs not available at the time.
+Colored annotation "N/A" indicates that the sub-efficiency of the corresponding year cannot be computed for the following reasons: V$_\text{f}$E, Droop: depend on current, which was below 350mA at the time; CE, SE: warm white spectrum LEDs not available at the time.
 Physical limits are indicated by black horizontal bars.
 The possible range for the physical limit of V$_\text{f}$E exceeds 100\% and depends on electrical device parameters, which are discussed in \cite{david2016electrical}.
-The range is given by an upward pointing black arrow.
-Color of vertical bars indicates improvement through either technology spillovers or other improvements (technology breakthroughs or process learning).
-Spillovers are annotated and listed in an inset table.
-Efficiency acronyms: forward voltage efficiency (V$_\text{f}$E), internal quantum efficiency (IQE), light extraction efficiency (LEE), conversion efficiency (CE), spectral efficiency (SE), power conversion efficiency ($\eta_{PC}$).
-Spillover acronyms: patterned sapphire substrate (PSS), indium tin oxide (ITO).
+The range of this limit is indicated by an upward pointing black arrow.
+Efficiency acronyms are listed in table \ref{tab:eff}.
 Data for historical improvements of sub efficiencies from Weinold et al.
@@
 
-The main author gratefully acknowledges funding by the \textit{Swiss Study Foundation} of Zurich, Switzerland.
-The authors further gratefully acknowledge project funding by the \textit{Alfred P.
-Sloan Foundation} of New York, USA.
+We express our gratitude to all inteviewees for their willingness to participate in this study and invaluable contributions.
+This research is funded by the \textit{Alfred P.
+Sloan Foundation} of New York, USA, grant number 253128.
+Michael Weinold further gratefully acknowledges support from the \textit{Swiss Study Foundation} of Zurich, Switzerland.
 

### 2021-02-11 SPIE/article/main.tex
more corrections

@@
 
-    The early development history of light-emitting diodes is well documented \cite{Steigerwald2002}\cite{Nakamura2013}\cite{feezell2018invention}.
-Major works dealing with device efficiency improvements have documented overall progress \cite{azevedo2009transition}\cite{tsao2010solid}\cite{pattison2017solid}.
-Despite the excellent overview these publications provide, they did not systematically investigate the sources and impact of related technology improvements.
-Investigations regarding cost reductions are rarely reported in scientific literature.
-The focus is instead on the energy economics implications of decreased electricity consumption \cite{tsao2010world}\cite{Tsao2010SSL}.
-
 \clearpage
-\section{METHODS}
+\section{Methods and Data}
 \label{sec:methods}
 
-    \subsection{Manufacturing Cost Model}
+\subsection{Choice of Metrics}
 
-        To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed.
-It covers the entire manufacturing process of GaN-on-sapphire-based phosphor-converted low-to-mid power light-emitting diode packages of different chip architectures.
-We considered the classical p-side-up lateral current spreading architecture, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture.
-An excerpt of the device architectures considered is shown in figure \ref{fig:chip_arch}.
-The model was constructed for the years 2003, 2012 and 2020.
-It was populated with equipment data from European and North American firms, selected for a virtual North American manufacturing location.
-Process specific step parameters were derived from scientific literature, company publications, archived product catalogs and patent literature.
-Where necessary, this data was augmented using information gathered during semi-structured interviews with experts from industry and academia.
-Details of the manufacturing process, as well as changes in the same were gathered from detailed patent analysis, augmented by semi-structured interviews with experts from industry and academia.
-For 2012, additional data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}.
-Our model includes both the wafer treatment process as well as the packaging process.
-While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach.
-The main aim of the model is not to faithfully represent real-world manufacturing such as those in present-day manufacturing locations in Asia, but rather to show the effect of technological change and manufacturing process improvements on total cost.
+    To describe the effect of technological breakthroughs and manufacturing process improvements, a set of metrics had to be identified.
+The ensemble of metrics is required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
+They must further have the ability to directly describe the effect of single technological breakthroughs, evolution in device architecture and manufacturing process improvements on device performance.
+
+    These requirements precluded the use of some established metrics.
+For instance, a highly cited and frequently updated metric is the total luminous flux per light-emitting diode package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
+In combination with the cost per total flux, it is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
+\cite{haitz1999case}.
+However, while it is often used to showcase technological progress in light-emitting diode design and manufacturing, care must be taken to consider its limitations.
+
+    Firstly, the metric today retains only limited significance as a proxy for technological development in light-emitting diodes.
+This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
+Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
+Secondly, disregarding these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
+Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
+
+    Figure \ref{fig:haitz} shows an updated and expanded overview of the flux per device and the cost per flux of highest performing light-emitting diodes, both at the chip and package level, to demonstrate the limitations of these metrics.
+It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
 
@@
         \begin{center}
-                \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
+            \includegraphics[width=\textwidth]{haitz_law_white.pdf}
         \end{center}
-            \caption{Cutaway side views of the evolution of chip architectures for classical chip designs (lateral current spreading).
-Note that dimensions are not to scale and smaller features are greatly exaggerated for visibility.
-Years correspond to earliest identified patent priority date.
-Dashed boxes indicate chip designs not brought to large scale production.
-Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
-            \label{fig:chip_arch}
+        \caption{Historical increase in flux for highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
+Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
+Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, blue colored datapoints corresponding to the right ordinate.}
+        \label{fig:haitz}
     \end{figure}
 
-        To dis-aggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak et al.
-in a cost model for photovoltaic modules \cite{kavlak2018evaluating}.
-It is based on the logarithmic derivative of the total differential of the cost function.
-The contribution of changes in the cost model $\Delta C_z$ to total manufacturing cost $C$ can thus be approximated.
-For the detailed derivation, we refer to the supplementary material of the original publication.
-
-    \subsection{Data Sources and Performance Calculations}+
-
-        XXX mention patent analysis
-        XXX mention interviews
-        XXX mention company publications
-        XXX move any relevant sentences from cost model subsection here
-
-        To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in devices, we gathered data on overall device performance and identified the associated device architecture, manufacturing processes and types of down conversion phosphors used.
-We systematically identified the architecture and manufacturing process improvements related to each physical loss channel, gathered associated efficiency data from literature or computed the respective values from raw data.
+    \clearpage
+    Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency.
+This metric is defined as the product of all device sub-efficiencies associated with an ensemble of different loss channels $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$.
+Sub-efficiencies directly capture the effect of particular technological breakthroughs in device design and manufacturing process improvements.
+Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
 
@@
 
+    Performance improvements in metrics related to consumer experience have also played a role in the adoption of light-emitting diode based luminaires \cite{cowan2011understanding}.
+Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
+For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources.
+The work on analyzing consumer experience metrics is ongoing and will be presented in our future work.
+Notably, we exclude flicker from this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
+We instead refer to a recent publication by Weinold \cite{weinold2020long}.
+
+    \subsection{Data Sources and Performance Calculations}
+    \label{subsec:data}
+
+        To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in light-emitting diode manufacturing, we gathered data from a variety of sources.
+The aim was to collect historical data on overall device performance and historical development in sub-efficiencies.
+Where possible, we gathered  efficiency data from literature or computed the respective values from raw data.
+We further had to identify the associated device architectures, manufacturing processes and types of down conversion phosphors used.
+Data was extracted from patent literature, scientific publications, industry reports and technology forecasts.
+Additional information and data was provided by experts from both the light-emitting diode manufacturing industry and the manufacturing equipment industry during semi-structured interviews.
+
         For instance, to gather historical data on the reduction of losses associated with electrical droop, device performance data was calculated based on the information extracted from datasheets, as shown in figure \ref{fig:droop}.
@@
 
-\subsection{Choice of Metrics}
+    \subsection{Manufacturing Cost Model}
 
-    Once technological breakthroughs and process improvements had been identified, the next step was to quantify their effect.
-The ensemble of metrics is required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
-They must further have the ability to directly describe the effect of single technological breakthroughs, device architectures and manufacturing process improvements on device performance.
-
-    These requirements precluded the use of some established metrics.
-For instance, a highly cited and frequently updated metric is the total luminous flux per light-emitting diode package \cite{Liu2009,haitz2011solid,cho2017white,Fontoynont2018}.
-In combination with the cost per total flux, it is sometimes referred to as \textit{"Haitz's Law"}, in reference to an early report on LED development by Haitz et al.
-\cite{haitz1999case}.
-However, while it is often used to showcase technological progress in light-emitting diode design and manufacturing, care must be taken to consider its limitations.
-
-    Firstly, the metric today retains only limited significance as a proxy for technological development in light-emitting diodes.
-This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
-Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
-Secondly, disregarding these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
-Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
-
-    Figure \ref{fig:haitz} shows an updated and expanded overview of best performing devices, both at the chip and package level, to demonstrate the limitations of these metrics.
-It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
+        To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed.
+It covers the entire manufacturing process of GaN-on-sapphire-based phosphor-converted low-to-mid power light-emitting diode packages of different chip architectures.
+We considered the classical p-side-up lateral current spreading architecture, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture.
+An excerpt of the device architectures considered is shown in figure \ref{fig:chip_arch}.
+The earliest commercial warm-white phosphor converted blue light emitting diodes were introduced in 2003, so the model was constructed for the years 2003, 2012 and 2020.
+It was populated with equipment data from European and North American firms, selected for a virtual North American manufacturing location.
+Details of the manufacturing processes and process-specific step parameters were derived from the sources mentiond in section \ref{subsec:data}.
+For 2012, additional data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}.
+Our model includes both the wafer treatment process as well as the packaging process.
+While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach.
+The main aim of the model is not to faithfully represent real-world manufacturing such as those in present-day manufacturing locations in Asia, but rather to show the effect of technological change and manufacturing process improvements on total cost.
 
@@
             \begin{center}
-            \includegraphics[width=\textwidth]{haitz_law_white.pdf}
+                \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
             \end{center}
-        \caption{Historical increase in flux for highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
-Note that flux per single chips has not increased as starkly as the flux for multi-chip packages.
-Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
-Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, blue colored datapoints corresponding to the right ordinate.
-Sources: Data from Weinold et al.
-\cite{weinold2020technology}}
-        \label{fig:haitz}
+            \caption{Cutaway side views of the evolution of chip architectures for classical chip designs (lateral current spreading).
+Note that dimensions are not to scale.
+Years correspond to earliest identified patent priority date.
+Dashed boxes indicate chip designs not brought to large scale production.
+Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
+            \label{fig:chip_arch}
         \end{figure}
 
-    Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency.
-This metric is defined as the product of all device sub-efficiencies associated with an ensemble of different loss channels $\eta_L = \prod_{i=(V_f,\dots,S)} \eta_i$.
-Sub-efficiencies directly capture the effect of particular technological breakthroughs in device design and manufacturing process improvements.
-Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
-
-    Performance improvements in metrics related to consumer experience have also played a role in the adoption of light-emitting diode based luminaires \cite{cowan2011understanding}.
-Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
-For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources.
-The work on analyzing consumer experience metrics is ongoing and will be presented in our future work.
-Notably, we exclude flicker from this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
-We instead refer to a recent publication by Weinold \cite{weinold2020long}.
+        To dis-aggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak et al.
+in a cost model for photovoltaic modules \cite{kavlak2018evaluating}.
+It is based on the logarithmic derivative of the total differential of the cost function.
+For the detailed derivation, we refer to the supplementary material of the original publication.
 
@@
 As previous studies have noted, no single loss channel dominates the overall the efficiency\cite{tsao2010solid}.
-We note, however, that those loss channels with a fixed physical limit become significantly more dominant in 2016 and 2020.
+Figure \ref{fig:efficiency} additionally shows the physical limits for the loss channels.
+We note, that those loss channels with a fixed physical limit below 100\% have become significantly more dominant in 2016 and 2020.
 For instance, Stokes loss, describing the energy dissipated upon conversion from short wavelength to long wavelength photons, becomes more dominant.
-Figure \ref{fig:efficiency} additionally shows the physical limits for the loss channels.
 Notably, sub-efficiencies for the most current devices are only $\sim10\%$ below the physical limit.
@@
 
-     XXXXXXXXXXXXXXXX
+     XXX video text
 
-\subsection{Manufacturing Cost Reductions}
-
-    Using our cost model, we estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, to have decreased from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
-
-    A preliminary analysis of components of the cost model suggests that an increase in wafer size in the manufacturing process is responsible for the largest part of the cost reduction.
-The number of die per wafer associated with this is displayed in the inset plot of figure \ref{fig:cost}.
-The wafer diameter used in production has been steadily increasing since 2002.
-The model assumes diameters and associated number of die per wafer  $d(2002)=50$mm$\rightarrow851$ to $d(2020)=200$mm$\rightarrow26,838$.
-Note that the decrease in exclusion zone and cutting width also contribute to the increase in die per wafer.
-
-    As the number of die per wafer increases, the packaging steps carry a larger share of the total cost.
-This is because these steps are limited by the throughput of the associated equipment.
-The wafer processing steps are limited by the number of die per wafer and only to a lesser extend the throughput of equipment.
-
-\subsection{Future Work}
-
-\clearpage
     \begin{figure} [ht]
@@
 
+\subsection{Manufacturing Cost Reductions}
+
+    Using our cost model, we estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, to have decreased from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
+
+    A preliminary analysis of components of the cost model suggests that an increase in wafer size in the manufacturing process is responsible for the largest part of the cost reduction.
+The number of die per wafer associated with this is displayed in the inset plot of figure \ref{fig:cost}.
+The wafer diameter used in production has been steadily increasing since 2002.
+The model assumes diameters and associated number of die per wafer  $d(2002)=50$mm$\rightarrow851$ to $d(2020)=200$mm$\rightarrow26,838$.
+Note that the decrease in exclusion zone and cutting width also contribute to the increase in die per wafer.
+
+    As the number of die per wafer increases, the packaging steps carry a larger share of the total cost.
+This is because these steps are limited by the throughput of the associated equipment.
+The wafer processing steps are limited by the number of die per wafer and only to a lesser extend the throughput of equipment.
+
 \begin{figure} [ht]
@@
 \end{figure}
+
+    \subsection{Future Work}
 

### 2021-02-11 SPIE/article/main.tex
more corrections

@@
 
-mention: technology had important improvements in past 25 years
-mention anniversiary of technology
-mention adoption in EU and US
-mention commoditization
-lack of understanding of sources of improvements across 25 years
-mention works by tsao (and at least two others)
+    Within only 25 years of the introduction of the first commercial white light-emitting diode, efficiency has increased by three orders of magnitude while manufacturing costs have fallen by two orders of magnitude.
+After legislation aimed at increasing the adoption of solid-state light sources has been implemented around the world, the solid state lighting industry has become a rare success story in the global drive for the transformation of energy systems.
+Estimates for the electrical energy saved annually from the adoption of energy efficiency lamps ranges from 131 TWh/year in 2020 for the EU \cite{eu2019impactass} to 442 TWh/year in 2020 for the US \cite{yamada2015adoption,guidehouse2020adoption}.
 
-    Within only 25 years of the introduction of the first commercial white light-emitting diode, efficiency has increased by three orders of magnitude while manufacturing costs have fallen by two orders of magnitude.
-The total market penetration in general illumination has recently exceeded 50\% in the European Union and the United States \cite{eu2019impactass}\cite{stratunl2018}.
-Today, the solid-state lighting markets in the US and the EU together are valued at 58.7 Bn.\$(2020) \cite{gvr2020market_us}\cite{gvr2020market_eu}.
-Legislation and policy aimed at increasing the adoption of solid-state light sources has been implemented around the world \cite{2009-125-EC_regulation-2012_impact}.
-Estimates for the electrical energy saved annually from the adoption of energy efficiency lamps ranges from 131 TWh/year in 2020 for the EU to 442 TWh/year in 2020 for the US.
+    Understanding the sources and impact of these dramatic improvements on LED manufacturing is essential for researchers, industry professionals, as well as policymakers in the sphere of energy, science and innovation, as they may provide valuable insights for the acceleration of innovation in other demand-side energy technologies.
+Our research aims to XXX
 

### 2021-02-11 SPIE/article/main.tex
additional corrections

@@
 
-    Within only 25 years of the introduction of the first commercial white light-emitting diode, efficiency has increased by three orders of magnitude while manufacturing costs have fallen by two orders of magnitude.
-After legislation aimed at increasing the adoption of solid-state light sources has been implemented around the world, the solid state lighting industry has become a rare success story in the global drive for the transformation of energy systems.
-Estimates for the electrical energy saved annually from the adoption of energy efficiency lamps ranges from 131 TWh/year in 2020 for the EU \cite{eu2019impactass} to 442 TWh/year in 2020 for the US \cite{yamada2015adoption,guidehouse2020adoption}.
+    Within just 25 years of the introduction of first commercial white light-emitting diodes (LED), the solid state lighting industry has become a rare success story in the global drive for the transformation of energy systems.
+Today, the solid-state lighting markets in the US and the EU together are valued at 58.7 Bn.\$(2020) \cite{gvr2020market_us,gvr2020market_eu} with general illumination market penetration exceeding 50\% \cite{eu2019impactass,stratunl2018}.
+Estimates for the electrical energy saved annually from SSL adoption ranges from 131 TWh/year in 2020 for the EU \cite{eu2019impactass} to 442 TWh/year in 2020 for the US \cite{yamada2015adoption,guidehouse2020adoption}, which is on par with the amount of energy produced by all deployed solar photovoltaic installations in these regions.
+Legislation aimed at increasing the adoption of solid-state light sources that has been implemented around the world creates favorable conditions for further expansion of the industry.
+However, this revolution in lighting would not have been possible without dramatic improvements in the overall LED device efficiency and reductions in LED manufacturing costs.
+In last twenty years, efficiency has increased by three orders of magnitude [CHECK IF THAT IS APPROPRIATE, AS "ORDERS OF MAGNITUDE" FOR VARIABLES MEASURED IN PERCENTAGES ARE NOT VERY MEANINGFUL], while manufacturing costs have fallen by two orders of magnitude.
 
     Understanding the sources and impact of these dramatic improvements on LED manufacturing is essential for researchers, industry professionals, as well as policymakers in the sphere of energy, science and innovation, as they may provide valuable insights for the acceleration of innovation in other demand-side energy technologies.
-Our research aims to XXX
+With this aim, in this paper we identify and quantify the principal sources of performance improvements and cost reductions in phosphor-converted white LED lighting devices over the last 20 years.
 
-\clearpage
 \section{Methods and Data}
@@
 
-    Firstly, the metric today retains only limited significance as a proxy for technological development in light-emitting diodes.
-This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
-Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
-Secondly, disregarding these limitations, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
-Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
-
-    Figure \ref{fig:haitz} shows an updated and expanded overview of the flux per device and the cost per flux of highest performing light-emitting diodes, both at the chip and package level, to demonstrate the limitations of these metrics.
-It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
-
     \begin{figure} [ht]
@@
 
-    \clearpage
+    Firstly, the metric today retains only limited significance as an appropriate for technological development in light-emitting diodes.
+This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
+Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
+Secondly, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
+Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
+
+    Figure \ref{fig:haitz} shows an updated and expanded overview of the flux per device and the cost per flux of highest performing light-emitting diodes, both at the chip and package level, to demonstrate the limitations of these metrics.
+It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
+
     Instead, progress in light-emitting diode technology is best described by the overall device efficiency, or lamp efficiency.
@@
 Notably, we exclude flicker from this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
-We instead refer to a recent publication by Weinold \cite{weinold2020long}.
+For further discussion of flicker in this context, see a recent publication by Weinold \cite{weinold2020long}.
 
@@
 
-     XXX video text
-
     \begin{figure} [ht]
@@
 
+    Notably, efficiency improvements in sub-efficiencies with a good understanding of the underlying physical loss channels, such as light-extraction efficiency, forward voltage efficiency and spectral efficiency were mostly a result of targeted research and development.
+For instance, optical simulations could be performed to determine ways of modifying the device architecture to improve light extraction efficiency.
+On the other hand, efficiency improvements in sub-efficiencies with a limited physical understanding internal quantum efficiency and the associated electrical droop at high currents were mostly due to process improvements in the manufacturing process.
+For instance, the reasons for electrical droop have, until recently, been a topic of hot debate, but significant improvements in this channel were still achieved by varying metal-organic vapour deposition (MOCVD) growth parameters in large ensembles of wafer batches.
+
 \subsection{Manufacturing Cost Reductions}
@@
 
+    XXX some numbers from the figure
+
 \begin{figure} [ht]
@@
     \end{center}
-    \caption{Modeled manufacturing cost for the specified architecture of light-emitting diode packages split by manufacturing step category, following the categories and color scheme used in the reports by the US DoE \cite{doe2016solid}.
-The inset plot shows the number of die per wafer corresponding to wafers of different sizes, associated street width and exclusion zone.
+    \caption{Left panel: Modeled manufacturing cost for low power classical chip architecture (lateral current spreading) the specified architecture of light-emitting diode packages split by manufacturing step category.
+Right panel: The number of die per wafer corresponding to wafers of different sizes, associated street width and exclusion zone.
 Calculated using approximations introduced by de Vries \cite{deVries2005}.}
@@
 
-    \subsection{Future Work}
+\section{CONCLUSIONS}
+
+    In this study, we found that technological breakthroughs and process innovation contributed to performance improvements across the entire range of LED device sub-efficiencies, resulting in the overall lamp efficiency increase for the highest performing devices at a test current of 350mA from 5.8\% in 2002 to 38.7\% in 2020.
+Notably, we found that efficiency improvements in loss channels with a good underlying understanding of physical mechanism were driven mostly by research and development, while improvements in channels with a lack of such understanding required extensive experimentation and learning-by-doing in the manufacturing setting.
+It shows the importance of further research into the underlying solid-state physics of light-emitting diodes.
+A deeper understanding of the underlying effects and mechanisms associated with device loss channels will potentially enable further advances in LED technology, with spectral efficiency showing the largest potential for further improvements.
+
+    We also construct the manufacturing cost model for low-power and mid-power LED packages produced at a virtual US location using state-of-the-art equipment.
+We estimate that the cost of LED manufacturing in these conditions has dropped from 1.11$(2020) in 2002 to 0.05$(2020) in 2020, a 95.5\% decrease.
+The largest contribution to overall historical cost reductions came from increase in wafer size.
+The largest remaining contributor to the manufacturing cost is the LED packaging, contributing 60\% to the cost structure in 2020.
 

### 2021-02-11 SPIE/article/main.tex
further edits

@@
 \usepackage{tabularx}
+\usepackage{wrapfig}
 \setlength{\extrarowheight}{2pt}
@@
 Estimates for the electrical energy saved annually from SSL adoption ranges from 131 TWh/year in 2020 for the EU \cite{eu2019impactass} to 442 TWh/year in 2020 for the US \cite{yamada2015adoption,guidehouse2020adoption}, which is on par with the amount of energy produced by all deployed solar photovoltaic installations in these regions.
-Legislation aimed at increasing the adoption of solid-state light sources that has been implemented around the world creates favorable conditions for further expansion of the industry.
-However, this revolution in lighting would not have been possible without dramatic improvements in the overall LED device efficiency and reductions in LED manufacturing costs.
-In last twenty years, efficiency has increased by three orders of magnitude [CHECK IF THAT IS APPROPRIATE, AS "ORDERS OF MAGNITUDE" FOR VARIABLES MEASURED IN PERCENTAGES ARE NOT VERY MEANINGFUL], while manufacturing costs have fallen by two orders of magnitude.
+This revolution in lighting would not have been possible without dramatic improvements in the overall device efficiency and reductions in manufacturing costs.
 
@@
 
-        For instance, to gather historical data on the reduction of losses associated with electrical droop, device performance data was calculated based on the information extracted from datasheets, as shown in figure \ref{fig:droop}.
-This data was cross-referenced with information on the different chip architectures and manufacturing processes used for each device.
-This data was then augmented by data provided in industry reports and company publications \cite{osram2014osram}.
-Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gather a complete picture of the historical development of light-emitting diode efficiency.
-
         \begin{figure} [ht]
@@
         \end{figure}
+
+        For instance, to gather historical data on the reduction of losses associated with electrical droop, device performance data was calculated based on the information extracted from datasheets, as shown in figure \ref{fig:droop}.
+This data was cross-referenced with information on the different chip architectures and manufacturing processes used for each device.
+This data was then augmented by data provided in industry reports and company publications \cite{osram2014osram}.
+Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gather a complete picture of the historical development of light-emitting diode efficiency.
 
@@
 The range of this limit is indicated by an upward pointing black arrow.
-Efficiency acronyms are listed in table \ref{tab:eff}.
-Data for historical improvements of sub efficiencies from Weinold et al.
-\cite{weinold2020technology}}
+Efficiency acronyms are listed in table \ref{tab:eff}.}
         \label{fig:efficiency}
@@
 Note that the decrease in exclusion zone and cutting width also contribute to the increase in die per wafer.
-
 As the number of die per wafer increases, the packaging steps carry a larger share of the total cost.
@@
 
-    XXX some numbers from the figure
-
-\begin{figure} [ht]
-    \begin{center}
-        \includegraphics[width=0.85\textwidth]{SPIE/article/costmodel_calibration.pdf}
-    \end{center}
-    \caption{Left panel: Modeled manufacturing cost for low power classical chip architecture (lateral current spreading) the specified architecture of light-emitting diode packages split by manufacturing step category.
-Right panel: The number of die per wafer corresponding to wafers of different sizes, associated street width and exclusion zone.
-Calculated using approximations introduced by de Vries \cite{deVries2005}.}
-    \label{fig:cost}
-\end{figure}
+    \begin{table}[h!]
+        \caption{Manufacturing cost of low-power classical chips (lateral current spreading) estimated with the cost model introduced in section \ref{sec:methods}.
+Relative contributions of different manufacturing process steps are shown in figure \ref{fig:cost}.}
+        \bigskip
+            \centering
+            \begin{tabularx}{\textwidth}{|l|l|X|X|X|}
+            	\hline
+            		\textit{Year} & \textit{Unit} & \textit{Cost (Wafer Proc.)} & \textit{Cost (Packaging)} & \textit{Total Cost} \\
+                \hline
+                    2002 & \$(2020) & 0.975 & 0.138 & 1.113 \\
+                \hline
+                    2002-2012 & \% & -62.05 & -47.10 & -90.12 \\
+                \hline
+                    2012 & \$(2020) & 0.037 & 0.073 & 0.110 \\
+                \hline
+                    2012-2020 & \% & -67.56 & -36.98 & -47.27 \\
+                \hline
+                    2020 & \$(2020) & 0.012 & 0.046 & 0.058 \\
+                \hline
+            \end{tabularx}
+            \label{tab:cost}
+        \end{table}
 
@@
     We also construct the manufacturing cost model for low-power and mid-power LED packages produced at a virtual US location using state-of-the-art equipment.
-We estimate that the cost of LED manufacturing in these conditions has dropped from 1.11$(2020) in 2002 to 0.05$(2020) in 2020, a 95.5\% decrease.
+We estimate that the cost of LED manufacturing in these conditions has dropped from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
 The largest contribution to overall historical cost reductions came from increase in wafer size.

### 2021-02-11 SPIE/article/main.tex
incorporated comments from Sergey

@@
 
-    Within just 25 years of the introduction of first commercial white light-emitting diodes (LED), the solid state lighting industry has become a rare success story in the global drive for the transformation of energy systems.
+    Within just 25 years of the introduction of first commercial white light-emitting diodes (LED), the solid state lighting (SSL) industry has become a rare success story in the global drive for the transformation of energy systems.
 Today, the solid-state lighting markets in the US and the EU together are valued at 58.7 Bn.\$(2020) \cite{gvr2020market_us,gvr2020market_eu} with general illumination market penetration exceeding 50\% \cite{eu2019impactass,stratunl2018}.
@@
 \subsection{Choice of Metrics}
+\label{subsec:metrics}
 
-    To describe the effect of technological breakthroughs and manufacturing process improvements, a set of metrics had to be identified.
-The ensemble of metrics is required to describe all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
-They must further have the ability to directly describe the effect of single technological breakthroughs, evolution in device architecture and manufacturing process improvements on device performance.
+    To quantify the effect of technological breakthroughs and manufacturing process improvements, a set of metrics has to be identified that describes all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
+They must further have the ability to directly capture the effect of single technological breakthroughs, evolution in device architecture and manufacturing process improvements on device performance.
 
@@
         \end{center}
-        \caption{Historical increase in flux for highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
+        \caption{Historical increase in flux for the highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
 Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
@@
 Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
-Secondly, the total flux per device would only be a proxy for technological improvements in light-emitting diodes, if data was given for single light-emitting diode chips, instead of multi-chip packages.
+Secondly, the total flux per device would only be a proxy for technological improvements in light-emitting diodes if data was given for single light-emitting diode chips instead of multi-chip packages.
 Historically, publications have sometimes failed to make this distinction, listing datapoints for both device levels in the same graph without supporting information.
@@
         \caption{List of the device sub-efficiencies used in our methodology.
-We follow the definitions used by previous authors, such as Tsao et al.
-\cite{tsao2010solid} and Pattison et al.
-\cite{pattison2017solid}.
-The historical development of the sub-efficiencies is displayed in figure \ref{fig:efficiency}.}
+We follow the definitions used by previous authors  \cite{tsao2010solid,pattison2017solid}.
+Developments of the sub-efficiencies are displayed in figure \ref{fig:efficiency}.}
         \bigskip
@@
 
-    Performance improvements in metrics related to consumer experience have also played a role in the adoption of light-emitting diode based luminaires \cite{cowan2011understanding}.
+    Performance improvements in metrics related to consumer experience have also played a role in the adoption of LED-based luminaires \cite{cowan2011understanding}.
 Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
 For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources.
-The work on analyzing consumer experience metrics is ongoing and will be presented in our future work.
+The analysis of consumer experience metrics is ongoing and will be presented in our future work.
 Notably, we exclude flicker from this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
@@
 
-        To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in light-emitting diode manufacturing, we gathered data from a variety of sources.
-The aim was to collect historical data on overall device performance and historical development in sub-efficiencies.
-Where possible, we gathered  efficiency data from literature or computed the respective values from raw data.
+        To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in LED manufacturing, we gathered historical data on device cost and performance across different sub-efficiencies, as well as on technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
+The data was collected from a variety of sources such as patent literature, scientic articles, company publications, industry reports, and technology roadmaps and forecasts.
+Additional information was provided by experts from both the light-emitting diode manufacturing industry and the manufacturing equipment industry during semi-structured interviews.
 We further had to identify the associated device architectures, manufacturing processes and types of down conversion phosphors used.
-Data was extracted from patent literature, scientific publications, industry reports and technology forecasts.
-Additional information and data was provided by experts from both the light-emitting diode manufacturing industry and the manufacturing equipment industry during semi-structured interviews.
 
@@
 
-        For instance, to gather historical data on the reduction of losses associated with electrical droop, device performance data was calculated based on the information extracted from datasheets, as shown in figure \ref{fig:droop}.
+        For instance, to gather historical data on the reduction of losses associated with electrical droop, device performance data was calculated based on the information extracted from datasheets by methods in figure \ref{fig:droop}.
 This data was cross-referenced with information on the different chip architectures and manufacturing processes used for each device.
-This data was then augmented by data provided in industry reports and company publications \cite{osram2014osram}.
-Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gather a complete picture of the historical development of light-emitting diode efficiency.
+Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gain a complete picture of the historical developments in light-emitting diode efficiency.
 
@@
 We considered the classical p-side-up lateral current spreading architecture, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture.
-An excerpt of the device architectures considered is shown in figure \ref{fig:chip_arch}.
-The earliest commercial warm-white phosphor converted blue light emitting diodes were introduced in 2003, so the model was constructed for the years 2003, 2012 and 2020.
+An excerpt of the evolution of the classical chip architecture considered in the cost mode is shown in figure \ref{fig:chip_arch}.
+The earliest commercial warm-white phosphor converted blue light emitting diodes were introduced in 2002, so the model was constructed for the years 2002, 2012 and 2020.
 It was populated with equipment data from European and North American firms, selected for a virtual North American manufacturing location.
-Details of the manufacturing processes and process-specific step parameters were derived from the sources mentiond in section \ref{subsec:data}.
+Details of the manufacturing processes and process-specific step parameters were derived from the sources mentioned in section \ref{subsec:data}.
 For 2012, additional data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}.
-Our model includes both the wafer treatment process as well as the packaging process.
+Our model includes both the wafer treatment process with further ongoing work also focusing on the packaging process.
 While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach.
@@
 Years correspond to earliest identified patent priority date.
-Dashed boxes indicate chip designs not brought to large scale production.
-Adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
+The boxed illustration indicates a chip design not brought to large scale production.
+Drawings adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
             \label{fig:chip_arch}
@@
 
-        To dis-aggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak et al.
-in a cost model for photovoltaic modules \cite{kavlak2018evaluating}.
+        To disaggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak with colleagues at MIT in a cost model for photovoltaic modules \cite{kavlak2018evaluating}.
 It is based on the logarithmic derivative of the total differential of the cost function.
@@
 Figure \ref{fig:efficiency} additionally shows the physical limits for the loss channels.
-We note, that those loss channels with a fixed physical limit below 100\% have become significantly more dominant in 2016 and 2020.
+We find that those loss channels with a fixed physical limit below 100\% have become significantly more dominant in 2016 and 2020.
 For instance, Stokes loss, describing the energy dissipated upon conversion from short wavelength to long wavelength photons, becomes more dominant.
@@
 The overall lamp efficiency $\eta_L$ is displayed as the rightmost column.
-This figure takes as inputs the state-of-the-art sub-efficiencies discussed in section \ref{sec:methods}.
+This figure takes as inputs the state-of-the-art sub-efficiencies discussed in section \ref{subsec:metrics}.
 Horizontal colored bars give state-of-the-art sub-efficiencies for five years: \textcolor{blue}{1997}, \textcolor{teal}{2002}, \textcolor{orange}{2010}, \textcolor{magenta}{2016} and \textcolor{red}{2020}.
@@
 For instance, optical simulations could be performed to determine ways of modifying the device architecture to improve light extraction efficiency.
-On the other hand, efficiency improvements in sub-efficiencies with a limited physical understanding internal quantum efficiency and the associated electrical droop at high currents were mostly due to process improvements in the manufacturing process.
+On the other hand, efficiency improvements in sub-efficiencies with a limited physical understanding, such as internal quantum efficiency and the associated electrical droop at high currents, were mostly due to manufacturing process improvements.
 For instance, the reasons for electrical droop have, until recently, been a topic of hot debate, but significant improvements in this channel were still achieved by varying metal-organic vapour deposition (MOCVD) growth parameters in large ensembles of wafer batches.
@@
 
-    Using our cost model, we estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, to have decreased from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
+    Using our cost model, we estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, have declined from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
 
     A preliminary analysis of components of the cost model suggests that an increase in wafer size in the manufacturing process is responsible for the largest part of the cost reduction.
-The number of die per wafer associated with this is displayed in the inset plot of figure \ref{fig:cost}.
 The wafer diameter used in production has been steadily increasing since 2002.
-The model assumes diameters and associated number of die per wafer  $d(2002)=50$mm$\rightarrow851$ to $d(2020)=200$mm$\rightarrow26,838$.
+The model assumes diameters and associated number of die per wafer changing from $d(2002)=50$mm$\rightarrow851$ to $d(2020)=200$mm$\rightarrow26,838$.
 Note that the decrease in exclusion zone and cutting width also contribute to the increase in die per wafer.
-As the number of die per wafer increases, the packaging steps carry a larger share of the total cost.
+Analysis of the relative contributions of different manufacturing process steps to the cost structure, suggests that as the number of die per wafer increases, the packaging steps now carry the largest share of the total cost at 60\% in 2020.
 This is because these steps are limited by the throughput of the associated equipment.
-The wafer processing steps are limited by the number of die per wafer and only to a lesser extend the throughput of equipment.
+The wafer processing step, which was the largest contributor to the total cost in 2002, is defined by the number of die per wafer and only to a lesser extend the throughput of equipment.
 
@@
             \centering
-            \begin{tabularx}{\textwidth}{|l|l|X|X|X|}
+            \begin{tabularx}{\textwidth}{|l|l|X|X|X|X|X|l|}
             	\hline
-            		\textit{Year} & \textit{Unit} & \textit{Cost (Wafer Proc.)} & \textit{Cost (Packaging)} & \textit{Total Cost} \\
+            		\textit{Year} & \textit{Unit} & \textit{Substrate} & \textit{Epitaxy} & \textit{Wafer Proc.} & \textit{Packaging} & \textit{Phosphor} & \textit{Total} \\
                 \hline
-                    2002 & \$(2020) & 0.975 & 0.138 & 1.113 \\
+                    2002 & \$(2020) & 0.0760 & 0.1638 & 0.7351 & 0.1203 & 0.0175 & 1.1131 \\
                 \hline
-                    2002-2012 & \% & -62.05 & -47.10 & -90.12 \\
+                    2002-2012 & $\Delta$ \% & -98.8 & -94.7 & -96.2 & -47.8 & -42.2 & -90.1 \\
                 \hline
-                    2012 & \$(2020) & 0.037 & 0.073 & 0.110 \\
+                    2012 & \$(2020) & 0.0008 & 0.0086 & 0.0276 & 0.0627 & 0.0101 & 0.0111 \\
                 \hline
-                    2012-2020 & \% & -67.56 & -36.98 & -47.27 \\
+                    2012-2020 & $\Delta$ \% & -81.8 & -34.37 & -61.5 & -50.3 & -56.4 & -52.7 \\
                 \hline
-                    2020 & \$(2020) & 0.012 & 0.046 & 0.058 \\
+                    2020 & \$(2020) & 0.0001 & 0.0056 & 0.0106 & 0.0311 & 0.0044 & 0.0528 \\
                 \hline
@@
 
-    We also construct the manufacturing cost model for low-power and mid-power LED packages produced at a virtual US location using state-of-the-art equipment.
-We estimate that the cost of LED manufacturing in these conditions has dropped from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
+    We also constructed the manufacturing cost model for low-power and mid-power LED packages produced at a virtual US location using state-of-the-art equipment.
+We estimated that the cost of LED manufacturing in these conditions dropped from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
 The largest contribution to overall historical cost reductions came from increase in wafer size.
-The largest remaining contributor to the manufacturing cost is the LED packaging, contributing 60\% to the cost structure in 2020.
+The largest remaining contributor to the manufacturing cost is the LED packaging, accounting for 60\% of the cost structure in 2020.
 

### 2021-02-12 SPIE/article/main.tex
added corrections by Sergey

@@
 We find that technological breakthroughs and process innovation contributed to performance improvements across the entire range of LED device sub-efficiencies, resulting in the overall increase in the lamp efficiency for the highest performing devices at a test current of 350mA from 5.8\% to 38.7\% between 2002 and 2020.
-We further develop a bottom-up manufacturing cost model with process-step resolution that captured improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale, to analyse progress in LED manufacturing cost structure between early manufacturing in 2003 and 2020.
+We further develop a bottom-up manufacturing cost model with process-step resolution that captures improvements in throughput, yield and related costs of all relevant manufacturing steps, as well as economies of scale, to analyse progress in LED manufacturing cost structure between early manufacturing in 2002 and mature industry in 2020.
 We estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location using state-of-the-art equipment has dropped from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
-We also find that the largest contribution to overall cost reductions has come from increase in wafer size, and that in 2020, chip packaging remains the largest contributor to cost at 60\%.
+We also find that the largest contribution to overall cost reduction has come from increase in wafer size, and that in 2020, LED chip packaging is the largest contributor to cost at 60\%.
 
@@
 Today, the solid-state lighting markets in the US and the EU together are valued at 58.7 Bn.\$(2020) \cite{gvr2020market_us,gvr2020market_eu} with general illumination market penetration exceeding 50\% \cite{eu2019impactass,stratunl2018}.
-Estimates for the electrical energy saved annually from SSL adoption ranges from 131 TWh/year in 2020 for the EU \cite{eu2019impactass} to 442 TWh/year in 2020 for the US \cite{yamada2015adoption,guidehouse2020adoption}, which is on par with the amount of energy produced by all deployed solar photovoltaic installations in these regions.
-This revolution in lighting would not have been possible without dramatic improvements in the overall device efficiency and reductions in manufacturing costs.
+Estimates for the electrical energy saved annually from SSL adoption range from 131 TWh/year in 2020 for the EU \cite{eu2019impactass} to 442 TWh/year in 2020 for the US \cite{yamada2015adoption,guidehouse2020adoption}, which is on par with the amount of energy produced by all deployed solar photovoltaic installations in these regions.
+This revolution in lighting would not have been possible without dramatic improvements in the overall LED device efficiency and reductions in LED manufacturing costs.
 
-    Understanding the sources and impact of these dramatic improvements on LED manufacturing is essential for researchers, industry professionals, as well as policymakers in the sphere of energy, science and innovation, as they may provide valuable insights for the acceleration of innovation in other demand-side energy technologies.
+    Understanding the sources and impact of these improvements on LED manufacturing is essential for researchers, industry professionals, as well as policymakers in the sphere of energy, science and innovation, as they may provide valuable insights for the acceleration of innovation in other demand-side energy technologies.
 With this aim, in this paper we identify and quantify the principal sources of performance improvements and cost reductions in phosphor-converted white LED lighting devices over the last 20 years.
@@
     To quantify the effect of technological breakthroughs and manufacturing process improvements, a set of metrics has to be identified that describes all dimensions of device performance, including the overall physical device efficiency and the sub-efficiencies related to different physical loss channels.
-They must further have the ability to directly capture the effect of single technological breakthroughs, evolution in device architecture and manufacturing process improvements on device performance.
+They must further have the ability to directly capture the effect of individual technological breakthroughs, evolution in device architecture, and manufacturing process improvements on device performance.
 
@@
         \end{center}
-        \caption{Historical increase in flux for the highest performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
-Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
-Note the logarithmic ordinates and the black colored datapoints corresponding to the left ordinate, blue colored datapoints corresponding to the right ordinate.}
+        \caption{Historical increase in flux for the highest-performing white light-emitting diode chips and (multi-chip) packages, inspired by \textit{"Haitz's Law"}\cite{haitz1999case}.
+Shown are the datapoints for best commercial performers gathered from press releases, datasheets and industry periodicals.
+Note the logarithmic ordinates, black-colored datapoints corresponding to the left ordinate, and blue-colored datapoints corresponding to the right ordinate.}
         \label{fig:haitz}
@@
 
-    Firstly, the metric today retains only limited significance as an appropriate for technological development in light-emitting diodes.
+    Firstly, today, the metric retains only limited significance as an appropriate proxy for technological progress in light-emitting diodes.
 This is because it is not desirable to increase the total flux per device beyond a certain point in many applications.
-Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010} and economical considerations where multiple LED die in a single package can achieve the same brightness as a single high-brightness LED die.
+Reasons for limiting the total flux per device may include lighting design considerations to reduce glare \cite{khan2015led}, device efficiency considerations to avoid electrical droop at high operating currents associated with high brightness \cite{Piprek2010}, and economical considerations that multiple LED chips in a single package can achieve the same brightness as a single high-brightness LED die.
 Secondly, the total flux per device would only be a proxy for technological improvements in light-emitting diodes if data was given for single light-emitting diode chips instead of multi-chip packages.
@@
     Figure \ref{fig:haitz} shows an updated and expanded overview of the flux per device and the cost per flux of highest performing light-emitting diodes, both at the chip and package level, to demonstrate the limitations of these metrics.
-It is evident that the historical improvement in total flux per package for single chips is not as pronounced as for multi-chip packages.
+It is evident that the historical improvements in total flux per package for single chips are not as pronounced as for multi-chip packages.
 
@@
 Sub-efficiencies directly capture the effect of particular technological breakthroughs in device design and manufacturing process improvements.
-Table \ref{tab:eff} lists the relevant sub-efficiencies considered along with mathematical definitions.
+Table \ref{tab:eff} lists the relevant sub-efficiencies considered in our work, along with their mathematical definitions.
 
         \begin{table}[h!]
-        \caption{List of the device sub-efficiencies used in our methodology.
-We follow the definitions used by previous authors  \cite{tsao2010solid,pattison2017solid}.
-Developments of the sub-efficiencies are displayed in figure \ref{fig:efficiency}.}
+        \caption{List of LED device sub-efficiencies used in our methodology.
+We follow the definitions used by Tsao et al.
+\cite{tsao2010solid} and Pattison et al.
+\cite{pattison2017solid}.
+Historical developments in the sub-efficiencies are displayed in figure \ref{fig:efficiency}.}
         \bigskip
@@
     Performance improvements in metrics related to consumer experience have also played a role in the adoption of LED-based luminaires \cite{cowan2011understanding}.
-Broadly described as the quality of light, these include metrics related to the emitted spectrum as well flicker, the temporal modulation of light.
-For instance, breakthroughs in the development of downconversion phosphors enabled a greater range in the color temperature of light sources.
-The analysis of consumer experience metrics is ongoing and will be presented in our future work.
-Notably, we exclude flicker from this context because it is not an inherent property of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
+Broadly described as the quality of light, these include metrics related to the emitted spectrum, as well as flicker, the temporal modulation of light.
+For instance, breakthroughs in the development of down-conversion phosphors enabled a greater range in the color temperature of light sources.
+The analysis of consumer experience metrics is ongoing and will be presented in our future research.
+Notably, we exclude flicker from this work because it is an inherent property not of the light-emitting diodes themselves, but rather the electrical ballasts in the luminaires.
 For further discussion of flicker in this context, see a recent publication by Weinold \cite{weinold2020long}.
@@
         To gain a detailed understanding of the sources and magnitude of efficiency improvements and cost reductions in LED manufacturing, we gathered historical data on device cost and performance across different sub-efficiencies, as well as on technological breakthroughs and manufacturing innovation for phosphor-converted white light-emitting diodes for the past 20 years.
-The data was collected from a variety of sources such as patent literature, scientic articles, company publications, industry reports, and technology roadmaps and forecasts.
-Additional information was provided by experts from both the light-emitting diode manufacturing industry and the manufacturing equipment industry during semi-structured interviews.
-We further had to identify the associated device architectures, manufacturing processes and types of down conversion phosphors used.
+The data was collected from a variety of sources such as patent literature, scientific articles, company publications, industry reports and roadmaps, and technical periodicals.
+Additional information was provided by experts from academia, the LED manufacturing industry, and the manufacturing equipment industry during semi-structured interviews.
+We further had to identify the associated device architectures, manufacturing processes and types of down-conversion phosphors used.
 
@@
 
-        For instance, to gather historical data on the reduction of losses associated with electrical droop, device performance data was calculated based on the information extracted from datasheets by methods in figure \ref{fig:droop}.
+        In cases where efficiency data was not readily available, we calculated the respective values from raw performance data.
+For instance, to calculate historical data on the reduction of losses associated with electrical droop, we used device performance data extracted from datasheets by the method shown in figure \ref{fig:droop}.
 This data was cross-referenced with information on the different chip architectures and manufacturing processes used for each device.
-Data was gathered in a similar way for all sub-efficiencies listed in table \ref{tab:eff}, to gain a complete picture of the historical developments in light-emitting diode efficiency.
+In a similar way, data was gathered for all sub-efficiencies listed in table \ref{tab:eff} to gain a complete picture of the historical developments in light-emitting diode efficiency.
 
     \subsection{Manufacturing Cost Model}
+    \label{subsec:costmodel}
 
-        To quantify changes in the manufacturing cost of devices, a bottom-up manufacturing cost model with process step resolution was constructed.
+        To quantify changes in the manufacturing cost of LED devices, a bottom-up manufacturing cost model with process step resolution was constructed.
 It covers the entire manufacturing process of GaN-on-sapphire-based phosphor-converted low-to-mid power light-emitting diode packages of different chip architectures.
-We considered the classical p-side-up lateral current spreading architecture, as well as a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture.
-An excerpt of the evolution of the classical chip architecture considered in the cost mode is shown in figure \ref{fig:chip_arch}.
+We designed it to accommodate the classical p-side-up lateral current spreading architecture, with two further architectures\textemdash a packaged flip-chip vertical current spreading architecture and a chip-scale package flip chip architecture\textemdash under development.
+An excerpt of the evolution of the classical chip architecture considered in the cost model is shown in figure \ref{fig:chip_arch}.
 The earliest commercial warm-white phosphor converted blue light emitting diodes were introduced in 2002, so the model was constructed for the years 2002, 2012 and 2020.
@@
 For 2012, additional data for the model was adapted from the \textit{LEDCOM} cost model prepared for the US Department of Energy by Stephen Bland of SB Consulting\cite{ledcomv2}.
-Our model includes both the wafer treatment process with further ongoing work also focusing on the packaging process.
-While the model offers great flexibility to researchers in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach.
-The main aim of the model is not to faithfully represent real-world manufacturing such as those in present-day manufacturing locations in Asia, but rather to show the effect of technological change and manufacturing process improvements on total cost.
+Our model includes the wafer treatment process with further ongoing work also focusing on the chip packaging process.
+While the model offers great flexibility in adapting the manufacturing process parameters and chip architectures used, it is important to note the limitations of this approach.
+The main aim of the model is not to faithfully represent real-world manufacturing such as in present-day manufacturing locations in Asia, but rather to show the effect of technological change and manufacturing process improvements on total cost.
 
@@
 Years correspond to earliest identified patent priority date.
-The boxed illustration indicates a chip design not brought to large scale production.
+Illustration in the box indicates a chip design not brought to large scale production.
 Drawings adapted from patents \cite{nagahama2013nitride,tanaka2010semiconductor,wierer2006photonic}}
@@
 
-        To disaggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak with colleagues at MIT in a cost model for photovoltaic modules \cite{kavlak2018evaluating}.
+        To disaggregate the contribution of changes in single variables to changes in the total manufacturing cost, we used an approach introduced by Kavlak and colleagues in a cost model for photovoltaic modules \cite{kavlak2018evaluating}.
 It is based on the logarithmic derivative of the total differential of the cost function.
@@
 
-     Results for performance improvements are presented in Figure \ref{fig:efficiency}.
+     Main results for performance improvements are presented in Figure \ref{fig:efficiency}.
 Overall efficiency in best performing devices improved from $\eta_L=5.8\%$ in 2002 to $\eta_L = 38.7\%$ in 2020.
@@
 We find that those loss channels with a fixed physical limit below 100\% have become significantly more dominant in 2016 and 2020.
-For instance, Stokes loss, describing the energy dissipated upon conversion from short wavelength to long wavelength photons, becomes more dominant.
+An example of this tendency is the increased importance of Stokes loss that describes the energy dissipated upon conversion from short wavelength to long wavelength photons.
 Notably, sub-efficiencies for the most current devices are only $\sim10\%$ below the physical limit.
-The exception is spectral efficiency, which at $\sim20\%$ below the physical limit shows larger potential for improvement.
+The exception is spectral efficiency, which at $\sim17\%$ below the physical limit shows larger potential for improvement.
 
      Our comparison between the improvements in device sub-efficiencies between 2002 and 2020 shows that the aggregate efficiency improvement was not driven primarily by improvements in a single sub-efficiency.
-Instead, there has been consistent progress across the ensemble of loss channels corresponding to the device sub-efficiencies in the past 18 years: forward voltage efficiency ($70\%\rightarrow99.5\%$), internal quantum efficiency ($55\%\rightarrow90\%)$, electrical droop ($65\%\rightarrow90\%$), light-extraction efficiency ($60\%\rightarrow90\%$) and spectral efficiency ($74\% \rightarrow83\%$).
+Instead, there has been consistent progress across the ensemble of loss channels corresponding to the device sub-efficiencies in the past 18 years, including forward voltage efficiency ($70\%\rightarrow99.5\%$), internal quantum efficiency ($55\%\rightarrow90\%)$, electrical droop ($65\%\rightarrow90\%$), light-extraction efficiency ($60\%\rightarrow90\%$) and spectral efficiency ($74\% \rightarrow83\%$).
 
@@
         \end{center}
-        \caption{Impact of technology breakthroughs and manufacturing process improvements on historical developments in sub-efficiencies of phosphor-converted warm white light-emitting diodes with test currents of at least $I_\text{test}=350$mA.
+        \caption{Historical changes in sub-efficiencies of phosphor-converted warm white light-emitting diodes with test currents of at least $I_\text{test}=350$mA.
 The overall lamp efficiency $\eta_L$ is displayed as the rightmost column.
@@
 Physical limits are indicated by black horizontal bars.
-The possible range for the physical limit of V$_\text{f}$E exceeds 100\% and depends on electrical device parameters, which are discussed in \cite{david2016electrical}.
-The range of this limit is indicated by an upward pointing black arrow.
+Possible range for the physical limit of V$_\text{f}$E exceeds 100\%, depends on electrical device parameters, and is indicated by an upward pointing black arrow.
 Efficiency acronyms are listed in table \ref{tab:eff}.}
@@
 
-    Notably, efficiency improvements in sub-efficiencies with a good understanding of the underlying physical loss channels, such as light-extraction efficiency, forward voltage efficiency and spectral efficiency were mostly a result of targeted research and development.
+    Notably, improvements in sub-efficiencies with a good understanding of the underlying physical loss channels, such as light-extraction efficiency, forward voltage efficiency and spectral efficiency, were mostly a result of targeted research and development.
 For instance, optical simulations could be performed to determine ways of modifying the device architecture to improve light extraction efficiency.
-On the other hand, efficiency improvements in sub-efficiencies with a limited physical understanding, such as internal quantum efficiency and the associated electrical droop at high currents, were mostly due to manufacturing process improvements.
-For instance, the reasons for electrical droop have, until recently, been a topic of hot debate, but significant improvements in this channel were still achieved by varying metal-organic vapour deposition (MOCVD) growth parameters in large ensembles of wafer batches.
+On the other hand, efficiency improvements in sub-efficiencies with a relatively limited understanding of underlying physical mechanisms, such as internal quantum efficiency and the associated electrical droop at high currents, were mostly due to manufacturing process improvements.
+For instance, the reasons for electrical droop have, until recently, been a topic of hot debate.
+Despite that, significant improvements in this channel were still achieved by varying metal-organic vapour deposition (MOCVD) growth parameters in large ensembles of wafer batches.
 
@@
 
-    Using our cost model, we estimate that the cost of manufacturing low-power and mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, have declined from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
+    Results of manufacturing cost modeling for three reference years (2002, 2012, and 2020), broken down by five main cost components in our model\textemdash substrate, epitaxy, wafer processing, chip packaging, and phosphor\textemdash are presented in table \ref{tab:cost}.
+Using our cost model, we estimate that the cost of manufacturing low-to-mid-power light-emitting diode packages at a US location, using state-of-the-art equipment, have declined from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% overall cost reduction.
 
-    A preliminary analysis of components of the cost model suggests that an increase in wafer size in the manufacturing process is responsible for the largest part of the cost reduction.
-The wafer diameter used in production has been steadily increasing since 2002.
+    A preliminary analysis of the cost model components suggests that an increase in the wafer size is responsible for the largest part of the overall cost reduction.
+The wafer diameter $d$ used in production has been steadily increasing since 2002.
 The model assumes diameters and associated number of die per wafer changing from $d(2002)=50$mm$\rightarrow851$ to $d(2020)=200$mm$\rightarrow26,838$.
-Note that the decrease in exclusion zone and cutting width also contribute to the increase in die per wafer.
-Analysis of the relative contributions of different manufacturing process steps to the cost structure, suggests that as the number of die per wafer increases, the packaging steps now carry the largest share of the total cost at 60\% in 2020.
-This is because these steps are limited by the throughput of the associated equipment.
-The wafer processing step, which was the largest contributor to the total cost in 2002, is defined by the number of die per wafer and only to a lesser extend the throughput of equipment.
+Related decreases in exclusion zone and cutting width have also contributed to the increase in die per wafer number.
+Further analysis of the relative contributions of different manufacturing process steps to the cost structure suggests that as the number of die per wafer increases, the packaging steps now carry the largest share of the total cost at 60\% in 2020.
+This is because the output of these steps is limited by the throughput of the associated equipment.
+The wafer processing steps, which made the largest contribution to the total cost in 2002, depend on the number of die per wafer and only to a lesser extent on the throughput of equipment, which explains why they became relatively less important in the cost structure by 2020.
 
     \begin{table}[h!]
-        \caption{Manufacturing cost of low-power classical chips (lateral current spreading) estimated with the cost model introduced in section \ref{sec:methods}.
-Relative contributions of different manufacturing process steps are shown in figure \ref{fig:cost}.}
+        \caption{Manufacturing cost of low-power classical chips (lateral current spreading) estimated with the cost model introduced in section \ref{subsec:costmodel}.}
         \bigskip
@@
 
-    In this study, we found that technological breakthroughs and process innovation contributed to performance improvements across the entire range of LED device sub-efficiencies, resulting in the overall lamp efficiency increase for the highest performing devices at a test current of 350mA from 5.8\% in 2002 to 38.7\% in 2020.
-Notably, we found that efficiency improvements in loss channels with a good underlying understanding of physical mechanism were driven mostly by research and development, while improvements in channels with a lack of such understanding required extensive experimentation and learning-by-doing in the manufacturing setting.
+    In this study, we found that technological breakthroughs and manufacturing process innovation contributed to performance improvements across the entire range of LED device sub-efficiencies, resulting in the overall lamp efficiency increase for the highest performing devices at a test current of 350mA from 5.8\% in 2002 to 38.7\% in 2020.
+Notably, we found that efficiency improvements in loss channels with a good understanding of underlying physical mechanisms were driven mostly by research and development, while improvements in channels with a lack of such understanding required extensive experimentation and learning-by-doing in the manufacturing setting.
 It shows the importance of further research into the underlying solid-state physics of light-emitting diodes.
-A deeper understanding of the underlying effects and mechanisms associated with device loss channels will potentially enable further advances in LED technology, with spectral efficiency showing the largest potential for further improvements.
+A deeper understanding of the effects and mechanisms associated with device loss channels will potentially enable further advances in LED technology, with spectral efficiency in particular showing promise for further improvements.
 
-    We also constructed the manufacturing cost model for low-power and mid-power LED packages produced at a virtual US location using state-of-the-art equipment.
-We estimated that the cost of LED manufacturing in these conditions dropped from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020, a 95.5\% decrease.
-The largest contribution to overall historical cost reductions came from increase in wafer size.
-The largest remaining contributor to the manufacturing cost is the LED packaging, accounting for 60\% of the cost structure in 2020.
+    We also constructed the manufacturing cost model for low-to-mid-power LED packages produced at a virtual US location using state-of-the-art equipment.
+We estimated that the cost of LED manufacturing in these conditions decreased by 95.5\% from 1.11\$(2020) in 2002 to 0.05\$(2020) in 2020.
+The largest contribution to the overall cost reductions came from increase in the wafer size, while the largest remaining contributor to the manufacturing cost is the LED packaging, which accounts for 60\% of the cost structure in 2020.
 
@@
 
-We express our gratitude to all inteviewees for their willingness to participate in this study and invaluable contributions.
+We express our deep gratitude to all interviewees for their willingness to participate in this study and invaluable contributions.
 This research is funded by the \textit{Alfred P.
-Sloan Foundation} of New York, USA, grant number 253128.
+Sloan Foundation}, grant number 253128.
 Michael Weinold further gratefully acknowledges support from the \textit{Swiss Study Foundation} of Zurich, Switzerland.

### 2021-02-12 SPIE/article/main.tex
final version submitted to SPIE

@@
 \author[a,b]{Michael Weinold}
-\author[b]{Sergey Kolesnikov}
-\author[b,c]{Laura Diaz Anadon}
-\affil[a]{Chair of Entrepreneurial Risks, ETH Zurich, Scheuchzerstrasse 7, CH-8092 Zurich, CH}
-\affil[b]{Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, CB3 9EP, UK}
+\author[a]{Sergey Kolesnikov}
+\author[a,c]{Laura Diaz Anadon}
+\affil[a]{Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, CB3 9EP, UK}
+\affil[b]{Chair of Entrepreneurial Risks, ETH Zurich, Scheuchzerstrasse 7, CH-8092 Zurich, CH}
 \affil[c]{Belfer Center for Science and International Affairs, Harvard Kennedy School, Harvard University, Cambridge, MA 02138, USA}
@@
 
-We express our deep gratitude to all interviewees for their willingness to participate in this study and invaluable contributions.
+We would like to thank Gabriel Chan, Anna Goldstein and Venkatesh Narayanamurti for many helpful discussions and their feedback.
+We also express our deep gratitude to all interviewees for their willingness to participate in this study and invaluable contributions.
 This research is funded by the \textit{Alfred P.

### 2021-02-18 IEEE/article/contribution_variables.tex
added Contribution of Variables paper to IEEE folder

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}

\begin{document}

In their 2018 paper on solar photovoltaics (10.1016/j.enpol.2018.08.015; P.9 Main Body,
P.1 in Supplementary Material), the authors have a cost function C which describes
the cost associated with manufacturing one solar cell at a certain point in time. This
function depends on manufacturing variables x, y, which change over time (e.g. price
of silicon, price of chemicals, etc.). Variables are known only at discrete points in time
(t0, t1).

\begin{equation}
    C(x(t),y(t))
\end{equation}

They want to determine the contribution of a single variable x to the total change of
the cost function between two points in time $\Delta C(t0, t1)$ (eg. X\% of the cost reductions
in manufacturing solar cells in 2000 vs. 2020 are due to a change in variable x).
They start by writing out the differential of the cost function C as

\begin{equation}
    dC(x(t),y(t)) = \frac{\partial C}{\partial x} \frac{\text{d}y}{\text{d}t} \text{d}t + \frac{\partial C}{\partial y} \frac{\text{d}y}{\text{d}t} \text{d}t
\end{equation}

A sample cost model equation can be written as

\begin{equation}
    \Delta C_x = \int_{t=t_0}^{t_1} \frac{\partial C}{\partial x} \frac{\text{d}x}{\text{d}t} \text{d}t
\end{equation}

and then for C(t) assume a constant $C(t) \approx \tilde{C}$, which is approximately chosen to be $\tilde{C}=\frac{\Delta \tilde{C}}{\Delta \ln \tilde{C}}$ such that $\Delta C_x+\Delta C_y=\Delta C$

where the contribution of the change in variable x over time $t_0 < t < t_1$ is then

Here they say
If it were possible to observe the (...) variables x in continuous time, (...) [this equation]
would provide all that is needed to compute the contribution of each variable x.
Using logarithmic differentiation, they go on to rewrite the expression as

\begin{equation}
    \Delta C_x = \int_{t=t_0}^{t_1} C(t) \frac{\partial \ln C}{\partial x} \frac{\text{d}x}{\text{d}t} \text{d}t
\end{equation}

Questions:
Given a function
\begin{equation}
    C(x(t), y(t)) = x * y
\end{equation}

and discrete data for variables x(t), y(t) at the points t0, t1, what is the contribution
(total or percentage) of the change in variable x to the change $\Delta_C$

in the function C?
The authors of the paper assume that given continuous variable functions x(t), y(t), equation
(1) would suffice to calculate the contributions. But even if the time dependence
of variables were known (eg. daily data on the price of silicon, etc.), then integrating
would not yield what the authors are actually looking for.
They are interested in the contribution of single variables to the total change in C (eg.
what percentage of total manufacturing cost reductions are due to decrease in silicon
price). But integrating using

\begin{equation}
    \Delta C_x = \int_{t=t_0}^{t_1} \frac{\partial C}{\partial x} \frac{\text{d}x}{\text{d}t}
\end{equation}

is dependent on the path of curves x(t), y(t). This would yield different results for different
time dependency of variables. A variable x(t) would yield a different Cx than a
variable x'(t), which is not what the authors seek to describe.

%
\begin{equation}
    C = \frac{1}{DY}(ab+cd)
\end{equation}
%
where
%
\begin{align*}
    a &\dots \text{step 1 variable 1} \\
    b &\dots \text{step 1 variable 2} \\
    c &\dots \text{step 2 variable 1} \\
    d &\dots \text{step 2 variable 2} \\
    D &\dots \text{global variable 1} \\
    Y &\dots \text{global variable 2}
\end{align*}
%

\end{document}

### 2021-03-16 SSL Paper 3 (XXX)/article/main.tex
cleaned up paper 3 file structure

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%This is the LaTeX ARTICLE template for RSC journals
%Copyright The Royal Society of Chemistry 2016
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[twoside,twocolumn,9pt]{article}
\usepackage{extsizes}
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage[version=3]{mhchem}
\usepackage[left=1.5cm, right=1.5cm, top=1.785cm, bottom=2.0cm]{geometry}
\usepackage{balance}
\usepackage{mathptmx}
\usepackage{sectsty}
\usepackage{graphicx} 
\usepackage{lastpage}
\usepackage[format=plain,justification=justified,singlelinecheck=false,font={stretch=1.125,small,sf},labelfont=bf,labelsep=space]{caption}
\usepackage{float}
\usepackage{fancyhdr}
\usepackage{fnpos}
\usepackage[english]{babel}
\addto{\captionsenglish}{%
  \renewcommand{\refname}{Notes and references}
}
\usepackage{array}
\usepackage{droidsans}
\usepackage{charter}
\usepackage[T1]{fontenc}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{setspace}
\usepackage[compact]{titlesec}
\usepackage{hyperref}
%%%Please don't disable any packages in the preamble, as this may cause the template to display incorrectly.%%%

\usepackage{epstopdf}%This line makes .eps figures into .pdf - please comment out if not required.

\definecolor{cream}{RGB}{222,217,201}

\begin{document}

\pagestyle{fancy}
\thispagestyle{plain}
\fancypagestyle{plain}{
%%%HEADER%%%
\renewcommand{\headrulewidth}{0pt}
}
%%%END OF HEADER%%%

%%%PAGE SETUP - Please do not change any commands within this section%%%
\makeFNbottom
\makeatletter
\renewcommand\LARGE{\@setfontsize\LARGE{15pt}{17}}
\renewcommand\Large{\@setfontsize\Large{12pt}{14}}
\renewcommand\large{\@setfontsize\large{10pt}{12}}
\renewcommand\footnotesize{\@setfontsize\footnotesize{7pt}{10}}
\makeatother

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand\footnoterule{\vspace*{1pt}% 
\color{cream}\hrule width 3.5in height 0.4pt \color{black}\vspace*{5pt}} 
\setcounter{secnumdepth}{5}

\makeatletter 
\renewcommand\@biblabel[1]{#1}            
\renewcommand\@makefntext[1]% 
{\noindent\makebox[0pt][r]{\@thefnmark\,}#1}
\makeatother 
\renewcommand{\figurename}{\small{Fig.}~}
\sectionfont{\sffamily\Large}
\subsectionfont{\normalsize}
\subsubsectionfont{\bf}
\setstretch{1.125} %In particular, please do not alter this line.
\setlength{\skip\footins}{0.8cm}
\setlength{\footnotesep}{0.25cm}
\setlength{\jot}{10pt}
\titlespacing*{\section}{0pt}{4pt}{4pt}
\titlespacing*{\subsection}{0pt}{15pt}{1pt}
%%%END OF PAGE SETUP%%%

%%%FOOTER%%%
\fancyfoot{}
\fancyfoot[LO,RE]{\vspace{-7.1pt}\includegraphics[height=9pt]{head_foot/LF}}
\fancyfoot[CO]{\vspace{-7.1pt}\hspace{13.2cm}\includegraphics{head_foot/RF}}
\fancyfoot[CE]{\vspace{-7.2pt}\hspace{-14.2cm}\includegraphics{head_foot/RF}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{0pt} 
\renewcommand{\footrulewidth}{0pt}
\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}
%%%END OF FOOTER%%%

%%%FIGURE SETUP - please do not change any commands within this section%%%
\makeatletter 
\newlength{\figrulesep} 
\setlength{\figrulesep}{0.5\textfloatsep}

\newcommand{\topfigrule}{\vspace*{-1pt}% 
\noindent{\color{cream}\rule[-\figrulesep]{\columnwidth}{1.5pt}} }

\newcommand{\botfigrule}{\vspace*{-2pt}% 
\noindent{\color{cream}\rule[\figrulesep]{\columnwidth}{1.5pt}} }

\newcommand{\dblfigrule}{\vspace*{-1pt}% 
\noindent{\color{cream}\rule[-\figrulesep]{\textwidth}{1.5pt}} }

\makeatother
%%%END OF FIGURE SETUP%%%

%%%TITLE, AUTHORS AND ABSTRACT%%%
\twocolumn[
  \begin{@twocolumnfalse}
{\includegraphics[height=30pt]{head_foot/journal_name}\hfill\raisebox{0pt}[0pt][0pt]{\includegraphics[height=55pt]{head_foot/RSC_LOGO_CMYK}}\\[1ex]
\includegraphics[width=18.5cm]{head_foot/header_bar}}\par
\vspace{1em}
\sffamily
\begin{tabular}{m{4.5cm} p{13.5cm} }

\includegraphics{head_foot/DOI} & \noindent\LARGE{\textbf{This is the title$^\dag$}} \\%Article title goes here instead of the text "This is the title"
\vspace{0.3cm} & \vspace{0.3cm} \\

 & \noindent\large{Full Name,$^{\ast}$\textit{$^{a}$} Full Name,\textit{$^{b\ddag}$} and Full Name\textit{$^{a}$}} \\%Author names go here instead of "Full name", etc.

\includegraphics{head_foot/dates} & \noindent\normalsize{The abstract should be a single paragraph which summarises the content of the article. Any references in the abstract should be written out in full \textit{e.g.}\ [Surname \textit{et al., Journal Title}, 2000, \textbf{35}, 3523].} \\%The abstrast goes here instead of the text "The abstract should be..."

\end{tabular}

 \end{@twocolumnfalse} \vspace{0.6cm}

  ]
%%%END OF TITLE, AUTHORS AND ABSTRACT%%%

%%%FONT SETUP - please do not change any commands within this section
\renewcommand*\rmdefault{bch}\normalfont\upshape
\rmfamily
\section*{}
\vspace{-1cm}

%%%FOOTNOTES%%%

\footnotetext{\textit{$^{a}$~Address, Address, Town, Country. Fax: XX XXXX XXXX; Tel: XX XXXX XXXX; E-mail: xxxx@aaa.bbb.ccc}}
\footnotetext{\textit{$^{b}$~Address, Address, Town, Country. }}

%Please use \dag to cite the ESI in the main text of the article.
%If you article does not have ESI please remove the the \dag symbol from the title and the footnotetext below.
\footnotetext{\dag~Electronic Supplementary Information (ESI) available: [details of any supplementary information available should be included here]. See DOI: 00.0000/00000000.}
%additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required

\footnotetext{\ddag~Additional footnotes to the title and authors can be included \textit{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P. Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}

%%%END OF FOOTNOTES%%%

%%%MAIN TEXT%%%%
The main text of the article\cite{Mena2000} should appear here.

\subsection{This is the subsection heading style}
Section headings can be typeset with and without numbers.\cite{Abernethy2003}

\subsubsection{This is the subsubsection style.~~} These headings should end in a full point.

\paragraph{This is the next level heading.~~} For this level please use \texttt{\textbackslash paragraph}. These headings should also end in a full point.

\section{Graphics and tables}
\subsection{Graphics}
Graphics should be inserted on the page where they are first mentioned (unless they are equations, which appear in the flow of the text).\cite{Cotton1999}

\begin{figure}[h]
\centering
  \includegraphics[height=3cm]{example1}
  \caption{An example figure caption.}
  \label{fgr:example}
\end{figure}

\begin{figure*}
 \centering
 \includegraphics[height=3cm]{example2}
 \caption{A two-column figure.}
 \label{fgr:example2col}
\end{figure*}

\subsection{Tables}
Tables typeset in RSC house style do not include vertical lines. Table footnote symbols are lower-case italic letters and are typeset at the bottom of the table. Table captions do not end in a full point.\cite{Arduengo1992,Eisenstein2005}

\begin{table}[h]
\small
  \caption{\ An example of a caption to accompany a table}
  \label{tbl:example}
  \begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}}lll}
    \hline
    Header one (units) & Header two & Header three \\
    \hline
    1 & 2 & 3 \\
    4 & 5 & 6 \\
    7 & 8 & 9 \\
    10 & 11 & 12 \\
    \hline
  \end{tabular*}
\end{table}

Adding notes to tables can be complicated.  Perhaps the easiest method is to generate these manually.\footnote[4]{Footnotes should appear here. These might include comments relevant to but not central to the matter under discussion, limited experimental and spectral data, and crystallographic data.}

\begin{table*}
\small
  \caption{\ An example of a caption to accompany a table \textendash\ table captions do not end in a full point}
  \label{tbl:example}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lllllll}
    \hline
    Header one & Header two & Header three & Header four & Header five & Header six  & Header seven\\
    \hline
    1 & 2 & 3 & 4 & 5 & 6  & 7\\
    8 & 9 & 10 & 11 & 12 & 13 & 14 \\
    15 & 16 & 17 & 18 & 19 & 20 & 21\\
    \hline
  \end{tabular*}
\end{table*}

\section{Equations}

Equations can be typeset inline \textit{e.g.}\ $ y = mx + c$ or displayed with and without numbers:

 \[ A = \pi r^2 \]

\begin{equation}
  \frac{\gamma}{\epsilon x} r^2 = 2r
\end{equation}

You can also put lists into the text. You can have bulleted or numbered lists of almost any kind. 
The \texttt{mhchem} package can also be used so that formulae are easy to input: \texttt{\textbackslash ce\{H2SO4\}} gives \ce{H2SO4}.

For footnotes in the main text of the article please number the footnotes to avoid duplicate symbols. \textit{e.g.}\ \texttt{\textbackslash footnote[num]\{your text\}}. The corresponding author $\ast$ counts as footnote 1, ESI as footnote 2, \textit{e.g.}\ if there is no ESI, please start at [num]=[2], if ESI is cited in the title please start at [num]=[3] \textit{etc.} Please also cite the ESI within the main body of the text using \dag.

\section{Conclusions}
The conclusions section should come in this section at the end of the article, before the Conflicts of interest statement.

\section*{Conflicts of interest}
In accordance with our policy on \href{http://www.rsc.org/journals-books-databases/journal-authors-reviewers/author-responsibilities/#code-of-conduct}{Conflicts of interest} please ensure that a conflicts of interest statement is included in your manuscript here.  Please note that this statement is required for all submitted manuscripts.  If no conflicts exist, please state that ``There are no conflicts to declare''.

\section*{Acknowledgements}
The Acknowledgements come at the end of an article after Conflicts of interest and before the Notes and references.

%%%END OF MAIN TEXT%%%

%The \balance command can be used to balance the columns on the final page if desired. It should be placed anywhere within the first column of the last page.

\balance

%If notes are included in your references you can change the title from 'References' to 'Notes and references' using the following command:
%\renewcommand\refname{Notes and references}

%%%REFERENCES%%%
\bibliography{rsc} %You need to replace "rsc" on this line with the name of your .bib file
\bibliographystyle{rsc} %the RSC's .bst file

\end{document}

### 2021-03-18 1_SSL_SPIE/article/main.tex
corrected spaces in folder names that caused compilation errors

--- 2021-02-12 SPIE/article/main.tex
@@
-%\documentclass[]{spie}  %>>> use for US letter paper
 \documentclass[a4paper,nocompress]{spie}  %>>> use this instead for A4 paper
-%\documentclass[nocompress]{spie}  %>>> to avoid compression of citations
 
@@
 \usepackage{tabularx}
-\usepackage{wrapfig}
-\setlength{\extrarowheight}{2pt}
-\usepackage{chemformula}
+%\usepackage{wrapfig}
+%\setlength{\extrarowheight}{2pt}
+%\usepackage{chemformula}
 
@@
 \author[a,b]{Michael Weinold}
-\author[a]{Sergey Kolesnikov}
-\author[a,c]{Laura Diaz Anadon}
-\affil[a]{Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, CB3 9EP, UK}
-\affil[b]{Chair of Entrepreneurial Risks, ETH Zurich, Scheuchzerstrasse 7, CH-8092 Zurich, CH}
-\affil[c]{Belfer Center for Science and International Affairs, Harvard Kennedy School, Harvard University, Cambridge, MA 02138, USA}
 
@@
 % Option to view page numbers
-\pagestyle{empty} % change to \pagestyle{plain} for page numbers   
-\setcounter{page}{301} % Set start page numbering at e.g. 301
+%\pagestyle{empty} % change to \pagestyle{plain} for page numbers   
+%\setcounter{page}{301} % Set start page numbering at e.g. 301
 
 \begin{document}
+
 \maketitle
@@
             \begin{center}
-                \includegraphics[width=0.85\textwidth]{SPIE/article/droop_lumileds.pdf}
+                \includegraphics[width=0.85\textwidth]{droop_lumileds.pdf}
             \end{center}
@@
             \begin{center}
-                \includegraphics[width=\textwidth]{SPIE/article/chip_architectures.pdf}
+                \includegraphics[width=\textwidth]{chip_architectures.pdf}
             \end{center}
@@
         \begin{center}
-            \includegraphics[width=0.85\textwidth]{SPIE/article/breakthroughs_efficiency.pdf}
+            \includegraphics[width=0.85\textwidth]{breakthroughs_efficiency.pdf}
         \end{center}
@@
 % References
-\bibliography{report} % bibliography data in report.bib
+\bibliography{main} % bibliography data in main.bib
 \bibliographystyle{spiebib} % makes bibtex use spiebib.bst

### 2021-03-18 4_SSL_XXX/article/main.tex
corrected spaces in folder names that caused compilation errors

--- 2021-02-18 IEEE/article/contribution_variables.tex
@@
 \usepackage{amsmath}
+
+\usepackage{babel,csquotes,xpatch}
+\usepackage[backend=biber]{biblatex}
+
+\bibliography{./contribution_variables.bib}
 
@@
 
+\clearpage
+
+Types of decomposition analysis:
+
+Structured decomposition analysis (SDA): "Decomposition analysis is a so-called comparative-static technique and can be utilised
+on various types of statistics including macro variables of the full economy.
+Adding the term “Structural” to get “Structural Decomposition Analysis” indicates the inclusion of the detailed structure of production as well as final demand by input output tables and models.
+In analyses of the complex interaction between the economy and the environment it is very important to get all the details in the production and consumption structure provided by the detailed input-output tables included in the analysis."(https://circabc.europa.eu/sd/a/01283b06-6c5c-4728-b640-19e2a39b7673/09%20DK%20247%20SDA.pdf)
+
+\clearpage
+\printbibliography
+
 \end{document}

### 2021-03-18 1_SSL_SPIE/article/main.tex
added missing author names

--- 2021-02-12 SPIE/article/main.tex
@@
 % Option to view page numbers
-\pagestyle{empty} % change to \pagestyle{plain} for page numbers   
-\setcounter{page}{301} % Set start page numbering at e.g. 301
+\pagestyle{plain} % change to \pagestyle{plain} for page numbers   
+\setcounter{page}{1} % Set start page numbering at e.g. 301
 
@@
 % References
-\bibliography{report} % bibliography data in report.bib
+\bibliography{main} % bibliography data in report.bib
 \bibliographystyle{spiebib} % makes bibtex use spiebib.bst

### 2022-07-28 2_SSL_EES/article/article_supplementary.tex
added supplementary information file

\documentclass{article}
% formatting
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% cross-referencing
\usepackage{hyperref}
\usepackage{url}
\usepackage{doi}
% tables
\usepackage{booktabs}
% fonts
\usepackage{amsfonts}
\usepackage{microtype}
% figures
\usepackage{graphicx}
% math
\usepackage{amsmath}

% template
\usepackage{style_files/arxiv}

\title{Something something progress in light}

\date{July 2022}

\author{
    \href{https://orcid.org/0000-0003-4859-2650}
    {\includegraphics[scale=0.06]{2_SSL_EES/article/figures/orcid.pdf}
    \hspace{1mm}
    Michael Weinold} \\
	ETH Zurich\\
    Switzerland \\
	\texttt{michael.weinold@alumni.ethz.ch} \\
	\AND
	Dr. Sergey Kolesnikov\\
	University of Cambridge \\
	United Kingdom \\
    \And
	Laura Diaz Anadon \\
	University of Cambridge \\
	United Kingdom \\
}

% Uncomment to remove the date
%\date{}

% Uncomment to override  the `A preprint' in the header
\renewcommand{\headeright}{Supplementary Information}
\renewcommand{\undertitle}{Supplementary Information}
%\renewcommand{\shorttitle}{\textit{arXiv} Template}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={A template for the arxiv style},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={David S.~Hippocampus, Elias D.~Striatum},
pdfkeywords={First keyword, Second keyword, More},
}

\begin{document}
\maketitle

\section{Manufacturing Cost Model: Details on computation}

\subsection{Total manufacturing cost}

The manufacturing process of semiconductor devices can be categorized by the level of integration at which steps are implemented. While the upstream steps in the manufacturing flow are carried out at the wafer level, the packaging steps are typically carried out at the die/package level. The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.

\begin{equation}
\label{eqn:cost_sum}
    C \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
\end{equation}

where

\begin{align*}
    P_S &\dots \text{sapphire substrate price} \\
    C_w &\dots \text{wafer processing cost} \\
    C_p &\dots \text{die processing cost}
\end{align*}

The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.

\begin{align}
	C_w &= \sum_i C_i \\
	C_p &= \sum_j C_j
\end{align}

The cost of a single process step $C_i$ can now be written as

\begin{equation}
\label{eqn:cost_wafer}
    C_i \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
\end{equation}

where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the indices $x,y$ run over all materials.

\begin{align*}
        D &\dots \text{good die per wafer} \label{def:cost_wafer_first} \\
        y &\dots \text{process step yield} \\
        u &\dots \text{equipment utilization} \\
        p &\dots \text{power consumption} \\
        e &\dots \text{hourly electricity cost} \\
        m &\dots \text{hourly maintenance cost} \\
        d &\dots \text{hourly depreciation cost}\\
        l &\dots \text{hourly labour cost} \\
        o &\dots \text{hourly overhead cost} \\
        t_i &\dots \text{time per run} \\
        w &\dots \text{wafers per run} \\
        A &\dots \text{wafer area} \\
        v_x &\dots \text{volume of substance per wafer} \\
        p_x &\dots \text{price of substance $x$ per volume}\\
\end{align*}

The number of die per wafer $D$ depends on the total usable wafer area. The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.

\begin{equation}
	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
\end{equation}

Determining the usable wafer area as a of these three parameters requires a numerical solution. However, following discussions in literature \cite{de2005investigation}, we approximate the number of good die per wafer as

\begin{equation}
label{eqn:dpw}
	D=\frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
\end{equation}

where

\begin{align*}
    d &\dots \text{wafer diameter} \\
    e &\dots \text{wafer edge exclusion zone width} \\
    a &\dots \text{die area} \\
    s &\dots \text{cutting street width} \\
\end{align*}

which gives us for the cost of a manufacturing step $C_i$ in the wafer processing category

\begin{equation}
\label{eqn:cost_wafer_full}
\begin{split}
    C_i \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
    &  \frac{1}{y_i}  \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
\end{split}
\end{equation}

and the cost of a manufacturing step $C_j$ in the packaging category

\begin{equation}
\label{eqn:cost_die}
    C_j \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((e*p) + l + m + d + o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x p_x \bigg\}
\end{equation}

where $c_j$ is $\text{throughput}^{-1}$. The total cost is thus
\begin{equation}
\label{eqn:cost_total}
\begin{split}
    C= P_s &+ \sum_i \bigg \{ \frac{1}{D} \frac{1}{y_i} \bigg[ \frac{t_i}{w_i u_i} \bigg((e*p) + l + m + d +o \bigg)_i  + \sum_{x} v_x p_x \bigg] \bigg \} + \\
    & + \sum_j \bigg \{ \frac{1}{y_j} \bigg[ \frac{c_j}{u_j}  \bigg((e*p) + l + m + d + o \bigg)_i + \sum_{x} a v_x p_x \bigg ] \bigg\}
\end{split}
\end{equation}

Note that in keeping with the categorization introduced by the United States Department of energy, certain steps from these two categories are reported separately. In the wafer processing category, the epitaxy step is reported separately due to its complexity and the large share of cost carried. In the wafer processing category, the phosphor step is reported separately.

\subsection{Yielded Cost}

Devices may be damaged or otherwise rendered unusable during the manufacturing process. The ratio between the number of good devices per step and the number of handled devices per step is known as the yield. Optimizing this yield is critical for reducing manufacturing cost \cite{Kumar2006}. This is because cumulative yield quickly drops as the yield from manufacturing steps with below 100\% yield is multiplied. We must thus consider not only the manufacturing cost per process step, but also the cost including the yield \cite{becker2001use}\cite{becker2001using}. While there are different mathematical approaches to including yield, we follow the definition in  \cite{becker2001use}. We write for the yielded cost $C_{Y_i}$ of a step $i$

\begin{equation}
\begin{split}
\label{eqn:C_2}
    C_{Y_1} &= \frac{C_1}{Y_1} \\
    C_{Y_2} &= \frac{C_1 + C_2}{Y_1 Y_2} - C_{Y_1} = \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} = \frac{1}{Y_1 Y_2} \bigg ( C_1 (1-Y_2) +C_2 \bigg)\\
    C_{Y_i} &= \frac{ \sum_{x \leq i} C_x }{ \prod_{x \leq i} Y_x } - \frac{ \sum_{x<i} C_x }{ \prod_{x<i} Y_x }
\end{split}
\end{equation}

If a step is applied more than once, we can conveniently rewrite this in a form suited to computation within the \textit{Excel} worksheet. Assuming step $2$ is used twice, we get for the yielded cost of this step an equation of the form

\begin{align}
\label{eqn:C_2^2}
    C_{Y_2}^{(2 \times)} &= \bigg( \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} \bigg) + \bigg( \frac{C_1 + 2 C_2}{Y_1 Y_2^2} - \frac{C_1 + C_2}{Y_1 Y_2}     \bigg) \\
    &= \frac{1}{Y_1 Y_2^2} \bigg( C_1 (1-Y_2^2) +2C_2 \bigg)
\end{align}

which can be compared to equation \ref{eqn:C_2} for a more general form.

\begin{equation}
    C_2^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
\end{equation}

Omitting a proof by induction, we get for the general term for a step $i>1$ repeated $n$ times an equation of the form

\begin{equation}
\label{eqn:yielded_cost}
    C_{Y_{i>1}}^{(n \times)} = \frac{1}{Y_i^{n-1} \prod_{x \leq i} Y_x} \bigg( nC_i + \sum_{x < i} C_x (1-Y_i^n) \bigg)
\end{equation}

This cummulative approach to yielded cost is different from the approach taken in the original \textit{LEDCOM} model. It uses what in literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}. The yielded cost of a single process step $f$ is described by

\begin{equation}
	f_\text{single} = \frac{i+s}{y}
\end{equation}

where

\begin{align*}
    i &\dots \text{material cost of previous step} \\
    s &\dots \text{step cost}
\end{align*}

For process steps that are performed more than once, a series expression is used

\begin{align}
\label{eqn:series}
    f_{\times 2} &=  \frac{\frac{i+s}{y}+s}{y} \\
    f_{\times n} &= \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
\end{align}

This itemized approach serves as a convenient approximation, but its cumulative contributions do not equal the total yielded cost of the entire manufacturing process

\begin{equation}
    f_\text{total} = \frac{\sum_i^n s_i}{\prod_i^ny_i} \neq \sum_i^n f_i
\end{equation}

where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in equation \ref{eqn:series}.

\subsection{Contribution of Variables}

To quantify the drivers of cost reductions across all process steps, one would need to identify the magnitude of contribution to cost reductions made by single variables in equations \ref{eqn:cost_wafer} and \ref{eqn:cost_die}. Mathematically, given a function $C$ which describes the cost associated with manufacturing a single unit at time $t$,

\begin{equation}
C=ab+cd
\end{equation}

and manufacturing variables $a,b,c,d$, we are looking for the contribution $\Delta C_{a}$ made by a single variable $a$ to the total change in the cost function $\Delta C$ between points $t_0,t_1$ such, that

\begin{equation}
\Delta C = C(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta C_i
\end{equation}

The infinitesimal contribution to the total cost by the infinitesimal change in a cost model variable is defined through the total differential of the cost function \ref{eqn:cost_sum} as

\begin{equation}
\text{d}C(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial C }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial C }{\partial x_i}  \Delta x_i
\end{equation}

where $x_i$ is an arbitrary cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to total cost $C$ over the period $t_0 < t < t_1 $ is then

\begin{equation}
\Delta C_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial C }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
\label{eqn:integral_1}
\end{equation}

However, data on the cost model variables is not available in continuous time. Disaggregating the contribution of single variables to the total cost reduction is thus not straightforward in our model. This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}. The following discussion follows an approach developed by Kavlak et al. \cite{kavlak2018evaluating}. For the detailed derivation, we refer to this publication.

The cost function $C$ as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as

\begin{equation}
C(\vec{r}) = C(r_1,r_2, \dots) = \sum_i C_i
\end{equation}

where

\begin{equation}
C_i(\vec{r}) = C_i^0 \prod_w g_{iw}(r_w)
\end{equation}

Using logarithmic differentiation, the integral from equation \ref{eqn:integral_1} can be rewritten as

\begin{equation}
\Delta C_x = \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
\end{equation}

where for $C(t)$ a constant $C(t) \approx \tilde{C} $ can be chosen such that $\Delta C_{x_i} = \Delta C$. In practice, this constant value can be approximated as $\tilde{C} \approx \frac{2}{3} C_i^\text{geo} + \frac{1}{3} \overline{C_i}$. The contribution of a single cost model variable $r_z$ can then be written as

\begin{equation}
\Delta C_z (t_1,t_2) \approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
\end{equation}

This approach was used to determine the effect of spillovers on device performance, displayed in figure \ref{fig:breakthroughs_efficiency}. Due to time constraints, this approach could not be used to determine the effect of spillovers on manufacturing cost. Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of equation \ref{eqn:cost_sum}, could this method.

\newpage
\section{Complete list of sources for figures}

\subsection{Figure XX (Historical development of luminous efficacy)}

The complete list of sources is organized by the different technologies described in the figure:

\subsubsection{LED}

Own research, compare:

M. Weinold. ‘Light-Emitting Diode (LED) Technology History’.
In: Zenodo Open Repository. doi: 10.5281/zenodo.3685284

\subsubsection{CFL (<1984)}

\cite{Bouwknegt1982}\cite{Vrenken1983}

\subsubsection{CFL (1984-2011)}

\cite{eger2018origin}

\subsubsection{CFL (>2011)}

\cite{Guan2015}

\subsubsection{Fire, Incandescent, HID}

\cite{azevedo2009transition} augmented by own calculations based on \cite{benesch1905beleuchtungswesen}

\subsubsection{Max. efficacy}

\cite{Murphy2012} Note that this is in contrast to the original value given in \cite{azevedo2009transition}. 
Note that a humorous illustration of this trend in this figure has been provided by Randall Munroe in \href{https://xkcd.com/1603/}{https://xkcd.com/1603/}.

\subsection{Figure XX (Historical development of lamp prices)}

Consumer organization magazines:

\textit{Stiftung Warentest (Germany)}: \cite{Warentest2008}\cite{Warentest2009_1}\cite{Warentest2009_2}\cite{Warentest2010_1}\cite{Warentest2010_2}\cite{Warentest2011}\cite{Warentest2012}\cite{Warentest2013}\cite{Warentest2014_1}\cite{Warentest2014_2}\cite{Warentest2015}\cite{Warentest2016_1}\cite{Warentest2016_2}\cite{Warentest2018}

\textit{Konsument (Austria)}: \cite{Konsument2010}

\textit{Which (UK):} \cite{Which2020})

Industry periodicals:

\cite{PM2020}

Government reports:

\cite{council2013assessment}

Support by German consumer organization Stiftung Warentest in providing access to their archive is gratefully acknowledged.
Figure XX (Historical evolution of light-emitting diode chip architectures)

Scientific publications:

\cite{plossl2010wafer}\cite{bierhuizen2007performance}\cite{gencc2019distributed}\cite{chong2014performance}

Patents:
\href{https://worldwide.espacenet.com/patent/search?q=pn%3DJPH11168235A}{JPH11168235A}, 
\href{https://worldwide.espacenet.com/patent/search?q=pn%3DDE19921987A1}{DE19921987A1}, 
\href{https://worldwide.espacenet.com/patent/search?q=pn%3DJPH11340514A}{JPH11340514A}, 
\href{https://worldwide.espacenet.com/patent/search?q=pn%3DJP2001044498A}{JP2001044498A}, 
\href{https://worldwide.espacenet.com/patent/search?q=pn%3DUS2005045893A1}{US2005045893A1}, 
\href{https://worldwide.espacenet.com/patent/search?q=pn%3DUS2002093023A1}{US2002093023A1}, 
\href{https://worldwide.espacenet.com/patent/search?q=pn%3DUS2006273339A1}{US2006273339A1}

Other publications:

https://web.archive.org/web/20170801160530/
https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf
https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf
http://web.archive.org/web/20160425025936/
https://www.slideshare.net/Yole_Developpement/yole-led-packagingjanuary2013reportsample

Figure XX (Historical developments in the device sub-efficiencies):

The complete list of sources for the figure provided below is organized by the sub-efficiencies (panels) described in the figure.

Forward Voltage (panel a):

\cite{nichia2001data}\cite{lumi2002data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data}

Internal Quantum Efficiency (panel b):

Data points: own research, compare:

M. Weinold. ‘Light-Emitting Diode (LED) Internal Quantum Efficiency History’.
In: Zenodo Open Repository. doi: 10.5281/zenodo.XXXXXXXX

Average performance and future targets set by the U.S. Department of Energy (DoE) derived from industry round-table discussions:

\cite{doe_ssl_multiyear_2006}\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2009}\cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016}

Note that the artifact in the DoE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}

Droop (panel c):

Data calculated from luminous intensity curves of respective device datasheets. Sources for datapoints:

\cite{datasheet_osram_topled}\cite{osram2008data}\cite{osram2008gdplus}\cite{osram2018csp}\cite{datasheet_lumileds_lux1}\cite{lumi2008data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{samsung2018data} \\ Sources for DoE data: \cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016}

Light-Extraction Efficiency (panel d):

\cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012}\cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010}\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011}

\newpage
\bibliographystyle{IEEEtran}
\bibliography{bibliography}

\end{document}

### 2022-07-28 2_SSL_EES/article/article_supplementary.tex
beautification and fixed bib file

@@
 
-\cite{Murphy2012} Note that this is in contrast to the original value given in \cite{azevedo2009transition}. 
+\cite{Murphy2012}
+
+Note that this is in contrast to the original value given in \cite{azevedo2009transition}. 
 Note that a humorous illustration of this trend in this figure has been provided by Randall Munroe in \href{https://xkcd.com/1603/}{https://xkcd.com/1603/}.
@@
 
-Consumer organization magazines:
+The complete list of sources is organized by different sources:
 
-\textit{Stiftung Warentest (Germany)}: \cite{Warentest2008}\cite{Warentest2009_1}\cite{Warentest2009_2}\cite{Warentest2010_1}\cite{Warentest2010_2}\cite{Warentest2011}\cite{Warentest2012}\cite{Warentest2013}\cite{Warentest2014_1}\cite{Warentest2014_2}\cite{Warentest2015}\cite{Warentest2016_1}\cite{Warentest2016_2}\cite{Warentest2018}
+\subsubsection{Consumer organization magazines}
 
-\textit{Konsument (Austria)}: \cite{Konsument2010}
+Stiftung Warentest (Germany): \cite{Warentest2008}\cite{Warentest2009_1}\cite{Warentest2009_2}\cite{Warentest2010_1}\cite{Warentest2010_2}\cite{Warentest2011}\cite{Warentest2012}\cite{Warentest2013}\cite{Warentest2014_1}\cite{Warentest2014_2}\cite{Warentest2015}\cite{Warentest2016_1}\cite{Warentest2016_2}\cite{Warentest2018}
 
-\textit{Which (UK):} \cite{Which2020})
+Konsument (Austria): \cite{Konsument2010}
 
-Industry periodicals:
+Which (UK): \cite{Which2020}
+
+\subsubsection{Industry periodicals}
 
@@
 Support by German consumer organization Stiftung Warentest in providing access to their archive is gratefully acknowledged.
-Figure XX (Historical evolution of light-emitting diode chip architectures)
 
-Scientific publications:
+\subsection{Figure XX (Historical evolution of light-emitting diode chip architectures)}
+
+\subsubsection{Scientific publications}
+
+The complete list of sources is organized by different sources:
 
@@
 
-Patents:
-\href{https://worldwide.espacenet.com/patent/search?q=pn%3DJPH11168235A}{JPH11168235A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn%3DDE19921987A1}{DE19921987A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn%3DJPH11340514A}{JPH11340514A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn%3DJP2001044498A}{JP2001044498A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn%3DUS2005045893A1}{US2005045893A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn%3DUS2002093023A1}{US2002093023A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn%3DUS2006273339A1}{US2006273339A1}
+\subsubsection{Patents}
 
-Other publications:
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11168235A}{JPH11168235A}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DDE19921987A1}{DE19921987A1}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11340514A}{JPH11340514A}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJP2001044498A}{JP2001044498A}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2005045893A1}{US2005045893A1}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2002093023A1}{US2002093023A1}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2006273339A1}{US2006273339A1}
 
-https://web.archive.org/web/20170801160530/
-https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf
-https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf
-http://web.archive.org/web/20160425025936/
-https://www.slideshare.net/Yole_Developpement/yole-led-packagingjanuary2013reportsample
+\subsubsection{Other publications}
 
-Figure XX (Historical developments in the device sub-efficiencies):
+\href{https://web.archive.org/web/20170801160530/https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf}{2015 presentation (www.energy.gov)}
+\href{https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf}{2016 presentation (www.energy.gov)}
+\href{http://web.archive.org/web/20160425025936/https://www.slideshare.net/Yole_Developpement/yole-led-packagingjanuary2013reportsample}{2013 presentation (www.slideshare.net)}
 
-The complete list of sources for the figure provided below is organized by the sub-efficiencies (panels) described in the figure.
+\subsection{Figure XX (Historical developments in the device sub-efficiencies)}
 
-Forward Voltage (panel a):
+The complete list of sources is organized by the sub-efficiencies (panels) described in the figure:
+
+\subsubsection{Forward Voltage (panel a)}
 
@@
 
-Internal Quantum Efficiency (panel b):
+\subsubsection{Internal Quantum Efficiency (panel b)}
 
@@
 
-Average performance and future targets set by the U.S. Department of Energy (DoE) derived from industry round-table discussions:
+Average performance and future targets set by the U.S. Department of Energy (DOE) derived from industry round-table discussions:
 
@@
 
-Note that the artifact in the DoE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}
+Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}
 
-Droop (panel c):
+\subsubsection{Droop (panel c)}
 
-Data calculated from luminous intensity curves of respective device datasheets. Sources for datapoints:
+Data calculated from luminous intensity curves of respective device datasheets.
 
-\cite{datasheet_osram_topled}\cite{osram2008data}\cite{osram2008gdplus}\cite{osram2018csp}\cite{datasheet_lumileds_lux1}\cite{lumi2008data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{samsung2018data} \\ Sources for DoE data: \cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016}
+Sources for datapoints: \cite{datasheet_osram_topled}\cite{osram2008data}\cite{osram2008gdplus}\cite{osram2018csp}\cite{datasheet_lumileds_lux1}\cite{lumi2008data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{samsung2018data}
 
-Light-Extraction Efficiency (panel d):
+Sources for DOE data: \cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016}
 
-\cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012}\cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010}\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011}
+\subsubsection{Light-Extraction Efficiency (panel d)}
+
+\cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012}\cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010} \newline
+\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011}
 

### 2022-07-28 2_SSL_EES/article/article_supplementary.tex
minor changes

@@
 
-\cite{Murphy2012}
+\cite{Murphy2012} (in contrast to the original value given in \cite{azevedo2009transition})
 
-Note that this is in contrast to the original value given in \cite{azevedo2009transition}. 
 Note that a humorous illustration of this trend in this figure has been provided by Randall Munroe in \href{https://xkcd.com/1603/}{https://xkcd.com/1603/}.
@@
 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11168235A}{JPH11168235A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DDE19921987A1}{DE19921987A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11340514A}{JPH11340514A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJP2001044498A}{JP2001044498A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2005045893A1}{US2005045893A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2002093023A1}{US2002093023A1}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11168235A}{JPH11168235A} \\
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DDE19921987A1}{DE19921987A1} \\
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11340514A}{JPH11340514A} \\
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJP2001044498A}{JP2001044498A} \\ 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2005045893A1}{US2005045893A1} \\ 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2002093023A1}{US2002093023A1} \\
 \href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2006273339A1}{US2006273339A1}
@@
 
-\href{https://web.archive.org/web/20170801160530/https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf}{2015 presentation (www.energy.gov)}
-\href{https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf}{2016 presentation (www.energy.gov)}
+\href{https://web.archive.org/web/20170801160530/https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf}{2015 presentation (www.energy.gov)} \\
+\href{https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf}{2016 presentation (www.energy.gov)} \\
 \href{http://web.archive.org/web/20160425025936/https://www.slideshare.net/Yole_Developpement/yole-led-packagingjanuary2013reportsample}{2013 presentation (www.slideshare.net)}

### 2022-07-29 2_SSL_EES/article/article_main.tex
added supplementary information file
started adding content to main paper

--- 2021-03-16 SSL Paper 3 (XXX)/article/main.tex
@@
 
+\usepackage{layout}
+
 \begin{document}
@@
 }
-%%%END OF HEADER%%%
+%%%END OF HEADER
 
@@
 \fancyfoot{}
-\fancyfoot[LO,RE]{\vspace{-7.1pt}\includegraphics[height=9pt]{head_foot/LF}}
-\fancyfoot[CO]{\vspace{-7.1pt}\hspace{13.2cm}\includegraphics{head_foot/RF}}
-\fancyfoot[CE]{\vspace{-7.2pt}\hspace{-14.2cm}\includegraphics{head_foot/RF}}
+\fancyfoot[LO,RE]{\vspace{-7.1pt}\includegraphics[height=9pt]{auxiliary/rsc_template_head_foot/LF}}
+\fancyfoot[CO]{\vspace{-7.1pt}\hspace{13.2cm}\includegraphics{auxiliary/rsc_template_head_foot/RF}}
+\fancyfoot[CE]{\vspace{-7.2pt}\hspace{-14.2cm}\includegraphics{auxiliary/rsc_template_head_foot/RF}}
 \fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
@@
   \begin{@twocolumnfalse}
-{\includegraphics[height=30pt]{head_foot/journal_name}\hfill\raisebox{0pt}[0pt][0pt]{\includegraphics[height=55pt]{head_foot/RSC_LOGO_CMYK}}\\[1ex]
-\includegraphics[width=18.5cm]{head_foot/header_bar}}\par
+{\includegraphics[height=30pt]{auxiliary/rsc_template_head_foot/journal_name}\hfill\raisebox{0pt}[0pt][0pt]{\includegraphics[height=55pt]{auxiliary/rsc_template_head_foot/RSC_LOGO_CMYK}}\\[1ex]
+\includegraphics[width=18.5cm]{auxiliary/rsc_template_head_foot/header_bar}}\par
 \vspace{1em}
@@
 
-\includegraphics{head_foot/DOI} & \noindent\LARGE{\textbf{This is the title$^\dag$}} \\%Article title goes here instead of the text "This is the title"
+\includegraphics{auxiliary/rsc_template_head_foot/DOI} & \noindent\LARGE{\textbf{Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers}} \\
 \vspace{0.3cm} & \vspace{0.3cm} \\
 
- & \noindent\large{Full Name,$^{\ast}$\textit{$^{a}$} Full Name,\textit{$^{b\ddag}$} and Full Name\textit{$^{a}$}} \\%Author names go here instead of "Full name", etc.
+ & \noindent\large{Michael Weinold \textit{$^{a b \ddag}$}, Sergey Kolesnikov,\textit{$^b$} and Laura Diaz Anadon\textit{$^{b}$}} \\
 
-\includegraphics{head_foot/dates} & \noindent\normalsize{The abstract should be a single paragraph which summarises the content of the article.
-Any references in the abstract should be written out in full \textit{e.g.}\ [Surname \textit{et al., Journal Title}, 2000, \textbf{35}, 3523].} \\%The abstrast goes here instead of the text "The abstract should be..."
+\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{First commercial white light-emitting diodes (LEDs) were introduced to the market in 1996 .
+Since then, white LEDs have experienced major improvements in performance and decreases in cost, resulting in rapid market expansion of LED-based solid-state lighting (SSL), one of the key current solutions in global climate change mitigation efforts.
+Despite the significance of the technology, the extent and sources of these improvements have not been systematically investigated.
+Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies for climate change mitigation.
+With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from a literature review, interviews with experts in industry and academia, and results from our cost and performance modelling.
+We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020.
+During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S.
+location using state-of-the-art equipment has dropped from $1.1\$$ to $0.05\$$ (in 2020 USD) between 2003 and 2020, a $95.5\%$ decrease.
+We also show that technology spillovers—i.e., knowledge originating in other technologies—affected all performance dimensions of LEDs.
+They contributed no less than $8.5\%$ of the total efficiency improvements of devices and were responsible for $\sim 100\%$ of the improvements in consumer experience metrics.
+The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}
 
@@
 
-\footnotetext{\textit{$^{a}$~Address, Address, Town, Country. Fax: XX XXXX XXXX; Tel: XX XXXX XXXX; E-mail: xxxx@aaa.bbb.ccc}}
-\footnotetext{\textit{$^{b}$~Address, Address, Town, Country. }}
+\footnotetext{\textit{$^{a}$~ETH Zurich, Zurich, Switzerland. E-mail: michael.weinold@alumni.ethz.ch}}
+\footnotetext{\textit{$^{b}$~Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom. }}
 
@@
 
-\footnotetext{\ddag~Additional footnotes to the title and authors can be included \textit{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P.
-Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}
+\footnotetext{\ddag~Present address: Paul Scherrer Institut, Laboratory for Energy Analysis, Group for Technology Assessment, Switzerland.}
 
@@
 %%%MAIN TEXT%%%%
-The main text of the article\cite{Mena2000} should appear here.
 
-\subsection{This is the subsection heading style}
-Section headings can be typeset with and without numbers.\cite{Abernethy2003}
+\section{Introduction}
 
-\subsubsection{This is the subsubsection style.~~} These headings should end in a full point.
+A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}.
+According to the United Nations, by the end of 2021, more than 130 countries have set or are considering setting a net zero emissions target by or around 2050 \cite{un2021climate}.
+The European Union, for example, has set a goal of net zero emissions by 2050—a goal it aims to meet with the help of the Green New Deal along other EU and national policies as part of the \textit{European Green Deal} \cite{eu2020green}.
 
-\paragraph{This is the next level heading.~~} For this level please use \texttt{\textbackslash paragraph}. These headings should also end in a full point.
+Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side technologies \cite{rgeVorsatz2009}.
+To ensure rapid adoption of these technologies, improvements in their costs, performance, and consumer experience are needed.
+This requires understanding how cost reductions and performance improvements can be achieved in these technologies \cite{Stephan2021}\cite{kolesnikov2020novel} \cite{Ziegler2021}.
 
-\section{Graphics and tables}
-\subsection{Graphics}
-Graphics should be inserted on the page where they are first mentioned (unless they are equations, which appear in the flow of the text).\cite{Cotton1999}
+There are a range of mechanisms that contribute to technological change in a technology over time, including targeted research and development (R\&D), economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global}\cite{national2016power}\cite{iea2020perspectives}.
+The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{kolesnikov2020novel} \cite{Ziegler2021}.
+Innovation is also driven by forces of supply and demand.
+Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.
 
-\begin{figure}[h]
-\centering
-  \includegraphics[height=3cm]{example1}
-  \caption{An example figure caption.}
-  \label{fgr:example}
-\end{figure}
+The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies} \cite{Anadon2016}.
+Among these drivers, the role of knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area \cite{Stephan2021}.
+While the exact definition of spillovers in literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012} we follow Stephan et al.
+\cite{Stephan2021} and consider knowledge to be external – thus a spillover –  if it has been developed for application in other technologies, sectors, or scientific disciplines.
+There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be a fruitful approach to understand \cite{Pichler2020} and shape  \cite{Clark2016} \cite{ Stephan2021} \cite{Sun2021} the future evolution of technologies.
 
-\begin{figure*}
- \centering
- \includegraphics[height=3cm]{example2}
- \caption{A two-column figure.}
- \label{fgr:example2col}
-\end{figure*}
-
-\subsection{Tables}
-Tables typeset in RSC house style do not include vertical lines.
-Table footnote symbols are lower-case italic letters and are typeset at the bottom of the table.
-Table captions do not end in a full point.\cite{Arduengo1992,Eisenstein2005}
-
-\begin{table}[h]
-\small
-  \caption{\ An example of a caption to accompany a table}
-  \label{tbl:example}
-  \begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}}lll}
-    \hline
-    Header one (units) & Header two & Header three \\
-    \hline
-    1 & 2 & 3 \\
-    4 & 5 & 6 \\
-    7 & 8 & 9 \\
-    10 & 11 & 12 \\
-    \hline
-  \end{tabular*}
-\end{table}
-
-Adding notes to tables can be complicated.
-Perhaps the easiest method is to generate these manually.\footnote[4]{Footnotes should appear here.
-These might include comments relevant to but not central to the matter under discussion, limited experimental and spectral data, and crystallographic data.}
-
-\begin{table*}
-\small
-  \caption{\ An example of a caption to accompany a table \textendash\ table captions do not end in a full point}
-  \label{tbl:example}
-  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lllllll}
-    \hline
-    Header one & Header two & Header three & Header four & Header five & Header six  & Header seven\\
-    \hline
-    1 & 2 & 3 & 4 & 5 & 6  & 7\\
-    8 & 9 & 10 & 11 & 12 & 13 & 14 \\
-    15 & 16 & 17 & 18 & 19 & 20 & 21\\
-    \hline
-  \end{tabular*}
-\end{table*}
-
-\section{Equations}
-
-Equations can be typeset inline \textit{e.g.}\ $ y = mx + c$ or displayed with and without numbers:
-
- \[ A = \pi r^2 \]
-
-\begin{equation}
-  \frac{\gamma}{\epsilon x} r^2 = 2r
-\end{equation}
-
-You can also put lists into the text. You can have bulleted or numbered lists of almost any kind. 
-The \texttt{mhchem} package can also be used so that formulae are easy to input: \texttt{\textbackslash ce\{H2SO4\}} gives \ce{H2SO4}.
-
-For footnotes in the main text of the article please number the footnotes to avoid duplicate symbols.
-\textit{e.g.}\ \texttt{\textbackslash footnote[num]\{your text\}}.
-The corresponding author $\ast$ counts as footnote 1, ESI as footnote 2, \textit{e.g.}\ if there is no ESI, please start at [num]=[2], if ESI is cited in the title please start at [num]=[3] \textit{etc.} Please also cite the ESI within the main body of the text using \dag.
+Among demand-side technologies, the provision of lighting is a particularly important area  for climate change mitigation efforts, as it currently accounts for $15-19\%$ of global electricity consumption \cite{Zissis2016}\cite{doe_electricity}.
+It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.
+As shown in Figure 1, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
+For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W.
+Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices.
+During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure 2.
 
 \section{Conclusions}
-The conclusions section should come in this section at the end of the article, before the Conflicts of interest statement.
+Our study has analysed in a systematic and granular way the sources of dramatic cost reductions and efficiency improvements in white light-emitting diodes since their introduction to the market in 1996.
+We find that the total LED device efficiency increased from $5.8\%$ to $38.7\%$ between 2003 and 2020 has been predominantly driven by LED technology innovations that have affected all physical energy loss channels and corresponding device sub-efficiencies.
+We also find that among those innovations, at least nine were driven by technology originating in areas of science and technology outside LEDs and solid-state lighting—i.e., those nine innovations can be referred to as technology spillovers .
+These spillovers were responsible for $8.5\%$ of the total efficiency improvements and nearly $100\%$ of the improvements in consumer experience metrics.
+
+Our manufacturing cost model shows that a $95.5\%$ decrease in the cost of producing white classic-chip LEDs (from $1.11\$$ to $0.05\$$ in 2020 USD) between 2003 and 2020.
+While efficiency improvements have played an important role in these cost reductions, the largest components of such cost reductions have been driven by increases in the wafer size and yields across different manufacturing steps.
+In contrast with efficiency improvements, these increases were mainly a result of learning-by-doing and manufacturing equipment improvements rather than LED technology innovations or spillovers.
+
+Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of efficiency improvements, highlights the critical role played by a growing deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs.
+Specifically, deep physical understanding of LED device efficiency loss channels enabled important breakthroughs and spillovers in LEDs and will enable further innovations to improve LED and SSL efficiency in the future, as expected by eminent experts in the field \cite{Weisbuch2020}.
+This suggests that additional research in these areas and a more deliberate search for spillovers may accelerate these expected future advances.
+Our findings on spillover enablers show that these efforts can be further supported or even accelerated by knowledge exchange events and long-term partnerships between academia and industry, dedicated mission-driven public R\&D funding, and freedom of search in academia.
+This reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles} \cite{narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
+
+There are various important avenues of future research that are opened up by our analysis.
+First, future work could expand the cost model by collecting and including data for a broader set of chip architectures and analyzing the impact of individual innovations and spillovers on costs as opposed to performance.
+Second, a deeper dive on the role of learning-by-doing is needed both in the cost and performance analysis to identify different types of learning by doing and how it came about.
+Third, building on the work on LED sub-efficiencies and physical limits, future efforts could focus on identifying priority areas for further efficiency improvements in LEDs and SSL in general.
+Finally, by comparing the drivers of innovation, technology spillovers, cost reductions and performance improvements at a granular level across different clean energy technologies, we can identify patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
 
 \section*{Conflicts of interest}
-In accordance with our policy on \href{http://www.rsc.org/journals-books-databases/journal-authors-reviewers/author-responsibilities/#code-of-conduct}{Conflicts of interest} please ensure that a conflicts of interest statement is included in your manuscript here.
-Please note that this statement is required for all submitted manuscripts.
-If no conflicts exist, please state that ``There are no conflicts to declare''.
+There are no conflicts to declare.
 
 \section*{Acknowledgements}
-The Acknowledgements come at the end of an article after Conflicts of interest and before the Notes and references.
+This research is supported by a grant of the Alfred P.
+Sloan Foundation titled \href{http://web.archive.org/web/20200623065733/https://sloan.org/grant-detail/8567}{\textit{“What factors drive innovation in energy technologies?
+The role of technology spillovers and government investment”}}.
+Michael Weinold further gratefully acknowledges support by the \href{https://www.studienstiftung.ch/}{Swiss Study Foundation}.
+We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and \href{https://www.ceenrg.landecon.cam.ac.uk/}{C-EENRG} Seminar Series of the University of Cambridge for many helpful discussions and feedback.
+We also express our deep gratitude to all interviewees for their willingness to participate in this study and their valuable contributions.
 
@@
 %%%REFERENCES%%%
-\bibliography{rsc} %You need to replace "rsc" on this line with the name of your .bib file
-\bibliographystyle{rsc} %the RSC's .bst file
+\bibliography{auxiliary/rsc_style_files/bibliography_rsc}
+\bibliographystyle{bibliography} %the RSC's .bst file
 

### 2022-08-03 2_SSL_EES/article/article_supplementary.tex
started adding content to main paper
added table to supplementary information

@@
 \usepackage{amsmath}
+% tables
+\usepackage{tabularx}
+\usepackage{colortbl}
+\definecolor{silver}{rgb}{0.75, 0.75, 0.75}
 
 % template
-\usepackage{style_files/arxiv}
+\usepackage{auxiliary/arxiv_template_style_files/arxiv}
 
@@
     \href{https://orcid.org/0000-0003-4859-2650}
-    {\includegraphics[scale=0.06]{2_SSL_EES/article/figures/orcid.pdf}
+    {\includegraphics[scale=0.06]{2_SSL_EES/article/auxiliary/rsc_template_figures/orcid.pdf}
     \hspace{1mm}
@@
 
+\section{Contribution of spillovers}
+
+    \begin{table}[h!]
+        \centering
+    	\begin{tabularx}{\textwidth}{|l|l|X|X|X|X|}
+    		\hline
+    			Year & $\eta_L$[\%]  & $\Delta \eta_L^{abs}$[\%] & $\Delta \eta_L^{rel}$[\%]  & $\Delta \eta_L^{abs}(S/O)$[\%]  & $\Delta \eta_L^{rel}(S/O)$[\%]  \\
+    		\hline
+    			2002 & 5.8 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
+    		\hline
+    		    \multicolumn{2}{|c|}{2002-2010} & 6.8 & 117.4 & 1.0 & 14.4 \\
+    		\hline
+    			2010 & 12.5 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
+    		\hline
+    		    \multicolumn{2}{|c|}{2010-2016} & 19.9 & 158 & 1.6 & 8.0 \\
+    		\hline
+    			2016 & 32.5 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
+    		\hline
+    		    \multicolumn{2}{|c|}{2016-2020} & 6.2 & 19.0 & 0.3 & 3.9 \\
+    		\hline
+    			2020 & 38.7 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
+    		\hline
+    		    \multicolumn{2}{|c|}{2002-2020} & 32.9 & 566.4 & 2.8 & 8.5 \\
+            \hline
+    	\end{tabularx}
+    	\caption{Contribution of technology spillovers to improvements in the total lamp efficiency $\eta_L$ for warm white light-emitting diodes in the period 2002-2020.
+Note numbers are rounded.
+\\ $\Delta \eta_L^{abs}\overset{eg.}{=}\eta_L(2010)-\eta_L(2002)$ is absolute change in total lamp efficiency.
+\\ $\Delta \eta_L^{rel}\overset{eg.}{=}(\eta_L(2010)-\eta_L(2002))/(\eta_L(2002)/100)$ is relative change in lamp efficiency.}
+    	\label{tab:spillover-data}
+    \end{table}
+
 \newpage

### 2022-08-06 2_SSL_EES/article/article_main.tex
added introduction and previous literature

@@
 According to the United Nations, by the end of 2021, more than 130 countries have set or are considering setting a net zero emissions target by or around 2050 \cite{un2021climate}.
-The European Union, for example, has set a goal of net zero emissions by 2050—a goal it aims to meet with the help of the Green New Deal along other EU and national policies as part of the \textit{European Green Deal} \cite{eu2020green}.
+The European Union, for example, has set a goal of net zero emissions by 2050—a goal it aims to meet with the help of the European Green Deal, alongside other EU and national policies \cite{eu2020green}.
 
-Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side technologies \cite{rgeVorsatz2009}.
-To ensure rapid adoption of these technologies, improvements in their costs, performance, and consumer experience are needed.
-This requires understanding how cost reductions and performance improvements can be achieved in these technologies \cite{Stephan2021}\cite{kolesnikov2020novel} \cite{Ziegler2021}.
+Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}.
+To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed.
+This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021}\cite{Ziegler2021}.
 
-There are a range of mechanisms that contribute to technological change in a technology over time, including targeted research and development (R\&D), economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global}\cite{national2016power}\cite{iea2020perspectives}.
-The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{kolesnikov2020novel} \cite{Ziegler2021}.
+There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global}\cite{national2016power}\cite{iea2020perspectives}.
+The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}.
 Innovation is also driven by forces of supply and demand.
@@
 
+\begin{figure*}[h]
+ \centering
+ \includegraphics[height=7cm]{2_SSL_EES/article/figures/history_efficacy.pdf}
+ \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020.
+Data shown for best performers by year of market introduction, with dashed lines for visual aid.
+The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
+\cite{Murphy2012}.
+The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year.
+For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines.
+Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot.
+Abbreviations: HID – High-Intensity Discharge; CFL – Compact Fluorescent Lamp; Hal.
+– Halogen, Incd.
+– Incandescent.
+Source: own synthesis of published data based on a visual approach proposed by Azevedo et al.
+\cite{azevedo2009transition}.
+See Supplementary Information, section X, for the full list of data sources.
+}
+ \label{fgr:history_efficecy}
+\end{figure*}
+
 The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies} \cite{Anadon2016}.
-Among these drivers, the role of knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area \cite{Stephan2021}.
-While the exact definition of spillovers in literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012} we follow Stephan et al.
+Among these drivers, the role of external knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area \cite{Stephan2021}.
+While the exact definition of spillovers in literature depends on the context and the research question \cite{liu2003does}\cite{Nemet2012}, we follow Stephan et al.
 \cite{Stephan2021} and consider knowledge to be external – thus a spillover –  if it has been developed for application in other technologies, sectors, or scientific disciplines.
-There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be a fruitful approach to understand \cite{Pichler2020} and shape  \cite{Clark2016} \cite{ Stephan2021} \cite{Sun2021} the future evolution of technologies.
+There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016} \cite{ Stephan2021} \cite{Sun2021} the future evolution of technologies.
 
@@
 It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.
-As shown in Figure 1, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
+As shown in Figure \ref{fgr:history_efficecy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
 For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W.
 Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices.
-During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure 2.
+During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure \ref{fgr:cost_lamp_small}.
+
+\begin{figure}
+\centering
+  \includegraphics[height=5cm]{2_SSL_EES/article/figures/cost_lamp_small.pdf}
+  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
+Annual price per flux decline for the represented period is $27.3\%$, in line with previous estimates \cite{Gerke2020}.
+Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
+Department of Energy (DoE) \cite{ council2013assessment}.
+Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs.
+Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications.
+Data on CFL sales prices collected from \cite{eger2018origin} as well as various consumer watchdog databases and publications.
+Incandescent light bulb price is assumed constant based on the average in the covered time period.
+See Supplementary information, section XX, for the full list of data sources.}
+  \label{fgr:cost_lamp_small}
+\end{figure}
+
+LED sales prices collected from various consumer watchdog databases and publications.
+Data on CFL sales prices collected from \cite{eger2018origin} as well as various consumer watchdog databases and publications.
+Incandescent light bulb price is assumed constant based on the average in the covered time period.
+See Supplementary information, section …, for the full list of data sources.
+
+These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long}\cite{Mills2014}\cite{stegmaier2021incandescent}\cite{grubb2021new}.
+As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year for the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
+Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}.
+For example, durability, low up-front cost and high efficiency of LED light sources have led to their wide-spread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
+LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical}, \cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
+
+Despite this impressive history, the sources of LED innovation have not received as much attention from researchers as innovation in supply-side energy technologies, such as solar photovoltaics \cite{kavlak2018evaluating} or wind energy \cite{qiu2012price}\cite{jennings2020policy}, or in lithium-ion batteries for transportation \cite{Ziegler2021}\cite{Stephan2021}.
+To the best of our knowledge, no study has comprehensively discussed the sources or extent of progress across various metrics of LED cost or performance since the introduction of first commercial white LED products.
+Understanding the extent to which individual innovations and knowledge spillovers contributed to improvements in LED technology, how this effect compares to other sources of improvements such as economies of scale, and how these innovations occurred (i.e., by what mechanisms and actors) will provide valuable lessons both for other demand-side technologies and overall for accelerating clean energy innovation.
+
+To address these questions, in this paper we identify a set of metrics suitable for tracking the historical progress of LED lighting technology.
+We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
+Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry.
+This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.
+Data on performance improvements associated with individual innovations identified in patents, scientific publications, industry reports, and interviews is then augmented by our calculations of device efficiency, including contributions of specific spillovers to the overall LED efficiency (see Supplementary Information for more detail).
+We further calculate LED manufacturing costs using own bottom-up cost model with a process-step resolution.
+Finally, we discuss what our observations and conclusions mean for the understanding of the innovation process in LED technology and for broader policy and industry efforts to accelerate clean energy innovation.
+
+\section{Previous literature}
+
+The historical development of light-emitting diodes has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015} \cite{Nakamura2015}.
+The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006} \cite{krames2007led} \cite{laubsch2009high} \cite{hahn2014development}, or contributions of different sub-efficiencies to overall device efficiency for 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
+
+Several descriptive papers include very useful information on some of the scientific breakthroughs that have contributed to LED progress \cite{krames2007status}, \cite{Phillips2007}, \cite{Bierhuizen2007}, \cite{Nakamura2013}, \cite{feezell2018invention} and \cite{Taki2019}.
+Additional  studies providing more detail on specific advances in LED technology were also published by authors working for key industry actors Lumileds\cite{MuellerMach2005}, \cite{Shchekin2006}, \cite{lumi2015lumi}, \cite{Bhardwaj2017} and Osram (now amsOsram) \cite{Haerle2004}, \cite{Baur2009}, \cite{laubsch2009high}, \cite{hahn2014development}.
+However, these studies typically include limited information on the macro chip-design and cover disparate aspects of the technology over different time periods.
+Another important source of historical information comes from patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds \cite{margalith2011thin}, Samsung \cite{jung2014phos} \cite{cha2019semiconductor} and Soraa \cite{cich2017high}.
+
+We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure 3, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+Despite the amount of literature published on the topic of LED history, to the best of our knowledge, no publication comprehensively aggregates known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
+In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating} \cite{kolesnikov2020novel} \cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not been studied in the context of lighting yet.
+This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012}, \cite{Creutzig2018}.
 
@@
 %%%REFERENCES%%%
-\bibliography{auxiliary/rsc_style_files/bibliography_rsc}
-\bibliographystyle{bibliography} %the RSC's .bst file
+\bibliography{bibliography}
+\bibliographystyle{auxiliary/rsc_style_files/bibliography_rsc} %the RSC's .bst file
 

### 2022-08-06 2_SSL_EES/article/article_main.tex
added figure

@@
 
+\begin{figure*}
+ \centering
+ \includegraphics[height=17cm]{2_SSL_EES/article/figures/chip_architecture_overview.pdf}
+ \caption{Historical evolution of light-emitting diode chip architectures.
+Rows from top to bottom: classical chip with lateral current spreading; Osram’s thin GaN flip-chip (vertical) architecture; flip-chip architecture; chip-scale package flip-chip architecture.
+Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
+The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip.
+Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity.
+Years indicated correspond to the earliest identified patent priority date.
+Black frames around certain designs indicate chip designs not brought to large-scale production.
+Abbreviations: GaN – Gallium Nitride, ITO – Indium Tin Oxide, MQW – Multiple Quantum Well.
+Source: adapted and compiled from multiple patents and industry publications.
+See Supplementary Information, section…, for the full list of sources and references.}
+ \label{fgr:chip_architecture_overview}
+\end{figure*}
+
 \section{Conclusions}

### 2022-08-06 2_SSL_EES/article/article_supplementary.tex
added figures to supplementary information

@@
 
-\title{Something something progress in light}
+\title{Rapid technological progress in \\ white light-emitting diodes and its sources in innovation and technology spillovers}
 
@@
 
+\section{Cost model sensitivity analysis}
+
+\begin{figure}
+	\centering
+    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_sensitivity.pdf}
+	\caption{Panels a-b: Sensitivity analysis of selected cost model parameters. Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.}
+	\label{fig:costmodel_calibration}
+\end{figure}
+
+Sensitivity analysis of selected cost model parameters. Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.
+
+A preliminary sensitivity analysis (panels a-b of Figure XX) confirms these preliminary findings, showing the largest deviation for changes in these parameters.
+
+\section{Cost model comparison with reported industry data}
+
+\begin{figure}
+	\centering
+    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_calibration.pdf}
+	\caption{Modelled manufacturing cost for light-emitting diode packages split by manufacturing step category.
+The top graph shows the data from LEDCOM model and round table discussions published by the United States Department of Energy (DoE).
+The bottom graph shows the results of the cost model developed in this thesis.
+Hatched bars are projections given by the sources.
+Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.}
+	\label{fig:costmodel_calibration}
+\end{figure}
+
+We note differences between our model and costs reported to the US Department of Energy (DoE) by industry.
+For instance, the share of the epitaxy step is larger in the DoE model.
+This can in part be explained by our model relying on state-of-the-art equipment, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+The packaging part of the manufacturing process is increasing in importance, which has been confirmed by researchers and industry reports on wafer-level packaging.
+
 \newpage
@@
 
-Note that a humorous illustration of this trend in this figure has been provided by Randall Munroe in \href{https://xkcd.com/1603/}{https://xkcd.com/1603/}.
+Note that a humorous illustration of the trend of dramatic improvements in luminous flux shown in this figure has been provided by cartoonist Randall Munroe in \href{https://xkcd.com/1603/}{https://xkcd.com/1603/}.
 

### 2023-01-18 2_SSL_EES/article/article_main.tex
updated introduction

@@
 
-\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{First commercial white light-emitting diodes (LEDs) were introduced to the market in 1996 .
+\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{First commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
 Since then, white LEDs have experienced major improvements in performance and decreases in cost, resulting in rapid market expansion of LED-based solid-state lighting (SSL), one of the key current solutions in global climate change mitigation efforts.
@@
 Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies for climate change mitigation.
-With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from a literature review, interviews with experts in industry and academia, and results from our cost and performance modelling.
+With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from a literature review, interviews with experts in industry and academia, and cost and performance modelling.
 We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020.
 During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S.
-location using state-of-the-art equipment has dropped from $1.1\$$ to $0.05\$$ (in 2020 USD) between 2003 and 2020, a $95.5\%$ decrease.
-We also show that technology spillovers—i.e., knowledge originating in other technologies—affected all performance dimensions of LEDs.
-They contributed no less than $8.5\%$ of the total efficiency improvements of devices and were responsible for $\sim 100\%$ of the improvements in consumer experience metrics.
+location using state-of-the-art equipment has dropped from 1.1\$ to 0.05\$ (in 2020 USD) between 2003 and 2020, a 95.5\% decrease.
+We also show that technology spillovers — i.e., knowledge originating in other technologies—affected all performance dimensions of LEDs, contributing no less than 8.5\% of the total efficiency improvements and $\sim$100\% of the improvements in consumer experience metrics.
 The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}
@@
 The European Union, for example, has set a goal of net zero emissions by 2050—a goal it aims to meet with the help of the European Green Deal, alongside other EU and national policies \cite{eu2020green}.
+The United Kingdom similarly has adopted a similar strategy to achieve net-zero by 2050 \cite{noauthor_ieairena_2023}.
 
@@
 To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed.
-This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021}\cite{Ziegler2021}.
+This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021} \cite{Ziegler2021}.
 
 There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global}\cite{national2016power}\cite{iea2020perspectives}.
-The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}.
+The role of these factors at different stages of the innovation life cycle \cite{ grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}.
 Innovation is also driven by forces of supply and demand.
@@
  \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020.
-Data shown for best performers by year of market introduction, with dashed lines for visual aid.
-The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
+Data shown for best performers by year of market introduction, with dashed lines.
+A multi-order polynomial fit was used to indicate the average rate of change as dashed lines.
+The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
 \cite{Murphy2012}.
@@
 \cite{azevedo2009transition}.
-See Supplementary Information, section X, for the full list of data sources.
+See Supplementary Information, section , for the full list of data sources.
 }
- \label{fgr:history_efficecy}
+ \label{fgr:history_efficacy}
 \end{figure*}
 
-The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies} \cite{Anadon2016}.
+The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies}\cite{Anadon2016}.
 Among these drivers, the role of external  knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area  \cite{Stephan2021}.
-While the exact definition of spillovers in literature depends on the context and the research question \cite{liu2003does}\cite{Nemet2012}, we follow Stephan et al.
+While the exact definition of spillovers in literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012}, we follow Stephan et al.
 \cite{Stephan2021} and consider knowledge to be external – thus a spillover –  if it has been developed for application in other technologies, sectors, or scientific disciplines.
-There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016} \cite{ Stephan2021} \cite{Sun2021} the future evolution of technologies.
+There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016}\cite{Stephan2021}\cite{Sun2021}\cite{kolesnikov_framework_2022} the future evolution of technologies.
 
-Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently accounts for $15-19\%$ of global electricity consumption \cite{Zissis2016}\cite{doe_electricity}.
+Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently  accounts for 15-19\% of global electricity consumption \cite{Zissis2016}\cite{doe_electricity}.
 It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.
-As shown in Figure \ref{fgr:history_efficecy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
+As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
 For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W.
 Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices.
-During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure \ref{fgr:cost_lamp_small}.
+During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure 2.
 
@@
 We also express our deep gratitude to all interviewees for their willingness to participate in this study and their valuable contributions.
+Support by German consumer organization \textit{Stiftung Warentest} in providing access to their archive is gratefully acknowledged.
 

### 2023-01-18 2_SSL_EES/article/article_main.tex
updated supplementary information

@@
 It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.
-As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
+As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years , which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
 For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W.
 Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices.
-During the same period, LED retail prices have fallen by two orders of magnitude, as shown in Figure 2.
+Figure \ref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline during the period of from 2008-2020 of 27.3\%, in line with previous estimates \cite{Gerke2020}.
 
@@
   \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
-Annual price per flux decline for the represented period is $27.3\%$, in line with previous estimates \cite{Gerke2020}.
 Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
-Department of Energy (DoE) \cite{ council2013assessment}.
+Department of Energy (DoE) \cite{council2013assessment}.
 Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs.
@@
 Incandescent light bulb price is assumed constant based on the average in the covered time period.
-See Supplementary information, section XX, for the full list of data sources.}
+See Supplementary information, section …, for the full list of data sources.}
   \label{fgr:cost_lamp_small}
@@
 
-LED sales prices collected from various consumer watchdog databases and publications.
-Data on CFL sales prices collected from \cite{eger2018origin} as well as various consumer watchdog databases and publications.
-Incandescent light bulb price is assumed constant based on the average in the covered time period.
-See Supplementary information, section …, for the full list of data sources.
-
-These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long}\cite{Mills2014}\cite{stegmaier2021incandescent}\cite{grubb2021new}.
+These  dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies  \cite{weinold2020long}\cite{Mills2014}\cite{Stegmaier2021}\cite{grubb2021new}.
 As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year for the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
@@
 To address these questions, in this paper we identify a set of metrics suitable for tracking the historical progress of LED lighting technology.
-We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
-Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry.
+We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing  .
+Given the proprietary nature of knowledge in the SSL industry , we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{Tansey} with eminent experts from academia, public research institutions and industry.
 This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.

### 2023-01-18 2_SSL_EES/article/article_supplementary.tex
updated supplementary information

@@
-\documentclass{article}
+\documentclass[10pt]{article}
+% page setup
+\usepackage[a4paper, total={6in, 8.75in}]{geometry}
+\usepackage{parskip}
 % formatting
@@
 \usepackage{colortbl}
+% boxes
+\usepackage{fancybox}
+
+%%% custom definitions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \definecolor{silver}{rgb}{0.75, 0.75, 0.75}
 
-% template
-\usepackage{auxiliary/arxiv_template_style_files/arxiv}
+%%% document metadata %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\title{Rapid technological progress in \\ white light-emitting diodes and its sources in innovation and technology spillovers}
+\title{Rapid technological progress in white light-emitting diodes \\ and its sources in innovation and technology spillovers  }
+\date{February 2023}
 
-\date{July 2022}
-
-\author{
-    \href{https://orcid.org/0000-0003-4859-2650}
-    {\includegraphics[scale=0.06]{2_SSL_EES/article/auxiliary/rsc_template_figures/orcid.pdf}
-    \hspace{1mm}
-    Michael Weinold} \\
-	ETH Zurich\\
-    Switzerland \\
-	\texttt{michael.weinold@alumni.ethz.ch} \\
-	\AND
-	Dr. Sergey Kolesnikov\\
-	University of Cambridge \\
-	United Kingdom \\
-    \And
-	Laura Diaz Anadon \\
-	University of Cambridge \\
-	United Kingdom \\
+\hypersetup{
+    pdftitle={Supplementary Information},
+    pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon}
 }
 
-% Uncomment to remove the date
-%\date{}
+%%% document body %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% Uncomment to override  the `A preprint' in the header
-\renewcommand{\headeright}{Supplementary Information}
-\renewcommand{\undertitle}{Supplementary Information}
-%\renewcommand{\shorttitle}{\textit{arXiv} Template}
+\begin{document}
 
-%%% Add PDF metadata to help others organize their library
-%%% Once the PDF is generated, you can check the metadata with
-%%% $ pdfinfo template.pdf
-\hypersetup{
-pdftitle={A template for the arxiv style},
-pdfsubject={q-bio.NC, q-bio.QM},
-pdfauthor={David S.~Hippocampus, Elias D.~Striatum},
-pdfkeywords={First keyword, Second keyword, More},
+\setlength{\fboxsep}{10pt}
+\fbox{
+    \parbox{\textwidth}{
+        \textbf{\textsc{Supplementary Information}} for: \\
+        M. Weinold, S. Kolesnikov, L.D. Anadon \\
+        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Energy and Environmental Science} (2023)
+    }
 }
 
-\begin{document}
-\maketitle
+\tableofcontents
+
+\newpage
 

### 2023-01-18 2_SSL_EES/article/article_main.tex
updated acknowledgements section

@@
 \section*{Acknowledgements}
-This research is supported by a grant of the Alfred P.
-Sloan Foundation titled \href{http://web.archive.org/web/20200623065733/https://sloan.org/grant-detail/8567}{\textit{“What factors drive innovation in energy technologies?
-The role of technology spillovers and government investment”}}.
-Michael Weinold further gratefully acknowledges support by the \href{https://www.studienstiftung.ch/}{Swiss Study Foundation}.
-We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and \href{https://www.ceenrg.landecon.cam.ac.uk/}{C-EENRG} Seminar Series of the University of Cambridge for many helpful discussions and feedback.
-We also express our deep gratitude to all interviewees for their willingness to participate in this study and their valuable contributions.
-Support by German consumer organization \textit{Stiftung Warentest} in providing access to their archive is gratefully acknowledged.
+This research is supported by the grant from the Alfred P.
+Sloan Foundation titled \textit{“What factors drive innovation in energy technologies?
+The role of technology spillovers and government investment”}.
+Michael Weinold further gratefully acknowledges financial support from the Swiss Study Foundation.
+We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
+We also express our deep gratitude to all interviewees for their willingness to participate in this study and their invaluable contributions.
 

### 2023-01-18 2_SSL_EES/article/article_main.tex
updated two figures in the introduction section

@@
 \centering
-  \includegraphics[height=5cm]{2_SSL_EES/article/figures/cost_lamp_small.pdf}
+  \includegraphics[height=6.5cm]{2_SSL_EES/article/figures/cost_lamp_small.pdf}
   \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
@@
 
-The historical development of light-emitting diodes has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015} \cite{Nakamura2015}.
-The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006} \cite{krames2007led} \cite{laubsch2009high} \cite{hahn2014development}, or contributions of different sub-efficiencies to overall device efficiency for 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
+The historical development of light-emitting diodes has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
+The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006}\cite{krames2007led}\cite{laubsch2009high}\cite{hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
 
@@
 
-We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure 3, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure \ref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
 Despite the amount of literature published on the topic of LED history, to the best of our knowledge, no publication comprehensively aggregates known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
-In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating} \cite{kolesnikov2020novel} \cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not been studied in the context of lighting yet.
-This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012}, \cite{Creutzig2018}.
+In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not been studied in the context of lighting yet.
+This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012}\cite{Creutzig2018}.
 

### 2023-01-18 2_SSL_EES/article/article_main.tex
updated figure description
added content to main document

@@
 Data shown for best performers by year of market introduction, with dashed lines.
-A multi-order polynomial fit was used to indicate the average rate of change as dashed lines.
+Dashed lines indicate the average improvement calculated from a 3rd-order polynomial fit to the data.
 The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
@@
 
-Several descriptive papers include very useful information on some of the scientific breakthroughs that have contributed to LED progress \cite{krames2007status}, \cite{Phillips2007}, \cite{Bierhuizen2007}, \cite{Nakamura2013}, \cite{feezell2018invention} and \cite{Taki2019}.
-Additional  studies providing more detail on specific advances in LED technology were also published by authors working for key industry actors Lumileds\cite{MuellerMach2005}, \cite{Shchekin2006}, \cite{lumi2015lumi}, \cite{Bhardwaj2017} and Osram (now amsOsram) \cite{Haerle2004}, \cite{Baur2009}, \cite{laubsch2009high}, \cite{hahn2014development}.
-However, these studies typically include limited information on the macro chip-design and cover disparate aspects of the technology over different time periods.
+Several descriptive publications provide a very useful overview of some of the scientific breakthroughs  that have contributed to LED progress \cite{krames2007status}\cite{Phillips2007}\cite{Bierhuizen2007}\cite{Nakamura2013}\cite{feezell2018invention}\cite{Taki2019} .
+Additional  studies providing more detail the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram) \cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
+However, these studies typically include limited information on the macro-level chip-design and cover disparate aspects of the technology over different time periods.
 Another important source of historical information comes from patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds \cite{margalith2011thin}, Samsung \cite{jung2014phos} \cite{cha2019semiconductor} and Soraa \cite{cich2017high}.
@@
 \end{figure*}
+
+\section{Metrics for technological progress in LEDs}
+
+Investigating the sources of rapid progress in LED technology over the past decades, in which white LEDs have come to dominate the lighting market \cite{zissis2021}, requires selecting appropriate metrics for tracking and quantifying this progress.
+The choice of metrics affects both what data sources can be used in the analysis and which research methodologies can be used to calculate and analyse such metrics.
+We select the metrics based on the following two general criteria.
+First, we focus only on metrics widely accepted and reported in industry because metrics proposed in the scientific literature but not reported by device manufacturers cannot be used to compare the performance of commercial LED devices over time.
+Second, the chosen metrics must be useful for understanding the impact of individual technological improvements on relevant performance and cost characteristics.
+
+Historically, the progress in LED technology is commonly described by pointing to impressive improvements in LED device performance (i.e., brightness and electrical efficiency) and manufacturing cost reductions \cite{Taki2019}.
+However, solid-state lighting performance is characterized not only by LED brightness or electrical efficiency.
+Consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, has also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}.
+Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
+Next, we introduce and discuss the metrics that we use to track progress in each of these three areas.
+
+\subsection{Device Performance Metrics}
+
+Initially, as LED devices developed, the primary metric of progress in solid-state lighting was typically the luminous flux, as reflected in the so-called “Haitz’s Law” \cite{haitz1999case}\cite{haitz2011solid}.
+The luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
+In 2000, highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm) \cite{haitz2011solid}.
+Today, however, highest performing devices yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
+LED brightness has become sufficient to enable the construction of lamps from multiple LED devices, with design improvements focusing on higher efficiency and quality of light instead of brightness.
+Even though some scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, by now, it fails to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
+
+\subsection{Consumer Experience Metrics}
+
+The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
+Any metric relevant to customer experience can thus be calculated from the spectrum alone.
+The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the downconversion phosphor used in the device.
+It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
+Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: Colour Rendering Index (CRI) and Colour Temperature.
+We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union /cite{weinold2020long}.
+This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
+It is thus beyond the scope of this work.
+
+The Colour Rendering Index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
+The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
+CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
+However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de-facto industry standard for describing colour fidelity of light sources \cite{DoE2016LED}.
+High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
+It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
+However, these niche applications remain outside our focus on general illumination.
+Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
+
+The Colour Temperature of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
+Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
+Early white LEDs produced only cool white light \cite{mueller2000light}.
+The introduction of first commercial warm white LED light bulbs played a significant part in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
+For this reason, we also adopt colour temperature as a metric for tracking progress in consumer experience in LED lighting.
+
+\subsection{Manufacturing Cost}
+
+In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must first highlight  the complexity of this task.
+Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
+Using sales price information instead of the cost for the same purpose could be in some cases a promising alternative, as prices can be easily obtained for recent LED products directly from manufacturers.
+However, historical data on prices for LEDs is as difficult to obtain as for costs.
+In addition, sales price includes components not relevant to the progress in technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
+Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
+
+\section{Methods and Data Collection}
+
+The evolution of LED device architecture and performance as well as the progress in understanding the underlying physical phenomena are well covered in the scholarly literature and patents, as described in section 2.
+However, information provided in such sources is insufficient for our goals on at least three accounts.
+First, existing work focusses  only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
+Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
+Second, not all relevant innovations are patented \cite{Pakes_1980}\cite{ Fontana_2013}.
+In the case of LED patents in particular, our interviews with industry experts suggest that the propensity to patent is the highest for knowledge related to macroscopic device architecture and chemical composition of phosphors, and the lowest for knowledge related to manufacturing process improvements and microscopic chip architecture that is difficult to reconstruct by reverse engineering.
+This means that relying only on patent literature would bias results by unduly emphasizing some focus areas and de-emphasizing others.
+Third, scientific publications and patents typically focus on experimental devices, rather than commercial products.
+While new LED features, designs and manufacturing methods reported in these sources can potentially result in significant performance gains or cost reductions, it is difficult to ascertain if these improvements have since been adopted in industry.
+Furthermore, information on LED manufacturing cost and the effect of process improvements on the total cost is highly proprietary.
+Estimates are occasionally reported in the scientific literature and company publications, but these often do not disclose which parts of the manufacturing process are responsible for the largest contribution to the overall cost, or which improvements led to cost reductions.
+
+To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis that combines information obtained from a review of the primary scientific literature and relevant patents with our own computations of device sub-efficiencies, information gained from industry publications, semi-structured interviews with experts from academia and industry, and bottom-up manufacturing cost modelling.
+We provide the details of implementation of each of these methods below.
+
+\subsection{Systematic Literature Review}
+
+We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
+This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in the “Metrics” section; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
+
+Relevant  sources for the review were found in an iterative search process that involved two components.
+The first was the search in specialized patent and publication databases such as Google Scholar, Google Patents, and Scopus , as well as company websites.
+The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the “Previous literature” section and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+We also relied on backward citations for the identification of technology spillovers,   considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
+
+\subsection{Semi-Structured Interviews}
+
+To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews with eleven eminent experts  from academia, industry and public research sector .
+Experts were selected based on their engagement in  different sub-fields of LED research and manufacturing, as well as the recommendations from other interviewed experts, in essence expanding the list of experts that emerged from the literature review by what is known as a snow-ball sampling method.
+All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
+A summary of the background of interviewed experts is provided in Table X in the supplementary information.
+
+The primary, structured part of the interviews explored which innovations were deemed most relevant to the evolution of device performance, consumer experience and manufacturing cost of LED packages.
+Thereafter, interviewees were asked to consider the extent to which these innovations may have originated outside of their respective field of expertise and the LED industry more broadly—i.e., which of the innovations may be considered spillovers.
+The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
+Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in subsection 4.4) and verify device performance data.
 

### 2023-01-18 2_SSL_EES/article/article_supplementary.tex
added content to main document

@@
 
+\section{Anonymized List of Interviewees}
+
+\begin{table}[!ht]
+    \centering
+    \begin{tabular}{|l|l|l|l|l|}
+    \hline
+        \# & Sector  & Role  & Country  & Expertise \\ \hline
+        1 & R\&D (Academia) & Researcher & UK & Epitaxy \\ \hline
+        2 & R\&D (Industry) & Consultant & USA & Device Architecture \\ \hline
+        3 & R\&D (Industry) & Head of R\&D & Germany & Epitaxy \\ \hline
+        4 & R\&D (Academia) & Professor & Austria & Phosphors \\ \hline
+        5 & R\&D (Industry) & Head of R\&D & USA & Device Architecture \\ \hline
+        6 & Consulting & Consultant & USA & Device Architecture \\ \hline
+        7 & R\&D (Academia) & Professor & Germany & Phosphors \\ \hline
+        8 & R\&D (Government) & R\&D Manager  & USA & Device Architecture \\ \hline
+        9 & Consulting & Consultant & USA & Device Applications \\ \hline
+        10 & R\&D (Academia) & Professor & France & Device Physics \\ \hline
+        11 & R\&D (Industry) & Head of R\&D & USA & Device Architecture \\ \hline
+        12 & R\&D (Industry) & Head of R\&D & Germany & Phosphors \\ \hline
+        13 & R\&D (Industry) & Head of R\&D & USA  & Phosphors \\ \hline
+    \end{tabular}
+    \label{tab:interviews}
+\end{table}
+
 \newpage

### 2023-01-18 2_SSL_EES/article/article_supplementary.tex
updated interviewees table

--- 2023-01-18 2_SSL_EES/article/article_supplementary.tex
@@
 
-The number of die per wafer $D$ depends on the total usable wafer area.
-The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
+The number of die per wafer $D$ depends on the total USAble wafer area.
+The USAble area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
 
 \begin{equation}
-	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
+	A_{\text{USAble}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
 \end{equation}
 
-Determining the usable wafer area as a of these three parameters requires a numerical solution.
+Determining the USAble wafer area as a of these three parameters requires a numerical solution.
 However, following discussions in literature \cite{de2005investigation}, we approximate the number of good die per wafer as
@@
 
-Devices may be damaged or otherwise rendered unusable during the manufacturing process.
+Devices may be damaged or otherwise rendered unUSAble during the manufacturing process.
 The ratio between the number of good devices per step and the number of handled devices per step is known as the yield.
@@
 
+\section{Anonymized List of Interviewees}
+
+\begin{table}[!ht]
+    \centering
+    \begin{tabular}{|l|l|l|l|l|}
+    \hline
+        \# & sector & role & country & expertise \\ \hline
+        1 & academia & sr. researcher & UK & epitaxy \\ \hline
+        2 & industry & consultant, form. sr. researcher & USA & device architecture \\ \hline
+        3 & industry & consultant, form. head of R\&D & Germany & epitaxy \\ \hline
+        4 & academia & professor & Austria & phosphors \\ \hline
+        5 & industry & consultant, form. head of R\&D & USA & device architecture \\ \hline
+        6 & consulting & consultant, form. sr. technical advisor & USA & device architecture \\ \hline
+        7 & academia & professor & Germany & phosphors \\ \hline
+        8 & government & R\&D manager & USA & device architecture \\ \hline
+        9 & consulting & consultant & USA & device applications \\ \hline
+        10 & academia & professor & France & device physics \\ \hline
+        11 & industry & sr scientist, form. head of R\&D & USA & device architecture \\ \hline
+        12 & industry & principal scientist & Germany & phosphors \\ \hline
+        13 & industry & form. head of R\&D & USA & phosphors \\ \hline
+    \end{tabular}
+    \label{tab:interviews}
+    \caption{list of domain experts with whom semi-structured interviews were conducted. \\ Abbreviations: form. - former, sr. - senior}
+\end{table}
+
 \newpage

### 2023-01-18 2_SSL_EES/article/article_main.tex
updated subefficiencies description

@@
 
+\subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
+
+Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
+\cite{tsao2010solid}} $V_fE$ describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
+These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
+It is defined as
+
+\begin{equation}
+    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
+\end{equation}
+
+where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Light Extraction Efficiency $\eta_{LE}$}
+
+Device light-extraction efficiency describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+
+\begin{equation}
+    \eta_{LE}=\frac{\text{\# of photons out}}{\text{\# of photons created}} = \frac{P_{opt}}{P_{int}}
+\end{equation}
+
+where $P_{opt}$ is the optical power of the device and $P_{int}$ is the internal power of the device\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Internal Quantum Efficiency $\eta_{IQ}$}
+
+Internal quantum efficiency describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
+\begin{equation}
+\label{eqn:ieq-eqe}
+    \eta_{EQ} = \eta_{IQ} \times \eta_{LE}
+\end{equation}
+
+and defined as
+
+\begin{equation}
+    \eta_{IQ} = \frac{\# \text{of photons created}}{\# \text{of electron-hole pairs in}}
+\end{equation}
+
+and depends on the current density of the device\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Droop $\eta_{droop}$}
+
+(Efficiency) droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
+This is often treated separately from internal quantum efficiency in literature due to its importance in high-power devices.
+It is defined as
+
+\begin{equation}
+    \eta_{droop} = 1 - \frac{\eta_{IQE}}{\eta_{IQE}(A \rightarrow 0)}
+\end{equation}
+
+where $\eta_{IQE}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
+In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$ as
+
+\begin{equation}
+\label{eqn:droop}
+    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
+\end{equation}
+
+Thereby, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Conversion Efficiency $\eta_{CE}$}
+
+(Light) conversion efficiency describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
+These losses are the sum of the Stokes loss and well as scattering/absorption losses.
+It is defined as
+
+\begin{equation}
+    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i}
+\end{equation}
+
+where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength (Blue, Red, Orange, Yellow, Green)\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Spectral Efficiency $\eta_{S}$}
+
+Spectral efficiency describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
+Photons converted into the infrared or ultraviolet are lost to illumination purposes.
+It is defined as
+
+\begin{equation}
+    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
+\end{equation}
+
+where $K$ is the luminous efficacy of radiation of the light source, can be computed from the spectrum and the luminosity function, which describes the sensitivity of the human eye.$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
+
 \subsection{Consumer Experience Metrics}

### 2023-01-19 2_SSL_EES/article/article_main.tex
added figures to results section

@@
 \subsection{Device Performance Metrics}
+\label{sec:device_performance_metrics}
 
@@
 
-\subsubsection{Conversion Efficiency $\eta_{CE}$}
+\subsubsection{Conversion Efficiency $\eta_{C}$}
 
@@
 
+\subsection{Performance Metrics Calculations}
+
+The contribution of individual technology innovations and spillovers to the progress in the overall device efficiency over time is estimated by the index decomposition analysis, which is commonly used in energy economics \cite{Ang2019}.
+Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
+It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia index, a method in statistical economics \cite{Diewert1988}, and subsequently refined \cite{Ang1997}.
+
+\section{Results}
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
+ \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003 (top panel) and 2010 (bottom panel), normalized to 1 Watt of electric power input (yellow bar on the left).
+Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}.
+Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts).
+Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
+The following corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) – flip-chip with YGAG phosphor; 2010 (bottom panel) – flip-chip with 258 phosphor.
+Details on phosphors are provided in Table XX and SI section XX.
+Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.6\%$ in 2010.
+Abbreviations: Scat.
+= Scattering.
+Source: own elaboration based on data from sources listed in SI, section XX.}
+ \label{fgr:waterfall_1}
+\end{figure*}
+
 \section{Conclusions}

### 2023-01-19 2_SSL_EES/article/article_main.tex
added spillover table

@@
 %%%Please don't disable any packages in the preamble, as this may cause the template to display incorrectly.%%%
+
+\usepackage{tabularx}
 
@@
 
-%Please use \dag to cite the ESI in the main text of the article.
+%Please use \dag to\cite the ESI in the main text of the article.
 %If you article does not have ESI please remove the the \dag symbol from the title and the footnotetext below.
@@
  \centering
- \includegraphics[height=7cm]{2_SSL_EES/article/figures/history_efficacy.pdf}
+ \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/history_efficacy.pdf}
  \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020.
@@
 Relevant  sources for the review were found in an iterative search process that involved two components.
-The first was the search in specialized patent and publication databases such as Google Scholar, Google Patents, and Scopus , as well as company websites.
+The first was the search in specialized patent and publication databases such as Google Scholar, Google Patents, and Scopus, as well as company websites.
 The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the “Previous literature” section and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
@@
 \end{figure*}
+
+Some text here
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
+ \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
+Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
+Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for five years: 1997, 2003, 2010, 2016, and 2020.
+Physical limits on sub-efficiencies are indicated by black horizontal lines.
+The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
+$\eta_Vf$ – forward voltage efficiency; $\eta_IQ$ – internal quantum efficiency; $\eta_Droop$ – efficiency droop; $\eta_LE$ – light extraction efficiency; $\eta_(C_ER)$ – conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ – conversion efficiency for yellow/green phosphors; $\eta_S$ – spectral efficiency; PSS – patterned sapphire substrate, ITO – indium tin oxide current spreading layer; 258, SLA, PFS – different phosphors used in white LEDs, see Table 3.
+"N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
+Source: own elaboration based on data from sources listed in SI, section XX.}
+ \label{fgr:breakthroughs_efficiency}
+\end{figure*}
+
+\subsection{Improvements in Consumer Experience Metrics}
+
+Historical improvements in consumer experience metrics for phosphor-converted warm white LEDs are shown in Figure \ref{fgr:consumer_experience}.
+A high colour rendering index (CRI) of an LED in combination with a specific range of possible colour temperatures is generally desirable in general illumination applications.
+The values of these metrics are determined by LED device spectra (provided and discussed in SI section X), which, in turn, depend on the means of light down conversion in the device.
+Due to that, we were able to establish the links between all major improvements in LED consumer experience metrics identified in this study and individual LED inventions associated either with phosphors or quantum dots used for light down conversion.
+The list of these inventions is provided in Table 3.
+
+Notably, from detailed descriptions of the history of inventions in this list, which we provide in SI section 3, we find that only a single invention related to LED consumer experience improvements was originally developed specifically for application in solid-state lighting: the 2016 SALON phosphor compound \cite{seibald2019phosphor}\cite{Hoerder2019}.
+All other inventions in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL.
+We discuss the details of corresponding technology spillovers in section XX below.
+
+\begin{table*}[h]
+    \small
+    \centering
+    \begin{tabularx}{\textwidth}{|l|l|l|X|X|l|}
+    \hline
+        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{SI Sec.} \\ \hline
+        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.1 \\ \hline
+        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.1 \\ \hline
+        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3.2 \\ \hline
+        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3.4 \\ \hline
+        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3.3 \\ \hline
+        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3.2 \\ \hline
+        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2 \\ \hline
+    \end{tabularx}
+    \label{tab:phosphors}
+    \caption{LED light down conversion inventions related to improvements in consumer experience metrics identified in this study.
+Note: Year column represents the earliest reported application of invention in white LEDs.
+These years are different from the years in Figure \ref{fgr:consumer_experience} that correspond to the earliest publication of spectral data for representative LED products  that relied on those inventions.
+Section in SI refers to a section in the Supplementary Information that contains references and a detailed description of the history of corresponding invention.
+Abbreviations: Desig.
+- Designation, Sec.
+- Section.}
+\end{table*}
+
+\begin{figure*}[h]
+ \centering
+ \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
+ \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
+Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
+Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for five years: 1997, 2003, 2010, 2016, and 2020.
+Physical limits on sub-efficiencies are indicated by black horizontal lines.
+The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
+$\eta_Vf$ – forward voltage efficiency; $\eta_IQ$ – internal quantum efficiency; $\eta_Droop$ – efficiency droop; $\eta_LE$ – light extraction efficiency; $\eta_(C_ER)$ – conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ – conversion efficiency for yellow/green phosphors; $\eta_S$ – spectral efficiency; PSS – patterned sapphire substrate, ITO – indium tin oxide current spreading layer; 258, SLA, PFS – different phosphors used in white LEDs, see Table 3.
+"N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
+Source: own elaboration based on data from sources listed in SI, section XX.}
+ \label{fgr:consumer_experience}
+\end{figure*}
+
+\subsection{Technology Spillovers}
+
+\begin{table*}[h]
+    \small
+    \centering
+    \begin{tabularx}{\textwidth}{|l|l|l|X|X|X|l|X|}
+    \hline
+        \textbf{Disc.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
+        1926 & 1994 & 1996 & LED phosphors & Use of phosphors for light down conversion in LEDs & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric}\cite{shimizu1994sheet}\cite{cho2017white} & Enabled light down conversion in LEDs \\ \hline
+        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new}\cite{bando1996}\cite{bando1998development}\cite{shimizu1999light}\cite{cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
+        1967 & 1996 & <2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical}\cite{bando1998development}\cite{shimizu1999light}\cite{Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
+        1982 & 1996 & 2015 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered} & $\eta_{LE}$, $\eta_{IQ}$ \\ \hline
+        1971 & 1999 & <2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) &\cite{margalith1999indium} & $\eta_{Vf}$ \\ \hline
+        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997}\cite{mueller2004phosphor}\cite{MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
+        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo}\cite{simmonsfinal}\cite{ledprof_nexxusqd}\cite{bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
+        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & Chemistry (S), Materials science (S) &\cite{paulusz1973efficient}\cite{radkov2009red}\cite{Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
+        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New}\cite{schmidt2013new}\cite{Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
+    \end{tabularx}
+    \label{tab:spillovers}
+    \caption{Technology spillovers involved in white LED technology innovations identified in this study.
+Note: LED innovations are ordered by the year in which a technology spillover into LED occurred, provided in the Spillover to LED column.
+Initial discovery is the year of the earliest identified discovery of the original idea or invention outside the LED domain.
+Commercial application is the year of the first recorded application of that idea or invention in a commercial LED product.
+Origin column represents knowledge domains in which spillovers initially emerged, where (S) denotes a scientific discipline and (T) is an area of technology.
+References column lists literature sources for the represented innovations and spillovers.
+Area of improvement represents the impact of spillovers on different aspects of white LED technology, e.g., improvements in particular sub-efficiencies.
+Abbreviations: Disc.
+- Year of Discovery, S/O - Year of Spillover to LED, Comm.
+- Year of Commercial Application., Ref.
+- References.}
+\end{table*}
 

### 2023-01-19 2_SSL_EES/article/article_main.tex
added consumer experience figure

@@
  \centering
- \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
- \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
-Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
-Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for five years: 1997, 2003, 2010, 2016, and 2020.
-Physical limits on sub-efficiencies are indicated by black horizontal lines.
-The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
-$\eta_Vf$ – forward voltage efficiency; $\eta_IQ$ – internal quantum efficiency; $\eta_Droop$ – efficiency droop; $\eta_LE$ – light extraction efficiency; $\eta_(C_ER)$ – conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ – conversion efficiency for yellow/green phosphors; $\eta_S$ – spectral efficiency; PSS – patterned sapphire substrate, ITO – indium tin oxide current spreading layer; 258, SLA, PFS – different phosphors used in white LEDs, see Table 3.
-"N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
-Source: own elaboration based on data from sources listed in SI, section XX.}
+ \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_consumer-experience.pdf}
+ \caption{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.
+Circle dots show data for the earliest identified representative white LED products with published spectral data that used phosphors listed in Table 3, each indicated by the phosphor label and publication year.
+Corresponding spectral data used to calculate colour temperature values is provided in Figure SI1 in SI section 2.
+Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison.
+Horizontal lines represent typical colour temperatures of “traditional” light sources, shown for reference.
+Desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures.
+Horizontal red arrow indicates desirable higher values of colour rendering index (CRI).
+LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG – Nichia, 1998 \cite{bando1998development}; YGAG – Lumileds, 2002 \cite{Mueller2002}; 258 – Lumileds, 2005 \cite{MuellerMach2005}; SLA – Lumileds, 2014 \cite{Pust2014}; PFS – GE, 2015 \cite{Murphy2015}; QD – Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON – Osram, 2019 \cite{Hoerder2019}.}
  \label{fgr:consumer_experience}

### 2023-01-19 2_SSL_EES/article/article_main.tex
added equation

@@
 
-The contribution of individual technology innovations and spillovers to the progress in the overall device efficiency over time is estimated by the index decomposition analysis, which is commonly used in energy economics \cite{Ang2019}.
+The contribution of individual technology innovations and spillovers to the progress in the overall device efficiency over time is estimated by the index decomposition analysis, which is commonly used in energy economics \cite{Ang1997}.
 Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
-It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia index, a method in statistical economics \cite{Diewert1988}, and subsequently refined \cite{Ang1997}.
+It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
+
+According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change  in a single sub-efficiency variable a between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
+
+\begin{align}
+    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
+    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
+\end{align}
+
+where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
+These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:
+
+\begin{equation}
+    \Delta a + \Delta b  = \Delta F
+\end{equation}
+
+To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered.
+However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance.
+For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies?
+where possible and necessary.
+Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications.
+In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra.
+This approach allowed us to fully quantify the improvements related to phosphor development in LEDs.
 
@@
 The following corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) – flip-chip with YGAG phosphor; 2010 (bottom panel) – flip-chip with 258 phosphor.
-Details on phosphors are provided in Table XX and SI section XX.
-Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.6\%$ in 2010.
+Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX.
+Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.7\%$ in 2010.
 Abbreviations: Scat.
@@
  \label{fgr:waterfall_1}
+\end{figure*}
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
+ \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
+ \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2016 (top panel) and 2020 (bottom panel), normalized to 1 Watt of electric power input (yellow bar on the left).
+Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}.
+Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) a
+ t the point of the channel and absolute values (in Watts).
+Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
+The following corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2016 (top panel) – flip-chip with PFS phosphor; 2020 (bottom panel) – flip-chip with SALON phosphor.
+Details on phosphors are provided in Table \ref{tab:phosphors} and SI section 2.
+Overall LED package efficiency $\eta_L= 32.5\%$ in 2016, $\eta_L= 38.7\%$ in 2020.
+Abbreviations: Scat.
+= Scattering.
+Source: own elaboration based on data from sources listed in SI, section XX.}
+ \label{fgr:waterfall_2}
 \end{figure*}
@@
     \end{tabularx}
-    \label{tab:phosphors}
     \caption{LED light down conversion inventions related to improvements in consumer experience metrics identified in this study.
@@
 - Section.}
+    \label{tab:phosphors}
 \end{table*}
@@
     \end{tabularx}
-    \label{tab:spillovers}
     \caption{Technology spillovers involved in white LED technology innovations identified in this study.
@@
 - References.}
+    \label{tab:spillovers}
 \end{table*}

### 2023-01-19 2_SSL_EES/article/article_main.tex
updated supplementary information

@@
  \centering
- \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
- \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
  \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003 (top panel) and 2010 (bottom panel), normalized to 1 Watt of electric power input (yellow bar on the left).
@@
 We also express our deep gratitude to all interviewees for their willingness to participate in this study and their invaluable contributions.
+Support by German consumer organization Stiftung Warentest in providing access to their archive is gratefully acknowledged.
 

### 2023-01-19 2_SSL_EES/article/article_supplementary.tex
updated supplementary information

@@
 \usepackage{graphicx}
+\usepackage{float}
 % math
@@
 \usepackage{fancybox}
+% tikz
+\usepackage{tikz}
+\usetikzlibrary{backgrounds}
+\usetikzlibrary{arrows}
+\usetikzlibrary{shapes,shapes.geometric,shapes.misc}
+
+% this style is applied by default to any tikzpicture included via \tikzfig
+\tikzstyle{tikzfig}=[baseline=-0.25em,scale=0.5]
+
+% these are dummy properties used by TikZiT, but ignored by LaTex
+\pgfkeys{/tikz/tikzit fill/.initial=0}
+\pgfkeys{/tikz/tikzit draw/.initial=0}
+\pgfkeys{/tikz/tikzit shape/.initial=0}
+\pgfkeys{/tikz/tikzit category/.initial=0}
+
+% standard layers used in .tikz files
+\pgfdeclarelayer{edgelayer}
+\pgfdeclarelayer{nodelayer}
+\pgfsetlayers{background,edgelayer,nodelayer,main}
+
+% style for blank nodes
+\tikzstyle{none}=[inner sep=0mm]
+
+% include a .tikz file
+\newcommand{\tikzfig}[1]{%
+{\tikzstyle{every picture}=[tikzfig]
+\IfFileExists{#1.tikz}
+  {\input{#1.tikz}}
+  {%
+    \IfFileExists{./figures/#1.tikz}
+      {\input{./figures/#1.tikz}}
+      {\tikz[baseline=-0.5em]{\node[draw=red,font=\color{red},fill=red!10!white] {\textit{#1}};}}%
+  }}%
+}
+
+% the same as \tikzfig, but in a {center} environment
+\newcommand{\ctikzfig}[1]{%
+\begin{center}\rm
+  \tikzfig{#1}
+\end{center}}
+
+% fix strange self-loops, which are PGF/TikZ default
+\tikzstyle{every loop}=[]
+\input{./auxiliary/tikz_style_files/main.tikzstyles}
 
@@
 
-\section{Manufacturing Cost Model: Details on computation}
+\section{Manufacturing Cost Model}
 
-\subsection{Total manufacturing cost}
+\subsection{Structure of the Model}
+
+\begin{figure}[h!]
+    \tikzfig{./figures/costmodel}
+    \centering
+    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to cost model outputs.
+For description of cost variables, see definitions for equation \ref{eqn:cost_wafer}.}
+    \label{fig:costmodel-schematic}
+    \vspace{-8pt}
+\end{figure}
+
+The cost model adapted for this publication is a microeconomic manufacturing cost model.
+Within the timeframe and scope laid out in the main publication, it returns the total manufacturing cost of phosphor converted warm white light-emitting diode packages.
+In this computation, it considers the main economic factors associated with operating and maintaining manufacturing equipment.
+It does not consider costs associated with research and development or those associated with the construction of manufacturing facilities.
+It considers market trends through their effect on manufacturing parameters.
+
+A schematic diagram of the cost model developed in this thesis is presented in figure \ref{fig:costmodel-schematic}.
+The cost model is process step based.
+It is split between the upstream part of the manufacturing process and the downstream part of the manufacturing process.
+The former is performed at the wafer level.
+The latter is performed at the package level.
+The model takes as inputs both parameters specific to manufacturing process steps (\textit{"step parameters"}) and parameters applying to all manufacturing steps (\textit{"global steps"}).
+The cost for each process step is then computed.
+The cost model returns the total and step manufacturing costs.
+It further considers the yield per step and returns the cumulative yield, the yielded cost per step and the yielded total manufacturing cost.
+
+\subsection{Computation of Manufacturing Cost}
 
@@
 
-\subsection{Yielded Cost}
+\subsection{Computation of Yielded Cost}
 
@@
 
-\subsection{Contribution of Variables}
+\subsection{Computation of the Contribution of Variables}
 
@@
 
-\section{Contribution of spillovers}
+\subsection{Preliminary Sensitivity Analysis}
 
-    \begin{table}[h!]
-        \centering
-    	\begin{tabularx}{\textwidth}{|l|l|X|X|X|X|}
-    		\hline
-    			Year & $\eta_L$[\%]  & $\Delta \eta_L^{abs}$[\%] & $\Delta \eta_L^{rel}$[\%]  & $\Delta \eta_L^{abs}(S/O)$[\%]  & $\Delta \eta_L^{rel}(S/O)$[\%]  \\
-    		\hline
-    			2002 & 5.8 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
-    		\hline
-    		    \multicolumn{2}{|c|}{2002-2010} & 6.8 & 117.4 & 1.0 & 14.4 \\
-    		\hline
-    			2010 & 12.5 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
-    		\hline
-    		    \multicolumn{2}{|c|}{2010-2016} & 19.9 & 158 & 1.6 & 8.0 \\
-    		\hline
-    			2016 & 32.5 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
-    		\hline
-    		    \multicolumn{2}{|c|}{2016-2020} & 6.2 & 19.0 & 0.3 & 3.9 \\
-    		\hline
-    			2020 & 38.7 & \multicolumn{4}{c|}{\cellcolor{silver}} \\
-    		\hline
-    		    \multicolumn{2}{|c|}{2002-2020} & 32.9 & 566.4 & 2.8 & 8.5 \\
-            \hline
-    	\end{tabularx}
-    	\caption{Contribution of technology spillovers to improvements in the total lamp efficiency $\eta_L$ for warm white light-emitting diodes in the period 2002-2020.
-Note numbers are rounded.
-\\ $\Delta \eta_L^{abs}\overset{eg.}{=}\eta_L(2010)-\eta_L(2002)$ is absolute change in total lamp efficiency.
-\\ $\Delta \eta_L^{rel}\overset{eg.}{=}(\eta_L(2010)-\eta_L(2002))/(\eta_L(2002)/100)$ is relative change in lamp efficiency.}
-    	\label{tab:spillover-data}
-    \end{table}
-
-\section{Cost model sensitivity analysis}
-
-\begin{figure}
+\begin{figure}[h]
 	\centering
@@
 	\caption{Panels a-b: Sensitivity analysis of selected cost model parameters. Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.}
-	\label{fig:costmodel_calibration}
+	\label{fig:sensitivity}
 \end{figure}
 
-Sensitivity analysis of selected cost model parameters. Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.
+A sensitivity analysis of the cost model has been performed using parameter variations listed in table \ref{tab:sensitivity}.
+The amount of parameter variation was chosen to encompass the identified range of the value of the parameter in different manufacturing setups of each considered year.
+As an example, the lower range in the variation for metal cost (+50\%,-0\%) was chosen because the metal prices in the model are price quotes from the United State Geological Survey price database.
+Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
+The upper range was chosen, because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
+The results of the analysis are shown in figure \ref{fig:sensitivity}.
+The cost model is generally more sensitive to variation of parameters at smaller wafer diameters.
+The most sensitive parameters are global parameters, such as yield or average equipment throughput.
 
-A preliminary sensitivity analysis (panels a-b of Figure XX) confirms these preliminary findings, showing the largest deviation for changes in these parameters.
+\begin{table}[H]
+\small
+    \begin{tabularx}{\textwidth}{ |X|l|l|l|l|l|l|l|X|}
+        \hline
+            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & \textit{Source} \\
+        \hline
+            Cleanroom Cost & \$/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & Figure \ref{fig:electricity+cleanroom_prices} \\
+        \hline
+            Manhours & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
+        \hline
+            Equip. Discount & \% of \$ & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I, \cite{Appleyard2001} \\
+        \hline
+            Overall Yield & \% & 100\% & +25,-25 & 100\% & +25,-15 & 100\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
+        \hline
+            Inspec. Yield Savings & \%/inspec. & 0.5\% & +80,-80 & 0.5\% & +80,-80 & 0.5\% & +80,-80 & \cite{mckinseyyield} \\
+        \hline
+            Overall Throughput & UPH or h$^{-1}$ & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & Datasheets \\
+        \hline
+            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I, Figure \ref{fig:wafer_size} \\
+        \hline
+            Edge Exclusion & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
+        \hline
+            Cutting Width & $\mu$m & 100 & +50,-25 & 75 & +33.3,-20 & 20 & +300,-50 & \cite{masaki2000division}\cite{ils2005width} \newline \cite{photonics2010width}\cite{discowidth} \\
+        \hline
+            Metal Prices & \$/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
+        \hline
+            Electricity Price & \$/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
+        \hline
+            Saph. Subst. Price & \$ & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & Figure \ref{fig:sapphire_prices} \\
+        \hline
+            Phosphor Prices & \$/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
+        \hline
+        \end{tabularx}
+    \caption{Cost model sensitivity analysis parameter list.
+For the sensitivity analysis performed using these variables, see figure \ref{fig:sensitivity}.
+Units for values in columns \textit{2002}-\textit{2020} are indicated in column \textit{Units}.
+If values in columns \textit{2003}-\textit{2020} are instead are given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.}
+    \label{tab:sensitivity}
+\end{table}
 
-\section{Cost model comparison with reported industry data}
+\subsection{Comparison with Reported Industry Data}
 
-\begin{figure}
+We note differences between our model and costs reported to the US Department of Energy (DoE) by industry.
+For instance, the share of the epitaxy step is larger in the DoE model.
+This can in part be explained by our model relying on state-of-the-art equipment, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+The packaging part of the manufacturing process is increasing in importance, which has been confirmed by researchers and industry reports on wafer-level packaging.
+
+\begin{figure}[h]
 	\centering
@@
 
-We note differences between our model and costs reported to the US Department of Energy (DoE) by industry.
-For instance, the share of the epitaxy step is larger in the DoE model.
-This can in part be explained by our model relying on state-of-the-art equipment, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
-The packaging part of the manufacturing process is increasing in importance, which has been confirmed by researchers and industry reports on wafer-level packaging.
+\section{Phosphors}
+
+\subsection{Spectral Data of Identifies LED Phosphor Innovations}
+
+\begin{figure}[h]
+	\centering
+    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/phosphor_spectrum-comparison.pdf}
+	\caption{Modelled manufacturing cost for light-emitting diode packages split by manufacturing step category.
+The top graph shows the data from LEDCOM model and round table discussions published by the United States Department of Energy (DoE).
+The bottom graph shows the results of the cost model developed in this thesis.
+Hatched bars are projections given by the sources.
+Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.}
+	\label{fig:costmodel_calibration}
+\end{figure}
+
+\subsection{History of Identifies Phosphor Inventions}
+
+\subsubsection{YAG and YGAG Phosphors}
+
+Prior to Nakamura’s invention of highly efficient blue LEDs in the 1990s, Japan’s Nichia Corporation had not sold commercially successful semiconductor products, instead specializing in phosphors for cathode ray tubes (CRT) and fluorescent lamps \cite{nakamura2013blue}  .
+Nevertheless, extensive firm expertise in this area helped Nichia’s Yoshinori Shimizu formulate the principles of using CRT phosphors to convert blue light from Nakamura’s LEDs into white light in 1994 \cite{shimizu1994sheet},\cite{cho2017white}.
+By 1996, Shimizu and his colleagues developed \cite{bando1996}, \cite{shimizu1999light} the first practical LED application of a well-known Yttrium Aluminium Garnet (YAG) CRT phosphor activated with cerium \cite{blasse1967new}, enabling the first commercial white LED products manufactured and sold by Nichia since late 1996 \cite{bando1998development}, \cite{cho2017white}.
+
+Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 8 in the main text).
+An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development},\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
+Used in combination with red-emitting sulfide phosphors, by 2002 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}.
+However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased.
+New chemically stable and non-toxic red phosphors were needed.
+
+\subsubsection{258, SLA and SALON phosphors}
+
+In 1997, Hubert Huppertz and Wolfgang Schnick, working at the University of Bayreuth in Germany, synthesized the first compound in a new class of rare earth nitridosilicate materials \cite{Huppertz1997} later dubbed “258” due to a proportion of elements in its chemical formula.
+The luminescent properties of these materials were identified by the Schnick’s group, by then at Ludwig-Maximilians University of Munich, in 2000 \cite{Hppe2000} after a suggestion made to Schnick at a conference following earlier reports of good luminosity properties of europium-doped compounds \cite{Qiua1998}.
+U.S.-based LED manufacturer Lumileds applied for a patent for the first class of red LED phosphors based on the 258 nitridosilicate chemistry in 2002 \cite{mueller2004phosphor}.
+The first use of the 258 phosphor in a commercial “Luxeon” LED package was then reported in a joint publication co-authored by inventors from Lumileds and researchers from the Schnick’s group in 2005 \cite{MuellerMach2005}.
+
+Further efforts in LED phosphor development were directed towards synthesizing a red narrow-band phosphor.
+Narrow LED emission peak widths yield the highest luminous efficacy of radiation, as in this case less light is emitted in the far-red range of the spectrum in which the human eye is not very sensitive.
+After synthesizing several narrow-band phosphors emitting in yellow \cite{Hppe2004} and cyan \cite{Kechele2009}, the Schnick’s group identified the local cubic cation coordination structure of the cyan phosphor compound as the reason for its narrow band width \cite{lumi2016narrow}.
+A search for a structurally analogous nitride compound with the narrow red instead of the cyan emission was undertaken.
+After several unsuccessful attempts, the sought-after cuboidal nitride compound was found in a 2008 publication led by Francis DiSalvo \cite{Park2008Sr}.
+Based on information provided in this work, Schnick and colleagues synthesized and studied the spectral properties of a new narrow band red SLA phosphor in 2013 \cite{schmidt2013new}\cite{Pust2014}\cite{schmidt2017phosphors}.
+The material was introduced in commercial LED devices by Lumileds in 2015 \cite{lumi2016narrow_whitepaper}.
+
+The most recent red narrow-band phosphor breakthrough included in Table 3 and Figure 8 in the main text, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}, \cite{Hoerder2019} , \cite{Hoerder2020}.
+The first patent application for this phosphor was filed in 2016.
+The SALON phosphor is a derivate of the SLA phosphor.
+Therefore, it is the only innovation related to consumer experience metrics identified in our study that seemingly not involved technology spillovers.
+
+\subsubsection{PFS Phosphor}
+
+\subsubsection{Quantum Dots for Light Down-Conversion}
 
@@
     \label{tab:interviews}
-    \caption{list of domain experts with whom semi-structured interviews were conducted. \\ Abbreviations: form. - former, sr. - senior}
+    \caption{List of domain experts with whom semi-structured interviews were conducted. \\ Abbreviations: form. - former, sr. - senior}
 \end{table}
@@
 \cite{council2013assessment}
-
-Support by German consumer organization Stiftung Warentest in providing access to their archive is gratefully acknowledged.
 

### 2023-01-19 2_SSL_EES/article/article_main.tex
added content to the main document

@@
 
+\subsection{Bottom-Up Manufacturing Cost Model}
+
+We address the limitations of data availability for the LED manufacturing cost by conducting bottom-up process-step resolution LED manufacturing cost modelling.
+In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
+
+We identified two existing cost models for LED manufacturing in the prior literature.
+The first is the LED Cosim model developed by the technology consultancy System Plus Consulting \cite{cosim2013} \cite{cosim2019}.
+Due to prohibitive licensing costs, this model could not be used in our study.
+The second model is the publicly available 2012 LEDCOM cost model prepared for the U.S.
+Department of Energy by Stephen Bland  \cite{ledcomv2}.
+It was set up to capture the manufacturing process for phosphor-converted LEDs of different power levels.
+However, given the remarkable changes in the performance, cost and architecture of LEDs since 2012, as shown in Figures 1-3, this model cannot fully capture the recent advances in LED technology, necessitating a significant modification to bring it up to date with the current state-of-the-art in LED manufacturing.
+
+Given this situation, we developed our own bottom-up LED manufacturing cost model.
+The model structure is generally based on the 2012 \textit{LEDCOM} cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+For each of these three periods, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
+Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips.
+The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
+In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
+These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
+
 \section{Results}

### 2023-01-19 2_SSL_EES/article/article_main.tex
added equation

@@
 
+The cost model we developed relies on a cumulative approach to yielded cost\cite{becker2001use}.
+In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
+
+\begin{equation}
+    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
+\end{equation}
+
+This cost metric is cumulative by definition, thus
+
+\begin{equation}
+    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
+\end{equation}
+
+Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
+Additional information on details of the computation of cost metrics and the different manufacturing steps is presented in the Supplementary Information.
+
 \section{Results}
@@
  \centering
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/test.pdf}
- \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003 (top panel) and 2010 (bottom panel), normalized to 1 Watt of electric power input (yellow bar on the left).
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
+ \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
 Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}.
@@
 Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
-The following corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) – flip-chip with YGAG phosphor; 2010 (bottom panel) – flip-chip with 258 phosphor.
+The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) – flip-chip with YGAG phosphor; 2010 (bottom panel) – flip-chip with 258 phosphor; 2016 (top panel) – flip-chip with PFS phosphor; 2020 (bottom panel) – flip-chip with SALON phosphor.
 Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX.
@@
 Source: own elaboration based on data from sources listed in SI, section XX.}
- \label{fgr:waterfall_1}
-\end{figure*}
-
-\begin{figure*}[h!]
- \centering
- \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
- \includegraphics[width=16.5cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
- \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2016 (top panel) and 2020 (bottom panel), normalized to 1 Watt of electric power input (yellow bar on the left).
-Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}.
-Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) a
- t the point of the channel and absolute values (in Watts).
-Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
-The following corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2016 (top panel) – flip-chip with PFS phosphor; 2020 (bottom panel) – flip-chip with SALON phosphor.
-Details on phosphors are provided in Table \ref{tab:phosphors} and SI section 2.
-Overall LED package efficiency $\eta_L= 32.5\%$ in 2016, $\eta_L= 38.7\%$ in 2020.
-Abbreviations: Scat.
-= Scattering.
-Source: own elaboration based on data from sources listed in SI, section XX.}
- \label{fgr:waterfall_2}
+ \label{fgr:waterfall}
 \end{figure*}

### 2023-01-20 2_SSL_EES/article/article_main.tex
updated figure caption

@@
 Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot.
-Abbreviations: HID – High-Intensity Discharge; CFL – Compact Fluorescent Lamp; Hal.
-– Halogen, Incd.
-– Incandescent.
+Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal.
+- Halogen, Incd.
+- Incandescent.
 Source: own synthesis of published data based on a visual approach proposed by Azevedo et al.
@@
 While the exact definition of spillovers in literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012}, we follow Stephan et al.
-\cite{Stephan2021} and consider knowledge to be external – thus a spillover –  if it has been developed for application in other technologies, sectors, or scientific disciplines.
+\cite{Stephan2021} and consider knowledge to be external - thus a spillover -  if it has been developed for application in other technologies, sectors, or scientific disciplines.
 There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016}\cite{Stephan2021}\cite{Sun2021}\cite{kolesnikov_framework_2022} the future evolution of technologies.
@@
 Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
-Department of Energy (DoE) \cite{council2013assessment}.
+Department of Energy (DOE) \cite{council2013assessment}.
 Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs.
@@
 Black frames around certain designs indicate chip designs not brought to large-scale production.
-Abbreviations: GaN – Gallium Nitride, ITO – Indium Tin Oxide, MQW – Multiple Quantum Well.
+Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
 Source: adapted and compiled from multiple patents and industry publications.
@@
 Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
-\cite{tsao2010solid}} $V_fE$ describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
+\cite{tsao2010solid}} (VfE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
 These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
@@
 
-Device light-extraction efficiency describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+Device light-extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
 
@@
 
-Internal quantum efficiency describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
+Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
 \begin{equation}
@@
 
-(Light) conversion efficiency describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
+(Light) conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
 These losses are the sum of the Stokes loss and well as scattering/absorption losses.
@@
 
-Spectral efficiency describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
+Spectral efficiency (SE) describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
 Photons converted into the infrared or ultraviolet are lost to illumination purposes.
@@
 CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
-However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de-facto industry standard for describing colour fidelity of light sources \cite{DoE2016LED}.
+However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de-facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
 High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
@@
  \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
  \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
@@
 Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
-The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) – flip-chip with YGAG phosphor; 2010 (bottom panel) – flip-chip with 258 phosphor; 2016 (top panel) – flip-chip with PFS phosphor; 2020 (bottom panel) – flip-chip with SALON phosphor.
+The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) - flip-chip with YGAG phosphor; 2010 (bottom panel) - flip-chip with 258 phosphor; 2016 (top panel) - flip-chip with PFS phosphor; 2020 (bottom panel) - flip-chip with SALON phosphor.
 Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX.
@@
 The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
-$\eta_Vf$ – forward voltage efficiency; $\eta_IQ$ – internal quantum efficiency; $\eta_Droop$ – efficiency droop; $\eta_LE$ – light extraction efficiency; $\eta_(C_ER)$ – conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ – conversion efficiency for yellow/green phosphors; $\eta_S$ – spectral efficiency; PSS – patterned sapphire substrate, ITO – indium tin oxide current spreading layer; 258, SLA, PFS – different phosphors used in white LEDs, see Table 3.
+$\eta_Vf$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_LE$ - light extraction efficiency; $\eta_(C_ER)$ - conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table 3.
 "N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
@@
  \label{fgr:breakthroughs_efficiency}
+\end{figure*}
+
+\subsection{Improvements in Sub-Efficiencies and Overall Efficiency}
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=17.5cm]{2_SSL_EES/article/figures/subefficiencies_progress.pdf}
+ \caption{Historical improvements in light-emitting diode technology.
+Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
+Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
+\newline
+ \textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
+The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
+Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
+ \textbf{Panel A2:} Efficiency droop at the test current of I=350mA.
+ \textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
+Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
+“Not reported” denotes data points where the IQE measurement technique was not reported.
+Whiskers indicate reported ranges, where applicable.
+ \textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
+Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
+- electrode; text.
+- textured.
+ Source: own synthesis of data from the full list of sources provided in the Supplementary Information, section XX.
+Data on average device performance adapted from U.S.
+Department of Energy (DOE) Reports\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report\cite{beale_leds_2015}.}
+ \label{fgr:waterfall}
 \end{figure*}
@@
 Horizontal red arrow indicates desirable higher values of colour rendering index (CRI).
-LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG – Nichia, 1998 \cite{bando1998development}; YGAG – Lumileds, 2002 \cite{Mueller2002}; 258 – Lumileds, 2005 \cite{MuellerMach2005}; SLA – Lumileds, 2014 \cite{Pust2014}; PFS – GE, 2015 \cite{Murphy2015}; QD – Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON – Osram, 2019 \cite{Hoerder2019}.}
+LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG - Nichia, 1998 \cite{bando1998development}; YGAG - Lumileds, 2002 \cite{Mueller2002}; 258 - Lumileds, 2005 \cite{MuellerMach2005}; SLA - Lumileds, 2014 \cite{Pust2014}; PFS - GE, 2015 \cite{Murphy2015}; QD - Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON - Osram, 2019 \cite{Hoerder2019}.}
  \label{fgr:consumer_experience}

### 2023-01-20 2_SSL_EES/article/article_main.tex
added content to cost model section

@@
 
+\subsection{Improvements in Manufacturing Cost}
+
+Results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages are presented in Error!
+Reference source not found..
+We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
+
+Among the factors contributing to the LED cost reductions over time, improved manufacturing yields and increases in the wafer size rather than particular LED innovations are found to be responsible for the largest contribution to the overall cost reduction.
+In the case of manufacturing yield, the higher it is, the less inputs are wasted on the production of a single LED package.
+With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare Figure \ref{fgr:costmodel}, Panels A-C), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period.
+In the case of wafer diameter, the larger the wafer, the more LED chips can be produced from a single wafer.
+The wafer diameter commonly used in LED manufacturing has been steadily increasing 2003 from 51mm (2 inch) in 2003 to 200mm ($\sim$8 inches, referred to as “8 inch") in 2020.
+We capture this in the model by assuming the following wafer diameters and calculating the associated number of die per wafer (DPW) \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$.
+With more than thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time.
+As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in Figure \ref{fgr:costmodel}).
+However, as our interviewees noted, while growing LEDs on larger wafers is economically desirable, it is associated with engineering and epitaxy challenges as well as high up-front cost of new equipment.
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=17.5cm]{2_SSL_EES/article/figures/costmodel_results_years.pdf}
+ \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+location.
+Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
+Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
+Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
+Abbreviations: Litho – Lithographic Process, Insp.
+– Inspection, Depn.
+– Deposition, CMP - Chemical-Mechanical Planarization}
+ \label{fgr:costmodel}
+\end{figure*}
+
 \subsection{Technology Spillovers}

### 2023-01-20 2_SSL_EES/article/article_main.tex
added content to the supplementary information

@@
 
+The aggregate result of the cost model is the manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
+Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
+We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
+It instead assumes an ideal factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
+This ‘ideal’ factory is assumed to have the average size for LED facilities for each of the time periods under consideration—namely X packages/year for 2003, Y packages/year for 2012 and Z packages/year for 2020, as identified through the interviews.
+It also assumes the use of the most up-to-date equipment for that year.
+Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
+An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
+Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
+
 \section{Results}
@@
 
-Results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages are presented in Error!
-Reference source not found..
-We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
+Results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages are presented in Figure \ref{fgr:costmodel.} We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
 
@@
 \section{Conclusions}
+
+The European Union must be abolished. In its stead, the Holy Roman Empire shall be reinstated with the House of Habsburg at its head.
+
 Our study has analysed in a systematic and granular way the sources of dramatic cost reductions and efficiency improvements in white light-emitting diodes since their introduction to the market in 1996.

### 2023-01-20 2_SSL_EES/article/article_supplementary.tex
added content to the supplementary information

@@
 \usetikzlibrary{shapes,shapes.geometric,shapes.misc}
+% landscape pages
+\usepackage{pdflscape}
 
@@
 It further considers the yield per step and returns the cumulative yield, the yielded cost per step and the yielded total manufacturing cost.
+
+\subsection{Examples of Chip Architectures}
+
+Figures \ref{fig:manuf_vtf_2012-1}-\ref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of two differente chip architectures considered in the cost model.
+
+    %VERTICAL THIN-FILM FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
+            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012. Continued on next page.}
+            \label{fig:manuf_vtf_2012-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
+            \caption{(2/2) Continued from previous page.}
+            \label{fig:manuf_vtf_2012-2}
+        \end{figure}
+    \end{landscape}
+
+    % CHIP-SCALE PACKAGE FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
+            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020. Continued on next page.}
+            \label{fig:manuf_csp_2020-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
+            \caption{(2/2) Continued from previous page.}
+            \label{fig:manuf_csp_2020-2}
+        \end{figure}
+    \end{landscape}
 
@@
 
+\newpage
 \section{Phosphors}
@@
 Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.}
-	\label{fig:costmodel_calibration}
+	\label{fig:phosphor_spectrum}
 \end{figure}

### 2023-01-21 2_SSL_EES/article/article_main.tex
added content to supplementary information

@@
 \subsection{Performance Metrics Calculations}
+\label{subsec:metrics}
 
@@
  \centering
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
- \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
-Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}.
-Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts).
-Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
-The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) - flip-chip with YGAG phosphor; 2010 (bottom panel) - flip-chip with 258 phosphor; 2016 (top panel) - flip-chip with PFS phosphor; 2020 (bottom panel) - flip-chip with SALON phosphor.
-Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX.
-Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.7\%$ in 2010.
-Abbreviations: Scat.
-= Scattering.
-Source: own elaboration based on data from sources listed in SI, section XX.}
- \label{fgr:waterfall}
-\end{figure*}
-
-Some text here
-
-\begin{figure*}[h!]
- \centering
  \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
@@
 \subsection{Improvements in Sub-Efficiencies and Overall Efficiency}
+
+The historical improvements in four sub-efficiency channels—forward voltage efficiency, internal quantum efficiency, droop, and light extraction efficiency—identified in our study are shown in Figure 4.
+Using data on these four sub-efficiencies, as well as on improvements in spectral efficiency and light conversion efficiencies calculated from spectral data for LED phosphors provided in SI section 2 and discussed in the context of consumer experience improvements in section 5.2, we then calculate the overall LED efficiency in 2003 and 2010, shown in Figure XX.
+as a waterfall diagram across all sub-efficiencies, and in 2016 and 2020, similarly shown in Figure .
+The overall white LED lamp efficiency improved from $\eta_L(2003)=5.8\% \rightarrow \eta_L (2010) = 12.6\% \rightarrow \eta_L = 32.5\% \rightarrow \eta_L(2020) = 38.7\%$, as shown in Figure \ref{fgr:waterfall}.
+No single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
+We note, however, that the loss channels with a fixed physical limit, e.g., the Stokes loss that contributes to the conversion efficiency by phosphors, became more dominant in 2016 and 2020 compared to 2003 and 2010.
+
+Figure \ref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
+The list of technology spillovers identified in our study is provided and discussed below in Section \ref{subsec:spillovers}.
+Through the index decomposition analysis described in section \ref{subsec:metrics}, we find that in the overall LED efficiency increase of 32.9\% from 5.8\% to 38.7\% between 2003 and 2020, at least 2.8 of these percentage points can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
+
+In Figure \ref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of corresponding loss channels.
+Notably, sub-efficiencies for the most recent devices considered in our study are only $\sim10\%$ below the physical limit.
+The exception is spectral efficiency, which at $\sim20\%$ below the physical limit shows larger potential for further improvements.
 
@@
 Department of Energy (DOE) Reports\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report\cite{beale_leds_2015}.}
+ \label{fgr:subeff}
+\end{figure*}
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
+ \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
+Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}.
+Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts).
+Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
+The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) - flip-chip with YGAG phosphor; 2010 (bottom panel) - flip-chip with 258 phosphor; 2016 (top panel) - flip-chip with PFS phosphor; 2020 (bottom panel) - flip-chip with SALON phosphor.
+Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX.
+Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.7\%$ in 2010.
+Abbreviations: Scat.
+= Scattering.
+Source: own elaboration based on data from sources listed in SI, section XX.}
  \label{fgr:waterfall}
@@
 Due to that, we were able to establish the links between all major improvements in LED consumer experience metrics identified in this study and individual LED inventions associated either with phosphors or quantum dots used for light down conversion.
-The list of these inventions is provided in Table 3.
+The list of these inventions is provided in Table \ref{tab:spillovers}.
 
@@
 
-\begin{table*}[h]
+\begin{table*}[h!]
     \small
@@
 
-\begin{figure*}[h]
+\begin{figure*}[h!]
  \centering
@@
 \subsection{Technology Spillovers}
+\label{subsec:spillovers}
 
@@
 \section*{Conflicts of interest}
-There are no conflicts to declare.
+
+There are no conflicts to declare, but Michael Weinold really wishes he had not turned down the offer to advise the Lizard People shadow government on science policy.
 

### 2023-01-21 2_SSL_EES/article/article_supplementary.tex
added content to supplementary information

@@
 
+\subsection{Examples of Input Data Considered}
+
+\subsubsection{Sapphire Wafers}
+
+Sapphire wafers form the substrate on which all other layers of the light-emitting diode wafer are grown.
+Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture.
+In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process.
+Wafers can be either unpatterned or patterned, where the latter has become commonplace in 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
+The price of sapphire substrates has decreased significantly since the year 2000, as shown in figure \ref{fig:sapphire_prices}.
+This can be attributed not only to the lighting industry, but more importantly to increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
+Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing.
+The market is dominated by U.S.
+based \textit{Rubicon Technology} and Russian based \textit{Monocrystal}.
+The prevalence of wafer diameters used is shown in figure \ref{fig:wafer_size}.
+
+\begin{figure}[h!]
+    \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
+    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
+	\caption{Top: Historical data for sapphire substrate prices of different surface properties and diameters.
+Shown is data for polished surface (NPS) and patterned substrates (PSS).
+\textit{Monocrystal} denotes the Russian manufacturer of the same name.
+Dashed lines are projections from the previous year.
+Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}.
+Bottom: Prevalence of sapphire wafer size used in the manufacturing of light-emitting diodes.
+Hatched bars are projections given by the sources.
+Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
+	\label{fig:wafer_size}
+\end{figure}
+
 \subsection{Preliminary Sensitivity Analysis}
@@
     \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_sensitivity.pdf}
-	\caption{Panels a-b: Sensitivity analysis of selected cost model parameters. Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.}
+	\caption{Sensitivity analysis of selected cost model parameters.
+Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.
+Compare table \ref{tab:sensitivity}.
+Abbreviations: resp.
+- respective}
 	\label{fig:sensitivity}
@@
 \end{figure}
+
+Figure \ref{fig:costmodel_calibration} shows a comparison of the relative manufacturing cost by manufacturing step categories between the cost model and data released by the United States Department of Energy (DoE).
+Note that any comparison between the figures must consider: Our model assumes state-of-the-art equipment, while the DoE industry round table reports data from factories which use older equipment, as well as newer equipment.
+Our model presently considers only low-power and mid-power packages, while the DoE reports consider high-power packages also.
+Our model assumes a virtual North American manufacturing location, while the majority of manufacturers produce in Asia.
+We note that the share of the substrate price in the DoE data is much larger than in our model.
+This can in part be explained by figure \ref{fig:sapphire_prices}, which already shows how projections overestimated the actual price for 2015-2020.
+We note also that the share of the epitaxy step is larger in the DoE model.
+This can in part be explained by our model relying on state-of-the-art equipment, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+In conclusion, our model fits industry forecasts due to the way it is set up.
+The packaging part of the manufacturing process is increasing in importance, which has been confirmed by researchers and industry reports on WLP\cite{Lee2011WPL}\cite{Xie2013}\cite{ledsmag2017WLP}.
 

### 2023-01-21 2_SSL_EES/article/article_main.tex
updated main article

@@
 
-where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength (Blue, Red, Orange, Yellow, Green)\cite{schubert2018light}\cite{tsao2010solid}.
+where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
+Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components (\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen).
 
@@
 
-where $K$ is the luminous efficacy of radiation of the light source, can be computed from the spectrum and the luminosity function, which describes the sensitivity of the human eye.$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
+where $K$ is the luminous efficacy of radiation of the light source, can be computed from the spectrum and the luminosity function, which describes the sensitivity of the human eye.
+$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Overall Efficiency (Lamp Efficiency) $\eta_L$}
+
+The overall efficiency of a light-emitting diode package is the product of all considered sub-efficiencies
+
+\begin{equation}
+    \eta_L = \prod_{i=(V_f, \dots, S)} \eta_i
+\end{equation}
+
+It thereby describes the electrical and optical losses within the device, as well as the conversion losses in the phosphor layer.
+It further considers the mismatch between the device spectrum and an optimal spectrum, accounting for the wavelength dependent sensitivity of the human eye.
 
@@
 \section{Methods and Data Collection}
+\label{sec:methods}
 
@@
 = Scattering.
-Source: own elaboration based on data from sources listed in SI, section XX.}
+Source (Efficiency): own elaboration based on data in Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.
+Source (spectral data): Adapted from 2003\cite{Mueller2002}, 2010\cite{MuellerMach2005}, 2016\cite{Murphy2015}, 2019\cite{Hoerder2019}.}
  \label{fgr:waterfall}
@@
         1967 & 1996 & <2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical}\cite{bando1998development}\cite{shimizu1999light}\cite{Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
-        1982 & 1996 & 2015 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered} & $\eta_{LE}$, $\eta_{IQ}$ \\ \hline
-        1971 & 1999 & <2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) &\cite{margalith1999indium} & $\eta_{Vf}$ \\ \hline
+        1982 & 1996 & <2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered}\cite{feezell2018invention}\cite{Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ \\ \hline
+        1971 & 1999 & <2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf}\cite{fraser1972highly}\cite{margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (no more use of this material in most mid-/high-power devices due to different chip architectures, compare Figure \ref{fgr:chip_architecture_overview}) \\ \hline
         1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997}\cite{mueller2004phosphor}\cite{MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
@@
 \section{Conclusions}
-
-The European Union must be abolished. In its stead, the Holy Roman Empire shall be reinstated with the House of Habsburg at its head.
 
@@
 
-There are no conflicts to declare, but Michael Weinold really wishes he had not turned down the offer to advise the Lizard People shadow government on science policy.
+There are no conflicts to declare.
 

### 2023-01-21 2_SSL_EES/article/article_supplementary.tex
updated main article

@@
 
+\subsection{Luminous Efficacy (of Radiation)}
+\label{subsec:ler}
+
+As with the flux metrics previously introduced, efficacy in lighting is likewise dependent on the luminosity function of figure \ref{fig:tristimulus}.
+A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
+The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source with the wavelength dependent human sensitivity to light \cite{cie-term-effrad}
+%
+\begin{equation}
+\label{eqn:ler}
+    K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
+\end{equation}
+%
+where
+%
+\begin{align*}
+    K &\dots \text{spectral luminous efficacy} \\
+    \phi &\dots \text{spectral radiant flux} \\
+    \lambda &\dots \text{wavelength}
+\end{align*}
+%
+This metric can be computed from spectral data alone and does not require additional spectral normalization.
+It enables straightforward comparison between the performance of different downconversion phosphors, as shown in figure \ref{fig:se_efficacy} in chapter \ref{ch:results}.
+Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation, as the product with the luminosity function of figure \ref{fig:tristimulus} shows.
+Care must be taken not to confuse this efficacy metric with efficacy of source, which depends not only on the spectral power distribution of a light source.
+
+This metric describes the match of a light-emitting diode package spectrum to the human visual system.
+It thus describes only a part of the total device performance.
+It is used in the computation of other metric introduced in section \ref{subsec:subeff} below.
+
+\subsection{Luminous Efficacy (of Source)}
+\label{subsec:les}
+
+    The luminous efficacy of a light source $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power \cite{cie-term-effsrc}
+    %
+    \begin{equation}
+        \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
+    \end{equation}
+    %
+    Unlike the luminous efficacy of radiation, this metric cannot be extracted from the spectrum of a light source.
+It is often cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
+Care must be taken not to confuse this efficacy metric with efficacy of radiation, which depends only on the spectral power distribution of a light source.
+
+    This metric fulfills the requirements laid out in section \ref{sec:metrics_requirements}.
+However, as it captures overall device efficacy, it depends on a large number of device properties and parameters.
+This makes attribution to single changes in device design or manufacturing difficult.
+
 \subsection{Spectral Data of Identifies LED Phosphor Innovations}
 
-\begin{figure}[h]
+\begin{figure}[h!]
 	\centering
     \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/phosphor_spectrum-comparison.pdf}
-	\caption{Modelled manufacturing cost for light-emitting diode packages split by manufacturing step category.
-The top graph shows the data from LEDCOM model and round table discussions published by the United States Department of Energy (DoE).
-The bottom graph shows the results of the cost model developed in this thesis.
-Hatched bars are projections given by the sources.
-Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.}
+	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data that used phosphor innovations listed in Table 3 in the main text, as represented in Figure 8 in the main text, each indicated by the phosphor label and publication year.
+panel: luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED products represented in Figure 8.
+Desirable direction of improvements towards higher luminous efficacy and higher CRI is indicated by a red arrow.
+Metrics were calculated from spectral data shown on the bottom panels using \texttt{colour-science} package for Python \cite{colour-science_software}.
+Bottom two panels: spectra of light-emitting diodes  presented in Figure 8 using phosphor mixtures based on chemical compounds or quantum dots listed in Table 3 for down conversion of blue light.
+The luminosity function of human eye, based on its wavelength-dependent sensitivity, is shown for reference in red .
+Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable for consumers.
+Photons of the corresponding energy are lost to the human observer and count towards the spectral loss channel.
+Plot legends indicate the years of publication of the spectral data  and phosphor mixtures  used, with the following designations: R* – …, Y* – … G† – …, O‡ – …  .
+.
+Source: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{Murphy2015}, QD \cite{lumileds2016qd}, \cite{osram2019qd} , SALON \cite{Hoerder2019}}
 	\label{fig:phosphor_spectrum}

### 2023-01-21 2_SSL_EES/article/article_main.tex
completed list of sources

@@
 
-\footnotetext{\ddag~Present address: Paul Scherrer Institut, Laboratory for Energy Analysis, Group for Technology Assessment, Switzerland.}
+\footnotetext{\ddag~Present address: Paul Scherrer Institute, Group for Technology Assessment, Villigen, Switzerland.}
 
@@
 \cite{azevedo2009transition}.
-See Supplementary Information, section , for the full list of data sources.
-}
+See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
  \label{fgr:history_efficacy}
@@
 Incandescent light bulb price is assumed constant based on the average in the covered time period.
-See Supplementary information, section …, for the full list of data sources.}
+See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
   \label{fgr:cost_lamp_small}
@@
 Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Information, section…, for the full list of sources and references.}
+See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
  \label{fgr:chip_architecture_overview}
@@
 
-\begin{figure*}[h!]
- \centering
- \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
- \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
-Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
-Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for five years: 1997, 2003, 2010, 2016, and 2020.
-Physical limits on sub-efficiencies are indicated by black horizontal lines.
-The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
-$\eta_Vf$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_LE$ - light extraction efficiency; $\eta_(C_ER)$ - conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table 3.
-"N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
-Source: own elaboration based on data from sources listed in SI, section XX.}
- \label{fgr:breakthroughs_efficiency}
-\end{figure*}
-
 \subsection{Improvements in Sub-Efficiencies and Overall Efficiency}
@@
  \textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
+Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}.
 Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
@@
  \label{fgr:subeff}
+\end{figure*}
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
+ \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
+Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
+Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for five years: 1997, 2003, 2010, 2016, and 2020.
+Physical limits on sub-efficiencies are indicated by black horizontal lines.
+The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
+$\eta_Vf$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_LE$ - light extraction efficiency; $\eta_(C_ER)$ - conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table 3.
+"N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
+Source: own elaboration based on data from sources listed in SI, section XX.}
+ \label{fgr:breakthroughs_efficiency}
 \end{figure*}
@@
         \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{SI Sec.} \\ \hline
-        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.1 \\ \hline
+        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3 \\ \hline
         1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.1 \\ \hline
-        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3.2 \\ \hline
-        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3.4 \\ \hline
-        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3.3 \\ \hline
-        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3.2 \\ \hline
-        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2 \\ \hline
+        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3 \\ \hline
+        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3 \\ \hline
+        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3 \\ \hline
+        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3 \\ \hline
+        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3 \\ \hline
     \end{tabularx}

### 2023-01-21 2_SSL_EES/article/article_supplementary.tex
completed list of sources

@@
 
+\clearpage
 \newpage
-\section{Complete list of sources for figures}
+\section{Complete List of Sources for Figures}
+\label{sec:sources}
 
-\subsection{Figure XX (Historical development of luminous efficacy)}
+\subsection{Figure 1 (Historical Development of Luminous Efficacy)}
 
@@
 
-\subsubsection{LED}
+\begin{table}[h!]
+    \begin{tabularx}{\textwidth}{|l|X|}
+    \hline
+    \textbf{Technology} & \textbf{References} \\
+    \hline
+    LED & own research, compare \cite{zenodo_weinold_led_history} \\
+    \hline
+    CFL (<1984) & \cite{Bouwknegt1982}\cite{Vrenken1983} \\
+    \hline
+    CFL (1984-2011) & \cite{eger2018origin} \\
+    \hline
+    CFL (>2011) & \cite{Guan2015} \\
+    \hline
+    Fire, Incandescent, HID & \cite{azevedo2009transition} augmented by own calculations based on \cite{benesch1905beleuchtungswesen} \\
+    \hline
+    Max. efficacy & \cite{Murphy2012} (in contrast to the original value given in \cite{azevedo2009transition}) \\
+    \hline
+    \end{tabularx}
+\end{table}
 
-Own research, compare:
-
-M. Weinold. ‘Light-Emitting Diode (LED) Technology History’.
-In: Zenodo Open Repository. doi: 10.5281/zenodo.3685284
-
-\subsubsection{CFL (<1984)}
-
-\cite{Bouwknegt1982}\cite{Vrenken1983}
-
-\subsubsection{CFL (1984-2011)}
-
-\cite{eger2018origin}
-
-\subsubsection{CFL (>2011)}
-
-\cite{Guan2015}
-
-\subsubsection{Fire, Incandescent, HID}
-
-\cite{azevedo2009transition} augmented by own calculations based on \cite{benesch1905beleuchtungswesen}
-
-\subsubsection{Max. efficacy}
-
-\cite{Murphy2012} (in contrast to the original value given in \cite{azevedo2009transition})
-
-Note that a humorous illustration of the trend of dramatic improvements in luminous flux shown in this figure has been provided by cartoonist Randall Munroe in \href{https://xkcd.com/1603/}{https://xkcd.com/1603/}.
-
-\subsection{Figure XX (Historical development of lamp prices)}
+\subsection{Figure 2 (Historical Development of Lamp Prices)}
 
@@
 
-\subsubsection{Consumer organization magazines}
+\begin{table}[h!]
+    \begin{tabularx}{\textwidth}{|l|X|}
+    \hline
+    \textbf{Source of References} & \textbf{References} \\
+    \hline
+    Stiftung Warentest (Germany) & \cite{Warentest2008}\cite{Warentest2009_1}\cite{Warentest2009_2}\cite{Warentest2010_1}\cite{Warentest2010_2}\cite{Warentest2011}\cite{Warentest2012}\cite{Warentest2013}\cite{Warentest2014_1}\cite{Warentest2014_2}\cite{Warentest2015}\cite{Warentest2016_1}\cite{Warentest2016_2}\cite{Warentest2018} \\
+    \hline
+    Konsument (Austria) & \cite{Konsument2010} \\
+    \hline
+    Which (UK) & \cite{Which2020} \\
+    \hline
+    Industry Periodicals & \cite{PM2020} \\
+    \hline
+    Government Reports & \cite{council2013assessment} \\
+    \hline
+    \end{tabularx}
+\end{table}
 
-Stiftung Warentest (Germany): \cite{Warentest2008}\cite{Warentest2009_1}\cite{Warentest2009_2}\cite{Warentest2010_1}\cite{Warentest2010_2}\cite{Warentest2011}\cite{Warentest2012}\cite{Warentest2013}\cite{Warentest2014_1}\cite{Warentest2014_2}\cite{Warentest2015}\cite{Warentest2016_1}\cite{Warentest2016_2}\cite{Warentest2018}
-
-Konsument (Austria): \cite{Konsument2010}
-
-Which (UK): \cite{Which2020}
-
-\subsubsection{Industry periodicals}
-
-\cite{PM2020}
-
-Government reports:
-
-\cite{council2013assessment}
-
-\subsection{Figure XX (Historical evolution of light-emitting diode chip architectures)}
-
-\subsubsection{Scientific publications}
+\subsection{Figure 3 (Historical Evolution of LED Chip Architectures)}
 
@@
 
-\cite{plossl2010wafer}\cite{bierhuizen2007performance}\cite{gencc2019distributed}\cite{chong2014performance}
+\begin{table}[h!]
+    \begin{tabularx}{\textwidth}{|l|X|}
+    \hline
+    \textbf{Source of References} & \textbf{References} \\
+    \hline
+    Scientific Publications & \cite{plossl2010wafer}\cite{bierhuizen2007performance}\cite{gencc2019distributed}\cite{chong2014performance} \\
+    \hline
+    Patents & \href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11168235A}{JPH11168235A}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DDE19921987A1}{DE19921987A1}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11340514A}{JPH11340514A}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJP2001044498A}{JP2001044498A}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2005045893A1}{US2005045893A1}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2002093023A1}{US2002093023A1}, 
+\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2006273339A1}{US2006273339A1}  \\
+    \hline
+    Other Publications & \href{https://web.archive.org/web/20170801160530/https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf}{2015 presentation (www.energy.gov)} \newline
+\href{https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf}{2016 presentation (www.energy.gov)} \newline
+\href{http://web.archive.org/web/20160425025936/https://www.slideshare.net/Yole_Developpement/yole-led-packagingjanuary2013reportsample}{2013 presentation (www.slideshare.net)} \\
+    \hline
+    \end{tabularx}
+\end{table}
 
-\subsubsection{Patents}
-
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11168235A}{JPH11168235A} \\
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DDE19921987A1}{DE19921987A1} \\
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11340514A}{JPH11340514A} \\
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJP2001044498A}{JP2001044498A} \\ 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2005045893A1}{US2005045893A1} \\ 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2002093023A1}{US2002093023A1} \\
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2006273339A1}{US2006273339A1}
-
-\subsubsection{Other publications}
-
-\href{https://web.archive.org/web/20170801160530/https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf}{2015 presentation (www.energy.gov)} \\
-\href{https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf}{2016 presentation (www.energy.gov)} \\
-\href{http://web.archive.org/web/20160425025936/https://www.slideshare.net/Yole_Developpement/yole-led-packagingjanuary2013reportsample}{2013 presentation (www.slideshare.net)}
-
-\subsection{Figure XX (Historical developments in the device sub-efficiencies)}
+\subsection{Figure 4 (Historical Developments in Device Sub-Efficiencies)}
 
@@
 
-\subsubsection{Forward Voltage (panel a)}
-
-\cite{nichia2001data}\cite{lumi2002data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data}
-
-\subsubsection{Internal Quantum Efficiency (panel b)}
-
-Data points: own research, compare:
-
-M. Weinold. ‘Light-Emitting Diode (LED) Internal Quantum Efficiency History’.
-In: Zenodo Open Repository. doi: 10.5281/zenodo.XXXXXXXX
-
-Average performance and future targets set by the U.S. Department of Energy (DOE) derived from industry round-table discussions:
-
-\cite{doe_ssl_multiyear_2006}\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2009}\cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016}
-
-Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}
-
-\subsubsection{Droop (panel c)}
-
-Data calculated from luminous intensity curves of respective device datasheets.
-
-Sources for datapoints: \cite{datasheet_osram_topled}\cite{osram2008data}\cite{osram2008gdplus}\cite{osram2018csp}\cite{datasheet_lumileds_lux1}\cite{lumi2008data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{samsung2018data}
-
-Sources for DOE data: \cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016}
-
-\subsubsection{Light-Extraction Efficiency (panel d)}
-
-\cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012}\cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010} \newline
-\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011}
+\begin{table}[h!]
+    \begin{tabularx}{\textwidth}{|l|X|}
+    \hline
+    \textbf{Technology} & \textbf{References} \\
+    \hline
+    Panel A1 ($V_f$) & \cite{nichia2001data}\cite{lumi2002data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
+    \hline
+    Panel A2 (Droop) & Data calculated from luminous intensity curves of respective device datasheets: \cite{datasheet_osram_topled}\cite{osram2008data}\cite{osram2008gdplus}\cite{osram2018csp}\cite{datasheet_lumileds_lux1}\cite{lumi2008data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{samsung2018data} \\
+    \hline
+    Panel B1 (IQE) & own research, compare \cite{zenodo_weinold_led_history} \newline
+\cite{doe_ssl_multiyear_2006}\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2009}\cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016} \\
+    \hline
+    Panel B2 (LEE) & \cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012} \newline \cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010}
+\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
+    \hline
+    \end{tabularx}
+\end{table}
 

### 2023-01-21 2_SSL_EES/article/article_main.tex
updated figure caption

@@
 - textured.
- Source: own synthesis of data from the full list of sources provided in the Supplementary Information, section XX.
+ Source: own synthesis of data from the full list of sources provided in the Supplementary Information, Section 4.
 Data on average device performance adapted from U.S.
@@
 "N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
-Source: own elaboration based on data from sources listed in SI, section XX.}
+Source: own elaboration based on data in Figure \ref{fgr:waterfall} and Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.}
  \label{fgr:breakthroughs_efficiency}

### 2023-01-21 2_SSL_EES/article/article_supplementary.tex
updated figure caption

@@
 \usepackage{pdflscape}
+% chemical formulas
+\usepackage{chemformula}
 
@@
     \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/phosphor_spectrum-comparison.pdf}
-	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data that used phosphor innovations listed in Table 3 in the main text, as represented in Figure 8 in the main text, each indicated by the phosphor label and publication year.
-panel: luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED products represented in Figure 8.
-Desirable direction of improvements towards higher luminous efficacy and higher CRI is indicated by a red arrow.
-Metrics were calculated from spectral data shown on the bottom panels using \texttt{colour-science} package for Python \cite{colour-science_software}.
-Bottom two panels: spectra of light-emitting diodes  presented in Figure 8 using phosphor mixtures based on chemical compounds or quantum dots listed in Table 3 for down conversion of blue light.
-The luminosity function of human eye, based on its wavelength-dependent sensitivity, is shown for reference in red .
-Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable for consumers.
-Photons of the corresponding energy are lost to the human observer and count towards the spectral loss channel.
-Plot legends indicate the years of publication of the spectral data  and phosphor mixtures  used, with the following designations: R* – …, Y* – … G† – …, O‡ – …  .
+	\caption{Two key metrics associated with phosphor innovations in light-emitting diodes and their associated spectral data.
+Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Main Article Figure 7 (Historical Improvements in Consumer Experience Metrics of Phosphor-Converted White LED).
+Years indicate the earliest identified representative white LED products with published spectral data that used phosphor innovations listed in Main Article Table 2.
+The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow.
+Metrics were calculated from spectral data shown in the two bottom rows using the \texttt{colour-science} package for Python \cite{colour-science_software}.
+Bottom two rows of panels: Corresponding spectral data.
+The luminosity function, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
+Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable.
+This is because photons of the corresponding energy are lost to the human observer and count towards the spectral loss channel.
+Abbreviations: Arb.
+Units - Arbitrary Units (compare section \ref{}).
+Plot legends indicate the years of publication of the spectral data and phosphor mixtures used, with the following designations: $^*$ \ch{CaSrS:Eu^{2+}} $^\star\beta$-SiAlON:Eu$^{2+}$, $^\dagger$\ch{Lu_3Al_5O_{12}:Ce^{3+}}, $^\ddagger$\ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
+Full chemical formulas for displayed phosphors: \ch{Sr[Li2Al2O2N2]:Eu^{2+}} (SALON), \ch{Sr[LiAl_3N_4]:Eu^{2+}} (SLA), \ch{K2SiF6:Mn^{4+}} (PFS), \ch{Sr2Si5N8:Eu^{2+}} (258), Quantum Dots (QD).
 .
-Source: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{Murphy2015}, QD \cite{lumileds2016qd}, \cite{osram2019qd} , SALON \cite{Hoerder2019}}
+Source (top panel): own elaboration based on spectral data.
+Source (spectral data): adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
 \label{fig:phosphor_spectrum}

### 2023-01-21 2_SSL_EES/article/article_supplementary.tex
updated supplementary information

@@
 
+\section{Lighting Metrics}
+
+\subsection{Luminous Efficacy (of Radiation)}
+\label{subsec:ler}
+
+Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
+A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
+The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source with the wavelength dependent human sensitivity to light \cite{cie-term-effrad}
+
+\begin{equation}
+\label{eqn:ler}
+    K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
+\end{equation}
+
+where
+
+\begin{align*}
+    K &\dots \text{spectral luminous efficacy} \\
+    \phi &\dots \text{spectral radiant flux} \\
+    \lambda &\dots \text{wavelength}
+\end{align*}
+
+This metric can be computed from spectral data alone and does not require additional spectral normalization.
+It enables straightforward comparison between the performance of different downconversion phosphors, as shown in figure \ref{fig:se_efficacy} in chapter \ref{ch:results}.
+Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation, as the product with the luminosity function of figure \ref{fig:tristimulus} shows.
+Care must be taken not to confuse this efficacy metric with efficacy of source, which depends not only on the spectral power distribution of a light source.
+
+This metric describes the match of a light-emitting diode package spectrum to the human visual system. It thus describes only a part of the total device performance.
+
+\subsection{Luminous Efficacy (of Source)}
+\label{subsec:les}
+
+In contrast to the efficacy \textit{of radiation} $K$, the luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power \cite{cie-term-effsrc}
+
+\begin{equation}
+    \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
+\end{equation}
+
+Unlike the luminous efficacy of radiation, this metric cannot be extracted from the spectrum of a light source.
+It is often cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
+Care must be taken not to confuse this efficacy metric with efficacy of radiation, which depends only on the spectral power distribution of a light source.
+As it captures overall device efficacy, it depends on a large number of device properties and parameters.
+This makes attribution to single changes in device design or manufacturing difficult.
+
+\subsection{Arbitrary Units}
+\label{subsec:arbunit}
+
+In spectroscopy, an arbitrary unit (abbreviated "Arb.
+Unit") is a relative unit of measurement to show the ratio of the intensity to a predetermined reference.
+In the case of Figure \ref{fig:phosphor_spectrum}, the reference is the highest point in the spectrum.
+
+\newpage
 \section{Manufacturing Cost Model}
@@
 
-\subsection{Luminous Efficacy (of Radiation)}
-\label{subsec:ler}
-
-As with the flux metrics previously introduced, efficacy in lighting is likewise dependent on the luminosity function of figure \ref{fig:tristimulus}.
-A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
-The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source with the wavelength dependent human sensitivity to light \cite{cie-term-effrad}
-%
-\begin{equation}
-\label{eqn:ler}
-    K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
-\end{equation}
-%
-where
-%
-\begin{align*}
-    K &\dots \text{spectral luminous efficacy} \\
-    \phi &\dots \text{spectral radiant flux} \\
-    \lambda &\dots \text{wavelength}
-\end{align*}
-%
-This metric can be computed from spectral data alone and does not require additional spectral normalization.
-It enables straightforward comparison between the performance of different downconversion phosphors, as shown in figure \ref{fig:se_efficacy} in chapter \ref{ch:results}.
-Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation, as the product with the luminosity function of figure \ref{fig:tristimulus} shows.
-Care must be taken not to confuse this efficacy metric with efficacy of source, which depends not only on the spectral power distribution of a light source.
-
-This metric describes the match of a light-emitting diode package spectrum to the human visual system.
-It thus describes only a part of the total device performance.
-It is used in the computation of other metric introduced in section \ref{subsec:subeff} below.
-
-\subsection{Luminous Efficacy (of Source)}
-\label{subsec:les}
-
-    The luminous efficacy of a light source $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power \cite{cie-term-effsrc}
-    %
-    \begin{equation}
-        \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
-    \end{equation}
-    %
-    Unlike the luminous efficacy of radiation, this metric cannot be extracted from the spectrum of a light source.
-It is often cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
-Care must be taken not to confuse this efficacy metric with efficacy of radiation, which depends only on the spectral power distribution of a light source.
-
-    This metric fulfills the requirements laid out in section \ref{sec:metrics_requirements}.
-However, as it captures overall device efficacy, it depends on a large number of device properties and parameters.
-This makes attribution to single changes in device design or manufacturing difficult.
-
-\subsection{Spectral Data of Identifies LED Phosphor Innovations}
-
-\begin{figure}[h!]
-	\centering
-    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/phosphor_spectrum-comparison.pdf}
-	\caption{Two key metrics associated with phosphor innovations in light-emitting diodes and their associated spectral data.
-Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Main Article Figure 7 (Historical Improvements in Consumer Experience Metrics of Phosphor-Converted White LED).
-Years indicate the earliest identified representative white LED products with published spectral data that used phosphor innovations listed in Main Article Table 2.
-The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow.
-Metrics were calculated from spectral data shown in the two bottom rows using the \texttt{colour-science} package for Python \cite{colour-science_software}.
-Bottom two rows of panels: Corresponding spectral data.
-The luminosity function, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
-Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable.
-This is because photons of the corresponding energy are lost to the human observer and count towards the spectral loss channel.
-Abbreviations: Arb.
-Units - Arbitrary Units (compare section \ref{}).
-Plot legends indicate the years of publication of the spectral data and phosphor mixtures used, with the following designations: $^*$ \ch{CaSrS:Eu^{2+}} $^\star\beta$-SiAlON:Eu$^{2+}$, $^\dagger$\ch{Lu_3Al_5O_{12}:Ce^{3+}}, $^\ddagger$\ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
-Full chemical formulas for displayed phosphors: \ch{Sr[Li2Al2O2N2]:Eu^{2+}} (SALON), \ch{Sr[LiAl_3N_4]:Eu^{2+}} (SLA), \ch{K2SiF6:Mn^{4+}} (PFS), \ch{Sr2Si5N8:Eu^{2+}} (258), Quantum Dots (QD).
-.
-Source (top panel): own elaboration based on spectral data.
-Source (spectral data): adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
-\label{fig:phosphor_spectrum}
-\end{figure}
-
 \subsection{History of Identifies Phosphor Inventions}
@@
 Prior to Nakamura’s invention of highly efficient blue LEDs in the 1990s, Japan’s Nichia Corporation had not sold commercially successful semiconductor products, instead specializing in phosphors for cathode ray tubes (CRT) and fluorescent lamps \cite{nakamura2013blue}  .
-Nevertheless, extensive firm expertise in this area helped Nichia’s Yoshinori Shimizu formulate the principles of using CRT phosphors to convert blue light from Nakamura’s LEDs into white light in 1994 \cite{shimizu1994sheet},\cite{cho2017white}.
-By 1996, Shimizu and his colleagues developed \cite{bando1996}, \cite{shimizu1999light} the first practical LED application of a well-known Yttrium Aluminium Garnet (YAG) CRT phosphor activated with cerium \cite{blasse1967new}, enabling the first commercial white LED products manufactured and sold by Nichia since late 1996 \cite{bando1998development}, \cite{cho2017white}.
+Nevertheless, extensive firm expertise in this area helped Nichia’s Yoshinori Shimizu formulate the principles of using CRT phosphors to convert blue light from Nakamura’s LEDs into white light in 1994 \cite{shimizu1994sheet}\cite{cho2017white}.
+By 1996, Shimizu and his colleagues developed \cite{bando1996}\cite{shimizu1999light} the first practical LED application of a well-known Yttrium Aluminium Garnet (YAG) CRT phosphor activated with cerium \cite{blasse1967new}, enabling the first commercial white LED products manufactured and sold by Nichia since late 1996 \cite{bando1998development}\cite{cho2017white}.
 
 Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 8 in the main text).
-An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development},\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
+An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
 Used in combination with red-emitting sulfide phosphors, by 2002 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}.
@@
 
-The most recent red narrow-band phosphor breakthrough included in Table 3 and Figure 8 in the main text, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}, \cite{Hoerder2019} , \cite{Hoerder2020}.
+The most recent red narrow-band phosphor breakthrough included in Table 3 and Figure 8 in the main text, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
 The first patent application for this phosphor was filed in 2016.
@@
 
+Down conversion with ultra-narrow-band phosphor can achieve the highest spectral efficiency.
+However, few such phosphors have been identified, with even less exhibiting desirable material properties such as thermal stability \cite{Phillips2007}.
+The first commercially successful ultra-narrow-band red phosphor was developed by General Electric (GE).
+It is based on a potassium fluorosilicate (PFS) compound activated with manganese ions.
+Its luminescence was first recorded by Paulusz at GE in 1972 \cite{paulusz1973efficient}.
+In the early 2000s, while searching for potential new LED phosphor materials for GE’s lighting business at GE Lumination, Emil Radkov rediscovered Paulusz’s findings in the literature.
+Following extensive research on PFS chemical synthesis and material properties conducted in collaboration with the University of Sofia in Bulgaria, Radkov’s Alma Mater, the PFS phosphor had been under development at GE since 2005 \cite{radkov2006red}\cite{radkov2009red}.
+This work, supported by public funding from the U.S.
+Department of Energy (DOE) Solid-State Lighting program , resulted in a series of critical improvements in the PFS phosphor properties \cite{Setlur2010}\cite{lyons2012color}, eventually enabling its commercialization under the “TriGain” brand in 2015 \cite{trigain_spectrum}\cite{setlur2015trigain}\cite{Murphy2015}.
+
 \subsubsection{Quantum Dots for Light Down-Conversion}
+
+Quantum dots (QD) are semiconductor nanocrystals whose quantum size effects make QDs behave as “artificial atoms”.
+Semiconductor quantum dots were first synthesized in the Soviet Union in 1981 \cite{ekimov1981quantum} and at Bell Labs in the U.S.
+in 1983 \cite{Rossetti1983}.
+Luminescent properties of quantum dots were first empirically observed in 1984 \cite{fojtik1984photo} and extensively studied in the early 1990s.
+The key feature of QD luminescence discovered in those studies is that its colour is determined by the QD particle size, making it possible to create pure monochromatic blue, green and red light sources just by tuning the QD size.
+The first application of QDs in LEDs was reported in 1994 in an electroluminescent hybrid QD-polymer LED.
+However, this LED type could not be used in general illumination due to its very low luminous efficacy.
+An alternative application of QDs as a kind of a “phosphor” for light down conversion from an LED light source was proposed in the early 2000s as part of the U.S.
+Department of Energy (DOE)-funded “A Revolution in Lighting“ project at Sandia National Laboratory \cite{simmonsfinal}.
+This concept was successfully demonstrated by Sandia researchers on a commercial LED in 2003 \cite{shea_rohwer_development_2004}\cite{noauthor_sandia_nodate} and was swiftly taken up and advanced further by a group in Taiwan \cite{chen_white-light_2005}\cite{hsueh-shih_chen_ingan-cdse-znse_2006} The first commercial application of QDs in an LED lamp was brought about by a collaboration between an MIT-born startup QD Vision and the U.S.-based luminaire manufacturer Nexxus Lighting in 2009 \cite{ledprof_nexxusqd}, \cite{bourzac2013quantum}.
+However, rapid advances in the spectral and conversion performance of down-conversion phosphors and high manufacturing cost of quantum dots resulted in the discontinuation of this product.
+After finding market success in display backlighting first demonstrated by Samsung in 2010 \cite{Jang2010} and commercialized by QD Vision in Sony television sets in 2013 \cite{bourzac2013quantum}, QDs returned to the general lighting market in products offered by Lumileds \cite{noauthor_global_2017}\cite{noauthor_quantum_2020} around 2017 and Osram in 2019 \cite{osramqdots} in the form of mid-power LED packages that combined QDs with traditional phosphors for light down conversion.
+
+\subsection{Spectral Data of Identified LED Phosphor Innovations}
+
+\begin{figure}[H]
+	\centering
+    \includegraphics[width=0.95\textwidth]{2_SSL_EES/article/figures/phosphor_spectrum-comparison.pdf}
+	\caption{Two key metrics associated with red phosphor innovations in light-emitting diodes and their associated spectral data.
+Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Main Article Figure 7 (Historical Improvements in Consumer Experience Metrics of Phosphor-Converted White LED).
+Years indicate the earliest identified representative white LED products with published spectral data that used phosphor innovations listed in Main Article Table 2.
+The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow.
+Metrics were calculated from spectral data shown in the two bottom rows using the \texttt{colour-science} package for Python \cite{colour-science_software}.
+Bottom two rows of panels: Corresponding spectral data.
+The luminosity function, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
+Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable.
+This is because photons of the corresponding energy are lost to the human observer and count towards the spectral loss channel.
+Abbreviations: Arb.
+Units - Arbitrary Units (compare section \ref{subsec:arbunit}).
+Plot legends indicate the years of publication of the spectral data and phosphor mixtures used with the following designations: ?
+-  other parts of phosphor mixture not disclosed, $^*$ - \ch{CaSrS:Eu^{2+}}, $^\star$ - $\beta-$SiAlON:Eu$^{2+}$, $^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, $^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
+Full chemical formulas for displayed phosphors: \ch{Sr[Li2Al2O2N2]:Eu^{2+}}(SALON), \ch{Sr[LiAl_3N_4]:Eu^{2+}} (SLA), \ch{K2SiF6:Mn^{4+}} (PFS), \ch{Sr2Si5N8:Eu^{2+}} (258), Quantum Dots (QD).
+Source (top panel): own elaboration based on spectral data.
+Source (spectral data): adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
+\label{fig:phosphor_spectrum}
+\end{figure}
 
@@
 
-\begin{table}[!ht]
+\begin{table}[h!]
+\small
     \centering
@@
     \hline
-        \# & sector & role & country & expertise \\ \hline
+        \textbf{\#} & \textbf{sector} & \textbf{role} & \textbf{country} & \textbf{expertise} \\ \hline
         1 & academia & sr. researcher & UK & epitaxy \\ \hline

### 2023-01-26 2_SSL_EES/article/article_main.tex
updated main document and supplementary information

@@
 \cite{azevedo2009transition}.
-See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
+See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
  \label{fgr:history_efficacy}
@@
 Incandescent light bulb price is assumed constant based on the average in the covered time period.
-See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
+See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
   \label{fgr:cost_lamp_small}
@@
 For example, durability, low up-front cost and high efficiency of LED light sources have led to their wide-spread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
-LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical}, \cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
+LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical}\cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
 
@@
 Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Information, section 4 (Complete List of Sources for Figures), for the full list of sources and references.}
+See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
  \label{fgr:chip_architecture_overview}
@@
 - textured.
- Source: own synthesis of data from the full list of sources provided in the Supplementary Information, Section 4.
+ Source: own synthesis of data from the full list of sources provided in the Supplementary Information, Section 5.
 Data on average device performance adapted from U.S.
@@
 Source (Efficiency): own elaboration based on data in Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.
-Source (spectral data): Adapted from 2003\cite{Mueller2002}, 2010\cite{MuellerMach2005}, 2016\cite{Murphy2015}, 2019\cite{Hoerder2019}.}
+Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
  \label{fgr:waterfall}
@@
 
+Our findings are further supported by sensitivity analysis, presented in the Supplementary Information section SI4, where we find that sensitivity of the model to variation in its main parameters decreases over time with the increase of the number of DPW.
+In SI section 5 we also provide a comparison of our model with past cost calculations and projections published by US DOE on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
+
 \begin{figure*}[h!]
@@
 
+\section{Discussion}
+
+\subsection{Performance Improvements}
+
+Our findings show that the overall white LED efficiency has increased from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020.
+This efficiency increase was predominantly the result of a series of innovations in the LED architecture and materials as well as changes in the manufacturing process.
+Our comparison of improvements in device sub-efficiencies between 2003 and 2020 shows that the overall efficiency improvement was not dominated by any single sub-efficiency channel.
+Instead, there has been consistent progress across all device sub-efficiencies in the recorded period.
+Some sub-efficiencies, such as forward voltage efficiency, internal quantum efficiency and light extraction efficiency, are now within $\sim~10\%$ of their respective physical limits.
+A marked trend is the increasing relative contribution of losses in channels with physical limits on efficiency due to the underlying physics, such as light conversion efficiency of phosphors determined by the Stokes shift.
+Spectral efficiency also remains at $\sim17\%$ from its physical limit, showing potential for further improvements.
+Research on improving LED performance across efficiency loss channels continues \cite{cho2017white}\cite{Weisbuch2020}.
+
+We also find that there has been a significant progress in consumer experience metrics, which was driven by innovations in the means of conversion of blue light generated by conventional GaN LEDs into white light required for general illumination.
+The first commercial white LEDs produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{BANDO1998}.
+After a series of breakthrough innovations in phosphors represented in Figure 8 and described in SI section 3, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
+
+Overall, we find that LED technology innovations made crucial contributions to the progress in both LED performance and consumer experience metrics across the entire period covered by our study.
+Our interviews have also revealed an important role of incremental process improvements and learning-by-doing\cite{WRIGHT_1936}\cite{Arrow_1962} in the progress in LED efficiency.
+However, further research is needed to quantify the contribution of learning by doing to this progress.
+
+\subsection{Cost Reductions}
+
+Our LED manufacturing cost model shows that the cost of manufacturing low-to-mid power GaN-based white LED packages with classical chip architecture at a U.S.
+location using state-of-the-art equipment has decreased by 95.5\% from 1.11\$(2020) in 2003 to 0.05\$(2020) in 2020.
+In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost happened mostly due to economies of scale stemming from an increase in the wafer size and higher yield per manufacturing step.
+Both were enabled by advances in manufacturing equipment performance and incremental process improvements from learning-by-doing.
+
+Notably, our bottom-up cost model is constructed to provide process-step resolution across three different key chip architectures: classical chips, flip chips, and chip-scale package flip chips.
+However, in this study we were able to collect data and compare the outcomes only for the classical chip architecture.
+Collecting the full set of data needed to populate the model for the remaining two architectures would require access to proprietary information from industry.
+With this limitation, tracking manufacturing cost declines across three key LED chip architectures remains a topic for future work.
+
+\subsection{Contribution of Technology Spillovers}
+
+We find that 9 technology spillovers identified in our study, listed in Table \ref{tab:spillovers}, affected all dimensions of white LED performance.
+Three spillovers associated with the use of YAG/YGAG phosphors in LEDs played the key role in the first commercial white LED lighting products, essentially enabling the solid-state lighting market and industry of today.
+Spillovers also had a significant effect on physical device performance, cumulatively contributing to 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020.
+Technology spillovers were particularly important for consumer experience performance of white LED lighting sources, with corresponding spillover-driven innovations responsible for $\sim$100\% of the improvements in consumer experience metrics.
+
+Following the framework for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021}, we gathered information from the interviews, historical records, literature reviews and industry publications to analyze the sources, mechanisms, and enablers of the identified spillovers into the white LED technology listed in Table 4.
+Among the spillover sources, we find that all 9 spillovers had origins in basic science disciplines such as various branches of chemistry, materials science, optics and photonics, and solid-state physics.
+Five spillovers also utilized technical knowledge and expertise in cathode ray tubes, fluorescent lighting, optoelectronic devices, nanotechnology, and nature-inspired material design.
+
+Among the spillover mechanisms, six spillovers (involved in all phosphors except PFS and SLA, plus ITO) were a result of application of external scientific and technical knowledge already available to researchers and inventors.
+Three remaining spillovers (involved in PFS and SLA phosphors, plus PSS) occurred as an outcome of targeted search for relevant external knowledge outside the LED domain.
+In addition, at least two spillovers (involved in 258 and PFS phosphors) occurred through direct R\&D collaboration.
+
+Among important enabling factors for the identified spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
+
+We find that, on average, it took 26 years from the initial scientific discovery or invention to the moment of its spillover into the LED domain, with this time varying from 5 to almost 70 years.
+In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED market.
+
 \section{Conclusions}
 
-Our study has analysed in a systematic and granular way the sources of dramatic cost reductions and efficiency improvements in white light-emitting diodes since their introduction to the market in 1996.
-We find that the total LED device efficiency increased from $5.8\%$ to $38.7\%$ between 2003 and 2020 has been predominantly driven by LED technology innovations that have affected all physical energy loss channels and corresponding device sub-efficiencies.
-We also find that among those innovations, at least nine were driven by technology originating in areas of science and technology outside LEDs and solid-state lighting—i.e., those nine innovations can be referred to as technology spillovers .
-These spillovers were responsible for $8.5\%$ of the total efficiency improvements and nearly $100\%$ of the improvements in consumer experience metrics.
+Our study has analysed in a systematic and granular way the sources of dramatic cost reductions and performance improvements in white light-emitting diodes since their introduction to the market in 1996.
+We find that the total LED device efficiency increase from 5.8\% to 38.7\% between 2003 and 2020, as well as improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that affected all physical energy loss channels and corresponding device sub-efficiencies.
+We also find that among those innovations, at least nine were driven by knowledge spillovers originating in areas of science and technology beyond LEDs or solid-state lighting.
+These spillover-driven innovations were responsible for 8.5\% of the total efficiency improvements and nearly 100\% of the improvements in consumer experience metrics.
 
-Our manufacturing cost model shows that a $95.5\%$ decrease in the cost of producing white classic-chip LEDs (from $1.11\$$ to $0.05\$$ in 2020 USD) between 2003 and 2020.
-While efficiency improvements have played an important role in these cost reductions, the largest components of such cost reductions have been driven by increases in the wafer size and yields across different manufacturing steps.
+Our manufacturing cost model shows a 95.5\% decrease in the cost of producing white classic-chip LEDs (from 1.11\$(2020) to 0.05\$(2020)) between 2003 and 2020, driven mostly by increases in the wafer size and yields across different manufacturing steps over time.
 In contrast with efficiency improvements, these increases were mainly a result of learning-by-doing and manufacturing equipment improvements rather than LED technology innovations or spillovers.
 
-Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of efficiency improvements, highlights the critical role played by a growing deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs.
-Specifically, deep physical understanding of LED device efficiency loss channels enabled important breakthroughs and spillovers in LEDs and will enable further innovations to improve LED and SSL efficiency in the future, as expected by eminent experts in the field \cite{Weisbuch2020}.
-This suggests that additional research in these areas and a more deliberate search for spillovers may accelerate these expected future advances.
-Our findings on spillover enablers show that these efforts can be further supported or even accelerated by knowledge exchange events and long-term partnerships between academia and industry, dedicated mission-driven public R\&D funding, and freedom of search in academia.
-This reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles} \cite{narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
+Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of improvements in LED efficiency and consumer experience metrics, highlights the critical role played by a  deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.
+Specifically, deep physical understanding of LED device efficiency loss channels enabled important innovations in LEDs and will continue enabling future improvements in LED and SSL efficiency, as expected by eminent experts in the field \cite{Weisbuch2020}.
+This suggests that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology.
+These knowledge spillovers can be enabled or even accelerated by knowledge exchange events and long-term partnerships between academia and industry, dedicated mission-driven public R\&D funding, and a freedom of search in academia.
+This further reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles}\cite{narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
 
 There are various important avenues of future research that are opened up by our analysis.
-First, future work could expand the cost model by collecting and including data for a broader set of chip architectures and analyzing the impact of individual innovations and spillovers on costs as opposed to performance.
-Second, a deeper dive on the role of learning-by-doing is needed both in the cost and performance analysis to identify different types of learning by doing and how it came about.
+First, future work could expand the cost model by collecting and including data for a broader set of chip architectures and analyzing the impact of individual innovations and spillovers on the costs as opposed to performance.
+Second, a deeper dive on the role of learning-by-doing is needed both in the cost and performance analysis.
 Third, building on the work on LED sub-efficiencies and physical limits, future efforts could focus on identifying priority areas for further efficiency improvements in LEDs and SSL in general.
@@
 \section*{Acknowledgements}
-This research is supported by the grant from the Alfred P.
+This research was supported by the grant from the Alfred P.
 Sloan Foundation titled \textit{“What factors drive innovation in energy technologies?
 The role of technology spillovers and government investment”}.
-Michael Weinold further gratefully acknowledges financial support from the Swiss Study Foundation.
+Michael Weinold further gratefully acknowledges support from the Swiss Study Foundation.
 We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
-We also express our deep gratitude to all interviewees for their willingness to participate in this study and their invaluable contributions.
-Support by German consumer organization Stiftung Warentest in providing access to their archive is gratefully acknowledged.
+We also express our deepest gratitude to all interviewees for their willingness to participate in this study and share their invaluable insights with us.
 

### 2023-01-26 2_SSL_EES/article/article_supplementary.tex
updated main document and supplementary information

@@
 \section{Manufacturing Cost Model}
+\label{sec:costmodel}
 
@@
 Hatched bars are projections given by the sources.
-Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.}
+Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.
+Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
+Sources for bottom panel: own elaboration based on the cost model described in Section \ref{sec:costmodel}.
+Sources for Inset Plot: own elaborations based on numerical approximations for die-per-wafer estimations provided in \cite{de2005investigation}, informed by expert interviews.}
 	\label{fig:costmodel_calibration}

### 2023-01-26 2_SSL_EES/article/article_main.tex
added content to main publication

@@
 
-The historical improvements in four sub-efficiency channels—forward voltage efficiency, internal quantum efficiency, droop, and light extraction efficiency—identified in our study are shown in Figure 4.
-Using data on these four sub-efficiencies, as well as on improvements in spectral efficiency and light conversion efficiencies calculated from spectral data for LED phosphors provided in SI section 2 and discussed in the context of consumer experience improvements in section 5.2, we then calculate the overall LED efficiency in 2003 and 2010, shown in Figure XX.
-as a waterfall diagram across all sub-efficiencies, and in 2016 and 2020, similarly shown in Figure .
-The overall white LED lamp efficiency improved from $\eta_L(2003)=5.8\% \rightarrow \eta_L (2010) = 12.6\% \rightarrow \eta_L = 32.5\% \rightarrow \eta_L(2020) = 38.7\%$, as shown in Figure \ref{fgr:waterfall}.
-No single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
-We note, however, that the loss channels with a fixed physical limit, e.g., the Stokes loss that contributes to the conversion efficiency by phosphors, became more dominant in 2016 and 2020 compared to 2003 and 2010.
+The historical improvements in four sub-efficiency channels—forward voltage efficiency, internal quantum efficiency, droop, and light extraction efficiency—identified in our study are shown in Figure \ref{fgr:subeff}.
+Using data on these four sub-efficiencies, as well as on improvements in spectral efficiency and light conversion efficiencies calculated from spectral data for LED phosphors provided in SI section 2 and discussed in the context of consumer experience improvements in section 5.2, we then calculate the overall LED efficiency in 2003 and 2010, shown in Figure 5 as a waterfall diagram across all sub-efficiencies, and in 2016 and 2020, similarly shown in Figure 6.
+The overall white LED lamp efficiency improved from $\eta_L=5.8\%$ in 2003 to $12.6\%$ in 2010, $32.5\%$ in 2016 and finally to $38.7\%$ in 2020.
+As shown in the panels of Figure \ref{fgr:waterfall}, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
+We note, however, that the loss channels with a fixed physical limit, e.g., Stokes loss that contributes to the conversion efficiency by phosphors, became more dominant in 2016 and 2020 compared to 2003 and 2010.
 
 Figure \ref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
-The list of technology spillovers identified in our study is provided and discussed below in Section \ref{subsec:spillovers}.
-Through the index decomposition analysis described in section \ref{subsec:metrics}, we find that in the overall LED efficiency increase of 32.9\% from 5.8\% to 38.7\% between 2003 and 2020, at least 2.8 of these percentage points can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
+The list of technology spillovers identified in our study is provided in Table \ref{tab:spillovers} and discussed in Section \ref{subsec:spillovers}.
+Through the index decomposition analysis described in Section \ref{subsec:metrics}, we find that in the overall LED efficiency increase of 32.9\% from 5.8\% to 38.7\% between 2003 and 2020, at least 2.8\% of which can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
 In Figure \ref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of corresponding loss channels.
-Notably, sub-efficiencies for the most recent devices considered in our study are only $\sim10\%$ below the physical limit.
-The exception is spectral efficiency, which at $\sim20\%$ below the physical limit shows larger potential for further improvements.
+We find that there has been consistent progress across all device sub-efficiencies in the recorded period.
+Specifically, between 2003 and 2020, forward voltage efficiency increased from 70\% to 99.5\%, internal quantum efficiency from 55\% to 90\%, electrical droop from 65\% to 90\%, light extraction efficiency from 60\% to 90\%, spectral efficiency from 74\% to 83\%, conversion efficiency (red) from 11\% to 45\%, conversion efficiency (green) from 19\% to 61\%.
+Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits.
+The exception is spectral efficiency, which at $\sim17\%$ below the physical limit shows larger potential for further improvements.
 

### 2023-01-29 2_SSL_EES/article/article_supplementary.tex
added innovations table to supplementary information

@@
 \newpage
+\section{Identified LED Innovations}
+
+\begin{table}[H]
+    \begin{tabularx}{\textwidth}{ |l|X|X|l|l| }
+        \multicolumn{5}{c}{Forward Voltage Efficiency (VFE) $\eta_{Vf}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            1983 & Indium Tin Oxide \newline Current Spreading Layer & Contact Resistance & Yes (T,S) & \cite{margalith1999indium}\\
+        \hline
+            1998 & Digited Electrodes & Contact Resistance & Weak & \cite{steigerwald2001electrode} \\
+        \hline
+            ongoing & Epitaxy Improvements \newline and better Doping & Polarization and \newline Bulk Resistance & No & I \\
+        \hline
+        $\sim$2000 & Silver p-Contacts & Contact Resistance & No & \cite{kondoh2001nitride} \\
+        \hline
+        \multicolumn{5}{c}{Light Extraction Efficiency (LEE) $\eta_{LE}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            $<$ 2003 & Optimization for  \newline Cavity Effects & Other & No & I, \cite{Shen2003} \\
+        \hline
+            (1973) 1993 & "Chip \newline Randomization" & Total Reflection \newline Absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
+        \hline
+            1993 & Thin-Film \newline Chip Architecture & Absorption & No & I, \cite{Schnitzer1993} \\
+        \hline
+            $<2001$ & Patterned Sapphire \newline Substrate  & Total Reflection \newline Absorption & Yes (T,S) & I, \cite{Tadatomo2001} \\
+        \hline
+            ongoing & Chip Design for \newline high LEE & Total Reflection \newline Absorption & No & I, \cite{Haerle2004} \\
+        \hline
+            $\sim$2000 & Silver p-Contacts & Absorption & No & I, \cite{kondoh2001nitride} \\
+        \hline
+        \multicolumn{5}{c}{Internal Quantum Efficiency (IQE) $\eta_{IQ}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            1994 & DH & $>$ Radiative Recombination \newline Probability & No & \cite{Nakamura1994} \\
+        \hline
+            1996 & MQW & $>$ Radiative Recombination \newline Probability & No & \cite{Koike1996} \\
+        \hline
+            ongoing & Active Region \newline Doping & Both & No & I, \cite{schubert2018light} \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & Both & No & I \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements & Both & No & I \\
+        \hline
+        \multicolumn{5}{c}{(Phosphor) Conversion Efficiency (CE) $\eta_{C}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            ongoing & Lower Current Density & Current Density & No & I \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect Getting \newline Underlayer & Defect Density & No & I, \cite{haller2017burying} \\
+        \hline
+        \end{tabularx}
+    \caption{Technology breakthroughs and spillovers related to improvements in forward voltage efficiency, light extraction efficiency, internal quantum efficiency and conversion efficiency.
+Year column indicates first identified use in solid-state lighting.
+S/O column indicates if the breakthrough has been identified as a spillover.
+Source column additionally indicates if the spillover has been identified as part of an expert interview (I).
+Spillover sources are indicated in brackets and follow the taxonomy developed in \cite{Stephan_2021}: Other technology (T), other sector (S), other scientific discipline (D).}
+    \label{tab:subeff_vfe}
+\end{table}
+
+\newpage
 \section{Phosphors}

### 2023-02-04 2_SSL_EES/article/article_main.tex
finalized figure captions

@@
  \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/history_efficacy.pdf}
- \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history from 1850 to 2020.
-Data shown for best performers by year of market introduction, with dashed lines.
-Dashed lines indicate the average improvement calculated from a 3rd-order polynomial fit to the data.
+ \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.
+Data points indicate best performers by year of market introduction.
+Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data.
 The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
@@
  \centering
- \includegraphics[width=17.5cm]{2_SSL_EES/article/figures/subefficiencies_progress.pdf}
+ \includegraphics[width=17cm]{2_SSL_EES/article/figures/subefficiencies_progress.pdf}
  \caption{Historical improvements in light-emitting diode technology.
@@
  \textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
+The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given.
 Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
@@
 Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
-Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for five years: 1997, 2003, 2010, 2016, and 2020.
+Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for the four years used in Figure \ref{fgr:waterfall}: 2003, 2010, 2016, and 2020.
+To provide additional historical context, data for 1997 is included for selected sub-efficiencies.
+"N/A" denotes sub-efficiencies where 1997 data could not be calculated to different reasons: $\eta_{Vf}$, $\eta_{droop}$ depend on the device current, which was below 350mA in 1997, making comparison with contemporary devices difficult.
+$\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997.
 Physical limits on sub-efficiencies are indicated by black horizontal lines.
 The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
-$\eta_Vf$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_LE$ - light extraction efficiency; $\eta_(C_ER)$ - conversion efficiency for red phosphors; $\eta_(C_(EY/G))$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table 3.
-"N/A" denotes sub-efficiencies that could not be calculated for 1997 due to $\eta_Vf$, $\eta_droop$ depending on current, which was below 350mA at the time, and warm white spectrum LEDs not available at the time for $\eta_(C_ER)$,$\eta_(C_(EY/G))$  and $\eta_S$ The physical limit on $\eta_Vf$ exceeds 100\% and depends on electrical device parameters \cite{david2016electrical}.
+$\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.
+Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}.
 Source: own elaboration based on data in Figure \ref{fgr:waterfall} and Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.}
@@
 Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
-The corresponding LED architectures from Figure \ref{fgr:chip_architecture_overview} are shown for reference: 2003 (top panel) - flip-chip with YGAG phosphor; 2010 (bottom panel) - flip-chip with 258 phosphor; 2016 (top panel) - flip-chip with PFS phosphor; 2020 (bottom panel) - flip-chip with SALON phosphor.
-Details on phosphors are provided in Table \ref{tab:phosphors} and SI section XX.
-Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003 and $\eta_L = 12.7\%$ in 2010.
+The corresponding LED architectures and associated down-conversion phosphors are shown for reference: 2003 - flip-chip with YGAG phosphor; 2010 - flip-chip with 258 phosphor; 2016 - flip-chip with PFS phosphor; 2020 - flip-chip with SALON phosphor.
+A more complete overview of LED chip architectures is provided in Figure \ref{fgr:chip_architecture_overview}.
+Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and development history collected in Section 4 in the supplementary information.
+Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003, $\eta_L = 12.7\%$ in 2010, $\eta_L = 32.6\%$ in 2016 and $\eta_L = 38.8\%$ in 2010.
 Abbreviations: Scat.
@@
     \end{tabularx}
-    \caption{LED light down conversion inventions related to improvements in consumer experience metrics identified in this study.
-Note: Year column represents the earliest reported application of invention in white LEDs.
-These years are different from the years in Figure \ref{fgr:consumer_experience} that correspond to the earliest publication of spectral data for representative LED products  that relied on those inventions.
-Section in SI refers to a section in the Supplementary Information that contains references and a detailed description of the history of corresponding invention.
+    \caption{LED down-conversion materials (phosphors and quantum dots) related to improvements in consumer experience metrics identified in this study.
+The \textit{year} column represents the earliest reported application of invention in white LEDs.
+These differ from the years used in Figure \ref{fgr:consumer_experience} that correspond to the earliest publication of spectral data for representative LED products that relied on those materials.
+The \textit{SI Sec.} column refers to the corresponding section in the Supplementary Information that contains references and a detailed description of the history of corresponding invention.
 Abbreviations: Desig.
@@
  \caption{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.
-Circle dots show data for the earliest identified representative white LED products with published spectral data that used phosphors listed in Table 3, each indicated by the phosphor label and publication year.
-Corresponding spectral data used to calculate colour temperature values is provided in Figure SI1 in SI section 2.
+Data points show the color temperature and colour rendering index performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year.
+Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 9 in section 4 of the supplementary information.
 Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison.
 Horizontal lines represent typical colour temperatures of “traditional” light sources, shown for reference.
-Desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures.
+The desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures.
 Horizontal red arrow indicates desirable higher values of colour rendering index (CRI).
@@
 Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
+For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the supplementary information.
 Abbreviations: Litho – Lithographic Process, Insp.

### 2023-02-04 2_SSL_EES/article/article_supplementary.tex
finalized figure captions

--- 2023-01-26 2_SSL_EES/article/article_supplementary.tex
@@
 \usepackage{chemformula}
+
+\renewcommand{\figurename}{Figure SI}
+\renewcommand{\tablename}{Table SI}
 
@@
 
+\section{Identified LED Innovations}
+
+\begin{table}[H]
+    \begin{tabularx}{\textwidth}{ |l|X|X|l|l| }
+        \multicolumn{5}{c}{Forward Voltage Efficiency (VFE) $\eta_{Vf}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            1983 & Indium Tin Oxide \newline Current Spreading Layer & Contact Resistance & Yes (T,S) & \cite{margalith1999indium}\\
+        \hline
+            1998 & Digited Electrodes & Contact Resistance & Weak & \cite{steigerwald2001electrode} \\
+        \hline
+            ongoing & Epitaxy Improvements \newline and better Doping & Polarization and \newline Bulk Resistance & No & I \\
+        \hline
+        $\sim$2000 & Silver p-Contacts & Contact Resistance & No & \cite{kondoh2001nitride} \\
+        \hline
+        \multicolumn{5}{c}{Light Extraction Efficiency (LEE) $\eta_{LE}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            $<$ 2003 & Optimization for  \newline Cavity Effects & Other & No & I, \cite{Shen2003} \\
+        \hline
+            (1973) 1993 & "Chip \newline Randomization" & Total Reflection \newline Absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
+        \hline
+            1993 & Thin-Film \newline Chip Architecture & Absorption & No & I, \cite{Schnitzer1993} \\
+        \hline
+            $<2001$ & Patterned Sapphire \newline Substrate  & Total Reflection \newline Absorption & Yes (T,S) & I, \cite{Tadatomo2001} \\
+        \hline
+            ongoing & Chip Design for \newline high LEE & Total Reflection \newline Absorption & No & I, \cite{Haerle2004} \\
+        \hline
+            $\sim$2000 & Silver p-Contacts & Absorption & No & I, \cite{kondoh2001nitride} \\
+        \hline
+        \multicolumn{5}{c}{Internal Quantum Efficiency (IQE) $\eta_{IQ}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            1994 & DH & $>$ Radiative Recombination \newline Probability & No & \cite{Nakamura1994} \\
+        \hline
+            1996 & MQW & $>$ Radiative Recombination \newline Probability & No & \cite{Koike1996} \\
+        \hline
+            ongoing & Active Region \newline Doping & Both & No & I, \cite{schubert2018light} \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & Both & No & I \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements & Both & No & I \\
+        \hline
+        \multicolumn{5}{c}{(Phosphor) Conversion Efficiency (CE) $\eta_{C}$} \\
+        \hline
+            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+        \hline
+            ongoing & Lower Current Density & Current Density & No & I \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect Getting \newline Underlayer & Defect Density & No & I, \cite{haller2017burying} \\
+        \hline
+        \end{tabularx}
+    \caption{Technology breakthroughs and spillovers related to a subset of the sub-efficiencies considered.
+For a list of innovations in spectral efficiency, compare Table 1 in the main article.
+Listed are improvements in forward voltage efficiency, light extraction efficiency, internal quantum efficiency and conversion efficiency.
+No  Year column indicates first identified use in solid-state lighting.
+S/O column indicates if the breakthrough has been identified as a spillover.
+Source column additionally indicates if the spillover has been identified as part of an expert interview (I).
+Spillover sources are indicated in brackets and follow the taxonomy developed in \cite{Stephan_2021}: Other technology (T), other sector (S), other scientific discipline (D).
+}
+    \label{tab:innovations}
+\end{table}
+
 \newpage

### 2023-02-04 2_SSL_EES/article/article_main.tex
updated first sections

@@
 
-\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{First commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
-Since then, white LEDs have experienced major improvements in performance and decreases in cost, resulting in rapid market expansion of LED-based solid-state lighting (SSL), one of the key current solutions in global climate change mitigation efforts.
+\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{The first commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
+Since then, white LEDs have experienced major improvements in performance and efficiency as well as significant decreases in cost, resulting in a rapid expansion of the market share of LED-based solid-state lighting (SSL), making the technology one of the key solutions in global climate change mitigation efforts.
 Despite the significance of the technology, the extent and sources of these improvements have not been systematically investigated.
 Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies for climate change mitigation.
-With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from a literature review, interviews with experts in industry and academia, and cost and performance modelling.
-We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020.
+With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from cost and performance modelling, augmented by a literature review and interviews with experts in industry and academia.
+We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.8\%$ in 2020.
 During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S.
 location using state-of-the-art equipment has dropped from 1.1\$ to 0.05\$ (in 2020 USD) between 2003 and 2020, a 95.5\% decrease.
-We also show that technology spillovers — i.e., knowledge originating in other technologies—affected all performance dimensions of LEDs, contributing no less than 8.5\% of the total efficiency improvements and $\sim$100\% of the improvements in consumer experience metrics.
+We further show that technology spillovers — i.e., knowledge originating in other technologies — affected all performance dimensions of LEDs, contributing no less than 8.5\% to total efficiency improvements and $\sim$100\% to the improvements in consumer experience metrics, thereby playing a key role in the widespread adoption of the technology.
 The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}
@@
 %If you article does not have ESI please remove the the \dag symbol from the title and the footnotetext below.
-\footnotetext{\dag~Electronic Supplementary Information (ESI) available: [details of any supplementary information available should be included here]. See DOI: 00.0000/00000000.}
+\footnotetext{\dag~Electronic Supplementary Information (ESI) available: [details of any Supplementary Information available should be included here]. See DOI: 00.0000/00000000.}
 %additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required
@@
 A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}.
-According to the United Nations, by the end of 2021, more than 130 countries have set or are considering setting a net zero emissions target by or around 2050 \cite{un2021climate}.
-The European Union, for example, has set a goal of net zero emissions by 2050—a goal it aims to meet with the help of the European Green Deal, alongside other EU and national policies \cite{eu2020green}.
-The United Kingdom similarly has adopted a similar strategy to achieve net-zero by 2050 \cite{noauthor_ieairena_2023}.
+According to the United Nations, by the end of 2021, more than 130 countries have set or are considering setting a net zero emissions target by 2050 \cite{un2021climate}.
+The European Union, for example, has set a target of net zero emissions by 2050.
+It aims to meet this goal with the help of the European Green Deal, alongside other EU and national policies \cite{eu2020green}.
+The United Kingdom similarly has adopted a national strategy to achieve net-zero by 2050 \cite{noauthor_ieairena_2023}.
 
@@
 Data points indicate best performers by year of market introduction.
+Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye.
 Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data.
@@
 To address these questions, in this paper we identify a set of metrics suitable for tracking the historical progress of LED lighting technology.
-We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing  .
-Given the proprietary nature of knowledge in the SSL industry , we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{Tansey} with eminent experts from academia, public research institutions and industry.
+We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
+Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{Tansey} with eminent experts from academia, public research institutions and industry.
 This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.
@@
 
-Several descriptive publications provide a very useful overview of some of the scientific breakthroughs  that have contributed to LED progress \cite{krames2007status}\cite{Phillips2007}\cite{Bierhuizen2007}\cite{Nakamura2013}\cite{feezell2018invention}\cite{Taki2019} .
-Additional  studies providing more detail the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram) \cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
+Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status}\cite{Phillips2007}\cite{Bierhuizen2007}\cite{Nakamura2013}\cite{feezell2018invention}\cite{Taki2019}.
+Additional studies providing more detail the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
 However, these studies typically include limited information on the macro-level chip-design and cover disparate aspects of the technology over different time periods.
-Another important source of historical information comes from patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds \cite{margalith2011thin}, Samsung \cite{jung2014phos} \cite{cha2019semiconductor} and Soraa \cite{cich2017high}.
+Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos}\cite{cha2019semiconductor} and Soraa\cite{cich2017high}.
 
@@
 
-Historically, the progress in LED technology is commonly described by pointing to impressive improvements in LED device performance (i.e., brightness and electrical efficiency) and manufacturing cost reductions \cite{Taki2019}.
-However, solid-state lighting performance is characterized not only by LED brightness or electrical efficiency.
-Consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, has also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}.
+Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}.
+However,metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, has also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}.
 Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
@@
 
-Initially, as LED devices developed, the primary metric of progress in solid-state lighting was typically the luminous flux, as reflected in the so-called “Haitz’s Law” \cite{haitz1999case}\cite{haitz2011solid}.
-The luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
+During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
+Progress in this metric was described by “Haitz’s Law”, an observation on the increasing brightness and reduction in retail price of devices \cite{haitz1999case}\cite{haitz2011solid}.
+The luminous flux (total brightness) of early white LEDs was in facto too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
 In 2000, highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm) \cite{haitz2011solid}.
@@
 Even though some scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, by now, it fails to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
+
+Were therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology:
 
@@
 
-where $K$ is the luminous efficacy of radiation of the light source, can be computed from the spectrum and the luminosity function, which describes the sensitivity of the human eye.
+where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
 $K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
 
-\subsubsection{Overall Efficiency (Lamp Efficiency) $\eta_L$}
+\subsubsection{Overall Efficiency ("Lamp Efficiency") $\eta_L$}
 
-The overall efficiency of a light-emitting diode package is the product of all considered sub-efficiencies
+Finally, the overall efficiency of a light-emitting diode package is the product of all considered sub-efficiencies
 
@@
 Any metric relevant to customer experience can thus be calculated from the spectrum alone.
-The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the downconversion phosphor used in the device.
+The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the down-conversion phosphor used in the device.
 It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
 Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: Colour Rendering Index (CRI) and Colour Temperature.
-We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union /cite{weinold2020long}.
+We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
 This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
 It is thus beyond the scope of this work.
+
+\subsubsection{Color Rendering Index (CRI)}
 
@@
 Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
+
+\subsubsection{Colour Temperature}
 
@@
 All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
-A summary of the background of interviewed experts is provided in Table X in the supplementary information.
+A summary of the background of interviewed experts is provided in Table X in the Supplementary Information.
 
@@
 A more complete overview of LED chip architectures is provided in Figure \ref{fgr:chip_architecture_overview}.
-Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and development history collected in Section 4 in the supplementary information.
+Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and development history collected in Section 4 in the Supplementary Information.
 Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003, $\eta_L = 12.7\%$ in 2010, $\eta_L = 32.6\%$ in 2016 and $\eta_L = 38.8\%$ in 2010.
@@
 Data points show the color temperature and colour rendering index performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year.
-Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 9 in section 4 of the supplementary information.
+Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 9 in section 4 of the Supplementary Information.
 Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison.
@@
 Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
-For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the supplementary information.
+For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information.
 Abbreviations: Litho – Lithographic Process, Insp.

### 2023-02-04 2_SSL_EES/article/article_main.tex
minor edits

@@
 
-\section{Previous literature}
+\section{Previous Literature}
+\label{sec:prev_lit}
 
@@
 
-The evolution of LED device architecture and performance as well as the progress in understanding the underlying physical phenomena are well covered in the scholarly literature and patents, as described in section 2.
-However, information provided in such sources is insufficient for our goals on at least three accounts.
-First, existing work focusses  only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
+The evolution of LED device architecture and performance as well as the progress in understanding the underlying physical phenomena are well covered in the scholarly literature and patents, as described in Section \ref{sec:prev_lit}.
+However, information provided in such sources is insufficient for our goals on at least three accounts: First, existing work focuses only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
 Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
@@
 
-To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis that combines information obtained from a review of the primary scientific literature and relevant patents with our own computations of device sub-efficiencies, information gained from industry publications, semi-structured interviews with experts from academia and industry, and bottom-up manufacturing cost modelling.
+To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis that combines information obtained from a review of the primary scientific literature, device datasheets and relevant patents with our own computations of device sub-efficiencies, information gained from industry publications, semi-structured interviews with experts from academia and industry, and bottom-up manufacturing cost modelling.
 We provide the details of implementation of each of these methods below.
@@
 We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
-This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in the “Metrics” section; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
+This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in Section \ref{subsec:metrics}; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
 
 Relevant sources for the review were found in an iterative search process that involved two components.
-The first was the search in specialized patent and publication databases such as Google Scholar, Google Patents, and Scopus, as well as company websites.
-The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the “Previous literature” section and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+The first was the search in specialized patent and publication databases as well as company websites.
+The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the Section \ref{sec:prev_lit} and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
 We also relied on backward citations for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
@@
 
-To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews with eleven eminent experts  from academia, industry and public research sector .
-Experts were selected based on their engagement in  different sub-fields of LED research and manufacturing, as well as the recommendations from other interviewed experts, in essence expanding the list of experts that emerged from the literature review by what is known as a snow-ball sampling method.
+To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews with eleven eminent experts from academia, industry and the public research sector.
+Experts were selected based on their engagement in  different sub-fields of LED research and manufacturing, as well as the recommendations from other interviewed experts, in essence expanding the list of experts that emerged from the initial literature review.
 All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
-A summary of the background of interviewed experts is provided in Table X in the Supplementary Information.
+A summary of the background of interviewed experts is provided in Table SI 3 in Section 5 of the Supplementary Information.
 
@@
 The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
-Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in subsection 4.4) and verify device performance data.
+Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in Section \ref{sec:costmodel}) and verify device performance data.
 
@@
 
-The contribution of individual technology innovations and spillovers to the progress in the overall device efficiency over time is estimated by the index decomposition analysis, which is commonly used in energy economics \cite{Ang1997}.
+The contribution of individual technology innovations and spillovers to the progress in overall device efficiency over time is estimated by index decomposition analysis.
+Mathematically, this involves breaking down a chosen performance indicator into its constituent components, each representing a specific factor that contributes to the change in the indicator \cite{Ang1997}.
 Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
-It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
+It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
 
@@
 However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance.
-For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies?
-where possible and necessary.
+For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies where possible and necessary.
 Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications.
@@
 \subsection{Bottom-Up Manufacturing Cost Model}
+\label{sec:costmodel}
 
@@
 Given this situation, we developed our own bottom-up LED manufacturing cost model.
-The model structure is generally based on the 2012 \textit{LEDCOM} cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
 The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
@@
 
-The historical improvements in four sub-efficiency channels—forward voltage efficiency, internal quantum efficiency, droop, and light extraction efficiency—identified in our study are shown in Figure \ref{fgr:subeff}.
-Using data on these four sub-efficiencies, as well as on improvements in spectral efficiency and light conversion efficiencies calculated from spectral data for LED phosphors provided in SI section 2 and discussed in the context of consumer experience improvements in section 5.2, we then calculate the overall LED efficiency in 2003 and 2010, shown in Figure 5 as a waterfall diagram across all sub-efficiencies, and in 2016 and 2020, similarly shown in Figure 6.
-The overall white LED lamp efficiency improved from $\eta_L=5.8\%$ in 2003 to $12.6\%$ in 2010, $32.5\%$ in 2016 and finally to $38.7\%$ in 2020.
-As shown in the panels of Figure \ref{fgr:waterfall}, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
-We note, however, that the loss channels with a fixed physical limit, e.g., Stokes loss that contributes to the conversion efficiency by phosphors, became more dominant in 2016 and 2020 compared to 2003 and 2010.
+The historical development of a selection of the sub-efficiencies introduced in Section \ref{sec:device_performance_metrics} is shown in Figure \ref{fgr:subeff}.
+Using this data, together with data on the remaining sub-efficiencies not shown in this figure, we calculated the overall LED efficiency in for four years: 2003, 2010, 2016 and 2020.
+Figure \ref{fgr:waterfall} shows how improvements in sub-efficiencies led to improvements in overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.6\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
+As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
+We note, however, that the loss channels with a fixed physical limit, e.g., Stokes loss that contributes to the conversion efficiency by phosphors, has become more dominant in 2016 and 2020 compared to 2003 and 2010.
+This is a direct result of the large efficiency improvements of upstream sub-efficiencies.
 
@@
 The list of technology spillovers identified in our study is provided in Table \ref{tab:spillovers} and discussed in Section \ref{subsec:spillovers}.
-Through the index decomposition analysis described in Section \ref{subsec:metrics}, we find that in the overall LED efficiency increase of 32.9\% from 5.8\% to 38.7\% between 2003 and 2020, at least 2.8\% of which can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
+Through the index decomposition analysis described in Section \ref{subsec:metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
@@
 Historical improvements in consumer experience metrics for phosphor-converted warm white LEDs are shown in Figure \ref{fgr:consumer_experience}.
-A high colour rendering index (CRI) of an LED in combination with a specific range of possible colour temperatures is generally desirable in general illumination applications.
-The values of these metrics are determined by LED device spectra (provided and discussed in SI section X), which, in turn, depend on the means of light down conversion in the device.
-Due to that, we were able to establish the links between all major improvements in LED consumer experience metrics identified in this study and individual LED inventions associated either with phosphors or quantum dots used for light down conversion.
-The list of these inventions is provided in Table \ref{tab:spillovers}.
+In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable.
+Both metrics are determined by LED device spectra, which, in turn, depend on the properties of the down-conversion materials in the device.
+We were thus able to establish the links between all major improvements in the two consumer experience metrics identified in this study and individual LED inventions associated either with phosphors or quantum dots.
+The list of these inventions is provided in Table \ref{tab:spillovers}, while details of the device spectra are provided and discussed in Section 5 in the Supplementary Information.
 
-Notably, from detailed descriptions of the history of inventions in this list, which we provide in SI section 3, we find that only a single invention related to LED consumer experience improvements was originally developed specifically for application in solid-state lighting: the 2016 SALON phosphor compound \cite{seibald2019phosphor}\cite{Hoerder2019}.
+Notably, from detailed descriptions of the history of inventions in this list, which we provide in Section 5 of the Supplementary Information, we find that only a single invention related to LED consumer experience improvements was originally developed specifically for application in solid-state lighting: the 2016 SALON phosphor compound \cite{seibald2019phosphor}\cite{Hoerder2019}.
 All other inventions in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL.
-We discuss the details of corresponding technology spillovers in section XX below.
+We discuss the details of the corresponding technology spillovers in Section \ref{subsec:spillovers} below.
 
@@
 
-Results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages are presented in Figure \ref{fgr:costmodel.} We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
+Results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages are presented in Figure \ref{fgr:costmodel}.
+We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
 
@@
 \end{figure*}
-
-\subsection{Technology Spillovers}
-\label{subsec:spillovers}
 

### 2023-02-04 2_SSL_EES/article/article_main.tex
various minor changes

@@
 During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
-Progress in this metric was described by “Haitz’s Law”, an observation on the increasing brightness and reduction in retail price of devices \cite{haitz1999case}\cite{haitz2011solid}.
-The luminous flux (total brightness) of early white LEDs was in facto too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
-In 2000, highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm) \cite{haitz2011solid}.
+This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
+In 2000, highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case}\cite{haitz2011solid}.
 Today, however, highest performing devices yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
-LED brightness has become sufficient to enable the construction of lamps from multiple LED devices, with design improvements focusing on higher efficiency and quality of light instead of brightness.
-Even though some scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, by now, it fails to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
+LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead higher efficiency and quality of light instead of brightness.
+Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, by now, it fails to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
+Specifically,  the stagnating levels of luminous flux in highest performing devices fail to capture the majority of research, which serves to improve electrical efficiency at constant brightness.
 
@@
 Using this data, together with data on the remaining sub-efficiencies not shown in this figure, we calculated the overall LED efficiency in for four years: 2003, 2010, 2016 and 2020.
-Figure \ref{fgr:waterfall} shows how improvements in sub-efficiencies led to improvements in overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.6\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
+Figure \ref{fgr:waterfall} shows how improvements in sub-efficiencies led to improvements in overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.7\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
 As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
@@
 The list of technology spillovers identified in our study is provided in Table \ref{tab:spillovers} and discussed in Section \ref{subsec:spillovers}.
+The complete list of all LED innovations is provided in Table SI 2 in Section 3 of the Supplementary Information.
 Through the index decomposition analysis described in Section \ref{subsec:metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
@@
 
-Our findings are further supported by sensitivity analysis, presented in the Supplementary Information section SI4, where we find that sensitivity of the model to variation in its main parameters decreases over time with the increase of the number of DPW.
-In SI section 5 we also provide a comparison of our model with past cost calculations and projections published by US DOE on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
+Our findings are further supported by apreliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the model to variation in its main parameters decreases over time with the increase of the number of DPW.
+We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.
 
@@
 
-Our findings show that the overall white LED efficiency has increased from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.7\%$ in 2020.
+Our findings show that the overall white LED efficiency has increased from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.8\%$ in 2020.
 This efficiency increase was predominantly the result of a series of innovations in the LED architecture and materials as well as changes in the manufacturing process.
@@
 Instead, there has been consistent progress across all device sub-efficiencies in the recorded period.
-Some sub-efficiencies, such as forward voltage efficiency, internal quantum efficiency and light extraction efficiency, are now within $\sim~10\%$ of their respective physical limits.
+Some sub-efficiencies, such as forward voltage efficiency, internal quantum efficiency and light extraction efficiency, are now within $\sim10\%$ of their respective physical limits.
 A marked trend is the increasing relative contribution of losses in channels with physical limits on efficiency due to the underlying physics, such as light conversion efficiency of phosphors determined by the Stokes shift.
 Spectral efficiency also remains at $\sim17\%$ from its physical limit, showing potential for further improvements.
-Research on improving LED performance across efficiency loss channels continues \cite{cho2017white}\cite{Weisbuch2020}.
+Research on improving LED performance across efficiency loss channels continues\cite{cho2017white}\cite{Weisbuch2020}.
 
-We also find that there has been a significant progress in consumer experience metrics, which was driven by innovations in the means of conversion of blue light generated by conventional GaN LEDs into white light required for general illumination.
+We further find that there has been significant progress in performance across the set of consumer experience metrics, which was driven by innovations related to the conversion of blue light generated by conventional GaN LEDs into the white light required for general illumination.
 The first commercial white LEDs produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{BANDO1998}.
-After a series of breakthrough innovations in phosphors represented in Figure 8 and described in SI section 3, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
+After a series of breakthrough innovations in phosphors represented in Figure \ref{fgr:consumer_experience} and described in Section 4 of the Supplementary Information, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
 
@@
 location using state-of-the-art equipment has decreased by 95.5\% from 1.11\$(2020) in 2003 to 0.05\$(2020) in 2020.
-In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost happened mostly due to economies of scale stemming from an increase in the wafer size and higher yield per manufacturing step.
+In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost resulted from an increase in the wafer size used in manufacturing and higher yields across manufacturing steps.
 Both were enabled by advances in manufacturing equipment performance and incremental process improvements from learning-by-doing.
@@
 
-Our study has analysed in a systematic and granular way the sources of dramatic cost reductions and performance improvements in white light-emitting diodes since their introduction to the market in 1996.
-We find that the total LED device efficiency increase from 5.8\% to 38.7\% between 2003 and 2020, as well as improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that affected all physical energy loss channels and corresponding device sub-efficiencies.
+Our study has analysed the sources of the dramatic cost reductions and performance improvements in white light-emitting diodes since their introduction to the market in 1996 in a systematic and granular way.
+We find that the total LED device efficiency increase from 5.8\% to 38.8\% between 2003 and 2020, as well as improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that affected all physical energy loss channels and corresponding device sub-efficiencies.
 We also find that among those innovations, at least nine were driven by knowledge spillovers originating in areas of science and technology beyond LEDs or solid-state lighting.

### 2023-02-04 2_SSL_EES/article/article_main.tex
minor changes

@@
 
-In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must first highlight  the complexity of this task.
+In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must  highlight the complexity of this task.
 Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
-Using sales price information instead of the cost for the same purpose could be in some cases a promising alternative, as prices can be easily obtained for recent LED products directly from manufacturers.
-However, historical data on prices for LEDs is as difficult to obtain as for costs.
-In addition, sales price includes components not relevant to the progress in technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
+Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
+However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
+In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
 Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
+Instead, we describe progress in the economics of manufacturing devices by using yielded cost as the primary metric.
+
+The cost model we developed and describe in Section \ref{sec:costmodel} relies on a cumulative approach to yielded cost\cite{becker2001use}.
+In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
+
+\begin{equation}
+    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
+\end{equation}
+
+This cost metric is cumulative by definition, thus
+
+\begin{equation}
+    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
+\end{equation}
+
+Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
+Additional information on details of the computation of cost metrics and the different manufacturing steps is presented in Section 2 of the Supplementary Information.
 
@@
 
-The cost model we developed relies on a cumulative approach to yielded cost\cite{becker2001use}.
-In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
-
-\begin{equation}
-    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
-\end{equation}
-
-This cost metric is cumulative by definition, thus
-
-\begin{equation}
-    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
-\end{equation}
-
-Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
-Additional information on details of the computation of cost metrics and the different manufacturing steps is presented in the Supplementary Information.
-
 The aggregate result of the cost model is the manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
@@
 It instead assumes an ideal factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
-This ‘ideal’ factory is assumed to have the average size for LED facilities for each of the time periods under consideration—namely X packages/year for 2003, Y packages/year for 2012 and Z packages/year for 2020, as identified through the interviews.
 It also assumes the use of the most up-to-date equipment for that year.
@@
 $\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997.
+Note that indium tin oxide (ITO) affects different sub-efficiencies depending on the chip architecture in question.
+For instance, in modern flip-chip architectures, light-extraction efficiency no longer depends on ITO, as can be seen from Figure \ref{fig:chip_architecture_overview}.
 Physical limits on sub-efficiencies are indicated by black horizontal lines.
@@
 Our findings are further supported by apreliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the model to variation in its main parameters decreases over time with the increase of the number of DPW.
-We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.
+We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model\cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables\cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.
 
@@
         1967 & 1996 & <2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical}\cite{bando1998development}\cite{shimizu1999light}\cite{Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
-        1982 & 1996 & <2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered}\cite{feezell2018invention}\cite{Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ \\ \hline
-        1971 & 1999 & <2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf}\cite{fraser1972highly}\cite{margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (no more use of this material in most mid-/high-power devices due to different chip architectures, compare Figure \ref{fgr:chip_architecture_overview}) \\ \hline
+        1982 & 1996 & <2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered}\cite{feezell2018invention}\cite{Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare Figure \ref{fig:chip_architecture_overview})\\ \hline
+        1971 & 1999 & <2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf}\cite{fraser1972highly}\cite{margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare Figure \ref{fig:chip_architecture_overview}) \\ \hline
         1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997}\cite{mueller2004phosphor}\cite{MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline

### 2023-03-13 2_SSL_EES/article/article_main.tex
minor changes

@@
 Among these drivers, the role of external  knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area  \cite{Stephan2021}.
-While the exact definition of spillovers in literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012}, we follow Stephan et al.
+While the exact definition of spillovers in the literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012}, we follow Stephan et al.
 \cite{Stephan2021} and consider knowledge to be external - thus a spillover -  if it has been developed for application in other technologies, sectors, or scientific disciplines.
@@
 It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.
-As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years , which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
+As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
 For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W.
@@
 These  dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies  \cite{weinold2020long}\cite{Mills2014}\cite{Stegmaier2021}\cite{grubb2021new}.
-As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year for the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
+As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
 Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}.
-For example, durability, low up-front cost and high efficiency of LED light sources have led to their wide-spread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
+For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
 LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical}\cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
@@
 We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
-Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{Tansey} with eminent experts from academia, public research institutions and industry.
+Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry.
 This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.
 Data on performance improvements associated with individual innovations identified in patents, scientific publications, industry reports, and interviews is then augmented  by our calculations of device efficiency, including contributions of specific spillovers to the overall LED efficiency (see Supplementary Information for more detail).
-We further calculate LED manufacturing costs using own bottom-up cost model with a process-step resolution.
-Finally, we discuss what our observations and conclusions mean for the understanding of the innovation process in LED technology and for broader policy and industry efforts to accelerate clean energy innovation.
+We further calculate LED manufacturing costs using a bottom-up cost model with process-step resolution that we developed.
+Finally, we discuss what our observations may mean for our understanding of the innovation process in innovation process in LED technology and for broader policy and industry efforts to accelerate clean energy innovation.
 
@@
 
-The historical development of light-emitting diodes has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
+The historical development XXXX development into powerful lghting devices from very dim novelty XXXX of light-emitting diodes received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
 The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006}\cite{krames2007led}\cite{laubsch2009high}\cite{hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
@@
 Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status}\cite{Phillips2007}\cite{Bierhuizen2007}\cite{Nakamura2013}\cite{feezell2018invention}\cite{Taki2019}.
-Additional studies providing more detail the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
-However, these studies typically include limited information on the macro-level chip-design and cover disparate aspects of the technology over different time periods.
+Additional studies provide more detail regarding the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
+However, these studies typically include limited information on the macro-level chip design and cover disparate aspects of the technology over different time periods.
 Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos}\cite{cha2019semiconductor} and Soraa\cite{cich2017high}.
@@
 We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure \ref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
-Despite the amount of literature published on the topic of LED history, to the best of our knowledge, no publication comprehensively aggregates known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
-In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not been studied in the context of lighting yet.
+Despite the amount of literature published on the topic of LED history, we are not aware of a prior publication that comprehensively and consistently aggregates and analyzes known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
+In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
 This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012}\cite{Creutzig2018}.

### 2023-03-14 2_SSL_EES/article/article_main.tex
minor edits

@@
  \caption{Historical evolution of light-emitting diode chip architectures.
-Rows from top to bottom: classical chip with lateral current spreading; Osram’s thin GaN flip-chip (vertical) architecture; flip-chip architecture; chip-scale package flip-chip architecture.
+Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
 Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
@@
 Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}.
-However,metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, has also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}.
+However, metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}.
 Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
@@
 This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
-In 2000, highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
+In 2000, the highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
 Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case}\cite{haitz2011solid}.
-Today, however, highest performing devices yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
-LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead higher efficiency and quality of light instead of brightness.
-Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, by now, it fails to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
-Specifically,  the stagnating levels of luminous flux in highest performing devices fail to capture the majority of research, which serves to improve electrical efficiency at constant brightness.
+Today, however, the devices with the highest performance yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
+LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
+Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
+Specifically,  the stagnating levels of luminous flux in highest performing devices does not capture a major research focus, which is improving electrical efficiency at constant brightness.
 
-Were therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology:
+We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology.
 
@@
 
-(Efficiency) droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
-This is often treated separately from internal quantum efficiency in literature due to its importance in high-power devices.
+Efficiency droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
+This is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
 It is defined as
@@
 
-Thereby, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
+Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
 
@@
 
-(Light) conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
+Light conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
 These losses are the sum of the Stokes loss and well as scattering/absorption losses.
@@
 where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
-Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components (\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen).
+Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components, i.e.
+\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.
 
@@
 This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
-It is thus beyond the scope of this work.
+As a result, it is beyond the scope of this work.
 
@@
 CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
-However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de-facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
+However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
 High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
@@
 
-According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change  in a single sub-efficiency variable a between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
+According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
 
@@
 In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra.
-This approach allowed us to fully quantify the improvements related to phosphor development in LEDs.
+This approach allowed us to  quantify the improvements related to phosphor development in LEDs.
 
@@
 
-We address the limitations of data availability for the LED manufacturing cost by conducting bottom-up process-step resolution LED manufacturing cost modelling.
+We address the limitations of data availability for the LED manufacturing cost by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
 In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
@@
 The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-For each of these three periods, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
+For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
 Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips.
@@
 
-The aggregate result of the cost model is the manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
+The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
 Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
 We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
-It instead assumes an ideal factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
+It instead assumes an hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
 It also assumes the use of the most up-to-date equipment for that year.
@@
 
-In Figure \ref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of corresponding loss channels.
+In Figure \ref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels.
 We find that there has been consistent progress across all device sub-efficiencies in the recorded period.
@@
 Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits.
-The exception is spectral efficiency, which at $\sim17\%$ below the physical limit shows larger potential for further improvements.
+The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements.
 
@@
 Note that indium tin oxide (ITO) affects different sub-efficiencies depending on the chip architecture in question.
-For instance, in modern flip-chip architectures, light-extraction efficiency no longer depends on ITO, as can be seen from Figure \ref{fig:chip_architecture_overview}.
+For instance, in modern flip-chip architectures, light-extraction efficiency no longer depends on ITO, as can be seen from Figure \ref{fgr:chip_architecture_overview}.
 Physical limits on sub-efficiencies are indicated by black horizontal lines.
@@
 The \textit{year} column represents the earliest reported application of invention in white LEDs.
-These differ from the years used in Figure \ref{fgr:consumer_experience} that correspond to the earliest publication of spectral data for representative LED products that relied on those materials.
+These differ from the years used in Figure \ref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for representative LED products that relied on those materials.
 The \textit{SI Sec.} column refers to the corresponding section in the Supplementary Information that contains references and a detailed description of the history of corresponding invention.
@@
 
-Results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages are presented in Figure \ref{fgr:costmodel}.
+Figure \ref{fgr:costmodel} shows key results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages.
 We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
@@
 We capture this in the model by assuming the following wafer diameters and calculating the associated number of die per wafer (DPW) \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$.
-With more than thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time.
+With more than a thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time.
 As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in Figure \ref{fgr:costmodel}).
@@
 
-Our findings are further supported by apreliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the model to variation in its main parameters decreases over time with the increase of the number of DPW.
+Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
 We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model\cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables\cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.
@@
 location.
+Consecutive steps shown on the x-axis.
 Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
@@
 Our comparison of improvements in device sub-efficiencies between 2003 and 2020 shows that the overall efficiency improvement was not dominated by any single sub-efficiency channel.
-Instead, there has been consistent progress across all device sub-efficiencies in the recorded period.
+Instead, there has been consistent progress across all device sub-efficiencies during the time period under consideration.
 Some sub-efficiencies, such as forward voltage efficiency, internal quantum efficiency and light extraction efficiency, are now within $\sim10\%$ of their respective physical limits.
@@
 We further find that there has been significant progress in performance across the set of consumer experience metrics, which was driven by innovations related to the conversion of blue light generated by conventional GaN LEDs into the white light required for general illumination.
-The first commercial white LEDs produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{BANDO1998}.
-After a series of breakthrough innovations in phosphors represented in Figure \ref{fgr:consumer_experience} and described in Section 4 of the Supplementary Information, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
+The first commercial white LEDs produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{bando1998development}.
+After a series of breakthrough innovations in phosphors shown in Figure \ref{fgr:consumer_experience} and described in Section 4 of the Supplementary Information, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
 
@@
 In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost resulted from an increase in the wafer size used in manufacturing and higher yields across manufacturing steps.
+This can be seen in the reduction of the contribution of wafer processing steps to  totaal cost in Figure \ref{fgr:costmodel}.
 Both were enabled by advances in manufacturing equipment performance and incremental process improvements from learning-by-doing.
@@
 \subsection{Contribution of Technology Spillovers}
+\label{subsec:spillovers}
 
-We find that 9 technology spillovers identified in our study, listed in Table \ref{tab:spillovers}, affected all dimensions of white LED performance.
+We find that nine technology spillovers identified in our study, listed in Table \ref{tab:spillovers}, affected all dimensions of white LED performance.
 Three spillovers associated with the use of YAG/YGAG phosphors in LEDs played the key role in the first commercial white LED lighting products, essentially enabling the solid-state lighting market and industry of today.
@@
 Following the framework for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021}, we gathered information from the interviews, historical records, literature reviews and industry publications to analyze the sources, mechanisms, and enablers of the identified spillovers into the white LED technology listed in Table 4.
-Among the spillover sources, we find that all 9 spillovers had origins in basic science disciplines such as various branches of chemistry, materials science, optics and photonics, and solid-state physics.
+Among the spillover sources, we find that all nine spillovers had origins in basic science disciplines such as various branches of chemistry, materials science, optics and photonics, and solid-state physics.
 Five spillovers also utilized technical knowledge and expertise in cathode ray tubes, fluorescent lighting, optoelectronic devices, nanotechnology, and nature-inspired material design.
@@
 Our manufacturing cost model shows a 95.5\% decrease in the cost of producing white classic-chip LEDs (from 1.11\$(2020) to 0.05\$(2020)) between 2003 and 2020, driven mostly by increases in the wafer size and yields across different manufacturing steps over time.
-In contrast with efficiency improvements, these increases were mainly a result of learning-by-doing and manufacturing equipment improvements rather than LED technology innovations or spillovers.
+In contrast with efficiency improvements, these learning by doing, economies of scale and improvements were facilitated by both improvements in the performance as well as policies creating and growing market demand, such as incandescent light bans and subsidies for LEDs.
 
 Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of improvements in LED efficiency and consumer experience metrics, highlights the critical role played by a deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.
-Specifically, deep physical understanding of LED device efficiency loss channels enabled important innovations in LEDs and will continue enabling future improvements in LED and SSL efficiency, as expected by eminent experts in the field \cite{Weisbuch2020}.
+Specifically, deep physical understanding of LED device efficiency loss channels enabled important innovations in LEDs that increased several sub-efficiencies in LEDs and will continue, as expected by eminent experts in the field \cite{Weisbuch2020}.
 This suggests that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology.

### 2023-03-14 2_SSL_EES/article/article_main.tex
minor edits

@@
 \includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{The first commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
-Since then, white LEDs have experienced major improvements in performance and efficiency as well as significant decreases in cost, resulting in a rapid expansion of the market share of LED-based solid-state lighting (SSL), making the technology one of the key solutions in global climate change mitigation efforts.
-Despite the significance of the technology, the extent and sources of these improvements have not been systematically investigated.
-Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies for climate change mitigation.
-With this aim, we gather systematic evidence on cost and performance improvements in white LEDs from cost and performance modelling, augmented by a literature review and interviews with experts in industry and academia.
+Since then, white LEDs have experienced major improvements in performance and efficiency.
+Significant decreases in cost have been driven by a rapid expansion of the market share of LED-based solid-state lighting (SSL), which in turn has driven down costs even further.
+Today, the technology is one of the key solutions in global climate change mitigation efforts.
+However, despite its significance, the extent and sources of the underlying improvements have not been systematically investigated.
+Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies.
+With this aim, we gather and analyse systematic evidence on cost and performance improvements in white LEDs from cost and performance modelling, augmented by a literature review and interviews with experts in industry and academia.
 We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.8\%$ in 2020.
@@
 location using state-of-the-art equipment has dropped from 1.1\$ to 0.05\$ (in 2020 USD) between 2003 and 2020, a 95.5\% decrease.
-We further show that technology spillovers — i.e., knowledge originating in other technologies — affected all performance dimensions of LEDs, contributing no less than 8.5\% to total efficiency improvements and $\sim$100\% to the improvements in consumer experience metrics, thereby playing a key role in the widespread adoption of the technology.
+We further show that technology spillovers — i.e., knowledge originating in other technologies, sectors, or scientific disciplines — affected all performance dimensions of LEDs, contributing no less than 8.5\% to total efficiency improvements and $\sim$100\% to the improvements in consumer experience metrics identified, thereby playing a key role in the widespread adoption of the technology.
 The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}

### 2023-03-14 2_SSL_EES/article/article_supplementary.tex
minor edits

@@
 \usepackage{microtype}
-% figures
+% s
 \usepackage{graphicx}
@@
 \usepackage{chemformula}
+% smart references
+\usepackage{cleveref}
 
@@
 \renewcommand{\tablename}{Table SI}
+\renewcommand{\theequation}{SI \arabic{equation}}
 
@@
 
-\section{Lighting Metrics}
+\section{Additional Performance Metrics}
 
-\subsection{Luminous Efficacy (of Radiation)}
+\subsection{Luminous Efficacy of Radiation}
 \label{subsec:ler}
@@
 This metric can be computed from spectral data alone and does not require additional spectral normalization.
-It enables straightforward comparison between the performance of different downconversion phosphors, as shown in figure \ref{fig:se_efficacy} in chapter \ref{ch:results}.
-Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation, as the product with the luminosity function of figure \ref{fig:tristimulus} shows.
+It enables straightforward comparison between the performance of different downconversion phosphors, as shown in  \cref{fig:se_efficacy} in chapter \cref{ch:results}.
+Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation, as the product with the luminosity function of  \cref{fig:tristimulus} shows.
 Care must be taken not to confuse this efficacy metric with efficacy of source, which depends not only on the spectral power distribution of a light source.
 
-This metric describes the match of a light-emitting diode package spectrum to the human visual system. It thus describes only a part of the total device performance.
+This metric describes the match of a light-emitting diode package spectrum to the human visual system.
 
-\subsection{Luminous Efficacy (of Source)}
+\subsection{Luminous Efficacy of Source}
 \label{subsec:les}
@@
 
-\subsection{Arbitrary Units}
-\label{subsec:arbunit}
-
-In spectroscopy, an arbitrary unit (abbreviated "Arb.
-Unit") is a relative unit of measurement to show the ratio of the intensity to a predetermined reference.
-In the case of Figure \ref{fig:phosphor_spectrum}, the reference is the highest point in the spectrum.
-
 \newpage
@@
     \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to cost model outputs.
-For description of cost variables, see definitions for equation \ref{eqn:cost_wafer}.}
+For description of cost variables, see definitions for equation \cref{eqn:cost_wafer}.}
     \label{fig:costmodel-schematic}
@@
 
-A schematic diagram of the cost model developed in this thesis is presented in figure \ref{fig:costmodel-schematic}.
-The cost model is process step based.
+A schematic diagram of the cost model developed is presented in  \cref{fig:costmodel-schematic}.
+The cost model is process step-based.
 It is split between the upstream part of the manufacturing process and the downstream part of the manufacturing process.
@@
 The latter is performed at the package level.
-The model takes as inputs both parameters specific to manufacturing process steps (\textit{"step parameters"}) and parameters applying to all manufacturing steps (\textit{"global steps"}).
+The model takes as inputs both parameters specific to manufacturing process steps (\textit{"manufacturing step parameters"}) and parameters affecting all manufacturing steps (\textit{"global parameters"}).
 The cost for each process step is then computed.
-The cost model returns the total and step manufacturing costs.
+The cost model returns returns the costs of individual manufacturing steps as well as the total manufacturing cost.
 It further considers the yield per step and returns the cumulative yield, the yielded cost per step and the yielded total manufacturing cost.
 
-\subsection{Examples of Chip Architectures}
+\subsection{Manufacturing Process Steps by Chip Architecture}
 
-Figures \ref{fig:manuf_vtf_2012-1}-\ref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of two differente chip architectures considered in the cost model.
+Figures \cref{fig:manuf_vtf_2012-1}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
 
@@
             \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
-            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012. Continued on next page.}
+            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
+Abbreviations: MQW - multiple quantum well; LLO - laser lift-off.
+Continued on next page.}
             \label{fig:manuf_vtf_2012-1}
@@
             \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
-            \caption{(2/2) Continued from previous page.}
+            \caption{(2/2) Continued from previous page. Abbreviations: Encaps. - encapsulation.}
             \label{fig:manuf_vtf_2012-2}
@@
             \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
-            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020. Continued on next page.}
+            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+Abbreviations: MQW - multiple quantum well.
+Continued on next page.}
             \label{fig:manuf_csp_2020-1}
@@
             \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
-            \caption{(2/2) Continued from previous page.}
+            \caption{(2/2) Continued from previous page. Abbreviations: Encaps. - encapsulation.}
             \label{fig:manuf_csp_2020-2}
@@
 
-The manufacturing process of semiconductor devices can be categorized by the level of integration at which steps are implemented.
+The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level.
 While the upstream steps in the manufacturing flow are carried out at the wafer level, the packaging steps are typically carried out at the die/package level.
@@
 \label{eqn:cost_sum}
-    C \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
+    C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
 \end{equation}
@@
 \label{eqn:cost_wafer}
-    C_i \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
 \end{equation}
@@
 \begin{align*}
-        D &\dots \text{good die per wafer} \label{def:cost_wafer_first} \\
+        D &\dots \text{functional die per wafer (tested successfully)} \label{def:cost_wafer_first} \\
         y &\dots \text{process step yield} \\
-        u &\dots \text{equipment utilization} \\
+        u &\dots \text{equipment utilization (in relation to theoretical equipment capacity)} \\
         p &\dots \text{power consumption} \\
@@
         A &\dots \text{wafer area} \\
-        v_x &\dots \text{volume of substance per wafer} \\
-        p_x &\dots \text{price of substance $x$ per volume}\\
+        v_x &\dots \text{volume of material $x$ per wafer} \\
+        p_x &\dots \text{price of material $x$ per volume}\\
 \end{align*}
 
-The number of die per wafer $D$ depends on the total USAble wafer area.
-The USAble area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
+The number of die per wafer $D$ depends on the total usable wafer area.
+The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
 
 \begin{equation}
-	A_{\text{USAble}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
+	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
 \end{equation}
 
-Determining the USAble wafer area as a of these three parameters requires a numerical solution.
-However, following discussions in literature \cite{de2005investigation}, we approximate the number of good die per wafer as
+Determining the usable wafer area as a function of these three parameters requires a numerical solution.
+However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
 
 \begin{equation}
-label{eqn:dpw}
+\label{eqn:dpw}
 	D=\frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
@@
 
-which gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
+    \caption{Number of die per wafer for the different wafer sized used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW to $d(2020)=200$mm$\rightarrow26,838$DPW.
+Source: own elucidations, based on equation \cref{eqn:dpw}.}
+    \label{fig:dpw}
+\end{figure}
+
+Results are plotted in Figure \cref{fig:dpw}, showing the increase in die per wafer (DPW).
+Equation \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
 
@@
 \begin{split}
-    C_i \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
     &  \frac{1}{y_i}  \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
@@
 \label{eqn:cost_die}
-    C_j \bigg[ \frac{ \$(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((e*p) + l + m + d + o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x p_x \bigg\}
+    C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((e*p) + l + m + d + o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x p_x \bigg\}
 \end{equation}
@@
 
-Note that in keeping with the categorization introduced by the United States Department of energy, certain steps from these two categories are reported separately.
+Note that in keeping with the categorization introduced by the United States Department of Energy (cf.
+eg.
+\cite{doe_ssl_rnd_2019}), certain steps from these two categories are reported separately.
 In the wafer processing category, the epitaxy step is reported separately due to its complexity and the large share of cost carried.
@@
 
-Devices may be damaged or otherwise rendered unUSAble during the manufacturing process.
+Devices may be damaged or otherwise rendered unusable during the manufacturing process.
 The ratio between the number of good devices per step and the number of handled devices per step is known as the yield.
@@
 
-which can be compared to equation \ref{eqn:C_2} for a more general form.
+which can be compared to equation \cref{eqn:C_2} to find a more general form:
 
 \begin{equation}
-    C_2^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
+    C_{Y_2}^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
 \end{equation}
 
-Omitting a proof by induction, we get for the general term for a step $i>1$ repeated $n$ times an equation of the form
+It can be shown by induction that the general term for a step $i>1$ repeated $n$ times gives an equation of the form:
 
@@
 
-This cummulative approach to yielded cost is different from the approach taken in the original \textit{LEDCOM} model.
-It uses what in literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}.
-The yielded cost of a single process step $f$ is described by
+This cumulative approach to yielded cost is different from the approach taken in the original \textit{LEDCOM} model.
+It uses what in the literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}.
+Using this approach, the yielded cost of a single process step $f$ is described by
 
@@
 
-where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in equation \ref{eqn:series}.
+where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in equation \cref{eqn:series}.
 
@@
 
-To quantify the drivers of cost reductions across all process steps, one would need to identify the magnitude of contribution to cost reductions made by single variables in equations \ref{eqn:cost_wafer} and \ref{eqn:cost_die}.
+To quantify the drivers of cost reductions across all process steps, one would need to identify the magnitude of contribution to cost reductions made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
 Mathematically, given a function $C$ which describes the cost associated with manufacturing a single unit at time $t$,
@@
 
-The infinitesimal contribution to the total cost by the infinitesimal change in a cost model variable is defined through the total differential of the cost function \ref{eqn:cost_sum} as
+The infinitesimal contribution to the total cost by the infinitesimal change in a cost model variable is defined through the total differential of the cost function \cref{eqn:cost_sum} as
 
@@
 
-Using logarithmic differentiation, the integral from equation \ref{eqn:integral_1} can be rewritten as
+Using logarithmic differentiation, the integral from equation \cref{eqn:integral_1} can be rewritten as
 
@@
 
-This approach was used to determine the effect of spillovers on device performance, displayed in figure \ref{fig:breakthroughs_efficiency}.
+This approach was used to determine the effect of spillovers on device performance, displayed in  \cref{fig:breakthroughs_efficiency}.
 Due to time constraints, this approach could not be used to determine the effect of spillovers on manufacturing cost.
-Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of equation \ref{eqn:cost_sum}, could this method.
+Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of equation \cref{eqn:cost_sum}, could this method.
 
@@
 Wafers can be either unpatterned or patterned, where the latter has become commonplace in 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
-The price of sapphire substrates has decreased significantly since the year 2000, as shown in figure \ref{fig:sapphire_prices}.
+The price of sapphire substrates has decreased significantly since the year 2000, as shown in  \cref{fig:sapphire_prices}.
 This can be attributed not only to the lighting industry, but more importantly to increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
@@
 based \textit{Rubicon Technology} and Russian based \textit{Monocrystal}.
-The prevalence of wafer diameters used is shown in figure \ref{fig:wafer_size}.
+The prevalence of wafer diameters used is shown in  \cref{fig:wafer_size}.
 
@@
 Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.
-Compare table \ref{tab:sensitivity}.
+Compare table \cref{tab:sensitivity}.
 Abbreviations: resp.
@@
 
-A sensitivity analysis of the cost model has been performed using parameter variations listed in table \ref{tab:sensitivity}.
+A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
 The amount of parameter variation was chosen to encompass the identified range of the value of the parameter in different manufacturing setups of each considered year.
@@
 The upper range was chosen, because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
-The results of the analysis are shown in figure \ref{fig:sensitivity}.
+The results of the analysis are shown in  \cref{fig:sensitivity}.
 The cost model is generally more sensitive to variation of parameters at smaller wafer diameters.
@@
         \hline
-            Cleanroom Cost & \$/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & Figure \ref{fig:electricity+cleanroom_prices} \\
+            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & Figure \cref{fig:electricity+cleanroom_prices} \\
         \hline
@@
         \hline
-            Equip. Discount & \% of \$ & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I, \cite{Appleyard2001} \\
+            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I, \cite{Appleyard2001} \\
         \hline
@@
         \hline
-            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I, Figure \ref{fig:wafer_size} \\
+            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I, Figure \cref{fig:wafer_size} \\
         \hline
@@
         \hline
-            Metal Prices & \$/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
+            Metal Prices & \text{USD}/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
         \hline
-            Electricity Price & \$/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
+            Electricity Price & \text{USD}/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
         \hline
-            Saph. Subst. Price & \$ & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & Figure \ref{fig:sapphire_prices} \\
+            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & Figure \cref{fig:sapphire_prices} \\
         \hline
-            Phosphor Prices & \$/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
+            Phosphor Prices & \text{USD}/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
         \hline
@@
     \caption{Cost model sensitivity analysis parameter list.
-For the sensitivity analysis performed using these variables, see figure \ref{fig:sensitivity}.
+For the sensitivity analysis performed using these variables, see  \cref{fig:sensitivity}.
 Units for values in columns \textit{2002}-\textit{2020} are indicated in column \textit{Units}.
@@
 Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
-Sources for bottom panel: own elaboration based on the cost model described in Section \ref{sec:costmodel}.
+Sources for bottom panel: own elaboration based on the cost model described in Section \cref{sec:costmodel}.
 Sources for Inset Plot: own elaborations based on numerical approximations for die-per-wafer estimations provided in \cite{de2005investigation}, informed by expert interviews.}
@@
 
-Figure \ref{fig:costmodel_calibration} shows a comparison of the relative manufacturing cost by manufacturing step categories between the cost model and data released by the United States Department of Energy (DoE).
-Note that any comparison between the figures must consider: Our model assumes state-of-the-art equipment, while the DoE industry round table reports data from factories which use older equipment, as well as newer equipment.
+Figure \cref{fig:costmodel_calibration} shows a comparison of the relative manufacturing cost by manufacturing step categories between the cost model and data released by the United States Department of Energy (DoE).
+Note that any comparison between the s must consider: Our model assumes state-of-the-art equipment, while the DoE industry round table reports data from factories which use older equipment, as well as newer equipment.
 Our model presently considers only low-power and mid-power packages, while the DoE reports consider high-power packages also.
@@
 We note that the share of the substrate price in the DoE data is much larger than in our model.
-This can in part be explained by figure \ref{fig:sapphire_prices}, which already shows how projections overestimated the actual price for 2015-2020.
+This can in part be explained by  \cref{fig:sapphire_prices}, which already shows how projections overestimated the actual price for 2015-2020.
 We note also that the share of the epitaxy step is larger in the DoE model.
@@
 Abbreviations: Arb.
-Units - Arbitrary Units (compare section \ref{subsec:arbunit}).
+Units - Arbitrary Units (compare section \cref{subsec:arbunit}).
 Plot legends indicate the years of publication of the spectral data and phosphor mixtures used with the following designations: ?
@@
 Source (top panel): own elaboration based on spectral data.
+Arb.
+Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum.
 Source (spectral data): adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
@@
 
-The complete list of sources is organized by the different technologies described in the figure:
+The complete list of sources is organized by the different technologies described in the :
 
@@
 
-The complete list of sources is organized by the sub-efficiencies (panels) described in the figure:
+The complete list of sources is organized by the sub-efficiencies (panels) described in the :
 

### 2023-03-14 2_SSL_EES/article/article_supplementary.tex
added figure
updated figure

@@
 % smart references
-\usepackage{cleveref}
+\usepackage[capitalise, nameinlink]{cleveref}
 
@@
 \renewcommand{\tablename}{Table SI}
-\renewcommand{\theequation}{SI \arabic{equation}}
+
+% custom equation numbering
+% https://tex.stackexchange.com/a/184641/
+\newcounter{defcounter}
+\setcounter{defcounter}{0}
+\newenvironment{myequation}{%
+\addtocounter{equation}{-1}
+\refstepcounter{defcounter}
+\renewcommand\theequation{SI\thedefcounter}
+\begin{equation}}
+{\end{equation}}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \label{eqn:ler}
     K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
     \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
-\end{equation}
+\end{myequation}
 
@@
 \begin{figure}[h!]
-    \tikzfig{./figures/costmodel}
     \centering
-    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to cost model outputs.
-For description of cost variables, see definitions for equation \cref{eqn:cost_wafer}.}
+    \includegraphics[width=0.95\textwidth]{./figures/cost_model.pdf}
+    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
+For the description of cost variables, see definitions for equation \cref{eqn:cost_wafer}.}
     \label{fig:costmodel-schematic}
-    \vspace{-8pt}
 \end{figure}
@@
 It considers market trends through their effect on manufacturing parameters.
+
+Calling something here as 'upstream' and 'downstream' is kind of confusing. I suggest to replace the text highlighted by this comment with the following simplifying reformulation: 
+'It is split between the two stages of the manufacturing process: first stage combining operations at the wafer level, followed by the LED packaging stage.'
 
@@
 
-Figures \cref{fig:manuf_vtf_2012-1}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
+\cref{fig:manuf_vtf_2012-1}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
+
+    %CLASSICAL CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
+            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003.
+Abbreviations: MQW - multiple quantum well; PLLC - plastic leaded chip carrier.
+Continued on next page.}
+            \label{fig:manuf_classical_2003}
+        \end{figure}
+    \end{landscape}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \label{eqn:cost_sum}
     C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{align}
-	C_w &= \sum_i C_i \\
-	C_p &= \sum_j C_j
-\end{align}
+\begin{myequation}
+        C_w = \sum_i C_i \\
+	C_p = \sum_j C_j
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \label{eqn:cost_wafer}
     C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \label{eqn:dpw}
 	D=\frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
-\end{equation}
+\end{myequation}
 
@@
 
-Results are plotted in Figure \cref{fig:dpw}, showing the increase in die per wafer (DPW).
-Equation \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
+Results are plotted in \cref{fig:dpw}, showing the increase in die per wafer (DPW). \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
 
-\begin{equation}
+\begin{myequation}
 \label{eqn:cost_wafer_full}
@@
 \end{split}
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \label{eqn:cost_die}
     C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((e*p) + l + m + d + o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x p_x \bigg\}
-\end{equation}
+\end{myequation}
 
 where $c_j$ is $\text{throughput}^{-1}$. The total cost is thus
-\begin{equation}
+\begin{myequation}
 \label{eqn:cost_total}
@@
 
-\begin{equation}
-\begin{split}
+\begin{myequation}
 \label{eqn:C_2}
-    C_{Y_1} &= \frac{C_1}{Y_1} \\
-    C_{Y_2} &= \frac{C_1 + C_2}{Y_1 Y_2} - C_{Y_1} = \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} = \frac{1}{Y_1 Y_2} \bigg ( C_1 (1-Y_2) +C_2 \bigg)\\
-    C_{Y_i} &= \frac{ \sum_{x \leq i} C_x }{ \prod_{x \leq i} Y_x } - \frac{ \sum_{x<i} C_x }{ \prod_{x<i} Y_x }
-\end{split}
-\end{equation}
+    C_{Y_1} = \frac{C_1}{Y_1}
+\end{myequation}
+\begin{myequation}
+    C_{Y_2} = \frac{C_1 + C_2}{Y_1 Y_2} - C_{Y_1} = \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} = \frac{1}{Y_1 Y_2} \bigg ( C_1 (1-Y_2) +C_2 \bigg)
+\end{myequation}
+\begin{myequation}
+    C_{Y_i} = \frac{ \sum_{x \leq i} C_x }{ \prod_{x \leq i} Y_x } - \frac{ \sum_{x<i} C_x }{ \prod_{x<i} Y_x }
+\end{myequation}
 
@@
 
-\begin{align}
+\begin{myequation}
 \label{eqn:C_2^2}
-    C_{Y_2}^{(2 \times)} &= \bigg( \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} \bigg) + \bigg( \frac{C_1 + 2 C_2}{Y_1 Y_2^2} - \frac{C_1 + C_2}{Y_1 Y_2}     \bigg) \\
-    &= \frac{1}{Y_1 Y_2^2} \bigg( C_1 (1-Y_2^2) +2C_2 \bigg)
-\end{align}
+    C_{Y_2}^{(2 \times)} = \bigg( \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} \bigg) + \bigg( \frac{C_1 + 2 C_2}{Y_1 Y_2^2} - \frac{C_1 + C_2}{Y_1 Y_2}     \bigg)
+\end{myequation}
+\begin{myequation}
+    = \frac{1}{Y_1 Y_2^2} \bigg( C_1 (1-Y_2^2) +2C_2 \bigg)
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
     C_{Y_2}^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \label{eqn:yielded_cost}
     C_{Y_{i>1}}^{(n \times)} = \frac{1}{Y_i^{n-1} \prod_{x \leq i} Y_x} \bigg( nC_i + \sum_{x < i} C_x (1-Y_i^n) \bigg)
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 	f_\text{single} = \frac{i+s}{y}
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{align}
+\begin{myequation}
 \label{eqn:series}
-    f_{\times 2} &=  \frac{\frac{i+s}{y}+s}{y} \\
-    f_{\times n} &= \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
-\end{align}
+    f_{\times 2} =  \frac{\frac{i+s}{y}+s}{y} \\
+    f_{\times n} = \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
     f_\text{total} = \frac{\sum_i^n s_i}{\prod_i^ny_i} \neq \sum_i^n f_i
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 C=ab+cd
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \Delta C = C(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta C_i
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \text{d}C(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial C }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial C }{\partial x_i}  \Delta x_i
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \Delta C_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial C }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
 \label{eqn:integral_1}
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 C(\vec{r}) = C(r_1,r_2, \dots) = \sum_i C_i
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 C_i(\vec{r}) = C_i^0 \prod_w g_{iw}(r_w)
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \Delta C_x = \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
-\end{equation}
+\end{myequation}
 
@@
 
-\begin{equation}
+\begin{myequation}
 \Delta C_z (t_1,t_2) \approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
-\end{equation}
+\end{myequation}
 
@@
 
-Figure \cref{fig:costmodel_calibration} shows a comparison of the relative manufacturing cost by manufacturing step categories between the cost model and data released by the United States Department of Energy (DoE).
+\cref{fig:costmodel_calibration} shows a comparison of the relative manufacturing cost by manufacturing step categories between the cost model and data released by the United States Department of Energy (DoE).
 Note that any comparison between the s must consider: Our model assumes state-of-the-art equipment, while the DoE industry round table reports data from factories which use older equipment, as well as newer equipment.

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
updated figure, table and equation naming
updated bibliography

@@
 
-\renewcommand{\figurename}{Figure SI}
-\renewcommand{\tablename}{Table SI}
+\renewcommand{\thefigure}{SI\arabic{figure}}
+\renewcommand{\thetable}{SI\arabic{table}}
 
@@
 
-\section{Additional Performance Metrics}
+\section{Luminous Efficacy Metrics}
+
+Figure 1 in the main publication uses luminous efficacy of radiation as the primary metric to describe progress in lighting technologies.
+Care must be taken not to confuse this metric with luminous efficacy of source, which is used in \ref{fig:phosphor_spectrum}.
 
@@
 
+This metric describes the match of a light-emitting diode package spectrum to the human visual system.
 Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
 A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
-The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source with the wavelength dependent human sensitivity to light \cite{cie-term-effrad}
+The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
 
@@
 This metric can be computed from spectral data alone and does not require additional spectral normalization.
-It enables straightforward comparison between the performance of different downconversion phosphors, as shown in  \cref{fig:se_efficacy} in chapter \cref{ch:results}.
-Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation, as the product with the luminosity function of  \cref{fig:tristimulus} shows.
-Care must be taken not to confuse this efficacy metric with efficacy of source, which depends not only on the spectral power distribution of a light source.
-
-This metric describes the match of a light-emitting diode package spectrum to the human visual system.
+It enables straightforward comparison between the performance of different downconversion phosphors, as shown in the top panel of \cref{fig:phosphor_spectrum}.
+Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation.
+Care must be taken not to confuse this efficacy metric with the \textit{efficacy of source} described in the following subsection.
 
@@
 
-In contrast to the efficacy \textit{of radiation} $K$, the luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power \cite{cie-term-effsrc}
+The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power \cite{cie-term-effsrc}
 
@@
 
-Unlike the luminous efficacy of radiation, this metric cannot be extracted from the spectrum of a light source.
-It is often cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
-Care must be taken not to confuse this efficacy metric with efficacy of radiation, which depends only on the spectral power distribution of a light source.
-As it captures overall device efficacy, it depends on a large number of device properties and parameters.
-This makes attribution to single changes in device design or manufacturing difficult.
+This metric is often cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
+Care must be taken not to confuse this efficacy metric with the \textit{luminous efficacy of radiation}, which depends only on the spectral characteristics of a light source.
+As the luminous efficacy of a light source $\eta$ captures the overall device efficacy, it depends on a large number of other device properties and parameters.
+This makes attribution of changes in this metric to individual changes in device design or manufacturing difficult.
+For this reason, we do not use this metric in our study.
 
@@
     \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
-For the description of cost variables, see definitions for equation \cref{eqn:cost_wafer}.}
+For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.}
     \label{fig:costmodel-schematic}
@@
 
-Calling something here as 'upstream' and 'downstream' is kind of confusing. I suggest to replace the text highlighted by this comment with the following simplifying reformulation: 
-'It is split between the two stages of the manufacturing process: first stage combining operations at the wafer level, followed by the LED packaging stage.'
-
-A schematic diagram of the cost model developed is presented in  \cref{fig:costmodel-schematic}.
+A schematic diagram of the cost model is presented in \cref{fig:costmodel-schematic}.
 The cost model is process step-based.
-It is split between the upstream part of the manufacturing process and the downstream part of the manufacturing process.
-The former is performed at the wafer level.
-The latter is performed at the package level.
-The model takes as inputs both parameters specific to manufacturing process steps (\textit{"manufacturing step parameters"}) and parameters affecting all manufacturing steps (\textit{"global parameters"}).
+It is split between the two stages of the manufacturing process: the first stage combining operations at the wafer level, followed by the LED packaging stage.
+The model takes as inputs parameters specific to individual manufacturing process steps (\textit{"manufacturing step parameters"}) and parameters affecting all manufacturing steps (\textit{"global parameters"}).
 The cost for each process step is then computed.
@@
 
-\cref{fig:manuf_vtf_2012-1}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
+\cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
 
@@
             \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
-            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003.
-Abbreviations: MQW - multiple quantum well; PLLC - plastic leaded chip carrier.
-Continued on next page.}
+            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003. Abbreviations: MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
             \label{fig:manuf_classical_2003}
@@
             \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
-            \caption{(2/2) Continued from previous page. Abbreviations: Encaps. - encapsulation.}
+            \caption{(2/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+Continued from previous page.
+Abbreviations: Encaps.
+- encapsulation.}
             \label{fig:manuf_csp_2020-2}
@@
     \caption{Number of die per wafer for the different wafer sized used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW to $d(2020)=200$mm$\rightarrow26,838$DPW.
-Source: own elucidations, based on equation \cref{eqn:dpw}.}
+Source: own elucidations, based on  \cref{eqn:dpw}.}
     \label{fig:dpw}
@@
 
-which can be compared to equation \cref{eqn:C_2} to find a more general form:
+which can be compared to  \cref{eqn:C_2} to find a more general form:
 
@@
 
-where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in equation \cref{eqn:series}.
+where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in  \cref{eqn:series}.
 
@@
 
-Using logarithmic differentiation, the integral from equation \cref{eqn:integral_1} can be rewritten as
+Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
 
@@
 Due to time constraints, this approach could not be used to determine the effect of spillovers on manufacturing cost.
-Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of equation \cref{eqn:cost_sum}, could this method.
+Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of  \cref{eqn:cost_sum}, could this method.
 
@@
 
-The complete list of sources is organized by the different technologies described in the :
+The list of sources for Figure 1 in the main article is organized by different technologies:
 
@@
     \hline
-    Max. efficacy & \cite{Murphy2012} (in contrast to the original value given in \cite{azevedo2009transition}) \\
+    Max. efficacy & \cite{Murphy2012} \\
     \hline
@@
 
-The complete list of sources is organized by different sources:
+The sources for Figure 2 in the main article are grouped by the provider of data:
 
@@
     \hline
-    \textbf{Source of References} & \textbf{References} \\
+    \textbf{Provider of References} & \textbf{References} \\
     \hline
@@
 
-The complete list of sources is organized by different sources:
+The sources for Figure 3 in the main article are grouped by their type:
 
@@
     \hline
-    \textbf{Source of References} & \textbf{References} \\
+    \textbf{Type of Source} & \textbf{References} \\
     \hline
@@
 
-The complete list of sources is organized by the sub-efficiencies (panels) described in the :
+The sources for Figure 4 in the main article are grouped by sub-efficiencies represented on different panels of the figure:
 
@@
     \hline
-    \textbf{Technology} & \textbf{References} \\
+    \textbf{Panel (Sub-Eff.)} & \textbf{References} \\
     \hline

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
udpated tables

@@
 
-\section{Identified LED Innovations}
+\section{Non-Phosphor-Related LED Innovations}
+
+This section summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies, as discussed in section 5.1 and shown in Figures 4-6 in the main article.
+Note that phosphor-related LED innovations affecting consumer experience metrics, along with spectral efficiency, are presented separately in Table 1 in the main publication and discussed in detail in \cref{sec:innovation_phosphor}.
 
@@
         \hline
-            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+            \textit{Year} & \textit{Innovation} & \textit{Area of Improvement} & \textit{Spillover} & \textit{Source} \\
         \hline
-            1983 & Indium Tin Oxide \newline Current Spreading Layer & Contact Resistance & Yes (T,S) & \cite{margalith1999indium}\\
+            1999 & Indium Tin Oxide \newline Current spreading layer & Contact resistance & Yes & \cite{margalith1999indium}\\
         \hline
-            1998 & Digited Electrodes & Contact Resistance & Weak & \cite{steigerwald2001electrode} \\
+            1998 & Digitated electrodes & Contact resistance & No & \cite{steigerwald2001electrode} \\
         \hline
-            ongoing & Epitaxy Improvements \newline and better Doping & Polarization and \newline Bulk Resistance & No & I \\
+            ongoing & Epitaxy improvements \newline and better doping & polarization and \newline bulk resistance & No & I \\
         \hline
-        $\sim$2000 & Silver p-Contacts & Contact Resistance & No & \cite{kondoh2001nitride} \\
+            1998 & Silver p-Contacts & Contact resistance & No & \cite{kondoh2001nitride} \\
         \hline
-        \multicolumn{5}{c}{Light Extraction Efficiency (LEE) $\eta_{LE}$} \\
+        \multicolumn{5}{c}{Light extraction efficiency (LEE) $\eta_{LE}$} \\
         \hline
-            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
         \hline
-            $<$ 2003 & Optimization for  \newline Cavity Effects & Other & No & I, \cite{Shen2003} \\
+            $<$ 2003 & Optimization for \newline cavity effects & Other & No & I, \cite{Shen2003} \\
         \hline
-            (1973) 1993 & "Chip \newline Randomization" & Total Reflection \newline Absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
+            1993 & Chip surface randomization & Total reflection \newline Absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
         \hline
-            1993 & Thin-Film \newline Chip Architecture & Absorption & No & I, \cite{Schnitzer1993} \\
+            1993 & Thin-Film \newline Chip architecture & Absorption & No & I, \cite{Schnitzer1993} \\
         \hline
-            $<2001$ & Patterned Sapphire \newline Substrate  & Total Reflection \newline Absorption & Yes (T,S) & I, \cite{Tadatomo2001} \\
+            1996 & Patterned sapphire \newline Substrate  & Total reflection \newline Absorption & Yes & I, \cite{Tadatomo2001} \\
         \hline
-            ongoing & Chip Design for \newline high LEE & Total Reflection \newline Absorption & No & I, \cite{Haerle2004} \\
+            ongoing & Chip design for \newline high LEE & total reflection \newline Absorption & No & I, \cite{Haerle2004} \\
         \hline
-            $\sim$2000 & Silver p-Contacts & Absorption & No & I, \cite{kondoh2001nitride} \\
+            $\sim$2000 & Silver p-contacts & Absorption & No & I, \cite{kondoh2001nitride} \\
         \hline
-        \multicolumn{5}{c}{Internal Quantum Efficiency (IQE) $\eta_{IQ}$} \\
+        \multicolumn{5}{c}{Internal quantum efficiency (IQE) $\eta_{IQ}$} \\
         \hline
-            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
         \hline
-            1994 & DH & $>$ Radiative Recombination \newline Probability & No & \cite{Nakamura1994} \\
+            1994 & DH & Higher radiative recombination \newline Probability & No & \cite{Nakamura1994} \\
         \hline
-            1996 & MQW & $>$ Radiative Recombination \newline Probability & No & \cite{Koike1996} \\
+            1996 & multiple quantum well & Higher radiative recombination \newline Probability & No & \cite{Koike1996} \\
         \hline
-            ongoing & Active Region \newline Doping & Both & No & I, \cite{schubert2018light} \\
+            ongoing & Active region \newline Doping & Both & No & I, \cite{schubert2018light} \\
         \hline
@@
         \hline
-            ongoing & Chip Architecture \newline Improvements & Both & No & I \\
+            ongoing & Chip architecture \newline Improvements & Both & No & I \\
         \hline
-        \multicolumn{5}{c}{(Phosphor) Conversion Efficiency (CE) $\eta_{C}$} \\
+        \multicolumn{5}{c}{Light Conversion efficiency (CE) $\eta_{C}$} \\
         \hline
-            \textit{Year} & \textit{Breakthrough} & \textit{Affects} & \textit{S/O} & \textit{Source} \\
+            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
         \hline
-            ongoing & Lower Current Density & Current Density & No & I \\
+            ongoing & Lower current density & Current density & No & I \\
         \hline
-            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{bhardwaj2016progress} \\
+            ongoing & Epitaxy \newline Improvements & Charge distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
         \hline
-            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{Wildeson2017} \\
+            ongoing & Chip architecture \newline improvements  & Charge distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
         \hline
-            $<$ 2017 & Defect Getting \newline Underlayer & Defect Density & No & I, \cite{haller2017burying} \\
+            $<$ 2017 & Defect getting \newline underlayer & Defect Density & No & I, \cite{haller2017burying} \\
         \hline
         \end{tabularx}
-    \caption{Technology breakthroughs and spillovers related to a subset of the sub-efficiencies considered.
-For a list of innovations in spectral efficiency, compare Table 1 in the main article.
-Listed are improvements in forward voltage efficiency, light extraction efficiency, internal quantum efficiency and conversion efficiency.
-No  Year column indicates first identified use in solid-state lighting.
-S/O column indicates if the breakthrough has been identified as a spillover.
-Source column additionally indicates if the spillover has been identified as part of an expert interview (I).
-Spillover sources are indicated in brackets and follow the taxonomy developed in \cite{Stephan_2021}: Other technology (T), other sector (S), other scientific discipline (D).
-}
+    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.
+The table includes innovations and technology improvements affecting forward voltage efficiency, light extraction efficiency, internal quantum efficiency, and light conversion efficiency.
+For a list of innovations affecting spectral efficiency, see Table 1 in the main article.
+Year column indicates the first instance of application of corresponding invention in white LEDs.
+Spillover column indicates if the innovation involved technology spillovers.
+Source column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.
+Source column additionally indicates if the spillover has been identified as part of an expert interview (I).}
     \label{tab:innovations}
@@
 
-\newpage
-\section{Phosphors}
+\section{Phosphor-Related LED Innovations}
+\label{sec:innovation_phosphor}
 
-\subsection{History of Identifies Phosphor Inventions}
+This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiences and consumer experience metrics and discussed in \cref{subsec:metrics}, Table 1 and Figures 5-7 in the main article.
+
+\subsection{History of Identified Phosphor-Related LED Innovations"}
 
@@
 
-Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 8 in the main text).
+Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 7 in the main article).
 An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
@@
 
-The most recent red narrow-band phosphor breakthrough included in Table 3 and Figure 8 in the main text, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
-The first patent application for this phosphor was filed in 2016.
+The most recent red narrow-band phosphor innovation included in Table 1 and Figure 7 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
+The first patent application for this phosphor was filed in 2016 .
 The SALON phosphor is a derivate of the SLA phosphor.
@@
 It is based on a potassium fluorosilicate (PFS) compound activated with manganese ions.
-Its luminescence was first recorded by Paulusz at GE in 1972 \cite{paulusz1973efficient}.
+Its luminescence was first recorded by Adrian Paulusz at GE in 1972 \cite{paulusz1973efficient}.
 In the early 2000s, while searching for potential new LED phosphor materials for GE’s lighting business at GE Lumination, Emil Radkov rediscovered Paulusz’s findings in the literature.
@@
 This work, supported by public funding from the U.S.
-Department of Energy (DOE) Solid-State Lighting program , resulted in a series of critical improvements in the PFS phosphor properties \cite{Setlur2010}\cite{lyons2012color}, eventually enabling its commercialization under the “TriGain” brand in 2015 \cite{trigain_spectrum}\cite{setlur2015trigain}\cite{Murphy2015}.
+Department of Energy (DOE) Solid-State Lighting program \cite{doesslprogram}, resulted in a series of critical improvements in the PFS phosphor properties \cite{Setlur2010}\cite{lyons2012color}, eventually enabling its commercialization under the “TriGain” brand in 2015 \cite{trigain_spectrum}\cite{setlur2015trigain}\cite{Murphy2015}.
 
@@
 Department of Energy (DOE)-funded “A Revolution in Lighting“ project at Sandia National Laboratory \cite{simmonsfinal}.
-This concept was successfully demonstrated by Sandia researchers on a commercial LED in 2003 \cite{shea_rohwer_development_2004}\cite{noauthor_sandia_nodate} and was swiftly taken up and advanced further by a group in Taiwan \cite{chen_white-light_2005}\cite{hsueh-shih_chen_ingan-cdse-znse_2006} The first commercial application of QDs in an LED lamp was brought about by a collaboration between an MIT-born startup QD Vision and the U.S.-based luminaire manufacturer Nexxus Lighting in 2009 \cite{ledprof_nexxusqd}, \cite{bourzac2013quantum}.
+This concept was successfully demonstrated by Sandia researchers on a commercial LED in 2003 \cite{shea_rohwer_development_2004}\cite{noauthor_sandia_nodate} and was swiftly taken up and advanced further by a group in Taiwan \cite{Chen_2005}\cite{Hsueh_Shih_Chen_2006} The first commercial application of QDs in an LED lamp was brought about by a collaboration between an MIT-born startup QD Vision and the U.S.-based luminaire manufacturer Nexxus Lighting in 2009 \cite{ledprof_nexxusqd}, \cite{bourzac2013quantum}.
 However, rapid advances in the spectral and conversion performance of down-conversion phosphors and high manufacturing cost of quantum dots resulted in the discontinuation of this product.
@@
 
-\subsection{Spectral Data of Identified LED Phosphor Innovations}
+\subsection{Spectral Data for Identified LED Phosphor Innovations}
 
@@
     \includegraphics[width=0.95\textwidth]{2_SSL_EES/article/figures/phosphor_spectrum-comparison.pdf}
-	\caption{Two key metrics associated with red phosphor innovations in light-emitting diodes and their associated spectral data.
-Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Main Article Figure 7 (Historical Improvements in Consumer Experience Metrics of Phosphor-Converted White LED).
-Years indicate the earliest identified representative white LED products with published spectral data that used phosphor innovations listed in Main Article Table 2.
+	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 7 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year.
+Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 7 in the main article.
 The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow.
-Metrics were calculated from spectral data shown in the two bottom rows using the \texttt{colour-science} package for Python \cite{colour-science_software}.
+Metrics were calculated from spectral data shown in the two bottom panels using the \texttt{colour-science} package for Python \cite{colour-science_software}.
 Bottom two rows of panels: Corresponding spectral data.
-The luminosity function, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
-Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable.
-This is because photons of the corresponding energy are lost to the human observer and count towards the spectral loss channel.
+The luminosity function \cite{cie-term-lumeff}, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
+Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel.
 Abbreviations: Arb.
 Units - Arbitrary Units (compare section \cref{subsec:arbunit}).
-Plot legends indicate the years of publication of the spectral data and phosphor mixtures used with the following designations: ?
--  other parts of phosphor mixture not disclosed, $^*$ - \ch{CaSrS:Eu^{2+}}, $^\star$ - $\beta-$SiAlON:Eu$^{2+}$, $^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, $^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
-Full chemical formulas for displayed phosphors: \ch{Sr[Li2Al2O2N2]:Eu^{2+}}(SALON), \ch{Sr[LiAl_3N_4]:Eu^{2+}} (SLA), \ch{K2SiF6:Mn^{4+}} (PFS), \ch{Sr2Si5N8:Eu^{2+}} (258), Quantum Dots (QD).
+Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: O$^\ddagger$ -  other parts of phosphor mixture not disclosed, G$^*$ - \ch{CaSrS:Eu^{2+}}, Y$^\star$ - $\beta-$SiAlON:Eu$^{2+}$, $^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, $^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
 Source (top panel): own elaboration based on spectral data.
@@
 Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum.
-Source (spectral data): adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
+Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
 \label{fig:phosphor_spectrum}
@@
 
-\section{Anonymized List of Interviewees}
+\section{List of Interviewees}
 
@@
     \hline
-        \textbf{\#} & \textbf{sector} & \textbf{role} & \textbf{country} & \textbf{expertise} \\ \hline
-        1 & academia & sr. researcher & UK & epitaxy \\ \hline
-        2 & industry & consultant, form. sr. researcher & USA & device architecture \\ \hline
-        3 & industry & consultant, form. head of R\&D & Germany & epitaxy \\ \hline
-        4 & academia & professor & Austria & phosphors \\ \hline
-        5 & industry & consultant, form. head of R\&D & USA & device architecture \\ \hline
-        6 & consulting & consultant, form. sr. technical advisor & USA & device architecture \\ \hline
-        7 & academia & professor & Germany & phosphors \\ \hline
-        8 & government & R\&D manager & USA & device architecture \\ \hline
-        9 & consulting & consultant & USA & device applications \\ \hline
-        10 & academia & professor & France & device physics \\ \hline
-        11 & industry & sr scientist, form. head of R\&D & USA & device architecture \\ \hline
-        12 & industry & principal scientist & Germany & phosphors \\ \hline
-        13 & industry & form. head of R\&D & USA & phosphors \\ \hline
+        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
+        1 & Academia & Sr. researcher & UK & Epitaxy \\ \hline
+        2 & Industry & Consultant, former sr. researcher & USA & Device architecture \\ \hline
+        3 & Industry & Consultant, former head of R\&D & Germany & Epitaxy \\ \hline
+        4 & Academia & Professor & Austria & Phosphors \\ \hline
+        5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
+        6 & consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
+        7 & Academia & Professor & Germany & Phosphors \\ \hline
+        8 & government & R\&D manager & USA & Device architecture \\ \hline
+        9 & consulting & Consultant & USA & Device applications \\ \hline
+        10 & Academia & Professor & France & Device physics \\ \hline
+        11 & Industry & Sr scientist, former head of R\&D & USA & Device architecture \\ \hline
+        12 & Industry & Principal scientist & Germany & Phosphors \\ \hline
+        13 & Industry & Former head of R\&D & USA & Phosphors \\ \hline
     \end{tabular}
     \label{tab:interviews}
-    \caption{List of domain experts with whom semi-structured interviews were conducted. \\ Abbreviations: form. - former, sr. - senior}
+    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
 \end{table}

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
various edits

@@
 
-\subsection{Examples of Input Data Considered}
+\subsection{Example of Input Data Considered: Sapphire Wafers}
 
-\subsubsection{Sapphire Wafers}
-
-Sapphire wafers form the substrate on which all other layers of the light-emitting diode wafer are grown.
+Sapphire wafers form the substrate on which all other layers of the light-emitting diode are grown.
 Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture.
 In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process.
-Wafers can be either unpatterned or patterned, where the latter has become commonplace in 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
-The price of sapphire substrates has decreased significantly since the year 2000, as shown in  \cref{fig:sapphire_prices}.
-This can be attributed not only to the lighting industry, but more importantly to increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
+Wafers can be either unpatterned or patterned, where the latter has become commonplace by 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
+The price of sapphire substrates has decreased significantly since the year 2000, as shown in the bottom panel of \cref{fig:wafers}.
+This can be attributed not only to the lighting industry, but more importantly to increased supply as a result of increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
 Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing.
-The market is dominated by U.S.
-based \textit{Rubicon Technology} and Russian based \textit{Monocrystal}.
-The prevalence of wafer diameters used is shown in  \cref{fig:wafer_size}.
+The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}.
+he evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.
 
 \begin{figure}[h!]
+    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
     \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
-    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
 	\caption{Top: Historical data for sapphire substrate prices of different surface properties and diameters.
@@
 Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
-	\label{fig:wafer_size}
+	\label{fig:wafers}
 \end{figure}
@@
         \hline
-            1993 & Chip surface randomization & Total reflection \newline Absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
+            1993 & Chip surface randomization & Total reflection and absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
         \hline
-            1993 & Thin-Film \newline Chip architecture & Absorption & No & I, \cite{Schnitzer1993} \\
+            1993 & Thin-film chip architecture & Absorption & No & I, \cite{Schnitzer1993} \\
         \hline
-            1996 & Patterned sapphire \newline Substrate  & Total reflection \newline Absorption & Yes & I, \cite{Tadatomo2001} \\
+            1996 & Patterned sapphire substrate  & Total reflection and absorption & Yes & I, \cite{Tadatomo2001} \\
         \hline
-            ongoing & Chip design for \newline high LEE & total reflection \newline Absorption & No & I, \cite{Haerle2004} \\
+            ongoing & Chip design for high LEE & Total reflection and absorption & No & I, \cite{Haerle2004} \\
         \hline
@@
         \hline
-            1994 & DH & Higher radiative recombination \newline Probability & No & \cite{Nakamura1994} \\
+            1994 & DH & Higher radiative recombination probability & No & \cite{Nakamura1994} \\
         \hline
-            1996 & multiple quantum well & Higher radiative recombination \newline Probability & No & \cite{Koike1996} \\
+            1996 & multiple quantum well & Higher radiative recombination probability & No & \cite{Koike1996} \\
         \hline

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
more edits

@@
 The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level.
-While the upstream steps in the manufacturing flow are carried out at the wafer level, the packaging steps are typically carried out at the die/package level.
 The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
@@
 \begin{align*}
-    P_S &\dots \text{sapphire substrate price} \\
-    C_w &\dots \text{wafer processing cost} \\
+    P_S &\dots \text{sapphire substrate price per die} \\
+    C_w &\dots \text{wafer processing cost per die} \\
     C_p &\dots \text{die processing cost}
@@
 \begin{myequation}
-        C_w = \sum_i C_i \\
+        C_w = \sum_i C_i
+\end{myequation}
+\begin{myequation}
 	C_p = \sum_j C_j
@@
 \label{eqn:cost_wafer}
-    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x \c_x \bigg\}
 \end{myequation}
 
-where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the indices $x,y$ run over all materials.
+where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
 
 \begin{align*}
-        D &\dots \text{functional die per wafer (tested successfully)} \label{def:cost_wafer_first} \\
+        D &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
         y &\dots \text{process step yield} \\
-        u &\dots \text{equipment utilization (in relation to theoretical equipment capacity)} \\
+        u &\dots \text{equipment utilization (relative to theoretical equipment capacity)} \\
         p &\dots \text{power consumption} \\
-        e &\dots \text{hourly electricity cost} \\
-        m &\dots \text{hourly maintenance cost} \\
-        d &\dots \text{hourly depreciation cost}\\
-        l &\dots \text{hourly labour cost} \\
-        o &\dots \text{hourly overhead cost} \\
+        c_e &\dots \text{hourly electricity cost} \\
+        c_m &\dots \text{hourly maintenance cost} \\
+        c_d &\dots \text{hourly depreciation cost}\\
+        c_l &\dots \text{hourly labour cost} \\
+        c_o &\dots \text{hourly overhead cost} \\
         t_i &\dots \text{time per run} \\
@@
         v_x &\dots \text{volume of material $x$ per wafer} \\
-        p_x &\dots \text{price of material $x$ per volume}\\
+        c_x &\dots \text{cost of material $x$ per volume}\\
 \end{align*}
@@
 \label{eqn:dpw}
-	D=\frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
+	DPW = \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
 \end{myequation}
@@
     \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
-    \caption{Number of die per wafer for the different wafer sized used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW to $d(2020)=200$mm$\rightarrow26,838$DPW.
-Source: own elucidations, based on  \cref{eqn:dpw}.}
+    \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2002)=50$mm$\rightarrow5370$DPW, $d(2020)=100$mm$, \rightarrow26,838$DPW.
+Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, according to \cref{fig:wafers}.}
     \label{fig:dpw}
@@
 
-Results are plotted in \cref{fig:dpw}, showing the increase in die per wafer (DPW). \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
-
+Results are plotted in \cref{fig:dpw}, showing the increase in die per wafer (DPW) over time. \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
 \begin{myequation}
@@
     C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
-    &  \frac{1}{y_i}  \bigg\{ \bigg((e*p) + l + m + d +o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x p_x \bigg\}
+    &  \frac{1}{y_i}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x \c_x \bigg\}
 \end{split}
 \end{myequation}
-
 and the cost of a manufacturing step $C_j$ in the packaging category
-
 \begin{myequation}
 \label{eqn:cost_die}
-    C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((e*p) + l + m + d + o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x p_x \bigg\}
+    C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x c_x \bigg\}
 \end{myequation}
-
 where $c_j$ is $\text{throughput}^{-1}$. The total cost is thus
@@
 \end{myequation}
-
 Note that in keeping with the categorization introduced by the United States Department of Energy (cf.
@@
 While there are different mathematical approaches to including yield, we follow the definition in \cite{becker2001use}.
-We write for the yielded cost $C_{Y_i}$ of a step $i$
+We write for the yielded cost $C_{Y_i}$ of a step $i$ with associated cost (before considering yield) $C_1$ and yield $Y_1$:
 
@@
 
-It can be shown by induction that the general term for a step $i>1$ repeated $n$ times gives an equation of the form:
+It can be shown by induction that the general form of a term for a step $i>1$ repeated $n$ times can be expressed as:
 
@@
 It uses what in the literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}.
-Using this approach, the yielded cost of a single process step $f$ is described by
+In this approach, the yielded cost of a single process step $f$ is described as
 
@@
 \begin{myequation}
+    f_{\times 2} =  \frac{\frac{i+s}{y}+s}{y}
+\end{myequation}
+\begin{myequation}
+    f_{\times n} = \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
 \label{eqn:series}
-    f_{\times 2} =  \frac{\frac{i+s}{y}+s}{y} \\
-    f_{\times n} = \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
 \end{myequation}
@@
 where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in  \cref{eqn:series}.
-
-\subsection{Computation of the Contribution of Variables}
-
-To quantify the drivers of cost reductions across all process steps, one would need to identify the magnitude of contribution to cost reductions made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
-Mathematically, given a function $C$ which describes the cost associated with manufacturing a single unit at time $t$,
-
-\begin{myequation}
-C=ab+cd
-\end{myequation}
-
-and manufacturing variables $a,b,c,d$, we are looking for the contribution $\Delta C_{a}$ made by a single variable $a$ to the total change in the cost function $\Delta C$ between points $t_0,t_1$ such, that
-
-\begin{myequation}
-\Delta C = C(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta C_i
-\end{myequation}
-
-The infinitesimal contribution to the total cost by the infinitesimal change in a cost model variable is defined through the total differential of the cost function \cref{eqn:cost_sum} as
-
-\begin{myequation}
-\text{d}C(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial C }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial C }{\partial x_i}  \Delta x_i
-\end{myequation}
-
-where $x_i$ is an arbitrary cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to total cost $C$ over the period $t_0 < t < t_1 $ is then
-
-\begin{myequation}
-\Delta C_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial C }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
-\label{eqn:integral_1}
-\end{myequation}
-
-However, data on the cost model variables is not available in continuous time.
-Disaggregating the contribution of single variables to the total cost reduction is thus not straightforward in our model.
-This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}.
-The following discussion follows an approach developed by Kavlak et al.
-\cite{kavlak2018evaluating}.
-For the detailed derivation, we refer to this publication.
-
-The cost function $C$ as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
-
-\begin{myequation}
-C(\vec{r}) = C(r_1,r_2, \dots) = \sum_i C_i
-\end{myequation}
-
-where
-
-\begin{myequation}
-C_i(\vec{r}) = C_i^0 \prod_w g_{iw}(r_w)
-\end{myequation}
-
-Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
-
-\begin{myequation}
-\Delta C_x = \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
-\end{myequation}
-
-where for $C(t)$ a constant $C(t) \approx \tilde{C} $ can be chosen such that $\Delta C_{x_i} = \Delta C$.
-In practice, this constant value can be approximated as $\tilde{C} \approx \frac{2}{3} C_i^\text{geo} + \frac{1}{3} \overline{C_i}$.
-The contribution of a single cost model variable $r_z$ can then be written as
-
-\begin{myequation}
-\Delta C_z (t_1,t_2) \approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
-\end{myequation}
-
-This approach was used to determine the effect of spillovers on device performance, displayed in  \cref{fig:breakthroughs_efficiency}.
-Due to time constraints, this approach could not be used to determine the effect of spillovers on manufacturing cost.
-Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of  \cref{eqn:cost_sum}, could this method.
 
@@
 
-\subsection{Comparison with Reported Industry Data}
+\subsection{Comparison with Reported Industry Data and DOE Projections}
 
-We note differences between our model and costs reported to the US Department of Energy (DoE) by industry.
-For instance, the share of the epitaxy step is larger in the DoE model.
-This can in part be explained by our model relying on state-of-the-art equipment, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
-The packaging part of the manufacturing process is increasing in importance, which has been confirmed by researchers and industry reports on wafer-level packaging.
+We conducted a comparison of the LED manufacturing cost structure produced by our cost model with previously reported US DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions (see \cref{fig:costmodel_calibration}).
+We note some differences between the results of our model and the cost structure reported or projected by the DOE.
+For instance, the share of the epitaxy step is consistently larger in the DOE data.
+This can in part be explained by our model relying on state-of-the-art equipment at a virtual US-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+In addition, the manufacturing lines of the majority of manufacturers is located in Asia.
+We also note that the share of the substrate price in the DOE data is much larger than in our model, which can in part be explained by the overestimation of the actual price of sapphire wafers in earlier projections for 2015-2020.
+Finally, the relative importance of the packaging part of the manufacturing process is very similar in our model and in the DOE results in 2012.
+However, in the DOE projections it significantly decreases by 2020, while in our model it retains and even increases its share of the total cost.
+This trend has been independently confirmed by researchers and industry reports on wafer-level packaging \cite{Lee2011WPL}\cite{Xie2013}\cite{ledsmag2017WLP}, showing better performance of our cost model compared to earlier DOE model projections
 
@@
 The packaging part of the manufacturing process is increasing in importance, which has been confirmed by researchers and industry reports on WLP\cite{Lee2011WPL}\cite{Xie2013}\cite{ledsmag2017WLP}.
+
+\section{Computation of Contribution of Individual Variables}
+
+To quantify the drivers of cost reductions across all process steps, one would need to identify the magnitude of contribution to cost reductions made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
+Mathematically, given a function $C$ which describes the cost associated with manufacturing a single unit at time $t$,
+
+\begin{myequation}
+C=ab+cd
+\end{myequation}
+
+and manufacturing variables $a,b,c,d$, we are looking for the contribution $\Delta C_{a}$ made by a single variable $a$ to the total change in the cost function $\Delta C$ between points $t_0,t_1$ such, that
+
+\begin{myequation}
+\Delta C = C(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta C_i
+\end{myequation}
+
+The infinitesimal contribution to the total cost by the infinitesimal change in a cost model variable is defined through the total differential of the cost function \cref{eqn:cost_sum} as
+
+\begin{myequation}
+\text{d}C(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial C }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial C }{\partial x_i}  \Delta x_i
+\end{myequation}
+
+where $x_i$ is an arbitrary cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to total cost $C$ over the period $t_0 < t < t_1 $ is then
+
+\begin{myequation}
+\Delta C_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial C }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
+\label{eqn:integral_1}
+\end{myequation}
+
+However, data on the cost model variables is not available in continuous time.
+Disaggregating the contribution of single variables to the total cost reduction is thus not straightforward in our model.
+This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}.
+The following discussion follows an approach developed by Kavlak et al.
+\cite{kavlak2018evaluating}.
+For the detailed derivation, we refer to this publication.
+
+The cost function $C$ as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
+
+\begin{myequation}
+C(\vec{r}) = C(r_1,r_2, \dots) = \sum_i C_i
+\end{myequation}
+
+where
+
+\begin{myequation}
+C_i(\vec{r}) = C_i^0 \prod_w g_{iw}(r_w)
+\end{myequation}
+
+Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
+
+\begin{myequation}
+\Delta C_x = \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
+\end{myequation}
+
+where for $C(t)$ a constant $C(t) \approx \tilde{C} $ can be chosen such that $\Delta C_{x_i} = \Delta C$.
+In practice, this constant value can be approximated as $\tilde{C} \approx \frac{2}{3} C_i^\text{geo} + \frac{1}{3} \overline{C_i}$.
+The contribution of a single cost model variable $r_z$ can then be written as
+
+\begin{myequation}
+\Delta C_z (t_1,t_2) \approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
+\end{myequation}
+
+This approach was used to determine the effect of spillovers on device performance, displayed in  \cref{fig:breakthroughs_efficiency}.
+Due to time constraints, this approach could not be used to determine the effect of spillovers on manufacturing cost.
+Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of  \cref{eqn:cost_sum}, could this method.
 

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
updated figures

@@
 \label{eqn:cost_wafer}
-    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x \c_x \bigg\}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
 \end{myequation}
@@
     \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2002)=50$mm$\rightarrow5370$DPW, $d(2020)=100$mm$, \rightarrow26,838$DPW.
-Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, according to \cref{fig:wafers}.}
+Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews.
+For detailed statistics on changes in wafer size, compare also \cref{fig:wafers}.}
     \label{fig:dpw}
@@
     C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
-    &  \frac{1}{y_i}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x \c_x \bigg\}
+    &  \frac{1}{y_i}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
 \end{split}
@@
 \begin{myequation}
+\label{eqn:series}
     f_{\times n} = \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
-\label{eqn:series}
 \end{myequation}
@@
     \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_calibration.pdf}
-	\caption{Modelled manufacturing cost for light-emitting diode packages split by manufacturing step category.
-The top graph shows the data from LEDCOM model and round table discussions published by the United States Department of Energy (DoE).
-The bottom graph shows the results of the cost model developed in this thesis.
-Hatched bars are projections given by the sources.
-Inset plot shows the number of die per wafer (DPW) calculated for different wafer sizes used in the model.
+	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel).
+Hatched bars are projections given by US DOE for 2018 and 2020.
 Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
-Sources for bottom panel: own elaboration based on the cost model described in Section \cref{sec:costmodel}.
-Sources for Inset Plot: own elaborations based on numerical approximations for die-per-wafer estimations provided in \cite{de2005investigation}, informed by expert interviews.}
+Sources for bottom panel: own elaboration based on the cost model described in Section \cref{sec:costmodel}.}
 	\label{fig:costmodel_calibration}
 \end{figure}
-
-\cref{fig:costmodel_calibration} shows a comparison of the relative manufacturing cost by manufacturing step categories between the cost model and data released by the United States Department of Energy (DoE).
-Note that any comparison between the s must consider: Our model assumes state-of-the-art equipment, while the DoE industry round table reports data from factories which use older equipment, as well as newer equipment.
-Our model presently considers only low-power and mid-power packages, while the DoE reports consider high-power packages also.
-Our model assumes a virtual North American manufacturing location, while the majority of manufacturers produce in Asia.
-We note that the share of the substrate price in the DoE data is much larger than in our model.
-This can in part be explained by  \cref{fig:sapphire_prices}, which already shows how projections overestimated the actual price for 2015-2020.
-We note also that the share of the epitaxy step is larger in the DoE model.
-This can in part be explained by our model relying on state-of-the-art equipment, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
-In conclusion, our model fits industry forecasts due to the way it is set up.
-The packaging part of the manufacturing process is increasing in importance, which has been confirmed by researchers and industry reports on WLP\cite{Lee2011WPL}\cite{Xie2013}\cite{ledsmag2017WLP}.
 

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
updated references

@@
     \hline
-    Industry Periodicals & \cite{PM2020} \\
+    Industry Periodical & \cite{PM2020} \\
     \hline
-    Government Reports & \cite{council2013assessment} \\
+    Government Report & \cite{council2013assessment} \\
     \hline
@@
     \hline
-    Other Publications & \href{https://web.archive.org/web/20170801160530/https://www.energy.gov/sites/prod/files/2015/02/f19/craford_innovation_sanfrancisco2015.pdf}{2015 presentation (www.energy.gov)} \newline
-\href{https://web.archive.org/web/20170715230721/https://www.energy.gov/sites/prod/files/2016/02/f29/sun_china_raleigh2016.pdf}{2016 presentation (www.energy.gov)} \newline
-\href{http://web.archive.org/web/20160425025936/https://www.slideshare.net/Yole_Developpement/yole-led-packagingjanuary2013reportsample}{2013 presentation (www.slideshare.net)} \\
+    Other Publications & \cite{craford2015}\cite{sun2016}\cite{yole2013packaging} \\
     \hline

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
completed phosphor section

@@
     \parbox{\textwidth}{
-        \textbf{\textsc{Supplementary Information}} for: \\
+        \textit{\textsc{Supplementary Information}} for: \\
         M. Weinold, S. Kolesnikov, L.D. Anadon \\
@@
 
-This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiences and consumer experience metrics and discussed in \cref{subsec:metrics}, Table 1 and Figures 5-7 in the main article.
+This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics and discussed in section 5.2, Table 1 and Figures 5-7 in the main article.
 
@@
 Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel.
-Abbreviations: Arb.
-Units - Arbitrary Units (compare section \cref{subsec:arbunit}).
-Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: O$^\ddagger$ -  other parts of phosphor mixture not disclosed, G$^*$ - \ch{CaSrS:Eu^{2+}}, Y$^\star$ - $\beta-$SiAlON:Eu$^{2+}$, $^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, $^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
+Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: ?
+-  other parts of phosphor mixture not disclosed, G$^*$ - \ch{CaSrS:Eu^{2+}}, Y$^\star$ - $\beta-$SiAlON:Eu$^{2+}$, G$^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, O$^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
 Source (top panel): own elaboration based on spectral data.
@@
     \hline
-    \textbf{Technology} & \textbf{References} \\
+    \textit{Technology} & \textit{References} \\
     \hline
@@
     \hline
-    \textbf{Provider of References} & \textbf{References} \\
+    \textit{Provider of References} & \textit{References} \\
     \hline
@@
     \hline
-    \textbf{Type of Source} & \textbf{References} \\
+    \textit{Type of Source} & \textit{References} \\
     \hline
@@
     \hline
-    Patents & \href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11168235A}{JPH11168235A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DDE19921987A1}{DE19921987A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJPH11340514A}{JPH11340514A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DJP2001044498A}{JP2001044498A}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2005045893A1}{US2005045893A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2002093023A1}{US2002093023A1}, 
-\href{https://worldwide.espacenet.com/patent/search?q=pn\%3DUS2006273339A1}{US2006273339A1}  \\
+    Patents & \cite{patent1999uemura}\cite{patent1998takaoka}\cite{patent1999komaki}\cite{patent1999komaki}\cite{ludowise2006resonant}\cite{camras2005iii}\cite{steigerwald2004contacting} \\
     \hline
@@
     \hline
-    \textbf{Panel (Sub-Eff.)} & \textbf{References} \\
+    \textit{Panel (Sub-Eff.)} & \textit{References} \\
     \hline

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
more edits

@@
 
-To quantify the drivers of cost reductions across all process steps, one would need to identify the magnitude of contribution to cost reductions made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
-Mathematically, given a function $C$ which describes the cost associated with manufacturing a single unit at time $t$,
+To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
+Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
 
 \begin{myequation}
-C=ab+cd
+F=ab+cd
 \end{myequation}
 
-and manufacturing variables $a,b,c,d$, we are looking for the contribution $\Delta C_{a}$ made by a single variable $a$ to the total change in the cost function $\Delta C$ between points $t_0,t_1$ such, that
+and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
 
 \begin{myequation}
-\Delta C = C(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta C_i
+\Delta F = F(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta F_i
 \end{myequation}
 
-The infinitesimal contribution to the total cost by the infinitesimal change in a cost model variable is defined through the total differential of the cost function \cref{eqn:cost_sum} as
+The infinitesimal contribution to the total funcation value by the infinitesimal change in an input variable is defined through the total differential of the function as
 
 \begin{myequation}
-\text{d}C(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial C }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial C }{\partial x_i}  \Delta x_i
+\text{d}F(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial F }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial F }{\partial x_i}  \Delta x_i
 \end{myequation}
 
-where $x_i$ is an arbitrary cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to total cost $C$ over the period $t_0 < t < t_1 $ is then
+where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
 
 \begin{myequation}
-\Delta C_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial C }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
+\Delta F_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial F }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
 \label{eqn:integral_1}
@@
 
-However, data on the cost model variables is not available in continuous time.
-Disaggregating the contribution of single variables to the total cost reduction is thus not straightforward in our model.
+However, data on the input variables is not available in continuous time.
+Disaggregating the contribution of single variables to the change in the cost or performance is thus not straightforward in our model.
 This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}.
-The following discussion follows an approach developed by Kavlak et al.
+In our work, we propose to address this problem by following an approach developed by Kavlak et al.
 \cite{kavlak2018evaluating}.
@@
 
-The cost function $C$ as a function of a vector of cost model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
+The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
 
 \begin{myequation}
-C(\vec{r}) = C(r_1,r_2, \dots) = \sum_i C_i
+    F(\vec{r}) = F(r_1,r_2, \dots) = \sum_i F_i
 \end{myequation}
@@
 \begin{myequation}
-C_i(\vec{r}) = C_i^0 \prod_w g_{iw}(r_w)
+    F_i(\vec{r}) = F_i^0 \prod_w g_{iw}(r_w)
 \end{myequation}
@@
 \begin{myequation}
-\Delta C_x = \int_{t=t_0}^{t_1} C(t) \frac{ \partial \ln C }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
+    \Delta F_x = \int_{t=t_0}^{t_1} F(t) \frac{ \partial \ln F }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
 \end{myequation}
 
-where for $C(t)$ a constant $C(t) \approx \tilde{C} $ can be chosen such that $\Delta C_{x_i} = \Delta C$.
-In practice, this constant value can be approximated as $\tilde{C} \approx \frac{2}{3} C_i^\text{geo} + \frac{1}{3} \overline{C_i}$.
+where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$.
+In practice, this constant value can be approximated through the geometric and arithmetic means $\tilde{F} \approx \frac{2}{3} F_i^\text{geo} + \frac{1}{3} \overline{F_i}$.
 The contribution of a single cost model variable $r_z$ can then be written as
@@
 \begin{myequation}
-\Delta C_z (t_1,t_2) \approx \sum_i \tilde{C_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
+    \Delta F_z (t_1,t_2) \approx \sum_i \tilde{F_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
 \end{myequation}
 
-This approach was used to determine the effect of spillovers on device performance, displayed in  \cref{fig:breakthroughs_efficiency}.
-Due to time constraints, this approach could not be used to determine the effect of spillovers on manufacturing cost.
-Future work to quantify the impact of different cost drivers, including breakthroughs and changes in the manufacturing variables of  \cref{eqn:cost_sum}, could this method.
+We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article.
+Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.
 

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
various edits

@@
     \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
-For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.}
+For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.
+For the computation of total cost, see \cref{subsec:cost_computation}.
+For the computation of changes in total cost, see \cref{sec:contribution_variables}.}
     \label{fig:costmodel-schematic}
@@
 \subsection{Computation of Manufacturing Cost}
+\label{subsec:cost_computation}
 
@@
 The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
-
 \begin{myequation}
@@
 \end{myequation}
-
 where
-
 \begin{align*}
@@
 \end{align*}
-
 The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.
-
 \begin{myequation}
@@
 \end{myequation}
-
 The cost of a single process step $C_i$ can now be written as
-
 \begin{myequation}
@@
 \end{myequation}
-
 where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
-
 \begin{align*}
@@
 \end{align*}
-
 The number of die per wafer $D$ depends on the total usable wafer area.
 The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
-
 \begin{myequation}
@@
 \end{myequation}
-
 Determining the usable wafer area as a function of these three parameters requires a numerical solution.
 However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
-
 \begin{myequation}
@@
 
+\newpage
 \subsection{Preliminary Sensitivity Analysis}
 
-\begin{figure}[h]
+\begin{figure}[]
 	\centering
     \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_sensitivity.pdf}
-	\caption{Sensitivity analysis of selected cost model parameters.
-Note the decreasing sensitivity to cost model parameters as the number of die per wafer increases.
-Compare table \cref{tab:sensitivity}.
+	\caption{Sensitivity analysis for selected cost model parameters.
+Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in \cref{tab:sensitivity}.
+Panel b shows the resulting range of variation in the manufacturing cost.
+Ranges are represented for three years considered in the cost model, where corresponding data is available, by whiskers of the following colour: 2003 – blue, 2012 – green, 2020 – red.
+Note the overall trend of decreasing sensitivity to cost model parameters as the number of die per wafer increases over time.
 Abbreviations: resp.
-- respective}
+- respective; yield sav.
+from insp.
+- yield savings from inspection.}
 	\label{fig:sensitivity}
@@
 A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
-The amount of parameter variation was chosen to encompass the identified range of the value of the parameter in different manufacturing setups of each considered year.
-As an example, the lower range in the variation for metal cost (+50\%,-0\%) was chosen because the metal prices in the model are price quotes from the United State Geological Survey price database.
+The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
+As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen because the prices quotes from the United State Geological Survey price database.
 Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
-The upper range was chosen, because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
+The upper range was chosen because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
 The results of the analysis are shown in  \cref{fig:sensitivity}.
-The cost model is generally more sensitive to variation of parameters at smaller wafer diameters.
+The cost model is generally more sensitive to the variation in parameters at smaller wafer diameters.
 The most sensitive parameters are global parameters, such as yield or average equipment throughput.
 
-\begin{table}[H]
+\begin{table}[]
 \small
-    \begin{tabularx}{\textwidth}{ |X|l|l|l|l|l|l|l|X|}
+    \begin{tabularx}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
         \hline
-            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & \textit{Source} \\
+            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & Source \\
         \hline
-            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & Figure \cref{fig:electricity+cleanroom_prices} \\
+            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & \cite{mddi1997cleanroom}\cite{ledcomv2} \newline \cite{bakshi2009euv}\cite{gajera2006process} \\
         \hline
-            Manhours & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
+            Person-hour & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
         \hline
-            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I, \cite{Appleyard2001} \\
+            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I \newline \cite{Appleyard_2001} \\
         \hline
@@
         \hline
-            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I, Figure \cref{fig:wafer_size} \\
+            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I and \cref{fig:wafers} \\
         \hline
@@
         \hline
-            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & Figure \cref{fig:sapphire_prices} \\
+            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & \cref{fig:wafers} \\
         \hline
@@
     \caption{Cost model sensitivity analysis parameter list.
-For the sensitivity analysis performed using these variables, see  \cref{fig:sensitivity}.
-Units for values in columns \textit{2002}-\textit{2020} are indicated in column \textit{Units}.
-If values in columns \textit{2003}-\textit{2020} are instead are given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.}
+The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.
+The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
+If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
+Abbreviations: FTE - full-time equivalent; UPH - units per hour; Equip.
+Discount - equipment discount (sales rebate for large purchases); Inspec.
+Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).}
     \label{tab:sensitivity}
@@
 \section{Computation of Contribution of Individual Variables}
+\label{sec:contribution_variables}
 
@@
 Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
-
 \begin{myequation}
@@
 \end{myequation}
-
 and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
-
 \begin{myequation}
@@
 \end{myequation}
-
 The infinitesimal contribution to the total funcation value by the infinitesimal change in an input variable is defined through the total differential of the function as
-
 \begin{myequation}
@@
 \end{myequation}
-
 where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
-
 \begin{myequation}
@@
 \end{myequation}
-
 However, data on the input variables is not available in continuous time.
@@
 The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
-
 \begin{myequation}
@@
 \end{myequation}
-
 where
-
 \begin{myequation}
@@
 \end{myequation}
-
 Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
-
 \begin{myequation}
@@
 \end{myequation}
-
 where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$.
@@
 The contribution of a single cost model variable $r_z$ can then be written as
-
 \begin{myequation}
@@
 \end{myequation}
-
 We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article.
@@
 The most recent red narrow-band phosphor innovation included in Table 1 and Figure 7 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
-The first patent application for this phosphor was filed in 2016 .
+The first U.S.
+patent application for this phosphor was filed in 2017 \cite{seibald2019phosphor}.
 The SALON phosphor is a derivate of the SLA phosphor.

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
final edits (?)

@@
 
+\newpage
 \subsection{Example of Input Data Considered: Sapphire Wafers}
@@
 
+\newpage
 \subsection{Comparison with Reported Industry Data and DOE Projections}
@@
 
+\newpage
 \section{Computation of Contribution of Individual Variables}
@@
 
-This section summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies, as discussed in section 5.1 and shown in Figures 4-6 in the main article.
+\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies, as discussed in section 5.1 and shown in Figures 4-6 in the main article.
 Note that phosphor-related LED innovations affecting consumer experience metrics, along with spectral efficiency, are presented separately in Table 1 in the main publication and discussed in detail in \cref{sec:innovation_phosphor}.
@@
         \hline
-            $<$ 2003 & Optimization for \newline cavity effects & Other & No & I, \cite{Shen2003} \\
+            $<$ 2003 & Optimization for \newline cavity effects & reduces self-interference of quantum well & No & I, \cite{Shen2003} \\
         \hline
@@
         \hline
-            1994 & DH & Higher radiative recombination probability & No & \cite{Nakamura1994} \\
+            1994 & double heterostructure & Higher radiative recombination probability & No & \cite{Nakamura1994} \\
         \hline
@@
         \hline
-            ongoing & Active region \newline Doping & Both & No & I, \cite{schubert2018light} \\
+            ongoing & Active region \newline Doping & radiative recombination probability & No & I, \cite{schubert2018light} \\
         \hline
-            ongoing & Epitaxy \newline Improvements & Both & No & I \\
+            ongoing & Epitaxy \newline Improvements & radiative recombination probability & No & I \\
         \hline
-            ongoing & Chip architecture \newline Improvements & Both & No & I \\
+            ongoing & Chip architecture \newline Improvements & radiative recombination probability & No & I \\
         \hline
@@
 For a list of innovations affecting spectral efficiency, see Table 1 in the main article.
-Year column indicates the first instance of application of corresponding invention in white LEDs.
-Spillover column indicates if the innovation involved technology spillovers.
-Source column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.
-Source column additionally indicates if the spillover has been identified as part of an expert interview (I).}
+\textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
+'ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
+\textit{Spillover} column indicates if the innovation involved technology spillovers.
+\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
     \label{tab:innovations}
@@
 
-\subsection{History of Identified Phosphor-Related LED Innovations"}
+\subsection{History of Identified Phosphor-Related LED Innovations}
 

### 2023-03-24 2_SSL_EES/article/article_main.tex
vaaarious fixes

@@
 \includegraphics[width=18.5cm]{auxiliary/rsc_template_head_foot/header_bar}}\par
-\vspace{1em}
+\vspace{1em}3
 \sffamily
@@
 
- & \noindent\large{Michael Weinold \textit{$^{a b \ddag}$}, Sergey Kolesnikov,\textit{$^b$} and Laura Diaz Anadon\textit{$^{b}$}} \\
+ & \noindent\large{Michael Weinold \textit{$^{a, b \ddag}$}, Sergey Kolesnikov,\textit{$^b$} and Laura Diaz Anadon\textit{$^{b, c}$}} \\
 
@@
 \footnotetext{\textit{$^{a}$~ETH Zurich, Zurich, Switzerland. E-mail: michael.weinold@alumni.ethz.ch}}
-\footnotetext{\textit{$^{b}$~Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom. }}
+\footnotetext{\textit{$^{b}$~Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom.}}
+\footnotetext{\textit{$^{c}$~Belfer Center for Science and International Affairs, Harvard University, Cambridge, MA, United States.}}
 
@@
 \cite{Stephan2021} and consider knowledge to be external - thus a spillover -  if it has been developed for application in other technologies, sectors, or scientific disciplines.
-There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016}\cite{Stephan2021}\cite{Sun2021}\cite{kolesnikov_framework_2022} the future evolution of technologies.
+There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016}\cite{Stephan2021}\cite{Sun2021}\cite{kolesnikov2022technology} the future evolution of technologies.
 
@@
 
-The historical development XXXX development into powerful lghting devices from very dim novelty XXXX of light-emitting diodes received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
+The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
 The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006}\cite{krames2007led}\cite{laubsch2009high}\cite{hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
@@
 All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
-A summary of the background of interviewed experts is provided in Table SI 3 in Section 5 of the Supplementary Information.
+A summary of the background of interviewed experts is provided in Table SI 3 in Section 6 of the Supplementary Information.
 
@@
 The list of technology spillovers identified in our study is provided in Table \ref{tab:spillovers} and discussed in Section \ref{subsec:spillovers}.
-The complete list of all LED innovations is provided in Table SI 2 in Section 3 of the Supplementary Information.
+The complete list of all LED innovations is provided in Table SI 2 in Section 4 of the Supplementary Information.
 Through the index decomposition analysis described in Section \ref{subsec:metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
@@
         \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{SI Sec.} \\ \hline
-        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3 \\ \hline
-        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.1 \\ \hline
-        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3 \\ \hline
-        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3 \\ \hline
-        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3 \\ \hline
-        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3 \\ \hline
-        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3 \\ \hline
+        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 5.1.1 \\ \hline
+        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 5.1.1 \\ \hline
+        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 5.1.2 \\ \hline
+        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 5.1.4 \\ \hline
+        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 5.1.3 \\ \hline
+        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 5.1.2 \\ \hline
+        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 5.1.2 \\ \hline
     \end{tabularx}
@@
 Data points show the color temperature and colour rendering index performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year.
-Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 9 in section 4 of the Supplementary Information.
+Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 11 in section 5 of the Supplementary Information.
 Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison.

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
vaaarious fixes

@@
 \label{eqn:cost_wafer}
-    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{D}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{DPW}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
 \end{myequation}
@@
 \begin{align*}
-        D &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
+        DPW &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
         y &\dots \text{process step yield} \\
@@
 \begin{split}
-    C= P_s &+ \sum_i \bigg \{ \frac{1}{D} \frac{1}{y_i} \bigg[ \frac{t_i}{w_i u_i} \bigg((e*p) + l + m + d +o \bigg)_i  + \sum_{x} v_x p_x \bigg] \bigg \} + \\
+    C= P_s &+ \sum_i \bigg \{ \frac{1}{DPW} \frac{1}{y_i} \bigg[ \frac{t_i}{w_i u_i} \bigg((e*p) + l + m + d +o \bigg)_i  + \sum_{x} v_x p_x \bigg] \bigg \} + \\
     & + \sum_j \bigg \{ \frac{1}{y_j} \bigg[ \frac{c_j}{u_j}  \bigg((e*p) + l + m + d + o \bigg)_i + \sum_{x} a v_x p_x \bigg ] \bigg\}
@@
 \begin{table}[H]
+    \caption{Banana}
     \begin{tabularx}{\textwidth}{ |l|X|X|l|l| }
@@
         \hline
-            ongoing & Epitaxy improvements \newline and better doping & polarization and \newline bulk resistance & No & I \\
+            Ongoing & Epitaxy improvements \newline and better doping & Polarization and \newline bulk resistance & No & I \\
         \hline
@@
         \hline
-            $<$ 2003 & Optimization for \newline cavity effects & reduces self-interference of quantum well & No & I, \cite{Shen2003} \\
+            $<$ 2003 & Optimization for \newline cavity effects & Reduces self-interference of quantum well & No & I, \cite{Shen2003} \\
         \hline
@@
         \hline
-            ongoing & Chip design for high LEE & Total reflection and absorption & No & I, \cite{Haerle2004} \\
+            Ongoing & Chip design for high LEE & Total reflection and absorption & No & I, \cite{Haerle2004} \\
         \hline
@@
         \hline
-            1994 & double heterostructure & Higher radiative recombination probability & No & \cite{Nakamura1994} \\
+            1994 & Double heterostructure & Higher Radiative recombination probability & No & \cite{Nakamura1994} \\
         \hline
-            1996 & multiple quantum well & Higher radiative recombination probability & No & \cite{Koike1996} \\
+            1996 & Multiple quantum well & Higher Radiative recombination probability & No & \cite{Koike1996} \\
         \hline
-            ongoing & Active region \newline Doping & radiative recombination probability & No & I, \cite{schubert2018light} \\
+            Ongoing & Active region \newline Doping & Radiative recombination probability & No & I, \cite{schubert2018light} \\
         \hline
-            ongoing & Epitaxy \newline Improvements & radiative recombination probability & No & I \\
+            Ongoing & Epitaxy \newline Improvements & Radiative recombination probability & No & I \\
         \hline
-            ongoing & Chip architecture \newline Improvements & radiative recombination probability & No & I \\
+            Ongoing & Chip architecture \newline Improvements & Radiative recombination probability & No & I \\
         \hline
@@
         \hline
-            ongoing & Lower current density & Current density & No & I \\
+            Ongoing & Lower current density & Current density & No & I \\
         \hline
-            ongoing & Epitaxy \newline Improvements & Charge distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
+            Ongoing & Epitaxy \newline Improvements & Charge distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
         \hline
-            ongoing & Chip architecture \newline improvements  & Charge distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
+            Ongoing & Chip architecture \newline improvements  & Charge distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
         \hline
@@
         \end{tabularx}
-    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.
+    LED innovations and technology improvements affecting key white LED device sub-efficiencies.
 The table includes innovations and technology improvements affecting forward voltage efficiency, light extraction efficiency, internal quantum efficiency, and light conversion efficiency.
@@
 \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
-'ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
+'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
 \textit{Spillover} column indicates if the innovation involved technology spillovers.
-\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
+\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.
     \label{tab:innovations}

### 2023-03-24 2_SSL_EES/article/article_supplementary.tex
final commit????

@@
 \small
+    \caption{Cost model sensitivity analysis parameter list. The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.}
+    \vspace{5mm}
     \begin{tabularx}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
@@
         \end{tabularx}
-    \caption{Cost model sensitivity analysis parameter list.
-The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.
-The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
+    \vspace{5mm}
+
+    Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
 If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
@@
 Discount - equipment discount (sales rebate for large purchases); Inspec.
-Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).}
+Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).
     \label{tab:sensitivity}
@@
 \begin{table}[H]
-    \caption{Banana}
+    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.
+The table includes innovations and technology improvements affecting forward voltage efficiency, light extraction efficiency, internal quantum efficiency, and light conversion efficiency.
+For a list of innovations affecting spectral efficiency, see Table 1 in the main article.}
+    \vspace{5mm}
     \begin{tabularx}{\textwidth}{ |l|X|X|l|l| }
@@
         \end{tabularx}
-    LED innovations and technology improvements affecting key white LED device sub-efficiencies.
-The table includes innovations and technology improvements affecting forward voltage efficiency, light extraction efficiency, internal quantum efficiency, and light conversion efficiency.
-For a list of innovations affecting spectral efficiency, see Table 1 in the main article.
-\textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
+    \vspace{5mm}
+
+     Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
 'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
@@
     \centering
+    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
+    \vspace{5mm}
     \begin{tabular}{|l|l|l|l|l|}
@@
     \label{tab:interviews}
-    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
 \end{table}

### 2023-04-11 2_SSL_EST/article/article_draft_est.tex
started work on new draft

--- 2023-03-24 2_SSL_EES/article/article_main.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%This is the LaTeX ARTICLE template for RSC journals
-%Copyright The Royal Society of Chemistry 2016
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% This is a (brief) model paper using the achemso class
+%% The document class accepts keyval options, which should include
+%% the target journal and optionally the manuscript type. 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[journal=jacsat,manuscript=article]{achemso}
 
-\documentclass[twoside,twocolumn,9pt]{article}
-\usepackage{extsizes}
-\usepackage[super,sort&compress,comma]{natbib} 
-\usepackage[version=3]{mhchem}
-\usepackage[left=1.5cm, right=1.5cm, top=1.785cm, bottom=2.0cm]{geometry}
-\usepackage{balance}
-\usepackage{mathptmx}
-\usepackage{sectsty}
-\usepackage{graphicx} 
-\usepackage{lastpage}
-\usepackage[format=plain,justification=justified,singlelinecheck=false,font={stretch=1.125,small,sf},labelfont=bf,labelsep=space]{caption}
-\usepackage{float}
-\usepackage{fancyhdr}
-\usepackage{fnpos}
-\usepackage[english]{babel}
-\addto{\captionsenglish}{%
-  \renewcommand{\refname}{Notes and references}
-}
-\usepackage{array}
-\usepackage{droidsans}
-\usepackage{charter}
-\usepackage[T1]{fontenc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Place any additional packages needed here.  Only include packages
+%% which are essential, to avoid problems later. Do NOT use any
+%% packages which require e-TeX (for example etoolbox): the e-TeX
+%% extensions are not currently available on the ACS conversion
+%% servers.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
+\usepackage{tabularx}
 \usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{setspace}
-\usepackage[compact]{titlesec}
-\usepackage{hyperref}
-%%%Please don't disable any packages in the preamble, as this may cause the template to display incorrectly.%%%
 
-\usepackage{tabularx}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% If issues arise when submitting your manuscript, you may want to
+%% un-comment the next line.  This provides information on the
+%% version of every file you have used.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%\listfiles
 
-\usepackage{epstopdf}%This line makes .eps figures into .pdf - please comment out if not required.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Place any additional macros here.  Please use \newcommand* where
+%% possible, and avoid layout-changing macros (which are not used
+%% when typesetting).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
 
-\definecolor{cream}{RGB}{222,217,201}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Meta-data block
+%% ---------------
+%% Each author should be given as a separate \author command.
+%%
+%% Corresponding authors should have an e-mail given after the author
+%% name as an \email command. Phone and fax numbers can be given
+%% using \phone and \fax, respectively; this information is optional.
+%%
+%% The affiliation of authors is given after the authors; each
+%% \affiliation command applies to all preceding authors not already
+%% assigned an affiliation.
+%%
+%% The affiliation takes an option argument for the short name.  This
+%% will typically be something like "University of Somewhere".
+%%
+%% The \altaffiliation macro should be used for new address, etc.
+%% On the other hand, \alsoaffiliation is used on a per author basis
+%% when authors are associated with multiple institutions.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\usepackage{layout}
+\author{Michael Weinold}
+\email{michael.weinold@alumni.ethz.ch}
+\affiliation[ETHZ]{ETH Zurich, Switzerland}
+\alsoaffiliation[UCAM]
+{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, United Kingdom}
+\altaffiliation{Current address: Technology Assessment Group, Paul Scherrer Institute, Switzerland}
 
+\author{Sergey Kolesnikov}
+\affiliation[UCAM]
+{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, United Kingdom}
+
+\author{Laura Diaz Anadon}
+\affiliation[UCAM]{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, United Kingdom}
+\alsoaffiliation[UHarv]
+{Belfer Center for Science and International Affairs, Harvard University, USA}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The document title should be given as usual. Some journals require
+%% a running title from the author: this should be supplied as an
+%% optional argument to \title.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\title[]{Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Some journals require a list of abbreviations or keywords to be
+%% supplied. These should be set up here, and will be printed after
+%% the title and author information, if needed.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\abbreviations{IR,NMR,UV}
+\keywords{American Chemical Society, \LaTeX}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The manuscript does not need to include \maketitle, which is
+%% executed automatically.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-\pagestyle{fancy}
-\thispagestyle{plain}
-\fancypagestyle{plain}{
-%%%HEADER%%%
-\renewcommand{\headrulewidth}{0pt}
-}
-%%%END OF HEADER
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The "tocentry" environment can be used to create an entry for the
+%% graphical table of contents. It is given here as some journals
+%% require that it is printed as part of the abstract page. It will
+%% be automatically moved as appropriate.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{tocentry}
 
-%%%PAGE SETUP - Please do not change any commands within this section%%%
-\makeFNbottom
-\makeatletter
-\renewcommand\LARGE{\@setfontsize\LARGE{15pt}{17}}
-\renewcommand\Large{\@setfontsize\Large{12pt}{14}}
-\renewcommand\large{\@setfontsize\large{10pt}{12}}
-\renewcommand\footnotesize{\@setfontsize\footnotesize{7pt}{10}}
-\makeatother
+Some journals require a graphical entry for the Table of Contents.
+This should be laid out ``print ready'' so that the sizing of the
+text is correct.
 
-\renewcommand{\thefootnote}{\fnsymbol{footnote}}
-\renewcommand\footnoterule{\vspace*{1pt}% 
-\color{cream}\hrule width 3.5in height 0.4pt \color{black}\vspace*{5pt}} 
-\setcounter{secnumdepth}{5}
+Inside the \texttt{tocentry} environment, the font used is Helvetica
+8\,pt, as required by \emph{Journal of the American Chemical
+Society}.
 
-\makeatletter 
-\renewcommand\@biblabel[1]{#1}            
-\renewcommand\@makefntext[1]% 
-{\noindent\makebox[0pt][r]{\@thefnmark\,}#1}
-\makeatother 
-\renewcommand{\figurename}{\small{Fig.}~}
-\sectionfont{\sffamily\Large}
-\subsectionfont{\normalsize}
-\subsubsectionfont{\bf}
-\setstretch{1.125} %In particular, please do not alter this line.
-\setlength{\skip\footins}{0.8cm}
-\setlength{\footnotesep}{0.25cm}
-\setlength{\jot}{10pt}
-\titlespacing*{\section}{0pt}{4pt}{4pt}
-\titlespacing*{\subsection}{0pt}{15pt}{1pt}
-%%%END OF PAGE SETUP%%%
+The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
+permitted for  \emph{Journal of the American Chemical Society}
+graphical table of content entries. The box will not resize if the
+content is too big: instead it will overflow the edge of the box.
 
-%%%FOOTER%%%
-\fancyfoot{}
-\fancyfoot[LO,RE]{\vspace{-7.1pt}\includegraphics[height=9pt]{auxiliary/rsc_template_head_foot/LF}}
-\fancyfoot[CO]{\vspace{-7.1pt}\hspace{13.2cm}\includegraphics{auxiliary/rsc_template_head_foot/RF}}
-\fancyfoot[CE]{\vspace{-7.2pt}\hspace{-14.2cm}\includegraphics{auxiliary/rsc_template_head_foot/RF}}
-\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
-\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
-\fancyhead{}
-\renewcommand{\headrulewidth}{0pt} 
-\renewcommand{\footrulewidth}{0pt}
-\setlength{\arrayrulewidth}{1pt}
-\setlength{\columnsep}{6.5mm}
-\setlength\bibsep{1pt}
-%%%END OF FOOTER%%%
+This box and the associated title will always be printed on a
+separate page at the end of the document.
 
-%%%FIGURE SETUP - please do not change any commands within this section%%%
-\makeatletter 
-\newlength{\figrulesep} 
-\setlength{\figrulesep}{0.5\textfloatsep}
+\end{tocentry}
 
-\newcommand{\topfigrule}{\vspace*{-1pt}% 
-\noindent{\color{cream}\rule[-\figrulesep]{\columnwidth}{1.5pt}} }
-
-\newcommand{\botfigrule}{\vspace*{-2pt}% 
-\noindent{\color{cream}\rule[\figrulesep]{\columnwidth}{1.5pt}} }
-
-\newcommand{\dblfigrule}{\vspace*{-1pt}% 
-\noindent{\color{cream}\rule[-\figrulesep]{\textwidth}{1.5pt}} }
-
-\makeatother
-%%%END OF FIGURE SETUP%%%
-
-%%%TITLE, AUTHORS AND ABSTRACT%%%
-\twocolumn[
-  \begin{@twocolumnfalse}
-{\includegraphics[height=30pt]{auxiliary/rsc_template_head_foot/journal_name}\hfill\raisebox{0pt}[0pt][0pt]{\includegraphics[height=55pt]{auxiliary/rsc_template_head_foot/RSC_LOGO_CMYK}}\\[1ex]
-\includegraphics[width=18.5cm]{auxiliary/rsc_template_head_foot/header_bar}}\par
-\vspace{1em}3
-\sffamily
-\begin{tabular}{m{4.5cm} p{13.5cm} }
-
-\includegraphics{auxiliary/rsc_template_head_foot/DOI} & \noindent\LARGE{\textbf{Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers}} \\
-\vspace{0.3cm} & \vspace{0.3cm} \\
-
- & \noindent\large{Michael Weinold \textit{$^{a, b \ddag}$}, Sergey Kolesnikov,\textit{$^b$} and Laura Diaz Anadon\textit{$^{b, c}$}} \\
-
-\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{The first commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The abstract environment will automatically gobble the contents
+%% if an abstract is not used by the target journal.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
+  The first commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
 Since then, white LEDs have experienced major improvements in performance and efficiency.
@@
 We further show that technology spillovers — i.e., knowledge originating in other technologies, sectors, or scientific disciplines — affected all performance dimensions of LEDs, contributing no less than 8.5\% to total efficiency improvements and $\sim$100\% to the improvements in consumer experience metrics identified, thereby playing a key role in the widespread adoption of the technology.
-The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}
+The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.
+\end{abstract}
 
-\end{tabular}
-
- \end{@twocolumnfalse} \vspace{0.6cm}
-
-  ]
-%%%END OF TITLE, AUTHORS AND ABSTRACT%%%
-
-%%%FONT SETUP - please do not change any commands within this section
-\renewcommand*\rmdefault{bch}\normalfont\upshape
-\rmfamily
-\section*{}
-\vspace{-1cm}
-
-%%%FOOTNOTES%%%
-
-\footnotetext{\textit{$^{a}$~ETH Zurich, Zurich, Switzerland. E-mail: michael.weinold@alumni.ethz.ch}}
-\footnotetext{\textit{$^{b}$~Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom.}}
-\footnotetext{\textit{$^{c}$~Belfer Center for Science and International Affairs, Harvard University, Cambridge, MA, United States.}}
-
-%Please use \dag to\cite the ESI in the main text of the article.
-%If you article does not have ESI please remove the the \dag symbol from the title and the footnotetext below.
-\footnotetext{\dag~Electronic Supplementary Information (ESI) available: [details of any Supplementary Information available should be included here]. See DOI: 00.0000/00000000.}
-%additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required
-
-\footnotetext{\ddag~Present address: Paul Scherrer Institute, Group for Technology Assessment, Villigen, Switzerland.}
-
-%%%END OF FOOTNOTES%%%
-
-%%%MAIN TEXT%%%%
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Start the main part of the manuscript here.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 
-\begin{figure*}[h]
- \centering
- \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/history_efficacy.pdf}
- \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.
-Data points indicate best performers by year of market introduction.
-Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye.
-Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data.
-The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
-\cite{Murphy2012}.
-The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year.
-For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines.
-Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot.
-Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal.
-- Halogen, Incd.
-- Incandescent.
-Source: own synthesis of published data based on a visual approach proposed by Azevedo et al.
-\cite{azevedo2009transition}.
-See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
- \label{fgr:history_efficacy}
-\end{figure*}
-
 The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies}\cite{Anadon2016}.
@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
-  \includegraphics[height=6.5cm]{2_SSL_EES/article/figures/cost_lamp_small.pdf}
+  \includegraphics[height=6.5cm]{2_SSL_EST/article/figures/cost_lamp_small.pdf}
   \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
@@
 
-\section{Previous Literature}
-\label{sec:prev_lit}
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/history_efficacy.pdf}
+ \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.
+Data points indicate best performers by year of market introduction.
+Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye.
+Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data.
+The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
+\cite{Murphy2012}.
+The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year.
+For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines.
+Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot.
+Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal.
+- Halogen, Incd.
+- Incandescent.
+Source: own synthesis of published data based on a visual approach proposed by Azevedo et al.
+\cite{azevedo2009transition}.
+See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
+ \label{fgr:history_efficacy}
+\end{figure}
 
-The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
-The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006}\cite{krames2007led}\cite{laubsch2009high}\cite{hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
-
-Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status}\cite{Phillips2007}\cite{Bierhuizen2007}\cite{Nakamura2013}\cite{feezell2018invention}\cite{Taki2019}.
-Additional studies provide more detail regarding the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
-However, these studies typically include limited information on the macro-level chip design and cover disparate aspects of the technology over different time periods.
-Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos}\cite{cha2019semiconductor} and Soraa\cite{cich2017high}.
-
-We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure \ref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
-Despite the amount of literature published on the topic of LED history, we are not aware of a prior publication that comprehensively and consistently aggregates and analyzes known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
-In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
-This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012}\cite{Creutzig2018}.
-
-\begin{figure*}
+\clearpage
+\begin{figure}[h!]
  \centering
- \includegraphics[height=17cm]{2_SSL_EES/article/figures/chip_architecture_overview.pdf}
+ \includegraphics[height=15cm]{2_SSL_EST/article/figures/chip_architecture_overview.pdf}
  \caption{Historical evolution of light-emitting diode chip architectures.
@@
  \label{fgr:chip_architecture_overview}
-\end{figure*}
+\end{figure}
 
-\section{Metrics for technological progress in LEDs}
+\section{Metrics for Technological Progress in LEDs}
 
@@
 
-\section{Methods and Data Collection}
-\label{sec:methods}
-
-The evolution of LED device architecture and performance as well as the progress in understanding the underlying physical phenomena are well covered in the scholarly literature and patents, as described in Section \ref{sec:prev_lit}.
-However, information provided in such sources is insufficient for our goals on at least three accounts: First, existing work focuses only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
-Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
-Second, not all relevant innovations are patented \cite{Pakes_1980}\cite{ Fontana_2013}.
-In the case of LED patents in particular, our interviews with industry experts suggest that the propensity to patent is the highest for knowledge related to macroscopic device architecture and chemical composition of phosphors, and the lowest for knowledge related to manufacturing process improvements and microscopic chip architecture that is difficult to reconstruct by reverse engineering.
-This means that relying only on patent literature would bias results by unduly emphasizing some focus areas and de-emphasizing others.
-Third, scientific publications and patents typically focus on experimental devices, rather than commercial products.
-While new LED features, designs and manufacturing methods reported in these sources can potentially result in significant performance gains or cost reductions, it is difficult to ascertain if these improvements have since been adopted in industry.
-Furthermore, information on LED manufacturing cost and the effect of process improvements on the total cost is highly proprietary.
-Estimates are occasionally reported in the scientific literature and company publications, but these often do not disclose which parts of the manufacturing process are responsible for the largest contribution to the overall cost, or which improvements led to cost reductions.
-
-To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis that combines information obtained from a review of the primary scientific literature, device datasheets and relevant patents with our own computations of device sub-efficiencies, information gained from industry publications, semi-structured interviews with experts from academia and industry, and bottom-up manufacturing cost modelling.
-We provide the details of implementation of each of these methods below.
-
-\subsection{Systematic Literature Review}
-
-We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
-This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in Section \ref{subsec:metrics}; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
-
-Relevant sources for the review were found in an iterative search process that involved two components.
-The first was the search in specialized patent and publication databases as well as company websites.
-The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the Section \ref{sec:prev_lit} and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
-We also relied on backward citations for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
-
-\subsection{Semi-Structured Interviews}
-
-To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews with eleven eminent experts from academia, industry and the public research sector.
-Experts were selected based on their engagement in  different sub-fields of LED research and manufacturing, as well as the recommendations from other interviewed experts, in essence expanding the list of experts that emerged from the initial literature review.
-All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
-A summary of the background of interviewed experts is provided in Table SI 3 in Section 6 of the Supplementary Information.
-
-The primary, structured part of the interviews explored which innovations were deemed most relevant to the evolution of device performance, consumer experience and manufacturing cost of LED packages.
-Thereafter, interviewees were asked to consider the extent to which these innovations may have originated outside of their respective field of expertise and the LED industry more broadly—i.e., which of the innovations may be considered spillovers.
-The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
-Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in Section \ref{sec:costmodel}) and verify device performance data.
-
-\subsection{Performance Metrics Calculations}
-\label{subsec:metrics}
-
-The contribution of individual technology innovations and spillovers to the progress in overall device efficiency over time is estimated by index decomposition analysis.
-Mathematically, this involves breaking down a chosen performance indicator into its constituent components, each representing a specific factor that contributes to the change in the indicator \cite{Ang1997}.
-Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
-It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
-
-According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
-
-\begin{align}
-    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
-    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
-\end{align}
-
-where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
-These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:
-
-\begin{equation}
-    \Delta a + \Delta b  = \Delta F
-\end{equation}
-
-To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered.
-However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance.
-For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies where possible and necessary.
-Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications.
-In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra.
-This approach allowed us to  quantify the improvements related to phosphor development in LEDs.
-
-\subsection{Bottom-Up Manufacturing Cost Model}
-\label{sec:costmodel}
-
-We address the limitations of data availability for the LED manufacturing cost by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
-In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
-
-We identified two existing cost models for LED manufacturing in the prior literature.
-The first is the LED Cosim model developed by the technology consultancy System Plus Consulting \cite{cosim2013} \cite{cosim2019}.
-Due to prohibitive licensing costs, this model could not be used in our study.
-The second model is the publicly available 2012 LEDCOM cost model prepared for the U.S.
-Department of Energy by Stephen Bland  \cite{ledcomv2}.
-It was set up to capture the manufacturing process for phosphor-converted LEDs of different power levels.
-However, given the remarkable changes in the performance, cost and architecture of LEDs since 2012, as shown in Figures 1-3, this model cannot fully capture the recent advances in LED technology, necessitating a significant modification to bring it up to date with the current state-of-the-art in LED manufacturing.
-
-Given this situation, we developed our own bottom-up LED manufacturing cost model.
-The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
-The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
-Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips.
-The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
-In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
-These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
-
-The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
-Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
-We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
-It instead assumes an hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
-It also assumes the use of the most up-to-date equipment for that year.
-Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
-An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
-Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
-
 \section{Results}
@@
 
-\begin{figure*}[h!]
+\begin{figure}[h!]
  \centering
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/subefficiencies_progress.pdf}
+ \includegraphics[width=17cm]{2_SSL_EST/article/figures/subefficiencies_progress.pdf}
  \caption{Historical improvements in light-emitting diode technology.
@@
 Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
-\newline
 \textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
@@
  \label{fgr:subeff}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[h!]
+\begin{figure}[h!]
  \centering
- \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
+ \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/breakthroughs_efficiency.pdf}
  \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
@@
  \label{fgr:breakthroughs_efficiency}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[h!]
+\begin{figure}[h!]
  \centering
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
+ \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2010.pdf}
+ \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2016.pdf}
+ \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2020.pdf}
  \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
@@
  \label{fgr:waterfall}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{table*}[h!]
+\begin{table}[h!]
     \small
@@
     \label{tab:phosphors}
-\end{table*}
+\end{table}
 
-\begin{figure*}[h!]
+\begin{figure}[h!]
  \centering
- \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_consumer-experience.pdf}
+ \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/breakthroughs_consumer-experience.pdf}
  \caption{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.
@@
  \label{fgr:consumer_experience}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}[h!]
- \centering
- \includegraphics[width=17.5cm]{2_SSL_EES/article/figures/costmodel_results_years.pdf}
- \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
-location.
-Consecutive steps shown on the x-axis.
-Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
-Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
-Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
-For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information.
-Abbreviations: Litho – Lithographic Process, Insp.
-– Inspection, Depn.
-– Deposition, CMP - Chemical-Mechanical Planarization}
- \label{fgr:costmodel}
-\end{figure*}
-
-\begin{table*}[h]
+\begin{table}[h]
     \small
@@
     \label{tab:spillovers}
-\end{table*}
+\end{table}
 
@@
 
-\section*{Conflicts of interest}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The "Acknowledgement" section can be given in all manuscript
+%% classes.  This should be given within the "acknowledgement"
+%% environment, which will make the correct section or running title.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{acknowledgement}
 
-There are no conflicts to declare.
-
-\section*{Acknowledgements}
 This research was supported by the grant from the Alfred P.
@@
 The role of technology spillovers and government investment”}.
-Michael Weinold further gratefully acknowledges support from the Swiss Study Foundation.
-We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
-We also express our deepest gratitude to all interviewees for their willingness to participate in this study and share their invaluable insights with us.
+Michael Weinold gratefully acknowledges support from the Swiss Study Foundation.
+The authors thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
+The authors further express their gratitude to all interviewees for their willingness to participate in this study and share their insights.
 
-%%%END OF MAIN TEXT%%%
+\end{acknowledgement}
 
-%The \balance command can be used to balance the columns on the final page if desired. It should be placed anywhere within the first column of the last page.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The same is true for Supporting Information, which should use the
+%% suppinfo environment.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{suppinfo}
 
-\balance
+This will usually read something like: ``Experimental procedures and
+characterization data for all new compounds. The class will
+automatically add a sentence pointing to the information on-line:
 
-%If notes are included in your references you can change the title from 'References' to 'Notes and references' using the following command:
-%\renewcommand\refname{Notes and references}
+\end{suppinfo}
 
-%%%REFERENCES%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The appropriate \bibliography command should be placed here.
+%% Notice that the class file automatically sets \bibliographystyle
+%% and also names the section correctly.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliography{bibliography}
-\bibliographystyle{auxiliary/rsc_style_files/bibliography_rsc} %the RSC's .bst file
 

### 2023-04-11 2_SSL_EST/article/article_supplementary.tex
started work on new draft

--- 2023-03-24 2_SSL_EES/article/article_supplementary.tex
@@
 	\centering
-    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_sensitivity.pdf}
+    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_sensitivity.pdf}
 	\caption{Sensitivity analysis for selected cost model parameters.
@@
 	\centering
-    \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/costmodel_calibration.pdf}
+    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_calibration.pdf}
 	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel).
@@
 	\centering
-    \includegraphics[width=0.95\textwidth]{2_SSL_EES/article/figures/phosphor_spectrum-comparison.pdf}
+    \includegraphics[width=0.95\textwidth]{2_SSL_EST/article/figures/phosphor_spectrum-comparison.pdf}
 	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 7 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year.
@@
 
-\section{List of Interviewees}
+\section{List of InterviewEST}
 
@@
 \clearpage
-\newpage
+\section{Previous Literature}
+\label{sec:prev_lit}
+
+The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
+The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006}\cite{krames2007led}\cite{laubsch2009high}\cite{hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
+
+Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status}\cite{Phillips2007}\cite{Bierhuizen2007}\cite{Nakamura2013}\cite{feezell2018invention}\cite{Taki2019}.
+Additional studies provide more detail regarding the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
+However, these studies typically include limited information on the macro-level chip design and cover disparate aspects of the technology over different time periods.
+Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos}\cite{cha2019semiconductor} and Soraa\cite{cich2017high}.
+
+We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure \ref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+Despite the amount of literature published on the topic of LED history, we are not aware of a prior publication that comprehensively and consistently aggregates and analyzes known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
+In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
+This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012}\cite{Creutzig2018}.
+
+\clearpage
 \section{Complete List of Sources for Figures}

### 2023-04-14 2_SSL_EST/article/article_draft_est.tex
edits

--- 2023-03-24 2_SSL_EES/article/article_main.tex
@@
 
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/history_efficacy.pdf}
+ \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.
+Data points indicate best performers by year of market introduction.
+Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye.
+Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data.
+The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
+\cite{Murphy2012}.
+The magnified plot shows the progress in cool white LEDs from 1996 to 2020, with the dashed line indicating a linear rate of efficacy improvement of 10lm/W per year.
+For comparison, efficacies of best performers in legacy lighting technologies for 2020 are shown as coloured horizontal lines.
+Note the logarithmic scale of the vertical axis on the main plot and the linear scale on the magnified plot.
+Abbreviations: HID - High-Intensity Discharge; CFL - Compact Fluorescent Lamp; Hal.
+- Halogen, Incd.
+- Incandescent.
+Source: own synthesis of published data based on a visual approach proposed by Azevedo et al.
+\cite{azevedo2009transition}.
+See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
+ \label{fgr:history_efficacy}
+\end{figure}
+
+\clearpage
+\begin{figure}[h!]
+ \centering
+ \includegraphics[height=15cm]{2_SSL_EST/article/figures/chip_architecture_overview.pdf}
+ \caption{Historical evolution of light-emitting diode chip architectures.
+Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
+Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
+The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip.
+Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity.
+Years indicated correspond to the earliest identified patent priority date.
+Black frames around certain designs indicate chip designs not brought to large-scale production.
+Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
+Source: adapted and compiled from multiple patents and industry publications.
+See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
+ \label{fgr:chip_architecture_overview}
+\end{figure}
+
 \section{Previous Literature}
@@
 
-\begin{figure*}
- \centering
- \includegraphics[height=17cm]{2_SSL_EES/article/figures/chip_architecture_overview.pdf}
- \caption{Historical evolution of light-emitting diode chip architectures.
-Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
-Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
-The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip.
-Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity.
-Years indicated correspond to the earliest identified patent priority date.
-Black frames around certain designs indicate chip designs not brought to large-scale production.
-Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
-Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
- \label{fgr:chip_architecture_overview}
-\end{figure*}
-
-\section{Metrics for technological progress in LEDs}
+\section{Metrics for Technological Progress in LEDs}
 
@@
 We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology.
+In line with previous publications \cite{schubert2018light}\cite{tsao2010solid}, use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
+The governing equations are provided in Section 1.3 in the Supplementary Information.
 
-\subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
-
-Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
-\cite{tsao2010solid}} (VfE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
-These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
-It is defined as
-
-\begin{equation}
-    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
-\end{equation}
-
-where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Light Extraction Efficiency $\eta_{LE}$}
-
-Device light-extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
-
-\begin{equation}
-    \eta_{LE}=\frac{\text{\# of photons out}}{\text{\# of photons created}} = \frac{P_{opt}}{P_{int}}
-\end{equation}
-
-where $P_{opt}$ is the optical power of the device and $P_{int}$ is the internal power of the device\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Internal Quantum Efficiency $\eta_{IQ}$}
-
-Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
-\begin{equation}
-\label{eqn:ieq-eqe}
-    \eta_{EQ} = \eta_{IQ} \times \eta_{LE}
-\end{equation}
-
-and defined as
-
-\begin{equation}
-    \eta_{IQ} = \frac{\# \text{of photons created}}{\# \text{of electron-hole pairs in}}
-\end{equation}
-
-and depends on the current density of the device\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Droop $\eta_{droop}$}
-
-Efficiency droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
-This is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
-It is defined as
-
-\begin{equation}
-    \eta_{droop} = 1 - \frac{\eta_{IQE}}{\eta_{IQE}(A \rightarrow 0)}
-\end{equation}
-
-where $\eta_{IQE}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
-In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$ as
-
-\begin{equation}
-\label{eqn:droop}
-    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
-\end{equation}
-
-Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Conversion Efficiency $\eta_{C}$}
-
-Light conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
-These losses are the sum of the Stokes loss and well as scattering/absorption losses.
-It is defined as
-
-\begin{equation}
-    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i}
-\end{equation}
-
-where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
-Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components, i.e.
-\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.
-
-\subsubsection{Spectral Efficiency $\eta_{S}$}
-
-Spectral efficiency (SE) describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
-Photons converted into the infrared or ultraviolet are lost to illumination purposes.
-It is defined as
-
-\begin{equation}
-    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
-\end{equation}
-
-where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
-$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Overall Efficiency ("Lamp Efficiency") $\eta_L$}
-
-Finally, the overall efficiency of a light-emitting diode package is the product of all considered sub-efficiencies
+Finally, the overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies
 
@@
 
+\clearpage
 \section{Results}
@@
 
-The historical development of a selection of the sub-efficiencies introduced in Section \ref{sec:device_performance_metrics} is shown in Figure \ref{fgr:subeff}.
-Using this data, together with data on the remaining sub-efficiencies not shown in this figure, we calculated the overall LED efficiency in for four years: 2003, 2010, 2016 and 2020.
+The historical development of the sub-efficiencies introduced in Section \ref{sec:device_performance_metrics} is shown in Figure SI12 in the Supplementary Information.
+Using this data, we calculated the overall LED efficiency in for four years: 2003, 2010, 2016 and 2020.
 Figure \ref{fgr:waterfall} shows how improvements in sub-efficiencies led to improvements in overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.7\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
@@
 
-\begin{figure*}[h!]
+\begin{figure}[h!]
  \centering
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/subefficiencies_progress.pdf}
- \caption{Historical improvements in light-emitting diode technology.
-Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
-Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
-\newline
- \textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
-The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
-Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
- \textbf{Panel A2:} Efficiency droop at the test current of I=350mA.
- \textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
-Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}.
-Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
-“Not reported” denotes data points where the IQE measurement technique was not reported.
-Whiskers indicate reported ranges, where applicable.
- \textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
-The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given.
-Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
-- electrode; text.
-- textured.
- Source: own synthesis of data from the full list of sources provided in the Supplementary Information, Section 5.
-Data on average device performance adapted from U.S.
-Department of Energy (DOE) Reports\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report\cite{beale_leds_2015}.}
- \label{fgr:subeff}
-\end{figure*}
-
-\begin{figure*}[h!]
- \centering
- \includegraphics[width=\textwidth]{2_SSL_EES/article/figures/breakthroughs_efficiency.pdf}
+ \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/breakthroughs_efficiency.pdf}
  \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
@@
 Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}.
-Source: own elaboration based on data in Figure \ref{fgr:waterfall} and Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.}
+Source: own elaboration based on data in Figure \ref{fgr:waterfall} and Figure SI12, as detailed in Section \ref{sec:methods}.}
  \label{fgr:breakthroughs_efficiency}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[h!]
+\begin{figure}[h!]
  \centering
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2003.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2010.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2016.pdf}
- \includegraphics[width=17cm]{2_SSL_EES/article/figures/waterfall_performance_2020.pdf}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2010.pdf}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2016.pdf}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2020.pdf}
+ \captionsetup{font=footnotesize}
  \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
@@
 = Scattering.
-Source (Efficiency): own elaboration based on data in Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.
+Source (Efficiency): own elaboration based on data in Figure SI12, as detailed in Section \ref{sec:methods}.
 Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
  \label{fgr:waterfall}
-\end{figure*}
+\end{figure}
 
@@
 
-Figure \ref{fgr:costmodel} shows key results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages.
  We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
+A detailed breakdown is provided in Figure SI13 in the Supplementary Information.
 
@@
 In the case of manufacturing yield, the higher it is, the less inputs are wasted on the production of a single LED package.
-With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare Figure \ref{fgr:costmodel}, Panels A-C), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period.
+With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020, it is not surprising that the total yielded LED manufacturing cost significantly declined over this period.
 In the case of wafer diameter, the larger the wafer, the more LED chips can be produced from a single wafer.
@@
 With more than a thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time.
-As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in Figure \ref{fgr:costmodel}).
+As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in Figure SI13).
 However, as our interviewees noted, while growing LEDs on larger wafers is economically desirable, it is associated with engineering and epitaxy challenges as well as high up-front cost of new equipment.
 
-Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
+Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
 We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model\cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables\cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.
 
-\begin{figure*}[h!]
- \centering
- \includegraphics[width=17.5cm]{2_SSL_EES/article/figures/costmodel_results_years.pdf}
- \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
-location.
-Consecutive steps shown on the x-axis.
-Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
-Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
-Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
-For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information.
-Abbreviations: Litho – Lithographic Process, Insp.
-– Inspection, Depn.
-– Deposition, CMP - Chemical-Mechanical Planarization}
- \label{fgr:costmodel}
-\end{figure*}
-
-\begin{table*}[h]
-    \small
+\clearpage
+\begin{table}[h!]
+    \tiny
     \centering
@@
         1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new}\cite{bando1996}\cite{bando1998development}\cite{shimizu1999light}\cite{cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
-        1967 & 1996 & <2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical}\cite{bando1998development}\cite{shimizu1999light}\cite{Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
-        1982 & 1996 & <2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered}\cite{feezell2018invention}\cite{Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare Figure \ref{fig:chip_architecture_overview})\\ \hline
-        1971 & 1999 & <2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf}\cite{fraser1972highly}\cite{margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare Figure \ref{fig:chip_architecture_overview}) \\ \hline
+        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical}\cite{bando1998development}\cite{shimizu1999light}\cite{Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
+        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered}\cite{feezell2018invention}\cite{Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare Figure \ref{fgr:chip_architecture_overview})\\ \hline
+        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf}\cite{fraser1972highly}\cite{margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare Figure \ref{fgr:chip_architecture_overview}) \\ \hline
         1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997}\cite{mueller2004phosphor}\cite{MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
@@
 In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost resulted from an increase in the wafer size used in manufacturing and higher yields across manufacturing steps.
-This can be seen in the reduction of the contribution of wafer processing steps to  totaal cost in Figure \ref{fgr:costmodel}.
+This can be seen in the reduction of the contribution of wafer processing steps to  total cost in Figure SI13.
 Both were enabled by advances in manufacturing equipment performance and incremental process improvements from learning-by-doing.
@@
 The role of technology spillovers and government investment”}.
-Michael Weinold further gratefully acknowledges support from the Swiss Study Foundation.
-We would like to thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
-We also express our deepest gratitude to all interviewees for their willingness to participate in this study and share their invaluable insights with us.
+Michael Weinold gratefully acknowledges support from the Swiss Study Foundation.
+The authors thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
+The authors further express their gratitude to all interviewees for their willingness to participate in this study and share their insights.
 
-%%%END OF MAIN TEXT%%%
+\end{acknowledgement}
 
-%The \balance command can be used to balance the columns on the final page if desired. It should be placed anywhere within the first column of the last page.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The same is true for Supporting Information, which should use the
+%% suppinfo environment.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{suppinfo}
 
-\balance
+Methodological details on the manufacturing cost model and the contribution of variables.
+Governing equations for relevant metrics in lighting.
+Detailed description of the technology spillovers in phosphor materials for light down-conversion.
+List of expert interviewees.
+Complete list of sources for Figures \ref{fgr:chip_architecture_overview} and \ref{fgr:history_efficacy}.
 
-%If notes are included in your references you can change the title from 'References' to 'Notes and references' using the following command:
-%\renewcommand\refname{Notes and references}
+\end{suppinfo}
 
-%%%REFERENCES%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The appropriate \bibliography command should be placed here.
+%% Notice that the class file automatically sets \bibliographystyle
+%% and also names the section correctly.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliography{bibliography}
-\bibliographystyle{auxiliary/rsc_style_files/bibliography_rsc} %the RSC's .bst file
 

### 2023-04-14 2_SSL_EST/article/article_supplementary.tex
edits

@@
 
-\section{Luminous Efficacy Metrics}
+\section{Metrics}
 
@@
 
+\subsection{Sub-Efficiencies}
+
+\subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
+
+Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
+\cite{tsao2010solid}} (VfE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
+These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
+It is defined as
+
+\begin{equation}
+    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
+\end{equation}
+
+where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Light Extraction Efficiency $\eta_{LE}$}
+
+Device light-extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+
+\begin{equation}
+    \eta_{LE}=\frac{\text{\# of photons out}}{\text{\# of photons created}} = \frac{P_{opt}}{P_{int}}
+\end{equation}
+
+where $P_{opt}$ is the optical power of the device and $P_{int}$ is the internal power of the device\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Internal Quantum Efficiency $\eta_{IQ}$}
+
+Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
+\begin{equation}
+\label{eqn:ieq-eqe}
+    \eta_{EQ} = \eta_{IQ} \times \eta_{LE}
+\end{equation}
+
+and defined as
+
+\begin{equation}
+    \eta_{IQ} = \frac{\# \text{of photons created}}{\# \text{of electron-hole pairs in}}
+\end{equation}
+
+and depends on the current density of the device\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Droop $\eta_{droop}$}
+
+Efficiency droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
+This is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
+It is defined as
+
+\begin{equation}
+    \eta_{droop} = 1 - \frac{\eta_{IQE}}{\eta_{IQE}(A \rightarrow 0)}
+\end{equation}
+
+where $\eta_{IQE}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
+In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$ as
+
+\begin{equation}
+\label{eqn:droop}
+    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
+\end{equation}
+
+Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
+
+\subsubsection{Conversion Efficiency $\eta_{C}$}
+
+Light conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
+These losses are the sum of the Stokes loss and well as scattering/absorption losses.
+It is defined as
+
+\begin{equation}
+    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i}
+\end{equation}
+
+where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
+Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components, i.e.
+\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.
+
+\subsubsection{Spectral Efficiency $\eta_{S}$}
+
+Spectral efficiency (SE) describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
+Photons converted into the infrared or ultraviolet are lost to illumination purposes.
+It is defined as
+
+\begin{equation}
+    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
+\end{equation}
+
+where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
+$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
+
 \newpage
-\section{Manufacturing Cost Model}
+\section{Methodology}
 \label{sec:costmodel}
 
-\subsection{Structure of the Model}
+\subsection{Manufacturing Cost Model}
+
+\subsubsection{Structure of the Model}
 
@@
 
-\subsection{Manufacturing Process Steps by Chip Architecture}
+\subsubsection{Manufacturing Process Steps by Chip Architecture}
 
@@
 
-\subsection{Computation of Manufacturing Cost}
+\subsubsection{Computation of Manufacturing Cost}
 \label{subsec:cost_computation}
@@
 
-\subsection{Computation of Yielded Cost}
+\subsubsection{Computation of Yielded Cost}
 
@@
 \newpage
-\subsection{Example of Input Data Considered: Sapphire Wafers}
+\subsubsection{Example of Input Data Considered: Sapphire Wafers}
 
@@
 \newpage
-\subsection{Preliminary Sensitivity Analysis}
+\subsubsection{Preliminary Sensitivity Analysis}
 
@@
 \newpage
-\subsection{Comparison with Reported Industry Data and DOE Projections}
+\subsubsection{Comparison with Reported Industry Data and DOE Projections}
 
@@
 \newpage
-\section{Computation of Contribution of Individual Variables}
+\subsection{Computation of Contribution of Individual Variables}
 \label{sec:contribution_variables}
@@
 
-\section{Non-Phosphor-Related LED Innovations}
+\section{LED Innovations}
+
+\subsection{Non-Phosphor-Related}
 
@@
 
-\section{Phosphor-Related LED Innovations}
+\subsection{Phosphor-Related}
 \label{sec:innovation_phosphor}
@@
 
-\subsection{History of Identified Phosphor-Related LED Innovations}
+\subsubsection{History of Identified Phosphor-Related LED Innovations}
 
-\subsubsection{YAG and YGAG Phosphors}
+\paragraph{YAG and YGAG Phosphors}
 
@@
 
-\subsubsection{258, SLA and SALON phosphors}
+\paragraph{258, SLA and SALON phosphors}
 
@@
 
-\subsubsection{PFS Phosphor}
+\paragraph{PFS Phosphor}
 
@@
 
-\subsubsection{Quantum Dots for Light Down-Conversion}
+\paragraph{Quantum Dots for Light Down-Conversion}
 
@@
 
-\subsection{Spectral Data for Identified LED Phosphor Innovations}
+\paragraph{Spectral Data for Identified LED Phosphor Innovations}
 
@@
 
-\section{List of InterviewEST}
+\section{List of Interviews}
 
@@
 \end{table}
+
+\clearpage
+\section{Results}
+
+\subsection{Historical Progress in Sub-Efficiencies}
+
+The historical progress in the different sub-efficiencies contributing to overall efficiency (lamp efficiency) is shown in \cref{fgr:subeff}. Context for figure panels is provided below:
+
+\textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
+The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
+Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
+\
+
+\textbf{Panel A2:} Efficiency droop at the test current of I=350mA. \
+
+\textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
+Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}.
+Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
+“Not reported” denotes data points where the IQE measurement technique was not reported.
+Whiskers indicate reported ranges, where applicable.
+\
+
+\textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
+The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given.
+Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
+- electrode; text.
+- textured.
+
+\begin{figure}[h!]
+ \centering
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/subefficiencies_progress.pdf}
+ \caption{Historical improvements in light-emitting diode technology.
+Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
+Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
+Source: own synthesis of data from the full list of sources provided in \cref{sec:sources}.
+Data on average device performance adapted from U.S.
+Department of Energy (DOE) Reports \cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report \cite{beale_leds_2015}.}
+ \label{fgr:subeff}
+\end{figure}
+
+\subsection{Manufacturing Cost Model}
+
+\begin{figure*}[h!]
+ \centering
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/costmodel_results_years.pdf}
+ \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+location.
+Consecutive steps shown on the x-axis.
+Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
+Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
+Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
+For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information.
+Abbreviations: Litho – Lithographic Process, Insp.
+– Inspection, Depn.
+– Deposition, CMP - Chemical-Mechanical Planarization}
+ \label{fgr:costmodel}
+\end{figure*}
 

### 2023-04-17 2_SSL_EST/article/article_draft_est.tex
moved sections around

@@
 
-The cost model we developed and describe in Section \ref{sec:costmodel} relies on a cumulative approach to yielded cost\cite{becker2001use}.
+The cost model we developed and describe in Section 2.3 of the Supplementary Information relies on a cumulative approach to yielded cost\cite{becker2001use}.
 In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
@@
 Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
-Second, not all relevant innovations are patented \cite{Pakes_1980}\cite{ Fontana_2013}.
+Second, not all relevant innovations are patented \cite{Pakes_1980}\cite{Fontana_2013}.
 In the case of LED patents in particular, our interviews with industry experts suggest that the propensity to patent is the highest for knowledge related to macroscopic device architecture and chemical composition of phosphors, and the lowest for knowledge related to manufacturing process improvements and microscopic chip architecture that is difficult to reconstruct by reverse engineering.
@@
 To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis that combines information obtained from a review of the primary scientific literature, device datasheets and relevant patents with our own computations of device sub-efficiencies, information gained from industry publications, semi-structured interviews with experts from academia and industry, and bottom-up manufacturing cost modelling.
-We provide the details of implementation of each of these methods below.
+The methodological details are discussed in Section 2 of the Supplementary Information.
 
@@
 The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
-Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in Section \ref{sec:costmodel}) and verify device performance data.
+Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in Section 2.3 of the Supplementary Information) and verify device performance data.
 
@@
 This approach allowed us to  quantify the improvements related to phosphor development in LEDs.
-
-\subsection{Bottom-Up Manufacturing Cost Model}
-\label{sec:costmodel}
-
-We address the limitations of data availability for the LED manufacturing cost by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
-In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
-
-We identified two existing cost models for LED manufacturing in the prior literature.
-The first is the LED Cosim model developed by the technology consultancy System Plus Consulting \cite{cosim2013} \cite{cosim2019}.
-Due to prohibitive licensing costs, this model could not be used in our study.
-The second model is the publicly available 2012 LEDCOM cost model prepared for the U.S.
-Department of Energy by Stephen Bland  \cite{ledcomv2}.
-It was set up to capture the manufacturing process for phosphor-converted LEDs of different power levels.
-However, given the remarkable changes in the performance, cost and architecture of LEDs since 2012, as shown in Figures 1-3, this model cannot fully capture the recent advances in LED technology, necessitating a significant modification to bring it up to date with the current state-of-the-art in LED manufacturing.
-
-Given this situation, we developed our own bottom-up LED manufacturing cost model.
-The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
-The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
-Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips.
-The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
-In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
-These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
-
-The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
-Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
-We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
-It instead assumes an hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
-It also assumes the use of the most up-to-date equipment for that year.
-Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
-An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
-Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
 
@@
 \begin{table}[h!]
-    \small
+    \footnotesize
     \centering

### 2023-04-17 2_SSL_EST/article/article_supplementary.tex
moved sections around

@@
 
+\subsection{Systematic Literature Review}
+
+We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
+This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in Section \ref{subsec:metrics}; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
+
+Relevant sources for the review were found in an iterative search process that involved two components.
+The first was the search in specialized patent and publication databases as well as company websites.
+The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the Section \ref{sec:prev_lit} and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+We also relied on backward citations for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
+
+\subsection{Semi-Structured Interviews}
+
+To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews with eleven eminent experts from academia, industry and the public research sector.
+Experts were selected based on their engagement in  different sub-fields of LED research and manufacturing, as well as the recommendations from other interviewed experts, in essence expanding the list of experts that emerged from the initial literature review.
+All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
+A summary of the background of interviewed experts is provided in Table SI 3 in Section 6 of the Supplementary Information.
+
+The primary, structured part of the interviews explored which innovations were deemed most relevant to the evolution of device performance, consumer experience and manufacturing cost of LED packages.
+Thereafter, interviewees were asked to consider the extent to which these innovations may have originated outside of their respective field of expertise and the LED industry more broadly—i.e., which of the innovations may be considered spillovers.
+The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
+Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in Section \ref{sec:costmodel}) and verify device performance data.
+
 \subsection{Manufacturing Cost Model}
 
+We address the limitations of data availability for the LED manufacturing cost by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
+In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
+
+We identified two existing cost models for LED manufacturing in the prior literature.
+The first is the LED Cosim model developed by the technology consultancy System Plus Consulting \cite{cosim2013} \cite{cosim2019}.
+Due to prohibitive licensing costs, this model could not be used in our study.
+The second model is the publicly available 2012 LEDCOM cost model prepared for the U.S.
+Department of Energy by Stephen Bland  \cite{ledcomv2}.
+It was set up to capture the manufacturing process for phosphor-converted LEDs of different power levels.
+However, given the remarkable changes in the performance, cost and architecture of LEDs since 2012, as shown in Figures 1-3, this model cannot fully capture the recent advances in LED technology, necessitating a significant modification to bring it up to date with the current state-of-the-art in LED manufacturing.
+
+Given this situation, we developed our own bottom-up LED manufacturing cost model.
+The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
+Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips.
+The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
+In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
+These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
+
+The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
+Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
+We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
+It instead assumes an hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
+It also assumes the use of the most up-to-date equipment for that year.
+Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
+An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
+Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
+
 \subsubsection{Structure of the Model}
 
-\begin{figure}[h!]
+\begin{figure}[H]
     \centering
@@
 
-\begin{figure*}[h!]
+\begin{figure*}[H]
  \centering
@@
 \clearpage
-\section{Previous Literature}
-\label{sec:prev_lit}
-
-The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015}\cite{Nakamura2015}.
-The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006}\cite{krames2007led}\cite{laubsch2009high}\cite{hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
-
-Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status}\cite{Phillips2007}\cite{Bierhuizen2007}\cite{Nakamura2013}\cite{feezell2018invention}\cite{Taki2019}.
-Additional studies provide more detail regarding the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005}\cite{Shchekin2006}\cite{lumi2015lumi}\cite{Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004}\cite{Baur2009}\cite{laubsch2009high}\cite{hahn2014development}.
-However, these studies typically include limited information on the macro-level chip design and cover disparate aspects of the technology over different time periods.
-Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos}\cite{cha2019semiconductor} and Soraa\cite{cich2017high}.
-
-We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure \ref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
-Despite the amount of literature published on the topic of LED history, we are not aware of a prior publication that comprehensively and consistently aggregates and analyzes known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
-In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
-This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012}\cite{Creutzig2018}.
-
-\clearpage
 \section{Complete List of Sources for Figures}

### 2023-04-20 2_SSL_EST/article/article_draft_est.tex
more annoying edits

--- 2023-04-11 2_SSL_EST/article/article_draft_est.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
 
@@
 \author{Michael Weinold}
-\email{michael.weinold@alumni.ethz.ch}
-\affiliation[ETHZ]{ETH Zurich, Switzerland}
+\email{mweinold@ethz.ch}
+\affiliation[PSI]{Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland}
+\alsoaffiliation[ETHZ]
+{Chair for Entrepreneurial Risk, ETH Zurich, Zurich, Switzerland}
 \alsoaffiliation[UCAM]
-{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, United Kingdom}
-\altaffiliation{Current address: Technology Assessment Group, Paul Scherrer Institute, Switzerland}
+{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom}
 
@@
 \affiliation[UCAM]
-{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, United Kingdom}
+{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom}
 
 \author{Laura Diaz Anadon}
-\affiliation[UCAM]{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, United Kingdom}
+\affiliation[UCAM]{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom}
 \alsoaffiliation[UHarv]
-{Belfer Center for Science and International Affairs, Harvard University, USA}
+{Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{tocentry}
-
-Some journals require a graphical entry for the Table of Contents.
-This should be laid out ``print ready'' so that the sizing of the
-text is correct.
-
-Inside the \texttt{tocentry} environment, the font used is Helvetica
-8\,pt, as required by \emph{Journal of the American Chemical
-Society}.
-
-The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
-permitted for  \emph{Journal of the American Chemical Society}
-graphical table of content entries. The box will not resize if the
-content is too big: instead it will overflow the edge of the box.
-
-This box and the associated title will always be printed on a
-separate page at the end of the document.
-
-\end{tocentry}
 
@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed.
-This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021} \cite{Ziegler2021}.
+This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021,Ziegler2021}.
 
-There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global}\cite{national2016power}\cite{iea2020perspectives}.
+There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global,national2016power,iea2020perspectives}.
 The role of these factors at different stages of the innovation life cycle \cite{ grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}.
@@
 
-The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies}\cite{Anadon2016}.
+The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies,Anadon2016}.
 Among these drivers, the role of external  knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area  \cite{Stephan2021}.
-While the exact definition of spillovers in the literature depends on the context and the research question \cite{Liu2003}\cite{Nemet2012}, we follow Stephan et al.
+While the exact definition of spillovers in the literature depends on the context and the research question \cite{Liu2003,Nemet2012}, we follow Stephan et al.
 \cite{Stephan2021} and consider knowledge to be external - thus a spillover -  if it has been developed for application in other technologies, sectors, or scientific disciplines.
-There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016}\cite{Stephan2021}\cite{Sun2021}\cite{kolesnikov2022technology} the future evolution of technologies.
+There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016,Stephan2021,Sun2021,kolesnikov2022technology} the future evolution of technologies.
 
-Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently   accounts for 15-19\% of global electricity consumption \cite{Zissis2016}\cite{doe_electricity}.
+Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently   accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}.
 It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.
@@
 Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices.
-Figure \ref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline during the period of from 2008-2020 of 27.3\%, in line with previous estimates \cite{Gerke2020}.
+Figure \ref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.
 
-\begin{figure}[h]
-\centering
-  \includegraphics[height=6.5cm]{2_SSL_EST/article/figures/cost_lamp_small.pdf}
-  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
-Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
-Department of Energy (DOE) \cite{council2013assessment}.
-Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs.
-Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications.
-Data on CFL sales prices collected from \cite{eger2018origin} as well as various consumer watchdog databases and publications.
-Incandescent light bulb price is assumed constant based on the average in the covered time period.
-See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
-  \label{fgr:cost_lamp_small}
-\end{figure}
-
-These  dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies  \cite{weinold2020long}\cite{Mills2014}\cite{Stegmaier2021}\cite{grubb2021new}.
-As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
-Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}.
-For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
-LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical}\cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
-
-Despite this impressive history, the sources of LED innovation have not received as much attention from researchers as innovation in supply-side  energy technologies, such as solar photovoltaics  \cite{kavlak2018evaluating} or wind energy \cite{qiu2012price}\cite{jennings2020policy}, or in lithium-ion batteries for transportation \cite{Ziegler2021}\cite{Stephan2021}.
-To the best of our knowledge, no study has comprehensively discussed the sources or extent of progress across various metrics of LED cost or performance since the introduction of first commercial white LED products.
-Understanding the extent to which individual innovations and knowledge spillovers  contributed to improvements in LED technology, how this effect compares to other sources of improvements such as economies of scale, and how these innovations occurred (i.e., by what mechanisms and actors) will provide valuable lessons both for other demand-side technologies and overall for accelerating clean energy innovation.
-
-To address these questions, in this paper we identify a set of metrics suitable for tracking the historical progress of LED lighting technology.
-We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
-Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry.
-This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.
-Data on performance improvements associated with individual innovations identified in patents, scientific publications, industry reports, and interviews is then augmented  by our calculations of device efficiency, including contributions of specific spillovers to the overall LED efficiency (see Supplementary Information for more detail).
-We further calculate LED manufacturing costs using a bottom-up cost model with process-step resolution that we developed.
-Finally, we discuss what our observations may mean for our understanding of the innovation process in innovation process in LED technology and for broader policy and industry efforts to accelerate clean energy innovation.
-
-\begin{figure}[h]
+\begin{figure}[h!]
  \centering
@@
 Data points indicate best performers by year of market introduction.
-Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye.
+Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Information, Section 1 for details).
 Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data.
@@
 \cite{azevedo2009transition}.
-See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
+See Supplementary Information, Section 4 for the full list of sources and references.}
  \label{fgr:history_efficacy}
 \end{figure}
+
+\begin{figure}[h!]
+\centering
+  \includegraphics[height=6.5cm]{2_SSL_EST/article/figures/cost_lamp_small.pdf}
+  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
+Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
+Department of Energy (DOE) \cite{council2013assessment}.
+Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs.
+Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications.
+Data on CFL sales prices collected from Eger \& Ehlhardt \cite{eger2018origin} as well as various consumer watchdog databases and publications.
+Incandescent light bulb price is assumed constant based on the average in the covered time period.
+See Supplementary Information, Section 4, for the full list of sources and references.}
+  \label{fgr:cost_lamp_small}
+\end{figure}
+
+These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies  \cite{weinold2020long,Mills2014,Stegmaier2021,grubb2021new}.
+As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
+Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}.
+For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
+LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical,Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
+
+Despite this impressive history, the sources of LED innovation have not received as much attention from researchers as innovation in supply-side  energy technologies, such as solar photovoltaics  \cite{kavlak2018evaluating} or wind energy \cite{qiu2012price,jennings2020policy}, or in lithium-ion batteries for transportation \cite{Ziegler2021,Stephan2021}.
+To the best of our knowledge, no study has comprehensively discussed the sources or extent of progress across various metrics of LED cost or performance since the introduction of first commercial white LED products.
+Understanding the extent to which individual innovations and knowledge spillovers  contributed to improvements in LED technology, how this effect compares to other sources of improvements such as economies of scale, and how these innovations occurred (i.e., by what mechanisms and actors) will provide valuable lessons both for other demand-side technologies and overall for accelerating clean energy innovation.
+
+To address these questions, in this paper we identify a set of metrics suitable for tracking the historical progress of LED lighting technology.
+We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
+Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry.
+This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.
+Data on performance improvements associated with individual innovations identified in patents, scientific publications, industry reports, and interviews is then augmented  by our calculations of device efficiency, including contributions of specific spillovers to the overall LED efficiency (see Supplementary Information for more detail).
+We further calculate LED manufacturing costs using a bottom-up cost model with process-step resolution that we developed.
+Finally, we discuss what our observations may mean for our understanding of the innovation process in innovation process in LED technology and for broader policy and industry efforts to accelerate clean energy innovation.
 
@@
 Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Information, Section 5 (Complete List of Sources for Figures), for the full list of sources and references.}
+See Supplementary Information, Section 4, for the full list of sources and references.}
  \label{fgr:chip_architecture_overview}
 \end{figure}
+
+\section{Previous Literature}
+\label{sec:prev_lit}
+
+The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015,Nakamura2015}.
+The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006,krames2007led,laubsch2009high,hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
+
+Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status,Phillips2007,Bierhuizen2007,Nakamura2013,feezell2018invention,Taki2019}.
+Additional studies provide more detail regarding the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005,Shchekin2006,lumi2015lumi,Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004,Baur2009,laubsch2009high,hahn2014development}.
+However, these studies typically include limited information on the macro-level chip design and cover disparate aspects of the technology over different time periods.
+Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos,cha2019semiconductor} and Soraa\cite{cich2017high}.
+
+We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure \ref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+Despite the amount of literature published on the topic of LED history, we are not aware of a prior publication that comprehensively and consistently aggregates and analyzes known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
+In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2020novel,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
+This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012,Creutzig2018}.
 
@@
 Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}.
-However, metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition}\cite{cho2017white}.
+However, metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000,Sandahl2006,CAIRD2008,murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
 Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
@@
 In 2000, the highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
-Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case}\cite{haitz2011solid}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
 Today, however, the devices with the highest performance yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
@@
 Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
-Specifically,  the stagnating levels of luminous flux in highest performing devices does not capture a major research focus, which is improving electrical efficiency at constant brightness.
+Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
 
 We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology.
-
-\subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
-
-Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
-\cite{tsao2010solid}} (VfE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
-These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
-It is defined as
-
-\begin{equation}
-    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
-\end{equation}
-
-where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Light Extraction Efficiency $\eta_{LE}$}
-
-Device light-extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
-
-\begin{equation}
-    \eta_{LE}=\frac{\text{\# of photons out}}{\text{\# of photons created}} = \frac{P_{opt}}{P_{int}}
-\end{equation}
-
-where $P_{opt}$ is the optical power of the device and $P_{int}$ is the internal power of the device\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Internal Quantum Efficiency $\eta_{IQ}$}
-
-Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
-\begin{equation}
-\label{eqn:ieq-eqe}
-    \eta_{EQ} = \eta_{IQ} \times \eta_{LE}
-\end{equation}
-
-and defined as
-
-\begin{equation}
-    \eta_{IQ} = \frac{\# \text{of photons created}}{\# \text{of electron-hole pairs in}}
-\end{equation}
-
-and depends on the current density of the device\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Droop $\eta_{droop}$}
-
-Efficiency droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
-This is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
-It is defined as
-
-\begin{equation}
-    \eta_{droop} = 1 - \frac{\eta_{IQE}}{\eta_{IQE}(A \rightarrow 0)}
-\end{equation}
-
-where $\eta_{IQE}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
-In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$ as
-
-\begin{equation}
-\label{eqn:droop}
-    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
-\end{equation}
-
-Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Conversion Efficiency $\eta_{C}$}
-
-Light conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
-These losses are the sum of the Stokes loss and well as scattering/absorption losses.
-It is defined as
-
-\begin{equation}
-    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i}
-\end{equation}
-
-where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
-Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components, i.e.
-\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.
-
-\subsubsection{Spectral Efficiency $\eta_{S}$}
-
-Spectral efficiency (SE) describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
-Photons converted into the infrared or ultraviolet are lost to illumination purposes.
-It is defined as
-
-\begin{equation}
-    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
-\end{equation}
-
-where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
-$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Overall Efficiency ("Lamp Efficiency") $\eta_L$}
-
-Finally, the overall efficiency of a light-emitting diode package is the product of all considered sub-efficiencies
+In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
+The equations defining these sub-efficiencies are provided in the Supplementary Information in Section 1.3.
+The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
 
@@
 
-It thereby describes the electrical and optical losses within the device, as well as the conversion losses in the phosphor layer.
-It further considers the mismatch between the device spectrum and an optimal spectrum, accounting for the wavelength dependent sensitivity of the human eye.
+This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
 
@@
 Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
-Instead, we describe progress in the economics of manufacturing devices by using yielded cost as the primary metric.
 
-The cost model we developed and describe in Section \ref{sec:costmodel} relies on a cumulative approach to yielded cost\cite{becker2001use}.
+We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
+In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
+
+We describe the details of our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model's limitations, in Supplementary Information section 2.4.
+The model structure is generally based on the 2012 LEDCOM cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
+
+The cost model we developed and describe in Section 2.4 of the Supplementary Information relies on a cumulative approach to yielded cost\cite{becker2001use}.
 In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
@@
 Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
-Additional information on details of the computation of cost metrics and the different manufacturing steps is presented in Section 2 of the Supplementary Information.
 
+\section{Methods and Data Collection}
+\label{sec:methods}
+
+The evolution of LED device architecture and performance as well as the progress in understanding the underlying physical phenomena are well covered in the scholarly literature and patents.
+However, information provided in such sources is insufficient for our goals on at least three accounts: First, existing work focuses only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
+Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
+Second, not all relevant innovations are patented \cite{Pakes_1980,Fontana_2013}.
+In the case of LED patents in particular, our interviews with industry experts suggest that the propensity to patent is the highest for knowledge related to macroscopic device architecture and chemical composition of phosphors, and the lowest for knowledge related to manufacturing process improvements and microscopic chip architecture that is difficult to reconstruct by reverse engineering.
+This means that relying only on patent literature would bias results by unduly emphasizing some focus areas and de-emphasizing others.
+Third, scientific publications and patents typically focus on experimental devices, rather than commercial products.
+While new LED features, designs and manufacturing methods reported in these sources can potentially result in significant performance gains or cost reductions, it is difficult to ascertain if these improvements have since been adopted in industry.
+Furthermore, information on LED manufacturing cost and the effect of process improvements on the total cost is highly proprietary.
+Estimates are occasionally reported in the scientific literature and company publications, but these often do not disclose which parts of the manufacturing process are responsible for the largest contribution to the overall cost, or which improvements led to cost reductions.
+
+To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis, the details of which are provided in Section 2 of the Supplementary Information document.
+Specifically, we combine information obtained from a systematic review of the primary scientific literature, device datasheets, relevant patents, and industry publications (SI Section 2.1) with information gained from semi-structured interviews with experts from academia and industry (SI Section 2.2), bottom-up manufacturing cost modelling (SI Section 2.3), and our own computations of device sub-efficiencies (SI Section 2.4).
+We then use this information to track the historical progress in white LED technology over time across the three groups of metrics identified above in the 'Metrics' section and identify its sources in innovation and technology spillovers.
+
+\clearpage
 \section{Results}
@@
 
-The historical development of a selection of the sub-efficiencies introduced in Section \ref{sec:device_performance_metrics} is shown in Figure \ref{fgr:subeff}.
-Using this data, together with data on the remaining sub-efficiencies not shown in this figure, we calculated the overall LED efficiency in for four years: 2003, 2010, 2016 and 2020.
-Figure \ref{fgr:waterfall} shows how improvements in sub-efficiencies led to improvements in overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.7\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
+The historical development of the sub-efficiencies introduced in Section Device Performance Metrics is shown in Figure SI11 in the Supplementary Information.
+Using this data, we calculated the overall LED efficiency for four years: 2003, 2010, 2016 and 2020.
+The waterfall diagrams in Figure \ref{fgr:waterfall} show how improvements in sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.7\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
 As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
@@
 Figure \ref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
-The list of technology spillovers identified in our study is provided in Table \ref{tab:spillovers} and discussed in Section \ref{subsec:spillovers}.
-The complete list of all LED innovations is provided in Table SI 2 in Section 4 of the Supplementary Information.
-Through the index decomposition analysis described in Section \ref{subsec:metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
+The full list of identified LED innovations considered in our study is provided in Table SI2 in Section 3 of the Supplementary Information document.
+The list of corresponding technology spillovers is provided in Table \ref{tab:spillovers} in the Discussion section below.
+Through the index decomposition analysis described in Section 2.4 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
@@
  \centering
- \includegraphics[width=17cm]{2_SSL_EST/article/figures/subefficiencies_progress.pdf}
- \caption{Historical improvements in light-emitting diode technology.
-Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
-Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
-\textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
-The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
-Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
-\textbf{Panel A2:} Efficiency droop at the test current of I=350mA.
-\textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
-Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}.
-Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
-“Not reported” denotes data points where the IQE measurement technique was not reported.
-Whiskers indicate reported ranges, where applicable.
-\textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
-The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given.
-Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
-- electrode; text.
-- textured.
-Source: own synthesis of data from the full list of sources provided in the Supplementary Information, Section 5.
-Data on average device performance adapted from U.S.
-Department of Energy (DOE) Reports\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report\cite{beale_leds_2015}.}
- \label{fgr:subeff}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2010.pdf}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2016.pdf}
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2020.pdf}
+ \captionsetup{font=footnotesize}
+ \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
+Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section 1.3 of the Supplementary Information.
+Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts).
+Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
+The corresponding LED architectures and associated down-conversion phosphors are shown for reference: 2003 - flip-chip with YGAG phosphor; 2010 - flip-chip with 258 phosphor; 2016 - flip-chip with PFS phosphor; 2020 - flip-chip with SALON phosphor.
+A more complete overview of LED chip architectures is provided in Figure \ref{fgr:chip_architecture_overview}.
+Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and invention history provided in Section 3 in the Supplementary Information.
+Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003, $\eta_L = 12.7\%$ in 2010, $\eta_L = 32.6\%$ in 2016 and $\eta_L = 38.8\%$ in 2010.
+Abbreviations: Scat.
+= Scattering.
+Source (Efficiency): own elaboration based on data in Figure SI11.
+Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
+ \label{fgr:waterfall}
 \end{figure}
@@
 $\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997.
-Note that indium tin oxide (ITO) affects different sub-efficiencies depending on the chip architecture in question.
-For instance, in modern flip-chip architectures, light-extraction efficiency no longer depends on ITO, as can be seen from Figure \ref{fgr:chip_architecture_overview}.
+Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures light extraction efficiency no longer depends on ITO, see Figure \ref{fgr:chip_architecture_overview}.
 Physical limits on sub-efficiencies are indicated by black horizontal lines.
@@
 Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}.
-Source: own elaboration based on data in Figure \ref{fgr:waterfall} and Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.}
+Source: own elaboration based on data represented in Figure \ref{fgr:waterfall} and Figure SI11 in the Supplementary Information, as detailed in Section Methods and Data Collection.}
  \label{fgr:breakthroughs_efficiency}
-\end{figure}
-
-\begin{figure}[h!]
- \centering
- \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2003.pdf}
- \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2010.pdf}
- \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2016.pdf}
- \includegraphics[width=17cm]{2_SSL_EST/article/figures/waterfall_performance_2020.pdf}
- \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
-Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section \ref{sec:device_performance_metrics}.
-Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts).
-Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
-The corresponding LED architectures and associated down-conversion phosphors are shown for reference: 2003 - flip-chip with YGAG phosphor; 2010 - flip-chip with 258 phosphor; 2016 - flip-chip with PFS phosphor; 2020 - flip-chip with SALON phosphor.
-A more complete overview of LED chip architectures is provided in Figure \ref{fgr:chip_architecture_overview}.
-Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and development history collected in Section 4 in the Supplementary Information.
-Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003, $\eta_L = 12.7\%$ in 2010, $\eta_L = 32.6\%$ in 2016 and $\eta_L = 38.8\%$ in 2010.
-Abbreviations: Scat.
-= Scattering.
-Source (Efficiency): own elaboration based on data in Figure \ref{fgr:subeff}, as detailed in Section \ref{sec:methods}.
-Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
- \label{fgr:waterfall}
 \end{figure}
@@
 We were thus able to establish the links between all major improvements in the two consumer experience metrics identified in this study and individual LED inventions associated either with phosphors or quantum dots.
-The list of these inventions is provided in Table \ref{tab:spillovers}, while details of the device spectra are provided and discussed in Section 5 in the Supplementary Information.
+The list of these inventions is provided in Table \ref{tab:spillovers}, while the detailed descriptions of these inventions and corresponding device spectra are provided in Section 3 in the Supplementary Information.
 
-Notably, from detailed descriptions of the history of inventions in this list, which we provide in Section 5 of the Supplementary Information, we find that only a single invention related to LED consumer experience improvements was originally developed specifically for application in solid-state lighting: the 2016 SALON phosphor compound \cite{seibald2019phosphor}\cite{Hoerder2019}.
+Notably, from detailed descriptions of the history of inventions in this list, which we provide in Section 3.2.2 of the Supplementary Information, we find that only a single invention related to LED consumer experience improvements was originally developed specifically for application in solid-state lighting: the 2016 SALON phosphor compound \cite{seibald2019phosphor,Hoerder2019}.
 All other inventions in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL.
-We discuss the details of the corresponding technology spillovers in Section \ref{subsec:spillovers} below.
+We discuss the details of the corresponding technology spillovers in the Discussion section below.
 
 \begin{table}[h!]
-    \small
+    \footnotesize
     \centering
@@
     \caption{LED down-conversion materials (phosphors and quantum dots) related to improvements in consumer experience metrics identified in this study.
-The \textit{year} column represents the earliest reported application of invention in white LEDs.
+The \textit{Year} column represents the earliest reported application of invention in white LEDs.
 These differ from the years used in Figure \ref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for representative LED products that relied on those materials.
@@
 Data points show the color temperature and colour rendering index performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year.
-Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 11 in section 5 of the Supplementary Information.
+Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 11 in Section 3 of the Supplementary Information.
 Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison.
@@
 
-\begin{table}[h]
-    \small
+\begin{figure*}[h!]
  \centering
+ \includegraphics[width=17.5cm]{2_SSL_EST/article/figures/costmodel_results_years.pdf}
+ \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+location.
+Consecutive steps shown on the x-axis.
+Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
+Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
+Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
+For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information.
+Abbreviations: Litho – Lithographic Process, Insp.
+– Inspection, Depn.
+– Deposition, CMP - Chemical-Mechanical Planarization}
+ \label{fgr:costmodel}
+\end{figure*}
+
+\clearpage
+\begin{table}[h!]
+    \tiny
+    \centering
+    \caption{Technology spillovers involved in white LED technology innovations identified in this study.}
     \begin{tabularx}{\textwidth}{|l|l|l|X|X|X|l|X|}
@@
         \textbf{Disc.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
-        1926 & 1994 & 1996 & LED phosphors & Use of phosphors for light down conversion in LEDs & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric}\cite{shimizu1994sheet}\cite{cho2017white} & Enabled light down conversion in LEDs \\ \hline
-        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new}\cite{bando1996}\cite{bando1998development}\cite{shimizu1999light}\cite{cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
-        1967 & 1996 & <2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical}\cite{bando1998development}\cite{shimizu1999light}\cite{Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
-        1982 & 1996 & <2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction}\cite{krames1998ordered}\cite{feezell2018invention}\cite{Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare Figure \ref{fig:chip_architecture_overview})\\ \hline
-        1971 & 1999 & <2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf}\cite{fraser1972highly}\cite{margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare Figure \ref{fig:chip_architecture_overview}) \\ \hline
-        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997}\cite{mueller2004phosphor}\cite{MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
-        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo}\cite{simmonsfinal}\cite{ledprof_nexxusqd}\cite{bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
-        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & Chemistry (S), Materials science (S) &\cite{paulusz1973efficient}\cite{radkov2009red}\cite{Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
-        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New}\cite{schmidt2013new}\cite{Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
+        1926 & 1994 & 1996 & LED phosphors & Use of phosphors for light down conversion in LEDs & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
+        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
+        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
+        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare Figure \ref{fgr:chip_architecture_overview})\\ \hline
+        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare Figure \ref{fgr:chip_architecture_overview}) \\ \hline
+        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
+        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo,simmonsfinal,ledprof_nexxusqd,bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
+        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & Chemistry (S), Materials science (S) &\cite{paulusz1973efficient,radkov2009red,Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
+        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
     \end{tabularx}
-    \caption{Technology spillovers involved in white LED technology innovations identified in this study.
-Note: LED innovations are ordered by the year in which a technology spillover into LED occurred, provided in the Spillover to LED column.
-Initial discovery is the year of the earliest identified discovery of the original idea or invention outside the LED domain.
-Commercial application is the year of the first recorded application of that idea or invention in a commercial LED product.
+    \begin{tabular}{@{}p{\linewidth}@{}}
+        Note: Disc.
+- Year of initial discovery; S/O - Year of spillover to LED; Comm.
+- Year of commercial application; Ref.
+- References.
+LED innovations are ordered by the year in which a technology spillover into LED occurred, provided in the S/O column.
+The year of initial discovery is the year of the earliest identified discovery of the original idea or invention outside the LED domain.
+The year of commercial application is the year of the first recorded application of that idea or invention in a commercial LED product.
 Origin column represents knowledge domains in which spillovers initially emerged, where (S) denotes a scientific discipline and (T) is an area of technology.
-References column lists literature sources for the represented innovations and spillovers.
-Area of improvement represents the impact of spillovers on different aspects of white LED technology, e.g., improvements in particular sub-efficiencies.
-Abbreviations: Disc.
-- Year of Discovery, S/O - Year of Spillover to LED, Comm.
-- Year of Commercial Application., Ref.
-- References.}
+Ref.
+column lists literature sources for the represented innovations and spillovers.
+Area of Improvement column represents the impact of spillovers on different aspects of white LED technology, e.g., improvements in particular sub-efficiencies.
+    \end{tabular}
     \label{tab:spillovers}
@@
 Spectral efficiency also remains at $\sim17\%$ from its physical limit, showing potential for further improvements.
-Research on improving LED performance across efficiency loss channels continues\cite{cho2017white}\cite{Weisbuch2020}.
+Research on improving LED performance across efficiency loss channels continues\cite{cho2017white,Weisbuch2020}.
 
@@
 The first commercial white LEDs produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{bando1998development}.
-After a series of breakthrough innovations in phosphors shown in Figure \ref{fgr:consumer_experience} and described in Section 4 of the Supplementary Information, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
+After a series of innovations in phosphors shown in Figure \ref{fgr:consumer_experience} and described in Section 3 of the Supplementary Information, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
 
 Overall, we find that LED technology innovations made crucial contributions to the progress in both LED performance and consumer experience metrics across the entire period covered by our study.
-Our interviews have also revealed an important role of incremental process improvements and learning-by-doing\cite{WRIGHT_1936}\cite{Arrow_1962} in the progress in LED efficiency.
+Our interviews have also revealed an important role of incremental process improvements and learning-by-doing\cite{WRIGHT_1936,Arrow_1962} in the progress in LED efficiency.
 However, further research is needed to quantify the contribution of learning by doing to this progress.
@@
 Our manufacturing cost model shows a 95.5\% decrease in the cost of producing white classic-chip LEDs (from 1.11\$(2020) to 0.05\$(2020)) between 2003 and 2020, driven mostly by increases in the wafer size and yields across different manufacturing steps over time.
-In contrast with efficiency improvements, these learning by doing, economies of scale and improvements were facilitated by both improvements in the performance as well as policies creating and growing market demand, such as incandescent light bans and subsidies for LEDs.
+In contrast with performance improvements, these cost reductions were thus a result of economies of scale and learning by doing,  which were in turn likely facilitated by policies creating and growing market demand for LED lighting, such as incandescent light bulb bans and subsidies for LEDs.
 
 Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of improvements in LED efficiency and consumer experience metrics, highlights the critical role played by a deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.
-Specifically, deep physical understanding of LED device efficiency loss channels enabled important innovations in LEDs that increased several sub-efficiencies in LEDs and will continue, as expected by eminent experts in the field \cite{Weisbuch2020}.
+Specifically, deep physical understanding of LED device efficiency loss channels enabled important innovations in LEDs that increased several sub-efficiencies in LEDs and will continue to do so, as expected by eminent experts in the field \cite{Weisbuch2020}.
 This suggests that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology.
 These knowledge spillovers can be enabled or even accelerated by knowledge exchange events and long-term partnerships between academia and industry, dedicated mission-driven public R\&D funding, and a freedom of search in academia.
-This further reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles}\cite{narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
+This further reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles,narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
 
@@
 
-This will usually read something like: ``Experimental procedures and
-characterization data for all new compounds. The class will
-automatically add a sentence pointing to the information on-line:
+Additional context and governing equations of metrics used to quantify progress in solid-state lighting, including the constituent sub-efficiencies of overall lamp efficiency.
+Detailed description of the multi-method approach used, including the systematic literature review, semic-structured interviews, the manufacturing cost model and performance calculations.
+Methodological details on dis-aggregation of the contribution of variables to overall lamp efficiency.
+Additional results, including a breakdown of manufacturing cost and a detailed description of the technology spillovers in phosphor materials for light down-conversion.
+Complete list of sources for Figures \ref{fgr:chip_architecture_overview} and \ref{fgr:history_efficacy}.
 

### 2023-04-20 2_SSL_EST/article/article_supplementary.tex

@@
         \textit{\textsc{Supplementary Information}} for: \\
-        M. Weinold, S. Kolesnikov, L.D. Anadon \\
-        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Energy and Environmental Science} (2023)
+        \newline 
+        M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
+        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Environmental Science \& Technology} (2023) \\
+        \newline
+        $^1$ Technology Assessment Group, Paul Scherrer Institute, Switzerland \\
+        $^2$ ETH Zurich, Zurich, Switzerland \\
+        $^3$ C-EENRG, Dept. of Land Economy, University of Cambridge, UK \\
+        $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
     }
@@
 Figure 1 in the main publication uses luminous efficacy of radiation as the primary metric to describe progress in lighting technologies.
-Care must be taken not to confuse this metric with luminous efficacy of source, which is used in \ref{fig:phosphor_spectrum}.
+Care must be taken not to confuse this metric with luminous efficacy of source, which is used in \cref{fig:phosphor_spectrum}.
 
@@
 
+This subsection provides definitions and equations for the LED device sub-efficiencies introduced in the Metrics section in the main text.
+
 \subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
@@
 
+This subsection provides the details for the main methods and data sources used in our mixed-methods research approach, including the systematic literature review, semi-structured interviews with eminent experts, manufacturing cost modelling, and performance metrics calculations.
+
 \subsection{Systematic Literature Review}
@@
 We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
-This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in Section \ref{subsec:metrics}; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
+This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in \cref{subsec:metrics}; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
 
@@
 The first was the search in specialized patent and publication databases as well as company websites.
-The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the Section \ref{sec:prev_lit} and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the \cref{sec:prev_lit} and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
 We also relied on backward citations for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
@@
 The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
-Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in Section \ref{sec:costmodel}) and verify device performance data.
+Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in \cref{sec:costmodel}) and verify device performance data.
+
+A summary of the background of interviewed experts is provided in \cref{tab:interviews}.
+
+\begin{table}[H]
+\small
+    \centering
+    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
+    \vspace{5mm}
+    \begin{tabular}{|l|l|l|l|l|}
+    \hline
+        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
+        1 & Academia & Sr. researcher & UK & Epitaxy \\ \hline
+        2 & Industry & Consultant, former sr. researcher & USA & Device architecture \\ \hline
+        3 & Industry & Consultant, former head of R\&D & Germany & Epitaxy \\ \hline
+        4 & Academia & Professor & Austria & Phosphors \\ \hline
+        5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
+        6 & consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
+        7 & Academia & Professor & Germany & Phosphors \\ \hline
+        8 & government & R\&D manager & USA & Device architecture \\ \hline
+        9 & consulting & Consultant & USA & Device applications \\ \hline
+        10 & Academia & Professor & France & Device physics \\ \hline
+        11 & Industry & Sr scientist, former head of R\&D & USA & Device architecture \\ \hline
+        12 & Industry & Principal scientist & Germany & Phosphors \\ \hline
+        13 & Industry & Former head of R\&D & USA & Phosphors \\ \hline
+    \end{tabular}
+    \label{tab:interviews}
+\end{table}
+
+\subsection{Performance Metrics Calculations}
+\label{subsec:metrics}
+
+The contribution of individual technology innovations and spillovers to the progress in overall device efficiency over time is estimated by index decomposition analysis.
+Mathematically, this involves breaking down a chosen performance indicator into its constituent components, each representing a specific factor that contributes to the change in the indicator \cite{Ang1997}.
+Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
+It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
+
+According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
+
+\begin{align}
+    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
+    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
+\end{align}
+
+where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
+These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:
+
+\begin{equation}
+    \Delta a + \Delta b  = \Delta F
+\end{equation}
+
+To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered.
+However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance.
+For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies where possible and necessary.
+Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications.
+In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra.
+This approach allowed us to  quantify the improvements related to phosphor development in LEDs.
 
@@
 
-We address the limitations of data availability for the LED manufacturing cost by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
-In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
-
-We identified two existing cost models for LED manufacturing in the prior literature.
-The first is the LED Cosim model developed by the technology consultancy System Plus Consulting \cite{cosim2013} \cite{cosim2019}.
-Due to prohibitive licensing costs, this model could not be used in our study.
-The second model is the publicly available 2012 LEDCOM cost model prepared for the U.S.
-Department of Energy by Stephen Bland  \cite{ledcomv2}.
-It was set up to capture the manufacturing process for phosphor-converted LEDs of different power levels.
-However, given the remarkable changes in the performance, cost and architecture of LEDs since 2012, as shown in Figures 1-3, this model cannot fully capture the recent advances in LED technology, necessitating a significant modification to bring it up to date with the current state-of-the-art in LED manufacturing.
-
-Given this situation, we developed our own bottom-up LED manufacturing cost model.
+\subsubsection{Our Approach to Cost Modeling}
+
 The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
@@
 
-\newpage
 \subsubsection{Example of Input Data Considered: Sapphire Wafers}
@@
 \newpage
-\subsubsection{Comparison with Reported Industry Data and DOE Projections}
-
-We conducted a comparison of the LED manufacturing cost structure produced by our cost model with previously reported US DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions (see \cref{fig:costmodel_calibration}).
-We note some differences between the results of our model and the cost structure reported or projected by the DOE.
-For instance, the share of the epitaxy step is consistently larger in the DOE data.
-This can in part be explained by our model relying on state-of-the-art equipment at a virtual US-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
-In addition, the manufacturing lines of the majority of manufacturers is located in Asia.
-We also note that the share of the substrate price in the DOE data is much larger than in our model, which can in part be explained by the overestimation of the actual price of sapphire wafers in earlier projections for 2015-2020.
-Finally, the relative importance of the packaging part of the manufacturing process is very similar in our model and in the DOE results in 2012.
-However, in the DOE projections it significantly decreases by 2020, while in our model it retains and even increases its share of the total cost.
-This trend has been independently confirmed by researchers and industry reports on wafer-level packaging \cite{Lee2011WPL}\cite{Xie2013}\cite{ledsmag2017WLP}, showing better performance of our cost model compared to earlier DOE model projections
-
-\begin{figure}[h]
-	\centering
-    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_calibration.pdf}
-	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel).
-Hatched bars are projections given by US DOE for 2018 and 2020.
-Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
-Sources for bottom panel: own elaboration based on the cost model described in Section \cref{sec:costmodel}.}
-	\label{fig:costmodel_calibration}
-\end{figure}
-
-\newpage
 \subsection{Computation of Contribution of Individual Variables}
@@
 
-\section{LED Innovations}
-
-\subsection{Non-Phosphor-Related}
+\clearpage
+\section{Additional Results}
+
+This section provides additional background and details for the results of our study presented in the Results section in the main text.
+
+\subsection{Historical Progress in Sub-Efficiencies}
+
+The historical progress in the different sub-efficiencies contributing to overall efficiency (lamp efficiency) is shown in \cref{fgr:subeff}. Context for figure panels is provided below:
+
+\textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
+The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
+Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
+\
+
+\textbf{Panel A2:} Efficiency droop at the test current of I=350mA. \
+
+\textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
+Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}.
+Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
+“Not reported” denotes data points where the IQE measurement technique was not reported.
+Whiskers indicate reported ranges, where applicable.
+\
+
+\textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
+The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given.
+Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
+- electrode; text.
+- textured.
+
+\begin{figure}[h!]
+ \centering
+ \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/subefficiencies_progress.pdf}
+ \caption{Historical improvements in light-emitting diode technology.
+Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
+Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
+Source: own synthesis of data from the full list of sources provided in \cref{sec:sources}.
+Data on average device performance adapted from U.S.
+Department of Energy (DOE) Reports \cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report \cite{beale_leds_2015}.}
+ \label{fgr:subeff}
+\end{figure}
+
+\subsection{LED Innovations}
+
+\subsubsection{Non-Phosphor-Related}
 
@@
 
-\begin{table}[H]
+\begin{table}[h!]
     \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.
@@
 
-\subsection{Phosphor-Related}
+\subsubsection{Phosphor-Related}
 \label{sec:innovation_phosphor}
@@
 This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics and discussed in section 5.2, Table 1 and Figures 5-7 in the main article.
-
-\subsubsection{History of Identified Phosphor-Related LED Innovations}
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
 	\centering
@@
 
-\section{List of Interviews}
-
-\begin{table}[h!]
-\small
+\clearpage
+\subsection{Cost Comparison with Reported Industry Data and DOE Projections}
+
+We conducted a comparison of the LED manufacturing cost structure produced by our cost model with previously reported US DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions (see \cref{fig:costmodel_calibration}).
+We note some differences between the results of our model and the cost structure reported or projected by the DOE.
+For instance, the share of the epitaxy step is consistently larger in the DOE data.
+This can in part be explained by our model relying on state-of-the-art equipment at a virtual US-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+In addition, the manufacturing lines of the majority of manufacturers is located in Asia.
+We also note that the share of the substrate price in the DOE data is much larger than in our model, which can in part be explained by the overestimation of the actual price of sapphire wafers in earlier projections for 2015-2020.
+Finally, the relative importance of the packaging part of the manufacturing process is very similar in our model and in the DOE results in 2012.
+However, in the DOE projections it significantly decreases by 2020, while in our model it retains and even increases its share of the total cost.
+This trend has been independently confirmed by researchers and industry reports on wafer-level packaging \cite{Lee2011WPL}\cite{Xie2013}\cite{ledsmag2017WLP}, showing better performance of our cost model compared to earlier DOE model projections
+
+\begin{figure}[h]
 	\centering
-    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
-    \vspace{5mm}
-    \begin{tabular}{|l|l|l|l|l|}
-    \hline
-        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
-        1 & Academia & Sr. researcher & UK & Epitaxy \\ \hline
-        2 & Industry & Consultant, former sr. researcher & USA & Device architecture \\ \hline
-        3 & Industry & Consultant, former head of R\&D & Germany & Epitaxy \\ \hline
-        4 & Academia & Professor & Austria & Phosphors \\ \hline
-        5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
-        6 & consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
-        7 & Academia & Professor & Germany & Phosphors \\ \hline
-        8 & government & R\&D manager & USA & Device architecture \\ \hline
-        9 & consulting & Consultant & USA & Device applications \\ \hline
-        10 & Academia & Professor & France & Device physics \\ \hline
-        11 & Industry & Sr scientist, former head of R\&D & USA & Device architecture \\ \hline
-        12 & Industry & Principal scientist & Germany & Phosphors \\ \hline
-        13 & Industry & Former head of R\&D & USA & Phosphors \\ \hline
-    \end{tabular}
-    \label{tab:interviews}
-\end{table}
-
-\clearpage
-\section{Results}
-
-\subsection{Historical Progress in Sub-Efficiencies}
-
-The historical progress in the different sub-efficiencies contributing to overall efficiency (lamp efficiency) is shown in \cref{fgr:subeff}. Context for figure panels is provided below:
-
-\textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
-The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
-Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
-\
-
-\textbf{Panel A2:} Efficiency droop at the test current of I=350mA. \
-
-\textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
-Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}.
-Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
-“Not reported” denotes data points where the IQE measurement technique was not reported.
-Whiskers indicate reported ranges, where applicable.
-\
-
-\textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
-The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given.
-Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
-- electrode; text.
-- textured.
-
-\begin{figure}[h!]
- \centering
- \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/subefficiencies_progress.pdf}
- \caption{Historical improvements in light-emitting diode technology.
-Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
-Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
-Source: own synthesis of data from the full list of sources provided in \cref{sec:sources}.
-Data on average device performance adapted from U.S.
-Department of Energy (DOE) Reports \cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report \cite{beale_leds_2015}.}
- \label{fgr:subeff}
-\end{figure}
-
-\subsection{Manufacturing Cost Model}
-
-\begin{figure*}[H]
- \centering
- \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/costmodel_results_years.pdf}
- \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
-location.
-Consecutive steps shown on the x-axis.
-Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
-Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
-Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
-For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information.
-Abbreviations: Litho – Lithographic Process, Insp.
-– Inspection, Depn.
-– Deposition, CMP - Chemical-Mechanical Planarization}
- \label{fgr:costmodel}
-\end{figure*}
+    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_calibration.pdf}
+	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel).
+Hatched bars are projections given by US DOE for 2018 and 2020.
+Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
+Sources for bottom panel: own elaboration based on the cost model described in Section \cref{sec:costmodel}.}
+	\label{fig:costmodel_calibration}
+\end{figure}
 

### 2023-04-20 2_SSL_EST/article/article_draft_est.tex

@@
  \centering
- \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/history_efficacy.pdf}
+ \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
  \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.
@@
 \centering
-  \includegraphics[height=6.5cm]{2_SSL_EST/article/figures/cost_lamp_small.pdf}
+  \includegraphics[height=6.5cm]{figures/cost_lamp_small.pdf}
   \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
@@
  \centering
- \includegraphics[height=15cm]{2_SSL_EST/article/figures/chip_architecture_overview.pdf}
+ \includegraphics[height=15cm]{figures/chip_architecture_overview.pdf}
  \caption{Historical evolution of light-emitting diode chip architectures.
@@
  \centering
- \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2003.pdf}
- \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2010.pdf}
- \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2016.pdf}
- \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/waterfall_performance_2020.pdf}
+ \includegraphics[width=15.5cm]{figures/waterfall_performance_2003.pdf}
+ \includegraphics[width=15.5cm]{figures/waterfall_performance_2010.pdf}
+ \includegraphics[width=15.5cm]{figures/waterfall_performance_2016.pdf}
+ \includegraphics[width=15.5cm]{figures/waterfall_performance_2020.pdf}
  \captionsetup{font=footnotesize}
@@
  \centering
- \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/breakthroughs_efficiency.pdf}
+ \includegraphics[width=\textwidth]{figures/breakthroughs_efficiency.pdf}
  \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
@@
  \centering
- \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/breakthroughs_consumer-experience.pdf}
+ \includegraphics[width=\textwidth]{figures/breakthroughs_consumer-experience.pdf}
  \caption{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.
@@
  \centering
- \includegraphics[width=17.5cm]{2_SSL_EST/article/figures/costmodel_results_years.pdf}
+ \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
  \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.

### 2023-04-20 2_SSL_EST/article/article_draft_est.tex
Updates from Overleaf

@@
 
-The cost model we developed and describe in Section 2.4 of the Supplementary Information relies on a cumulative approach to yielded cost\cite{becker2001use}.
-In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
-
-\begin{equation}
-    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
-\end{equation}
-
-This cost metric is cumulative by definition, thus
-
-\begin{equation}
-    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
-\end{equation}
-
-Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
-
 \section{Methods and Data Collection}
@@
 The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
-$\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_Droop$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.
+$\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.
 Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}.
@@
 We were thus able to establish the links between all major improvements in the two consumer experience metrics identified in this study and individual LED inventions associated either with phosphors or quantum dots.
-The list of these inventions is provided in Table \ref{tab:spillovers}, while the detailed descriptions of these inventions and corresponding device spectra are provided in Section 3 in the Supplementary Information.
+The list of these inventions is provided in Table \ref{tab:phosphors}, while the detailed descriptions of these inventions and corresponding device spectra are provided in Section 3 in the Supplementary Information.
 
@@
 In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost resulted from an increase in the wafer size used in manufacturing and higher yields across manufacturing steps.
-This can be seen in the reduction of the contribution of wafer processing steps to  total cost in Figure SI13.
+This can be seen in the reduction of the contribution of wafer processing steps to total cost in \ref{fgr:costmodel}.
 Both were enabled by advances in manufacturing equipment performance and incremental process improvements from learning-by-doing.
@@
 Additional context and governing equations of metrics used to quantify progress in solid-state lighting, including the constituent sub-efficiencies of overall lamp efficiency.
-Detailed description of the multi-method approach used, including the systematic literature review, semic-structured interviews, the manufacturing cost model and performance calculations.
+Detailed description of the multi-method approach used, including the systematic literature review, semi-structured interviews, the manufacturing cost model and performance calculations.
 Methodological details on dis-aggregation of the contribution of variables to overall lamp efficiency.

### 2023-04-20 2_SSL_EST/article/article_supplementary.tex
Updates from Overleaf

@@
         $^2$ ETH Zurich, Zurich, Switzerland \\
-        $^3$ C-EENRG, Dept. of Land Economy, University of Cambridge, UK \\
+        $^3$ CEENRG, Dept. of Land Economy, University of Cambridge, UK \\
         $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
@@
 
+\subsection{Luminous Efficacy of Radiation}
+\label{subsec:ler}
+
 Figure 1 in the main publication uses luminous efficacy of radiation as the primary metric to describe progress in lighting technologies.
 Care must be taken not to confuse this metric with luminous efficacy of source, which is used in \cref{fig:phosphor_spectrum}.
-
-\subsection{Luminous Efficacy of Radiation}
-\label{subsec:ler}
 
@@
 We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
-This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics selected above in \cref{subsec:metrics}; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
+This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics introduced in the main article; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
 
@@
 The first was the search in specialized patent and publication databases as well as company websites.
-The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in the \cref{sec:prev_lit} and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in section Previous Literature in the main article and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
 We also relied on backward citations for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
@@
 All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
-A summary of the background of interviewed experts is provided in Table SI 3 in Section 6 of the Supplementary Information.
+A summary of the background of interviewed experts is provided in \cref{tab:interviews}.
 
@@
 Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in \cref{sec:costmodel}) and verify device performance data.
-
-A summary of the background of interviewed experts is provided in \cref{tab:interviews}.
 
@@
         5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
-        6 & consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
+        6 & Consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
         7 & Academia & Professor & Germany & Phosphors \\ \hline
-        8 & government & R\&D manager & USA & Device architecture \\ \hline
-        9 & consulting & Consultant & USA & Device applications \\ \hline
+        8 & Government & R\&D manager & USA & Device architecture \\ \hline
+        9 & Consulting & Consultant & USA & Device applications \\ \hline
         10 & Academia & Professor & France & Device physics \\ \hline
@@
 Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
+
+The cost model we developed relies on a cumulative approach to yielded cost\cite{becker2001use}.
+In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
+
+\begin{equation}
+    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
+\end{equation}
+
+This cost metric is cumulative by definition, thus
+
+\begin{equation}
+    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
+\end{equation}
+
+Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
 

### 2023-04-20 2_SSL_EST/article/article_draft_est.tex
final edits

@@
 The list of corresponding technology spillovers is provided in Table \ref{tab:spillovers} in the Discussion section below.
-Through the index decomposition analysis described in Section 2.4 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
+Through the index decomposition analysis described in Section 2.5 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
@@
         \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{SI Sec.} \\ \hline
-        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 5.1.1 \\ \hline
-        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 5.1.1 \\ \hline
-        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 5.1.2 \\ \hline
-        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 5.1.4 \\ \hline
-        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 5.1.3 \\ \hline
-        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 5.1.2 \\ \hline
-        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 5.1.2 \\ \hline
+        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.2.2.1 \\ \hline
+        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.2.2.1 \\ \hline
+        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3.2.2.2 \\ \hline
+        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3.2.2.4 \\ \hline
+        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3.2.2.3 \\ \hline
+        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3.2.2.2 \\ \hline
+        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2.2.2 \\ \hline
     \end{tabularx}
@@
 
-Following the framework for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021}, we gathered information from the interviews, historical records, literature reviews and industry publications to analyze the sources, mechanisms, and enablers of the identified spillovers into the white LED technology listed in Table 4.
+Following the framework for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021}, we gathered information from the interviews, historical records, literature reviews and industry publications to analyze the sources, mechanisms, and enablers of the identified spillovers into the white LED technology listed in Table \ref{tab:spillovers}.
 Among the spillover sources, we find that all nine spillovers had origins in basic science disciplines such as various branches of chemistry, materials science, optics and photonics, and solid-state physics.
@@
 Additional results, including a breakdown of manufacturing cost and a detailed description of the technology spillovers in phosphor materials for light down-conversion.
-Complete list of sources for Figures \ref{fgr:chip_architecture_overview} and \ref{fgr:history_efficacy}.
+Complete list of sources for Figures \ref{fgr:history_efficacy} and \ref{fgr:chip_architecture_overview}.
 

### 2023-05-23 2_SSL_NE/archive/EES/article_draft_ees.tex
prepared files for nature energy submission

--- 2023-03-24 2_SSL_EES/article/article_main.tex
@@
 
-\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{The first commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
-Since then, white LEDs have experienced major improvements in performance and efficiency.
-Significant decreases in cost have been driven by a rapid expansion of the market share of LED-based solid-state lighting (SSL), which in turn has driven down costs even further.
-Today, the technology is one of the key solutions in global climate change mitigation efforts.
-However, despite its significance, the extent and sources of the underlying improvements have not been systematically investigated.
-Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies.
-With this aim, we gather and analyse systematic evidence on cost and performance improvements in white LEDs from cost and performance modelling, augmented by a literature review and interviews with experts in industry and academia.
-We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.8\%$ in 2020.
-During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S.
-location using state-of-the-art equipment has dropped from 1.1\$ to 0.05\$ (in 2020 USD) between 2003 and 2020, a 95.5\% decrease.
-We further show that technology spillovers — i.e., knowledge originating in other technologies, sectors, or scientific disciplines — affected all performance dimensions of LEDs, contributing no less than 8.5\% to total efficiency improvements and $\sim$100\% to the improvements in consumer experience metrics identified, thereby playing a key role in the widespread adoption of the technology.
-The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}
+\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{§}
 
@@
 
-\section{Metrics for technological progress in LEDs}
+\section{Metrics for Technological Progress in LEDs}
 

### 2023-05-23 2_SSL_NE/article/article_methods.tex
prepared files for nature energy submission

\documentclass[parskip=full]{article}

% functionality
\input{auxiliary/preamble/1_preamble_functionality_main}

% layout
\input{auxiliary/preamble/2_preamble_layout_report}

% bibliography
\input{auxiliary/preamble/1_preamble_functionality_bibliography}
\addbibresource{bibliography.bib}

%% METADATA %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypersetup{
    pdftitle={Hybrid Life-Cycle Methods},
    pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
}

%% MAIN DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
Banana
\end{document}

### 2023-05-23 2_SSL_NE/article/article_supplementary.tex
prepared files for nature energy submission

--- 2023-04-20 2_SSL_EST/article/article_supplementary.tex
@@
 
+\setcounter{secnumdepth}{4}
+
 % fix strange self-loops, which are PGF/TikZ default
@@
 The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}.
-he evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.
+The evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.
 
@@
 
+\clearpage
 \subsection{LED Innovations}
@@
 
-\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies, as discussed in section 5.1 and shown in Figures 4-6 in the main article.
+\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies.
 Note that phosphor-related LED innovations affecting consumer experience metrics, along with spectral efficiency, are presented separately in Table 1 in the main publication and discussed in detail in \cref{sec:innovation_phosphor}.
@@
 
-This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics and discussed in section 5.2, Table 1 and Figures 5-7 in the main article.
+This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics.
 
@@
 
-Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 7 in the main article).
+Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 6 in the main article).
 An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
@@
 
-The most recent red narrow-band phosphor innovation included in Table 1 and Figure 7 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
+The most recent red narrow-band phosphor innovation included in Table 1 and Figure 6 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
 The first U.S.
@@
 
+\clearpage
 \paragraph{Spectral Data for Identified LED Phosphor Innovations}
 
-\begin{figure}[h!]
+\hspace{1mm}
+
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.95\textwidth]{2_SSL_EST/article/figures/phosphor_spectrum-comparison.pdf}
-	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 7 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year.
-Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 7 in the main article.
+    \includegraphics[width=0.9\textwidth]{2_SSL_EST/article/figures/phosphor_spectrum-comparison.pdf}
+	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 6 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year.
+Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 6 in the main article.
 The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow.
@@
 
-\clearpage
+\newpage
 \subsection{Cost Comparison with Reported Industry Data and DOE Projections}
@@
 Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
-Sources for bottom panel: own elaboration based on the cost model described in Section \cref{sec:costmodel}.}
+Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel}.}
 	\label{fig:costmodel_calibration}
@@
 
-\subsection{Figure 4 (Historical Developments in Device Sub-Efficiencies)}
-
-The sources for Figure 4 in the main article are grouped by sub-efficiencies represented on different panels of the figure:
+\subsection{\cref{fgr:subeff} (Historical Developments in Device Sub-Efficiencies)}
+
+The sources for \cref{fgr:subeff} are grouped by sub-efficiencies represented on different panels of the figure:
 

### 2023-05-29 2_SSL_NE/article/article_main.tex

--- 2023-04-20 2_SSL_EST/article/article_draft_est.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Keywords}
+
+demand-side, solid-state lighting, innovation theory, efficiency, consumer experience
+
+\section{Synopsis}
+
+Understanding the underlying drivers of innovation in LEDs will enable better policy recommendations for a range of demand-side energy technologies, which will play a key role in climate change mitigation.
+
 \section{Introduction}
@@
 The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
-
-\section{Methods and Data Collection}
-\label{sec:methods}
-
-The evolution of LED device architecture and performance as well as the progress in understanding the underlying physical phenomena are well covered in the scholarly literature and patents.
-However, information provided in such sources is insufficient for our goals on at least three accounts: First, existing work focuses only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
-Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
-Second, not all relevant innovations are patented \cite{Pakes_1980,Fontana_2013}.
-In the case of LED patents in particular, our interviews with industry experts suggest that the propensity to patent is the highest for knowledge related to macroscopic device architecture and chemical composition of phosphors, and the lowest for knowledge related to manufacturing process improvements and microscopic chip architecture that is difficult to reconstruct by reverse engineering.
-This means that relying only on patent literature would bias results by unduly emphasizing some focus areas and de-emphasizing others.
-Third, scientific publications and patents typically focus on experimental devices, rather than commercial products.
-While new LED features, designs and manufacturing methods reported in these sources can potentially result in significant performance gains or cost reductions, it is difficult to ascertain if these improvements have since been adopted in industry.
-Furthermore, information on LED manufacturing cost and the effect of process improvements on the total cost is highly proprietary.
-Estimates are occasionally reported in the scientific literature and company publications, but these often do not disclose which parts of the manufacturing process are responsible for the largest contribution to the overall cost, or which improvements led to cost reductions.
-
-To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis, the details of which are provided in Section 2 of the Supplementary Information document.
-Specifically, we combine information obtained from a systematic review of the primary scientific literature, device datasheets, relevant patents, and industry publications (SI Section 2.1) with information gained from semi-structured interviews with experts from academia and industry (SI Section 2.2), bottom-up manufacturing cost modelling (SI Section 2.3), and our own computations of device sub-efficiencies (SI Section 2.4).
-We then use this information to track the historical progress in white LED technology over time across the three groups of metrics identified above in the 'Metrics' section and identify its sources in innovation and technology spillovers.
 

### 2023-05-29 2_SSL_NE/article/article_methods.tex

@@
 \begin{document}
-Banana
+
+\section{Methods and Data Collection}
+\label{sec:methods}
+
+The evolution of LED device architecture and performance as well as the progress in understanding the underlying physical phenomena are well covered in the scholarly literature and patents.
+However, information provided in such sources is insufficient for our goals on at least three accounts: First, existing work focuses only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
+Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
+Second, not all relevant innovations are patented \cite{Pakes_1980,Fontana_2013}.
+In the case of LED patents in particular, our interviews with industry experts suggest that the propensity to patent is the highest for knowledge related to macroscopic device architecture and chemical composition of phosphors, and the lowest for knowledge related to manufacturing process improvements and microscopic chip architecture that is difficult to reconstruct by reverse engineering.
+This means that relying only on patent literature would bias results by unduly emphasizing some focus areas and de-emphasizing others.
+Third, scientific publications and patents typically focus on experimental devices, rather than commercial products.
+While new LED features, designs and manufacturing methods reported in these sources can potentially result in significant performance gains or cost reductions, it is difficult to ascertain if these improvements have since been adopted in industry.
+Furthermore, information on LED manufacturing cost and the effect of process improvements on the total cost is highly proprietary.
+Estimates are occasionally reported in the scientific literature and company publications, but these often do not disclose which parts of the manufacturing process are responsible for the largest contribution to the overall cost, or which improvements led to cost reductions.
+
+To overcome the limitations of these different methods for understanding technological progress, in this study we rely on a multi-method approach to data collection and analysis, the details of which are provided in Section 2 of the Supplementary Information document.
+Specifically, we combine information obtained from a systematic review of the primary scientific literature, device datasheets, relevant patents, and industry publications (SI Section 2.1) with information gained from semi-structured interviews with experts from academia and industry (SI Section 2.2), bottom-up manufacturing cost modelling (SI Section 2.3), and our own computations of device sub-efficiencies (SI Section 2.4).
+We then use this information to track the historical progress in white LED technology over time across the three groups of metrics identified above in the 'Metrics' section and identify its sources in innovation and technology spillovers.
+
 \end{document}

### 2023-05-29 2_SSL_NE/article/article_supplementary.tex

@@
-\documentclass[10pt]{article}
-% page setup
-\usepackage[a4paper, total={6in, 8.75in}]{geometry}
-\usepackage{parskip}
-% formatting
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc} % use 8-bit T1 fonts
-% cross-referencing
-\usepackage{hyperref}
-\usepackage{url}
-\usepackage{doi}
-% tables
-\usepackage{booktabs}
-% fonts
-\usepackage{amsfonts}
-\usepackage{microtype}
-% s
-\usepackage{graphicx}
-\usepackage{float}
-% math
-\usepackage{amsmath}
-% tables
-\usepackage{tabularx}
-\usepackage{colortbl}
-% boxes
-\usepackage{fancybox}
-% tikz
-\usepackage{tikz}
-\usetikzlibrary{backgrounds}
-\usetikzlibrary{arrows}
-\usetikzlibrary{shapes,shapes.geometric,shapes.misc}
-% landscape pages
-\usepackage{pdflscape}
-% chemical formulas
-\usepackage{chemformula}
-% smart references
-\usepackage[capitalise, nameinlink]{cleveref}
-
+\documentclass[parskip=full]{article}
+
+% functionality
+\input{auxiliary/preamble/1_preamble_functionality_main}
+
+% layout
+\input{auxiliary/preamble/2_preamble_layout_report}
+
+% bibliography
+\input{auxiliary/preamble/1_preamble_functionality_bibliography}
+\addbibresource{bibliography.bib}
+
+%% CUSTOM SETUP %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% custom figure and table numbering
 \renewcommand{\thefigure}{SI\arabic{figure}}
@@
 
-% this style is applied by default to any tikzpicture included via \tikzfig
-\tikzstyle{tikzfig}=[baseline=-0.25em,scale=0.5]
-
-% these are dummy properties used by TikZiT, but ignored by LaTex
-\pgfkeys{/tikz/tikzit fill/.initial=0}
-\pgfkeys{/tikz/tikzit draw/.initial=0}
-\pgfkeys{/tikz/tikzit shape/.initial=0}
-\pgfkeys{/tikz/tikzit category/.initial=0}
-
-% standard layers used in .tikz files
-\pgfdeclarelayer{edgelayer}
-\pgfdeclarelayer{nodelayer}
-\pgfsetlayers{background,edgelayer,nodelayer,main}
-
-% style for blank nodes
-\tikzstyle{none}=[inner sep=0mm]
-
-% include a .tikz file
-\newcommand{\tikzfig}[1]{%
-{\tikzstyle{every picture}=[tikzfig]
-\IfFileExists{#1.tikz}
-  {\input{#1.tikz}}
-  {%
-    \IfFileExists{./figures/#1.tikz}
-      {\input{./figures/#1.tikz}}
-      {\tikz[baseline=-0.5em]{\node[draw=red,font=\color{red},fill=red!10!white] {\textit{#1}};}}%
-  }}%
+%% METADATA %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\hypersetup{
+    pdftitle={Hybrid Life-Cycle Methods},
+    pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
 }
 
-% the same as \tikzfig, but in a {center} environment
-\newcommand{\ctikzfig}[1]{%
-\begin{center}\rm
-  \tikzfig{#1}
-\end{center}}
-
-\setcounter{secnumdepth}{4}
-
-% fix strange self-loops, which are PGF/TikZ default
-\tikzstyle{every loop}=[]
-\input{./auxiliary/tikz_style_files/main.tikzstyles}
-
-%%% custom definitions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\definecolor{silver}{rgb}{0.75, 0.75, 0.75}
-
-%%% document metadata %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{Rapid technological progress in white light-emitting diodes \\ and its sources in innovation and technology spillovers  }
-\date{February 2023}
-
-\hypersetup{
-    pdftitle={Supplementary Information},
-    pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon}
-}
-
-%%% document body %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% MAIN DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 	\centering
-    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_sensitivity.pdf}
+    \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
 	\caption{Sensitivity analysis for selected cost model parameters.
@@
     \vspace{5mm}
-    \begin{tabularx}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
+    \begin{NiceTabularX}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
         \hline
@@
         \hline
-        \end{tabularx}
+        \end{NiceTabularX}
     \vspace{5mm}
@@
  \centering
- \includegraphics[width=15.5cm]{2_SSL_EST/article/figures/subefficiencies_progress.pdf}
+ \includegraphics[width=15.5cm]{./figures/subefficiencies_progress.pdf}
  \caption{Historical improvements in light-emitting diode technology.
@@
     \vspace{5mm}
-    \begin{tabularx}{\textwidth}{ |l|X|X|l|l| }
+    \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
         \multicolumn{5}{c}{Forward Voltage Efficiency (VFE) $\eta_{Vf}$} \\
@@
 	\centering
-    \includegraphics[width=0.9\textwidth]{2_SSL_EST/article/figures/phosphor_spectrum-comparison.pdf}
+    \includegraphics[width=0.9\textwidth]{./figures/phosphor_spectrum-comparison.pdf}
 	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 6 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year.
@@
 	\centering
-    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_calibration.pdf}
+    \includegraphics[width=\textwidth]{./figures/costmodel_calibration.pdf}
 	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel).
@@
 \begin{table}[h!]
-    \begin{tabularx}{\textwidth}{|l|X|}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
     \hline
@@
     \hline
-    \end{tabularx}
+    \end{NiceTabularX}
 \end{table}
@@
 \newpage
-\bibliographystyle{IEEEtran}
-\bibliography{bibliography}
+
+\printbibliography
 

### 2023-05-29 2_SSL_NE/article/article_main.tex
updated main document

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% This is a (brief) model paper using the achemso class
-%% The document class accepts keyval options, which should include
-%% the target journal and optionally the manuscript type. 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[journal=jacsat,manuscript=article]{achemso}
+\documentclass[parskip=full]{article}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Place any additional packages needed here.  Only include packages
-%% which are essential, to avoid problems later. Do NOT use any
-%% packages which require e-TeX (for example etoolbox): the e-TeX
-%% extensions are not currently available on the ACS conversion
-%% servers.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
-\usepackage{tabularx}
-\usepackage[usenames,dvipsnames]{xcolor}
+% functionality
+\input{auxiliary/preamble/1_preamble_functionality_main}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% If issues arise when submitting your manuscript, you may want to
-%% un-comment the next line.  This provides information on the
-%% version of every file you have used.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%\listfiles
+% layout
+\input{auxiliary/preamble/2_preamble_layout_report}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Place any additional macros here.  Please use \newcommand* where
-%% possible, and avoid layout-changing macros (which are not used
-%% when typesetting).
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% bibliography
+\input{auxiliary/preamble/1_preamble_functionality_bibliography}
+\addbibresource{bibliography.bib}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Meta-data block
-%% ---------------
-%% Each author should be given as a separate \author command.
-%%
-%% Corresponding authors should have an e-mail given after the author
-%% name as an \email command. Phone and fax numbers can be given
-%% using \phone and \fax, respectively; this information is optional.
-%%
-%% The affiliation of authors is given after the authors; each
-%% \affiliation command applies to all preceding authors not already
-%% assigned an affiliation.
-%%
-%% The affiliation takes an option argument for the short name.  This
-%% will typically be something like "University of Somewhere".
-%%
-%% The \altaffiliation macro should be used for new address, etc.
-%% On the other hand, \alsoaffiliation is used on a per author basis
-%% when authors are associated with multiple institutions.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% METADATA %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\author{Michael Weinold}
-\email{mweinold@ethz.ch}
-\affiliation[PSI]{Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland}
-\alsoaffiliation[ETHZ]
-{Chair for Entrepreneurial Risk, ETH Zurich, Zurich, Switzerland}
-\alsoaffiliation[UCAM]
-{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom}
+\hypersetup{
+    pdftitle={Hybrid Life-Cycle Methods},
+    pdfauthor={Michael Weinold, Guillaume Majeau-Bettez, Russell McKenna},
+}
 
-\author{Sergey Kolesnikov}
-\affiliation[UCAM]
-{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom}
+%% MAIN DOCUMENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\author{Laura Diaz Anadon}
-\affiliation[UCAM]{Cambridge Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, United Kingdom}
-\alsoaffiliation[UHarv]
-{Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The document title should be given as usual. Some journals require
-%% a running title from the author: this should be supplied as an
-%% optional argument to \title.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[]{Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Some journals require a list of abbreviations or keywords to be
-%% supplied. These should be set up here, and will be printed after
-%% the title and author information, if needed.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\abbreviations{IR,NMR,UV}
-\keywords{American Chemical Society, \LaTeX}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The manuscript does not need to include \maketitle, which is
-%% executed automatically.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The "tocentry" environment can be used to create an entry for the
-%% graphical table of contents. It is given here as some journals
-%% require that it is printed as part of the abstract page. It will
-%% be automatically moved as appropriate.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
+        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Environmental Science \& Technology} (2023) \\
+        \newline
+        $^1$ Technology Assessment Group, Paul Scherrer Institute, Switzerland \\
+        $^2$ ETH Zurich, Zurich, Switzerland \\
+        $^3$ CEENRG, Dept. of Land Economy, University of Cambridge, UK \\
+        $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The abstract environment will automatically gobble the contents
-%% if an abstract is not used by the target journal.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{abstract}
@@
     \centering
-    \begin{tabularx}{\textwidth}{|l|l|l|X|X|l|}
+    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|l|}
     \hline
@@
         2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2.2.2 \\ \hline
-    \end{tabularx}
+    \end{NiceTabularX}
     \caption{LED down-conversion materials (phosphors and quantum dots) related to improvements in consumer experience metrics identified in this study.
@@
     \caption{Technology spillovers involved in white LED technology innovations identified in this study.}
-    \begin{tabularx}{\textwidth}{|l|l|l|X|X|X|l|X|}
+    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|X|l|X|}
     \hline
@@
         2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
-    \end{tabularx}
+    \end{NiceTabularX}
     \begin{tabular}{@{}p{\linewidth}@{}}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The "Acknowledgement" section can be given in all manuscript
-%% classes.  This should be given within the "acknowledgement"
-%% environment, which will make the correct section or running title.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{acknowledgement}
-
 This research was supported by the grant from the Alfred P.
@@
 
-\end{acknowledgement}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The same is true for Supporting Information, which should use the
-%% suppinfo environment.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{suppinfo}
-
 Additional context and governing equations of metrics used to quantify progress in solid-state lighting, including the constituent sub-efficiencies of overall lamp efficiency.
@@
 
-\end{suppinfo}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The appropriate \bibliography command should be placed here.
-%% Notice that the class file automatically sets \bibliographystyle
-%% and also names the section correctly.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliography{bibliography}
+\newpage
+\printbibliography
 

### 2023-05-29 2_SSL_NE/article/article_main.tex
completed latex layout modifications

@@
 % layout
-\input{auxiliary/preamble/2_preamble_layout_report}
+\input{2_SSL_NE/article/auxiliary/preamble/2_preamble_layout_paperdraft}
 
@@
 
+\begin{center}
+    \large
+    \textbf{Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers}
+\end{center}
+
 M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
-        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Environmental Science \& Technology} (2023) \\
 \newline
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Start the main part of the manuscript here.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Keywords}
+\section*{Keywords}
 
 demand-side, solid-state lighting, innovation theory, efficiency, consumer experience
-
-\section{Synopsis}
-
-Understanding the underlying drivers of innovation in LEDs will enable better policy recommendations for a range of demand-side energy technologies, which will play a key role in climate change mitigation.
 

### 2023-05-29 2_SSL_NE/article/article_main.tex
added content to methods section

@@
     pdftitle={Hybrid Life-Cycle Methods},
-    pdfauthor={Michael Weinold, Guillaume Majeau-Bettez, Russell McKenna},
+    pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
 }
@@
 \newline
-$^1$ Technology Assessment Group, Paul Scherrer Institute, Switzerland \\
+$^1$ CEENRG, Dept. of Land Economy, University of Cambridge, Cambridge,  UK \\
 $^2$ ETH Zurich, Zurich, Switzerland \\
-$^3$ CEENRG, Dept. of Land Economy, University of Cambridge, UK \\
+$^3$ Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland \\
 $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA

### 2023-05-29 2_SSL_NE/article/article_methods.tex
added content to methods section

@@
 \hypersetup{
-    pdftitle={Hybrid Life-Cycle Methods},
+    pdftitle={Methods},
     pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
@@
 
-\section{Methods and Data Collection}
+\setlength{\fboxsep}{10pt}
+\fbox{
+    \parbox{\textwidth}{
+        \textit{\textsc{Methods Information}} for: \\
+        \newline 
+        M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
+        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Nature Energy} (2023) \\
+        \newline
+        $^1$ Technology Assessment Group, Paul Scherrer Institute, Switzerland \\
+        $^2$ ETH Zurich, Zurich, Switzerland \\
+        $^3$ CEENRG, Dept. of Land Economy, University of Cambridge, UK \\
+        $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
+    }
+}
+
+\section{Overview}
 \label{sec:methods}
@@
 
+\section{Methodology}
+\label{sec:costmodel}
+
+This subsection provides the details for the main methods and data sources used in our mixed-methods research approach, including the systematic literature review, semi-structured interviews with eminent experts, manufacturing cost modelling, and performance metrics calculations.
+
+\subsection{Systematic Literature Review}
+
+We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
+This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics introduced in the main article; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
+
+Relevant sources for the review were found in an iterative search process that involved two components.
+The first was the search in specialized patent and publication databases as well as company websites.
+The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in section Previous Literature in the main article and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+We also relied on backward citations for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
+
+\subsection{Semi-Structured Interviews}
+
+To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews with eleven eminent experts from academia, industry and the public research sector.
+Experts were selected based on their engagement in  different sub-fields of LED research and manufacturing, as well as the recommendations from other interviewed experts, in essence expanding the list of experts that emerged from the initial literature review.
+All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
+A summary of the background of interviewed experts is provided in \cref{tab:interviews}.
+
+The primary, structured part of the interviews explored which innovations were deemed most relevant to the evolution of device performance, consumer experience and manufacturing cost of LED packages.
+Thereafter, interviewees were asked to consider the extent to which these innovations may have originated outside of their respective field of expertise and the LED industry more broadly—i.e., which of the innovations may be considered spillovers.
+The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
+Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in \cref{sec:costmodel}) and verify device performance data.
+
+\begin{table}[H]
+\small
+    \centering
+    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
+    \vspace{5mm}
+    \begin{tabular}{|l|l|l|l|l|}
+    \hline
+        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
+        1 & Academia & Sr. researcher & UK & Epitaxy \\ \hline
+        2 & Industry & Consultant, former sr. researcher & USA & Device architecture \\ \hline
+        3 & Industry & Consultant, former head of R\&D & Germany & Epitaxy \\ \hline
+        4 & Academia & Professor & Austria & Phosphors \\ \hline
+        5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
+        6 & Consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
+        7 & Academia & Professor & Germany & Phosphors \\ \hline
+        8 & Government & R\&D manager & USA & Device architecture \\ \hline
+        9 & Consulting & Consultant & USA & Device applications \\ \hline
+        10 & Academia & Professor & France & Device physics \\ \hline
+        11 & Industry & Sr scientist, former head of R\&D & USA & Device architecture \\ \hline
+        12 & Industry & Principal scientist & Germany & Phosphors \\ \hline
+        13 & Industry & Former head of R\&D & USA & Phosphors \\ \hline
+    \end{tabular}
+    \label{tab:interviews}
+\end{table}
+
+\subsection{Performance Metrics Calculations}
+\label{subsec:metrics}
+
+The contribution of individual technology innovations and spillovers to the progress in overall device efficiency over time is estimated by index decomposition analysis.
+Mathematically, this involves breaking down a chosen performance indicator into its constituent components, each representing a specific factor that contributes to the change in the indicator \cite{Ang1997}.
+Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
+It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
+
+According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
+
+\begin{align}
+    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
+    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
+\end{align}
+
+where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
+These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:
+
+\begin{equation}
+    \Delta a + \Delta b  = \Delta F
+\end{equation}
+
+To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered.
+However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance.
+For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies where possible and necessary.
+Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications.
+In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra.
+This approach allowed us to  quantify the improvements related to phosphor development in LEDs.
+
+\subsection{Manufacturing Cost Model}
+
+\subsubsection{Our Approach to Cost Modeling}
+
+The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
+Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips.
+The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
+In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
+These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
+
+The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
+Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
+We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
+It instead assumes an hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
+It also assumes the use of the most up-to-date equipment for that year.
+Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
+An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
+Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
+
+The cost model we developed relies on a cumulative approach to yielded cost\cite{becker2001use}.
+In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
+
+\begin{equation}
+    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
+\end{equation}
+
+This cost metric is cumulative by definition, thus
+
+\begin{equation}
+    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
+\end{equation}
+
+Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
+
+\subsubsection{Structure of the Model}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.95\textwidth]{./figures/cost_model.pdf}
+    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
+For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.
+For the computation of total cost, see \cref{subsec:cost_computation}.
+For the computation of changes in total cost, see \cref{sec:contribution_variables}.}
+    \label{fig:costmodel-schematic}
+\end{figure}
+
+The cost model adapted for this publication is a microeconomic manufacturing cost model.
+Within the timeframe and scope laid out in the main publication, it returns the total manufacturing cost of phosphor converted warm white light-emitting diode packages.
+In this computation, it considers the main economic factors associated with operating and maintaining manufacturing equipment.
+It does not consider costs associated with research and development or those associated with the construction of manufacturing facilities.
+It considers market trends through their effect on manufacturing parameters.
+
+A schematic diagram of the cost model is presented in \cref{fig:costmodel-schematic}.
+The cost model is process step-based.
+It is split between the two stages of the manufacturing process: the first stage combining operations at the wafer level, followed by the LED packaging stage.
+The model takes as inputs parameters specific to individual manufacturing process steps (\textit{"manufacturing step parameters"}) and parameters affecting all manufacturing steps (\textit{"global parameters"}).
+The cost for each process step is then computed.
+The cost model returns returns the costs of individual manufacturing steps as well as the total manufacturing cost.
+It further considers the yield per step and returns the cumulative yield, the yielded cost per step and the yielded total manufacturing cost.
+
+\subsubsection{Manufacturing Process Steps by Chip Architecture}
+
+\cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
+
+    %CLASSICAL CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
+            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003. Abbreviations: MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
+            \label{fig:manuf_classical_2003}
+        \end{figure}
+    \end{landscape}
+
+    %VERTICAL THIN-FILM FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
+            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
+Abbreviations: MQW - multiple quantum well; LLO - laser lift-off.
+Continued on next page.}
+            \label{fig:manuf_vtf_2012-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
+            \caption{(2/2) Continued from previous page. Abbreviations: Encaps. - encapsulation.}
+            \label{fig:manuf_vtf_2012-2}
+        \end{figure}
+    \end{landscape}
+
+    % CHIP-SCALE PACKAGE FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
+            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+Abbreviations: MQW - multiple quantum well.
+Continued on next page.}
+            \label{fig:manuf_csp_2020-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
+            \caption{(2/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+Continued from previous page.
+Abbreviations: Encaps.
+- encapsulation.}
+            \label{fig:manuf_csp_2020-2}
+        \end{figure}
+    \end{landscape}
+
+\subsubsection{Computation of Manufacturing Cost}
+\label{subsec:cost_computation}
+
+The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level.
+The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
+\begin{equation}
+\label{eqn:cost_sum}
+    C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
+\end{equation}
+where
+\begin{align*}
+    P_S &\dots \text{sapphire substrate price per die} \\
+    C_w &\dots \text{wafer processing cost per die} \\
+    C_p &\dots \text{die processing cost}
+\end{align*}
+The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.
+\begin{equation}
+        C_w = \sum_i C_i
+\end{equation}
+\begin{equation}
+	C_p = \sum_j C_j
+\end{equation}
+The cost of a single process step $C_i$ can now be written as
+\begin{equation}
+\label{eqn:cost_wafer}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{DPW}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
+\end{equation}
+where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
+\begin{align*}
+        DPW &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
+        y &\dots \text{process step yield} \\
+        u &\dots \text{equipment utilization (relative to theoretical equipment capacity)} \\
+        p &\dots \text{power consumption} \\
+        c_e &\dots \text{hourly electricity cost} \\
+        c_m &\dots \text{hourly maintenance cost} \\
+        c_d &\dots \text{hourly depreciation cost}\\
+        c_l &\dots \text{hourly labour cost} \\
+        c_o &\dots \text{hourly overhead cost} \\
+        t_i &\dots \text{time per run} \\
+        w &\dots \text{wafers per run} \\
+        A &\dots \text{wafer area} \\
+        v_x &\dots \text{volume of material $x$ per wafer} \\
+        c_x &\dots \text{cost of material $x$ per volume}\\
+\end{align*}
+The number of die per wafer $D$ depends on the total usable wafer area.
+The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
+\begin{equation}
+	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
+\end{equation}
+Determining the usable wafer area as a function of these three parameters requires a numerical solution.
+However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
+\begin{equation}
+\label{eqn:dpw}
+	DPW = \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
+\end{equation}
+where
+\begin{align*}
+    d &\dots \text{wafer diameter} \\
+    e &\dots \text{wafer edge exclusion zone width} \\
+    a &\dots \text{die area} \\
+    s &\dots \text{cutting street width} \\
+\end{align*}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
+    \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2002)=50$mm$\rightarrow5370$DPW, $d(2020)=100$mm$, \rightarrow26,838$DPW.
+Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews.
+For detailed statistics on changes in wafer size, compare also \cref{fig:wafers}.}
+    \label{fig:dpw}
+\end{figure}
+
+Results are plotted in \cref{fig:dpw}, showing the increase in die per wafer (DPW) over time. \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
+\begin{equation}
+\label{eqn:cost_wafer_full}
+\begin{split}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
+    &  \frac{1}{y_i}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
+\end{split}
+\end{equation}
+and the cost of a manufacturing step $C_j$ in the packaging category
+\begin{equation}
+\label{eqn:cost_die}
+    C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x c_x \bigg\}
+\end{equation}
+where $c_j$ is $\text{throughput}^{-1}$. The total cost is thus
+\begin{equation}
+\label{eqn:cost_total}
+\begin{split}
+    C= P_s &+ \sum_i \bigg \{ \frac{1}{DPW} \frac{1}{y_i} \bigg[ \frac{t_i}{w_i u_i} \bigg((e*p) + l + m + d +o \bigg)_i  + \sum_{x} v_x p_x \bigg] \bigg \} + \\
+    & + \sum_j \bigg \{ \frac{1}{y_j} \bigg[ \frac{c_j}{u_j}  \bigg((e*p) + l + m + d + o \bigg)_i + \sum_{x} a v_x p_x \bigg ] \bigg\}
+\end{split}
+\end{equation}
+Note that in keeping with the categorization introduced by the United States Department of Energy (cf.
+eg.
+\cite{doe_ssl_rnd_2019}), certain steps from these two categories are reported separately.
+In the wafer processing category, the epitaxy step is reported separately due to its complexity and the large share of cost carried.
+In the wafer processing category, the phosphor step is reported separately.
+
+\subsubsection{Computation of Yielded Cost}
+
+Devices may be damaged or otherwise rendered unusable during the manufacturing process.
+The ratio between the number of good devices per step and the number of handled devices per step is known as the yield.
+Optimizing this yield is critical for reducing manufacturing cost \cite{Kumar2006}.
+This is because cumulative yield quickly drops as the yield from manufacturing steps with below 100\% yield is multiplied.
+We must thus consider not only the manufacturing cost per process step, but also the cost including the yield \cite{becker2001use}\cite{becker2001using}.
+While there are different mathematical approaches to including yield, we follow the definition in \cite{becker2001use}.
+We write for the yielded cost $C_{Y_i}$ of a step $i$ with associated cost (before considering yield) $C_1$ and yield $Y_1$:
+
+\begin{equation}
+\label{eqn:C_2}
+    C_{Y_1} = \frac{C_1}{Y_1}
+\end{equation}
+\begin{equation}
+    C_{Y_2} = \frac{C_1 + C_2}{Y_1 Y_2} - C_{Y_1} = \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} = \frac{1}{Y_1 Y_2} \bigg ( C_1 (1-Y_2) +C_2 \bigg)
+\end{equation}
+\begin{equation}
+    C_{Y_i} = \frac{ \sum_{x \leq i} C_x }{ \prod_{x \leq i} Y_x } - \frac{ \sum_{x<i} C_x }{ \prod_{x<i} Y_x }
+\end{equation}
+
+If a step is applied more than once, we can conveniently rewrite this in a form suited to computation within the \textit{Excel} worksheet.
+Assuming step $2$ is used twice, we get for the yielded cost of this step an equation of the form
+
+\begin{equation}
+\label{eqn:C_2^2}
+    C_{Y_2}^{(2 \times)} = \bigg( \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} \bigg) + \bigg( \frac{C_1 + 2 C_2}{Y_1 Y_2^2} - \frac{C_1 + C_2}{Y_1 Y_2}     \bigg)
+\end{equation}
+\begin{equation}
+    = \frac{1}{Y_1 Y_2^2} \bigg( C_1 (1-Y_2^2) +2C_2 \bigg)
+\end{equation}
+
+which can be compared to  \cref{eqn:C_2} to find a more general form:
+
+\begin{equation}
+    C_{Y_2}^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
+\end{equation}
+
+It can be shown by induction that the general form of a term for a step $i>1$ repeated $n$ times can be expressed as:
+
+\begin{equation}
+\label{eqn:yielded_cost}
+    C_{Y_{i>1}}^{(n \times)} = \frac{1}{Y_i^{n-1} \prod_{x \leq i} Y_x} \bigg( nC_i + \sum_{x < i} C_x (1-Y_i^n) \bigg)
+\end{equation}
+
+This cumulative approach to yielded cost is different from the approach taken in the original \textit{LEDCOM} model.
+It uses what in the literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}.
+In this approach, the yielded cost of a single process step $f$ is described as
+
+\begin{equation}
+    f_\text{single} = \frac{i+s}{y}
+\end{equation}
+
+where
+
+\begin{align*}
+    i &\dots \text{material cost of previous step} \\
+    s &\dots \text{step cost}
+\end{align*}
+
+For process steps that are performed more than once, a series expression is used
+
+\begin{equation}
+    f_{\times 2} =  \frac{\frac{i+s}{y}+s}{y}
+\end{equation}
+\begin{equation}
+\label{eqn:series}
+    f_{\times n} = \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
+\end{equation}
+
+This itemized approach serves as a convenient approximation, but its cumulative contributions do not equal the total yielded cost of the entire manufacturing process
+
+\begin{equation}
+    f_\text{total} = \frac{\sum_i^n s_i}{\prod_i^ny_i} \neq \sum_i^n f_i
+\end{equation}
+
+where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in  \cref{eqn:series}.
+
+\subsubsection{Example of Input Data Considered: Sapphire Wafers}
+
+Sapphire wafers form the substrate on which all other layers of the light-emitting diode are grown.
+Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture.
+In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process.
+Wafers can be either unpatterned or patterned, where the latter has become commonplace by 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
+The price of sapphire substrates has decreased significantly since the year 2000, as shown in the bottom panel of \cref{fig:wafers}.
+This can be attributed not only to the lighting industry, but more importantly to increased supply as a result of increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
+Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing.
+The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}.
+The evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.
+
+\begin{figure}[h!]
+    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
+    \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
+	\caption{Top: Historical data for sapphire substrate prices of different surface properties and diameters.
+Shown is data for polished surface (NPS) and patterned substrates (PSS).
+\textit{Monocrystal} denotes the Russian manufacturer of the same name.
+Dashed lines are projections from the previous year.
+Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}.
+Bottom: Prevalence of sapphire wafer size used in the manufacturing of light-emitting diodes.
+Hatched bars are projections given by the sources.
+Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
+	\label{fig:wafers}
+\end{figure}
+
+\newpage
+\subsubsection{Preliminary Sensitivity Analysis}
+
+\begin{figure}[]
+	\centering
+    \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
+	\caption{Sensitivity analysis for selected cost model parameters.
+Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in \cref{tab:sensitivity}.
+Panel b shows the resulting range of variation in the manufacturing cost.
+Ranges are represented for three years considered in the cost model, where corresponding data is available, by whiskers of the following colour: 2003 – blue, 2012 – green, 2020 – red.
+Note the overall trend of decreasing sensitivity to cost model parameters as the number of die per wafer increases over time.
+Abbreviations: resp.
+- respective; yield sav.
+from insp.
+- yield savings from inspection.}
+	\label{fig:sensitivity}
+\end{figure}
+
+A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
+The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
+As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen because the prices quotes from the United State Geological Survey price database.
+Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
+The upper range was chosen because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
+The results of the analysis are shown in  \cref{fig:sensitivity}.
+The cost model is generally more sensitive to the variation in parameters at smaller wafer diameters.
+The most sensitive parameters are global parameters, such as yield or average equipment throughput.
+
+\begin{table}[]
+\small
+    \caption{Cost model sensitivity analysis parameter list. The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.}
+    \vspace{5mm}
+    \begin{NiceTabularX}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
+        \hline
+            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & Source \\
+        \hline
+            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & \cite{mddi1997cleanroom}\cite{ledcomv2} \newline \cite{bakshi2009euv}\cite{gajera2006process} \\
+        \hline
+            Person-hour & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
+        \hline
+            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I \newline \cite{Appleyard_2001} \\
+        \hline
+            Overall Yield & \% & 100\% & +25,-25 & 100\% & +25,-15 & 100\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
+        \hline
+            Inspec. Yield Savings & \%/inspec. & 0.5\% & +80,-80 & 0.5\% & +80,-80 & 0.5\% & +80,-80 & \cite{mckinseyyield} \\
+        \hline
+            Overall Throughput & UPH or h$^{-1}$ & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & Datasheets \\
+        \hline
+            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I and \cref{fig:wafers} \\
+        \hline
+            Edge Exclusion & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
+        \hline
+            Cutting Width & $\mu$m & 100 & +50,-25 & 75 & +33.3,-20 & 20 & +300,-50 & \cite{masaki2000division}\cite{ils2005width} \newline \cite{photonics2010width}\cite{discowidth} \\
+        \hline
+            Metal Prices & \text{USD}/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
+        \hline
+            Electricity Price & \text{USD}/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
+        \hline
+            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & \cref{fig:wafers} \\
+        \hline
+            Phosphor Prices & \text{USD}/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
+        \hline
+        \end{NiceTabularX}
+    \vspace{5mm}
+
+    Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
+If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
+Abbreviations: FTE - full-time equivalent; UPH - units per hour; Equip.
+Discount - equipment discount (sales rebate for large purchases); Inspec.
+Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).
+    \label{tab:sensitivity}
+\end{table}
+
+\newpage
+\subsection{Computation of Contribution of Individual Variables}
+\label{sec:contribution_variables}
+
+To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
+Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
+\begin{equation}
+F=ab+cd
+\end{equation}
+and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
+\begin{equation}
+\Delta F = F(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta F_i
+\end{equation}
+The infinitesimal contribution to the total funcation value by the infinitesimal change in an input variable is defined through the total differential of the function as
+\begin{equation}
+\text{d}F(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial F }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial F }{\partial x_i}  \Delta x_i
+\end{equation}
+where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
+\begin{equation}
+\Delta F_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial F }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
+\label{eqn:integral_1}
+\end{equation}
+However, data on the input variables is not available in continuous time.
+Disaggregating the contribution of single variables to the change in the cost or performance is thus not straightforward in our model.
+This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}.
+In our work, we propose to address this problem by following an approach developed by Kavlak et al.
+\cite{kavlak2018evaluating}.
+For the detailed derivation, we refer to this publication.
+
+The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
+\begin{equation}
+    F(\vec{r}) = F(r_1,r_2, \dots) = \sum_i F_i
+\end{equation}
+where
+\begin{equation}
+    F_i(\vec{r}) = F_i^0 \prod_w g_{iw}(r_w)
+\end{equation}
+Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
+\begin{equation}
+    \Delta F_x = \int_{t=t_0}^{t_1} F(t) \frac{ \partial \ln F }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
+\end{equation}
+where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$.
+In practice, this constant value can be approximated through the geometric and arithmetic means $\tilde{F} \approx \frac{2}{3} F_i^\text{geo} + \frac{1}{3} \overline{F_i}$.
+The contribution of a single cost model variable $r_z$ can then be written as
+\begin{equation}
+    \Delta F_z (t_1,t_2) \approx \sum_i \tilde{F_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
+\end{equation}
+We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article.
+Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.
+
 \end{document}

### 2023-05-29 2_SSL_NE/article/article_supplementary.tex
added content to methods section

@@
         M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
-        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Environmental Science \& Technology} (2023) \\
+        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Nature Energy} (2023) \\
         \newline
@@
 \newpage
-\section{Methodology}
-\label{sec:costmodel}
-
-This subsection provides the details for the main methods and data sources used in our mixed-methods research approach, including the systematic literature review, semi-structured interviews with eminent experts, manufacturing cost modelling, and performance metrics calculations.
-
-\subsection{Systematic Literature Review}
-
-We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
-This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics introduced in the main article; 2) identifying individual innovations that contributed to this evolution and whether or not they could be spillovers, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of research or technology outside of solid-state lighting, making them a technology spillovers.
-
-Relevant sources for the review were found in an iterative search process that involved two components.
-The first was the search in specialized patent and publication databases as well as company websites.
-The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in section Previous Literature in the main article and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
-We also relied on backward citations for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
-
-\subsection{Semi-Structured Interviews}
-
-To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews with eleven eminent experts from academia, industry and the public research sector.
-Experts were selected based on their engagement in  different sub-fields of LED research and manufacturing, as well as the recommendations from other interviewed experts, in essence expanding the list of experts that emerged from the initial literature review.
-All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
-A summary of the background of interviewed experts is provided in \cref{tab:interviews}.
-
-The primary, structured part of the interviews explored which innovations were deemed most relevant to the evolution of device performance, consumer experience and manufacturing cost of LED packages.
-Thereafter, interviewees were asked to consider the extent to which these innovations may have originated outside of their respective field of expertise and the LED industry more broadly—i.e., which of the innovations may be considered spillovers.
-The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
-Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in \cref{sec:costmodel}) and verify device performance data.
-
-\begin{table}[H]
-\small
-    \centering
-    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
-    \vspace{5mm}
-    \begin{tabular}{|l|l|l|l|l|}
-    \hline
-        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
-        1 & Academia & Sr. researcher & UK & Epitaxy \\ \hline
-        2 & Industry & Consultant, former sr. researcher & USA & Device architecture \\ \hline
-        3 & Industry & Consultant, former head of R\&D & Germany & Epitaxy \\ \hline
-        4 & Academia & Professor & Austria & Phosphors \\ \hline
-        5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
-        6 & Consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
-        7 & Academia & Professor & Germany & Phosphors \\ \hline
-        8 & Government & R\&D manager & USA & Device architecture \\ \hline
-        9 & Consulting & Consultant & USA & Device applications \\ \hline
-        10 & Academia & Professor & France & Device physics \\ \hline
-        11 & Industry & Sr scientist, former head of R\&D & USA & Device architecture \\ \hline
-        12 & Industry & Principal scientist & Germany & Phosphors \\ \hline
-        13 & Industry & Former head of R\&D & USA & Phosphors \\ \hline
-    \end{tabular}
-    \label{tab:interviews}
-\end{table}
-
-\subsection{Performance Metrics Calculations}
-\label{subsec:metrics}
-
-The contribution of individual technology innovations and spillovers to the progress in overall device efficiency over time is estimated by index decomposition analysis.
-Mathematically, this involves breaking down a chosen performance indicator into its constituent components, each representing a specific factor that contributes to the change in the indicator \cite{Ang1997}.
-Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
-It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
-
-According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
-
-\begin{align}
-    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
-    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
-\end{align}
-
-where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
-These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:
-
-\begin{equation}
-    \Delta a + \Delta b  = \Delta F
-\end{equation}
-
-To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered.
-However, the scope of data reporting in scientific literature and industry publications is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance.
-For this reason, our data collection efforts were supplemented by performance calculations for individual sub-efficiencies where possible and necessary.
-Of the device sub-efficiencies, those related to the emission spectrum  were computed from the spectral data often reported in LED device specifications.
-In particular, we used the \texttt{colour-science} package for Python\cite{python-colour} to calculate the luminous efficacy of radiation, colour rendering performance and luminous efficacy of radiation of phosphor down-conversion of blue light on the basis of available LED emission spectra.
-This approach allowed us to  quantify the improvements related to phosphor development in LEDs.
-
-\subsection{Manufacturing Cost Model}
-
-\subsubsection{Our Approach to Cost Modeling}
-
-The model structure is generally based on the 2012 LEDCOM cost model, but we expand it significantly both in scope and in its ability to capture historical trends.
-The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the \textit{Internet Archive}, and expert interviews.
-Because the architecture of LED chips has changed significantly since the introduction of the first commercial devices in 1996, three different chip architectures were initially considered in the model: classical chips, flip chips, and chip-scale package flip chips.
-The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
-In addition, two LED life cycle analyses \cite{scholand2012life}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
-These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
-
-The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
-Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
-We also note that the purpose of our cost model is not to give specific estimates of LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
-It instead assumes an hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
-It also assumes the use of the most up-to-date equipment for that year.
-Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
-An important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we were able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
-Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
-
-The cost model we developed relies on a cumulative approach to yielded cost\cite{becker2001use}.
-In this approach, the yielded cost of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$:
-
-\begin{equation}
-    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
-\end{equation}
-
-This cost metric is cumulative by definition, thus
-
-\begin{equation}
-    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
-\end{equation}
-
-Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
-
-\subsubsection{Structure of the Model}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.95\textwidth]{./figures/cost_model.pdf}
-    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
-For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.
-For the computation of total cost, see \cref{subsec:cost_computation}.
-For the computation of changes in total cost, see \cref{sec:contribution_variables}.}
-    \label{fig:costmodel-schematic}
-\end{figure}
-
-The cost model adapted for this publication is a microeconomic manufacturing cost model.
-Within the timeframe and scope laid out in the main publication, it returns the total manufacturing cost of phosphor converted warm white light-emitting diode packages.
-In this computation, it considers the main economic factors associated with operating and maintaining manufacturing equipment.
-It does not consider costs associated with research and development or those associated with the construction of manufacturing facilities.
-It considers market trends through their effect on manufacturing parameters.
-
-A schematic diagram of the cost model is presented in \cref{fig:costmodel-schematic}.
-The cost model is process step-based.
-It is split between the two stages of the manufacturing process: the first stage combining operations at the wafer level, followed by the LED packaging stage.
-The model takes as inputs parameters specific to individual manufacturing process steps (\textit{"manufacturing step parameters"}) and parameters affecting all manufacturing steps (\textit{"global parameters"}).
-The cost for each process step is then computed.
-The cost model returns returns the costs of individual manufacturing steps as well as the total manufacturing cost.
-It further considers the yield per step and returns the cumulative yield, the yielded cost per step and the yielded total manufacturing cost.
-
-\subsubsection{Manufacturing Process Steps by Chip Architecture}
-
-\cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
-
-    %CLASSICAL CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
-            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003. Abbreviations: MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
-            \label{fig:manuf_classical_2003}
-        \end{figure}
-    \end{landscape}
-
-    %VERTICAL THIN-FILM FLIP-CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
-            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
-Abbreviations: MQW - multiple quantum well; LLO - laser lift-off.
-Continued on next page.}
-            \label{fig:manuf_vtf_2012-1}
-        \end{figure}
-    \end{landscape}
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
-            \caption{(2/2) Continued from previous page. Abbreviations: Encaps. - encapsulation.}
-            \label{fig:manuf_vtf_2012-2}
-        \end{figure}
-    \end{landscape}
-
-    % CHIP-SCALE PACKAGE FLIP-CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
-            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
-Abbreviations: MQW - multiple quantum well.
-Continued on next page.}
-            \label{fig:manuf_csp_2020-1}
-        \end{figure}
-    \end{landscape}
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
-            \caption{(2/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
-Continued from previous page.
-Abbreviations: Encaps.
-- encapsulation.}
-            \label{fig:manuf_csp_2020-2}
-        \end{figure}
-    \end{landscape}
-
-\subsubsection{Computation of Manufacturing Cost}
-\label{subsec:cost_computation}
-
-The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level.
-The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
-\begin{myequation}
-\label{eqn:cost_sum}
-    C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
-\end{myequation}
-where
-\begin{align*}
-    P_S &\dots \text{sapphire substrate price per die} \\
-    C_w &\dots \text{wafer processing cost per die} \\
-    C_p &\dots \text{die processing cost}
-\end{align*}
-The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.
-\begin{myequation}
-        C_w = \sum_i C_i
-\end{myequation}
-\begin{myequation}
-	C_p = \sum_j C_j
-\end{myequation}
-The cost of a single process step $C_i$ can now be written as
-\begin{myequation}
-\label{eqn:cost_wafer}
-    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{DPW}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
-\end{myequation}
-where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
-\begin{align*}
-        DPW &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
-        y &\dots \text{process step yield} \\
-        u &\dots \text{equipment utilization (relative to theoretical equipment capacity)} \\
-        p &\dots \text{power consumption} \\
-        c_e &\dots \text{hourly electricity cost} \\
-        c_m &\dots \text{hourly maintenance cost} \\
-        c_d &\dots \text{hourly depreciation cost}\\
-        c_l &\dots \text{hourly labour cost} \\
-        c_o &\dots \text{hourly overhead cost} \\
-        t_i &\dots \text{time per run} \\
-        w &\dots \text{wafers per run} \\
-        A &\dots \text{wafer area} \\
-        v_x &\dots \text{volume of material $x$ per wafer} \\
-        c_x &\dots \text{cost of material $x$ per volume}\\
-\end{align*}
-The number of die per wafer $D$ depends on the total usable wafer area.
-The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
-\begin{myequation}
-	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
-\end{myequation}
-Determining the usable wafer area as a function of these three parameters requires a numerical solution.
-However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
-\begin{myequation}
-\label{eqn:dpw}
-	DPW = \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
-\end{myequation}
-where
-\begin{align*}
-    d &\dots \text{wafer diameter} \\
-    e &\dots \text{wafer edge exclusion zone width} \\
-    a &\dots \text{die area} \\
-    s &\dots \text{cutting street width} \\
-\end{align*}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
-    \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2002)=50$mm$\rightarrow5370$DPW, $d(2020)=100$mm$, \rightarrow26,838$DPW.
-Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews.
-For detailed statistics on changes in wafer size, compare also \cref{fig:wafers}.}
-    \label{fig:dpw}
-\end{figure}
-
-Results are plotted in \cref{fig:dpw}, showing the increase in die per wafer (DPW) over time. \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
-\begin{myequation}
-\label{eqn:cost_wafer_full}
-\begin{split}
-    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
-    &  \frac{1}{y_i}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
-\end{split}
-\end{myequation}
-and the cost of a manufacturing step $C_j$ in the packaging category
-\begin{myequation}
-\label{eqn:cost_die}
-    C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x c_x \bigg\}
-\end{myequation}
-where $c_j$ is $\text{throughput}^{-1}$. The total cost is thus
-\begin{myequation}
-\label{eqn:cost_total}
-\begin{split}
-    C= P_s &+ \sum_i \bigg \{ \frac{1}{DPW} \frac{1}{y_i} \bigg[ \frac{t_i}{w_i u_i} \bigg((e*p) + l + m + d +o \bigg)_i  + \sum_{x} v_x p_x \bigg] \bigg \} + \\
-    & + \sum_j \bigg \{ \frac{1}{y_j} \bigg[ \frac{c_j}{u_j}  \bigg((e*p) + l + m + d + o \bigg)_i + \sum_{x} a v_x p_x \bigg ] \bigg\}
-\end{split}
-\end{myequation}
-Note that in keeping with the categorization introduced by the United States Department of Energy (cf.
-eg.
-\cite{doe_ssl_rnd_2019}), certain steps from these two categories are reported separately.
-In the wafer processing category, the epitaxy step is reported separately due to its complexity and the large share of cost carried.
-In the wafer processing category, the phosphor step is reported separately.
-
-\subsubsection{Computation of Yielded Cost}
-
-Devices may be damaged or otherwise rendered unusable during the manufacturing process.
-The ratio between the number of good devices per step and the number of handled devices per step is known as the yield.
-Optimizing this yield is critical for reducing manufacturing cost \cite{Kumar2006}.
-This is because cumulative yield quickly drops as the yield from manufacturing steps with below 100\% yield is multiplied.
-We must thus consider not only the manufacturing cost per process step, but also the cost including the yield \cite{becker2001use}\cite{becker2001using}.
-While there are different mathematical approaches to including yield, we follow the definition in \cite{becker2001use}.
-We write for the yielded cost $C_{Y_i}$ of a step $i$ with associated cost (before considering yield) $C_1$ and yield $Y_1$:
-
-\begin{myequation}
-\label{eqn:C_2}
-    C_{Y_1} = \frac{C_1}{Y_1}
-\end{myequation}
-\begin{myequation}
-    C_{Y_2} = \frac{C_1 + C_2}{Y_1 Y_2} - C_{Y_1} = \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} = \frac{1}{Y_1 Y_2} \bigg ( C_1 (1-Y_2) +C_2 \bigg)
-\end{myequation}
-\begin{myequation}
-    C_{Y_i} = \frac{ \sum_{x \leq i} C_x }{ \prod_{x \leq i} Y_x } - \frac{ \sum_{x<i} C_x }{ \prod_{x<i} Y_x }
-\end{myequation}
-
-If a step is applied more than once, we can conveniently rewrite this in a form suited to computation within the \textit{Excel} worksheet.
-Assuming step $2$ is used twice, we get for the yielded cost of this step an equation of the form
-
-\begin{myequation}
-\label{eqn:C_2^2}
-    C_{Y_2}^{(2 \times)} = \bigg( \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} \bigg) + \bigg( \frac{C_1 + 2 C_2}{Y_1 Y_2^2} - \frac{C_1 + C_2}{Y_1 Y_2}     \bigg)
-\end{myequation}
-\begin{myequation}
-    = \frac{1}{Y_1 Y_2^2} \bigg( C_1 (1-Y_2^2) +2C_2 \bigg)
-\end{myequation}
-
-which can be compared to  \cref{eqn:C_2} to find a more general form:
-
-\begin{myequation}
-    C_{Y_2}^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
-\end{myequation}
-
-It can be shown by induction that the general form of a term for a step $i>1$ repeated $n$ times can be expressed as:
-
-\begin{myequation}
-\label{eqn:yielded_cost}
-    C_{Y_{i>1}}^{(n \times)} = \frac{1}{Y_i^{n-1} \prod_{x \leq i} Y_x} \bigg( nC_i + \sum_{x < i} C_x (1-Y_i^n) \bigg)
-\end{myequation}
-
-This cumulative approach to yielded cost is different from the approach taken in the original \textit{LEDCOM} model.
-It uses what in the literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}.
-In this approach, the yielded cost of a single process step $f$ is described as
-
-\begin{myequation}
-    f_\text{single} = \frac{i+s}{y}
-\end{myequation}
-
-where
-
-\begin{align*}
-    i &\dots \text{material cost of previous step} \\
-    s &\dots \text{step cost}
-\end{align*}
-
-For process steps that are performed more than once, a series expression is used
-
-\begin{myequation}
-    f_{\times 2} =  \frac{\frac{i+s}{y}+s}{y}
-\end{myequation}
-\begin{myequation}
-\label{eqn:series}
-    f_{\times n} = \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
-\end{myequation}
-
-This itemized approach serves as a convenient approximation, but its cumulative contributions do not equal the total yielded cost of the entire manufacturing process
-
-\begin{myequation}
-    f_\text{total} = \frac{\sum_i^n s_i}{\prod_i^ny_i} \neq \sum_i^n f_i
-\end{myequation}
-
-where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in  \cref{eqn:series}.
-
-\subsubsection{Example of Input Data Considered: Sapphire Wafers}
-
-Sapphire wafers form the substrate on which all other layers of the light-emitting diode are grown.
-Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture.
-In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process.
-Wafers can be either unpatterned or patterned, where the latter has become commonplace by 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
-The price of sapphire substrates has decreased significantly since the year 2000, as shown in the bottom panel of \cref{fig:wafers}.
-This can be attributed not only to the lighting industry, but more importantly to increased supply as a result of increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
-Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing.
-The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}.
-The evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.
-
-\begin{figure}[h!]
-    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
-    \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
-	\caption{Top: Historical data for sapphire substrate prices of different surface properties and diameters.
-Shown is data for polished surface (NPS) and patterned substrates (PSS).
-\textit{Monocrystal} denotes the Russian manufacturer of the same name.
-Dashed lines are projections from the previous year.
-Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}.
-Bottom: Prevalence of sapphire wafer size used in the manufacturing of light-emitting diodes.
-Hatched bars are projections given by the sources.
-Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
-	\label{fig:wafers}
-\end{figure}
-
-\newpage
-\subsubsection{Preliminary Sensitivity Analysis}
-
-\begin{figure}[]
-	\centering
-    \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
-	\caption{Sensitivity analysis for selected cost model parameters.
-Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in \cref{tab:sensitivity}.
-Panel b shows the resulting range of variation in the manufacturing cost.
-Ranges are represented for three years considered in the cost model, where corresponding data is available, by whiskers of the following colour: 2003 – blue, 2012 – green, 2020 – red.
-Note the overall trend of decreasing sensitivity to cost model parameters as the number of die per wafer increases over time.
-Abbreviations: resp.
-- respective; yield sav.
-from insp.
-- yield savings from inspection.}
-	\label{fig:sensitivity}
-\end{figure}
-
-A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
-The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
-As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen because the prices quotes from the United State Geological Survey price database.
-Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
-The upper range was chosen because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
-The results of the analysis are shown in  \cref{fig:sensitivity}.
-The cost model is generally more sensitive to the variation in parameters at smaller wafer diameters.
-The most sensitive parameters are global parameters, such as yield or average equipment throughput.
-
-\begin{table}[]
-\small
-    \caption{Cost model sensitivity analysis parameter list. The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.}
-    \vspace{5mm}
-    \begin{NiceTabularX}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
-        \hline
-            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & Source \\
-        \hline
-            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & \cite{mddi1997cleanroom}\cite{ledcomv2} \newline \cite{bakshi2009euv}\cite{gajera2006process} \\
-        \hline
-            Person-hour & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
-        \hline
-            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I \newline \cite{Appleyard_2001} \\
-        \hline
-            Overall Yield & \% & 100\% & +25,-25 & 100\% & +25,-15 & 100\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
-        \hline
-            Inspec. Yield Savings & \%/inspec. & 0.5\% & +80,-80 & 0.5\% & +80,-80 & 0.5\% & +80,-80 & \cite{mckinseyyield} \\
-        \hline
-            Overall Throughput & UPH or h$^{-1}$ & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & Datasheets \\
-        \hline
-            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I and \cref{fig:wafers} \\
-        \hline
-            Edge Exclusion & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
-        \hline
-            Cutting Width & $\mu$m & 100 & +50,-25 & 75 & +33.3,-20 & 20 & +300,-50 & \cite{masaki2000division}\cite{ils2005width} \newline \cite{photonics2010width}\cite{discowidth} \\
-        \hline
-            Metal Prices & \text{USD}/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
-        \hline
-            Electricity Price & \text{USD}/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
-        \hline
-            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & \cref{fig:wafers} \\
-        \hline
-            Phosphor Prices & \text{USD}/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
-        \hline
-        \end{NiceTabularX}
-    \vspace{5mm}
-
-    Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
-If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
-Abbreviations: FTE - full-time equivalent; UPH - units per hour; Equip.
-Discount - equipment discount (sales rebate for large purchases); Inspec.
-Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).
-    \label{tab:sensitivity}
-\end{table}
-
-\newpage
-\subsection{Computation of Contribution of Individual Variables}
-\label{sec:contribution_variables}
-
-To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
-Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
-\begin{myequation}
-F=ab+cd
-\end{myequation}
-and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
-\begin{myequation}
-\Delta F = F(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta F_i
-\end{myequation}
-The infinitesimal contribution to the total funcation value by the infinitesimal change in an input variable is defined through the total differential of the function as
-\begin{myequation}
-\text{d}F(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial F }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial F }{\partial x_i}  \Delta x_i
-\end{myequation}
-where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
-\begin{myequation}
-\Delta F_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial F }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
-\label{eqn:integral_1}
-\end{myequation}
-However, data on the input variables is not available in continuous time.
-Disaggregating the contribution of single variables to the change in the cost or performance is thus not straightforward in our model.
-This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}.
-In our work, we propose to address this problem by following an approach developed by Kavlak et al.
-\cite{kavlak2018evaluating}.
-For the detailed derivation, we refer to this publication.
-
-The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
-\begin{myequation}
-    F(\vec{r}) = F(r_1,r_2, \dots) = \sum_i F_i
-\end{myequation}
-where
-\begin{myequation}
-    F_i(\vec{r}) = F_i^0 \prod_w g_{iw}(r_w)
-\end{myequation}
-Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
-\begin{myequation}
-    \Delta F_x = \int_{t=t_0}^{t_1} F(t) \frac{ \partial \ln F }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
-\end{myequation}
-where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$.
-In practice, this constant value can be approximated through the geometric and arithmetic means $\tilde{F} \approx \frac{2}{3} F_i^\text{geo} + \frac{1}{3} \overline{F_i}$.
-The contribution of a single cost model variable $r_z$ can then be written as
-\begin{myequation}
-    \Delta F_z (t_1,t_2) \approx \sum_i \tilde{F_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
-\end{myequation}
-We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article.
-Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.
 

### 2023-05-31 2_SSL_NE/article/article_main.tex
minor changes

@@
 \begin{abstract}
-  The first commercial white light-emitting diodes (LEDs) were introduced to the market in 1996.
-Since then, white LEDs have experienced major improvements in performance and efficiency.
-Significant decreases in cost have been driven by a rapid expansion of the market share of LED-based solid-state lighting (SSL), which in turn has driven down costs even further.
-Today, the technology is one of the key solutions in global climate change mitigation efforts.
-However, despite its significance, the extent and sources of the underlying improvements have not been systematically investigated.
-Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies.
-With this aim, we gather and analyse systematic evidence on cost and performance improvements in white LEDs from cost and performance modelling, augmented by a literature review and interviews with experts in industry and academia.
-We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.8\%$ in 2020.
-During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S.
-location using state-of-the-art equipment has dropped from 1.1\$ to 0.05\$ (in 2020 USD) between 2003 and 2020, a 95.5\% decrease.
-We further show that technology spillovers — i.e., knowledge originating in other technologies, sectors, or scientific disciplines — affected all performance dimensions of LEDs, contributing no less than 8.5\% to total efficiency improvements and $\sim$100\% to the improvements in consumer experience metrics identified, thereby playing a key role in the widespread adoption of the technology.
-The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.
+  Since their introduction in 1996, white LEDs have greatly improved in performance, efficiency, and manufacturing cost.
+These advancements are crucial for reducing carbon emissions as lighting accounts for nearly 20\% of global electricity consumption.
+Studying the rapid progress in LED technology can provide valuable insights for accelerating innovation in other clean energy technologies.
+However, the extent and sources of these improvements have not been systematically investigated.
+Through cost and performance modeling, along with expert interviews, we found that the efficiency of top-performing warm white LED packages increased from 5.8\% in 2003 to 38.8\% in 2020.
+Additionally, the manufacturing cost of low-to-mid-power LED packages decreased from \$1.1 to \$0.05 (in 2020 USD) between 2003 and 2020, a 95.5\% decrease.
+Technology spillovers from other sectors contributed to 8.5\% of efficiency improvements and nearly 100\% of consumer experience enhancements, playing a crucial role in widespread LED adoption.
 \end{abstract}
@@
 
-\section{Metrics for Technological Progress in LEDs}
-
-Investigating the sources of rapid progress in LED technology over the past decades, in which white LEDs have come to dominate the lighting market \cite{zissis2021}, requires selecting appropriate metrics for tracking and quantifying this progress.
-The choice of metrics affects both what data sources can be used in the analysis and which research methodologies can be used to calculate and analyse such metrics.
-We select the metrics based on the following two general criteria.
-First, we focus only on metrics widely accepted and reported in industry because metrics proposed in the scientific literature but not reported by device manufacturers cannot be used to compare the performance of commercial LED devices over time.
-Second, the chosen metrics must be useful for understanding the impact of individual technological improvements on relevant performance and cost characteristics.
-
-Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}.
-However, metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000,Sandahl2006,CAIRD2008,murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
-Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
-Next, we introduce and discuss the metrics that we use to track progress in each of these three areas.
-
-\subsection{Device Performance Metrics}
-\label{sec:device_performance_metrics}
-
-During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
-This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
-In 2000, the highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
-Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
-Today, however, the devices with the highest performance yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
-LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
-Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
-Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
-
-We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology.
-In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
-The equations defining these sub-efficiencies are provided in the Supplementary Information in Section 1.3.
-The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
-
-\begin{equation}
-    \eta_L = \prod_{i=(V_f, \dots, S)} \eta_i
-\end{equation}
-
-This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
-
-\subsection{Consumer Experience Metrics}
-
-The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
-Any metric relevant to customer experience can thus be calculated from the spectrum alone.
-The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the down-conversion phosphor used in the device.
-It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
-Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: Colour Rendering Index (CRI) and Colour Temperature.
-We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
-This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
-As a result, it is beyond the scope of this work.
-
-\subsubsection{Color Rendering Index (CRI)}
-
-The Colour Rendering Index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
-The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
-CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
-However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
-High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
-It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
-However, these niche applications remain outside our focus on general illumination.
-Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
-
-\subsubsection{Colour Temperature}
-
-The Colour Temperature of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
-Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
-Early white LEDs produced only cool white light \cite{mueller2000light}.
-The introduction of first commercial warm white LED light bulbs played a significant part in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
-For this reason, we also adopt colour temperature as a metric for tracking progress in consumer experience in LED lighting.
-
-\subsection{Manufacturing Cost}
-
-In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must highlight the complexity of this task.
-Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
-Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
-However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
-In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
-Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
-
-We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
-In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
-
-We describe the details of our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model's limitations, in Supplementary Information section 2.4.
-The model structure is generally based on the 2012 LEDCOM cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
-The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
-
-\clearpage
 \section{Results}

### 2023-05-31 2_SSL_NE/article/article_supplementary.tex
minor changes

@@
 
-\section{Metrics}
+\section{Metrics for Technological Progress in LEDs}
+
+This section provides additional background and details for the physical, economic and consumer experience metrics that were chosen to quantify multi-dimensional progress in light-emitting diode technology.
+
+\subsection{Overview}
+
+Investigating the sources of rapid progress in LED technology over the past decades, in which white LEDs have come to dominate the lighting market \cite{zissis2021}, requires selecting appropriate metrics for tracking and quantifying this progress.
+The choice of metrics affects both what data sources can be used in the analysis and which research methodologies can be used to calculate and analyse such metrics.
+We select the metrics based on the following two general criteria.
+First, we focus only on metrics widely accepted and reported in industry because metrics proposed in the scientific literature but not reported by device manufacturers cannot be used to compare the performance of commercial LED devices over time.
+Second, the chosen metrics must be useful for understanding the impact of individual technological improvements on relevant performance and cost characteristics.
+
+Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}.
+However, metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000,Sandahl2006,CAIRD2008,murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
+Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
+Next, we introduce and discuss the metrics that we use to track progress in each of these three areas.
+
+\subsection{Device Performance Metrics}
+\label{sec:device_performance_metrics}
+
+During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
+This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
+In 2000, the highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
+Today, however, the devices with the highest performance yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
+LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
+Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
+Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
+
+We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology.
+In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
+The equations defining these sub-efficiencies are provided in the Supplementary Information in Section 1.3.
+The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
+
+\begin{equation}
+    \eta_L = \prod_{i=(V_f, \dots, S)} \eta_i
+\end{equation}
+
+This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
+
+\subsection{Consumer Experience Metrics}
+
+The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
+Any metric relevant to customer experience can thus be calculated from the spectrum alone.
+The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the down-conversion phosphor used in the device.
+It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
+Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: Colour Rendering Index (CRI) and Colour Temperature.
+We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
+This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
+As a result, it is beyond the scope of this work.
+
+\subsubsection{Color Rendering Index (CRI)}
+
+The Colour Rendering Index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
+The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
+CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
+However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
+High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
+It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
+However, these niche applications remain outside our focus on general illumination.
+Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
+
+\subsubsection{Colour Temperature}
+
+The Colour Temperature of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
+Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
+Early white LEDs produced only cool white light \cite{mueller2000light}.
+The introduction of first commercial warm white LED light bulbs played a significant part in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
+For this reason, we also adopt colour temperature as a metric for tracking progress in consumer experience in LED lighting.
+
+\subsection{Manufacturing Cost}
+
+In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must highlight the complexity of this task.
+Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
+Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
+However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
+In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
+Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
+
+We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
+In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
+
+We describe the details of our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model's limitations, in Supplementary Information section 2.4.
+The model structure is generally based on the 2012 LEDCOM cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
 

### 2023-07-28 2_SSL_NE/article/article_main.tex
fixed file paths
minor edits

@@
 % functionality
-\input{auxiliary/preamble/1_preamble_functionality_main}
+\input{./auxiliary/preamble/1_preamble_functionality_main}
 
 % layout
-\input{2_SSL_NE/article/auxiliary/preamble/2_preamble_layout_paperdraft}
+\input{./auxiliary/preamble/2_preamble_layout_paperdraft}
 
 % bibliography
-\input{auxiliary/preamble/1_preamble_functionality_bibliography}
+\input{./auxiliary/preamble/1_preamble_functionality_bibliography}
 \addbibresource{bibliography.bib}
@@
 
-There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{johansson2012global,national2016power,iea2020perspectives}.
-The role of these factors at different stages of the innovation life cycle \cite{ grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}.
+There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{national2016power,iea2020perspectives}.
+The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}.
 Innovation is also driven by forces of supply and demand.
@@
 Among these drivers, the role of external  knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area  \cite{Stephan2021}.
-While the exact definition of spillovers in the literature depends on the context and the research question \cite{Liu2003,Nemet2012}, we follow Stephan et al.
+While the exact definition of spillovers in the literature depends on the context and the research question \cite{Nemet2012}, we follow Stephan et al.
 \cite{Stephan2021} and consider knowledge to be external - thus a spillover -  if it has been developed for application in other technologies, sectors, or scientific disciplines.
@@
 
-These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies  \cite{weinold2020long,Mills2014,Stegmaier2021,grubb2021new}.
+These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long,Stegmaier2021}.
 As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
@@
 For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
-LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{o2019optical,Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
+LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
 
@@
 We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
-Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry.
+Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews with eminent experts from academia, public research institutions and industry.
 This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.
@@
 
-The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Akasaki2015,Nakamura2015}.
+The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Nakamura2015}.
 The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006,krames2007led,laubsch2009high,hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
@@
 The waterfall diagrams in Figure \ref{fgr:waterfall} show how improvements in sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.7\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
-As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations\cite{tsao2010solid}.
+As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations \cite{tsao2010solid}.
 We note, however, that the loss channels with a fixed physical limit, e.g., Stokes loss that contributes to the conversion efficiency by phosphors, has become more dominant in 2016 and 2020 compared to 2003 and 2010.
@@
 Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
-We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model\cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables\cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} in Section 2 of the Supplementary Information.
+We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables in Section 2 of the Supplementary Information.
 

### 2023-08-10 2_SSL_NE/article/article_main.tex

@@
 \hypersetup{
-    pdftitle={Hybrid Life-Cycle Methods},
+    pdftitle={Main Article},
     pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
@@
 $^1$ CEENRG, Dept. of Land Economy, University of Cambridge, Cambridge,  UK \\
-$^2$ ETH Zurich, Zurich, Switzerland \\
+$^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
 $^3$ Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland \\

### 2023-08-10 2_SSL_NE/article/article_supplementary.tex

@@
 
+% nature energy custom code
+
+\newcounter{supplementarynote}
+\renewcommand{\thesupplementarynote}{\arabic{supplementarynote}}
+\newcommand{\supplementarynote}{%
+  \refstepcounter{supplementarynote}%
+  Supplementary Note~\thesupplementarynote:~%
+}
+\titleformat{\section}
+  {\normalfont\large\bfseries}
+  {\supplementarynote}
+  {0em} % Adjust this value to control the spacing
+  {}
+\renewcommand{\thesection}{\thesupplementarynote}
+
 % bibliography
@@
 % custom figure and table numbering
-\renewcommand{\thefigure}{SI\arabic{figure}}
+\renewcommand{\thefigure}{Supplementary Figure \arabic{figure}}
 \renewcommand{\thetable}{SI\arabic{table}}
@@
 \hypersetup{
-    pdftitle={Hybrid Life-Cycle Methods},
+    pdftitle={Supplement},
     pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
@@
         M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
-        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" \textit{Nature Energy} (2023) \\
+        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" (2023) \\
         \newline
         $^1$ Technology Assessment Group, Paul Scherrer Institute, Switzerland \\
-        $^2$ ETH Zurich, Zurich, Switzerland \\
+        $^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
         $^3$ CEENRG, Dept. of Land Economy, University of Cambridge, UK \\

### 2023-08-10 2_SSL_NE/article/article_main.tex
completed section 1

@@
 
-demand-side, solid-state lighting, innovation theory, efficiency, consumer experience
+solid-state lighting, light emitting diodes, innovation, technology spillovers, energy efficiency, cost reductions, consumer experience metrics
 
@@
 \end{figure}
-
-\section{Previous Literature}
-\label{sec:prev_lit}
-
-The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Nakamura2015}.
-The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006,krames2007led,laubsch2009high,hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
-
-Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status,Phillips2007,Bierhuizen2007,Nakamura2013,feezell2018invention,Taki2019}.
-Additional studies provide more detail regarding the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005,Shchekin2006,lumi2015lumi,Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004,Baur2009,laubsch2009high,hahn2014development}.
-However, these studies typically include limited information on the macro-level chip design and cover disparate aspects of the technology over different time periods.
-Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos,cha2019semiconductor} and Soraa\cite{cich2017high}.
-
-We summarize the information collected, analysed and classified on the progress in LED chip architectures and manufacturing processes from these and other disparate sources in Figure \ref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
-Despite the amount of literature published on the topic of LED history, we are not aware of a prior publication that comprehensively and consistently aggregates and analyzes known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
-In addition, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2020novel,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
-This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012,Creutzig2018}.
 

### 2023-08-10 2_SSL_NE/article/article_supplementary.tex
completed section 1

@@
 
+\section{Previous Literature on Technological Progress in LEDs}
+
+\section{Previous Literature}
+\label{sec:prev_lit}
+
+The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Nakamura2015}.
+The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006,krames2007led,laubsch2009high,hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
+
+Several descriptive publications provide a very useful overview of selected scientific breakthroughs that have contributed to LED progress \cite{krames2007status,Phillips2007,Bierhuizen2007,Nakamura2013,feezell2018invention,Taki2019}.
+Additional studies provide more detail regarding the origin and impact of specific advances in LED technology as well as their integration into device manufacturing were published by authors working for key industry actors Lumileds\cite{MuellerMach2005,Shchekin2006,lumi2015lumi,Bhardwaj2017} and Osram (now amsOsram)\cite{Haerle2004,Baur2009,laubsch2009high,hahn2014development}.
+However, these studies typically include limited information on the macro-level chip design and cover disparate aspects of the technology over different time periods.
+Another important source of historical information are patents, which cover the entirety of the device architecture or manufacturing process, including those by Lumileds\cite{margalith2011thin}, Samsung\cite{jung2014phos,cha2019semiconductor} and Soraa\cite{cich2017high}.
+
+Despite the amount of literature published on the topic of LED history, we are not aware of a prior publication that comprehensively and consistently aggregates and analyzes known chip design, manufacturing, and material improvements to show the overall effect of these improvements on device efficiency or manufacturing cost over time.
+Existing publications, including those mentioned above, are typically subject to one of the following key limitations:
+
+\begin{itemize}
+    \item Publications often look only at progress across a select few LED device sub-efficiencies, often during a very short time period and, most importantly, only at devices manufactured by one specific company, e.g., OSRAM Opto Semiconductors in \cite{hahn2014development} or Lumileds in \cite{Bhardwaj2017}.
+    \item Publications do not report on a number of potentially sensitive or proprietary advances, especially following the market entry of Chinese LED manufacturing companies as a result of the 12th Five-Year Plan (cf.
+\cite{guo2017china}).
+    \item Publications typically do not look at the overall LED efficiency and its underlying physical limits and tend to overlook the metrics of progress related to consumer experience. 
+\end{itemize}
+
+Due to these limitations, any recommendations on future improvements in LED technology based on these studies are fragmented at best.
+Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan_2021}, has not yet been studied in the context of lighting.
+
+In our study, we overcome the limitations of previous literature by, first, synthesizing data provided in multiple publications across an ensemble of various cost, performance and consumer experience metrics for tracking progress in LED technology over an extended period of time.
+Second, our key advantage and contribution compared to these publications is a multi-method approach that we take to fill in the data gaps and augment information obtained from the literature sources by our own performance calculations, a bottom-up cost model and a series of expert interviews.
+This enabled us to synthesize evidence from multiple sources and, for the first time, reconstruct a comprehensive picture of the historical progress in white LED technology throughout its entire history.
+
 \section{Metrics for Technological Progress in LEDs}
 
-This section provides additional background and details for the physical, economic and consumer experience metrics that were chosen to quantify multi-dimensional progress in light-emitting diode technology.
+This Supplementary Note provides additional background and details for the physical, economic and consumer experience metrics that were chosen to quantify multi-dimensional progress in light-emitting diode technology.
 
@@
     \hline
-    Panel B1 (IQE) & own research, compare \cite{zenodo_weinold_led_history} \newline
+    Panel B1 (IQE) & Data extracted from respective device datasheets and scientific publications (for a complete list, compare \cite{zenodo_weinold_led_history}).
+Additional data was extracted from government reports: \newline
 \cite{doe_ssl_multiyear_2006}\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2009}\cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016} \\

### 2023-08-11 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 \section{Previous Literature on Technological Progress in LEDs}
-
-\section{Previous Literature}
 \label{sec:prev_lit}
@@
 
-Historically, the progress in LED technology has commonly be described by pointing to impressive improvements in LED device performance, more specifically device brightness, electrical efficiency and manufacturing cost reductions \cite{Taki2019}.
-However, metrics related to consumer experience, including the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000,Sandahl2006,CAIRD2008,murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
-Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in: 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
+Historically, the progress in LED technology has been commonly described by pointing to impressive improvements in LED device performance, more specifically device brightness, as well as electrical efficiency, and manufacturing cost reductions \cite{Taki2019}.
+However, metrics related to consumer experience from using white LEDs as a lighting source, such as the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
+Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
 Next, we introduce and discuss the metrics that we use to track progress in each of these three areas.

### 2023-08-11 2_SSL_NE/article/article_methods.tex
added edits

@@
 Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model (described in \cref{sec:costmodel}) and verify device performance data.
-
-\begin{table}[H]
-\small
-    \centering
-    \caption{Anonymized list of LED experts interviewed for this study. Abbreviations: sr. - senior}
-    \vspace{5mm}
-    \begin{tabular}{|l|l|l|l|l|}
-    \hline
-        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
-        1 & Academia & Sr. researcher & UK & Epitaxy \\ \hline
-        2 & Industry & Consultant, former sr. researcher & USA & Device architecture \\ \hline
-        3 & Industry & Consultant, former head of R\&D & Germany & Epitaxy \\ \hline
-        4 & Academia & Professor & Austria & Phosphors \\ \hline
-        5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
-        6 & Consulting & Consultant, former sr. technical advisor & USA & Device architecture \\ \hline
-        7 & Academia & Professor & Germany & Phosphors \\ \hline
-        8 & Government & R\&D manager & USA & Device architecture \\ \hline
-        9 & Consulting & Consultant & USA & Device applications \\ \hline
-        10 & Academia & Professor & France & Device physics \\ \hline
-        11 & Industry & Sr scientist, former head of R\&D & USA & Device architecture \\ \hline
-        12 & Industry & Principal scientist & Germany & Phosphors \\ \hline
-        13 & Industry & Former head of R\&D & USA & Phosphors \\ \hline
-    \end{tabular}
-    \label{tab:interviews}
-\end{table}
 

### 2023-08-11 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 Due to these limitations, any recommendations on future improvements in LED technology based on these studies are fragmented at best.
-Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating}\cite{kolesnikov2020novel}\cite{nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan_2021}, has not yet been studied in the context of lighting.
+Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2020novel,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan_2021}, has not yet been studied in the context of lighting.
 
@@
 Historically, the progress in LED technology has been commonly described by pointing to impressive improvements in LED device performance, more specifically device brightness, as well as electrical efficiency, and manufacturing cost reductions \cite{Taki2019}.
-However, metrics related to consumer experience from using white LEDs as a lighting source, such as the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000}\cite{Sandahl2006}\cite{CAIRD2008}\cite{murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
+However, metrics related to consumer experience from using white LEDs as a lighting source, such as the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000,Sandahl2006,CAIRD2008,murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
 Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
@@
 In 2000, the highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
-Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
-Today, however, the devices with the highest performance yield in excess of 1600lm, the equivalent of a 100W incandescent bulb \cite{cree2020bright}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law” \cite{haitz1999case,haitz2011solid}.
+Today, however, the devices with the highest performance yield in excess of 1600 lm, the equivalent of a 100 W incandescent bulb \cite{cree2020bright}.
 LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
-Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving overall LED efficiency \cite{weinold2021compound}.
+Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving the overall LED efficiency \cite{weinold2021compound}.
 Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
@@
 \newpage
+
+\section{Methodological Details}
+
+This Supplementary Note provides additional details for the main methods and data sources used in our mixed-methods research approach described in the Methods section in the main text.
+
+\subsection{Background of Interviewed Experts}
+
+\begin{table}[H]
+\small
+    \centering
+    \caption{Anonymized list of interviewed LED experts.}
+    \vspace{5mm}
+    \begin{tabular}{|l|l|l|l|l|}
+    \hline
+        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
+        1 & Academia & Senior researcher & UK & Epitaxy \\ \hline
+        2 & Industry & Consultant, former senior researcher & USA & Device architecture \\ \hline
+        3 & Industry & Consultant, former head of R\&D & Germany & Epitaxy \\ \hline
+        4 & Academia & Professor & Austria & Phosphors \\ \hline
+        5 & Industry & Consultant, former head of R\&D & USA & Device architecture \\ \hline
+        6 & Consulting & Consultant, former senior technical advisor & USA & Device architecture \\ \hline
+        7 & Academia & Professor & Germany & Phosphors \\ \hline
+        8 & Government & R\&D manager & USA & Device architecture \\ \hline
+        9 & Consulting & Consultant & USA & Device applications \\ \hline
+        10 & Academia & Professor & France & Device physics \\ \hline
+        11 & Industry & Senior scientist, former head of R\&D & USA & Device architecture \\ \hline
+        12 & Industry & Principal scientist & Germany & Phosphors \\ \hline
+        13 & Industry & Former head of R\&D & USA & Phosphors \\ \hline
+    \end{tabular}
+    \label{tab:interviews}
+\end{table}
 

### 2023-08-11 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
 
-\subsection{Consumer Experience Metrics}
+\subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
 
-The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
-Any metric relevant to customer experience can thus be calculated from the spectrum alone.
-The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra  of the down-conversion phosphor used in the device.
-It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
-Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: Colour Rendering Index (CRI) and Colour Temperature.
-We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
-This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
-As a result, it is beyond the scope of this work.
+Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
+\cite{tsao2010solid}} (FVE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
+These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
+It is defined as
+%
+\begin{equation}
+    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
+\end{equation}
+%
+where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
 
-\subsubsection{Color Rendering Index (CRI)}
+\subsubsection{Light Extraction Efficiency $\eta_{LE}$}
 
-The Colour Rendering Index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
-The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
-CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
-However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
-High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
-It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
-However, these niche applications remain outside our focus on general illumination.
-Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
+LED light-extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+%
+\begin{equation}
+    \eta_{LE}=\frac{\text{\# of photons out}}{\text{\# of photons created}} = \frac{P_{opt}}{P_{int}}
+\end{equation}
+%
+where $P_{opt}$ is the optical power of the device and $P_{int}$ is the internal power of the device\cite{schubert2018light}\cite{tsao2010solid}.
 
-\subsubsection{Colour Temperature}
+\subsubsection{Internal Quantum Efficiency $\eta_{IQ}$}
 
-The Colour Temperature of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
-Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
-Early white LEDs produced only cool white light \cite{mueller2000light}.
-The introduction of first commercial warm white LED light bulbs played a significant part in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
-For this reason, we also adopt colour temperature as a metric for tracking progress in consumer experience in LED lighting.
+Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
+%
+\begin{equation}
+\label{eqn:ieq-eqe}
+    \eta_{EQ} = \eta_{IQ} \times \eta_{LE}
+\end{equation}
+%
+It is defined as
+%
+\begin{equation}
+    \eta_{IQ} = \frac{\# \text{of photons created}}{\# \text{of electron-hole pairs in}}
+\end{equation}
+%
+and depends on the current density of the device\cite{schubert2018light}\cite{tsao2010solid}.
 
-\subsection{Manufacturing Cost}
+\subsubsection{Droop $\eta_{droop}$}
 
-In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must highlight the complexity of this task.
-Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
-Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
-However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
-In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
-Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
+Efficiency droop describes the decrease of LED internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
+This energy loss channel is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
+It is defined as
+%
+\begin{equation}
+    \eta_{droop} = 1 - \frac{\eta_{IQ}}{\eta_{IQ}(A \rightarrow 0)}
+\end{equation}
+%
+where $\eta_{IQ}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
+In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$
+%
+\begin{equation}
+\label{eqn:droop}
+    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
+\end{equation}
+%
+Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
 
-We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
-In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
+\subsubsection{Light Conversion Efficiency $\eta_{C}$}
 
-We describe the details of our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model's limitations, in Supplementary Information section 2.4.
-The model structure is generally based on the 2012 LEDCOM cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
-The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
+Light conversion efficiency (LCE) describes the losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
+These losses are the sum of the Stokes loss and well as scattering/absorption losses.
+It is defined as
+%
+\begin{equation}
+    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i}
+\end{equation}
+%
+where $E_i$ denotes the total energy of light at the color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
+Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
+\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.
+
+\subsubsection{Spectral Efficiency $\eta_{S}$}
+
+Spectral efficiency (SE) describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
+Photons converted into the infrared or ultraviolet are lost to illumination purposes.
+It is defined as
+
+\begin{equation}
+    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
+\end{equation}
+
+where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
+$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
 
@@
 
-\subsection{Sub-Efficiencies}
+\subsection{Consumer Experience Metrics}
 
-This subsection provides definitions and equations for the LED device sub-efficiencies introduced in the Metrics section in the main text.
+The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
+Any metric relevant to customer experience can thus be calculated from the spectrum alone.
+The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra of the down-conversion phosphor used in the device.
+It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
+Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: colour rendering index (CRI) and colour temperature (CT).
+We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
+This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
+As a result, it is beyond the scope of this work.
 
-\subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
+\subsubsection{Color Rendering Index (CRI)}
 
-Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
-\cite{tsao2010solid}} (VfE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
-These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
-It is defined as
+The colour rendering index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
+The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
+CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
+\cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
+High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
+It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
+However, these niche applications remain outside our focus on general illumination.
+Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
 
-\begin{equation}
-    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
-\end{equation}
+\subsubsection{Colour Temperature}
 
-where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
+The colour temperature (CT) of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
+Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
+Early white LEDs produced only cool white light \cite{mueller2000light}.
+The introduction of first commercial warm white LED light bulbs played a significant role in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
+For this reason, we also adopt CT as a metric for tracking progress in consumer experience in LED lighting.
 
-\subsubsection{Light Extraction Efficiency $\eta_{LE}$}
+\subsection{Cost Metrics}
 
-Device light-extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must highlight the complexity of this task.
+Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
+Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
+However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
+In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
+Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
 
-\begin{equation}
-    \eta_{LE}=\frac{\text{\# of photons out}}{\text{\# of photons created}} = \frac{P_{opt}}{P_{int}}
-\end{equation}
+We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
+In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
 
-where $P_{opt}$ is the optical power of the device and $P_{int}$ is the internal power of the device\cite{schubert2018light}\cite{tsao2010solid}.
+We describe the details of our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model's limitations, in Supplementary Information section 2.4.
+The model structure is generally based on the 2012 LEDCOM cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
 
-\subsubsection{Internal Quantum Efficiency $\eta_{IQ}$}
-
-Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
-\begin{equation}
-\label{eqn:ieq-eqe}
-    \eta_{EQ} = \eta_{IQ} \times \eta_{LE}
-\end{equation}
-
-and defined as
-
-\begin{equation}
-    \eta_{IQ} = \frac{\# \text{of photons created}}{\# \text{of electron-hole pairs in}}
-\end{equation}
-
-and depends on the current density of the device\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Droop $\eta_{droop}$}
-
-Efficiency droop describes the decrease of device internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
-This is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
-It is defined as
-
-\begin{equation}
-    \eta_{droop} = 1 - \frac{\eta_{IQE}}{\eta_{IQE}(A \rightarrow 0)}
-\end{equation}
-
-where $\eta_{IQE}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
-In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$ as
-
-\begin{equation}
-\label{eqn:droop}
-    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
-\end{equation}
-
-Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
-
-\subsubsection{Conversion Efficiency $\eta_{C}$}
-
-Light conversion efficiency (LCE) describes losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
-These losses are the sum of the Stokes loss and well as scattering/absorption losses.
-It is defined as
-
-\begin{equation}
-    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i}
-\end{equation}
-
-where $E_i$ denotes the total energy of light at the color corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
-Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all components, i.e.
-\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.
-
-\subsubsection{Spectral Efficiency $\eta_{S}$}
-
-Spectral efficiency (SE) describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
-Photons converted into the infrared or ultraviolet are lost to illumination purposes.
-It is defined as
-
-\begin{equation}
-    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
-\end{equation}
-
-where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
-$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
-
-\newpage
-
+\clearpage
 \section{Methodological Details}

### 2023-08-12 2_SSL_NE/article/article_methods.tex
added edits

@@
 Yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
-
-\subsubsection{Structure of the Model}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.95\textwidth]{./figures/cost_model.pdf}
-    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
-For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.
-For the computation of total cost, see \cref{subsec:cost_computation}.
-For the computation of changes in total cost, see \cref{sec:contribution_variables}.}
-    \label{fig:costmodel-schematic}
-\end{figure}
-
-The cost model adapted for this publication is a microeconomic manufacturing cost model.
-Within the timeframe and scope laid out in the main publication, it returns the total manufacturing cost of phosphor converted warm white light-emitting diode packages.
-In this computation, it considers the main economic factors associated with operating and maintaining manufacturing equipment.
-It does not consider costs associated with research and development or those associated with the construction of manufacturing facilities.
-It considers market trends through their effect on manufacturing parameters.
-
-A schematic diagram of the cost model is presented in \cref{fig:costmodel-schematic}.
-The cost model is process step-based.
-It is split between the two stages of the manufacturing process: the first stage combining operations at the wafer level, followed by the LED packaging stage.
-The model takes as inputs parameters specific to individual manufacturing process steps (\textit{"manufacturing step parameters"}) and parameters affecting all manufacturing steps (\textit{"global parameters"}).
-The cost for each process step is then computed.
-The cost model returns returns the costs of individual manufacturing steps as well as the total manufacturing cost.
-It further considers the yield per step and returns the cumulative yield, the yielded cost per step and the yielded total manufacturing cost.
-
-\subsubsection{Manufacturing Process Steps by Chip Architecture}
-
-\cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering of the manufacturing process of three different chip architectures considered in the cost model.
-
-    %CLASSICAL CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
-            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003. Abbreviations: MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
-            \label{fig:manuf_classical_2003}
-        \end{figure}
-    \end{landscape}
-
-    %VERTICAL THIN-FILM FLIP-CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
-            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
-Abbreviations: MQW - multiple quantum well; LLO - laser lift-off.
-Continued on next page.}
-            \label{fig:manuf_vtf_2012-1}
-        \end{figure}
-    \end{landscape}
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
-            \caption{(2/2) Continued from previous page. Abbreviations: Encaps. - encapsulation.}
-            \label{fig:manuf_vtf_2012-2}
-        \end{figure}
-    \end{landscape}
-
-    % CHIP-SCALE PACKAGE FLIP-CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
-            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
-Abbreviations: MQW - multiple quantum well.
-Continued on next page.}
-            \label{fig:manuf_csp_2020-1}
-        \end{figure}
-    \end{landscape}
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
-            \caption{(2/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
-Continued from previous page.
-Abbreviations: Encaps.
-- encapsulation.}
-            \label{fig:manuf_csp_2020-2}
-        \end{figure}
-    \end{landscape}
-
-\subsubsection{Computation of Manufacturing Cost}
-\label{subsec:cost_computation}
-
-The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level.
-The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
-\begin{equation}
-\label{eqn:cost_sum}
-    C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
-\end{equation}
-where
-\begin{align*}
-    P_S &\dots \text{sapphire substrate price per die} \\
-    C_w &\dots \text{wafer processing cost per die} \\
-    C_p &\dots \text{die processing cost}
-\end{align*}
-The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.
-\begin{equation}
-        C_w = \sum_i C_i
-\end{equation}
-\begin{equation}
-	C_p = \sum_j C_j
-\end{equation}
-The cost of a single process step $C_i$ can now be written as
-\begin{equation}
-\label{eqn:cost_wafer}
-    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{DPW}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
-\end{equation}
-where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
-\begin{align*}
-        DPW &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
-        y &\dots \text{process step yield} \\
-        u &\dots \text{equipment utilization (relative to theoretical equipment capacity)} \\
-        p &\dots \text{power consumption} \\
-        c_e &\dots \text{hourly electricity cost} \\
-        c_m &\dots \text{hourly maintenance cost} \\
-        c_d &\dots \text{hourly depreciation cost}\\
-        c_l &\dots \text{hourly labour cost} \\
-        c_o &\dots \text{hourly overhead cost} \\
-        t_i &\dots \text{time per run} \\
-        w &\dots \text{wafers per run} \\
-        A &\dots \text{wafer area} \\
-        v_x &\dots \text{volume of material $x$ per wafer} \\
-        c_x &\dots \text{cost of material $x$ per volume}\\
-\end{align*}
-The number of die per wafer $D$ depends on the total usable wafer area.
-The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
-\begin{equation}
-	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
-\end{equation}
-Determining the usable wafer area as a function of these three parameters requires a numerical solution.
-However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
-\begin{equation}
-\label{eqn:dpw}
-	DPW = \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
-\end{equation}
-where
-\begin{align*}
-    d &\dots \text{wafer diameter} \\
-    e &\dots \text{wafer edge exclusion zone width} \\
-    a &\dots \text{die area} \\
-    s &\dots \text{cutting street width} \\
-\end{align*}
 

### 2023-08-12 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 It is defined as
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
@@
 The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
-
+%
 \begin{myequation}
@@
 \end{myequation}
-
+%
 where
-
+%
 \begin{align*}
@@
 \end{align*}
-
+%
 This metric can be computed from spectral data alone and does not require additional spectral normalization.
@@
 The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power \cite{cie-term-effsrc}
-
+%
 \begin{myequation}
@@
 \end{myequation}
-
+%
 This metric is often cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
@@
 
-In selecting metrics for tracking the progress in manufacturing cost reductions in LED, we must highlight the complexity of this task.
+In selecting the metrics for tracking progress in manufacturing cost reductions in LED, we must first highlight the complexity of this task.
 Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
@@
 In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
-Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress very difficult in practice.
+Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress difficult in practice.
 
@@
 
-We describe the details of our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model's limitations, in Supplementary Information section 2.4.
-The model structure is generally based on the 2012 LEDCOM cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
-The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-The aggregate result of the cost model is the manufacturing cost per LED package for each of the three years considered, which includes all costs associated with producing the chip, including running costs of the factory.
+We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:method_details}.
+The outcome of the cost model is the cumulative yielded manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
 
@@
 \section{Methodological Details}
+\label{sec:method_details}
 
@@
 
-\clearpage
-\section{Additional Results}
+\subsection{Manufacturing Cost Model Details}
 
-This section provides additional background and details for the results of our study presented in the Results section in the main text.
+\subsubsection{Structure of the Model}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.95\textwidth]{./figures/cost_model.pdf}
+    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
+For the description of cost variables, see definitions in \cref{eqn:cost_wafer}.
+For the computation of total cost, see \cref{subsec:cost_computation}.
+For the computation of changes in total cost, see \cref{sec:contribution_variables}.}
+    \label{fig:costmodel-schematic}
+\end{figure}
+
+The LED cost model developed for this publication is a microeconomic manufacturing cost model.
+Within the timeframe and scope laid out in the main text of this publication, it returns the total manufacturing cost of GaN-based phosphor-converted warm white light-emitting diode packages.
+The model considers main economic factors associated with operating and maintaining manufacturing equipment.
+It does not consider costs associated with research and development or those associated with the construction of manufacturing facilities.
+It considers market trends through their effect on manufacturing parameters.
+
+A schematic diagram of the cost model is presented in \cref{fig:costmodel-schematic}.
+The cost model is process step-based.
+It is split between the two stages of the manufacturing process: the first stage combining operations at the wafer level, followed by the LED packaging stage.
+The model takes as inputs parameters specific to individual manufacturing process steps (\textit{"manufacturing step parameters"}) and parameters affecting all manufacturing steps (\textit{"global parameters"}).
+The cost for each process step is then computed.
+The cost model returns the costs of individual manufacturing steps as well as the total manufacturing cost.
+It further considers the yield per step and returns the cumulative yield, the yielded cost per step and the yielded total manufacturing cost.
+
+\subsubsection{Manufacturing Process Steps by Chip Architecture}
+
+\cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering for the manufacturing process of three different chip architectures considered in the cost model.
+
+    %CLASSICAL CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
+            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003. MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
+            \label{fig:manuf_classical_2003}
+        \end{figure}
+    \end{landscape}
+
+    %VERTICAL THIN-FILM FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
+            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
+MQW - multiple quantum well; LLO - laser lift-off.
+Continued in \cref{fig:manuf_vtf_2012-2}}
+            \label{fig:manuf_vtf_2012-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
+            \caption{(2/2) Continued from \cref{fig:manuf_vtf_2012-1}. Encaps. - encapsulation.}
+            \label{fig:manuf_vtf_2012-2}
+        \end{figure}
+    \end{landscape}
+
+    % CHIP-SCALE PACKAGE FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
+            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+MQW - multiple quantum well.
+Continued in \cref{fig:manuf_csp_2020-2}.}
+            \label{fig:manuf_csp_2020-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
+            \caption{(2/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+Continued from \cref{fig:manuf_csp_2020-1}.
+Encaps.
+- encapsulation.}
+            \label{fig:manuf_csp_2020-2}
+        \end{figure}
+    \end{landscape}
+
+\subsubsection{Computation of Manufacturing Cost}
+\label{subsec:cost_computation}
+
+The manufacturing process of semiconductor devices can be categorized by the level at which manufacturing process steps are implemented, i.e., either at the wafer level or at the individual chip/package level.
+The total manufacturing cost per die is thus the sum of the total costs of all wafer processing steps and all die packaging steps.
+%
+\begin{equation}
+\label{eqn:cost_sum}
+    C \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = P_S + C_w + C_p
+\end{equation}
+%
+where
+%
+\begin{align*}
+    P_S &\dots \text{sapphire substrate price per die} \\
+    C_w &\dots \text{wafer processing cost per die} \\
+    C_p &\dots \text{die processing cost}
+\end{align*}
+%
+The total wafer processing cost and total die packaging costs are in turn the sum of all associated process steps.
+%
+\begin{equation}
+        C_w = \sum_i C_i
+\end{equation}
+%
+\begin{equation}
+	C_p = \sum_j C_j
+\end{equation}
+%
+The cost of a single process step $C_i$ can now be written as
+%
+\begin{equation}
+\label{eqn:cost_wafer}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] =\frac{1}{DPW}  \frac{1}{y_i}   \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
+\end{equation}
+%
+where the index $i$ runs over all wafer processing steps, the index $j$ runs over all die processing steps and the index $x$ run over all materials.
+%
+\begin{align*}
+        DPW &\dots \text{number of functional (i.e., successfully tested) die per wafer}\\
+        y &\dots \text{process step yield} \\
+        u &\dots \text{equipment utilization (relative to theoretical equipment capacity)} \\
+        p &\dots \text{power consumption} \\
+        c_e &\dots \text{hourly electricity cost} \\
+        c_m &\dots \text{hourly maintenance cost} \\
+        c_d &\dots \text{hourly depreciation cost}\\
+        c_l &\dots \text{hourly labour cost} \\
+        c_o &\dots \text{hourly overhead cost} \\
+        t_i &\dots \text{time per run} \\
+        w &\dots \text{wafers per run} \\
+        v_x &\dots \text{volume of material $x$ per wafer} \\
+        c_x &\dots \text{cost of material $x$ per volume}\\
+\end{align*}
+%
+The number of die per wafer $D$ depends on the total usable wafer area.
+The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
+%
+\begin{equation}
+	A_{\text{usable}}=A_{\text{wafer}}-A_{\text{cut}}-A_{\text{exclusion}}
+\end{equation}
+%
+Determining the usable wafer area as a function of these three parameters requires a numerical solution.
+However, following discussions in the literature \cite{de2005investigation}, we approximate the number of functional die per wafer\footnote{often \textit{"good die per wafer"} in the literature} as
+%
+\begin{equation}
+\label{eqn:dpw}
+	DPW = \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2}
+\end{equation}
+%
+where
+%
+\begin{align*}
+    d &\dots \text{wafer diameter} \\
+    e &\dots \text{wafer edge exclusion zone width} \\
+    a &\dots \text{die area} \\
+    s &\dots \text{cutting street width} \\
+\end{align*}
 
@@
 \clearpage
-\section{Complete List of Sources for Figures}
+\section{List of Sources for Figures}
 \label{sec:sources}
 
-\subsection{Figure 1 (Historical Development of Luminous Efficacy)}
-
-The list of sources for Figure 1 in the main article is organized by different technologies:
+\subsection{Figure 1 (Historical Progress in Luminous Efficacy of Lighting Technologies)}
 
 \begin{table}[h!]
+    \caption{List of sources for Figure 1 organized by technology.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}
@@
     \end{NiceTabularX}
+    \caption{Note: CFL - Compact Fluorescent Lamp. HID - High-Intensity Discharge.}
 \end{table}
 
-\subsection{Figure 2 (Historical Development of Lamp Prices)}
+\subsection{Figure 2 (Historical Development of LED Lamp Prices)}
 
@@
 \begin{table}[h!]
+    \caption{List of sources for Figure 2 grouped by data provider.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}
     \hline
-    \textit{Provider of References} & \textit{References} \\
+    \textit{Source} & \textit{References} \\
     \hline
@@
     \hline
-    Government Report & \cite{council2013assessment} \\
+    Government Report (US) & \cite{council2013assessment} \\
     \hline
@@
 
-The sources for Figure 3 in the main article are grouped by their type:
-
 \begin{table}[h!]
+\caption{List of sources for Figure 3 grouped by publication type.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}
     \hline
-    \textit{Type of Source} & \textit{References} \\
+    \textit{Publication Type} & \textit{References} \\
     \hline
@@
 
-\subsection{\cref{fgr:subeff} (Historical Developments in Device Sub-Efficiencies)}
-
-The sources for \cref{fgr:subeff} are grouped by sub-efficiencies represented on different panels of the figure:
+\subsection{\cref{fgr:subeff} (Historical Developments in LED Sub-Efficiencies)}
 
 \begin{table}[h!]
+    \caption{List of sources for \cref{fgr:subeff} grouped by LED sub-efficiencies.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}
@@
     \hline
-    Panel A1 ($V_f$) & \cite{nichia2001data}\cite{lumi2002data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
+    Panel A1 (VFE) & \cite{nichia2001data}\cite{lumi2002data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
     \hline

### 2023-08-12 2_SSL_NE/article/article_methods.tex
added edits

@@
 \newpage
-\subsubsection{Preliminary Sensitivity Analysis}
-
-\begin{figure}[]
-	\centering
-    \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
-	\caption{Sensitivity analysis for selected cost model parameters.
-Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in \cref{tab:sensitivity}.
-Panel b shows the resulting range of variation in the manufacturing cost.
-Ranges are represented for three years considered in the cost model, where corresponding data is available, by whiskers of the following colour: 2003 – blue, 2012 – green, 2020 – red.
-Note the overall trend of decreasing sensitivity to cost model parameters as the number of die per wafer increases over time.
-Abbreviations: resp.
-- respective; yield sav.
-from insp.
-- yield savings from inspection.}
-	\label{fig:sensitivity}
-\end{figure}
-
-A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
-The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
-As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen because the prices quotes from the United State Geological Survey price database.
-Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
-The upper range was chosen because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
-The results of the analysis are shown in  \cref{fig:sensitivity}.
-The cost model is generally more sensitive to the variation in parameters at smaller wafer diameters.
-The most sensitive parameters are global parameters, such as yield or average equipment throughput.
-
-\begin{table}[]
-\small
-    \caption{Cost model sensitivity analysis parameter list. The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.}
-    \vspace{5mm}
-    \begin{NiceTabularX}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
-        \hline
-            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & Source \\
-        \hline
-            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & \cite{mddi1997cleanroom}\cite{ledcomv2} \newline \cite{bakshi2009euv}\cite{gajera2006process} \\
-        \hline
-            Person-hour & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
-        \hline
-            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I \newline \cite{Appleyard_2001} \\
-        \hline
-            Overall Yield & \% & 100\% & +25,-25 & 100\% & +25,-15 & 100\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
-        \hline
-            Inspec. Yield Savings & \%/inspec. & 0.5\% & +80,-80 & 0.5\% & +80,-80 & 0.5\% & +80,-80 & \cite{mckinseyyield} \\
-        \hline
-            Overall Throughput & UPH or h$^{-1}$ & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & Datasheets \\
-        \hline
-            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I and \cref{fig:wafers} \\
-        \hline
-            Edge Exclusion & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
-        \hline
-            Cutting Width & $\mu$m & 100 & +50,-25 & 75 & +33.3,-20 & 20 & +300,-50 & \cite{masaki2000division}\cite{ils2005width} \newline \cite{photonics2010width}\cite{discowidth} \\
-        \hline
-            Metal Prices & \text{USD}/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
-        \hline
-            Electricity Price & \text{USD}/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
-        \hline
-            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & \cref{fig:wafers} \\
-        \hline
-            Phosphor Prices & \text{USD}/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
-        \hline
-        \end{NiceTabularX}
-    \vspace{5mm}
-
-    Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
-If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
-Abbreviations: FTE - full-time equivalent; UPH - units per hour; Equip.
-Discount - equipment discount (sales rebate for large purchases); Inspec.
-Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).
-    \label{tab:sensitivity}
-\end{table}
-
-\newpage
 \subsection{Computation of Contribution of Individual Variables}

### 2023-08-12 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 
-\subsection{Luminous Efficacy of Radiation}
+\subsubsection{Luminous Efficacy of Radiation}
 \label{subsec:ler}
 
-Figure 1 in the main publication uses luminous efficacy of radiation as the primary metric to describe progress in lighting technologies.
-Care must be taken not to confuse this metric with luminous efficacy of source, which is used in \cref{fig:phosphor_spectrum}.
-
-This metric describes the match of a light-emitting diode package spectrum to the human visual system.
+The luminous efficacy \textit{of radiation} $K$ describes the match of a light-emitting diode package spectrum to the human visual system.
 Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
@@
 \begin{align*}
-    K &\dots \text{spectral luminous efficacy} \\
+    K &\dots \text{luminosity function} \\
     \phi &\dots \text{spectral radiant flux} \\
@@
 Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation.
-Care must be taken not to confuse this efficacy metric with the \textit{efficacy of source} described in the following subsection.
 
-\subsection{Luminous Efficacy of Source}
+Care must be taken not to confuse this metric with luminous efficacy of source described in \cref{subsec:les}.
+
+\subsubsection{Luminous Efficacy of Source}
 \label{subsec:les}
 
-The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power \cite{cie-term-effsrc}
+The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power $P_{el}$ \cite{cie-term-effsrc}.
+It is often used to describe general progress in lighting technologies.
 %
@@
 %
-This metric is often cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
-Care must be taken not to confuse this efficacy metric with the \textit{luminous efficacy of radiation}, which depends only on the spectral characteristics of a light source.
+This metric is also frequently cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
+However, as noted in \cref{subsec:ler}, care must be taken not to confuse this efficacy metric with the \textit{luminous efficacy of radiation}, which depends only on the spectral characteristics of a light source.
 As the luminous efficacy of a light source $\eta$ captures the overall device efficacy, it depends on a large number of other device properties and parameters.
 This makes attribution of changes in this metric to individual changes in device design or manufacturing difficult.
-For this reason, we do not use this metric in our study.
+For this reason, we do not use this metric only to show historical progress across different lighting technologies in Figure 1 in the main text of the article.
 
@@
 
+\clearpage
+\section{Identified LED Innovations and Performance Improvements}
+
+This Supplementary Note provides details for the identified LED innovations and performance improvements that contributed to the historical progress in LED technology across efficiency and consumer experience metrics discussed in the main text.
+
 \subsection{Historical Progress in Sub-Efficiencies}
@@
 
-\clearpage
-\subsection{LED Innovations}
+\subsection{Non-Phosphor-Related LED Innovations}
 
-\subsubsection{Non-Phosphor-Related}
-
-\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies.
-Note that phosphor-related LED innovations affecting consumer experience metrics, along with spectral efficiency, are presented separately in Table 1 in the main publication and discussed in detail in \cref{sec:innovation_phosphor}.
+\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting four white LED sub-efficiencies: forward voltage efficiency, light extraction efficiency, internal quantum efficiency, droop and light conversion efficiency.
+Note that this table does not include phosphor-related LED innovations affecting spectral efficiency, as these are presented separately in Table 1 in the main text as affecting consumer experience metrics and discussed in detail in \cref{sec:innovation_phosphor}.
 
@@
     \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
+        \hline
         \multicolumn{5}{c}{Forward Voltage Efficiency (VFE) $\eta_{Vf}$} \\
@@
         \hline
+        \multicolumn{5}{c}{Droop $\eta_{droop}$} \\
+        \hline
+            ongoing & Lower Current Density & Current Density & No & I \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect Getting \newline Underlayer & Defect Density & No & I, \cite{haller2017burying} \\
+        \hline
         \multicolumn{5}{c}{Light Conversion efficiency (CE) $\eta_{C}$} \\
@@
         \hline
-            $<$ 2017 & Defect getting \newline underlayer & Defect Density & No & I, \cite{haller2017burying} \\
+            $<$ 2017 & Defect gettering \newline underlayer & Defect Density & No & I, \cite{haller2017burying} \\
         \hline
         \end{NiceTabularX}
-    \vspace{5mm}
-
-     Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
+    \caption*{Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
 'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
 \textit{Spillover} column indicates if the innovation involved technology spillovers.
-\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.
+\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
     \label{tab:innovations}
@@
 
-\subsubsection{Phosphor-Related}
+\subsection{Phosphor-Related LED Innovations}
 \label{sec:innovation_phosphor}
@@
 This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics.
+These include a detailed description of the history of those innovations, which we use to identify the details of involved technology spillovers.
+In addition, spectral data for the first identified LED products that used those innovations is presented in \cref{fig:phosphor_spectrum}, which we use to calculate consumer experience metrics, as well as LED spectral efficiency.
 
-\paragraph{YAG and YGAG Phosphors}
+\subsubsection{YAG and YGAG Phosphors}
 
@@
 
-\paragraph{258, SLA and SALON phosphors}
+\subsubsection{258, SLA and SALON phosphors}
 
@@
 
-\paragraph{PFS Phosphor}
+\subsubsection{PFS Phosphor}
 
@@
 
-\paragraph{Quantum Dots for Light Down-Conversion}
+\subsubsection{Quantum Dots for Light Down-Conversion}
 
@@
 \clearpage
-\paragraph{Spectral Data for Identified LED Phosphor Innovations}
-
-\hspace{1mm}
+\subsubsection{Spectral Data for Identified LED Phosphor Innovations}
 
@@
 \newpage
-\subsection{Cost Comparison with Reported Industry Data and DOE Projections}
+\section{Cost Model Sensitivity Analysis}
 
-We conducted a comparison of the LED manufacturing cost structure produced by our cost model with previously reported US DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions (see \cref{fig:costmodel_calibration}).
+\begin{figure}[ht!]
+	\centering
+    \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
+	\caption{Sensitivity analysis for selected cost model parameters.
+Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in Supplementary Table 3.
+Panel b shows the resulting range of variation in the manufacturing cost.
+Ranges are represented for three years considered in the cost model, where corresponding data is available, by whiskers of the following colour: 2003 – blue, 2012 – green, 2020 – red.
+Note the overall trend of decreasing sensitivity to cost model parameters as the number of die per wafer increases over time.
+Abbreviations: resp.
+- respective; Yield Sav.
+from Insp.
+- yield savings from inspection.}
+	\label{fig:sensitivity}
+\end{figure}
+
+A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
+The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
+As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen because the prices quotes from the United State Geological Survey price database \cite{usgeoprices}.
+Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
+The upper range was chosen because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
+The results of the analysis are shown in  \cref{fig:sensitivity}.
+The cost model is generally more sensitive to the variation in parameters at smaller wafer diameters.
+The most sensitive parameters are global parameters, such as yield or average equipment throughput.
+
+\begin{table}[]
+\small
+    \caption{Cost model sensitivity analysis parameter list. The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.}
+    \vspace{5mm}
+    \begin{NiceTabularX}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
+        \hline
+            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & Source \\
+        \hline
+            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & \cite{mddi1997cleanroom}\cite{ledcomv2} \newline \cite{bakshi2009euv}\cite{gajera2006process} \\
+        \hline
+            Person-hour & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
+        \hline
+            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I \newline \cite{Appleyard_2001} \\
+        \hline
+            Overall Yield & \% & 100\% & +25,-25 & 100\% & +25,-15 & 100\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
+        \hline
+            Inspec. Yield Savings & \%/inspec. & 0.5\% & +80,-80 & 0.5\% & +80,-80 & 0.5\% & +80,-80 & \cite{mckinseyyield} \\
+        \hline
+            Overall Throughput & UPH or h$^{-1}$ & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & Datasheets \\
+        \hline
+            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I and \cref{fig:wafers} \\
+        \hline
+            Edge Exclusion & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
+        \hline
+            Cutting Width & $\mu$m & 100 & +50,-25 & 75 & +33.3,-20 & 20 & +300,-50 & \cite{masaki2000division}\cite{ils2005width} \newline \cite{photonics2010width}\cite{discowidth} \\
+        \hline
+            Metal Prices & \text{USD}/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
+        \hline
+            Electricity Price & \text{USD}/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
+        \hline
+            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & \cref{fig:wafers} \\
+        \hline
+            Phosphor Prices & \text{USD}/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
+        \hline
+        \end{NiceTabularX}
+    \caption*{Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
+If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
+Abbreviations: FTE - full-time equivalent; UPH - units per hour; Equip.
+Discount - equipment discount (sales rebate for large purchases); Inspec.
+Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).}
+    \label{tab:sensitivity}
+\end{table}
+
+\clearpage
+\section{Cost Comparison with Reported Industry Data and DOE Projections}
+
+We conduct a comparison of the outcomes of our LED manufacturing cost model regarding the evolution of the cost structure of low-to-mid-power classic chip GaN phosphor-converted white LED with previously reported U.S.
+DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions (see \cref{fig:costmodel_calibration}).
 We note some differences between the results of our model and the cost structure reported or projected by the DOE.
 For instance, the share of the epitaxy step is consistently larger in the DOE data.
-This can in part be explained by our model relying on state-of-the-art equipment at a virtual US-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
-In addition, the manufacturing lines of the majority of manufacturers is located in Asia.
-We also note that the share of the substrate price in the DOE data is much larger than in our model, which can in part be explained by the overestimation of the actual price of sapphire wafers in earlier projections for 2015-2020.
+This can in part be explained by our model relying on state-of-the-art equipment at a virtual U.S.-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+In addition, the manufacturing lines of the majority of manufacturers are located in Asia.
+We also note that the share of the sapphire substrate price in the DOE data is much larger than in our model, which can in part be explained by the overestimation of the actual price of sapphire wafers in earlier projections made for 2015-2020.
 Finally, the relative importance of the packaging part of the manufacturing process is very similar in our model and in the DOE results in 2012.
 However, in the DOE projections it significantly decreases by 2020, while in our model it retains and even increases its share of the total cost.
-This trend has been independently confirmed by researchers and industry reports on wafer-level packaging \cite{Lee2011WPL}\cite{Xie2013}\cite{ledsmag2017WLP}, showing better performance of our cost model compared to earlier DOE model projections
+This trend has been independently confirmed by researchers and industry reports on wafer-level packaging \cite{Lee2011WPL,Xie2013,ledsmag2017WLP}, showing better performance of our cost model compared to earlier DOE model projections.
 
@@
     \includegraphics[width=\textwidth]{./figures/costmodel_calibration.pdf}
-	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel).
-Hatched bars are projections given by US DOE for 2018 and 2020.
+	\caption{Top panel: manufacturing cost calculations (2008-2016) and projections (2018, 2020) previously published by the U.S.
+DOE based on industry data and LEDCOM model.
+Bottom panel: manufacturing cost structure produced by our cost model for 2002, 2012 and 2020.
+Hatched bars are DOE projections.
 Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
@@
     \end{NiceTabularX}
-    \caption{Note: CFL - Compact Fluorescent Lamp. HID - High-Intensity Discharge.}
+    \caption*{Note: CFL - Compact Fluorescent Lamp. HID - High-Intensity Discharge.}
 \end{table}

### 2023-08-12 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 \begin{equation}
-    \eta_{LE}=\frac{\text{\# of photons out}}{\text{\# of photons created}} = \frac{P_{opt}}{P_{int}}
+    \eta_{LE} = \frac{P_{opt}}{P_{int}} = \frac{\text{\# of photons out/s}}{\text{\# of photons created/s}}
 \end{equation}
 %
-where $P_{opt}$ is the optical power of the device and $P_{int}$ is the internal power of the device\cite{schubert2018light}\cite{tsao2010solid}.
+where $P_{opt}$ is the optical power emitted from the device $P_{int}$ is the optical power emitted from the active region into other parts of the device (internal) \cite{schubert2018light}\cite{tsao2010solid}.
 
@@
 
-Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is related to the external quantum efficiency through
+Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is defined as
 %
 \begin{equation}
-\label{eqn:ieq-eqe}
-    \eta_{EQ} = \eta_{IQ} \times \eta_{LE}
+    \eta_{IQ} = \frac{\# \text{of photons created/s}}{\# \text{of electron-hole pairs in/s}}
 \end{equation}
-%
-It is defined as
-%
-\begin{equation}
-    \eta_{IQ} = \frac{\# \text{of photons created}}{\# \text{of electron-hole pairs in}}
-\end{equation}
-%
-and depends on the current density of the device\cite{schubert2018light}\cite{tsao2010solid}.
 
@@
 Light conversion efficiency (LCE) describes the losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
-These losses are the sum of the Stokes loss and well as scattering/absorption losses.
+These losses are the sum of the Stokes loss and  scattering/absorption losses.
 It is defined as
@@
 
-We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:method_details}.
+We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:costmodel_details}.
 The outcome of the cost model is the cumulative yielded manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
@@
 \subsection{Manufacturing Cost Model Details}
+\label{sec:costmodel_details}
 
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{./figures/cost_model.pdf}
+    \includegraphics[width=0.95\textwidth]{./figures/costmodel.pdf}
     \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.

### 2023-08-12 2_SSL_NE/article/article_methods.tex
added edits

@@
 
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
-    \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2002)=50$mm$\rightarrow5370$DPW, $d(2020)=100$mm$, \rightarrow26,838$DPW.
-Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews.
-For detailed statistics on changes in wafer size, compare also \cref{fig:wafers}.}
-    \label{fig:dpw}
-\end{figure}
-
-Results are plotted in \cref{fig:dpw}, showing the increase in die per wafer (DPW) over time. \cref{eqn:dpw} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
-\begin{equation}
-\label{eqn:cost_wafer_full}
-\begin{split}
-    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
-    &  \frac{1}{y_i}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
-\end{split}
-\end{equation}
-and the cost of a manufacturing step $C_j$ in the packaging category
-\begin{equation}
-\label{eqn:cost_die}
-    C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x c_x \bigg\}
-\end{equation}
-where $c_j$ is $\text{throughput}^{-1}$. The total cost is thus
-\begin{equation}
-\label{eqn:cost_total}
-\begin{split}
-    C= P_s &+ \sum_i \bigg \{ \frac{1}{DPW} \frac{1}{y_i} \bigg[ \frac{t_i}{w_i u_i} \bigg((e*p) + l + m + d +o \bigg)_i  + \sum_{x} v_x p_x \bigg] \bigg \} + \\
-    & + \sum_j \bigg \{ \frac{1}{y_j} \bigg[ \frac{c_j}{u_j}  \bigg((e*p) + l + m + d + o \bigg)_i + \sum_{x} a v_x p_x \bigg ] \bigg\}
-\end{split}
-\end{equation}
-Note that in keeping with the categorization introduced by the United States Department of Energy (cf.
-eg.
-\cite{doe_ssl_rnd_2019}), certain steps from these two categories are reported separately.
-In the wafer processing category, the epitaxy step is reported separately due to its complexity and the large share of cost carried.
-In the wafer processing category, the phosphor step is reported separately.
-
-\subsubsection{Computation of Yielded Cost}
-
-Devices may be damaged or otherwise rendered unusable during the manufacturing process.
-The ratio between the number of good devices per step and the number of handled devices per step is known as the yield.
-Optimizing this yield is critical for reducing manufacturing cost \cite{Kumar2006}.
-This is because cumulative yield quickly drops as the yield from manufacturing steps with below 100\% yield is multiplied.
-We must thus consider not only the manufacturing cost per process step, but also the cost including the yield \cite{becker2001use}\cite{becker2001using}.
-While there are different mathematical approaches to including yield, we follow the definition in \cite{becker2001use}.
-We write for the yielded cost $C_{Y_i}$ of a step $i$ with associated cost (before considering yield) $C_1$ and yield $Y_1$:
-
-\begin{equation}
-\label{eqn:C_2}
-    C_{Y_1} = \frac{C_1}{Y_1}
-\end{equation}
-\begin{equation}
-    C_{Y_2} = \frac{C_1 + C_2}{Y_1 Y_2} - C_{Y_1} = \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} = \frac{1}{Y_1 Y_2} \bigg ( C_1 (1-Y_2) +C_2 \bigg)
-\end{equation}
-\begin{equation}
-    C_{Y_i} = \frac{ \sum_{x \leq i} C_x }{ \prod_{x \leq i} Y_x } - \frac{ \sum_{x<i} C_x }{ \prod_{x<i} Y_x }
-\end{equation}
-
-If a step is applied more than once, we can conveniently rewrite this in a form suited to computation within the \textit{Excel} worksheet.
-Assuming step $2$ is used twice, we get for the yielded cost of this step an equation of the form
-
-\begin{equation}
-\label{eqn:C_2^2}
-    C_{Y_2}^{(2 \times)} = \bigg( \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} \bigg) + \bigg( \frac{C_1 + 2 C_2}{Y_1 Y_2^2} - \frac{C_1 + C_2}{Y_1 Y_2}     \bigg)
-\end{equation}
-\begin{equation}
-    = \frac{1}{Y_1 Y_2^2} \bigg( C_1 (1-Y_2^2) +2C_2 \bigg)
-\end{equation}
-
-which can be compared to  \cref{eqn:C_2} to find a more general form:
-
-\begin{equation}
-    C_{Y_2}^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
-\end{equation}
-
-It can be shown by induction that the general form of a term for a step $i>1$ repeated $n$ times can be expressed as:
-
-\begin{equation}
-\label{eqn:yielded_cost}
-    C_{Y_{i>1}}^{(n \times)} = \frac{1}{Y_i^{n-1} \prod_{x \leq i} Y_x} \bigg( nC_i + \sum_{x < i} C_x (1-Y_i^n) \bigg)
-\end{equation}
-
-This cumulative approach to yielded cost is different from the approach taken in the original \textit{LEDCOM} model.
-It uses what in the literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}.
-In this approach, the yielded cost of a single process step $f$ is described as
-
-\begin{equation}
-    f_\text{single} = \frac{i+s}{y}
-\end{equation}
-
-where
-
-\begin{align*}
-    i &\dots \text{material cost of previous step} \\
-    s &\dots \text{step cost}
-\end{align*}
-
-For process steps that are performed more than once, a series expression is used
-
-\begin{equation}
-    f_{\times 2} =  \frac{\frac{i+s}{y}+s}{y}
-\end{equation}
-\begin{equation}
-\label{eqn:series}
-    f_{\times n} = \frac{i + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
-\end{equation}
-
-This itemized approach serves as a convenient approximation, but its cumulative contributions do not equal the total yielded cost of the entire manufacturing process
-
-\begin{equation}
-    f_\text{total} = \frac{\sum_i^n s_i}{\prod_i^ny_i} \neq \sum_i^n f_i
-\end{equation}
-
-where $n$ is the total number of steps. This is due to the approximation introduced through the series approximation in  \cref{eqn:series}.
-
-\subsubsection{Example of Input Data Considered: Sapphire Wafers}
-
-Sapphire wafers form the substrate on which all other layers of the light-emitting diode are grown.
-Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture.
-In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process.
-Wafers can be either unpatterned or patterned, where the latter has become commonplace by 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
-The price of sapphire substrates has decreased significantly since the year 2000, as shown in the bottom panel of \cref{fig:wafers}.
-This can be attributed not only to the lighting industry, but more importantly to increased supply as a result of increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
-Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing.
-The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}.
-The evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.
-
-\begin{figure}[h!]
-    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
-    \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
-	\caption{Top: Historical data for sapphire substrate prices of different surface properties and diameters.
-Shown is data for polished surface (NPS) and patterned substrates (PSS).
-\textit{Monocrystal} denotes the Russian manufacturer of the same name.
-Dashed lines are projections from the previous year.
-Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}.
-Bottom: Prevalence of sapphire wafer size used in the manufacturing of light-emitting diodes.
-Hatched bars are projections given by the sources.
-Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
-	\label{fig:wafers}
-\end{figure}
-
-\newpage
-\subsection{Computation of Contribution of Individual Variables}
-\label{sec:contribution_variables}
-
-To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
-Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
-\begin{equation}
-F=ab+cd
-\end{equation}
-and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
-\begin{equation}
-\Delta F = F(t_1)-C(t_0) = \sum_{i=a, \dots, d} \Delta F_i
-\end{equation}
-The infinitesimal contribution to the total funcation value by the infinitesimal change in an input variable is defined through the total differential of the function as
-\begin{equation}
-\text{d}F(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial F }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial F }{\partial x_i}  \Delta x_i
-\end{equation}
-where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
-\begin{equation}
-\Delta F_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial F }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
-\label{eqn:integral_1}
-\end{equation}
-However, data on the input variables is not available in continuous time.
-Disaggregating the contribution of single variables to the change in the cost or performance is thus not straightforward in our model.
-This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}.
-In our work, we propose to address this problem by following an approach developed by Kavlak et al.
-\cite{kavlak2018evaluating}.
-For the detailed derivation, we refer to this publication.
-
-The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
-\begin{equation}
-    F(\vec{r}) = F(r_1,r_2, \dots) = \sum_i F_i
-\end{equation}
-where
-\begin{equation}
-    F_i(\vec{r}) = F_i^0 \prod_w g_{iw}(r_w)
-\end{equation}
-Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
-\begin{equation}
-    \Delta F_x = \int_{t=t_0}^{t_1} F(t) \frac{ \partial \ln F }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
-\end{equation}
-where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$.
-In practice, this constant value can be approximated through the geometric and arithmetic means $\tilde{F} \approx \frac{2}{3} F_i^\text{geo} + \frac{1}{3} \overline{F_i}$.
-The contribution of a single cost model variable $r_z$ can then be written as
-\begin{equation}
-    \Delta F_z (t_1,t_2) \approx \sum_i \tilde{F_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
-\end{equation}
-We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article.
-Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.
-
 \end{document}

### 2023-08-12 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 \end{align*}
+
+\cref{fig:dpw} shows the results of DPW calculations using \cref{eqn:dpw} for different wafer diameters $d$ that were most commonly used in LED manufacturing in $d(2002)=50$mm, $d(2010)=100$mm and $d(2020)=200$mm.
+We can see that DPW dramatically increases over time.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
+    \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2010)=100$mm$\rightarrow5370$DPW, $d(2020)=200$mm$ \rightarrow26,838$DPW.
+Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews.
+For detailed statistics on changes in wafer size, compare also \cref{fig:wafers}.}
+    \label{fig:dpw}
+\end{figure}
+
+\cref{eqn:cost_wafer_full} then gives us for the cost of a manufacturing step $C_i$ in the wafer processing category
+%
+\begin{equation}
+\label{eqn:cost_wafer_full}
+\begin{split}
+    C_i \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] &= \bigg (  \frac{\pi}{4}  \bigg ( \frac{d-2e}{\sqrt{a}+s/2} \bigg ) ^2 - \frac{\pi}{\sqrt{2}}\frac{d-2e}{(\sqrt{a}+s/2)^2} \bigg )^{-1} \times \\
+    &  \frac{1}{y_i}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i \bigg( \frac{t_i}{w_i u_i} \bigg) + \sum_{x} v_x c_x \bigg\}
+\end{split}
+\end{equation}
+%
+and the cost of a manufacturing step $C_j$ in the packaging category
+%
+\begin{equation}
+\label{eqn:cost_die}
+    C_j \bigg[ \frac{ \text{USD}(2020) }{ \text{die} } \bigg] = \frac{1}{y_j}  \bigg\{ \bigg((c_e*p) + c_l + c_m + c_d + c_o \bigg)_i  \frac{c_j}{u_j} + \sum_{x} a v_x c_x \bigg\}
+\end{equation}
+where $c_j$ is $\text{throughput}^{-1}$. The total cost is thus
+%
+\begin{equation}
+\label{eqn:cost_total}
+\begin{split}
+    C= P_s &+ \sum_i \bigg \{ \frac{1}{DPW} \frac{1}{y_i} \bigg[ \frac{t_i}{w_i u_i} \bigg((e*p) + l + m + d +o \bigg)_i  + \sum_{x} v_x p_x \bigg] \bigg \} + \\
+    & + \sum_j \bigg \{ \frac{1}{y_j} \bigg[ \frac{c_j}{u_j}  \bigg((e*p) + l + m + d + o \bigg)_i + \sum_{x} a v_x p_x \bigg ] \bigg\}
+\end{split}
+\end{equation}
+
+\subsubsection{Computation of Yielded Cost}
+
+Devices may be damaged or otherwise rendered unusable during the manufacturing process.
+The ratio between the number of good devices per step and the number of handled devices per step is known as the yield.
+Optimizing this yield is critical for reducing manufacturing cost \cite{Kumar2006}.
+This is because cumulative yield quickly drops as the yield from manufacturing steps with below 100\% yield compounds.
+We must thus consider not only the manufacturing cost per process step, but also the cost including the yield \cite{becker2001use}\cite{becker2001using}.
+While there are different mathematical approaches to including yield, we follow the definition in \cite{becker2001use}.
+We write for the yielded cost $C_{Y_i}$ of a step $i$ with associated cost (before considering yield) $C_1$ and yield $Y_1$, starting from steps $1,2,\dots,i$:
+%
+\begin{equation}
+\label{eqn:C_2}
+    C_{Y_1} = \frac{C_1}{Y_1}
+\end{equation}
+\begin{equation}
+    C_{Y_2} = \frac{C_1 + C_2}{Y_1 Y_2} - C_{Y_1} = \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} = \frac{1}{Y_1 Y_2} \bigg ( C_1 (1-Y_2) +C_2 \bigg)
+\end{equation}
+\begin{equation}
+    C_{Y_i} = \frac{ \sum_{x \leq i} C_x }{ \prod_{x \leq i} Y_x } - \frac{ \sum_{x<i} C_x }{ \prod_{x<i} Y_x }
+\end{equation}
+%
+If a step is applied more than once, we can conveniently rewrite this in a form suited to computation within the \textit{Excel} worksheet.
+Assuming step $2$ is used twice, we get for the yielded cost of this step an equation of the form
+%
+\begin{equation}
+\label{eqn:C_2^2}
+    C_{Y_2}^{(2 \times)} = \bigg( \frac{C_1 + C_2}{Y_1 Y_2} - \frac{C_1}{Y_1} \bigg) + \bigg( \frac{C_1 + 2 C_2}{Y_1 Y_2^2} - \frac{C_1 + C_2}{Y_1 Y_2}     \bigg)
+\end{equation}
+%
+\begin{equation}
+    = \frac{1}{Y_1 Y_2^2} \bigg( C_1 (1-Y_2^2) +2C_2 \bigg)
+\end{equation}
+%
+which can be compared to  \cref{eqn:C_2} to find a more general form:
+%
+\begin{equation}
+    C_{Y_2}^{(n \times)} = \frac{1}{Y_1 Y_2^n} \bigg( C_1 (1-Y_2^n)+nC_2\bigg)
+\end{equation}
+%
+It can be shown by induction that the general form of a term for a step $i>1$ repeated $n$ times can be expressed as:
+%
+\begin{equation}
+\label{eqn:yielded_cost}
+    C_{Y_{i>1}}^{(n \times)} = \frac{1}{Y_i^{n-1} \prod_{x \leq i} Y_x} \bigg( nC_i + \sum_{x < i} C_x (1-Y_i^n) \bigg)
+\end{equation}
+%
+This cumulative approach to yielded cost is different from the approach taken in the original \textit{LEDCOM} model \cite{ledcomv2}.
+It uses what in the literature is described as an \textit{"itemized approach"} to yielded cost \cite{becker2001use}.
+In this approach, the yielded cost of a single process step $f$ is described as
+%
+\begin{equation}
+    f_\text{single} = \frac{k+s}{y}
+\end{equation}
+%
+where
+%
+\begin{align*}
+    k &\dots \text{material cost of previous step} \\
+    s &\dots \text{step cost}
+\end{align*}
+%
+For process steps that are performed more than once, a series expression is used
+%
+\begin{equation}
+    f_{\times 2} =  \frac{\frac{k+s}{y}+s}{y}
+\end{equation}
+\begin{equation}
+\label{eqn:series}
+    f_{\times n} = \frac{k + s(1+y+y^2+ \dots + y^{n-1})}{y^n}
+\end{equation}
+%
+This itemized approach serves as a convenient approximation, but its cumulative contributions do not equal the total yielded cost of the entire manufacturing process
+%
+\begin{equation}
+    f_\text{total} = \frac{\sum_i^n s_k}{\prod_i^n y_i} \neq \sum_i^n f_i
+\end{equation}
+%
+where $n$ is the total number of steps. This is due to the approximation introduced in \cref{eqn:series}.
+
+\subsubsection{Example of Input Data Considered: Sapphire Wafers}
+
+Sapphire wafers form the substrate on which all other layers of the light-emitting diode are grown.
+Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture.
+In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process.
+Wafers can be either unpatterned or patterned, where the latter has become commonplace by 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}.
+The price of sapphire substrates has decreased significantly since the year 2000, as shown in the top panel of \cref{fig:wafers}.
+This can be attributed not only to their role in the solid state lighting industry, but more importantly to increased supply in response to increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}.
+
+Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing (see top panel of \cref{fig:wafers}).
+The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}.
+We use the market average for the sapphire wafer price in different years as a direct input into the cost model, and the dominant wafer diameter (see bottom panel in \cref{fig:wafers}) in different years as an input to calculate DPW in \cref{eqn:dpw}.
+
+\begin{figure}[h!]
+    \includegraphics[width=15cm]{./figures/sapphire_prices.pdf}
+    \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
+	\caption{Historical data for sapphire substrate prices of different surface properties and diameters.
+Shown is data for polished surface (NPS) and patterned substrates (PSS).
+\textit{Monocrystal} denotes the Russian manufacturer of the same name.
+Dashed lines are projections from the previous year.
+Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}.
+Bottom: Prevalence of sapphire wafer size used in the manufacturing of light-emitting diodes.
+Hatched bars are projections.
+Sources: \cite{veeco2013}\cite{Scholand2012}\cite{yole2015sapphire}}
+	\label{fig:wafers}
+\end{figure}
+
+\subsubsection{Computation of Contribution of Individual Variables}
+\label{sec:contribution_variables}
+
+To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
+Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
+%
+\begin{equation}
+F=ab+cd
+\end{equation}
+%
+and input variables $a,b,c,d$, we are looking for the contribution $\Delta F_{a}$ made by a single variable $a$ to the total change in the function value $\Delta F$ between points $t_0,t_1$ such, that
+%
+\begin{equation}
+\Delta F = F(t_1)-F(t_0) = \sum_{i=a, \dots, d} \Delta F_i
+\end{equation}
+%
+Index decomposition analysis has been developed to solve this problem \cite{Boyd1987}. A short derivation, based on \cite{kavlak2018evaluating} is provided here:
+
+The infinitesimal contribution to the total function value by the infinitesimal change in an input variable is defined through the total differential of the function as
+%
+\begin{equation}
+\text{d}F(x_1 (t), x_2(t), \dots) = \sum_i \frac{\partial F }{\partial x_i}     \frac{\text{d}x_i}{\text{d}t} = \sum_i \frac{\partial F }{\partial x_i}  \Delta x_i
+\end{equation}
+%
+where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
+%
+\begin{equation}
+\Delta F_{x_1} = \int_{t=t_1}^{t_2} \frac{\partial F }{\partial x_1} \frac{\text{d}x_1}{\text{d}t} \text{d}t
+\label{eqn:integral_1}
+\end{equation}
+%
+However, data on the input variables in our study is often not available in continuous time.
+Disaggregating the contribution of single variables to the change in the cost or performance is thus not straightforward in our model.
+This problem does not arise in cost models which compute cost changes directly, such as \cite{nemet2012solar} \cite{goodrich2013assessing}.
+In our work, we propose to address this problem by following an approach developed by Kavlak et al.
+\cite{kavlak2018evaluating}.
+For the detailed derivation, we refer to this publication.
+
+The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
+%
+\begin{equation}
+    F(\vec{r}) = F(r_1,r_2, \dots) = \sum_i F_i
+\end{equation}
+where
+\begin{equation}
+    F_i(\vec{r}) = F_i^0 \prod_w g_{iw}(r_w)
+\end{equation}
+%
+which can be written as a constant part $F$ and a part describing the dependence of the $i$-th cost component on the $w$-th variable.
+%
+Using logarithmic differentiation, the integral from  \cref{eqn:integral_1} can be rewritten as
+\begin{equation}
+    \Delta F_x = \int_{t=t_0}^{t_1} F(t) \frac{ \partial \ln F }{ \partial x } \frac{ \text{d} x }{ \text{d} t} \text{d} t
+\end{equation}
+%
+where for $F(t)$ a constant $F(t) \approx \tilde{F} $ can be chosen such that $\Delta F_{x_i} = \Delta F$.
+In practice, this constant value can be approximated through the geometric and arithmetic means $\tilde{F} \approx \frac{2}{3} F_i^\text{geo} + \frac{1}{3} \overline{F_i}$ \cite{kavlak2018evaluating}.
+The contribution of a single cost model variable $r_z$ can then be written as
+%
+\begin{equation}
+    \Delta F_z (t_1,t_2) \approx \sum_i \tilde{F_i} \ln \frac{g_{iz}(t_2)}{g_{iz}(t_1)}
+\end{equation}
+%
+We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article.
+Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.
 

### 2023-08-12 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 
-The historical progress in the different sub-efficiencies contributing to overall efficiency (lamp efficiency) is shown in \cref{fgr:subeff}. Context for figure panels is provided below:
+Empirical data on the historical progress in metrics related to four different sub-efficiencies contributing to the overall LED efficiency (lamp efficiency $\eta_L$) is shown in \cref{fgr:subeff}.
+Specifically, the figure includes data for forward voltage used to calculate to forward voltage efficiency $\eta_{V_f}$ (Panel A1); efficiency droop $\eta_droop$ (Panel A2); internal quantum efficiency $\eta_IQ$ (Panel B1); and light extraction efficiency $\eta_LE$ (Panel B2).
+Additional information and context for individual panels in Supplementary \cref{fgr:subeff} are provided below.
 
@@
 - textured.
+
+Notably, \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $η_{C_R}$ and $\eta_{C_{Y/G}}$.
+Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the empirical \cref{fgr:subeff} not relevant.
+Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum}.
 

### 2023-08-12 2_SSL_NE/article/article_main.tex
added edits

@@
 
+\section{Methods}
+
+\subsection{Multi-Method Approach}
+
+\section{Acknowledgements}
+
 This research was supported by the grant from the Alfred P.
-Sloan Foundation titled \textit{“What factors drive innovation in energy technologies?
-The role of technology spillovers and government investment”}.
+Sloan Foundation titled “What factors drive innovation in energy technologies?
+The role of technology spillovers and government investment”.
 Michael Weinold gratefully acknowledges support from the Swiss Study Foundation.
-The authors thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and C-EENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
-The authors further express their gratitude to all interviewees for their willingness to participate in this study and share their insights.
-
-Additional context and governing equations of metrics used to quantify progress in solid-state lighting, including the constituent sub-efficiencies of overall lamp efficiency.
-Detailed description of the multi-method approach used, including the systematic literature review, semi-structured interviews, the manufacturing cost model and performance calculations.
-Methodological details on dis-aggregation of the contribution of variables to overall lamp efficiency.
-Additional results, including a breakdown of manufacturing cost and a detailed description of the technology spillovers in phosphor materials for light down-conversion.
-Complete list of sources for Figures \ref{fgr:history_efficacy} and \ref{fgr:chip_architecture_overview}.
+The authors thank Venkatesh Narayanamurti, Gabriel Chan, Anna Goldstein, Didier Sornette, and participants of the SPIE West 2021 Conference and CEENRG Seminar Series of the University of Cambridge for many helpful discussions and feedback.
+The authors further express their deepest gratitude to all interviewees for their willingness to participate in this study and share their insights.
 

### 2023-08-12 2_SSL_NE/article/article_main.tex
added edits

@@
 
+The evolution of LED device architecture and performance, as well as the progress in understanding the underlying physical phenomena, are relatively well covered in scholarly literature and patents (see Supplementary Note 1 for a brief literature review).
+However, information provided in such sources is insufficient for our goals on at least three accounts: First, existing work focuses only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
+Scientific publications also do not always disclose the underlying device architecture or the features responsible for the gains in performance.
+Second, not all relevant innovations are patented \cite{Pakes_1980,Fontana_2013}.
+In the case of LED patents in particular, our interviews with industry experts suggest that the propensity to patent is the highest for knowledge related to macroscopic device architecture and chemical composition of phosphors, and the lowest for knowledge related to manufacturing process improvements and microscopic chip architecture that is difficult to reconstruct by reverse engineering.
+This means that relying only on patent literature would bias results by unduly emphasizing some focus areas and de-emphasizing others.
+Third, scientific publications and patents typically focus on experimental devices, rather than commercial products.
+While new LED features, designs and manufacturing methods reported in these sources can potentially result in significant performance gains or cost reductions, it is difficult to ascertain if these improvements have since been adopted in industry.
+Furthermore, information on LED manufacturing cost and the effect of process improvements on the total cost is highly proprietary.
+Estimates are occasionally reported in the scientific literature and company publications, but these often do not disclose which parts of the manufacturing process are responsible for the largest contribution to the overall cost, or which improvements led to cost reductions.
+
+To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a systematic review of the primary scientific literature, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies.
+We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in innovation and technology spillovers.
+We briefly describe our approach to each method below and provide further methodological details in Supplementary Note 3.
+
+\subsection{Systematic Literature Review}
+
+We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
+This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics; 2) identifying individual innovations that contributed to this evolution, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of science or technology outside of SSL, making them involved in technology spillovers.
+
+Relevant sources for the review were found in an iterative search process that involved two components.
+The first was the search in specialized patent and publication databases as well as company websites.
+The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in Supplementary Note 1 and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+We also relied on backward citations in these sources for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
+
+\subsection{Semi-Structured Interviews}
+
+To supplement our data collection efforts, verify our findings and identify additional spillovers, we conducted a series of elite semi-structured interviews \cite{tansey2009process} with thirteen eminent experts from academia, industry, and public research sector.
+Experts were initially selected based on their engagement in different sub-fields of LED research and manufacturing, based on the literature review, then the list was expanded by a ‘snowballing’ tactic based on recommendations from already-interviewed experts.
+All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
+An anonymized summary of the background of interviewed experts is provided in Supplementary Table 1.
+Notably, all our interviewed experts came from Europe or USA, with none representing Asia, which potentially may have biased our findings particularly for the earliest and latest periods of white LED history dominated by LED manufacturers in Japan and China, correspondingly.
+However, this was not an intentional bias, as none of the identified experts in Asia responded to our interview requests.
+
+In terms of the interview content, the primary, structured part of the interviews explored which innovations were deemed most relevant to the evolution of device performance, consumer experience and manufacturing cost of LED packages.
+Thereafter, interviewees were asked to consider the extent to which those innovations may have originated outside of their respective field of expertise and the LED industry more broadly—i.e., which of the innovations potentially involved technology spillovers.
+The remainder of the interview was focused on learning about particular aspects of the manufacturing processes relevant to cost and performance modelling, the current state of industry, and the circumstances surrounding the innovations and spillovers identified in the first part of the interview.
+Specific quantitative data was also provided by experts, helping fine-tune the parameters of the manufacturing cost model and verify device performance data.
+
+\subsection{Performance Metrics Calculations}
+
+The contribution of individual technology innovations and spillovers to the progress in overall device efficiency over time is estimated by index decomposition analysis.
+Mathematically, this involves breaking down a chosen performance indicator into its constituent components, each representing a specific factor that contributes to the change in the indicator \cite{Ang1997}.
+Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
+It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
+
+According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
+%
+\begin{align}
+    \Delta a &= \frac{a_{t=T} - a_{t=0}}{\ln(a_{t=T}) - \ln(a_{t=0})} \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big ) \\
+    & \stackrel{a_{t=0} \neq a_{t=T}}{=} L(F_{t=T}, F_{t=0}) \times \ln \big ( \frac{a_{t=T}}{a_{t=0}} \big )
+\end{align}
+%
+where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
+These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:
+%
+\begin{equation}
+    \Delta a + \Delta b  = \Delta F
+\end{equation}
+%
+To document historical improvements in LED device performance accurately, we need data on all sub-efficiencies for the selected device architectures and periods covered.
+However, the scope of data reporting in the literature is typically limited to selected metrics of interest, rather than the full ensemble of sub-efficiencies that determine the overall device performance.
+Where gaps in relevant data existed, we filled them with our own performance calculations for individual sub-efficiencies where possible.
+Specifically, sub-efficiencies related to the emission spectrum of phosphor down-converted LED devices (i.e., spectral efficiency and light conversion efficiency)  were computed from the spectral data, often reported in LED device specifications, using the \texttt{colour-science} package for Python \cite{colour-science_software}.
+We also used the same approach to calculate CRI and luminous efficacy of radiation of LED devices (see Supplementary Note 4 and Supplementary Figure 10 for the associated spectral data and calculation results).
+This approach allowed us to quantify the improvements related to consumer experience and phosphor development in LEDs.
+
+\subsection{Manufacturing Cost Model}
+
+The structure of our bottom-up white LED manufacturing cost model with process-step resolution is generally based on the 2012 \textit{LEDCOM} cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
+The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the Internet Archive , and expert interviews.
+Because the architecture of LED chips has changed significantly since the introduction of the first commercial white LED devices in 1996, three different chip architectures are considered in the model: classical chips, flip chips, and chip-scale package flip chips (see Figure 3 and Supplementary Figures 2-6 for the details of each architecture).
+The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks  and relevant patents.
+In addition, two LED life cycle analyses (Scholand and Dillon 2012a)(Casamayor, Su, and Ren 2018) were used to validate the model structure and extract some of the necessary quantitative model inputs.
+These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
+
+The cost model we developed is based on a cumulative approach to yielded cost\cite{becker2001use}.
+In this approach, the yielded cost $C_{Y_i}$ of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$.
+Thus, for each consecutive step, starting from $i=1$, we get
+%
+\begin{equation}
+    C_{Y_1} = \frac{C_1}{Y_1}, \ C_{Y_2} = C_{Y_{2 \rightarrow 3}} - C_{Y_1} = \frac{C_1(1-Y_2)+C_2}{Y_1Y_2}, \ C_{Y_3}=\dots
+\end{equation}
+%
+Notably, yielded cost per step is dependent on the step order and blind to downstream information \cite{becker2001use}.
+The yielded cost metric is also cumulative by definition, thus the total cumulative yielded cost is calculated as:
+%
+\begin{equation}
+    \sum_i C_{Y_i} = \frac{\sum_i C_i}{\prod_i Y_i}
+\end{equation}
+%
+The overall outcome of the cost model is thus the cumulative yielded  manufacturing cost per LED package for each of the three years considered.
+In our model, this metric includes all costs associated with producing the chip, including running costs of the factory.
+Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
+We also note that the purpose of our cost model is not to give specific estimates of the LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
+It instead assumes a hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
+It also assumes the use of the most up-to-date equipment for that year.
+Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
+
+Another important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we are able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
+Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
+
+Further details on our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, detailed calculations for the yielded costs, as well as the model’s limitations, are provided in Supplementary Note 3.
+
 \section{Acknowledgements}

### 2023-08-12 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 
-Notably, \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $η_{C_R}$ and $\eta_{C_{Y/G}}$.
+Notably, \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.
 Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the empirical \cref{fgr:subeff} not relevant.

### 2023-08-12 2_SSL_NE/article/article_main.tex
added edits

@@
  \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
- \caption{Historical development of the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.
-Data points indicate best performers by year of market introduction.
-Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Information, Section 1 for details).
-Dashed lines indicate an average improvement for each technology, computed from a 3rd-order polynomial fit to the data.
-The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_max^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
+ \caption{\textbf{Historical progress in the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.} Data points indicate best performers by year of market introduction.
+Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details).
+Dashed lines indicate a 3rd-order polynomial fit to the data trend.
+The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
 \cite{Murphy2012}.
@@
 \cite{azevedo2009transition}.
-See Supplementary Information, Section 4 for the full list of sources and references.}
+See Supplementary Note 7 for the full list of sources and references.}
  \label{fgr:history_efficacy}
@@
   \includegraphics[height=6.5cm]{figures/cost_lamp_small.pdf}
-  \caption{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, including light bulbs, spotlights and recessed lights from 2008 to 2020.
+  \caption{\textbf{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights.
 Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
 Department of Energy (DOE) \cite{council2013assessment}.
-Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs.
-Source: own synthesis of data on LED sales prices collected from various consumer watchdog databases and publications.
-Data on CFL sales prices collected from Eger \& Ehlhardt \cite{eger2018origin} as well as various consumer watchdog databases and publications.
-Incandescent light bulb price is assumed constant based on the average in the covered time period.
-See Supplementary Information, Section 4, for the full list of sources and references.}
+Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period.
+Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications.
+See Supplementary Note 7 for the full list of sources and references.}
   \label{fgr:cost_lamp_small}
@@
 
-These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long,Stegmaier2021}.
+These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long,stegmaier2021incandescent}.
 As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
@@
 
+\section{Main}
+
+\subsection{Evolution of LED Chip Architectures}
+
 \clearpage
@@
 \includegraphics[height=15cm]{figures/chip_architecture_overview.pdf}
- \caption{Historical evolution of light-emitting diode chip architectures.
-Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
+\caption{\textbf{Historical evolution of LED chip architectures.} Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
 Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
@@
 Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Information, Section 4, for the full list of sources and references.}
+See Supplementary Note 7 for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
@@
 
-\section{Results}
+We summarize the information collected and analysed on the progress in GaN-based LED chip architectures and manufacturing processes in \cref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+We use this information as input into LED performance calculations and manufacturing cost modeling.
 
-\subsection{Improvements in Sub-Efficiencies and Overall Efficiency}
+\subsection{Efficiency Improvements}
 
@@
 
-\subsection{Improvements in Consumer Experience Metrics}
+\subsection{Consumer Experience Improvements}
 
@@
 
-\subsection{Improvements in Manufacturing Cost}
+\subsection{Manufacturing Cost Improvements}
 

### 2023-08-12 2_SSL_NE/article/article_main.tex
added edits

@@
 
+Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 4, we find that only a single innovation related to LED consumer experience improvements was initially intended exclusively for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
+All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:phosphors}).
+This means that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
+
 Figure \ref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
@@
 
-Historical improvements in consumer experience metrics for phosphor-converted warm white LEDs are shown in Figure \ref{fgr:consumer_experience}.
+Historical improvements in consumer experience metrics for phosphor-converted warm white GaN-based LEDs are shown in \cref{fgr:consumer_experience}.
 In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable.
-Both metrics are determined by LED device spectra, which, in turn, depend on the properties of the down-conversion materials in the device.
-We were thus able to establish the links between all major improvements in the two consumer experience metrics identified in this study and individual LED inventions associated either with phosphors or quantum dots.
-The list of these inventions is provided in Table \ref{tab:phosphors}, while the detailed descriptions of these inventions and corresponding device spectra are provided in Section 3 in the Supplementary Information.
+Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into the white light required for general illumination .
+It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs.
+The first commercial white LED produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{bando1998development}.
+As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:innovations}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.
 
@@
     \centering
+    \caption{\textbf{LED innovations related to improvements in consumer experience metrics.}}
     \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|l|}
     \hline
-        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{SI Sec.} \\ \hline
+        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{Ref.} \\ \hline
         1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.2.2.1 \\ \hline
@@
     \end{NiceTabularX}
-    \caption{LED down-conversion materials (phosphors and quantum dots) related to improvements in consumer experience metrics identified in this study.
-The \textit{Year} column represents the earliest reported application of invention in white LEDs.
-These differ from the years used in Figure \ref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for representative LED products that relied on those materials.
-The \textit{SI Sec.} column refers to the corresponding section in the Supplementary Information that contains references and a detailed description of the history of corresponding invention.
-Abbreviations: Desig.
-- Designation, Sec.
-- Section.}
+    \caption{\textit{Year} column represents the earliest reported application of innovation in white LEDs.
+These differ from the years shown in \ref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for representative LED products that relied on those innovations.
+Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4.
+Desig.
+= Designation, Ref.
+- Reference.}
     \label{tab:phosphors}
@@
  \includegraphics[width=\textwidth]{figures/breakthroughs_consumer-experience.pdf}
- \caption{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.
-Data points show the color temperature and colour rendering index performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year.
-Corresponding spectral data used to calculate colour temperature values is provided in Figure SI 11 in Section 3 of the Supplementary Information.
+ \caption{\textbf{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.} Data points show the color temperature (CT) and colour rendering index (CRI) performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year.
+Corresponding spectral data used to calculate CT values is provided in Supplementary Note 4.
 Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison.
-Horizontal lines represent typical colour temperatures of “traditional” light sources, shown for reference.
+Horizontal lines represent typical CTs of “traditional” light sources, shown for reference.
 The desirable range of colour temperatures for home illumination, indicated by a vertical red arrow, lies between two horizontal orange lines representing typical incandescent light and warm daylight colour temperatures.
@@
 
-Figure \ref{fgr:costmodel} shows key results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages.
+Figure \ref{fgr:costmodel} shows main results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages.
 We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
@@
 In the case of manufacturing yield, the higher it is, the less inputs are wasted on the production of a single LED package.
-With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare Figure \ref{fgr:costmodel}, Panels A-C), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period.
+With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare Figure \ref{fgr:costmodel}, Panel D), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period.
+
 In the case of wafer diameter, the larger the wafer, the more LED chips can be produced from a single wafer.
-The wafer diameter commonly used in LED manufacturing has been steadily increasing 2003 from 51mm (2 inch) in 2003 to 200mm ($\sim$8 inches, referred to as “8 inch") in 2020.
-We capture this in the model by assuming the following wafer diameters and calculating the associated number of die per wafer (DPW) \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$.
+The wafer diameter commonly used in LED manufacturing has been steadily increasing from 50mm ($\sim2$ inch\footnote{Industry measures wafer size in mm.
+Still, a 50mm=1.9685Inch wafer is frequently referred to as a "2-inch wafer".}) in 2003 to 200mm ($\sim$8 inches, referred to as “8 inch") in 2020.
+We capture this effect in the model by calculating the associated number of die per wafer (DPW) for each representative wafer diameter d \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$ (see Supplementary Figure 7 in Supplementary Note 3).
 With more than a thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time.
@@
 Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
-We also provide a comparison of our model with past cost calculations and projections published by the US DOE on the basis of the LEDCOM cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables in Section 2 of the Supplementary Information.
+
+In Supplementary Note 6, we also provide a comparison of the outcomes of our model with past cost calculations and projections published by the U.S.
+Department of Energy (DOE) on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables.
 
@@
 
-\clearpage
+\subsection{Technology Spillovers}
+
+Following the frameworks for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021} and Kolesnikov et al.
+\cite{kolesnikov2022technology}, we synthesize information from the interviews, historical records, literature review and citations in patents and publications to identify which LED innovations involved knowledge originating in areas of science and technology beyond LED and SSL.
+We find that at least nine LED innovations, listed in Table 2, clearly involved such ‘external’ knowledge.
+We then analyze the sources, mechanisms and enablers of corresponding technology spillovers, as well as their contributions to progress in white LED technology.
+
+We find that three spillovers associated with the use of YAG/YGAG phosphors in LEDs played the key role in the development of the first commercial white LED lighting products, essentially enabling the solid-state lighting market and industry of today.
+As noted above, subsequent spillovers then had a significant effect on physical device performance, cumulatively contributing to 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020, as well as nearly all key improvements in consumer experience metrics.
+
+Among the spillover sources, we find that all nine spillovers had origins in the scientific disciplines such as various branches of chemistry, materials science, optics and photonics, and solid-state physics.
+Five spillovers also utilized technical knowledge and expertise in cathode ray tubes, fluorescent lighting, optoelectronic devices, nanotechnology, and nature-inspired material design.
+Among the spillover mechanisms, six spillovers (involved in all phosphors except PFS and SLA, plus ITO) were a result of application of external scientific and technical knowledge already available to researchers and inventors.
+Three remaining spillovers (involved in PFS and SLA phosphors, plus PSS) occurred as an outcome of targeted search for relevant external knowledge outside the LED domain.
+In addition, at least two spillovers (involved in 258 and PFS phosphors) occurred as a result of knowledge exchange in direct R\&D collaboration.
+
+Among important enabling factors for the identified spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
+Finally, we find that, on average, it took 26 years from the initial scientific discovery or invention to the moment of its spillover into the LED domain, with this time varying from 5 to almost 70 years.
+In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED market.
+
 \begin{table}[h!]
@@
     \centering
-    \caption{Technology spillovers involved in white LED technology innovations identified in this study.}
+    \caption{\textbf{Technology spillovers involved in white LED technology innovations identified in this study.}}
     \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|X|l|X|}
@@
     \end{NiceTabularX}
-    \begin{tabular}{@{}p{\linewidth}@{}}
-        Note: Disc.
-- Year of initial discovery; S/O - Year of spillover to LED; Comm.
-- Year of commercial application; Ref.
+    \caption{Note: Disc.
+- Year of initial discovery, identified  by the earliest literature source describing the discovery of original idea or invention outside the LED domain.
+S/O - Year of spillover to LED; Comm.
+- Year of commercial application, identified as the year of the first recorded application of that idea or invention in a commercial LED product.
+Ref.
 - References.
 LED innovations are ordered by the year in which a technology spillover into LED occurred, provided in the S/O column.
-The year of initial discovery is the year of the earliest identified discovery of the original idea or invention outside the LED domain.
-The year of commercial application is the year of the first recorded application of that idea or invention in a commercial LED product.
 Origin column represents knowledge domains in which spillovers initially emerged, where (S) denotes a scientific discipline and (T) is an area of technology.
 Ref.
-column lists literature sources for the represented innovations and spillovers.
-Area of Improvement column represents the impact of spillovers on different aspects of white LED technology, e.g., improvements in particular sub-efficiencies.
-    \end{tabular}
+column lists literature sources for the represented discoveries, innovations and spillovers.
+Area of Improvement column represents the impact of spillovers on different aspects of white LED technology, e.g., improvements in particular sub-efficiencies.}
     \label{tab:spillovers}
@@
 
-\subsection{Performance Improvements}
+We find that the total LED device efficiency increase from 5.8\% to 38.8\% between 2003 and 2020, as well as all major improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that reduced energy losses across all physical loss channels in LED devices and introduced new phosphor materials for LED light down conversion.
+Our interviews have also revealed an important role of incremental manufacturing process improvements and learning-by-doing \cite{WRIGHT_1936, Arrow_1962} in the progress in LED efficiency.
+However, further research is needed to quantify the contribution of learning-by-doing to this progress.
 
-Our findings show that the overall white LED efficiency has increased from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.8\%$ in 2020.
-This efficiency increase was predominantly the result of a series of innovations in the LED architecture and materials as well as changes in the manufacturing process.
-Our comparison of improvements in device sub-efficiencies between 2003 and 2020 shows that the overall efficiency improvement was not dominated by any single sub-efficiency channel.
-Instead, there has been consistent progress across all device sub-efficiencies during the time period under consideration.
-Some sub-efficiencies, such as forward voltage efficiency, internal quantum efficiency and light extraction efficiency, are now within $\sim10\%$ of their respective physical limits.
-A marked trend is the increasing relative contribution of losses in channels with physical limits on efficiency due to the underlying physics, such as light conversion efficiency of phosphors determined by the Stokes shift.
-Spectral efficiency also remains at $\sim17\%$ from its physical limit, showing potential for further improvements.
-Research on improving LED performance across efficiency loss channels continues\cite{cho2017white,Weisbuch2020}.
-
-We further find that there has been significant progress in performance across the set of consumer experience metrics, which was driven by innovations related to the conversion of blue light generated by conventional GaN LEDs into the white light required for general illumination.
-The first commercial white LEDs produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{bando1998development}.
-After a series of innovations in phosphors shown in Figure \ref{fgr:consumer_experience} and described in Section 3 of the Supplementary Information, LEDs today can be tuned for high colour rendering performance, high colour saturation and a range of desirable colour temperatures.
-
-Overall, we find that LED technology innovations made crucial contributions to the progress in both LED performance and consumer experience metrics across the entire period covered by our study.
-Our interviews have also revealed an important role of incremental process improvements and learning-by-doing\cite{WRIGHT_1936,Arrow_1962} in the progress in LED efficiency.
-However, further research is needed to quantify the contribution of learning by doing to this progress.
-
-\subsection{Cost Reductions}
-
-Our LED manufacturing cost model shows that the cost of manufacturing low-to-mid power GaN-based white LED packages with classical chip architecture at a U.S.
-location using state-of-the-art equipment has decreased by 95.5\% from 1.11\$(2020) in 2003 to 0.05\$(2020) in 2020.
-In contrast with LED performance improvements, where progress was driven mostly by LED technology innovations, the dramatic decline in LED manufacturing cost resulted from an increase in the wafer size used in manufacturing and higher yields across manufacturing steps.
-This can be seen in the reduction of the contribution of wafer processing steps to total cost in \ref{fgr:costmodel}.
-Both were enabled by advances in manufacturing equipment performance and incremental process improvements from learning-by-doing.
+Our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid power GaN-based white classic-chip LEDs from 1.11$ to 0.05$ (in 2020 USD) between 2003 and 2020.
+In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size , with the former factor contributing to XX\% and latter factor contributing to YY\% of the overall LED cost reduction over time .
+Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing.
+Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time.
+[PLACEHOLDER TO DISCUSS THE REASONS FOR THAT].
 
@@
 However, in this study we were able to collect data and compare the outcomes only for the classical chip architecture.
-Collecting the full set of data needed to populate the model for the remaining two architectures would require access to proprietary information from industry.
-With this limitation, tracking manufacturing cost declines across three key LED chip architectures remains a topic for future work.
+Collecting the full set of data needed to populate the model for the remaining two architectures requires access to proprietary information from industry.
+With this limitation, tracking manufacturing cost reductions across all three key LED chip architectures remains a topic for future work.
 
-\subsection{Contribution of Technology Spillovers}
-\label{subsec:spillovers}
+We find that technology spillovers into white LED technology have been significant drivers of improvements in LED efficiency and, particularly, consumer experience metrics.
+Our analysis of the sources, mechanisms and enablers of the identified technology spillovers has important practical implications, highlighting the critical role played by a deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.
+Specifically, deep physical understanding of LED device energy loss channels had enabled important innovations in LEDs that increased several sub-efficiencies in LEDs and will continue to do so, as expected by eminent experts in the field .
+This suggests that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology and its applications both in SSL and technology areas beyond general illumination.
+Corresponding knowledge spillovers can be enabled and deliberately stimulated by various measures such as knowledge exchange events and long-term partnerships between academia and industry, interdisciplinary training and hiring, dedicated mission-driven public R\&D funding, and ensuring certain freedom of search in academia.
+These observations also further reinforce broader arguments made against the dichotomy of basic and applied research \cite{narayanamurti2016cycles, narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
 
-We find that nine technology spillovers identified in our study, listed in Table \ref{tab:spillovers}, affected all dimensions of white LED performance.
-Three spillovers associated with the use of YAG/YGAG phosphors in LEDs played the key role in the first commercial white LED lighting products, essentially enabling the solid-state lighting market and industry of today.
-Spillovers also had a significant effect on physical device performance, cumulatively contributing to 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020.
-Technology spillovers were particularly important for consumer experience performance of white LED lighting sources, with corresponding spillover-driven innovations responsible for $\sim$100\% of the improvements in consumer experience metrics.
-
-Following the framework for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021}, we gathered information from the interviews, historical records, literature reviews and industry publications to analyze the sources, mechanisms, and enablers of the identified spillovers into the white LED technology listed in Table \ref{tab:spillovers}.
-Among the spillover sources, we find that all nine spillovers had origins in basic science disciplines such as various branches of chemistry, materials science, optics and photonics, and solid-state physics.
-Five spillovers also utilized technical knowledge and expertise in cathode ray tubes, fluorescent lighting, optoelectronic devices, nanotechnology, and nature-inspired material design.
-
-Among the spillover mechanisms, six spillovers (involved in all phosphors except PFS and SLA, plus ITO) were a result of application of external scientific and technical knowledge already available to researchers and inventors.
-Three remaining spillovers (involved in PFS and SLA phosphors, plus PSS) occurred as an outcome of targeted search for relevant external knowledge outside the LED domain.
-In addition, at least two spillovers (involved in 258 and PFS phosphors) occurred through direct R\&D collaboration.
-
-Among important enabling factors for the identified spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
-
-We find that, on average, it took 26 years from the initial scientific discovery or invention to the moment of its spillover into the LED domain, with this time varying from 5 to almost 70 years.
-In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED market.
+There are several important avenues of future research that are opened up by our analysis.
+First, future work could expand the cost model by collecting and including data for a broader set of chip architectures.
+Second, a deeper dive into the role of learning-by-doing is needed both in the cost and performance analysis.
+Third, building on the work on the physical limits in LED sub-efficiencies, future efforts could focus on identifying priority areas for further performance improvements in LEDs and SSL in general.
+Fourth, better understanding is needed about the role of demand-side factors, such as policies stimulating market demand for LED lighting (e.g., through incandescent light bulb bans or subsidies for LED adoption), not only in the rapid diffusion of LED-based lighting technologies, which is reasonably well understood by now \cite{Mills2014, Kamat2020, weinold2020technology, stegmaier2021incandescent, grubb2021new}, but also in facilitating learning-by-doing, innovation and technology spillovers that have contributed to the technological progress in white LEDs.
+Finally, the lessons learned about the drivers of innovation, technology spillovers, cost reductions and performance improvements in LEDs may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}.
+By comparing these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
 

### 2023-08-13 2_SSL_NE/article/article_main.tex
added edits

@@
 \begin{abstract}
-  Since their introduction in 1996, white LEDs have greatly improved in performance, efficiency, and manufacturing cost.
-These advancements are crucial for reducing carbon emissions as lighting accounts for nearly 20\% of global electricity consumption.
-Studying the rapid progress in LED technology can provide valuable insights for accelerating innovation in other clean energy technologies.
-However, the extent and sources of these improvements have not been systematically investigated.
-Through cost and performance modeling, along with expert interviews, we found that the efficiency of top-performing warm white LED packages increased from 5.8\% in 2003 to 38.8\% in 2020.
-Additionally, the manufacturing cost of low-to-mid-power LED packages decreased from \$1.1 to \$0.05 (in 2020 USD) between 2003 and 2020, a 95.5\% decrease.
-Technology spillovers from other sectors contributed to 8.5\% of efficiency improvements and nearly 100\% of consumer experience enhancements, playing a crucial role in widespread LED adoption.
+  Since their introduction to the market in 1996, white light-emitting diodes (LEDs) have greatly improved in performance, efficiency, and manufacturing cost.
+These advancements have been crucial for reducing global carbon emissions in the lighting sector which accounts for nearly 20\%  of global electricity consumption.
+Understanding the extent and sources of the rapid progress in white LED technology can thus provide valuable insights for accelerating innovation in other demand-side clean energy technologies.
+Through cost and performance modeling, literature review, patent analysis, and expert interviews, we find that the efficiency of top-performing  warm white GaN-based  LED packages increased from 5.8\% in 2003 to 38.8\% in 2020.
+Over the same period, the manufacturing cost of low-to-mid-power LED packages decreased by 95.5\% from \$1.1 to \$0.05 (in 2020 USD).
+Technology spillovers from other sectors contributed to 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
 \end{abstract}
@@
 A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}.
-According to the United Nations, by the end of 2021, more than 130 countries have set or are considering setting a net zero emissions target by 2050 \cite{un2021climate}.
+According to the Net Zero Tracker Database, by the end of 2023, more than 150 countries have set or are considering setting a net zero emissions target by 2050 \cite{zerotracker}.
 The European Union, for example, has set a target of net zero emissions by 2050.
@@
 
-Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}.
+Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies \cite{iea2020cleanenergy}, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}.
 To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed.

### 2023-08-13 2_SSL_NE/article/article_main.tex
added edits

@@
 
-There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{national2016power,iea2020perspectives}.
-The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}.
-Innovation is also driven by forces of supply and demand.
+There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{WRIGHT_1936, Arrow_1962}, and economies of scope \cite{johansson2012global, national2016power, iea2020perspectives}.
+The role of these factors at different stages of the innovation life cycle, from research and technology development to demonstration, market formation and diffusion \cite{grubler2012policies}, is an area of active research  \cite{kavlak2018evaluating, Ziegler2021}.
+Innovation is also driven by forces of supply and demand \cite{Mowery1979}.
 Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.
 
-The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies,Anadon2016}.
-Among these drivers, the role of external  knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area  \cite{Stephan2021}.
-While the exact definition of spillovers in the literature depends on the context and the research question \cite{Nemet2012}, we follow Stephan et al.
-\cite{Stephan2021} and consider knowledge to be external - thus a spillover -  if it has been developed for application in other technologies, sectors, or scientific disciplines.
-There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016,Stephan2021,Sun2021,kolesnikov2022technology} the future evolution of technologies.
+The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies, Anadon2016}.
+Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}.
+While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2020novel}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2020novel} by considering technology spillovers as the application of external knowledge in a technology where knowledge is considered external  if it has been developed for application in other technologies, sectors, or scientific disciplines.
+There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies.
 
 Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently   accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}.
-It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.
-As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
+It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements.
+As shown in  \cref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
 For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W.
@@
 
-These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long,stegmaier2021incandescent}.
+These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}.
 As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
@@
 
-Despite this impressive history, the sources of LED innovation have not received as much attention from researchers as innovation in supply-side  energy technologies, such as solar photovoltaics  \cite{kavlak2018evaluating} or wind energy \cite{qiu2012price,jennings2020policy}, or in lithium-ion batteries for transportation \cite{Ziegler2021,Stephan2021}.
-To the best of our knowledge, no study has comprehensively discussed the sources or extent of progress across various metrics of LED cost or performance since the introduction of first commercial white LED products.
-Understanding the extent to which individual innovations and knowledge spillovers  contributed to improvements in LED technology, how this effect compares to other sources of improvements such as economies of scale, and how these innovations occurred (i.e., by what mechanisms and actors) will provide valuable lessons both for other demand-side technologies and overall for accelerating clean energy innovation.
+Despite this impressive history, the sources of technological progress in white LEDs have not received as much attention from researchers as innovation  in supply-side energy technologies, such as solar photovoltaics \cite{kavlak2018evaluating, nemet2019solar}, wind energy \cite{qiu2012price, jennings2020policy}, or lithium-ion batteries \cite{Ziegler2021, Stephan2021}.
+To the best of our knowledge, no study has comprehensively discussed the sources or extent of historical progress across a range of metrics related to LED cost and performance since the introduction of the first commercial white LED products (see Supplementary Note 1 for a brief review of previous literature on this topic).
+This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012, Creutzig2018}.
+Understanding the extent to which individual innovations and technology spillovers contributed to improvements in white LED technology, how their effect compares to other sources of improvements such as economies of scale, and how these innovations and spillovers occurred (i.e., by what mechanisms and actors) will provide valuable lessons for innovation in other demand-side energy technologies and clean energy innovation in general.
 
-To address these questions, in this paper we identify a set of metrics suitable for tracking the historical progress of LED lighting technology.
-We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing.
-Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews with eminent experts from academia, public research institutions and industry.
-This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them.
-Data on performance improvements associated with individual innovations identified in patents, scientific publications, industry reports, and interviews is then augmented  by our calculations of device efficiency, including contributions of specific spillovers to the overall LED efficiency (see Supplementary Information for more detail).
-We further calculate LED manufacturing costs using a bottom-up cost model with process-step resolution that we developed.
-Finally, we discuss what our observations may mean for our understanding of the innovation process in innovation process in LED technology and for broader policy and industry efforts to accelerate clean energy innovation.
+To address these questions, first, we identify three sets of metrics suitable for tracking the historical progress of LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, including colour rendering index (CRI) and colour temperature (CT); and 3) LED device manufacturing cost.
+The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Note 2.
+We then trace the historical improvements in these metrics, along with associated changes in LED chip architectures, from the time of introduction of the first commercial warm white GaN-based LEDs in 2003 to 2020, the year with the most recent data available at the time of data collection.
+Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry (see Methods section and Supplementary Note 3 for details).
+This approach allows us to identify individual innovations in white LED technology, examine their knowledge origins to discern technology spillovers among them and quantify their contributions to the overall progress in LED efficiency and consumer experience metrics based on our modeling of LED performance and spectral metrics.
+We further develop a bottom-up LED manufacturing cost model with process-step resolution and use it to analyze the progress in LED manufacturing cost for the “classical” GaN-based LED chip architecture with lateral current spreading shown in Panel A in \cref{fgr:chip_architecture_overview}.
 

### 2023-08-13 2_SSL_NE/article/article_main.tex
added edits

@@
 A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}.
-According to the Net Zero Tracker Database, by the end of 2023, more than 150 countries have set or are considering setting a net zero emissions target by 2050 \cite{zerotracker}.
+According to the Net Zero Tracker Database, by the end of 2023, more than 150 countries have set or are considering setting a net zero emissions target by 2050 \cite{climateaction}.
 The European Union, for example, has set a target of net zero emissions by 2050.
@@
 Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices.
-Figure \ref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.
+\cref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.
 
@@
 Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
-Department of Energy (DOE) \cite{council2013assessment}.
+Department of Energy (DOE) \cite{national2013assessment}.
 Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period.
@@
 
-The historical development of the sub-efficiencies introduced in Section Device Performance Metrics is shown in Figure SI11 in the Supplementary Information.
-Using this data, we calculated the overall LED efficiency for four years: 2003, 2010, 2016 and 2020.
-The waterfall diagrams in Figure \ref{fgr:waterfall} show how improvements in sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to $12.7\%$ in 2010, $32.5\%$ in 2016 and finally to $38.8\%$ in 2020.
+The historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with LED innovations and manufacturing process improvements that have driven this progress, is described in Supplementary Note 4.
+Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020.
+The waterfall diagrams of electric power input losses in \cref{fgr:waterfall} show how improvements in individual sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to 12.7\% in 2010, 32.5\% in 2016 and finally to 38.8\% in 2020.
 As is evident from the figure, no single loss channel dominates in terms of its contribution to the overall efficiency, in line with previous observations \cite{tsao2010solid}.
-We note, however, that the loss channels with a fixed physical limit, e.g., Stokes loss that contributes to the conversion efficiency by phosphors, has become more dominant in 2016 and 2020 compared to 2003 and 2010.
-This is a direct result of the large efficiency improvements of upstream sub-efficiencies.
+We note, however, that the loss channels with a fixed physical limit on efficiency, e.g., Stokes loss that determines the light conversion efficiency by phosphors, became relatively more dominant in 2016 and 2020 compared to 2003 and 2010.
+This is a direct result of large efficiency improvements in upstream sub-efficiencies.
 
-Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 4, we find that only a single innovation related to LED consumer experience improvements was initially intended exclusively for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
-All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:phosphors}).
-This means that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
+\cref{fgr:subeff} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
+The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in \cref{sec:discussion}.
+Through the index decomposition analysis described in \cref{subsec:performance_metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to innovations that involved technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
-Figure \ref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
+\cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
 The full list of identified LED innovations considered in our study is provided in Table SI2 in Section 3 of the Supplementary Information document.
-The list of corresponding technology spillovers is provided in Table \ref{tab:spillovers} in the Discussion section below.
+The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in the Discussion section below.
 Through the index decomposition analysis described in Section 2.5 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
-In Figure \ref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels.
+In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels.
 We find that there has been consistent progress across all device sub-efficiencies in the recorded period.
@@
 Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits.
-The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements.
+The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements, which is important, given that efforts to improve LED performance across different efficiency loss channels through R\&D continue \cite{cho2017white, Weisbuch2020}.
 
@@
  \captionsetup{font=footnotesize}
- \caption{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020 (top to bottom panels), normalized to 1 Watt of electric power input (yellow bar on the left).
-Sub-efficiencies corresponding to each loss channel are listed below each column and described in Section 1.3 of the Supplementary Information.
+ \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020} (top to bottom panels).
+Losses are normalized to 1 Watt of electric power input (yellow bar on the left).
+Sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2.
 Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts).
-Percentages for red, green and blue loss channels indicate losses of remaining red/green/blue light energy.
-The corresponding LED architectures and associated down-conversion phosphors are shown for reference: 2003 - flip-chip with YGAG phosphor; 2010 - flip-chip with 258 phosphor; 2016 - flip-chip with PFS phosphor; 2020 - flip-chip with SALON phosphor.
-A more complete overview of LED chip architectures is provided in Figure \ref{fgr:chip_architecture_overview}.
-Details on phosphor chemical composition are provided in Table \ref{tab:phosphors} with additional data on the spectral performance and invention history provided in Section 3 in the Supplementary Information.
-Overall LED package efficiency is $\eta_L = 5.8\%$ in 2003, $\eta_L = 12.7\%$ in 2010, $\eta_L = 32.6\%$ in 2016 and $\eta_L = 38.8\%$ in 2010.
+Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy.
+The following LED architectures and down-conversion phosphor spectra used in calculations in each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor.
+Details for each architecture are provided in Figure 3.
+Details on the phosphors for light down conversion are provided in Table 1 and Supplementary Note 4.
 Abbreviations: Scat.
-= Scattering.
+= Scattering; Nonrad.
+= Nonradiative.
 Source (Efficiency): own elaboration based on data in Figure SI11.
@@
  \includegraphics[width=\textwidth]{figures/breakthroughs_efficiency.pdf}
- \caption{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs with test currents of at least 350mA.
-Vertical bars represent LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
-Horizontal coloured lines indicate state-of-the-art sub-efficiency levels for the four years used in Figure \ref{fgr:waterfall}: 2003, 2010, 2016, and 2020.
-To provide additional historical context, data for 1997 is included for selected sub-efficiencies.
-"N/A" denotes sub-efficiencies where 1997 data could not be calculated to different reasons: $\eta_{Vf}$, $\eta_{droop}$ depend on the device current, which was below 350mA in 1997, making comparison with contemporary devices difficult.
-$\eta_{C(R)}$,$\eta_{C(Y/G)}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997.
-Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures light extraction efficiency no longer depends on ITO, see Figure \ref{fgr:chip_architecture_overview}.
+ \caption{\textbf{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs.} LEDs with test currents of at least 350mA are considered.
+The following sub-efficiencies are represented: $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $η_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C_R}$ - conversion efficiency for red phosphors; $\eta_{C_{Y/G}}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency.
+The overall LED lamp efficiency $\eta_L$ is displayed in the rightmost column.
+Vertical bars represent cumulative contributions to efficiency improvements from LED technology innovations, with purple bars indicating innovations driven by technology spillovers (annotated and listed in an inset table) and grey bars indicating all other improvements identified in this study.
+Horizontal coloured lines indicate sub-efficiency levels of best performing LEDs for the four years used in \cref{fgr:waterfall}: 2003, 2010, 2016, and 2020.
+For additional historical context, data for 1997 is included whenever possible.
+"N/A" denotes sub-efficiencies where 1997 data could not be calculated for the following reasons: $\eta_{V_f}, \eta_{droop}$ depend on the device current, which was below 350mA in 1997, making a comparison with contemporary devices difficult.
+$\eta_{C_R}, \eta_{C_{Y/G}}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997.
+Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures $\eta_{LE}$ no longer depends on ITO, see \cref{fgr:chip_architecture_overview}.
 Physical limits on sub-efficiencies are indicated by black horizontal lines.
-The overall LED lamp efficiency, $\eta_L$, is displayed in the rightmost column.
-$\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C(R)}$ - conversion efficiency for red phosphors; $\eta_{C(Y/G)}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency; PSS - patterned sapphire substrate, ITO - indium tin oxide current spreading layer; 258, SLA, PFS - different phosphors used in white LEDs, see Table \ref{tab:phosphors}.
-Note that physical limit on $\eta_{V_f}$ is above 100\% due to quantum effects which depends on electrical device parameters \cite{david2016electrical}.
-Source: own elaboration based on data represented in Figure \ref{fgr:waterfall} and Figure SI11 in the Supplementary Information, as detailed in Section Methods and Data Collection.}
+Note that the physical limit of $\eta_{V_f}$ is above 100\% due to quantum effects \cite{david2016electrical}.
+Source: own elaboration based on data represented in \cref{fgr:waterfall}, \cref{tab:spillovers} and Supplementary Note 4.}
  \label{fgr:breakthroughs_efficiency}
@@
 
-Notably, from detailed descriptions of the history of inventions in this list, which we provide in Section 3.2.2 of the Supplementary Information, we find that only a single invention related to LED consumer experience improvements was originally developed specifically for application in solid-state lighting: the 2016 SALON phosphor compound \cite{seibald2019phosphor,Hoerder2019}.
-All other inventions in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL.
-We discuss the details of the corresponding technology spillovers in the Discussion section below.
+Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 4, we find that only a single innovation related to LED consumer experience improvements was initially intended exclusively for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
+All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:phosphors}).
+This means that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
 
@@
     \caption{\textit{Year} column represents the earliest reported application of innovation in white LEDs.
-These differ from the years shown in \ref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for representative LED products that relied on those innovations.
+These differ from the years shown in \cref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for LED products that relied on those innovations.
 Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4.
@@
  \includegraphics[width=\textwidth]{figures/breakthroughs_consumer-experience.pdf}
- \caption{\textbf{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.} Data points show the color temperature (CT) and colour rendering index (CRI) performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in Table \ref{tab:phosphors}, each indicated by the phosphor label and publication year.
+ \caption{\textbf{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.} Data points show the color temperature (CT) and colour rendering index (CRI) performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in \cref{tab:phosphors}, each indicated by the phosphor label and publication year.
 Corresponding spectral data used to calculate CT values is provided in Supplementary Note 4.
@@
 
-Figure \ref{fgr:costmodel} shows main results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages.
+\cref{fgr:costmodel} shows main results of our manufacturing cost modelling for low-to-mid-power classic chip GaN phosphor-converted white LED packages.
 We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
@@
 In the case of manufacturing yield, the higher it is, the less inputs are wasted on the production of a single LED package.
-With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare Figure \ref{fgr:costmodel}, Panel D), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period.
+With the total manufacturing yield dramatically improving from $\sim25\%$ in 2003 to $\sim75\%$ in 2020 (compare \cref{fgr:costmodel}, Panel D), it is not surprising that the total yielded LED manufacturing cost significantly declined over this period.
 
@@
 With more than a thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time.
-As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in Figure \ref{fgr:costmodel}).
+As the number of die per wafer increases, the packaging steps, which in the classic chip architecture must be performed separately for each individual LED chip, carry a significantly larger share of the total cost in 2020 than in 2003 (compare Panels A-C in \cref{fgr:costmodel}).
 However, as our interviewees noted, while growing LEDs on larger wafers is economically desirable, it is associated with engineering and epitaxy challenges as well as high up-front cost of new equipment.
@@
         1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
-        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare Figure \ref{fgr:chip_architecture_overview})\\ \hline
-        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare Figure \ref{fgr:chip_architecture_overview}) \\ \hline
+        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview})\\ \hline
+        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview}) \\ \hline
         1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
@@
 \section{Discussion}
+\label{sec:discussion}
 
@@
 Third, building on the work on the physical limits in LED sub-efficiencies, future efforts could focus on identifying priority areas for further performance improvements in LEDs and SSL in general.
-Fourth, better understanding is needed about the role of demand-side factors, such as policies stimulating market demand for LED lighting (e.g., through incandescent light bulb bans or subsidies for LED adoption), not only in the rapid diffusion of LED-based lighting technologies, which is reasonably well understood by now \cite{Mills2014, Kamat2020, weinold2020technology, stegmaier2021incandescent, grubb2021new}, but also in facilitating learning-by-doing, innovation and technology spillovers that have contributed to the technological progress in white LEDs.
+Fourth, better understanding is needed about the role of demand-side factors, such as policies stimulating market demand for LED lighting (e.g., through incandescent light bulb bans or subsidies for LED adoption), not only in the rapid diffusion of LED-based lighting technologies, which is reasonably well understood by now \cite{Mills2014, Kamat2020, weinold2021quantifying, stegmaier2021incandescent, grubb2021new}, but also in facilitating learning-by-doing, innovation and technology spillovers that have contributed to the technological progress in white LEDs.
 Finally, the lessons learned about the drivers of innovation, technology spillovers, cost reductions and performance improvements in LEDs may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}.
@@
 \subsection{Performance Metrics Calculations}
+\label{subsec:performance_metrics}
 

### 2023-08-16 2_SSL_NE/article/article_main.tex
updated figures

@@
  \caption{\textbf{Contribution of innovations and technology spillovers to the historical progress in sub-efficiencies of phosphor-converted warm white LEDs.} LEDs with test currents of at least 350mA are considered.
-The following sub-efficiencies are represented: $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $η_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C_R}$ - conversion efficiency for red phosphors; $\eta_{C_{Y/G}}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency.
+The following sub-efficiencies are represented: $\eta_{V_f}$ - forward voltage efficiency; $\eta_IQ$ - internal quantum efficiency; $\eta_{droop}$ - efficiency droop; $\eta_{LE}$ - light extraction efficiency; $\eta_{C_R}$ - conversion efficiency for red phosphors; $\eta_{C_{Y/G}}$ - conversion efficiency for yellow/green phosphors; $\eta_S$ - spectral efficiency.
 The overall LED lamp efficiency $\eta_L$ is displayed in the rightmost column.
@@
  \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
- \caption{Manufacturing cost structure modelled for a single low-to-mid power GaN, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+ \caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and \cref{costmodel_details}) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
 location.
-Consecutive steps shown on the x-axis.
-Panels A-C: Waterfall diagrams of LED manufacturing cost split by manufacturing process steps for years 2003, 2012 and 2020.
+Panels A-C: Waterfall diagrams of LED manufacturing cost structure split by manufacturing process steps for years 2003, 2012 and 2020.
 Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
-Panel F: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
-For a visual representation of the overview of the manufacturing process, compare the diagrams in Section 2 of the Supplementary Information.
+Panel D: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
+For a graphic representation of the manufacturing process, see the diagrams in \cref{subsec:manufacturing_steps}.
 Abbreviations: Litho – Lithographic Process, Insp.
 – Inspection, Depn.
-– Deposition, CMP - Chemical-Mechanical Planarization}
+– Deposition, CMP - Chemical-Mechanical Planarization.}
  \label{fgr:costmodel}

### 2023-08-16 2_SSL_NE/article/article_supplementary.tex
added edits
updated figures

@@
 % nature energy custom code
-
-\newcounter{supplementarynote}
-\renewcommand{\thesupplementarynote}{\arabic{supplementarynote}}
-\newcommand{\supplementarynote}{%
-  \refstepcounter{supplementarynote}%
-  Supplementary Note~\thesupplementarynote:~%
-}
-\titleformat{\section}
-  {\normalfont\large\bfseries}
-  {\supplementarynote}
-  {0em} % Adjust this value to control the spacing
-  {}
-\renewcommand{\thesection}{\thesupplementarynote}
+\makeatletter
+\renewcommand{\@seccntformat}[1]{Supplementary Note \csname the#1\endcsname\quad}
+\makeatother
 
@@
 
-Light conversion efficiency (LCE) describes the losses in the conversion process of blue light that is the basis for all phosphor-converted white LEDs.
-These losses are the sum of the Stokes loss and  scattering/absorption losses.
-It is defined as
+Light conversion efficiency (LCE) describes the sum of all losses in the process of converting the blue light that is the basis for all phosphor-converted white LEDs.
+Specifically, energy is lost due to three distinct effects.
+First, the inherent wavelength-conversion loss or \textit{Stokes shift} as the difference between the original blue wavelength and the down-converted red, yellow or green wavelength.
+Second, non-radiative processes in the phosphor material.
+Third, scattering and absorption processes in the phosphor material.
+These three effects are listed individually in \cref{fgr:waterfall}.
+The overall light conversion efficiency is defined as
 %
@@
 %
-where $E_i$ denotes the total energy of light at the color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
+where $E_i$ denotes the total energy of light at color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
 Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
@@
 \subsubsection{Manufacturing Process Steps by Chip Architecture}
+\label{subsec:manufacturing_steps}
 

### 2023-08-16 2_SSL_NE/article/article_main.tex
some more edits

@@
 
-Our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid power GaN-based white classic-chip LEDs from 1.11$ to 0.05$ (in 2020 USD) between 2003 and 2020.
-In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size , with the former factor contributing to XX\% and latter factor contributing to YY\% of the overall LED cost reduction over time .
+Our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid power GaN-based white classic-chip LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020.
+In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.
 Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing.

### 2023-08-17 2_SSL_NE/article/article_main.tex
added edits

@@
     pdftitle={Main Article},
-    pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
+    pdfauthor={Michael Philipp Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
 }
@@
 
-M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
+M. P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
 \newline
-$^1$ CEENRG, Dept. of Land Economy, University of Cambridge, Cambridge,  UK \\
+$^1$ Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, UK \\
 $^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
@@
  \includegraphics[width=\textwidth]{figures/history_efficacy.pdf}
- \caption{\textbf{Historical progress in the luminous efficacy ($K$) of the most widely-used lighting technologies in human history.} Data points indicate best performers by year of market introduction.
+ \caption{\textbf{Historical progress in the luminous efficacy ($\eta$) of the most widely-used lighting technologies in human history.} Data points indicate best performers by year of market introduction.
 Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details).
@@
 The following LED architectures and down-conversion phosphor spectra used in calculations in each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor.
-Details for each architecture are provided in Figure 3.
-Details on the phosphors for light down conversion are provided in Table 1 and Supplementary Note 4.
+Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.
+Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4.
 Abbreviations: Scat.
@@
 = Nonradiative.
-Source (Efficiency): own elaboration based on data in Figure SI11.
-Source (spectral data): Adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
+Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10,  adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
  \label{fgr:waterfall}
@@
     \end{NiceTabularX}
-    \caption{\textit{Year} column represents the earliest reported application of innovation in white LEDs.
+    \caption*{Note: \textit{Year} column represents the earliest reported application of innovation in white LEDs.
 These differ from the years shown in \cref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for LED products that relied on those innovations.
@@
     \end{NiceTabularX}
-    \caption{Note: Disc.
+    \caption*{Note: Disc.
 - Year of initial discovery, identified  by the earliest literature source describing the discovery of original idea or invention outside the LED domain.
@@
 
+In this study, we use a multi-method approach to synthesize evidence from multiple sources and, for the first time, reconstruct a comprehensive picture of the rapid technological progress in white LED technology throughout its history since the introduction to the market in 1996 and across an ensemble of metrics related to device performance, cost, and consumer experience.
+Improvements in these metrics are traced to sources in innovation, technology spillovers, and economies of scale in the manufacturing process.
+
 We find that the total LED device efficiency increase from 5.8\% to 38.8\% between 2003 and 2020, as well as all major improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that reduced energy losses across all physical loss channels in LED devices and introduced new phosphor materials for LED light down conversion.
@@
 In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.
-Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing.
 Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time.
-[PLACEHOLDER TO DISCUSS THE REASONS FOR THAT].
+There are two reasons for this:
+
+First, the \textit{functional unit} of our study is the LED chip itself, not the luminaire containing multiple chips.
+As efficiency and overall LED brightness has increased, a smaller number of chips could be used in luminaires to obtain the same level of overall flux.
+However, since we are investigating improvements at the level of single chips, we do not capture cost effects downstream of chip manufacturing.
+Nonetheless, we were previously able to show that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
+
+Second, we find that an increase in wafer size and higher overall yield together are responsible for a 98\% reduction of manufacturing cost per LED die between between 2003 and 2020.
+Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, which was confirmed by multiple interviews with experts from manufacturing.
+A plausible explanation for this finding is that the main focus of R&D activities in white LED technology in the considered period was likely on LED performance improvements rather than cost reductions.
 
@@
 By comparing these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
-
-\section{Conclusions}
-
-Our study has analysed the sources of the dramatic cost reductions and performance improvements in white light-emitting diodes since their introduction to the market in 1996 in a systematic and granular way.
-We find that the total LED device efficiency increase from 5.8\% to 38.8\% between 2003 and 2020, as well as improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that affected all physical energy loss channels and corresponding device sub-efficiencies.
-We also find that among those innovations, at least nine were driven by knowledge spillovers originating in areas of science and technology beyond LEDs or solid-state lighting.
-These spillover-driven innovations were responsible for 8.5\% of the total efficiency improvements and nearly 100\% of the improvements in consumer experience metrics.
-
-Our manufacturing cost model shows a 95.5\% decrease in the cost of producing white classic-chip LEDs (from 1.11\$(2020) to 0.05\$(2020)) between 2003 and 2020, driven mostly by increases in the wafer size and yields across different manufacturing steps over time.
-In contrast with performance improvements, these cost reductions were thus a result of economies of scale and learning by doing,  which were in turn likely facilitated by policies creating and growing market demand for LED lighting, such as incandescent light bulb bans and subsidies for LEDs.
-
-Our analysis of the sources, mechanisms and enablers of the identified technology spillovers which were significant drivers of improvements in LED efficiency and consumer experience metrics, highlights the critical role played by a deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.
-Specifically, deep physical understanding of LED device efficiency loss channels enabled important innovations in LEDs that increased several sub-efficiencies in LEDs and will continue to do so, as expected by eminent experts in the field \cite{Weisbuch2020}.
-This suggests that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology.
-These knowledge spillovers can be enabled or even accelerated by knowledge exchange events and long-term partnerships between academia and industry, dedicated mission-driven public R\&D funding, and a freedom of search in academia.
-This further reinforces arguments made against the dichotomy of basic research versus applied research \cite{narayanamurti2016cycles,narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
-
-There are various important avenues of future research that are opened up by our analysis.
-First, future work could expand the cost model by collecting and including data for a broader set of chip architectures and analyzing the impact of individual innovations and spillovers on the costs as opposed to performance.
-Second, a deeper dive on the role of learning-by-doing is needed both in the cost and performance analysis.
-Third, building on the work on LED sub-efficiencies and physical limits, future efforts could focus on identifying priority areas for further efficiency improvements in LEDs and SSL in general.
-Finally, by comparing the drivers of innovation, technology spillovers, cost reductions and performance improvements at a granular level across different clean energy technologies, we can identify patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
 

### 2023-08-17 2_SSL_NE/article/article_supplementary.tex
updated custom names
some more edits
added edits

@@
 
-% nature energy custom code
+% cross-referencing
+\input{2_SSL_NE/article/auxiliary/preamble/1_preamble_functionality_crossreference}
+
+% custom section designations
 \makeatletter
-\renewcommand{\@seccntformat}[1]{Supplementary Note \csname the#1\endcsname\quad}
+\renewcommand{\@seccntformat}[1]{Supplementary Note \csname the#1\endcsname: } % https://tex.stackexchange.com/a/693618
 \makeatother
+% custom equation designations
+\renewcommand{\figurename}{Supplementary Figure}
+\crefname{figure}{Supplementary Figure}{Supplementary Figures}
+\renewcommand{\tablename}{Supplementary Table}
+\crefname{table}{Supplementary Table}{Supplementary Tables}
 
@@
 
-% custom figure and table numbering
-\renewcommand{\thefigure}{Supplementary Figure \arabic{figure}}
-\renewcommand{\thetable}{SI\arabic{table}}
-
-% custom equation numbering
-% https://tex.stackexchange.com/a/184641/
-\newcounter{defcounter}
-\setcounter{defcounter}{0}
-\newenvironment{myequation}{%
-\addtocounter{equation}{-1}
-\refstepcounter{defcounter}
-\renewcommand\theequation{SI\thedefcounter}
-\begin{equation}}
-{\end{equation}}
-
 %% METADATA %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     pdftitle={Supplement},
-    pdfauthor={Michael Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
+    pdfauthor={Michael Philipp Weinold, Sergey Kolesnikov, Laura Diaz Anadon},
 }
@@
         \newline 
-        M. Weinold$^{1,2,3}$, S. Kolesnikov$^3$, L.D. Anadon$^{3,4}$ \\
-        "Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers" (2023) \\
+        M. P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
         \newline
-        $^1$ Technology Assessment Group, Paul Scherrer Institute, Switzerland \\
+        $^1$ Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, UK \\
         $^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
-        $^3$ CEENRG, Dept. of Land Economy, University of Cambridge, UK \\
+        $^3$ Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland \\
         $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
@@
 
+\clearpage
 \tableofcontents
 
-\newpage
-
-\section{Previous Literature on Technological Progress in LEDs}
+\clearpage
+\section{Previous Literature on Techological Progress in LEDs}
 \label{sec:prev_lit}
@@
 Due to these limitations, any recommendations on future improvements in LED technology based on these studies are fragmented at best.
-Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2020novel,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan_2021}, has not yet been studied in the context of lighting.
+Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2020novel,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
 
@@
 
-LED light-extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+LED light extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
 %
@@
 %
-where $P_{opt}$ is the optical power emitted from the device $P_{int}$ is the optical power emitted from the active region into other parts of the device (internal) \cite{schubert2018light}\cite{tsao2010solid}.
+where $P_{opt}$ is the optical power emitted from the device and  $P_{int}$ is the optical power emitted from the active region into other parts of the device (internal) \cite{schubert2018light}\cite{tsao2010solid}.
 
@@
 Specifically, energy is lost due to three distinct effects.
-First, the inherent wavelength-conversion loss or \textit{Stokes shift} as the difference between the original blue wavelength and the down-converted red, yellow or green wavelength.
-Second, non-radiative processes in the phosphor material.
-Third, scattering and absorption processes in the phosphor material.
-These three effects are listed individually in \cref{fgr:waterfall}.
+The first effect is the inherent wavelength-conversion loss or \textit{Stokes shift} as the difference between the original blue wavelength and the down-converted red, yellow or green wavelength.
+The second effect are non-radiative processes in the phosphor material.
+The third , scattering and absorption processes in the phosphor material.
+These three effects are listed individually in Figure 4.
 The overall light conversion efficiency is defined as
@@
 %
-where $E_i$ denotes the total energy of light at color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
+where $E_{\textcolor{blue}{B}}$ is the total energy of blue light before down-conversion and $E_i$ denotes the total energy of light at color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
 Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
@@
 
-Spectral efficiency (SE) describes losses in the conversion process due to the wavelength-dependent efficiency of the human eyes.
-Photons converted into the infrared or ultraviolet are lost to illumination purposes.
+Spectral efficiency (SE) describes losses in the light down conversion process due to the wavelength-dependent efficiency of the human eyes.
+Photons converted into the infrared or ultraviolet light are lost to illumination purposes.
 It is defined as
@@
 %
-where $K$ is the luminous efficacy of radiation of the light source, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
-$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
+where $K$ is the luminous efficacy of radiation of the light source, defined in \cref{eqn:ler}, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
+$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering index performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
 
@@
 
-The luminous efficacy \textit{of radiation} $K$ describes the match of a light-emitting diode package spectrum to the human visual system.
+The luminous efficacy \textit{of radiation} $K$ is a supplementary metric used in our calculations of SE, rather than directly in our analysis of the improvements in LED performance.
+Care must be taken not to confuse this metric with the luminous efficacy of a light source, described in \cref{subsec:les}.
+
+The luminous efficacy \textit{of radiation} $K$ describes the match of the spectrum of a light-emitting diode to the human visual system.
 Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
 A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
-The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
-%
-\begin{myequation}
+The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source $\phi$ with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
+%
+\begin{equation}
 \label{eqn:ler}
     K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
-\end{myequation}
+\end{equation}
 %
@@
 
-Care must be taken not to confuse this metric with luminous efficacy of source described in \cref{subsec:les}.
-
 \subsubsection{Luminous Efficacy of Source}
@@
 
-The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux and the consumed electrical power $P_{el}$ \cite{cie-term-effsrc}.
-It is often used to describe general progress in lighting technologies.
-%
-\begin{myequation}
+The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux $\phi$ and the consumed electrical power $P_{el}$ \cite{cie-term-effsrc}.
+%
+\begin{equation}
     \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
-\end{myequation}
+\end{equation}
 %
@@
 CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
-\cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DOE2016LED}.
+However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DoE2016LED}.
 High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.

### 2023-08-17 2_SSL_NE/article/article_main.tex
added edits

@@
 \addbibresource{bibliography.bib}
+
+% number all lines as pre journal submission guidelines
+\usepackage{lineno}
+\linenumbers
 
@@
 
-\cref{fgr:subeff} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
+\cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
 The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in \cref{sec:discussion}.
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
  \centering
@@
 
-\begin{table}[h!]
+\begin{table}[H]
     \footnotesize
@@
 We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
+Our model shows that for the wafer-level manufacturing steps illustrated in Supplementary Figures 2-6, yield improvements and an increase in  wafer size together are responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.
 
@@
 
-\begin{figure*}[h!]
+\begin{figure}[ht!]
 \centering
 \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
- \caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and \cref{costmodel_details}) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+\caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and (see Methods and Supplementary Note 3) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
 location.
@@
 Panel D: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
-For a graphic representation of the manufacturing process, see the diagrams in \cref{subsec:manufacturing_steps}.
+For a graphic representation of the manufacturing process, see the diagrams in Supplementary Figures 2-6.
 Abbreviations: Litho – Lithographic Process, Insp.
@@
 \label{fgr:costmodel}
-\end{figure*}
+\end{figure}
 
@@
 In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.
+
 Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time.
-There are two reasons for this:
+There are several explanations for this finding.
+First, the functional unit of our study is the LED chip itself, not the luminaire containing multiple chips.
+As LED efficiency and overall brightness have increased, a smaller number of chips can be used in luminaires to obtain the same level of the overall light flux.
+However, since we are investigating improvements at the level of single chips, we do not capture the cost effects downstream of chip manufacturing.
+We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
 
-First, the \textit{functional unit} of our study is the LED chip itself, not the luminaire containing multiple chips.
-As efficiency and overall LED brightness has increased, a smaller number of chips could be used in luminaires to obtain the same level of overall flux.
-However, since we are investigating improvements at the level of single chips, we do not capture cost effects downstream of chip manufacturing.
-Nonetheless, we were previously able to show that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
-
-Second, we find that an increase in wafer size and higher overall yield together are responsible for a 98\% reduction of manufacturing cost per LED die between between 2003 and 2020.
-Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, which was confirmed by multiple interviews with experts from manufacturing.
-A plausible explanation for this finding is that the main focus of R&D activities in white LED technology in the considered period was likely on LED performance improvements rather than cost reductions.
+Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.
+Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular LED innovations, which was confirmed by interviews with experts in LED manufacturing.
 
@@
 
+Finally, the lessons learned about the drivers of innovation, technology spillovers, cost reductions and performance improvements in LEDs may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}.
+By comparing these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
+
 There are several important avenues of future research that are opened up by our analysis.
@@
 Fourth, better understanding is needed about the role of demand-side factors, such as policies stimulating market demand for LED lighting (e.g., through incandescent light bulb bans or subsidies for LED adoption), not only in the rapid diffusion of LED-based lighting technologies, which is reasonably well understood by now \cite{Mills2014, Kamat2020, weinold2021quantifying, stegmaier2021incandescent, grubb2021new}, but also in facilitating learning-by-doing, innovation and technology spillovers that have contributed to the technological progress in white LEDs.
-Finally, the lessons learned about the drivers of innovation, technology spillovers, cost reductions and performance improvements in LEDs may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}.
-By comparing these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
 
 \section{Methods}
+\label{sec:methods}
 
@@
 
-To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a systematic review of the primary scientific literature, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies.
+To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a systematic review of the primary scientific literature \cite{Haddaway_2014}, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies.
 We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in innovation and technology spillovers.

### 2023-08-17 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 \input{2_SSL_NE/article/auxiliary/preamble/1_preamble_functionality_crossreference}
+
+% number all lines as pre journal submission guidelines
+\usepackage{lineno}
+\linenumbers
 
@@
 This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
-In 2000, the highest performing devices yielded around 10lm, just below the output of a candle as defined in the unit candela (1cd=12.57lm)\cite{haitz2011solid}.
-Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law” \cite{haitz1999case,haitz2011solid}.
+In 2000, the highest performing devices yielded around 10 lm, just below the output of a candle as defined in the unit candela (1 cd=12.57 lm) \cite{haitz2011solid}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
 Today, however, the devices with the highest performance yield in excess of 1600 lm, the equivalent of a 100 W incandescent bulb \cite{cree2020bright}.
@@
 
-We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe the different physical loss channels within the device to describe progress in light-emitting diode technology.
+We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe different physical loss channels within the device to describe progress in LED technology.
 In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
-The equations defining these sub-efficiencies are provided in the Supplementary Information in Section 1.3.
+The equations defining these sub-efficiencies are provided below.
 The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:

### 2023-08-18 2_SSL_NE/article/article_supplementary.tex
added edits

@@
         \newline 
-        M. P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
+        M.P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
         \newline
@@
             \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
-            \caption{Manufacturing process for a classical LED package with lateral current spreading, circa 2003. MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
+            \caption{\textbf{Manufacturing process for a classical LED package with lateral current spreading, circa 2003.} MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
         \label{fig:manuf_classical_2003}
@@
             \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
-            \caption{(1/2) Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
-MQW - multiple quantum well; LLO - laser lift-off.
+            \caption{\textbf{Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
+(1/2)} MQW - multiple quantum well; LLO - laser lift-off.
 Continued in \cref{fig:manuf_vtf_2012-2}}
@@
             \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
-            \caption{(2/2) Continued from \cref{fig:manuf_vtf_2012-1}. Encaps. - encapsulation.}
+            \caption{\textbf{Manufacturing process for a vertical thin-film package flip-chip LED chip with vertical current spreading, circa 2012 (2/2)}.
+Continued from \cref{fig:manuf_vtf_2012-1}.
+Encaps.
+- encapsulation.}
             \label{fig:manuf_vtf_2012-2}
@@
             \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
-            \caption{(1/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
-MQW - multiple quantum well.
+            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+(1/2)} MQW - multiple quantum well.
 Continued in \cref{fig:manuf_csp_2020-2}.}
@@
             \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
-            \caption{(2/2) Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+(2/2)}.
 Continued from \cref{fig:manuf_csp_2020-1}.
@@
 %
-The number of die per wafer $D$ depends on the total usable wafer area.
-The usable area depends on the wafer diameter, the cutting street width between the chips and the exclusion zone at the rim of the wafer.
+The number of die per wafer $D$ depends on the total usable wafer area $A_{\text{usable}}$.
+The usable area depends on the total area of the wafer $A_{\text{wafer}}$ , the cutting street width area between the chips $A_{\text{cut}}$  and the exclusion zone at the rim of the wafer $A_{\text{exclusion}}$ .
 %
@@
 
-\cref{fig:dpw} shows the results of DPW calculations using \cref{eqn:dpw} for different wafer diameters $d$ that were most commonly used in LED manufacturing in $d(2002)=50$mm, $d(2010)=100$mm and $d(2020)=200$mm.
+\cref{fig:dpw} shows the results of DPW calculations using \cref{eqn:dpw} for different wafer diameters $d$ that were most commonly used in LED manufacturing in 2003 (50mm), 2010 (100mm) and 2020 (200mm).
 We can see that DPW dramatically increases over time.
@@
     \includegraphics[width=5.5cm]{./figures/die-per-wafer.pdf}
-    \caption{Number of die per wafer for the different wafer sizes used in different years: (DPW) $d(2002)=50$mm$\rightarrow851$DPW, $d(2010)=100$mm$\rightarrow5370$DPW, $d(2020)=200$mm$ \rightarrow26,838$DPW.
+    \caption{\textbf{Number of die per wafer for the dominant wafer diameters used in different years.} $d(2003)=50$mm$\rightarrow851$ DPW, $d(2010)=100$mm$\rightarrow5370$ DPW, $d(2020)=200$mm$\rightarrow26,838$ DPW.
 Source: own estimates, based on \cref{eqn:dpw} and the dominant wafer size of the considered year, informed by expert interviews.
@@
 
-\subsubsection{Example of Input Data Considered: Sapphire Wafers}
+\subsubsection{Input Data on Sapphire Wafers}
 
@@
 
-Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing (see top panel of \cref{fig:wafers}).
+Wafer sizes used in manufacturing have also increased over the years due to favourable economics of large wafer processing (see bottom panel in \cref{fig:wafers}).
 The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}.
@@
     \includegraphics[width=14.5cm]{./figures/wafer_size.pdf}
-	\caption{Historical data for sapphire substrate prices of different surface properties and diameters.
-Shown is data for polished surface (NPS) and patterned substrates (PSS).
-\textit{Monocrystal} denotes the Russian manufacturer of the same name.
-Dashed lines are projections from the previous year.
-Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}.
-Bottom: Prevalence of sapphire wafer size used in the manufacturing of light-emitting diodes.
+	\caption{\textbf{Historical trends in sapphire wafer size and price.} Top panel: Historical data for sapphire substrate prices of different surface properties and diameters, 2004-2020.
+Shown is data for non-patterned (polished) surface (NPS) and patterned surface substrates (PSS).
+Monocrystal denotes the Russian manufacturer of the same name.
+Dashed lines are projections made in 2015.
+Sources: (Monocrystal 2020)(Yole Developpement 2011)(Yole Developpement 2015).
+Bottom panel: Prevalence of sapphire wafer size used in LED manufacturing in 2008-2020.
 Hatched bars are projections.
@@
 To quantify the drivers of changes in manufacturing cost or device performance among many contributing factors, one would need to identify the magnitude of contribution to these changes made by single variables in equations \cref{eqn:cost_wafer} and \cref{eqn:cost_die}.
-Mathematically, given a function $C$ which describes the manufacturing cost or performance function of a device at time $t$,
+Mathematically, given a function $F$ which describes the manufacturing cost or performance function of a device at time $t$,
 %
@@
 %
-Index decomposition analysis has been developed to solve this problem \cite{Boyd1987}. A short derivation, based on \cite{kavlak2018evaluating} is provided here:
+Index decomposition analysis has been developed to solve this problem \cite{Boyd1987}. A short derivation, based on \cite{kavlak2018evaluating} is provided below.
 
@@
 %
-where $x_i$ is an input cost variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
+where $x_i$ is an input variable. The contribution of the change $\Delta x_1$ in variable $x_1$ to the total function value $F$ over the period $t_0 < t < t_1 $ is then
 %
@@
 
-The function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
+In this approach, the function $F$ as a function of a vector of input model variables $\vec{r}=(r_1,r_2,\dots)$ is defined as
 %
@@
 %
-We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in section 4.3 and shown in Figure 5 in the main article.
+We use this approach to estimate the effect of individual innovations and spillovers on LED device performance, as discussed in relation to Figure 5 in the main article.
 Due to time constraints, we leave the quantification of the impact of different drivers of cost reductions with this methodological approach for our future work.
@@
 Empirical data on the historical progress in metrics related to four different sub-efficiencies contributing to the overall LED efficiency (lamp efficiency $\eta_L$) is shown in \cref{fgr:subeff}.
-Specifically, the figure includes data for forward voltage used to calculate to forward voltage efficiency $\eta_{V_f}$ (Panel A1); efficiency droop $\eta_droop$ (Panel A2); internal quantum efficiency $\eta_IQ$ (Panel B1); and light extraction efficiency $\eta_LE$ (Panel B2).
-Additional information and context for individual panels in Supplementary \cref{fgr:subeff} are provided below.
+Specifically, the figure includes data for forward voltage used to calculate to forward voltage efficiency $\eta_{V_f}$ (Panel A1); efficiency droop $\eta_{droop}$ (Panel A2); internal quantum efficiency $\eta_{IQ}$ (Panel B1); and light extraction efficiency $\eta_{LE}$ (Panel B2).
+Additional information and context for individual panels in \cref{fgr:subeff} are provided below.
 
@@
 
-\textbf{Panel B1:} Internal quantum efficiency, for different chip architectures, by type of measurement used.
-Note that the artifact in the DOE Average around 2013 is due to a change in definition for internal quantum efficiency laid out in \cite{doe_ssl_multiyear_2013}.
-Measurement methods: PL - Photoluminescence\cite{Shim_2018}, EL - Electroluminescence\cite{Getty_2009};  ABC Model\cite{Karpov_2014}, Ideality Factor\cite{Masui_2010}.
+\textbf{Panel B1:} Internal quantum efficiency (IQE) for different chip architectures by type of measurement used.
+Note that the bump in the DOE Average around 2013 is due to a change in IQE definition laid out in \cite{doe_ssl_multiyear_2013}.
+Measurement methods: PL - Photoluminescence \cite{Shim_2018}, EL - Electroluminescence \cite{Getty_2009}; ABC Model \cite{Karpov_2014}, Ideality Factor \cite{Masui_2010}.
 “Not reported” denotes data points where the IQE measurement technique was not reported.
-Whiskers indicate reported ranges, where applicable.
+Whiskers indicate reported ranges, where available.
 \
 
-\textbf{Panel B2:} Light extraction efficiency for different chip architectures, by type of specific technology used to improve light-extraction efficiency.
-The source of data points is shown for reference: simulation - ray-tracing computer simulation; experimental (absolute) - light-extraction efficiency given directly; experimenal (relative) - relative improvement over baseline chip architecture given.
+\textbf{Panel B2:} Light extraction efficiency (LEE) for different chip architectures by type of specific technology used to improve LEE.
+The types of data points are shown for reference: Simulation - ray-tracing computer simulation; Experimental (absolute) – direct LEE values; Experimental (relative) - relative improvement over baseline chip architecture.
 Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
@@
 Notably, \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.
-Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the empirical \cref{fgr:subeff} not relevant.
-Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum}.
+Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the (mostly) empirical \cref{fgr:subeff} not relevant.
+Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum} below (see also Methods section in the main text).
 
@@
 \includegraphics[width=15.5cm]{./figures/subefficiencies_progress.pdf}
-\caption{Historical improvements in light-emitting diode technology.
-Panels A1-A2 show they key metrics used to compute forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
-Panels B1-B2 show the sub-efficiencies internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
-Source: own synthesis of data from the full list of sources provided in \cref{sec:sources}.
+\caption{\textbf{Historical improvements in LED sub-efficiencies.} Panels A1-A2 show data for key metrics used to calculate forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
+Panels B1-B2 show data for internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
+Source: own synthesis of data from the list of sources provided in \cref{sec:sources}.
 Data on average device performance adapted from U.S.
@@
 
-\begin{table}[h!]
-    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.
-The table includes innovations and technology improvements affecting forward voltage efficiency, light extraction efficiency, internal quantum efficiency, and light conversion efficiency.
-For a list of innovations affecting spectral efficiency, see Table 1 in the main article.}
+\begin{table}[H]
+    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.}
     \vspace{5mm}
@@
         \hline
-        \multicolumn{5}{c}{Forward Voltage Efficiency (VFE) $\eta_{Vf}$} \\
+        \multicolumn{5}{c}{Forward Voltage Efficiency (FVE) $\eta_{Vf}$} \\
         \hline
@@
         \hline
-            1994 & Double heterostructure & Higher Radiative recombination probability & No & \cite{Nakamura1994} \\
-        \hline
-            1996 & Multiple quantum well & Higher Radiative recombination probability & No & \cite{Koike1996} \\
+            1994 & Double heterostructure & Higher radiative recombination probability & No & \cite{Nakamura1994} \\
+        \hline
+            1996 & Multiple quantum well & Higher radiative recombination probability & No & \cite{Koike1996} \\
         \hline
@@
         \hline
-            ongoing & Lower Current Density & Current Density & No & I \\
-        \hline
-            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{bhardwaj2016progress} \\
-        \hline
-            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect Density & No & I, \cite{Wildeson2017} \\
-        \hline
-            $<$ 2017 & Defect Getting \newline Underlayer & Defect Density & No & I, \cite{haller2017burying} \\
-        \hline
-        \multicolumn{5}{c}{Light Conversion efficiency (CE) $\eta_{C}$} \\
+            ongoing & Lower current Density & Current density & No & I \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect getting \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
+        \hline
+        \multicolumn{5}{c}{Light conversion efficiency (CE) $\eta_{C}$} \\
         \hline
@@
         \hline
-            $<$ 2017 & Defect gettering \newline underlayer & Defect Density & No & I, \cite{haller2017burying} \\
+            $<$ 2017 & Defect gettering \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
         \hline
@@
 An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
-Used in combination with red-emitting sulfide phosphors, by 2002 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}.
+Used in combination with red-emitting sulfide phosphors, by 2003 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2003}.
 However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased.
@@
 The luminescent properties of these materials were identified by the Schnick’s group, by then at Ludwig-Maximilians University of Munich, in 2000 \cite{Hppe2000} after a suggestion made to Schnick at a conference following earlier reports of good luminosity properties of europium-doped compounds \cite{Qiua1998}.
-U.S.-based LED manufacturer Lumileds applied for a patent for the first class of red LED phosphors based on the 258 nitridosilicate chemistry in 2002 \cite{mueller2004phosphor}.
+U.S.-based LED manufacturer Lumileds applied for a patent for the first class of red LED phosphors based on the 258 nitridosilicate chemistry in 2003 \cite{mueller2004phosphor}.
 The first use of the 258 phosphor in a commercial “Luxeon” LED package was then reported in a joint publication co-authored by inventors from Lumileds and researchers from the Schnick’s group in 2005 \cite{MuellerMach2005}.
@@
     \includegraphics[width=0.9\textwidth]{./figures/phosphor_spectrum-comparison.pdf}
-	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 6 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year.
+	\caption{\textbf{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data.} Corresponding LED products that used phosphor innovations listed in Table 1 in the main article are each indicated by the phosphor label and spectral data publication year.
 Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 6 in the main article.
-The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow.
-Metrics were calculated from spectral data shown in the two bottom panels using the \texttt{colour-science} package for Python \cite{colour-science_software}.
-Bottom two rows of panels: Corresponding spectral data.
+The desirable direction of improvements towards higher LER at higher CRI is indicated by a red arrow.
+Metrics were calculated from spectral data shown in the bottom panels using the \texttt{colour-science} package for Python \cite{colour-science_software}.
+Bottom two rows of panels: spectral data for the first identified LED products that used phosphor innovations listed in Table 1 in the main article.
 The luminosity function \cite{cie-term-lumeff}, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
-Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel.
+Peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel.
 Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: ?
@@
 Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum.
-Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
+Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2003}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
 \label{fig:phosphor_spectrum}
@@
 The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
-As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen because the prices quotes from the United State Geological Survey price database \cite{usgeoprices}.
+As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen based on the prices quotes from the United State Geological Survey price database \cite{usgeoprices}.
 Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
@@
 \small
-    \caption{Cost model sensitivity analysis parameter list. The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.}
+    \caption{List of LED manufacturing cost model parameters considered in sensitivity analysis}
     \vspace{5mm}
@@
         \end{NiceTabularX}
-    \caption*{Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
+    \caption*{The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.
+Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
 If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
@@
 Discount - equipment discount (sales rebate for large purchases); Inspec.
-Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).}
+Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).
+’I’ in the Source column indicates expert interviews as the source of information.}
     \label{tab:sensitivity}
@@
 \clearpage
-\section{Cost Comparison with Reported Industry Data and DOE Projections}
+\section{Cost Model Comparison with Industry Data and DOE Projections}
 
@@
     \includegraphics[width=\textwidth]{./figures/costmodel_calibration.pdf}
-	\caption{Top panel: manufacturing cost calculations (2008-2016) and projections (2018, 2020) previously published by the U.S.
+	\caption{\textbf{Cost structure comparison between our cost model, industry data and DOE projections.} Top panel: manufacturing cost calculations (2008-2016) and projections (2018, 2020) previously published by the U.S.
 DOE based on industry data and LEDCOM model.
@@
     \hline
-    LED & own research, compare \cite{zenodo_weinold_led_history} \\
+    LED & Own research, compare \cite{zenodo_weinold_led_history} \\
     \hline
@@
     \hline
-    Government Report (US) & \cite{council2013assessment} \\
+    Government Report (US) & \cite{national2013assessment} \\
     \hline
@@
     \hline
-    Panel A1 (VFE) & \cite{nichia2001data}\cite{lumi2002data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
+    Panel A1 (FVE) & \cite{nichia2001data}\cite{lumi2003data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
     \hline
@@
     Panel B2 (LEE) & \cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012} \newline \cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010}
-\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
+\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2003}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
     \hline

### 2023-08-18 2_SSL_NE/article/article_main.tex
added edits
more edits

@@
 
-M. P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
+M.P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
 \newline
@@
 We then trace the historical improvements in these metrics, along with associated changes in LED chip architectures, from the time of introduction of the first commercial warm white GaN-based LEDs in 2003 to 2020, the year with the most recent data available at the time of data collection.
-Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry (see Methods section and Supplementary Note 3 for details).
+Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a comprehensive review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry (see Methods section and Supplementary Note 3 for details).
 This approach allows us to identify individual innovations in white LED technology, examine their knowledge origins to discern technology spillovers among them and quantify their contributions to the overall progress in LED efficiency and consumer experience metrics based on our modeling of LED performance and spectral metrics.
@@
 \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
-\caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and (see Methods and Supplementary Note 3) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+\caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and Supplementary Note 3) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
 location.
@@
 
-To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a systematic review of the primary scientific literature \cite{Haddaway_2014}, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies.
+To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a comprehensive review of the primary scientific literature \cite{Haddaway_2014}, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies.
 We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in innovation and technology spillovers.
@@
 
-\subsection{Systematic Literature Review}
+\subsection{Comprehensive Literature Review}
 
-We collected data on LED performance and characteristics in a systematic literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
+We collected data on LED performance and characteristics in a comprehensive literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
 This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics; 2) identifying individual innovations that contributed to this evolution, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of science or technology outside of SSL, making them involved in technology spillovers.
@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
 

### 2023-08-18 2_SSL_NE/article/article_supplementary.tex
more edits

@@
     \hline
-        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Expertise} \\ \hline
+        \textit{\#} & \textit{Sector} & \textit{Role} & \textit{Country} & \textit{Area of Expertise} \\ \hline
         1 & Academia & Senior researcher & UK & Epitaxy \\ \hline
@@
     \includegraphics[width=0.95\textwidth]{./figures/costmodel.pdf}
-    \caption{Schematic diagram of the cost model showing inputs to each step and computational steps leading to the cost model outputs.
-For the description of cost variables, see definitions in \cref{eqn:cost_wafer}.
-For the computation of total cost, see \cref{subsec:cost_computation}.
-For the computation of changes in total cost, see \cref{sec:contribution_variables}.}
+    \caption{\textbf{Cost model structure.} The diagram shows inputs to each step and computational steps leading to the cost model outputs.
+For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.}
     \label{fig:costmodel-schematic}

### 2023-08-18 2_SSL_NE/article/article_main.tex
added edits

@@
 Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}.
-While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2020novel}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2020novel} by considering technology spillovers as the application of external knowledge in a technology where knowledge is considered external  if it has been developed for application in other technologies, sectors, or scientific disciplines.
+While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2021spillovers}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2021spillovers} by considering technology spillovers as the application of external knowledge in a technology where knowledge is considered external  if it has been developed for application in other technologies, sectors, or scientific disciplines.
 There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies.

### 2023-08-18 2_SSL_NE/article/article_supplementary.tex
added edits

@@
 \input{2_SSL_NE/article/auxiliary/preamble/2_preamble_layout_paperdraft}
-
-% cross-referencing
-\input{2_SSL_NE/article/auxiliary/preamble/1_preamble_functionality_crossreference}
 
@@
 Due to these limitations, any recommendations on future improvements in LED technology based on these studies are fragmented at best.
-Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2020novel,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
+Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2021spillovers,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
 
@@
 Dashed lines are projections made in 2015.
-Sources: (Monocrystal 2020)(Yole Developpement 2011)(Yole Developpement 2015).
+Sources: \cite{monocrystal2020private}\cite{yole2011sapphire}\cite{yole2015sapphire}.
 Bottom panel: Prevalence of sapphire wafer size used in LED manufacturing in 2008-2020.
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
 \centering
@@
 
+\clearpage
 \subsection{Non-Phosphor-Related LED Innovations}
@@
 An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
-Used in combination with red-emitting sulfide phosphors, by 2003 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2003}.
+Used in combination with red-emitting sulfide phosphors, by 2003 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}.
 However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased.
@@
 Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum.
-Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2003}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
+Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
 \label{fig:phosphor_spectrum}
@@
 Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
-Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel}.}
+Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel_details}.}
 	\label{fig:costmodel_calibration}
@@
     \hline
-    Panel A1 (FVE) & \cite{nichia2001data}\cite{lumi2003data}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
+    Panel A1 (FVE) & \cite{nichia2001data}\cite{lumi2003lumi}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
     \hline
@@
     Panel B2 (LEE) & \cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012} \newline \cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010}
-\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2003}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
+\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
     \hline

### 2023-08-18 2_SSL_NE/article/article_main.tex
final edits

@@
 The first commercial white LED produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{bando1998development}.
-As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:innovations}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.
+As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.
 
 Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 4, we find that only a single innovation related to LED consumer experience improvements was initially intended exclusively for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
-All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:phosphors}).
+All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:spillovers}).
 This means that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
@@
 Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4.
-Desig.
-= Designation, Ref.
-- Reference.}
+Desig.=Designation, Ref.=Reference.}
     \label{tab:phosphors}
@@
 The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks  and relevant patents.
-In addition, two LED life cycle analyses (Scholand and Dillon 2012a)(Casamayor, Su, and Ren 2018) were used to validate the model structure and extract some of the necessary quantitative model inputs.
+In addition, two LED life cycle analyses \cite{Scholand2012}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
 These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.

### 2023-10-05 2_SSL_NE/article/article_main.tex
added data availability statement

@@
 
+\subsection{Data Availability}
+
+The manufacturing cost model and associated input data is available are the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410658) \cite{zenodo_costmodel}.
+The Python scripts that were used to compute the light quality metrics of \cref{fgr:consumer_experience} are available in the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410789 \cite{zenodo_lightquality}.
+
 \section*{Acknowledgements}

### 2024-02-10 2_SSL_NE/article/article_main.tex
revision based on comments

@@
 Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details).
-Dashed lines indicate a 3rd-order polynomial fit to the data trend.
+Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend.
 The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
@@
 
+\section*{Corresponding authors}
+Correspondence to Michael P. Weinold
+
 \section*{Acknowledgements}
@@
 
+\section*{Contributions}
+All authors designed the study.
+M.P.W.
+and S.K.
+carried out the main experiments under the supervision of L.D.A.
+The data was analysed by M.P.W.
+All authors contributed to the discussion of the results and commented on the manuscript.
+The draft was written by M.P.W.
+and S.K.
+
+\section*{Ethocs Declarations}
+The authors declare no competing interests.
+
 \newpage

### 2024-03-05 2_SSL_NE/article/article_supplementary.tex
corrections for reviewer 1

@@
 \begin{equation}
-    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor{yellow}{Y},\textcolor{teal}{G}} E_i}
+    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor[RGB]{225, 180, 0}{Y},\textcolor{teal}{G}} E_i}
 \end{equation}
@@
 Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
-\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor{yellow}{Y}ellow, \textcolor{teal}{G}reen.
+\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor[RGB]{225, 180, 0}{Y}ellow, \textcolor{teal}{G}reen.
 
@@
     \caption{List of LED manufacturing cost model parameters considered in sensitivity analysis}
-    \vspace{5mm}
-    \begin{NiceTabularX}{\textwidth}{ |X|X|l|l|l|l|l|l|X|}
+    \begin{NiceTabularX}{1.1\textwidth}{ |X|X|l|l|l|l|l|l|X|}
         \hline
@@
         \hline
-            Person-hour & FTE & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & I \\
+            Labor Rate & \$/h & 30 & +50,-50 & 30 & +50,-50 & 30 & +50,-50 & I \\
         \hline
@@
         \hline
-            Overall Yield & \% & 100\% & +25,-25 & 100\% & +25,-15 & 100\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
+            Overall Yield & \% & 31\% & +25,-25 & 57\% & +25,-15 & 74\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
         \hline
@@
         \hline
-            Overall Throughput & UPH or h$^{-1}$ & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & Datasheets \\
+            Overall Throughput & UPH or h$^{-1}$ & varies & +50,-50 & varies & +50,-50 & varies & +50,-50 & Equipment Datasheets \\
         \hline
@@
         \hline
-            Metal Prices & \text{USD}/kg & 100\% & +50,-0 & 100\% & +50,-0 & 100\% & +50,-0 & Datasheets \\
-        \hline
-            Electricity Price & \text{USD}/kWh & 100\% & +50,-50 & 100\% & +50,-50 & 100\% & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
+            Metal Prices & \text{USD}/kg & varies & +50,-0 & varies & +50,-0 & varies & +50,-0 & Datasheets \\
+        \hline
+            Electricity Price & \text{USD}/kWh & 0.07 & +50,-50 & 0.08 & +50,-50 & 0.06 & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
         \hline
@@
         \hline
-            Phosphor Prices & \text{USD}/g & 100\% & +50,-50 & 150 & +0,-0 & 200 & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
+            Phosphor Prices & \text{USD}/g & varies & +50,-50 & varies & +0,-0 & varies & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
         \hline
@@
     \caption*{The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.
-Note: The units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
-If values in columns \textit{2003}-\textit{2020} are instead given in \%, this indicates that the parameters were varied by a set percentage from their respective model baselines.
-Abbreviations: FTE - full-time equivalent; UPH - units per hour; Equip.
+Note: Baseline values are provided only for \textit{global} model parameters in the columns \textit{2003, 2012, 2020}.
+For \textit{non-global} model parameters (those which vary by equipment) only the variation, in percent of the individual baseline values is provided.
+The corresponding units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
+The variation employed during the sensitivity analysis is provided in the columns $\pm [\%]$.
+Abbreviations: UPH - units per hour; Equip.
 Discount - equipment discount (sales rebate for large purchases); Inspec.
-Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow).
+Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow); Saph.
+Subst.
+- Sapphire Substrate.
 ’I’ in the Source column indicates expert interviews as the source of information.}

### 2024-03-13 2_SSL_NE/article/article_main.tex
corrections for reviewer 1
updated figures and captions

@@
   \includegraphics[height=6.5cm]{figures/cost_lamp_small.pdf}
-  \caption{\textbf{Historical development of retail sales prices (in 2020 USD) per luminous flux of LED-based luminaires, 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights.
+  \caption{\textbf{Historical development of retail sales prices per luminous flux for LED-based luminaires in the period 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights.
 Red curved and dashed lines represent average retail sales prices and price projections for LED based luminaires published by the U.S.
@@
 
-\subsection{Evolution of LED Chip Architectures}
+% should be integrated into some other sub-heading
+% \subsection{Evolution of LED Chip Architectures}
 
@@
 \cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
-The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in \cref{sec:discussion}.
-Through the index decomposition analysis described in \cref{subsec:performance_metrics}, we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to innovations that involved technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
-
-\cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
 The full list of identified LED innovations considered in our study is provided in Table SI2 in Section 3 of the Supplementary Information document.
@@
  \captionsetup{font=footnotesize}
- \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020} (top to bottom panels).
+ \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020}.
 Losses are normalized to 1 Watt of electric power input (yellow bar on the left).
@@
 It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs.
-The first commercial white LED produced by Nichia in 1996 used a YAG phosphor that could generate only cool white light \cite{bando1998development}.
+The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor that could generate only cool white light \cite{bando1998development}.
 As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.
@@
 
-Our findings are further supported by a preliminary sensitivity analysis, presented in Section 2 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
+Our findings are further supported by a preliminary sensitivity analysis, presented in Section 5 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
 
@@
     \caption{\textbf{Technology spillovers involved in white LED technology innovations identified in this study.}}
-    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|X|l|X|}
+    \begin{NiceTabularX}{1.1\textwidth}{|l|l|l|X|X|X|X|X|X|}
     \hline
-        \textbf{Disc.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
-        1926 & 1994 & 1996 & LED phosphors & Use of phosphors for light down conversion in LEDs & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
-        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
-        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Chemistry (S), Materials science (S) &\cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
-        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) &\cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview})\\ \hline
-        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview}) \\ \hline
-        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
-        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo,simmonsfinal,ledprof_nexxusqd,bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
-        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & Chemistry (S), Materials science (S) &\cite{paulusz1973efficient,radkov2009red,Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
-        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
+        \textbf{Disc.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Enabler} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
+        1926 & 1994 & 1996 & LED phosphors &  Use of phosphors for light down conversion in LEDs & ???
+& Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
+        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & ???
+& Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
+        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & ???
+&  Chemistry (S), Materials science (S) & \cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
+        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & targeted search & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T)  & \cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview})\\ \hline
+        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & targeted search & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview}) \\ \hline
+        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & research conference & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
+        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & ???
+& Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo,simmonsfinal,ledprof_nexxusqd,bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
+        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & ???
+& Chemistry (S), Materials science (S) &\cite{paulusz1973efficient,radkov2009red,Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
+        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & ???
+& Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
     \end{NiceTabularX}
@@
 where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
+In practice, this mean can be approximated, simplifying calculations.
 These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:
@@
 The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
-For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the Internet Archive , and expert interviews.
+For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the Internet Archive, and expert interviews.
 Because the architecture of LED chips has changed significantly since the introduction of the first commercial white LED devices in 1996, three different chip architectures are considered in the model: classical chips, flip chips, and chip-scale package flip chips (see Figure 3 and Supplementary Figures 2-6 for the details of each architecture).
 The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
-In addition, two LED life cycle analyses \cite{Scholand2012}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
+In addition, two LED life-cycle analyses \cite{Scholand2012}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
 These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.

### 2024-03-13 2_SSL_NE/article/article_supplementary.tex
updated figures and captions

@@
 % custom section designations
-\makeatletter
-\renewcommand{\@seccntformat}[1]{Supplementary Note \csname the#1\endcsname: } % https://tex.stackexchange.com/a/693618
-\makeatother
+% https://tex.stackexchange.com/a/712523
+\titleformat{\section}[display]{\large\bfseries\filright}{\bfseries Supplementary Note \thesection}{0pt}{}
+\crefname{section}{Supplementary Note}{Supplementary Notes}
 % custom equation designations
@@
 \clearpage
+
 \section{Previous Literature on Techological Progress in LEDs}
@@
 
-\section{Metrics for Technological Progress in LEDs}
-
-This Supplementary Note provides additional background and details for the physical, economic and consumer experience metrics that were chosen to quantify multi-dimensional progress in light-emitting diode technology.
-
-\subsection{Overview}
+\section{Metrics: Technological Progress in LEDs}
 
@@
 
-\subsection{Device Performance Metrics}
+\section{Metrics: Device Performance}
 \label{sec:device_performance_metrics}
@@
 
-\subsubsection{Forward Voltage Efficiency $\eta_{V_f}$}
+\textbf{Forward Voltage Efficiency $\eta_{V_f}$}
 
@@
 
-\subsubsection{Light Extraction Efficiency $\eta_{LE}$}
+\textbf{Light Extraction Efficiency $\eta_{LE}$}
 
@@
 
-\subsubsection{Internal Quantum Efficiency $\eta_{IQ}$}
+\textbf{Internal Quantum Efficiency $\eta_{IQ}$}
 
@@
 
-\subsubsection{Droop $\eta_{droop}$}
+\textbf{Droop $\eta_{droop}$}
 
@@
 
-\subsubsection{Light Conversion Efficiency $\eta_{C}$}
+\textbf{Light Conversion Efficiency $\eta_{C}$}
 
@@
 
-\subsubsection{Spectral Efficiency $\eta_{S}$}
+\textbf{Spectral Efficiency $\eta_{S}$}
 
@@
 
-\subsubsection{Luminous Efficacy of Radiation}
+\textbf{Luminous Efficacy of Radiation}
 \label{subsec:ler}
@@
 
-\subsubsection{Luminous Efficacy of Source}
+\textbf{Luminous Efficacy of Source}
 \label{subsec:les}
@@
 
-\subsection{Consumer Experience Metrics}
+\clearpage
+\section{Metrics: Consumer Experience}
 
@@
 
-\subsubsection{Color Rendering Index (CRI)}
+\textbf{Color Rendering Index (CRI)}
 
@@
 
-\subsubsection{Colour Temperature}
+\textbf{Colour Temperature}
 
@@
 
-\subsection{Cost Metrics}
+\textbf{Cost Metrics}
 
@@
 \clearpage
-\section{Methodological Details}
-\label{sec:method_details}
-
-This Supplementary Note provides additional details for the main methods and data sources used in our mixed-methods research approach described in the Methods section in the main text.
-
-\subsection{Background of Interviewed Experts}
+\section{Methods: Background of Interviewed Experts}
 
@@
     \caption{Anonymized list of interviewed LED experts.}
-    \vspace{5mm}
     \begin{tabular}{|l|l|l|l|l|}
@@
 
-\subsection{Manufacturing Cost Model Details}
-\label{sec:costmodel_details}
-
-\subsubsection{Structure of the Model}
+\clearpage
+\section{Methods: Cost Model Structure}
+\label{sec:costmodel_structure}
 
@@
 
-The LED cost model developed for this publication is a microeconomic manufacturing cost model.
+The LED cost model developed for this publication is a bottom-up microeconomic manufacturing cost model with process-step resolution.
 Within the timeframe and scope laid out in the main text of this publication, it returns the total manufacturing cost of GaN-based phosphor-converted warm white light-emitting diode packages.
-The model considers main economic factors associated with operating and maintaining manufacturing equipment.
-It does not consider costs associated with research and development or those associated with the construction of manufacturing facilities.
+The model assumes production at a \textit{"virtual"} North American factory which employs the highest performing equipment which could be identified.
+Performance here includes throughput, associated yield and resource consumption.
+It assumes North American utility prices as well as labour rates, which have remained constant over the time period of interest.
+The equipment used to model the manufacturing processes has been chosen from the catalogues of current market leaders in semiconductor equipment manufacturing.
+This choice was made based on the specialization of the equipment, as some manufacturers offer product lines aimed specifically at light-emitting diode manufacturing.
+When more than one type of manufacturing equipment was available, the higher volume variant was chosen.
+The model does not consider costs associated with research and development or those associated with the construction of manufacturing facilities.
 It considers market trends through their effect on manufacturing parameters.
@@
 
-\subsubsection{Manufacturing Process Steps by Chip Architecture}
-\label{subsec:manufacturing_steps}
+To provide some context on the scope of our model, a selection of similar efforts is listed in \cref{tab:costmodels}:
+
+\begin{table}[h!]
+    \caption{\textbf{Recent manufacturing cost models for clean energy technologies.}}
+    \begin{NiceTabularX}{\textwidth}{ |l|l|X|l|l|l|l|l|l|l| }
+    \CodeBefore
+       \rectanglecolor{blue!10}{13-1}{14-10}
+    \Body
+    \hline
+        \multirow{2}{*}{\textit{Year}} & \multirow{2}{*}{\textit{Tech.}} & \multirow{2}{*}{\textit{Purpose}} & \multirow{2}{*}{\textit{Period}} & \multirow{2}{*}{\textit{Base}} &  \multicolumn{4}{c|}{\textit{Parameters}} & \multirow{2}{*}{\textit{Source}} \\
+    \cline{6-9}
+         & & & & & \textit{M} & \textit{E} & \textit{G} & \textit{S} & \\
+    \hline
+        2018 & Solar & Cost reductions & 1980-2012 & steps & Y & N & Y & Y* & \cite{kavlak2018evaluating} \\
+    \hline
+        2013 & Solar & Geographic diff. in Cost & 2012 & Red. & Y & Y & Y & N & \cite{goodrich2013assessing} \\
+    \hline
+        2006 & Solar & Cost reductions & 1975-2001 & Red. & Y & N & Y & N & \cite{Nemetsolar} \\
+    \hline
+        2011 & IC & Cost reductions & 2011 & Steps & N & N & Y & Y & \cite{Gopalakrishnan2011} \\
+    \hline
+        2017 & Li-Ion & Technology comparison & 2017 & Steps & (Y) & (Y) & Y & Y & \cite{Ciez2017} \\
+    \hline
+        2015 & Li-Ion & Industry & 2015 & Steps & N & N & Y & Y &  \cite{Petri2015} \\
+    \hline
+        2012 & Li-Ion & Manufacturers & 2012-now & Steps & (Y) & N & Y & Y & \cite{Nelson2012} \\
+    \hline
+        2020 & Wind & Impact of breakthroughs & 2005-2017 & Steps & Y & Y & Y & Y & \cite{Elia2020} \\
+    \hline
+        2010 & Wind & Future cost savings & 2010 & Steps & N & N & Y & Y & \cite{Schubel2010} \\
+    \hline
+        2006 & Wind & Design changes & 1999-2006 & Ops. & N & N & Y & N & \cite{Fingersh2006} \\
+    \hline
+        2012 & LED & Cost reductions & 2012 & Steps & N & N & Y & Y & \cite{ledcomv2} \\
+    \hline
+        2020 & LED & Impact of breakthroughs & 2002-now & Steps & (Y) & (Y) & Y & Y & $^\text{this}_\text{work}$\\
+    \hline
+\end{NiceTabularX}
+\vspace{3mm}
+\caption*{The \textit{Base} column describes the base unit for computations.
+This can either be manufacturing steps ("Steps"), cost reductions ("Red.") or total cost of operation ("Ops.").
+The \textit{Parameters} column describes the parameters considered.
+Market parameters include market size, manufacturing equipment discount, economies of scale, etc.
+Economy parameters include taxes, subsidies, debt, inflation, etc.
+"Global" parameters include resource cost, labour cost, overall yield, etc.
+"Step" parameters include equipment performance, equipment cost, step yield etc.
+Models self-described as \textit{"bottom-up"} consider parameters $G,S$.
+Note that those models only capture the manufacturing situation a certain point in time.
+The exception to this are the model \cite{Elia2020} and the present model.
+Abbreviations: Solar Photovoltaics, excluding perovskite cells (Solar), integrated circuits (IC), base level for calculations (Base), Parameters: market (M), economy (E), "global" (G), "step" (S).
+Parameters not considered directly but instead through their impact on other parameters denoted (Y).}
+\label{tab:costmodels}
+\end{table}
+
+While generally comprehensive, the model is limited by the availability of performance data for the most recent manufacturing equipment, as well as the confidential nature of cost structures of Asia-based manufacturing lines.
+A complete treatment of the limitations of the model is provided in \cref{sec:costmodel_limitations}.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.85\textwidth]{./figures/SEMI_pyramid.pdf}
+    \caption{\textit{SEMI Standard 35} pyramid showing variables that enter into the computation of manufacturing cost.
+Variables not captured by our model are greyed out.
+Source: Adapted from \cite{standardE35} and \cite{helin200618th}.}
+    \label{fig:semi_pyramid}
+    %\vspace{-10pt}
+\end{figure}
+
+\clearpage
+\section{Methods: Cost Model Limitations}
+\label{sec:costmodel_limitations}
+
+\cref{sec:costmodel_structure} provided additional information on the structure of the cost model.
+As illustrated in \cref{tab:costmodels}, our model is comprehensive in its coverage of market, economic, manufacturing step and global manufacturing parameters.
+However, the model has limitations which must be considered when interpreting the results.
+
+\textbf{Availability of Equipment Performance Data}
+
+The data for the inputs used in the model were obtained through a combination of scientific literature, conference proceedings, industry reports, industry periodicals, manufacturing equipment datasheets and experts interviews.
+The model was therefore set up to consider four light-emitting diode chip architectures: A classical light-emitting diode chip architecture with lateral current spreading, two diﬀerent high-power thin-ﬁlm ﬂip-chip architectures and a chip-scale package thin-ﬁlm ﬂip chip architecture, all shown in \cref{sec:manufacturing_steps}.
+
+Unfortunately, despite the multitude of data sources, important performance parameters of  manufacturing equipment relevant to the latest generation of chips, such as those shown in \cref{fig:manuf_vtf_2012-1}-\cref{fig:manuf_csp_2020-2}, were unavailable due to their highly proprietary nature.
+Since full coverage of manufacturing equipment performance parameters could therefore not be assured for all chip architectures, costs were computed only for the aforementioned classical chip architecture.
+
+\textbf{Lack of Data for Calibration}
+
+While the model is setup up to capture the cost structure of a \textit{"virtual"} North American factory, the majority of production capacity is presently located in Asia.
+There, the share of domestically produced manufacturing equipment has increased from 10\% in 2008 to 65\% in 2015 \cite{tseng2015fab}.
+Recent investments into factories for high-power light-emitting diodes include Osram's Kulim plant in Malaysia \cite{osram2020kulim} and Lumileds' Penang plants in Malaysia \cite{lumileds2020penang}.
+In the production of low-power light-emitting diodes, the world's largest producer of LED packages by volume, MLS Co.
+operates plants in mainland China \cite{mls2020dev}.
+
+This means that it is difficult to validate model outputs directly with the help of industry data, which is provided only for Asia-based manufacturing lines.
+We compare cost model data to a selection of industry data in \cref{sec:cost_model_calibration} and provide an interpretation of the differing cost structure.
+Generally, we found that our model provides the best-case scenario for LED manufacturing, where the most performant (highest throughput, highest yield) equipment is used in each year.
+Interviews revealed that this is rarely the case in practice.
+One reason for this is economic pressure on western manufacturers after the introduction of significant Chinese subsidies as part of the country's Twelfth Five-Year Plan \cite{sanderson2014light}.
+
+However, because the purpose of our model is to show the impact of changes in the design and manufacturing process of devices, the geographic location and associated potentially lower overhead costs or slower adoption of specific equipment are of secondary importance to our model.
+This is why a US based manufacturing environment was assumed.
+
+\section{Manufacturing Process Steps by Chip Architecture}
+\label{sec:manufacturing_steps}
 
@@
     \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.}
-    \vspace{5mm}
     \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
@@
 \newpage
-\section{Cost Model Sensitivity Analysis}
+\section{Cost Model: Sensitivity Analysis}
 
@@
     \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
-	\caption{Sensitivity analysis for selected cost model parameters.
-Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in Supplementary Table 3.
+	\caption{\textbf{Sensitivity analysis for selected cost model parameters.} Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in Supplementary Table 3.
 Panel b shows the resulting range of variation in the manufacturing cost.
@@
 \begin{table}[]
+\centering
 \small
-    \caption{List of LED manufacturing cost model parameters considered in sensitivity analysis}
+    \caption{\textbf{List of cost model parameters considered in sensitivity analysis.}}
     \begin{NiceTabularX}{1.1\textwidth}{ |X|X|l|l|l|l|l|l|X|}
@@
         \hline
-            Edge Exclusion & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
+            Edge+Street Width & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
         \hline
@@
         \end{NiceTabularX}
+    \vspace{2mm}
     \caption*{The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.
@@
 \clearpage
-\section{Cost Model Comparison with Industry Data and DOE Projections}
-
-We conduct a comparison of the outcomes of our LED manufacturing cost model regarding the evolution of the cost structure of low-to-mid-power classic chip GaN phosphor-converted white LED with previously reported U.S.
-DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions (see \cref{fig:costmodel_calibration}).
+\section{Cost Model: Comparison with Industry Data/Projections}
+\label{sec:cost_model_calibration}
+
+In order to validate the outputs of the cost model regarding the evolution of the cost structure of low-to-mid-power classic chip GaN phosphor-converted white LED, we conducted a comparison with previously reported U.S.
+DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions.
+This comparison is shown in \cref{fig:costmodel_calibration}.
+
 We note some differences between the results of our model and the cost structure reported or projected by the DOE.
@@
 \clearpage
-\section{List of Sources for Figures}
+\section{Full List of Sources for Figures}
 \label{sec:sources}
-
-\subsection{Figure 1 (Historical Progress in Luminous Efficacy of Lighting Technologies)}
 
@@
     \end{NiceTabularX}
-    \caption*{Note: CFL - Compact Fluorescent Lamp. HID - High-Intensity Discharge.}
+    \vspace{2mm}
+    \caption*{Abbreviations: CFL - Compact Fluorescent Lamp, HID - High-Intensity Discharge}
 \end{table}
-
-\subsection{Figure 2 (Historical Development of LED Lamp Prices)}
-
-The sources for Figure 2 in the main article are grouped by the provider of data:
 
@@
 
-\subsection{Figure 3 (Historical Evolution of LED Chip Architectures)}
-
 \begin{table}[h!]
@@
 
-\subsection{\cref{fgr:subeff} (Historical Developments in LED Sub-Efficiencies)}
-
 \begin{table}[h!]
@@
 
-\printbibliography
+\printbibliography[title=Supplementary References]
 

### 2024-03-13 2_SSL_NE/article/article_main.tex
removed some references

@@
 \cite{azevedo2009transition}.
-See Supplementary Note 7 for the full list of sources and references.}
+See Supplementary Note 16 for the full list of sources and references.}
  \label{fgr:history_efficacy}
@@
 Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications.
-See Supplementary Note 7 for the full list of sources and references.}
+See Supplementary Note 16 for the full list of sources and references.}
   \label{fgr:cost_lamp_small}
@@
 For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
-LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
+LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
 
@@
 Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Note 7 for the full list of sources and references.}
+See Supplementary Note 16 for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
@@
 Horizontal red arrow indicates desirable higher values of colour rendering index (CRI).
-LED products based on the following phosphor innovations from the following LED manufacturers are represented : YAG - Nichia, 1998 \cite{bando1998development}; YGAG - Lumileds, 2002 \cite{Mueller2002}; 258 - Lumileds, 2005 \cite{MuellerMach2005}; SLA - Lumileds, 2014 \cite{Pust2014}; PFS - GE, 2015 \cite{Murphy2015}; QD - Lumileds, 2017 \cite{lumileds2016qd}, Osram, 2019 \cite{osram2019qd}; SALON - Osram, 2019 \cite{Hoerder2019}.}
+See Supplementary Note 16 for the full list of sources and references of all phosphor represented.}
  \label{fgr:consumer_experience}
@@
 Specifically, we use the additive logarithmic mean Divisia index method I (LMDI-I), also known as the Additive Sato-Vartia indicator \cite{deBoer2019}.
-It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics \cite{Diewert1988}, and subsequently refined.
+It was developed by Boyd in 1987 \cite{Boyd1987} on the basis of Divisia Index, a method in statistical economics, and subsequently refined.
 
@@
 Where gaps in relevant data existed, we filled them with our own performance calculations for individual sub-efficiencies where possible.
-Specifically, sub-efficiencies related to the emission spectrum of phosphor down-converted LED devices (i.e., spectral efficiency and light conversion efficiency)  were computed from the spectral data, often reported in LED device specifications, using the \texttt{colour-science} package for Python \cite{colour-science_software}.
+Specifically, sub-efficiencies related to the emission spectrum of phosphor down-converted LED devices (i.e., spectral efficiency and light conversion efficiency)  were computed from the spectral data, often reported in LED device specifications, using the \texttt{colour-science} package for Python.
 We also used the same approach to calculate CRI and luminous efficacy of radiation of LED devices (see Supplementary Note 4 and Supplementary Figure 10 for the associated spectral data and calculation results).
@@
 The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
-In addition, two LED life-cycle analyses \cite{Scholand2012}\cite{casamayor2018comparative} were used to validate the model structure and extract some of the necessary quantitative model inputs.
+In addition, two LED life-cycle analyses were used to validate the model structure and extract some of the necessary quantitative model inputs.
 These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
@@
 
-The manufacturing cost model and associated input data is available are the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410658) \cite{zenodo_costmodel}.
-The Python scripts that were used to compute the light quality metrics of \cref{fgr:consumer_experience} are available in the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410789 \cite{zenodo_lightquality}.
+The manufacturing cost model and associated input data is available are the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410658.
+The Python scripts that were used to compute the light quality metrics of \cref{fgr:consumer_experience} are available in the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410789.
 
@@
 
-\section*{Ethocs Declarations}
+\section*{Ethics Declarations}
 The authors declare no competing interests.

### 2024-03-13 2_SSL_NE/article/article_supplementary.tex
removed some references

@@
 
+\cite{Scholand2012}\cite{casamayor2018comparative} some more text on model inputs
+
 \begin{figure}[h!]
@@
 
-\section{Manufacturing Process Steps by Chip Architecture}
+\section{Diagrams: Chip Architecture Manufacturing Process Steps}
 \label{sec:manufacturing_steps}
@@
 
-\subsubsection{Computation of Manufacturing Cost}
-\label{subsec:cost_computation}
+\section{Cost Model: Computation of Yielded Cost}
+\label{sec:cost_computation}
 
@@
 
-\subsubsection{Computation of Yielded Cost}
+\textbf{Computation of Yielded Cost}
 
@@
 
-\subsubsection{Input Data on Sapphire Wafers}
+\textbf{Input Data on Sapphire Wafers}
 
@@
 
-\subsubsection{Computation of Contribution of Individual Variables}
+\section{Cost Model: Contribution of Variables}
 \label{sec:contribution_variables}
@@
 
+\newpage
+\section{Cost Model: Sensitivity Analysis}
+
+\begin{figure}[ht!]
+	\centering
+    \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
+	\caption{\textbf{Sensitivity analysis for selected cost model parameters.} Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in Supplementary Table 3.
+Panel b shows the resulting range of variation in the manufacturing cost.
+Ranges are represented for three years considered in the cost model, where corresponding data is available, by whiskers of the following colour: 2003 – blue, 2012 – green, 2020 – red.
+Note the overall trend of decreasing sensitivity to cost model parameters as the number of die per wafer increases over time.
+Abbreviations: resp.
+- respective; Yield Sav.
+from Insp.
+- yield savings from inspection.}
+	\label{fig:sensitivity}
+\end{figure}
+
+A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
+The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
+As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen based on the prices quotes from the United State Geological Survey price database \cite{usgeoprices}.
+Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
+The upper range was chosen because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
+The results of the analysis are shown in  \cref{fig:sensitivity}.
+The cost model is generally more sensitive to the variation in parameters at smaller wafer diameters.
+The most sensitive parameters are global parameters, such as yield or average equipment throughput.
+
+\begin{table}[]
+\centering
+\small
+    \caption{\textbf{List of cost model parameters considered in sensitivity analysis.}}
+    \begin{NiceTabularX}{1.1\textwidth}{ |X|X|l|l|l|l|l|l|X|}
+        \hline
+            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & Source \\
+        \hline
+            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & \cite{mddi1997cleanroom}\cite{ledcomv2} \newline \cite{bakshi2009euv}\cite{gajera2006process} \\
+        \hline
+            Labor Rate & \$/h & 30 & +50,-50 & 30 & +50,-50 & 30 & +50,-50 & I \\
+        \hline
+            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I \newline \cite{Appleyard_2001} \\
+        \hline
+            Overall Yield & \% & 31\% & +25,-25 & 57\% & +25,-15 & 74\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
+        \hline
+            Inspec. Yield Savings & \%/inspec. & 0.5\% & +80,-80 & 0.5\% & +80,-80 & 0.5\% & +80,-80 & \cite{mckinseyyield} \\
+        \hline
+            Overall Throughput & UPH or h$^{-1}$ & varies & +50,-50 & varies & +50,-50 & varies & +50,-50 & Equipment Datasheets \\
+        \hline
+            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I and \cref{fig:wafers} \\
+        \hline
+            Edge+Street Width & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
+        \hline
+            Cutting Width & $\mu$m & 100 & +50,-25 & 75 & +33.3,-20 & 20 & +300,-50 & \cite{masaki2000division}\cite{ils2005width} \newline \cite{photonics2010width}\cite{discowidth} \\
+        \hline
+            Metal Prices & \text{USD}/kg & varies & +50,-0 & varies & +50,-0 & varies & +50,-0 & Datasheets \\
+        \hline
+            Electricity Price & \text{USD}/kWh & 0.07 & +50,-50 & 0.08 & +50,-50 & 0.06 & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
+        \hline
+            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & \cref{fig:wafers} \\
+        \hline
+            Phosphor Prices & \text{USD}/g & varies & +50,-50 & varies & +0,-0 & varies & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
+        \hline
+        \end{NiceTabularX}
+    \vspace{2mm}
+    \caption*{The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.
+Note: Baseline values are provided only for \textit{global} model parameters in the columns \textit{2003, 2012, 2020}.
+For \textit{non-global} model parameters (those which vary by equipment) only the variation, in percent of the individual baseline values is provided.
+The corresponding units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
+The variation employed during the sensitivity analysis is provided in the columns $\pm [\%]$.
+Abbreviations: UPH - units per hour; Equip.
+Discount - equipment discount (sales rebate for large purchases); Inspec.
+Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow); Saph.
+Subst.
+- Sapphire Substrate.
+’I’ in the Source column indicates expert interviews as the source of information.}
+    \label{tab:sensitivity}
+\end{table}
+
+\clearpage
+\section{Cost Model: Comparison with Industry Data/Projections}
+\label{sec:cost_model_calibration}
+
+In order to validate the outputs of the cost model regarding the evolution of the cost structure of low-to-mid-power classic chip GaN phosphor-converted white LED, we conducted a comparison with previously reported U.S.
+DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions.
+This comparison is shown in \cref{fig:costmodel_calibration}.
+
+We note some differences between the results of our model and the cost structure reported or projected by the DOE.
+For instance, the share of the epitaxy step is consistently larger in the DOE data.
+This can in part be explained by our model relying on state-of-the-art equipment at a virtual U.S.-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+In addition, the manufacturing lines of the majority of manufacturers are located in Asia.
+We also note that the share of the sapphire substrate price in the DOE data is much larger than in our model, which can in part be explained by the overestimation of the actual price of sapphire wafers in earlier projections made for 2015-2020.
+Finally, the relative importance of the packaging part of the manufacturing process is very similar in our model and in the DOE results in 2012.
+However, in the DOE projections it significantly decreases by 2020, while in our model it retains and even increases its share of the total cost.
+This trend has been independently confirmed by researchers and industry reports on wafer-level packaging \cite{Lee2011WPL,Xie2013,ledsmag2017WLP}, showing better performance of our cost model compared to earlier DOE model projections.
+
+\begin{figure}[h]
+	\centering
+    \includegraphics[width=\textwidth]{./figures/costmodel_calibration.pdf}
+	\caption{\textbf{Cost structure comparison between our cost model, industry data and DOE projections.} Top panel: manufacturing cost calculations (2008-2016) and projections (2018, 2020) previously published by the U.S.
+DOE based on industry data and LEDCOM model.
+Bottom panel: manufacturing cost structure produced by our cost model for 2002, 2012 and 2020.
+Hatched bars are DOE projections.
+Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
+Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel_details}.}
+	\label{fig:costmodel_calibration}
+\end{figure}
+
 \clearpage
@@
 
-\subsection{Historical Progress in Sub-Efficiencies}
+\textbf{Historical Progress in Sub-Efficiencies}
 
@@
 \clearpage
-\subsection{Non-Phosphor-Related LED Innovations}
+\textbf{Non-Phosphor-Related LED Innovations}
 
@@
 
-\subsection{Phosphor-Related LED Innovations}
+\section{Phosphor-Related LED Innovations}
 \label{sec:innovation_phosphor}
@@
 
-\subsubsection{YAG and YGAG Phosphors}
+\textbf{YAG and YGAG Phosphors}
 
@@
 
-\subsubsection{258, SLA and SALON phosphors}
+\textbf{258, SLA and SALON phosphors}
 
@@
 
-\subsubsection{PFS Phosphor}
+\textbf{PFS Phosphor}
 
@@
 
-\subsubsection{Quantum Dots for Light Down-Conversion}
+\textbf{Quantum Dots for Light Down-Conversion}
 
@@
 \clearpage
-\subsubsection{Spectral Data for Identified LED Phosphor Innovations}
+\section{Spectral Data for Phosphors}
 
@@
 
-\newpage
-\section{Cost Model: Sensitivity Analysis}
-
-\begin{figure}[ht!]
-	\centering
-    \includegraphics[width=\textwidth]{./figures/costmodel_sensitivity.pdf}
-	\caption{\textbf{Sensitivity analysis for selected cost model parameters.} Panel a represents the tested range of variation of parameters indicated on horizontal axis along with their respective measurement units, as represented in Supplementary Table 3.
-Panel b shows the resulting range of variation in the manufacturing cost.
-Ranges are represented for three years considered in the cost model, where corresponding data is available, by whiskers of the following colour: 2003 – blue, 2012 – green, 2020 – red.
-Note the overall trend of decreasing sensitivity to cost model parameters as the number of die per wafer increases over time.
-Abbreviations: resp.
-- respective; Yield Sav.
-from Insp.
-- yield savings from inspection.}
-	\label{fig:sensitivity}
-\end{figure}
-
-A sensitivity analysis of the cost model has been performed using parameter variations listed in table \cref{tab:sensitivity}.
-The amount of parameter variation was chosen to encompass the identified range of parameter values in different manufacturing setups for each considered year.
-As an example, the lower range in the variation for the price of metals used in the model (+50\%,-0\%) was chosen based on the prices quotes from the United State Geological Survey price database \cite{usgeoprices}.
-Industry metals are not sold below the price of the raw material and often the markup is small compared to the price of the material.
-The upper range was chosen because a survey of industrial metal suppliers for semiconductor manufacturing showed that the largest markup was below 50\%.
-The results of the analysis are shown in  \cref{fig:sensitivity}.
-The cost model is generally more sensitive to the variation in parameters at smaller wafer diameters.
-The most sensitive parameters are global parameters, such as yield or average equipment throughput.
-
-\begin{table}[]
-\centering
-\small
-    \caption{\textbf{List of cost model parameters considered in sensitivity analysis.}}
-    \begin{NiceTabularX}{1.1\textwidth}{ |X|X|l|l|l|l|l|l|X|}
-        \hline
-            \textit{Parameter} & \textit{Unit} & \textit{2003} & $\pm [\%]$ & \textit{2012} & $\pm [\%]$ & \textit{2020} & $\pm [\%]$ & Source \\
-        \hline
-            Cleanroom Cost & \text{USD}/m$^2$ & 3000 & +16,-16 & 3000 & +16,-16 & 3000 & +16,-16 & \cite{mddi1997cleanroom}\cite{ledcomv2} \newline \cite{bakshi2009euv}\cite{gajera2006process} \\
-        \hline
-            Labor Rate & \$/h & 30 & +50,-50 & 30 & +50,-50 & 30 & +50,-50 & I \\
-        \hline
-            Equip. Discount & \% of \text{USD} & 0\% & +25,-25 & 0\% & +25,-25 & 0\% & +25,-25 & I \newline \cite{Appleyard_2001} \\
-        \hline
-            Overall Yield & \% & 31\% & +25,-25 & 57\% & +25,-15 & 74\% & +25,-25 & I, \cite{lumi2012yield}\cite{ledsmag2012} \newline \cite{systemplus2015reverse}\cite{ledcomv2} \\
-        \hline
-            Inspec. Yield Savings & \%/inspec. & 0.5\% & +80,-80 & 0.5\% & +80,-80 & 0.5\% & +80,-80 & \cite{mckinseyyield} \\
-        \hline
-            Overall Throughput & UPH or h$^{-1}$ & varies & +50,-50 & varies & +50,-50 & varies & +50,-50 & Equipment Datasheets \\
-        \hline
-            Wafer Diameter & mm & 100 & +0,-50 & 150 & +33.3,-33.3 & 200 & +0,-25 & I and \cref{fig:wafers} \\
-        \hline
-            Edge+Street Width & mm & 7 & +0,-50 & 5 & +40,-0 & 5 & +40,-20 & \cite{ledsmagexclusion}\cite{rubiconexclusion} \newline \cite{xiamenexclusion}\cite{american2007annual} \\
-        \hline
-            Cutting Width & $\mu$m & 100 & +50,-25 & 75 & +33.3,-20 & 20 & +300,-50 & \cite{masaki2000division}\cite{ils2005width} \newline \cite{photonics2010width}\cite{discowidth} \\
-        \hline
-            Metal Prices & \text{USD}/kg & varies & +50,-0 & varies & +50,-0 & varies & +50,-0 & Datasheets \\
-        \hline
-            Electricity Price & \text{USD}/kWh & 0.07 & +50,-50 & 0.08 & +50,-50 & 0.06 & +50,-50 & \cite{eia2000electric}\cite{eia2019electric} \\
-        \hline
-            Saph. Subst. Price & \text{USD} & 40 & +12.5,-12.5 & 10 & +100,-20 & 3 & +66.6,-33.3 & \cref{fig:wafers} \\
-        \hline
-            Phosphor Prices & \text{USD}/g & varies & +50,-50 & varies & +0,-0 & varies & +0,-0 & I, \cite{yole_phosphor_2012}\cite{yole2017phosphor} \\
-        \hline
-        \end{NiceTabularX}
-    \vspace{2mm}
-    \caption*{The results of the sensitivity analysis for the parameters in this table are presented in \cref{fig:sensitivity}.
-Note: Baseline values are provided only for \textit{global} model parameters in the columns \textit{2003, 2012, 2020}.
-For \textit{non-global} model parameters (those which vary by equipment) only the variation, in percent of the individual baseline values is provided.
-The corresponding units for the values in columns \textit{2003}-\textit{2020} are indicated in the column \textit{Units}.
-The variation employed during the sensitivity analysis is provided in the columns $\pm [\%]$.
-Abbreviations: UPH - units per hour; Equip.
-Discount - equipment discount (sales rebate for large purchases); Inspec.
-Yield Savings - yield savings from inspection (early detection and alleviation of issues in the manufacturing workflow); Saph.
-Subst.
-- Sapphire Substrate.
-’I’ in the Source column indicates expert interviews as the source of information.}
-    \label{tab:sensitivity}
-\end{table}
-
-\clearpage
-\section{Cost Model: Comparison with Industry Data/Projections}
-\label{sec:cost_model_calibration}
-
-In order to validate the outputs of the cost model regarding the evolution of the cost structure of low-to-mid-power classic chip GaN phosphor-converted white LED, we conducted a comparison with previously reported U.S.
-DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions.
-This comparison is shown in \cref{fig:costmodel_calibration}.
-
-We note some differences between the results of our model and the cost structure reported or projected by the DOE.
-For instance, the share of the epitaxy step is consistently larger in the DOE data.
-This can in part be explained by our model relying on state-of-the-art equipment at a virtual U.S.-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
-In addition, the manufacturing lines of the majority of manufacturers are located in Asia.
-We also note that the share of the sapphire substrate price in the DOE data is much larger than in our model, which can in part be explained by the overestimation of the actual price of sapphire wafers in earlier projections made for 2015-2020.
-Finally, the relative importance of the packaging part of the manufacturing process is very similar in our model and in the DOE results in 2012.
-However, in the DOE projections it significantly decreases by 2020, while in our model it retains and even increases its share of the total cost.
-This trend has been independently confirmed by researchers and industry reports on wafer-level packaging \cite{Lee2011WPL,Xie2013,ledsmag2017WLP}, showing better performance of our cost model compared to earlier DOE model projections.
-
-\begin{figure}[h]
-	\centering
-    \includegraphics[width=\textwidth]{./figures/costmodel_calibration.pdf}
-	\caption{\textbf{Cost structure comparison between our cost model, industry data and DOE projections.} Top panel: manufacturing cost calculations (2008-2016) and projections (2018, 2020) previously published by the U.S.
-DOE based on industry data and LEDCOM model.
-Bottom panel: manufacturing cost structure produced by our cost model for 2002, 2012 and 2020.
-Hatched bars are DOE projections.
-Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
-Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel_details}.}
-	\label{fig:costmodel_calibration}
-\end{figure}
-
 \clearpage
@@
 \begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
     \caption{List of sources for Figure 1 organized by technology.}
@@
 \begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
     \caption{List of sources for Figure 2 grouped by data provider.}
@@
 \begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
 \caption{List of sources for Figure 3 grouped by publication type.}
@@
 \begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+\caption{List of sources for Figure 6.}
+    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|}
+    \hline
+    \textit{Phosphor/QD} & \textit{Company} & \textit{YOI} & \textit{Reference} \\
+    \hline
+    YAG & Nichia & 1998 & \cite{bando1998development} \\
+    \hline
+    YGAG & Lumileds & 2002 &  \cite{Mueller2002} \\
+    \hline
+    258 & Lumileds & 2005 & \cite{MuellerMach2005} \\
+    \hline
+    SLA & Lumileds & 2014 & \cite{Pust2014} \\
+    \hline
+    PFS & GE & 2015 & \cite{Murphy2015} \\
+    \hline
+    QD & Lumileds & 2017 & \cite{lumileds2016qd} \\
+    \hline
+    Salon & Osram & 2019 & \cite{osram2019qd} \\
+    \hline
+    SALON & Osram & 2019 & \cite{Hoerder2019} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
     \caption{List of sources for \cref{fgr:subeff} grouped by LED sub-efficiencies.}

### 2024-03-13 2_SSL_NE/article/article_supplementary.tex
updated figures

@@
 \clearpage
-
-\section{Previous Literature on Techological Progress in LEDs}
+\section{Previous Literature on Technological Progress in LEDs}
 \label{sec:prev_lit}
@@
 
-\section{Metrics: Technological Progress in LEDs}
+\clearpage
+\section{Metrics for Technological Progress in LEDs}
+
+\textbf{Motivation for Methods Selection}
 
@@
 We select the metrics based on the following two general criteria.
-First, we focus only on metrics widely accepted and reported in industry because metrics proposed in the scientific literature but not reported by device manufacturers cannot be used to compare the performance of commercial LED devices over time.
+First, we focus only on metrics widely accepted and reported in industry.
+This is because metrics proposed in the scientific literature, but not reported by device manufacturers, cannot be used to compare the performance of commercial LED devices over time.
 Second, the chosen metrics must be useful for understanding the impact of individual technological improvements on relevant performance and cost characteristics.
@@
 Historically, the progress in LED technology has been commonly described by pointing to impressive improvements in LED device performance, more specifically device brightness, as well as electrical efficiency, and manufacturing cost reductions \cite{Taki2019}.
-However, metrics related to consumer experience from using white LEDs as a lighting source, such as the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000,Sandahl2006,CAIRD2008,murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
+However, metrics related to the quality of light, such as the perceived temperature of a white light source and its ability to faithfully render colours, have also played a significant part in the market adoption of new lighting technologies \cite{Menanteau2000,Sandahl2006,CAIRD2008,murphy2012governing} and received substantial attention in LED research and development efforts \cite{azevedo2009transition,cho2017white}.
 Therefore, a comprehensive analysis of the evolution in white LED technology must take into consideration advances in 1) physical device performance; 2) consumer experience; and 3) LED device manufacturing costs.
@@
 
+\textbf{Distinction between Metrics}
+
+We use a well-established distinction between metrics.
+Those metrics which are related to the spectrum of the light source only ("color quality metrics" \cite{DLC_SSL_Requirements}\cite{dilaura2011lighting}), we group under the term consumer experience metrics.
+Those metrics, which are determined by the number and energy of photons leaving the device (as a function of the number of electrons entering it), we group under the term device performance metrics.
+As such, our distinction is physical at the fundamental level.
+This follows existing literature on the technological progress in solid-state lighting \cite{tsao2010solid}.
+We describe these metrics in detail in \cref{sec:device_performance_metrics}-\cref{sec:metrics_consumer}.
+
+\textbf{"Haitz's Law"}
+
+A highly cited and frequently called-upon metric is \textit{flux/package}, as described in what has become known as Haitz's law.
+An observation rather than a law of nature, it describes increasing device performance and decreasing manufacturing cost.
+Originally introduced in 2000 \cite{haitz1999case}, it has been updated frequently since then.
+Unfortunately, the way Haitz's law is frequently used is  problematic due to the misleading conclusions it suggests.
+While the original version presented by Haitz displays the increasing performance of light-emitting diode chips, the graphs have been extended with datapoints from lamps which contain an unspecified number of chips.
+The metric therefore loses its value in describing technological progress at the chip-level.
+The continuing rise in performance inferred from this chart is disconnected from the actual, much lower, performance of single chips \cite{weinold2021compound}.
+This is shown in \cref{fig:haitz-org}-\cref{fig:haitz}.
+
+\begin{figure}[h!]
+    \includegraphics[width=\textwidth]{./figures/haitz_law_old_both.pdf}
+    \caption{Historical increase in flux for white light-emitting diode packages (\textit{"Haitz's Law"}), as it is often reproduced in literature.
+The original data by Haitz for red LEDs from 2000 is indicated for reference.
+Note the different range on the abscissa compared to \cref{fig:haitz}.
+Sources for red data: \cite{haitz1999case}\cite{haitz2011solid}.
+Sources for white-light data: figure \cref{fig:haitz}.}
+    \label{fig:haitz-org}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.85\textwidth]{./figures/haitz_law_white.pdf}
+    \caption{Historical increase in flux for white light-emitting diode chips and packages, inspired by \textit{"Haitz's Law"}.
+Note that the increase in flux per single chip is not increasing as starkly as the flux for multi-chip packages.
+Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
+\\ Sources for cost: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} \\ Sources for flux: \cite{haitz2007seoul}\cite{haitz2010osram}\cite{haitz2010lumi-1}\cite{haitz2010cree} \cite{haitz2011lumi-1}\cite{haitz2014nichia}\cite{haitz2015cree-1}\cite{haitz2015cree-2}\cite{haitz2018cree}\cite{haitz2019osram}\cite{haitz2019lumi-1}\cite{haitz2019lumi-1}\cite{haitz2020lumi-1} \cite{haitz2013cob}\cite{haitz2016cob}\cite{haitz2017cob}\cite{haitz2019cob}\cite{haitz2019lumi-1}\cite{haitz2019lumi-2}}
+    \label{fig:haitz}
+    \vspace{-20mm}
+\end{figure}
+
+\clearpage
 \section{Metrics: Device Performance}
@@
 \section{Metrics: Consumer Experience}
+\label{sec:metrics_consumer}
 

### 2024-03-15 2_SSL_NE/article/article_main.tex
various edits

@@
 
-These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}.
+These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent light bulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}.
 As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
@@
 
+\clearpage
 \section{Main}
@@
 % \subsection{Evolution of LED Chip Architectures}
-
-\clearpage
-\begin{figure}[h!]
-\centering
-\includegraphics[height=15cm]{figures/chip_architecture_overview.pdf}
-\caption{\textbf{Historical evolution of LED chip architectures.} Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
-Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
-The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip.
-Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity.
-Years indicated correspond to the earliest identified patent priority date.
-Black frames around certain designs indicate chip designs not brought to large-scale production.
-Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
-Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Note 16 for the full list of sources and references.}
-\label{fgr:chip_architecture_overview}
-\end{figure}
-
-We summarize the information collected and analysed on the progress in GaN-based LED chip architectures and manufacturing processes in \cref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
-We use this information as input into LED performance calculations and manufacturing cost modeling.
 
@@
  \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020}.
-Losses are normalized to 1 Watt of electric power input (yellow bar on the left).
-Sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2.
+Losses are normalized to 1W of electric power input.
+The sub-efficiencies corresponding to each loss channel are listed below each column and described in Supplementary Note 2.
 Numbers for each loss channel indicate energy losses both in relative terms of input power (in percent) at the point of the channel and absolute values (in Watts).
 Percentages for loss channels labeled by red, green and blue font indicate losses of corresponding remaining red/green/blue light energy.
-The following LED architectures and down-conversion phosphor spectra used in calculations in each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor.
+The following LED architectures and light-source spectra used in calculations for each considered year are shown for reference: 2003 – Flip-chip with YGAG phosphor; 2010 – Flip-chip with 258 phosphor; 2016 – Flip-chip with PFS phosphor; 2020 – Chip-scale package flip-chip with SALON phosphor.
+Note that the inset spectra clearly show progress in spectral efficiency.
+The red curve shows the wavelength-dependent sensitivity of the human eye, while the black curve shows the spectrum of the light source.
+When the spectrum has significant energy at the edges of the sensitivity curve, energy is lost.
 Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.
@@
 
+\clearpage
+\begin{figure}[h!]
+\centering
+\includegraphics[height=15cm]{figures/chip_architecture_overview.pdf}
+\caption{\textbf{Historical evolution of LED chip architectures.} Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
+Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
+The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip.
+Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity.
+Years indicated correspond to the earliest identified patent priority date.
+Black frames around certain designs indicate chip designs not brought to large-scale production.
+Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
+Source: adapted and compiled from multiple patents and industry publications.
+See Supplementary Note 16 for the full list of sources and references.}
+\label{fgr:chip_architecture_overview}
+\end{figure}
+
+We summarize the information collected and analysed on the progress in GaN-based LED chip architectures and manufacturing processes in \cref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+We use this information as input into LED performance calculations and manufacturing cost modeling.
+
 \subsection{Manufacturing Cost Model}

### 2024-03-15 2_SSL_NE/article/article_supplementary.tex
various edits

@@
 \begin{figure}[h!]
-    \includegraphics[width=\textwidth]{./figures/haitz_law_old_both.pdf}
-    \caption{Historical increase in flux for white light-emitting diode packages (\textit{"Haitz's Law"}), as it is often reproduced in literature.
+\vspace{-10mm}
+    \includegraphics[width=0.9\textwidth]{./figures/haitz_law_old_both.pdf}
+    \caption{\textit{"Haitz's Law"}, as it is often reproduced in literature.
 The original data by Haitz for red LEDs from 2000 is indicated for reference.
-Note the different range on the abscissa compared to \cref{fig:haitz}.
-Sources for red data: \cite{haitz1999case}\cite{haitz2011solid}.
-Sources for white-light data: figure \cref{fig:haitz}.}
+Note the different range on the abscissa compared to \cref{fig:haitz_new}.
+Sources: \cite{haitz1999case}\cite{haitz2011solid} and  \cref{fig:haitz_new}.}
     \label{fig:haitz-org}
@@
     \includegraphics[width=0.85\textwidth]{./figures/haitz_law_white.pdf}
-    \caption{Historical increase in flux for white light-emitting diode chips and packages, inspired by \textit{"Haitz's Law"}.
+    \caption{Adaption of \textit{"Haitz's Law"}.
 Note that the increase in flux per single chip is not increasing as starkly as the flux for multi-chip packages.
-Shown are datapoints for best commercial performers from press releases, datasheets and and industry periodicals.
-\\ Sources for cost: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} \\ Sources for flux: \cite{haitz2007seoul}\cite{haitz2010osram}\cite{haitz2010lumi-1}\cite{haitz2010cree} \cite{haitz2011lumi-1}\cite{haitz2014nichia}\cite{haitz2015cree-1}\cite{haitz2015cree-2}\cite{haitz2018cree}\cite{haitz2019osram}\cite{haitz2019lumi-1}\cite{haitz2019lumi-1}\cite{haitz2020lumi-1} \cite{haitz2013cob}\cite{haitz2016cob}\cite{haitz2017cob}\cite{haitz2019cob}\cite{haitz2019lumi-1}\cite{haitz2019lumi-2}}
-    \label{fig:haitz}
-    \vspace{-20mm}
-\end{figure}
-
-\clearpage
-\section{Metrics: Device Performance}
-\label{sec:device_performance_metrics}
-
-During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
-This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
-In 2000, the highest performing devices yielded around 10 lm, just below the output of a candle as defined in the unit candela (1 cd=12.57 lm) \cite{haitz2011solid}.
-Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
-Today, however, the devices with the highest performance yield in excess of 1600 lm, the equivalent of a 100 W incandescent bulb \cite{cree2020bright}.
-LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
-Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving the overall LED efficiency \cite{weinold2021compound}.
-Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
-
-We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe different physical loss channels within the device to describe progress in LED technology.
-In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
-The equations defining these sub-efficiencies are provided below.
-The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
-%
-\begin{equation}
-    \eta_L = \prod_{i=(V_f, \dots, S)} \eta_i
-\end{equation}
-%
-This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
-
-\textbf{Forward Voltage Efficiency $\eta_{V_f}$}
-
-Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
-\cite{tsao2010solid}} (FVE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
-These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
-It is defined as
-%
-\begin{equation}
-    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
-\end{equation}
-%
-where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Light Extraction Efficiency $\eta_{LE}$}
-
-LED light extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
-%
-\begin{equation}
-    \eta_{LE} = \frac{P_{opt}}{P_{int}} = \frac{\text{\# of photons out/s}}{\text{\# of photons created/s}}
-\end{equation}
-%
-where $P_{opt}$ is the optical power emitted from the device and  $P_{int}$ is the optical power emitted from the active region into other parts of the device (internal) \cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Internal Quantum Efficiency $\eta_{IQ}$}
-
-Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is defined as
-%
-\begin{equation}
-    \eta_{IQ} = \frac{\# \text{of photons created/s}}{\# \text{of electron-hole pairs in/s}}
-\end{equation}
-
-\textbf{Droop $\eta_{droop}$}
-
-Efficiency droop describes the decrease of LED internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
-This energy loss channel is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
-It is defined as
-%
-\begin{equation}
-    \eta_{droop} = 1 - \frac{\eta_{IQ}}{\eta_{IQ}(A \rightarrow 0)}
-\end{equation}
-%
-where $\eta_{IQ}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
-In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$
-%
-\begin{equation}
-\label{eqn:droop}
-    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
-\end{equation}
-%
-Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Light Conversion Efficiency $\eta_{C}$}
-
-Light conversion efficiency (LCE) describes the sum of all losses in the process of converting the blue light that is the basis for all phosphor-converted white LEDs.
-Specifically, energy is lost due to three distinct effects.
-The first effect is the inherent wavelength-conversion loss or \textit{Stokes shift} as the difference between the original blue wavelength and the down-converted red, yellow or green wavelength.
-The second effect are non-radiative processes in the phosphor material.
-The third , scattering and absorption processes in the phosphor material.
-These three effects are listed individually in Figure 4.
-The overall light conversion efficiency is defined as
-%
-\begin{equation}
-    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor[RGB]{225, 180, 0}{Y},\textcolor{teal}{G}} E_i}
-\end{equation}
-%
-where $E_{\textcolor{blue}{B}}$ is the total energy of blue light before down-conversion and $E_i$ denotes the total energy of light at color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
-Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
-\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor[RGB]{225, 180, 0}{Y}ellow, \textcolor{teal}{G}reen.
-
-\textbf{Spectral Efficiency $\eta_{S}$}
-
-Spectral efficiency (SE) describes losses in the light down conversion process due to the wavelength-dependent efficiency of the human eyes.
-Photons converted into the infrared or ultraviolet light are lost to illumination purposes.
-It is defined as
-%
-\begin{equation}
-    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
-\end{equation}
-%
-where $K$ is the luminous efficacy of radiation of the light source, defined in \cref{eqn:ler}, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
-$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering index performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Luminous Efficacy of Radiation}
-\label{subsec:ler}
-
-The luminous efficacy \textit{of radiation} $K$ is a supplementary metric used in our calculations of SE, rather than directly in our analysis of the improvements in LED performance.
-Care must be taken not to confuse this metric with the luminous efficacy of a light source, described in \cref{subsec:les}.
-
-The luminous efficacy \textit{of radiation} $K$ describes the match of the spectrum of a light-emitting diode to the human visual system.
-Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
-A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
-The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source $\phi$ with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
-%
-\begin{equation}
-\label{eqn:ler}
-    K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
-\end{equation}
-%
-where
-%
-\begin{align*}
-    K &\dots \text{luminosity function} \\
-    \phi &\dots \text{spectral radiant flux} \\
-    \lambda &\dots \text{wavelength}
-\end{align*}
-%
-This metric can be computed from spectral data alone and does not require additional spectral normalization.
-It enables straightforward comparison between the performance of different downconversion phosphors, as shown in the top panel of \cref{fig:phosphor_spectrum}.
-Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation.
-
-\textbf{Luminous Efficacy of Source}
-\label{subsec:les}
-
-The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux $\phi$ and the consumed electrical power $P_{el}$ \cite{cie-term-effsrc}.
-%
-\begin{equation}
-    \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
-\end{equation}
-%
-This metric is also frequently cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
-However, as noted in \cref{subsec:ler}, care must be taken not to confuse this efficacy metric with the \textit{luminous efficacy of radiation}, which depends only on the spectral characteristics of a light source.
-As the luminous efficacy of a light source $\eta$ captures the overall device efficacy, it depends on a large number of other device properties and parameters.
-This makes attribution of changes in this metric to individual changes in device design or manufacturing difficult.
-For this reason, we do not use this metric only to show historical progress across different lighting technologies in Figure 1 in the main text of the article.
-
-\clearpage
-\section{Metrics: Consumer Experience}
-\label{sec:metrics_consumer}
-
-The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
-Any metric relevant to customer experience can thus be calculated from the spectrum alone.
-The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra of the down-conversion phosphor used in the device.
-It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
-Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: colour rendering index (CRI) and colour temperature (CT).
-We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
-This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
-As a result, it is beyond the scope of this work.
-
-\textbf{Color Rendering Index (CRI)}
-
-The colour rendering index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
-The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
-CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
-However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DoE2016LED}.
-High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
-It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
-However, these niche applications remain outside our focus on general illumination.
-Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
-
-\textbf{Colour Temperature}
-
-The colour temperature (CT) of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
-Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
-Early white LEDs produced only cool white light \cite{mueller2000light}.
-The introduction of first commercial warm white LED light bulbs played a significant role in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
-For this reason, we also adopt CT as a metric for tracking progress in consumer experience in LED lighting.
-
-\textbf{Cost Metrics}
-
-In selecting the metrics for tracking progress in manufacturing cost reductions in LED, we must first highlight the complexity of this task.
-Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
-Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
-However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
-In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
-Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress difficult in practice.
-
-We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
-In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
-
-We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:costmodel_details}.
-The outcome of the cost model is the cumulative yielded manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
-
-\clearpage
-\section{Methods: Background of Interviewed Experts}
+Shown are datapoints for best commercial performers.
+See \cref{sec:sources} for the full list of sources and references.}
+    \label{fig:haitz_new}
+\end{figure}
+
+\section{Background of Interviewed Experts}
 
@@
 \clearpage
+\section{Metrics: Device Performance}
+\label{sec:device_performance_metrics}
+
+During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
+This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
+In 2000, the highest performing devices yielded around 10 lm, just below the output of a candle as defined in the unit candela (1 cd=12.57 lm) \cite{haitz2011solid}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
+Today, however, the devices with the highest performance yield in excess of 1600 lm, the equivalent of a 100 W incandescent bulb \cite{cree2020bright}.
+LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
+Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving the overall LED efficiency \cite{weinold2021compound}.
+Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
+
+We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe different physical loss channels within the device to describe progress in LED technology.
+In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
+The equations defining these sub-efficiencies are provided below.
+The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
+%
+\begin{equation}
+    \eta_L = \prod_{i=(V_f, \dots, S)} \eta_i
+\end{equation}
+%
+This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
+
+\textbf{Forward Voltage Efficiency $\eta_{V_f}$}
+
+Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
+\cite{tsao2010solid}} (FVE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
+These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
+It is defined as
+%
+\begin{equation}
+    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
+\end{equation}
+%
+where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Light Extraction Efficiency $\eta_{LE}$}
+
+LED light extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+%
+\begin{equation}
+    \eta_{LE} = \frac{P_{opt}}{P_{int}} = \frac{\text{\# of photons out/s}}{\text{\# of photons created/s}}
+\end{equation}
+%
+where $P_{opt}$ is the optical power emitted from the device and  $P_{int}$ is the optical power emitted from the active region into other parts of the device (internal) \cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Internal Quantum Efficiency $\eta_{IQ}$}
+
+Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is defined as
+%
+\begin{equation}
+    \eta_{IQ} = \frac{\# \text{of photons created/s}}{\# \text{of electron-hole pairs in/s}}
+\end{equation}
+
+\textbf{Droop $\eta_{droop}$}
+
+Efficiency droop describes the decrease of LED internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
+This energy loss channel is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
+It is defined as
+%
+\begin{equation}
+    \eta_{droop} = 1 - \frac{\eta_{IQ}}{\eta_{IQ}(A \rightarrow 0)}
+\end{equation}
+%
+where $\eta_{IQ}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
+In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$
+%
+\begin{equation}
+\label{eqn:droop}
+    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
+\end{equation}
+%
+Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Light Conversion Efficiency $\eta_{C}$}
+
+Light conversion efficiency (LCE) describes the sum of all losses in the process of converting the blue light that is the basis for all phosphor-converted white LEDs.
+Specifically, energy is lost due to three distinct effects.
+The first effect is the inherent wavelength-conversion loss or \textit{Stokes shift} as the difference between the original blue wavelength and the down-converted red, yellow or green wavelength.
+The second effect are non-radiative processes in the phosphor material.
+The third , scattering and absorption processes in the phosphor material.
+These three effects are listed individually in Figure 4.
+The overall light conversion efficiency is defined as
+%
+\begin{equation}
+    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor[RGB]{225, 180, 0}{Y},\textcolor{teal}{G}} E_i}
+\end{equation}
+%
+where $E_{\textcolor{blue}{B}}$ is the total energy of blue light before down-conversion and $E_i$ denotes the total energy of light at color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
+Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
+\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor[RGB]{225, 180, 0}{Y}ellow, \textcolor{teal}{G}reen.
+
+\textbf{Spectral Efficiency $\eta_{S}$}
+
+Spectral efficiency (SE) describes losses in the light down conversion process due to the wavelength-dependent efficiency of the human eyes.
+Photons converted into the infrared or ultraviolet light are lost to illumination purposes.
+It is defined as
+%
+\begin{equation}
+    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
+\end{equation}
+%
+where $K$ is the luminous efficacy of radiation of the light source, defined in \cref{eqn:ler}, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
+$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering index performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Luminous Efficacy of Radiation}
+\label{subsec:ler}
+
+The luminous efficacy \textit{of radiation} $K$ is a supplementary metric used in our calculations of SE, rather than directly in our analysis of the improvements in LED performance.
+Care must be taken not to confuse this metric with the luminous efficacy of a light source, described in \cref{subsec:les}.
+
+The luminous efficacy \textit{of radiation} $K$ describes the match of the spectrum of a light-emitting diode to the human visual system.
+Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
+A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
+The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source $\phi$ with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
+%
+\begin{equation}
+\label{eqn:ler}
+    K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
+\end{equation}
+%
+where
+%
+\begin{align*}
+    K &\dots \text{luminosity function} \\
+    \phi &\dots \text{spectral radiant flux} \\
+    \lambda &\dots \text{wavelength}
+\end{align*}
+%
+This metric can be computed from spectral data alone and does not require additional spectral normalization.
+It enables straightforward comparison between the performance of different downconversion phosphors, as shown in the top panel of \cref{fig:phosphor_spectrum}.
+Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation.
+
+\textbf{Luminous Efficacy of Source}
+\label{subsec:les}
+
+The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux $\phi$ and the consumed electrical power $P_{el}$ \cite{cie-term-effsrc}.
+%
+\begin{equation}
+    \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
+\end{equation}
+%
+This metric is also frequently cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
+However, as noted in \cref{subsec:ler}, care must be taken not to confuse this efficacy metric with the \textit{luminous efficacy of radiation}, which depends only on the spectral characteristics of a light source.
+As the luminous efficacy of a light source $\eta$ captures the overall device efficacy, it depends on a large number of other device properties and parameters.
+This makes attribution of changes in this metric to individual changes in device design or manufacturing difficult.
+For this reason, we do not use this metric only to show historical progress across different lighting technologies in Figure 1 in the main text of the article.
+
+\clearpage
+\section{Metrics: Consumer Experience}
+\label{sec:metrics_consumer}
+
+The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
+Any metric relevant to customer experience can thus be calculated from the spectrum alone.
+The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra of the down-conversion phosphor used in the device.
+It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
+Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: colour rendering index (CRI) and colour temperature (CT).
+We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
+This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
+As a result, it is beyond the scope of this work.
+
+\textbf{Color Rendering Index (CRI)}
+
+The colour rendering index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
+The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
+CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
+However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DoE2016LED}.
+High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
+It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
+However, these niche applications remain outside our focus on general illumination.
+Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
+
+\textbf{Colour Temperature}
+
+The colour temperature (CT) of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
+Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
+Early white LEDs produced only cool white light \cite{mueller2000light}.
+The introduction of first commercial warm white LED light bulbs played a significant role in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
+For this reason, we also adopt CT as a metric for tracking progress in consumer experience in LED lighting.
+
+\textbf{Cost Metrics}
+
+In selecting the metrics for tracking progress in manufacturing cost reductions in LED, we must first highlight the complexity of this task.
+Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
+Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
+However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
+In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
+Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress difficult in practice.
+
+We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
+In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
+
+We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:costmodel_details}.
+The outcome of the cost model is the cumulative yielded manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
+
+\clearpage
 \section{Methods: Cost Model Structure}
@@
 
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.95\textwidth]{./figures/costmodel.pdf}
-    \caption{\textbf{Cost model structure.} The diagram shows inputs to each step and computational steps leading to the cost model outputs.
-For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.}
-    \label{fig:costmodel-schematic}
-\end{figure}
-
 The LED cost model developed for this publication is a bottom-up microeconomic manufacturing cost model with process-step resolution.
 Within the timeframe and scope laid out in the main text of this publication, it returns the total manufacturing cost of GaN-based phosphor-converted warm white light-emitting diode packages.
-The model assumes production at a \textit{"virtual"} North American factory which employs the highest performing equipment which could be identified.
-Performance here includes throughput, associated yield and resource consumption.
+The model assumes production at a \textit{"virtual"} factory which employs the highest performing equipment available in each year.
+Performance in this context comprises throughput, associated yield and resource consumption.
 It assumes North American utility prices as well as labour rates, which have remained constant over the time period of interest.
+Compared to other parameters, these fixed costs have little impact on the overall cost, as shown in \cref{fig:sensitivity} and \cref{fig:costmodel_calibration}.
 The equipment used to model the manufacturing processes has been chosen from the catalogues of current market leaders in semiconductor equipment manufacturing.
@@
 
-To provide some context on the scope of our model, a selection of similar efforts is listed in \cref{tab:costmodels}:
-
-\begin{table}[h!]
-    \caption{\textbf{Recent manufacturing cost models for clean energy technologies.}}
-    \begin{NiceTabularX}{\textwidth}{ |l|l|X|l|l|l|l|l|l|l| }
-    \CodeBefore
-       \rectanglecolor{blue!10}{13-1}{14-10}
-    \Body
-    \hline
-        \multirow{2}{*}{\textit{Year}} & \multirow{2}{*}{\textit{Tech.}} & \multirow{2}{*}{\textit{Purpose}} & \multirow{2}{*}{\textit{Period}} & \multirow{2}{*}{\textit{Base}} &  \multicolumn{4}{c|}{\textit{Parameters}} & \multirow{2}{*}{\textit{Source}} \\
-    \cline{6-9}
-         & & & & & \textit{M} & \textit{E} & \textit{G} & \textit{S} & \\
-    \hline
-        2018 & Solar & Cost reductions & 1980-2012 & steps & Y & N & Y & Y* & \cite{kavlak2018evaluating} \\
-    \hline
-        2013 & Solar & Geographic diff. in Cost & 2012 & Red. & Y & Y & Y & N & \cite{goodrich2013assessing} \\
-    \hline
-        2006 & Solar & Cost reductions & 1975-2001 & Red. & Y & N & Y & N & \cite{Nemetsolar} \\
-    \hline
-        2011 & IC & Cost reductions & 2011 & Steps & N & N & Y & Y & \cite{Gopalakrishnan2011} \\
-    \hline
-        2017 & Li-Ion & Technology comparison & 2017 & Steps & (Y) & (Y) & Y & Y & \cite{Ciez2017} \\
-    \hline
-        2015 & Li-Ion & Industry & 2015 & Steps & N & N & Y & Y &  \cite{Petri2015} \\
-    \hline
-        2012 & Li-Ion & Manufacturers & 2012-now & Steps & (Y) & N & Y & Y & \cite{Nelson2012} \\
-    \hline
-        2020 & Wind & Impact of breakthroughs & 2005-2017 & Steps & Y & Y & Y & Y & \cite{Elia2020} \\
-    \hline
-        2010 & Wind & Future cost savings & 2010 & Steps & N & N & Y & Y & \cite{Schubel2010} \\
-    \hline
-        2006 & Wind & Design changes & 1999-2006 & Ops. & N & N & Y & N & \cite{Fingersh2006} \\
-    \hline
-        2012 & LED & Cost reductions & 2012 & Steps & N & N & Y & Y & \cite{ledcomv2} \\
-    \hline
-        2020 & LED & Impact of breakthroughs & 2002-now & Steps & (Y) & (Y) & Y & Y & $^\text{this}_\text{work}$\\
-    \hline
-\end{NiceTabularX}
-\vspace{3mm}
-\caption*{The \textit{Base} column describes the base unit for computations.
-This can either be manufacturing steps ("Steps"), cost reductions ("Red.") or total cost of operation ("Ops.").
-The \textit{Parameters} column describes the parameters considered.
-Market parameters include market size, manufacturing equipment discount, economies of scale, etc.
-Economy parameters include taxes, subsidies, debt, inflation, etc.
-"Global" parameters include resource cost, labour cost, overall yield, etc.
-"Step" parameters include equipment performance, equipment cost, step yield etc.
-Models self-described as \textit{"bottom-up"} consider parameters $G,S$.
-Note that those models only capture the manufacturing situation a certain point in time.
-The exception to this are the model \cite{Elia2020} and the present model.
-Abbreviations: Solar Photovoltaics, excluding perovskite cells (Solar), integrated circuits (IC), base level for calculations (Base), Parameters: market (M), economy (E), "global" (G), "step" (S).
-Parameters not considered directly but instead through their impact on other parameters denoted (Y).}
-\label{tab:costmodels}
-\end{table}
+\textit{Semi}, the global industry association representing the electronics and semiconductor manufacturing industry, has published standards for manufacturing cost modeling.
+These include term definitions and equations for calculating cost and the associated quantities, such as equipment depreciation.
+This standard has been widely adopted industry and provides a standardized comparative analysis of the economics of the manufacturing chain \cite{ragona2002cost}\cite{wouters2018industry}.
+An overview of the parameters used in the model, according to \textit{SEMI Standard 35} is provided in \cref{semi_pyramid}.
+Note that the full ensemble of inputs are required only for an industry cost model used by manufacturers for detailed forecasting.
+As per the scope of our model, we need only a selection of these inputs.
+Any inputs unrelated to the manufacturing process itself can be disregarded.
+These include taxes, equipment installation, delivery facilities, customer orders etc.
 
@@
 
-\cite{Scholand2012}\cite{casamayor2018comparative} some more text on model inputs
+\clearpage
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.95\textwidth]{./figures/costmodel.pdf}
+    \caption{\textbf{Cost model structure.} The diagram shows inputs to each step and computational steps leading to the cost model outputs.
+For the description of cost variables, see definitions for \cref{eqn:cost_wafer}.}
+    \label{fig:costmodel-schematic}
+\end{figure}
 
@@
     \includegraphics[width=0.85\textwidth]{./figures/SEMI_pyramid.pdf}
-    \caption{\textit{SEMI Standard 35} pyramid showing variables that enter into the computation of manufacturing cost.
-Variables not captured by our model are greyed out.
-Source: Adapted from \cite{standardE35} and \cite{helin200618th}.}
+    \caption{\textbf{\textit{SEMI Standard 35} pyramid showing variables that enter into the computation of manufacturing cost in our model.} Variables not captured by our model are greyed out.
+The pyramid has been adapted from \cite{standardE35} and \cite{helin200618th}.}
     \label{fig:semi_pyramid}
-    %\vspace{-10pt}
+    \vspace{-20mm}
 \end{figure}
@@
 
-\cref{sec:costmodel_structure} provided additional information on the structure of the cost model.
-As illustrated in \cref{tab:costmodels}, our model is comprehensive in its coverage of market, economic, manufacturing step and global manufacturing parameters.
-However, the model has limitations which must be considered when interpreting the results.
+As illustrated in \cref{fig:semi_pyramid}, our model is comprehensive in its coverage of market, economic, manufacturing step and global manufacturing parameters.
+However, the model has limitations which must be considered when interpreting the results:
 
@@
 
-While the model is setup up to capture the cost structure of a \textit{"virtual"} North American factory, the majority of production capacity is presently located in Asia.
+The model is set up up to capture the cost structure of a \textit{"virtual"} factory employing the most up-to-date equipment in every model year.
+It operates at a fixed-cost basis (labor cost, electricity cost, cleanroom cost, etc.), which was - due to limitations in data availability - taken to be based on a US-based semiconductor manufacturing plant.
+However, this does not correspond to any \textit{real, physical} location.
+The majority of production capacity is presently located in Asia.
 There, the share of domestically produced manufacturing equipment has increased from 10\% in 2008 to 65\% in 2015 \cite{tseng2015fab}.
@@
 operates plants in mainland China \cite{mls2020dev}.
-
-This means that it is difficult to validate model outputs directly with the help of industry data, which is provided only for Asia-based manufacturing lines.
-We compare cost model data to a selection of industry data in \cref{sec:cost_model_calibration} and provide an interpretation of the differing cost structure.
+This means that it is difficult to validate model outputs directly with the help of industry data, which (even for American and European companies) is provided only for Asia-based manufacturing lines.
+
 Generally, we found that our model provides the best-case scenario for LED manufacturing, where the most performant (highest throughput, highest yield) equipment is used in each year.
 Interviews revealed that this is rarely the case in practice.
-One reason for this is economic pressure on western manufacturers after the introduction of significant Chinese subsidies as part of the country's Twelfth Five-Year Plan \cite{sanderson2014light}.
-
-However, because the purpose of our model is to show the impact of changes in the design and manufacturing process of devices, the geographic location and associated potentially lower overhead costs or slower adoption of specific equipment are of secondary importance to our model.
-This is why a US based manufacturing environment was assumed.
+One reason for this is additional economic pressure on western manufacturers following the introduction of significant Chinese subsidies as part of the country's Twelfth Five-Year Plan \cite{sanderson2014light}.
+While in reality, older equipment with lower performance might be used, we model the use of the best available equipment for each year considered.
+
+Because the purpose of our model is to show the impact of changes in the design and manufacturing process of devices, rather than the geographic location and associated economic circumstances, we believe this to be a minor drawback.
+We therefore assumed a US-based manufacturing environment, where the associated overhead costs did not change during the time period considered.
+
+We do compare cost model data to a selection of industry data in \cref{sec:cost_model_calibration} and provide an interpretation of the differing cost structure.
 
@@
 In order to validate the outputs of the cost model regarding the evolution of the cost structure of low-to-mid-power classic chip GaN phosphor-converted white LED, we conducted a comparison with previously reported U.S.
-DOE calculations and projections based on the LEDCOM model and industry data provided as part of industry round table discussions.
-This comparison is shown in \cref{fig:costmodel_calibration}.
-
+DOE calculations and projections based on the \textit{LEDCOM} model and industry data provided as part of industry round table discussions:
+
+\begin{figure}[h]
+	\centering
+    \includegraphics[width=\textwidth]{./figures/costmodel_calibration.pdf}
+	\caption{\textbf{Cost structure comparison between our cost model, industry data and DOE projections.} Top panel: Absolute manufacturing cost, with two U.S.
+DOE projection for context.
+Center panel: manufacturing cost calculations (2008-2016) and projections (2018, 2020) previously published by the U.S.
+DOE based on industry data and LEDCOM model.
+Bottom panel: manufacturing cost structure produced by our cost model for 2002, 2012 and 2020.
+Hatched bars are DOE projections.
+Sources for top/center panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
+Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel_structure}.}
+	\label{fig:costmodel_calibration}
+\end{figure}
+
+\textbf{}
 We note some differences between the results of our model and the cost structure reported or projected by the DOE.
@@
 
-\begin{figure}[h]
-	\centering
-    \includegraphics[width=\textwidth]{./figures/costmodel_calibration.pdf}
-	\caption{\textbf{Cost structure comparison between our cost model, industry data and DOE projections.} Top panel: manufacturing cost calculations (2008-2016) and projections (2018, 2020) previously published by the U.S.
-DOE based on industry data and LEDCOM model.
-Bottom panel: manufacturing cost structure produced by our cost model for 2002, 2012 and 2020.
-Hatched bars are DOE projections.
-Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}.
-Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel_details}.}
-	\label{fig:costmodel_calibration}
-\end{figure}
-
-\clearpage
-\section{Identified LED Innovations and Performance Improvements}
-
-This Supplementary Note provides details for the identified LED innovations and performance improvements that contributed to the historical progress in LED technology across efficiency and consumer experience metrics discussed in the main text.
-
-\textbf{Historical Progress in Sub-Efficiencies}
-
-Empirical data on the historical progress in metrics related to four different sub-efficiencies contributing to the overall LED efficiency (lamp efficiency $\eta_L$) is shown in \cref{fgr:subeff}.
-Specifically, the figure includes data for forward voltage used to calculate to forward voltage efficiency $\eta_{V_f}$ (Panel A1); efficiency droop $\eta_{droop}$ (Panel A2); internal quantum efficiency $\eta_{IQ}$ (Panel B1); and light extraction efficiency $\eta_{LE}$ (Panel B2).
-Additional information and context for individual panels in \cref{fgr:subeff} are provided below.
-
-\textbf{Panel A1:} Device forward voltage at a test current of I=350mA.
-The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
-Data points for devices released in 2020 by various manufacturers are shown in an inset plot.
-\
-
-\textbf{Panel A2:} Efficiency droop at the test current of I=350mA. \
-
-\textbf{Panel B1:} Internal quantum efficiency (IQE) for different chip architectures by type of measurement used.
-Note that the bump in the DOE Average around 2013 is due to a change in IQE definition laid out in \cite{doe_ssl_multiyear_2013}.
-Measurement methods: PL - Photoluminescence \cite{Shim_2018}, EL - Electroluminescence \cite{Getty_2009}; ABC Model \cite{Karpov_2014}, Ideality Factor \cite{Masui_2010}.
-“Not reported” denotes data points where the IQE measurement technique was not reported.
-Whiskers indicate reported ranges, where available.
-\
-
-\textbf{Panel B2:} Light extraction efficiency (LEE) for different chip architectures by type of specific technology used to improve LEE.
-The types of data points are shown for reference: Simulation - ray-tracing computer simulation; Experimental (absolute) – direct LEE values; Experimental (relative) - relative improvement over baseline chip architecture.
-Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
-- electrode; text.
-- textured.
-
-Notably, \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.
-Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the (mostly) empirical \cref{fgr:subeff} not relevant.
-Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum} below (see also Methods section in the main text).
+\clearpage
+\section{Historical Progress in Sub-Efficiencies}
+
+Empirical data on the historical progress in metrics related to four different sub-efficiencies contributing to the overall LED efficiency (lamp efficiency $\eta_L$) is shown in \cref{fgr:subeff}:
 
@@
 \label{fgr:subeff}
-\end{figure}
-
-\clearpage
-\textbf{Non-Phosphor-Related LED Innovations}
-
-\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting four white LED sub-efficiencies: forward voltage efficiency, light extraction efficiency, internal quantum efficiency, droop and light conversion efficiency.
+\vspace{-20mm}
+\end{figure}
+
+\clearpage
+
+\textbf{Panel A1:} This panel shows device forward voltage at a test current of I=350mA, categorized by chip architecture.
+The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
+Forward voltage efficiency is computed from this metric as described in \cref{sec:device_performance_metrics}.
+Data points for devices released in 2020 by various manufacturers are shown in an inset plot for additional context.
+Note that differences between chip architectures in this plot have become small for all but one company.
+Chip architecture abbreviations: Acronyms: TFFC - thin-film flip-chip, FC - flip-chip, CSP-FC - chip-scale package flip-chip.
+Company abbreviations: LG - LG Innotek, SE - Seoul Semiconductor, SA - Samsung, OS - Osram, LL - Lumileds.
+\
+
+\textbf{Panel A2:} Efficiency droop at the test current of I=350mA, categorized by chip architecture.
+Droop efficiency is computed from this metric as described in \cref{sec:device_performance_metrics}.
+Data has been calculated from luminous intensity curves of respective device datasheets.
+\
+
+\textbf{Panel B1:} Internal quantum efficiency (IQE) for categorized by type of measurement used.
+Because it is inherently more difficult to measure IQE than is the case for other sub-efficiencies, this panel shows data categorized by different measurement methods.
+This is not the case for other panels of this figure.
+Note that the bump in the DOE Average around 2013 is due to a change in IQE definition laid out in \cite{doe_ssl_multiyear_2013}.
+Measurement methods abbreviations: PL - Photoluminescence \cite{Shim_2018}, EL - Electroluminescence \cite{Getty_2009}; ABC Model \cite{Karpov_2014}, Ideality Factor \cite{Masui_2010}.
+“Not reported” denotes data points where the IQE measurement technique was not reported.
+Whiskers indicate reported ranges, where available.
+\
+
+\textbf{Panel B2:} Light extraction efficiency (LEE) for different chip architectures by type of specific technology used to improve LEE.
+The sources for all data points are shown for reference: Simulation - ray-tracing computer simulation; Experimental (absolute) – direct LEE values; Experimental (relative) - relative improvement over baseline chip architecture.
+Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
+- electrode; text.
+- textured.
+
+\textbf{Additional Notes on \cref{fgr:subeff}}
+
+Data illustrated in \cref{fgr:subeff} was used to determine the performance of best-in-class devices across all sub-efficiencies. The results are shown in Figures 4-6 in the main text.
+
+\textbf{Additional Data on Phosphor Performance}
+
+Note that \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.
+Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the (mostly) empirical \cref{fgr:subeff} not relevant.
+Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum} below (see also Methods section in the main text).
+
+\clearpage
+\section{Non-Phosphor-Related LED Innovations}
+
+\cref{tab:innovations_1}-\cref{tab:innovations_2} summarize LED innovations and technology and manufacturing process improvements identified in this study as affecting four white LED sub-efficiencies: forward voltage efficiency, light extraction efficiency, internal quantum efficiency, droop and light conversion efficiency.
 Note that this table does not include phosphor-related LED innovations affecting spectral efficiency, as these are presented separately in Table 1 in the main text as affecting consumer experience metrics and discussed in detail in \cref{sec:innovation_phosphor}.
@@
 \begin{table}[H]
-    \caption{LED innovations and technology improvements affecting key white LED device sub-efficiencies.}
+    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note this table is continued in \cref{tab:innovations_2}.}
     \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
@@
         \hline
-        \multicolumn{5}{c}{Droop $\eta_{droop}$} \\
-        \hline
-            ongoing & Lower current Density & Current density & No & I \\
-        \hline
-            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
-        \hline
-            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
-        \hline
-            $<$ 2017 & Defect getting \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
-        \hline
-        \multicolumn{5}{c}{Light conversion efficiency (CE) $\eta_{C}$} \\
-        \hline
-            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
-        \hline
-            Ongoing & Lower current density & Current density & No & I \\
-        \hline
-            Ongoing & Epitaxy \newline Improvements & Charge distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
-        \hline
-            Ongoing & Chip architecture \newline improvements  & Charge distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
-        \hline
-            $<$ 2017 & Defect gettering \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
-        \hline
         \end{NiceTabularX}
+        \vspace{2mm}
     \caption*{Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
@@
 \textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
-    \label{tab:innovations}
+    \vspace{-20mm}
+    \label{tab:innovations_1}
 \end{table}
 
+\begin{table}[H]
+    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note that this table continues \cref{tab:innovations_1}.}
+    \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
+    \hline
+    \multicolumn{5}{c}{Droop $\eta_{droop}$} \\
+        \hline
+            ongoing & Lower current Density & Current density & No & I \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect getting \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
+        \hline
+        \multicolumn{5}{c}{Light conversion efficiency (CE) $\eta_{C}$} \\
+        \hline
+            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
+        \hline
+            Ongoing & Lower current density & Current density & No & I \\
+        \hline
+            Ongoing & Epitaxy \newline Improvements & Charge distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            Ongoing & Chip architecture \newline improvements  & Charge distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect gettering \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
+        \hline
+        \end{NiceTabularX}
+        \vspace{2mm}
+    \caption*{Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
+'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
+\textit{Spillover} column indicates if the innovation involved technology spillovers.
+\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
+    \label{tab:innovations_2}
+\end{table}
+
+\clearpage
 \section{Phosphor-Related LED Innovations}
@@
 
-\clearpage
 \section{Spectral Data for Phosphors}
@@
 \captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for Figure 1 organized by technology.}
+    \caption{List of sources for Figure 1.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}
@@
 \captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for Figure 2 grouped by data provider.}
+    \caption{List of sources for Figure 2.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}
@@
 \captionsetup{justification=raggedright,singlelinecheck=false}
-\caption{List of sources for Figure 3 grouped by publication type.}
+\caption{List of sources for Figure 3.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}
@@
 \captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for \cref{fgr:subeff} grouped by LED sub-efficiencies.}
+    \caption{List of sources for \cref{fig:haitz_new}.}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
+    \hline
+    \textit{Metric} & \textit{References} \\
+    \hline
+    Cost & \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} \\
+    \hline
+    Flux & \cite{haitz2007seoul}\cite{haitz2010osram}\cite{haitz2010lumi-1}\cite{haitz2010cree}\cite{haitz2011lumi-1}\cite{haitz2014nichia}\cite{haitz2015cree-1}\cite{haitz2015cree-2}\cite{haitz2018cree}\cite{haitz2019osram}\cite{haitz2019lumi-1}\cite{haitz2019lumi-1}\cite{haitz2020lumi-1}\cite{haitz2013cob}\cite{haitz2016cob}\cite{haitz2017cob}\cite{haitz2019cob}\cite{haitz2019lumi-1}\cite{haitz2019lumi-2} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+    \caption{List of sources for \cref{fgr:subeff}.}
     \begin{NiceTabularX}{\textwidth}{|l|X|}

### 2024-03-16 2_SSL_NE/article/article_main.tex
various edits

@@
 Over the same period, the manufacturing cost of low-to-mid-power LED packages decreased by 95.5\% from \$1.1 to \$0.05 (in 2020 USD).
-Technology spillovers from other sectors contributed to 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
+Technology spillovers from other sectors accounted for at least 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
 \end{abstract}
@@
 A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}.
-According to the Net Zero Tracker Database, by the end of 2023, more than 150 countries have set or are considering setting a net zero emissions target by 2050 \cite{climateaction}.
-The European Union, for example, has set a target of net zero emissions by 2050.
-It aims to meet this goal with the help of the European Green Deal, alongside other EU and national policies \cite{eu2020green}.
-The United Kingdom similarly has adopted a national strategy to achieve net-zero by 2050 \cite{noauthor_ieairena_2023}.
+While the Paris Agreement of 2015 requires signatory states to develop national emissions reduction strategies, these are not legally binding \cite{bodansky2016legal}.
+However, some countries have introduced more ambitious legislation, including the European Union, which has set a target of net zero emissions by 2050.
+It aims to meet this goal with the help of the European Green Deal, alongside other national policies \cite{eu2020green}.
 
@@
 We find that three spillovers associated with the use of YAG/YGAG phosphors in LEDs played the key role in the development of the first commercial white LED lighting products, essentially enabling the solid-state lighting market and industry of today.
-As noted above, subsequent spillovers then had a significant effect on physical device performance, cumulatively contributing to 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020, as well as nearly all key improvements in consumer experience metrics.
+As noted above, subsequent spillovers then had a significant effect on physical device performance, cumulatively contributing at least 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020, as well as nearly all key improvements in consumer experience metrics.
 
@@
 The overall outcome of the cost model is thus the cumulative yielded manufacturing cost per LED package for each of the three years considered.
-In our model, this metric includes all costs associated with producing the chip, including running costs of the factory.
-Costs associated with research and development, administrative overhead of the manufacturer or other investment costs are not considered.
-We also note that the purpose of our cost model is not to give specific estimates of the LED manufacturing cost for a factory of any size, specific geographic location or total annual manufacturing volume.
-It instead assumes a hypothetical factory with an assumed location in the United States and associated overhead costs related to the operation of the factory.
-It also assumes the use of the most up-to-date equipment for that year.
+In our model, this metric includes all costs associated with producing the chip, including operating costs of the factory.
+Costs associated with research and development, administrative overhead, or other investment costs are not considered.
+We note that the purpose of the cost model is not to provide specific estimates of the LED manufacturing cost for a factory of specific size, geographic location or manufacturing capacity.
+It is instead intended to capture the impact of specific improvements in the manufacturing process flow on overall cost.
+It therefore models a hypothetical factory operating the most up-to-date equipment for each model year.
+Fixed cost parameters are similar to those for a semiconductor factory operating in the United States.
 Even with these simplifying assumptions, the model reasonably identifies the impact that changes in single process steps can have on the total LED manufacturing cost.
 
-Another important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we are able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low- to mid-power devices.
-Populating the model with data for the remaining two architectures requires access to proprietary industry information, which we have not been able to get thus far.
+Another important limitation of our cost modelling efforts is that, even though the model captures three different chip architectures in its structure, in the present study we are able to collect, estimate and present the full set of quantitative inputs and outputs only for the classical chip architecture of low-to-mid-power devices.
+Populating the model with data for the remaining two chip architectures would require additional access to proprietary industry information.
 

### 2024-03-16 2_SSL_NE/article/article_supplementary.tex
various edits

@@
 % https://tex.stackexchange.com/a/712523
-\titleformat{\section}[display]{\large\bfseries\filright}{\bfseries Supplementary Note \thesection}{0pt}{}
+\titleformat{\section}[display]{\large\bfseries\filright}{\bfseries Supplementary Note \thesection}{-5pt}{}
 \crefname{section}{Supplementary Note}{Supplementary Notes}
@@
         \newline
-        $^1$ Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, UK \\
-        $^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
-        $^3$ Technology Assessment Group, Paul Scherrer Institute, Villigen, Switzerland \\
-        $^4$ Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
+        $^1$Centre for Environment, Energy and Natural Resource Governance \\ Department of Land Economy, University of Cambridge, Cambridge, UK \\
+        $^2$Group for Energy Systems Analysis, Department of Mechanical and Process Engineering \\ ETH Zurich, Zurich, Switzerland \\
+        $^3$Technology Assessment Group, Paul Scherrer Institute  Villigen, Switzerland \\
+        $^4$Belfer Center for Science and International Affairs, Harvard University, Cambridge MA, USA
     }
@@
 Within the timeframe and scope laid out in the main text of this publication, it returns the total manufacturing cost of GaN-based phosphor-converted warm white light-emitting diode packages.
-The model assumes production at a \textit{"virtual"} factory which employs the highest performing equipment available in each year.
+The model assumes a hypothetical factory which employs the highest performing equipment available in each year.
 Performance in this context comprises throughput, associated yield and resource consumption.
@@
 
-\textbf{Availability of Equipment Performance Data}
+\textbf{Availability of Equipment Performance Data for all Chip Architectures}
 
@@
 
-The model is set up up to capture the cost structure of a \textit{"virtual"} factory employing the most up-to-date equipment in every model year.
+The model is set up up to capture the cost structure of a hypothetical factory employing the most up-to-date equipment in every model year.
 It operates at a fixed-cost basis (labor cost, electricity cost, cleanroom cost, etc.), which was - due to limitations in data availability - taken to be based on a US-based semiconductor manufacturing plant.
@@
 Because the purpose of our model is to show the impact of changes in the design and manufacturing process of devices, rather than the geographic location and associated economic circumstances, we believe this to be a minor drawback.
-We therefore assumed a US-based manufacturing environment, where the associated overhead costs did not change during the time period considered.
-
+For fixed cost parameters, we therefore assumed a US-based manufacturing environment.
 We do compare cost model data to a selection of industry data in \cref{sec:cost_model_calibration} and provide an interpretation of the differing cost structure.
@@
 \cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering for the manufacturing process of three different chip architectures considered in the cost model.
+
+\vspace{-20mm}
 
@@
 For instance, the share of the epitaxy step is consistently larger in the DOE data.
-This can in part be explained by our model relying on state-of-the-art equipment at a virtual U.S.-based manufacturing location, while industry might not run low-power and mid-power chip production on these, more expensive, reactors.
+This can in part be explained by our model assuming the use of state-of-the-art equipment, while industry might not run low-power and mid-power chip production on these more expensive reactors.
 In addition, the manufacturing lines of the majority of manufacturers are located in Asia.

### 2024-03-16 2_SSL_NE/article/article_main.tex
updated supplementary information introduction

@@
 \newline
-$^1$ Centre for Environment, Energy and Natural Resource Governance, Department of Land Economy, University of Cambridge, Cambridge, UK \\
+$^1$ Centre for Environment, Energy and Natural Resource Governance Department of Land Economy, University of Cambridge, Cambridge, UK \\
 $^2$ Group for Energy Systems Analysis, Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\

### 2024-03-16 2_SSL_NE/article/article_supplementary.tex
updated supplementary information introduction

@@
         \newline 
+        \textbf{Rapid technological progress in white light-emitting diodes\\
+        and its sources in innovation and technology spillovers} \\
+        \newline
         M.P. Weinold$^{1,2,3}$, S. Kolesnikov$^1$, L.D. Anadon$^{1,4}$ \\
         \newline
-        $^1$Centre for Environment, Energy and Natural Resource Governance \\ Department of Land Economy, University of Cambridge, Cambridge, UK \\
-        $^2$Group for Energy Systems Analysis, Department of Mechanical and Process Engineering \\ ETH Zurich, Zurich, Switzerland \\
+        $^1$Centre for Environment, Energy and Natural Resource Governance \\
+        \hspace*{0.7mm} Department of Land Economy, University of Cambridge, Cambridge, UK \\
+        $^2$Group for Energy Systems Analysis \\
+        \hspace*{0.7mm} Department of Mechanical and Process Engineering, ETH Zurich, Zurich, Switzerland \\
         $^3$Technology Assessment Group, Paul Scherrer Institute  Villigen, Switzerland \\
@@
 
-\clearpage
 \tableofcontents

### 2024-03-16 2_SSL_NE/article/article_main.tex
various edits

@@
 Understanding the extent and sources of the rapid progress in white LED technology can thus provide valuable insights for accelerating innovation in other demand-side clean energy technologies.
-Through cost and performance modeling, literature review, patent analysis, and expert interviews, we find that the efficiency of top-performing  warm white GaN-based  LED packages increased from 5.8\% in 2003 to 38.8\% in 2020.
+Here we show, through cost and performance modeling based on data from literature review, patent analysis, and expert interviews, that the efficiency of top-performing warm white GaN-based LED packages increased from 5.8\% in 2003 to 38.8\% in 2020.
 Over the same period, the manufacturing cost of low-to-mid-power LED packages decreased by 95.5\% from \$1.1 to \$0.05 (in 2020 USD).
-Technology spillovers from other sectors accounted for at least 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
+We find that technology spillovers from other sectors accounted for at least 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
 \end{abstract}
@@
 
-\section{Introduction}
+\section{Main}
 \label{sec:intro}
@@
 However, some countries have introduced more ambitious legislation, including the European Union, which has set a target of net zero emissions by 2050.
-It aims to meet this goal with the help of the European Green Deal, alongside other national policies \cite{eu2020green}.
+It aims to meet this goal through an ambitious set of policies known as the European Green Deal \cite{eu2020green}.
 
-Achieving these ambitious and critically important targets will require both the deployment of new clean energy technologies \cite{iea2020cleanenergy}, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}.
+Achieving such ambitious and critically important targets will require both the deployment of new clean energy technologies \cite{iea2020cleanenergy}, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}.
 To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed.
@@
  \label{fgr:history_efficacy}
+ \vspace{-7mm}
 \end{figure}
 
-\begin{figure}[h!]
+\clearpage
+
+\begin{figure}[H]
 \centering
-  \includegraphics[height=6.5cm]{figures/cost_lamp_small.pdf}
+  \includegraphics[height=5cm]{figures/cost_lamp_small.pdf}
   \caption{\textbf{Historical development of retail sales prices per luminous flux for LED-based luminaires in the period 2008-2020.} Luminaires considered include light bulbs, spotlights and recessed lights.
@@
 These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent light bulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}.
-As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions.
+As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year in the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced in the same year by all photovoltaic installations in these regions.
 Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}.
-For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
+For example, durability, low up-front cost and high efficiency of LED light sources have led to early and widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}.
 LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.
@@
 
-To address these questions, first, we identify three sets of metrics suitable for tracking the historical progress of LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, including colour rendering index (CRI) and colour temperature (CT); and 3) LED device manufacturing cost.
+\clearpage
+
+To address these questions, we use three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost.
 The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Note 2.
-We then trace the historical improvements in these metrics, along with associated changes in LED chip architectures, from the time of introduction of the first commercial warm white GaN-based LEDs in 2003 to 2020, the year with the most recent data available at the time of data collection.
-Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a comprehensive review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry (see Methods section and Supplementary Note 3 for details).
-This approach allows us to identify individual innovations in white LED technology, examine their knowledge origins to discern technology spillovers among them and quantify their contributions to the overall progress in LED efficiency and consumer experience metrics based on our modeling of LED performance and spectral metrics.
-We further develop a bottom-up LED manufacturing cost model with process-step resolution and use it to analyze the progress in LED manufacturing cost for the “classical” GaN-based LED chip architecture with lateral current spreading shown in Panel A in \cref{fgr:chip_architecture_overview}.
+We find that  total LED device efficiency increased from 5.8\% to 38.8\% between 2003 and 2020, of which at least 8.5\% can be attributed specifically to technology spillovers.
+Our interviews have also revealed the important role of incremental manufacturing process improvements and learning-by-doing \cite{WRIGHT_1936, Arrow_1962} in this progress in LED efficiency.
+However, further research is needed to quantify the contribution of learning-by-doing to this progress.
+On the other hand, we find that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
+Finally, our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid-power classic-chip-architecture GaN-based phosphor-converted warm-white LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020.
+In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.
+Overall we find that technology spillovers are most prevalent in those device loss channels that are well understood at the physical level.
+Improvements in device loss channels that are governed by complex quantum effects at the atomic level, and can only be described heuristically, have come from incremental manufacturing process improvements and learning-by-doing.
+Directing research efforts to better understand fundamental physical or chemical processes related to novel clean energy technologies is therefore likely to increase the possibility of technology spillovers.
 
 \clearpage
-\section{Main}
 
-% should be integrated into some other sub-heading
-% \subsection{Evolution of LED Chip Architectures}
-
-\subsection{Efficiency Improvements}
+\section{Efficiency Improvements}
 
@@
 
-\subsection{Consumer Experience Improvements}
+\section{Consumer Experience Improvements}
 
@@
 
-\subsection{Manufacturing Cost Improvements}
+\section{Manufacturing Cost Improvements}
 
@@
 
-\subsection{Technology Spillovers}
+\section{Technology Spillovers}
 
@@
 Improvements in these metrics are traced to sources in innovation, technology spillovers, and economies of scale in the manufacturing process.
-
-We find that the total LED device efficiency increase from 5.8\% to 38.8\% between 2003 and 2020, as well as all major improvements in consumer experience metrics, have been predominantly driven by LED technology innovations that reduced energy losses across all physical loss channels in LED devices and introduced new phosphor materials for LED light down conversion.
-Our interviews have also revealed an important role of incremental manufacturing process improvements and learning-by-doing \cite{WRIGHT_1936, Arrow_1962} in the progress in LED efficiency.
-However, further research is needed to quantify the contribution of learning-by-doing to this progress.
-
-Our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid power GaN-based white classic-chip LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020.
-In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.
+Improvements in device efficiency are shown in \cref{fgr:waterfall} and \cref{fgr:breakthroughs_efficiency}.
+Improvements in consumer experience are shown in \cref{fgr:consumer_experience}.
+Improvements in manufacturing cost are shown in \cref{fgr:costmodel}.
 
@@
 We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
-
 Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.
@@
 
-We find that technology spillovers into white LED technology have been significant drivers of improvements in LED efficiency and, particularly, consumer experience metrics.
 Our analysis of the sources, mechanisms and enablers of the identified technology spillovers has important practical implications, highlighting the critical role played by a deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.

### 2024-03-22 2_SSL_NE/article/article_main.tex
added draft of domain definition

@@
 
+Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time.
+There are several explanations for this finding.
+First, the functional unit of our study is the LED chip itself, not the luminaire containing multiple chips.
+As LED efficiency and overall brightness have increased, a smaller number of chips can be used in luminaires to obtain the same level of the overall light flux.
+However, since we are investigating improvements at the level of single chips, we do not capture the cost effects downstream of chip manufacturing.
+We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
+Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.
+Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular LED innovations, which was confirmed by interviews with experts in LED manufacturing.
+
+Notably, our bottom-up cost model is constructed to provide process-step resolution across three different key chip architectures: classical chips, flip chips, and chip-scale package flip chips.
+However, in this study we were able to collect data and compare the outcomes only for the classical chip architecture.
+Collecting the full set of data needed to populate the model for the remaining two architectures requires access to proprietary information from industry.
+With this limitation, tracking manufacturing cost reductions across all three key LED chip architectures remains a topic for future work.
+
 Our findings are further supported by a preliminary sensitivity analysis, presented in Section 5 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
 
-In Supplementary Note 6, we also provide a comparison of the outcomes of our model with past cost calculations and projections published by the U.S.
-Department of Energy (DOE) on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported to the DOE as part of SSL round tables.
+In Supplementary Note 6, we also provide a comparison of our model results with past cost report and projections published by the U.S.
+Department of Energy (DOE) on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported as part of SSL round table meetings.
 
@@
 Improvements in manufacturing cost are shown in \cref{fgr:costmodel}.
-
-Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time.
-There are several explanations for this finding.
-First, the functional unit of our study is the LED chip itself, not the luminaire containing multiple chips.
-As LED efficiency and overall brightness have increased, a smaller number of chips can be used in luminaires to obtain the same level of the overall light flux.
-However, since we are investigating improvements at the level of single chips, we do not capture the cost effects downstream of chip manufacturing.
-We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
-Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.
-Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular LED innovations, which was confirmed by interviews with experts in LED manufacturing.
-
-Notably, our bottom-up cost model is constructed to provide process-step resolution across three different key chip architectures: classical chips, flip chips, and chip-scale package flip chips.
-However, in this study we were able to collect data and compare the outcomes only for the classical chip architecture.
-Collecting the full set of data needed to populate the model for the remaining two architectures requires access to proprietary information from industry.
-With this limitation, tracking manufacturing cost reductions across all three key LED chip architectures remains a topic for future work.
 

### 2024-03-22 2_SSL_NE/article/article_supplementary.tex
added draft of domain definition

@@
 % layout
-\input{2_SSL_NE/article/auxiliary/preamble/2_preamble_layout_paperdraft}
+\input{auxiliary/preamble/2_preamble_layout_paperdraft}
+
+% formatting
+\input{auxiliary/preamble/3_preamble_formatting_listings}
 
@@
 \clearpage
+\section{Domain Definition}
+\label{sec:domain}
+
+\textit{Elsevier Scopus}\footnote{\url{https://www.scopus.com/search/}} advanced search query:
+
+\begin{pseudocode}
+TITLE-ABS-KEY(("light-emitting" OR "light emitting") AND "diode")
+AND NOT
+TITLE-ABS-KEY(solar OR violet OR laser)
+AND NOT
+TITLE-ABS-KEY(perovskite OR GaAsP OR AlGaInP OR GaAlAs OR ZnO)
+AND NOT
+TITLE-ABS-KEY(organic OR polymer*)
+AND NOT
+SUBJAREA(ARTS OR BUSI OR DECI OR ECON OR PSYC OR SOCI)
+AND NOT
+SUBJAREA(MEDI OR NURS OR VETE OR DENT OR HEAL OR MULT)
+AND NOT
+SUBJAREA(AGRI OR BIOC OR IMMU OR NEUR OR PHAR)
+AND NOT
+SUBJAREA(envi OR EART OR comp)
+\end{pseudocode}
+
+Something about patents also?
+
+\clearpage
 \section{Full List of Sources for Figures}

### 2024-03-23 2_SSL_NE/article/article_main.tex
in the middle of issues with supplement toc

@@
 
-\clearpage
-\begin{figure}[h!]
-\centering
-\includegraphics[height=15cm]{figures/chip_architecture_overview.pdf}
-\caption{\textbf{Historical evolution of LED chip architectures.} Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
-Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
-The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip.
-Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity.
-Years indicated correspond to the earliest identified patent priority date.
-Black frames around certain designs indicate chip designs not brought to large-scale production.
-Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
-Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Note 16 for the full list of sources and references.}
-\label{fgr:chip_architecture_overview}
-\end{figure}
-
-We summarize the information collected and analysed on the progress in GaN-based LED chip architectures and manufacturing processes in \cref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
-We use this information as input into LED performance calculations and manufacturing cost modeling.
-
 \subsection{Manufacturing Cost Model}

### 2024-03-23 2_SSL_NE/article/article_supplementary.tex
in the middle of issues with supplement toc

--- 2024-03-16 2_SSL_NE/article/article_supplementary.tex
@@
 % https://tex.stackexchange.com/a/712523
-\titleformat{\section}[display]{\large\bfseries\filright}{\bfseries Supplementary Note \thesection}{-5pt}{}
-\crefname{section}{Supplementary Note}{Supplementary Notes}
+
+\newcounter{suppnotecounter}
+\newcommand{\suppnote}[1]{%
+  \stepcounter{suppnotecounter}%
+  \subsection*{\large\bfseries\filright Supplementary Note \arabic{suppnotecounter}: \\ #1}%
+  \addcontentsline{toc}{subsection}{#1}%
+  \label{suppnote:#1}
+}
+
+\newcounter{suppdisccounter}
+\newcommand{\suppdisc}[1]{%
+  \stepcounter{suppdisccounter}%
+  \subsection*{\large\bfseries\filright Supplementary Discussion \arabic{suppdisccounter}: \\ #1}%
+  \addcontentsline{toc}{subsection}{#1}%
+}
+
+\newcounter{suppmethcounter}
+\newcommand{\suppmeth}[1]{%
+  \stepcounter{suppmethcounter}%
+  \subsection*{\large\bfseries\filright Supplementary Method \arabic{suppmethcounter}: \\ #1}%
+  \addcontentsline{toc}{subsection}{#1}%
+}
+
+\newcounter{supprefcounter}
+\newcommand{\suppref}[1]{%
+  \stepcounter{supprefcounter}%
+  \subsection*{\large\bfseries\filright Supplementary References \arabic{supprefcounter}: \\ #1}%
+  \addcontentsline{toc}{subsection}{#1}%
+}
+
 % custom equation designations
@@
 \clearpage
-\section{Previous Literature on Technological Progress in LEDs}
+
+\addcontentsline{toc}{section}{Supplementary Notes}
+
+\suppnote{Previous Literature on Technological Progress in LEDs}
 \label{sec:prev_lit}
+
+\cref{sec:prev_lit}
 
@@
 
-\begin{itemize}
-    \item Publications often look only at progress across a select few LED device sub-efficiencies, often during a very short time period and, most importantly, only at devices manufactured by one specific company, e.g., OSRAM Opto Semiconductors in \cite{hahn2014development} or Lumileds in \cite{Bhardwaj2017}.
-    \item Publications do not report on a number of potentially sensitive or proprietary advances, especially following the market entry of Chinese LED manufacturing companies as a result of the 12th Five-Year Plan (cf.
+Publications often look only at progress across a select few LED device sub-efficiencies, often during a very short time period and, most importantly, only at devices manufactured by one specific company, e.g., OSRAM Opto Semiconductors in \cite{hahn2014development} or Lumileds in \cite{Bhardwaj2017}.
+Publications do not report on a number of potentially sensitive or proprietary advances, especially following the market entry of Chinese LED manufacturing companies as a result of the 12th Five-Year Plan (cf.
 \cite{guo2017china}).
-    \item Publications typically do not look at the overall LED efficiency and its underlying physical limits and tend to overlook the metrics of progress related to consumer experience. 
-\end{itemize}
+Publications typically do not look at the overall LED efficiency and its underlying physical limits and tend to overlook the metrics of progress related to consumer experience.
 
@@
 
-\clearpage
-\section{Metrics for Technological Progress in LEDs}
-
-\textbf{Motivation for Methods Selection}
+\suppnote{Manufacturing Process Steps}
+\label{sec:manufacturing_steps}
+
+\cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering for the manufacturing process of three different chip architectures considered in the cost model.
+
+\vspace{-20mm}
+
+    %CLASSICAL CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
+            \caption{\textbf{Manufacturing process for a classical LED package with lateral current spreading, circa 2003.} MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
+        \label{fig:manuf_classical_2003}
+        \end{figure}
+    \end{landscape}
+
+    %VERTICAL THIN-FILM FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
+            \caption{\textbf{Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
+(1/2)} MQW - multiple quantum well; LLO - laser lift-off.
+Continued in \cref{fig:manuf_vtf_2012-2}}
+            \label{fig:manuf_vtf_2012-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
+            \caption{\textbf{Manufacturing process for a vertical thin-film package flip-chip LED chip with vertical current spreading, circa 2012 (2/2)}.
+Continued from \cref{fig:manuf_vtf_2012-1}.
+Encaps.
+- encapsulation.}
+            \label{fig:manuf_vtf_2012-2}
+        \end{figure}
+    \end{landscape}
+
+    % CHIP-SCALE PACKAGE FLIP-CHIP
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
+            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+(1/2)} MQW - multiple quantum well.
+Continued in \cref{fig:manuf_csp_2020-2}.}
+            \label{fig:manuf_csp_2020-1}
+        \end{figure}
+    \end{landscape}
+
+    \begin{landscape}
+        \begin{figure}
+            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
+            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
+(2/2)}.
+Continued from \cref{fig:manuf_csp_2020-1}.
+Encaps.
+- encapsulation.}
+            \label{fig:manuf_csp_2020-2}
+        \end{figure}
+    \end{landscape}
+
+\clearpage
+
+\suppnote{Historical Progress in Chip Architectures}
+\label{sec:chip_architecture}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=15cm]{figures/chip_architecture_overview.pdf}
+\caption{\textbf{Historical evolution of LED chip architectures.} Panel A: classical chip with lateral current spreading; Panel B: Osram’s thin GaN flip-chip (vertical) architecture; Panel C: flip-chip architecture; Panel D: chip-scale package flip-chip architecture.
+Shown are side views of chips without packages, along a cutaway line best suited to the features of each architecture.
+The cutaway surface is indicated by a red arrow with an eye on the overlaid top view of each chip.
+Note that the dimensions are not to scale, and smaller features are greatly exaggerated for clarity.
+Years indicated correspond to the earliest identified patent priority date.
+Black frames around certain designs indicate chip designs not brought to large-scale production.
+Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
+Source: adapted and compiled from multiple patents and industry publications.
+See Supplementary Note 16 for the full list of sources and references.}
+\label{fgr:chip_architecture_overview}
+\end{figure}
+
+We summarize the information collected and analysed on the progress in GaN-based LED chip architectures and manufacturing processes in \cref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+We use this information as input into LED performance calculations and manufacturing cost modeling.
+
+\clearpage
+\suppnote{Historical Progress in Sub-Efficiencies}
+\label{sec:subeff_progress}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=15.5cm]{./figures/subefficiencies_progress.pdf}
+\caption{\textbf{Historical improvements in LED sub-efficiencies.} Panels A1-A2 show data for key metrics used to calculate forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
+Panels B1-B2 show data for internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
+Source: own synthesis of data from the list of sources provided in \cref{sec:sources}.
+For an illustration of the different chip architectures shown, compare \cref{sec:chip_architecture}.
+Data on average device performance adapted from U.S.
+Department of Energy (DOE) Reports \cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report \cite{beale_leds_2015}.}
+\label{fgr:subeff}
+\vspace{-20mm}
+\end{figure}
+
+\clearpage
+
+\textbf{Additional Notes on \cref{fgr:subeff}}
+
+Data illustrated in \cref{fgr:subeff} was used to determine the performance of best-in-class devices across all sub-efficiencies. The results are shown in Figures 4-6 in the main text.
+
+\textbf{Panel A1:} This panel shows device forward voltage at a test current of I=350mA, categorized by chip architecture.
+The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
+Forward voltage efficiency is computed from this metric as described in \cref{sec:device_performance_metrics}.
+Data points for devices released in 2020 by various manufacturers are shown in an inset plot for additional context.
+Note that differences between chip architectures in this plot have become small for all but one company.
+Chip architecture abbreviations: Acronyms: TFFC - thin-film flip-chip, FC - flip-chip, CSP-FC - chip-scale package flip-chip.
+Company abbreviations: LG - LG Innotek, SE - Seoul Semiconductor, SA - Samsung, OS - Osram, LL - Lumileds.
+\
+
+\textbf{Panel A2:} Efficiency droop at the test current of I=350mA, categorized by chip architecture.
+Droop efficiency is computed from this metric as described in \cref{sec:device_performance_metrics}.
+Data has been calculated from luminous intensity curves of respective device datasheets.
+\
+
+\textbf{Panel B1:} Internal quantum efficiency (IQE) for categorized by type of measurement used.
+Because it is inherently more difficult to measure IQE than is the case for other sub-efficiencies, this panel shows data categorized by different measurement methods.
+This is not the case for other panels of this figure.
+Note that the bump in the DOE Average around 2013 is due to a change in IQE definition laid out in \cite{doe_ssl_multiyear_2013}.
+Measurement methods abbreviations: PL - Photoluminescence \cite{Shim_2018}, EL - Electroluminescence \cite{Getty_2009}; ABC Model \cite{Karpov_2014}, Ideality Factor \cite{Masui_2010}.
+“Not reported” denotes data points where the IQE measurement technique was not reported.
+Whiskers indicate reported ranges, where available.
+\
+
+\textbf{Panel B2:} Light extraction efficiency (LEE) for different chip architectures by type of specific technology used to improve LEE.
+The sources for all data points are shown for reference: Simulation - ray-tracing computer simulation; Experimental (absolute) – direct LEE values; Experimental (relative) - relative improvement over baseline chip architecture.
+Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
+- electrode; text.
+- textured.
+
+\textbf{Additional Data on Phosphor Performance}
+
+Note that \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.
+Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the (mostly) empirical \cref{fgr:subeff} not relevant.
+Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum} below (see also Methods section in the main text).
+
+\clearpage
+\suppnote{Non-Phosphor-Related LED Innovations}
+
+\cref{tab:innovations_1}-\cref{tab:innovations_2} summarize LED innovations and technology and manufacturing process improvements identified in this study as affecting four white LED sub-efficiencies: forward voltage efficiency, light extraction efficiency, internal quantum efficiency, droop and light conversion efficiency.
+Note that this table does not include phosphor-related LED innovations affecting spectral efficiency, as these are presented separately in Table 1 in the main text as affecting consumer experience metrics and discussed in detail in \cref{sec:innovation_phosphor}.
+
+\begin{table}[H]
+    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note this table is continued in \cref{tab:innovations_2}.}
+    \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
+        \hline
+        \multicolumn{5}{c}{Forward Voltage Efficiency (FVE) $\eta_{Vf}$} \\
+        \hline
+            \textit{Year} & \textit{Innovation} & \textit{Area of Improvement} & \textit{Spillover} & \textit{Source} \\
+        \hline
+            1999 & Indium Tin Oxide \newline Current spreading layer & Contact resistance & Yes & \cite{margalith1999indium}\\
+        \hline
+            1998 & Digitated electrodes & Contact resistance & No & \cite{steigerwald2001electrode} \\
+        \hline
+            Ongoing & Epitaxy improvements \newline and better doping & Polarization and \newline bulk resistance & No & I \\
+        \hline
+            1998 & Silver p-Contacts & Contact resistance & No & \cite{kondoh2001nitride} \\
+        \hline
+        \multicolumn{5}{c}{Light extraction efficiency (LEE) $\eta_{LE}$} \\
+        \hline
+            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
+        \hline
+            $<$ 2003 & Optimization for \newline cavity effects & Reduces self-interference of quantum well & No & I, \cite{Shen2003} \\
+        \hline
+            1993 & Chip surface randomization & Total reflection and absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
+        \hline
+            1993 & Thin-film chip architecture & Absorption & No & I, \cite{Schnitzer1993} \\
+        \hline
+            1996 & Patterned sapphire substrate  & Total reflection and absorption & Yes & I, \cite{Tadatomo2001} \\
+        \hline
+            Ongoing & Chip design for high LEE & Total reflection and absorption & No & I, \cite{Haerle2004} \\
+        \hline
+            $\sim$2000 & Silver p-contacts & Absorption & No & I, \cite{kondoh2001nitride} \\
+        \hline
+        \multicolumn{5}{c}{Internal quantum efficiency (IQE) $\eta_{IQ}$} \\
+        \hline
+            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
+        \hline
+            1994 & Double heterostructure & Higher radiative recombination probability & No & \cite{Nakamura1994} \\
+        \hline
+            1996 & Multiple quantum well & Higher radiative recombination probability & No & \cite{Koike1996} \\
+        \hline
+            Ongoing & Active region \newline Doping & Radiative recombination probability & No & I, \cite{schubert2018light} \\
+        \hline
+            Ongoing & Epitaxy \newline Improvements & Radiative recombination probability & No & I \\
+        \hline
+            Ongoing & Chip architecture \newline Improvements & Radiative recombination probability & No & I \\
+        \hline
+        \end{NiceTabularX}
+        \vspace{2mm}
+    \caption*{Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
+'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
+\textit{Spillover} column indicates if the innovation involved technology spillovers.
+\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
+    \vspace{-20mm}
+    \label{tab:innovations_1}
+\end{table}
+
+\begin{table}[H]
+    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note that this table continues \cref{tab:innovations_1}.}
+    \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
+    \hline
+    \multicolumn{5}{c}{Droop $\eta_{droop}$} \\
+        \hline
+            ongoing & Lower current Density & Current density & No & I \\
+        \hline
+            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect getting \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
+        \hline
+        \multicolumn{5}{c}{Light conversion efficiency (CE) $\eta_{C}$} \\
+        \hline
+            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
+        \hline
+            Ongoing & Lower current density & Current density & No & I \\
+        \hline
+            Ongoing & Epitaxy \newline Improvements & Charge distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
+        \hline
+            Ongoing & Chip architecture \newline improvements  & Charge distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
+        \hline
+            $<$ 2017 & Defect gettering \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
+        \hline
+        \end{NiceTabularX}
+        \vspace{2mm}
+    \caption*{Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
+'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
+\textit{Spillover} column indicates if the innovation involved technology spillovers.
+\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
+    \label{tab:innovations_2}
+\end{table}
+
+\clearpage
+\suppnote{Phosphor-Related LED Innovations}
+\label{sec:innovation_phosphor}
+
+This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics.
+These include a detailed description of the history of those innovations, which we use to identify the details of involved technology spillovers.
+In addition, spectral data for the first identified LED products that used those innovations is presented in \cref{fig:phosphor_spectrum}, which we use to calculate consumer experience metrics, as well as LED spectral efficiency.
+
+\textbf{YAG and YGAG Phosphors}
+
+Prior to Nakamura’s invention of highly efficient blue LEDs in the 1990s, Japan’s Nichia Corporation had not sold commercially successful semiconductor products, instead specializing in phosphors for cathode ray tubes (CRT) and fluorescent lamps \cite{nakamura2013blue}  .
+Nevertheless, extensive firm expertise in this area helped Nichia’s Yoshinori Shimizu formulate the principles of using CRT phosphors to convert blue light from Nakamura’s LEDs into white light in 1994 \cite{shimizu1994sheet}\cite{cho2017white}.
+By 1996, Shimizu and his colleagues developed \cite{bando1996}\cite{shimizu1999light} the first practical LED application of a well-known Yttrium Aluminium Garnet (YAG) CRT phosphor activated with cerium \cite{blasse1967new}, enabling the first commercial white LED products manufactured and sold by Nichia since late 1996 \cite{bando1998development}\cite{cho2017white}.
+
+Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 6 in the main article).
+An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
+Used in combination with red-emitting sulfide phosphors, by 2003 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}.
+However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased.
+New chemically stable and non-toxic red phosphors were needed.
+
+\textbf{258, SLA and SALON phosphors}
+
+In 1997, Hubert Huppertz and Wolfgang Schnick, working at the University of Bayreuth in Germany, synthesized the first compound in a new class of rare earth nitridosilicate materials \cite{Huppertz1997} later dubbed “258” due to a proportion of elements in its chemical formula.
+The luminescent properties of these materials were identified by the Schnick’s group, by then at Ludwig-Maximilians University of Munich, in 2000 \cite{Hppe2000} after a suggestion made to Schnick at a conference following earlier reports of good luminosity properties of europium-doped compounds \cite{Qiua1998}.
+U.S.-based LED manufacturer Lumileds applied for a patent for the first class of red LED phosphors based on the 258 nitridosilicate chemistry in 2003 \cite{mueller2004phosphor}.
+The first use of the 258 phosphor in a commercial “Luxeon” LED package was then reported in a joint publication co-authored by inventors from Lumileds and researchers from the Schnick’s group in 2005 \cite{MuellerMach2005}.
+
+Further efforts in LED phosphor development were directed towards synthesizing a red narrow-band phosphor.
+Narrow LED emission peak widths yield the highest luminous efficacy of radiation, as in this case less light is emitted in the far-red range of the spectrum in which the human eye is not very sensitive.
+After synthesizing several narrow-band phosphors emitting in yellow \cite{Hppe2004} and cyan \cite{Kechele2009}, the Schnick’s group identified the local cubic cation coordination structure of the cyan phosphor compound as the reason for its narrow band width \cite{lumi2016narrow}.
+A search for a structurally analogous nitride compound with the narrow red instead of the cyan emission was undertaken.
+After several unsuccessful attempts, the sought-after cuboidal nitride compound was found in a 2008 publication led by Francis DiSalvo \cite{Park2008Sr}.
+Based on information provided in this work, Schnick and colleagues synthesized and studied the spectral properties of a new narrow band red SLA phosphor in 2013 \cite{schmidt2013new}\cite{Pust2014}\cite{schmidt2017phosphors}.
+The material was introduced in commercial LED devices by Lumileds in 2015 \cite{lumi2016narrow_whitepaper}.
+
+The most recent red narrow-band phosphor innovation included in Table 1 and Figure 6 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
+The first U.S.
+patent application for this phosphor was filed in 2017 \cite{seibald2019phosphor}.
+The SALON phosphor is a derivate of the SLA phosphor.
+Therefore, it is the only innovation related to consumer experience metrics identified in our study that seemingly not involved technology spillovers.
+
+\textbf{PFS Phosphor}
+
+Down conversion with ultra-narrow-band phosphor can achieve the highest spectral efficiency.
+However, few such phosphors have been identified, with even less exhibiting desirable material properties such as thermal stability \cite{Phillips2007}.
+The first commercially successful ultra-narrow-band red phosphor was developed by General Electric (GE).
+It is based on a potassium fluorosilicate (PFS) compound activated with manganese ions.
+Its luminescence was first recorded by Adrian Paulusz at GE in 1972 \cite{paulusz1973efficient}.
+In the early 2000s, while searching for potential new LED phosphor materials for GE’s lighting business at GE Lumination, Emil Radkov rediscovered Paulusz’s findings in the literature.
+Following extensive research on PFS chemical synthesis and material properties conducted in collaboration with the University of Sofia in Bulgaria, Radkov’s Alma Mater, the PFS phosphor had been under development at GE since 2005 \cite{radkov2006red}\cite{radkov2009red}.
+This work, supported by public funding from the U.S.
+Department of Energy (DOE) Solid-State Lighting program \cite{doesslprogram}, resulted in a series of critical improvements in the PFS phosphor properties \cite{Setlur2010}\cite{lyons2012color}, eventually enabling its commercialization under the “TriGain” brand in 2015 \cite{trigain_spectrum}\cite{setlur2015trigain}\cite{Murphy2015}.
+
+\textbf{Quantum Dots for Light Down-Conversion}
+
+Quantum dots (QD) are semiconductor nanocrystals whose quantum size effects make QDs behave as “artificial atoms”.
+Semiconductor quantum dots were first synthesized in the Soviet Union in 1981 \cite{ekimov1981quantum} and at Bell Labs in the U.S.
+in 1983 \cite{Rossetti1983}.
+Luminescent properties of quantum dots were first empirically observed in 1984 \cite{fojtik1984photo} and extensively studied in the early 1990s.
+The key feature of QD luminescence discovered in those studies is that its colour is determined by the QD particle size, making it possible to create pure monochromatic blue, green and red light sources just by tuning the QD size.
+The first application of QDs in LEDs was reported in 1994 in an electroluminescent hybrid QD-polymer LED.
+However, this LED type could not be used in general illumination due to its very low luminous efficacy.
+An alternative application of QDs as a kind of a “phosphor” for light down conversion from an LED light source was proposed in the early 2000s as part of the U.S.
+Department of Energy (DOE)-funded “A Revolution in Lighting“ project at Sandia National Laboratory \cite{simmonsfinal}.
+This concept was successfully demonstrated by Sandia researchers on a commercial LED in 2003 \cite{shea_rohwer_development_2004}\cite{noauthor_sandia_nodate} and was swiftly taken up and advanced further by a group in Taiwan \cite{Chen_2005}\cite{Hsueh_Shih_Chen_2006} The first commercial application of QDs in an LED lamp was brought about by a collaboration between an MIT-born startup QD Vision and the U.S.-based luminaire manufacturer Nexxus Lighting in 2009 \cite{ledprof_nexxusqd}, \cite{bourzac2013quantum}.
+However, rapid advances in the spectral and conversion performance of down-conversion phosphors and high manufacturing cost of quantum dots resulted in the discontinuation of this product.
+After finding market success in display backlighting first demonstrated by Samsung in 2010 \cite{Jang2010} and commercialized by QD Vision in Sony television sets in 2013 \cite{bourzac2013quantum}, QDs returned to the general lighting market in products offered by Lumileds \cite{noauthor_global_2017}\cite{noauthor_quantum_2020} around 2017 and Osram in 2019 \cite{osramqdots} in the form of mid-power LED packages that combined QDs with traditional phosphors for light down conversion.
+
+\suppnote{Spectral Data for Phosphors}
+
+\begin{figure}[H]
+	\centering
+    \includegraphics[width=0.9\textwidth]{./figures/phosphor_spectrum-comparison.pdf}
+	\caption{\textbf{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data.} Corresponding LED products that used phosphor innovations listed in Table 1 in the main article are each indicated by the phosphor label and spectral data publication year.
+Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 6 in the main article.
+The desirable direction of improvements towards higher LER at higher CRI is indicated by a red arrow.
+Metrics were calculated from spectral data shown in the bottom panels using the \texttt{colour-science} package for Python \cite{colour-science_software}.
+Bottom two rows of panels: spectral data for the first identified LED products that used phosphor innovations listed in Table 1 in the main article.
+The luminosity function \cite{cie-term-lumeff}, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
+Peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel.
+Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: ?
+-  other parts of phosphor mixture not disclosed, G$^*$ - \ch{CaSrS:Eu^{2+}}, Y$^\star$ - $\beta-$SiAlON:Eu$^{2+}$, G$^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, O$^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
+Source (top panel): own elaboration based on spectral data.
+Arb.
+Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum.
+Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
+\label{fig:phosphor_spectrum}
+\end{figure}
+
+\clearpage
+\suppref{Full List of Sources for Figures}
+\label{sec:sources}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+    \caption{List of sources for Figure 1.}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
+    \hline
+    \textit{Technology} & \textit{References} \\
+    \hline
+    LED & Own research, compare \cite{zenodo_weinold_led_history} \\
+    \hline
+    CFL (<1984) & \cite{Bouwknegt1982}\cite{Vrenken1983} \\
+    \hline
+    CFL (1984-2011) & \cite{eger2018origin} \\
+    \hline
+    CFL (>2011) & \cite{Guan2015} \\
+    \hline
+    Fire, Incandescent, HID & \cite{azevedo2009transition} augmented by own calculations based on \cite{benesch1905beleuchtungswesen} \\
+    \hline
+    Max. efficacy & \cite{Murphy2012} \\
+    \hline
+    \end{NiceTabularX}
+    \vspace{2mm}
+    \caption*{Abbreviations: CFL - Compact Fluorescent Lamp, HID - High-Intensity Discharge}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+    \caption{List of sources for Figure 2.}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
+    \hline
+    \textit{Source} & \textit{References} \\
+    \hline
+    Stiftung Warentest (Germany) & \cite{Warentest2008}\cite{Warentest2009_1}\cite{Warentest2009_2}\cite{Warentest2010_1}\cite{Warentest2010_2}\cite{Warentest2011}\cite{Warentest2012}\cite{Warentest2013}\cite{Warentest2014_1}\cite{Warentest2014_2}\cite{Warentest2015}\cite{Warentest2016_1}\cite{Warentest2016_2}\cite{Warentest2018} \\
+    \hline
+    Konsument (Austria) & \cite{Konsument2010} \\
+    \hline
+    Which (UK) & \cite{Which2020} \\
+    \hline
+    Industry Periodical & \cite{PM2020} \\
+    \hline
+    Government Report (US) & \cite{national2013assessment} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+\caption{List of sources for Figure 3.}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
+    \hline
+    \textit{Publication Type} & \textit{References} \\
+    \hline
+    Scientific Publications & \cite{plossl2010wafer}\cite{bierhuizen2007performance}\cite{gencc2019distributed}\cite{chong2014performance} \\
+    \hline
+    Patents & \cite{patent1999uemura}\cite{patent1998takaoka}\cite{patent1999komaki}\cite{patent1999komaki}\cite{ludowise2006resonant}\cite{camras2005iii}\cite{steigerwald2004contacting} \\
+    \hline
+    Other Publications & \cite{craford2015}\cite{sun2016}\cite{yole2013packaging} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+\caption{List of sources for Figure 6.}
+    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|}
+    \hline
+    \textit{Phosphor/QD} & \textit{Company} & \textit{YOI} & \textit{Reference} \\
+    \hline
+    YAG & Nichia & 1998 & \cite{bando1998development} \\
+    \hline
+    YGAG & Lumileds & 2002 &  \cite{Mueller2002} \\
+    \hline
+    258 & Lumileds & 2005 & \cite{MuellerMach2005} \\
+    \hline
+    SLA & Lumileds & 2014 & \cite{Pust2014} \\
+    \hline
+    PFS & GE & 2015 & \cite{Murphy2015} \\
+    \hline
+    QD & Lumileds & 2017 & \cite{lumileds2016qd} \\
+    \hline
+    Salon & Osram & 2019 & \cite{osram2019qd} \\
+    \hline
+    SALON & Osram & 2019 & \cite{Hoerder2019} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+    \caption{List of sources for \cref{fig:haitz_new}.}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
+    \hline
+    \textit{Metric} & \textit{References} \\
+    \hline
+    Cost & \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} \\
+    \hline
+    Flux & \cite{haitz2007seoul}\cite{haitz2010osram}\cite{haitz2010lumi-1}\cite{haitz2010cree}\cite{haitz2011lumi-1}\cite{haitz2014nichia}\cite{haitz2015cree-1}\cite{haitz2015cree-2}\cite{haitz2018cree}\cite{haitz2019osram}\cite{haitz2019lumi-1}\cite{haitz2019lumi-1}\cite{haitz2020lumi-1}\cite{haitz2013cob}\cite{haitz2016cob}\cite{haitz2017cob}\cite{haitz2019cob}\cite{haitz2019lumi-1}\cite{haitz2019lumi-2} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+    \caption{List of sources for \cref{fgr:subeff}.}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
+    \hline
+    \textit{Panel (Sub-Eff.)} & \textit{References} \\
+    \hline
+    Panel A1 (FVE) & \cite{nichia2001data}\cite{lumi2003lumi}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
+    \hline
+    Panel A2 (Droop) & Data calculated from luminous intensity curves of respective device datasheets: \cite{datasheet_osram_topled}\cite{osram2008data}\cite{osram2008gdplus}\cite{osram2018csp}\cite{datasheet_lumileds_lux1}\cite{lumi2008data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{samsung2018data} \\
+    \hline
+    Panel B1 (IQE) & Data extracted from respective device datasheets and scientific publications (for a complete list, compare \cite{zenodo_weinold_led_history}).
+Additional data was extracted from government reports: \newline
+\cite{doe_ssl_multiyear_2006}\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2009}\cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016} \\
+    \hline
+    Panel B2 (LEE) & \cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012} \newline \cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010}
+\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\clearpage
+\addcontentsline{toc}{section}{Supplementary Methods}
+
+\suppmeth{Domain Definition}
+\label{sec:domain}
+
+\textit{Elsevier Scopus}\footnote{\url{https://www.scopus.com/search/}} advanced search query:
+
+\begin{pseudocode}
+TITLE-ABS-KEY(("light-emitting" OR "light emitting") W/2 diode*)
+AND NOT
+TITLE-ABS-KEY(solar OR violet OR laser)
+AND NOT
+TITLE-ABS-KEY(perovskite OR GaAsP OR AlGaInP OR GaAlAs OR ZnO)
+AND NOT
+TITLE-ABS-KEY(organic OR OLED* OR polymer*)
+AND NOT
+SUBJAREA(ARTS OR BUSI OR DECI OR ECON OR PSYC OR SOCI)
+AND NOT
+SUBJAREA(MEDI OR NURS OR VETE OR DENT OR HEAL OR MULT)
+AND NOT
+SUBJAREA(AGRI OR BIOC OR IMMU OR NEUR OR PHAR)
+AND NOT
+SUBJAREA(envi OR EART OR comp)
+\end{pseudocode}
+
+\clearpage
+
+\suppmeth{Metrics for Quantifying Technological Progress in LEDs}
+
+\textbf{Motivation for Metrics Selection}
 
@@
 
-\section{Background of Interviewed Experts}
+\clearpage
+\suppmeth{Device Performance Metrics}
+\label{sec:device_performance_metrics}
+
+During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
+This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
+In 2000, the highest performing devices yielded around 10 lm, just below the output of a candle as defined in the unit candela (1 cd=12.57 lm) \cite{haitz2011solid}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
+Today, however, the devices with the highest performance yield in excess of 1600 lm, the equivalent of a 100 W incandescent bulb \cite{cree2020bright}.
+LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
+Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving the overall LED efficiency \cite{weinold2021compound}.
+Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
+
+We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe different physical loss channels within the device to describe progress in LED technology.
+In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
+The equations defining these sub-efficiencies are provided below.
+The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
+%
+\begin{equation}
+    \eta_L = \prod_{i=(V_f, \dots, S)} \eta_i
+\end{equation}
+%
+This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
+
+\textbf{Forward Voltage Efficiency $\eta_{V_f}$}
+
+Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
+\cite{tsao2010solid}} (FVE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
+These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
+It is defined as
+%
+\begin{equation}
+    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
+\end{equation}
+%
+where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Light Extraction Efficiency $\eta_{LE}$}
+
+LED light extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
+%
+\begin{equation}
+    \eta_{LE} = \frac{P_{opt}}{P_{int}} = \frac{\text{\# of photons out/s}}{\text{\# of photons created/s}}
+\end{equation}
+%
+where $P_{opt}$ is the optical power emitted from the device and  $P_{int}$ is the optical power emitted from the active region into other parts of the device (internal) \cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Internal Quantum Efficiency $\eta_{IQ}$}
+
+Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is defined as
+%
+\begin{equation}
+    \eta_{IQ} = \frac{\# \text{of photons created/s}}{\# \text{of electron-hole pairs in/s}}
+\end{equation}
+
+\textbf{Droop $\eta_{droop}$}
+
+Efficiency droop describes the decrease of LED internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
+This energy loss channel is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
+It is defined as
+%
+\begin{equation}
+    \eta_{droop} = 1 - \frac{\eta_{IQ}}{\eta_{IQ}(A \rightarrow 0)}
+\end{equation}
+%
+where $\eta_{IQ}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
+In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$
+%
+\begin{equation}
+\label{eqn:droop}
+    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
+\end{equation}
+%
+Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Light Conversion Efficiency $\eta_{C}$}
+
+Light conversion efficiency (LCE) describes the sum of all losses in the process of converting the blue light that is the basis for all phosphor-converted white LEDs.
+Specifically, energy is lost due to three distinct effects.
+The first effect is the inherent wavelength-conversion loss or \textit{Stokes shift} as the difference between the original blue wavelength and the down-converted red, yellow or green wavelength.
+The second effect are non-radiative processes in the phosphor material.
+The third , scattering and absorption processes in the phosphor material.
+These three effects are listed individually in Figure 4.
+The overall light conversion efficiency is defined as
+%
+\begin{equation}
+    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor[RGB]{225, 180, 0}{Y},\textcolor{teal}{G}} E_i}
+\end{equation}
+%
+where $E_{\textcolor{blue}{B}}$ is the total energy of blue light before down-conversion and $E_i$ denotes the total energy of light at color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
+Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
+\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor[RGB]{225, 180, 0}{Y}ellow, \textcolor{teal}{G}reen.
+
+\textbf{Spectral Efficiency $\eta_{S}$}
+
+Spectral efficiency (SE) describes losses in the light down conversion process due to the wavelength-dependent efficiency of the human eyes.
+Photons converted into the infrared or ultraviolet light are lost to illumination purposes.
+It is defined as
+%
+\begin{equation}
+    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
+\end{equation}
+%
+where $K$ is the luminous efficacy of radiation of the light source, defined in \cref{eqn:ler}, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
+$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering index performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
+
+\textbf{Luminous Efficacy of Radiation}
+\label{subsec:ler}
+
+The luminous efficacy \textit{of radiation} $K$ is a supplementary metric used in our calculations of SE, rather than directly in our analysis of the improvements in LED performance.
+Care must be taken not to confuse this metric with the luminous efficacy of a light source, described in \cref{subsec:les}.
+
+The luminous efficacy \textit{of radiation} $K$ describes the match of the spectrum of a light-emitting diode to the human visual system.
+Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
+A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
+The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source $\phi$ with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
+%
+\begin{equation}
+\label{eqn:ler}
+    K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
+\end{equation}
+%
+where
+%
+\begin{align*}
+    K &\dots \text{luminosity function} \\
+    \phi &\dots \text{spectral radiant flux} \\
+    \lambda &\dots \text{wavelength}
+\end{align*}
+%
+This metric can be computed from spectral data alone and does not require additional spectral normalization.
+It enables straightforward comparison between the performance of different downconversion phosphors, as shown in the top panel of \cref{fig:phosphor_spectrum}.
+Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation.
+
+\textbf{Luminous Efficacy of Source}
+\label{subsec:les}
+
+The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux $\phi$ and the consumed electrical power $P_{el}$ \cite{cie-term-effsrc}.
+%
+\begin{equation}
+    \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
+\end{equation}
+%
+This metric is also frequently cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
+However, as noted in \cref{subsec:ler}, care must be taken not to confuse this efficacy metric with the \textit{luminous efficacy of radiation}, which depends only on the spectral characteristics of a light source.
+As the luminous efficacy of a light source $\eta$ captures the overall device efficacy, it depends on a large number of other device properties and parameters.
+This makes attribution of changes in this metric to individual changes in device design or manufacturing difficult.
+For this reason, we do not use this metric only to show historical progress across different lighting technologies in Figure 1 in the main text of the article.
+
+\clearpage
+\suppmeth{Consumer Experience Metrics}
+\label{sec:metrics_consumer}
+
+The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
+Any metric relevant to customer experience can thus be calculated from the spectrum alone.
+The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra of the down-conversion phosphor used in the device.
+It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
+Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: colour rendering index (CRI) and colour temperature (CT).
+We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
+This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
+As a result, it is beyond the scope of this work.
+
+\textbf{Color Rendering Index (CRI)}
+
+The colour rendering index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
+The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
+CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
+However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DoE2016LED}.
+High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
+It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
+However, these niche applications remain outside our focus on general illumination.
+Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
+
+\textbf{Colour Temperature}
+
+The colour temperature (CT) of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
+Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
+Early white LEDs produced only cool white light \cite{mueller2000light}.
+The introduction of first commercial warm white LED light bulbs played a significant role in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
+For this reason, we also adopt CT as a metric for tracking progress in consumer experience in LED lighting.
+
+\textbf{Cost Metrics}
+
+In selecting the metrics for tracking progress in manufacturing cost reductions in LED, we must first highlight the complexity of this task.
+Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
+Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
+However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
+In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
+Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress difficult in practice.
+
+We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
+In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
+
+We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:costmodel_details}.
+The outcome of the cost model is the cumulative yielded manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
+
+\clearpage
+
+\suppmeth{Background of Interviewed Experts}
 
@@
 \clearpage
-\section{Metrics: Device Performance}
-\label{sec:device_performance_metrics}
-
-During the years following the market introduction of white LEDs, the primary metric of progress in solid-state lighting was typically luminous flux.
-This was because the luminous flux (total brightness) of early white LEDs was too small to allow for the economical combination of multiple LEDs into lamps for general illumination purposes.
-In 2000, the highest performing devices yielded around 10 lm, just below the output of a candle as defined in the unit candela (1 cd=12.57 lm) \cite{haitz2011solid}.
-Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
-Today, however, the devices with the highest performance yield in excess of 1600 lm, the equivalent of a 100 W incandescent bulb \cite{cree2020bright}.
-LED brightness has thereby become sufficient to enable the construction of lamps from multiple LED devices, with contemporary improvements focusing instead on higher efficiency and quality of light instead of brightness.
-Even though a large number of scientific publications and industry periodicals continue to focus on brightness as a metric of progress in lighting, we find that, at this point in time, this metric is insufficient to capture the complexity of the multitude of efficiency improvements that have been driving the overall LED efficiency \cite{weinold2021compound}.
-Specifically, the stagnating levels of luminous flux in the highest performing devices do not capture the outcomes of a major area of LED research, which is improving electrical efficiency at constant brightness.
-
-We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe different physical loss channels within the device to describe progress in LED technology.
-In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
-The equations defining these sub-efficiencies are provided below.
-The overall efficiency ("Lamp Efficiency") $\eta_L$ of a light-emitting diode package is the product of all considered sub-efficiencies:
-%
-\begin{equation}
-    \eta_L = \prod_{i=(V_f, \dots, S)} \eta_i
-\end{equation}
-%
-This metric describes the cumulative electrical and optical losses within the device, as well as the light conversion losses in the phosphor layer.
-
-\textbf{Forward Voltage Efficiency $\eta_{V_f}$}
-
-Device forward voltage efficiency\footnote{"Joule Efficiency" $\epsilon_J$ in Tsao et al.
-\cite{tsao2010solid}} (FVE) describes all electrical losses at the interface between the electrodes and the semiconductor and in the bulk.
-These losses can be due to tunneling and Ohmic resistance at the interface, as well as Ohmic resistance and other electrical losses within the bulk of the semiconductor.
-It is defined as
-%
-\begin{equation}
-    \eta_{V_f} = \frac{E_{h\nu}}{V_f}
-\end{equation}
-%
-where $E_{h\nu}$ denotes the photon energy and $V_f$ is the diode forward voltage \cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Light Extraction Efficiency $\eta_{LE}$}
-
-LED light extraction efficiency (LEE) describes the  losses due to absorption in the material after electron-hole recombination and the associated emission of a photon. It is defined as
-%
-\begin{equation}
-    \eta_{LE} = \frac{P_{opt}}{P_{int}} = \frac{\text{\# of photons out/s}}{\text{\# of photons created/s}}
-\end{equation}
-%
-where $P_{opt}$ is the optical power emitted from the device and  $P_{int}$ is the optical power emitted from the active region into other parts of the device (internal) \cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Internal Quantum Efficiency $\eta_{IQ}$}
-
-Internal quantum efficiency (IQE) describes non-radiative recombinatory processes in the semiconductor bulk. It is defined as
-%
-\begin{equation}
-    \eta_{IQ} = \frac{\# \text{of photons created/s}}{\# \text{of electron-hole pairs in/s}}
-\end{equation}
-
-\textbf{Droop $\eta_{droop}$}
-
-Efficiency droop describes the decrease of LED internal quantum efficiency at high current densities, which is caused by a number of different physical effects\cite{David2020}.
-This energy loss channel is often treated separately from internal quantum efficiency in the literature due to its importance in high-power devices.
-It is defined as
-%
-\begin{equation}
-    \eta_{droop} = 1 - \frac{\eta_{IQ}}{\eta_{IQ}(A \rightarrow 0)}
-\end{equation}
-%
-where $\eta_{IQ}(A \rightarrow 0)$ denotes the internal quantum efficiency at low current densities.
-In practice, droop is often given as the percentage difference between the ideal luminous intensity curve $\phi_{ideal}$ and the real luminous intensity curve $\phi$ at a set diode test current $A_{test}$
-%
-\begin{equation}
-\label{eqn:droop}
-    D = \frac{\phi_{ideal}(A_{test})-\phi(A_{test})}{\phi_{ideal}(A_{test})/100}
-\end{equation}
-%
-Therefore, according to this definition, a lack of droop corresponds to a droop efficiency of $\eta_{droop} = 100\%$\cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Light Conversion Efficiency $\eta_{C}$}
-
-Light conversion efficiency (LCE) describes the sum of all losses in the process of converting the blue light that is the basis for all phosphor-converted white LEDs.
-Specifically, energy is lost due to three distinct effects.
-The first effect is the inherent wavelength-conversion loss or \textit{Stokes shift} as the difference between the original blue wavelength and the down-converted red, yellow or green wavelength.
-The second effect are non-radiative processes in the phosphor material.
-The third , scattering and absorption processes in the phosphor material.
-These three effects are listed individually in Figure 4.
-The overall light conversion efficiency is defined as
-%
-\begin{equation}
-    \eta_{C} = \frac{E_{\textcolor{blue}{B}}}{\sum_{i=\textcolor{red}{R},\textcolor{orange}{O},\textcolor[RGB]{225, 180, 0}{Y},\textcolor{teal}{G}} E_i}
-\end{equation}
-%
-where $E_{\textcolor{blue}{B}}$ is the total energy of blue light before down-conversion and $E_i$ denotes the total energy of light at color $i$ corresponding to the down-converted photon wavelength \cite{schubert2018light}\cite{tsao2010solid}.
-Since every individual phosphor component in the device has its own associated conversion losses, the denominator sums over all phosphor components, i.e.
-\textcolor{red}{R}ed, \textcolor{orange}{O}range, \textcolor[RGB]{225, 180, 0}{Y}ellow, \textcolor{teal}{G}reen.
-
-\textbf{Spectral Efficiency $\eta_{S}$}
-
-Spectral efficiency (SE) describes losses in the light down conversion process due to the wavelength-dependent efficiency of the human eyes.
-Photons converted into the infrared or ultraviolet light are lost to illumination purposes.
-It is defined as
-%
-\begin{equation}
-    \eta_{S} = \frac{K}{K_{max}(CRI,CCT)}
-\end{equation}
-%
-where $K$ is the luminous efficacy of radiation of the light source, defined in \cref{eqn:ler}, which can be computed from the device spectrum and the luminosity function, which describes the sensitivity of the human eye.
-$K_{max}$ is the maximum luminous efficacy of radiation of a perfect light source with the same color rendering index performance and correlated color temperature as the light source in question\cite{schubert2018light}\cite{tsao2010solid}.
-
-\textbf{Luminous Efficacy of Radiation}
-\label{subsec:ler}
-
-The luminous efficacy \textit{of radiation} $K$ is a supplementary metric used in our calculations of SE, rather than directly in our analysis of the improvements in LED performance.
-Care must be taken not to confuse this metric with the luminous efficacy of a light source, described in \cref{subsec:les}.
-
-The luminous efficacy \textit{of radiation} $K$ describes the match of the spectrum of a light-emitting diode to the human visual system.
-Efficacy in lighting is dependent on the luminosity function, which describes the wavelength-dependent sensitivity of the human eye.
-A light source emitting very \textit{efficiently} in the infrared yet emitting no visible light has a very low \textit{efficacy}.
-The luminous efficacy of radiation $K$ is mathematically defined as the normalized, integrated product of the spectral radiant flux of a light source $\phi$ with the wavelength-dependent human sensitivity to light \cite{cie-term-effrad}
-%
-\begin{equation}
-\label{eqn:ler}
-    K [\text{lm/W}_{opt}]= \frac{\int_0^\infty K( \lambda ) \phi \text{d} \lambda}{\int_0^\infty \phi \text{d} \lambda}
-\end{equation}
-%
-where
-%
-\begin{align*}
-    K &\dots \text{luminosity function} \\
-    \phi &\dots \text{spectral radiant flux} \\
-    \lambda &\dots \text{wavelength}
-\end{align*}
-%
-This metric can be computed from spectral data alone and does not require additional spectral normalization.
-It enables straightforward comparison between the performance of different downconversion phosphors, as shown in the top panel of \cref{fig:phosphor_spectrum}.
-Light sources emitting in the far red or blue part of the spectrum have lower efficacy of radiation.
-
-\textbf{Luminous Efficacy of Source}
-\label{subsec:les}
-
-The luminous efficacy \textit{of a light source} $\eta$ is defined as the ratio between the emitted luminous flux $\phi$ and the consumed electrical power $P_{el}$ \cite{cie-term-effsrc}.
-%
-\begin{equation}
-    \eta [\text{lm/W}_{el}]= \frac{\phi}{P_{el}}
-\end{equation}
-%
-This metric is also frequently cited in device datasheets, scientific literature and textbooks when describing the performance of light-emitting diodes.
-However, as noted in \cref{subsec:ler}, care must be taken not to confuse this efficacy metric with the \textit{luminous efficacy of radiation}, which depends only on the spectral characteristics of a light source.
-As the luminous efficacy of a light source $\eta$ captures the overall device efficacy, it depends on a large number of other device properties and parameters.
-This makes attribution of changes in this metric to individual changes in device design or manufacturing difficult.
-For this reason, we do not use this metric only to show historical progress across different lighting technologies in Figure 1 in the main text of the article.
-
-\clearpage
-\section{Metrics: Consumer Experience}
-\label{sec:metrics_consumer}
-
-The perceived quality of light is entirely determined by the emission spectrum of a light source \cite{ies_handbook}.
-Any metric relevant to customer experience can thus be calculated from the spectrum alone.
-The spectrum of an LED light source is determined by the emission wavelength of the LED itself and the absorption and emission spectra of the down-conversion phosphor used in the device.
-It is typically included in the product datasheets provided by manufacturers, which enables the calculation of all relevant spectrum-based metrics for these devices.
-Based on the prevalence in scholarly literature and industry publications, for this study we choose two consumer experience metrics: colour rendering index (CRI) and colour temperature (CT).
-We do not consider flicker, the unintended high-frequency temporal modulation of light, which is another important consumer experience metric for lighting and a subject of recent regulation by the European Union \cite{weinold2020long}.
-This effect is caused not by LEDs themselves, but rather by inadequately designed electrical ballasts \cite{Lehman2014}.
-As a result, it is beyond the scope of this work.
-
-\textbf{Color Rendering Index (CRI)}
-
-The colour rendering index (CRI) of a light source describes its ability to render the colours of an object faithfully when compared to illumination under a reference light source, such as standard daylight \cite{khan2015led}.
-The way it is calculated is defined by the International Illumination Commission (CIE) \cite{cie_cri_1995}.
-CRI has certain limitations when applied to solid-state light sources \cite{david2013cri}.
-However, despite repeated attempts at constructing more elaborate colour rendering metrics \cite{Houser2013}, CRI has remained the de facto industry standard for describing colour fidelity of light sources \cite{DoE2016LED}.
-High colour rendering performance of lighting is a requirement in workplace environments, retail stores, clinical operating environments and art exhibitions \cite{khanh2017color}.
-It should be noted that some niche applications prioritize high colour saturation over high CRI, for instance, in food display or fabric retail applications \cite{david2013cri}.
-However, these niche applications remain outside our focus on general illumination.
-Due to a broad availability and importance of CRI data for consumers of various LED lighting sources, we adopt CRI as the key metric to track progress in consumer experience in LED lighting, despite its limitations.
-
-\textbf{Colour Temperature}
-
-The colour temperature (CT) of a light source describes the equivalent temperature of an ideal black body which emits light of a colour comparable to that of the light source \cite{commission2011cie}.
-Warm white light sources are widely used in general illumination, while cold white light sources are used in workplace illumination and outdoor lighting.
-Early white LEDs produced only cool white light \cite{mueller2000light}.
-The introduction of first commercial warm white LED light bulbs played a significant role in increasing adoption of LED-based lighting among consumers, as their spectrum more closely resembled the warm white colour temperature of incandescent light bulbs \cite{al2016optics}.
-For this reason, we also adopt CT as a metric for tracking progress in consumer experience in LED lighting.
-
-\textbf{Cost Metrics}
-
-In selecting the metrics for tracking progress in manufacturing cost reductions in LED, we must first highlight the complexity of this task.
-Access to manufacturing cost data at the chip level is usually restricted due to its proprietary nature and is available only for selected products.
-Using sales price information instead of the cost for the same purpose seems a promising alternative, as prices can be easily obtained directly from manufacturers for current products.
-However, historical data on prices for different chip architectures and different years is similarly difficult to obtain.
-In addition, sales price includes components not relevant to progress in the technology, such as profit margins and overhead costs, and is affected by policies such as rebates and purchase subsidies.
-Historical information on these factors and price components, as well as their impact on the manufacturing cost for each LED product under consideration is even harder to obtain, making the use of LED sales prices as a direct metric of technology progress difficult in practice.
-
-We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
-In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
-
-We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:costmodel_details}.
-The outcome of the cost model is the cumulative yielded manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
-
-\clearpage
-\section{Methods: Cost Model Structure}
+\suppmeth{Cost Model Structure and Limitations}
 \label{sec:costmodel_structure}
@@
 \clearpage
-\section{Methods: Cost Model Limitations}
-\label{sec:costmodel_limitations}
 
@@
 
-\section{Diagrams: Chip Architecture Manufacturing Process Steps}
-\label{sec:manufacturing_steps}
-
-\cref{fig:manuf_classical_2003}-\cref{fig:manuf_csp_2020-2} show a simplified rendering for the manufacturing process of three different chip architectures considered in the cost model.
-
-\vspace{-20mm}
-
-    %CLASSICAL CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=555pt]{./figures/classical_overview_2003.pdf}
-            \caption{\textbf{Manufacturing process for a classical LED package with lateral current spreading, circa 2003.} MQW - multiple quantum well; PLLC - plastic leaded chip carrier.}
-        \label{fig:manuf_classical_2003}
-        \end{figure}
-    \end{landscape}
-
-    %VERTICAL THIN-FILM FLIP-CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/vtf_overview_2012-1.pdf}
-            \caption{\textbf{Manufacturing process for a vertical thin-film  package flip-chip LED chip with vertical current spreading, circa 2012.
-(1/2)} MQW - multiple quantum well; LLO - laser lift-off.
-Continued in \cref{fig:manuf_vtf_2012-2}}
-            \label{fig:manuf_vtf_2012-1}
-        \end{figure}
-    \end{landscape}
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/vtf_overview_2012-2.pdf}
-            \caption{\textbf{Manufacturing process for a vertical thin-film package flip-chip LED chip with vertical current spreading, circa 2012 (2/2)}.
-Continued from \cref{fig:manuf_vtf_2012-1}.
-Encaps.
-- encapsulation.}
-            \label{fig:manuf_vtf_2012-2}
-        \end{figure}
-    \end{landscape}
-
-    % CHIP-SCALE PACKAGE FLIP-CHIP
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/csp_overview_2020-1.pdf}
-            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
-(1/2)} MQW - multiple quantum well.
-Continued in \cref{fig:manuf_csp_2020-2}.}
-            \label{fig:manuf_csp_2020-1}
-        \end{figure}
-    \end{landscape}
-
-    \begin{landscape}
-        \begin{figure}
-            \includegraphics[width=595pt]{./figures/csp_overview_2020-2.pdf}
-            \caption{\textbf{Manufacturing process for a chip scale package flip-chip LED chip with vertical current spreading, circa 2020.
-(2/2)}.
-Continued from \cref{fig:manuf_csp_2020-1}.
-Encaps.
-- encapsulation.}
-            \label{fig:manuf_csp_2020-2}
-        \end{figure}
-    \end{landscape}
-
-\section{Cost Model: Computation of Yielded Cost}
+\suppmeth{Yielded Cost}
 \label{sec:cost_computation}
@@
 
-\section{Cost Model: Contribution of Variables}
+\suppmeth{Contribution of Variables}
 \label{sec:contribution_variables}
@@
 
-\newpage
-\section{Cost Model: Sensitivity Analysis}
+\clearpage
+
+\addcontentsline{toc}{section}{Supplementary Discussion}
+
+\suppdisc{Cost Model Sensitivity Analysis}
 
@@
 \clearpage
-\section{Cost Model: Comparison with Industry Data/Projections}
+\suppdisc{Cost Data Comparison with Industry Projections}
 \label{sec:cost_model_calibration}
@@
 \clearpage
-\section{Historical Progress in Sub-Efficiencies}
-
-Empirical data on the historical progress in metrics related to four different sub-efficiencies contributing to the overall LED efficiency (lamp efficiency $\eta_L$) is shown in \cref{fgr:subeff}:
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=15.5cm]{./figures/subefficiencies_progress.pdf}
-\caption{\textbf{Historical improvements in LED sub-efficiencies.} Panels A1-A2 show data for key metrics used to calculate forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
-Panels B1-B2 show data for internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
-Source: own synthesis of data from the list of sources provided in \cref{sec:sources}.
-Data on average device performance adapted from U.S.
-Department of Energy (DOE) Reports \cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_rnd_2016}\cite{doe_ssl_rnd_2018} and an Osram company report \cite{beale_leds_2015}.}
-\label{fgr:subeff}
-\vspace{-20mm}
-\end{figure}
-
-\clearpage
-
-\textbf{Panel A1:} This panel shows device forward voltage at a test current of I=350mA, categorized by chip architecture.
-The physical limit for a blue light wavelength of 450nm without electric pumping is shown for reference.
-Forward voltage efficiency is computed from this metric as described in \cref{sec:device_performance_metrics}.
-Data points for devices released in 2020 by various manufacturers are shown in an inset plot for additional context.
-Note that differences between chip architectures in this plot have become small for all but one company.
-Chip architecture abbreviations: Acronyms: TFFC - thin-film flip-chip, FC - flip-chip, CSP-FC - chip-scale package flip-chip.
-Company abbreviations: LG - LG Innotek, SE - Seoul Semiconductor, SA - Samsung, OS - Osram, LL - Lumileds.
-\
-
-\textbf{Panel A2:} Efficiency droop at the test current of I=350mA, categorized by chip architecture.
-Droop efficiency is computed from this metric as described in \cref{sec:device_performance_metrics}.
-Data has been calculated from luminous intensity curves of respective device datasheets.
-\
-
-\textbf{Panel B1:} Internal quantum efficiency (IQE) for categorized by type of measurement used.
-Because it is inherently more difficult to measure IQE than is the case for other sub-efficiencies, this panel shows data categorized by different measurement methods.
-This is not the case for other panels of this figure.
-Note that the bump in the DOE Average around 2013 is due to a change in IQE definition laid out in \cite{doe_ssl_multiyear_2013}.
-Measurement methods abbreviations: PL - Photoluminescence \cite{Shim_2018}, EL - Electroluminescence \cite{Getty_2009}; ABC Model \cite{Karpov_2014}, Ideality Factor \cite{Masui_2010}.
-“Not reported” denotes data points where the IQE measurement technique was not reported.
-Whiskers indicate reported ranges, where available.
-\
-
-\textbf{Panel B2:} Light extraction efficiency (LEE) for different chip architectures by type of specific technology used to improve LEE.
-The sources for all data points are shown for reference: Simulation - ray-tracing computer simulation; Experimental (absolute) – direct LEE values; Experimental (relative) - relative improvement over baseline chip architecture.
-Abbreviations: TFFC - Thin-Film Flip-Chip; FC - Flip-Chip; CSP - Chip-Scale Package; ITO - Indium Tin Oxide; ZnO - Zinc Oxide; PSS - Patterned Sapphire Substrate; electr.
-- electrode; text.
-- textured.
-
-\textbf{Additional Notes on \cref{fgr:subeff}}
-
-Data illustrated in \cref{fgr:subeff} was used to determine the performance of best-in-class devices across all sub-efficiencies. The results are shown in Figures 4-6 in the main text.
-
-\textbf{Additional Data on Phosphor Performance}
-
-Note that \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.
-Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the (mostly) empirical \cref{fgr:subeff} not relevant.
-Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum} below (see also Methods section in the main text).
-
-\clearpage
-\section{Non-Phosphor-Related LED Innovations}
-
-\cref{tab:innovations_1}-\cref{tab:innovations_2} summarize LED innovations and technology and manufacturing process improvements identified in this study as affecting four white LED sub-efficiencies: forward voltage efficiency, light extraction efficiency, internal quantum efficiency, droop and light conversion efficiency.
-Note that this table does not include phosphor-related LED innovations affecting spectral efficiency, as these are presented separately in Table 1 in the main text as affecting consumer experience metrics and discussed in detail in \cref{sec:innovation_phosphor}.
-
-\begin{table}[H]
-    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note this table is continued in \cref{tab:innovations_2}.}
-    \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
-        \hline
-        \multicolumn{5}{c}{Forward Voltage Efficiency (FVE) $\eta_{Vf}$} \\
-        \hline
-            \textit{Year} & \textit{Innovation} & \textit{Area of Improvement} & \textit{Spillover} & \textit{Source} \\
-        \hline
-            1999 & Indium Tin Oxide \newline Current spreading layer & Contact resistance & Yes & \cite{margalith1999indium}\\
-        \hline
-            1998 & Digitated electrodes & Contact resistance & No & \cite{steigerwald2001electrode} \\
-        \hline
-            Ongoing & Epitaxy improvements \newline and better doping & Polarization and \newline bulk resistance & No & I \\
-        \hline
-            1998 & Silver p-Contacts & Contact resistance & No & \cite{kondoh2001nitride} \\
-        \hline
-        \multicolumn{5}{c}{Light extraction efficiency (LEE) $\eta_{LE}$} \\
-        \hline
-            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
-        \hline
-            $<$ 2003 & Optimization for \newline cavity effects & Reduces self-interference of quantum well & No & I, \cite{Shen2003} \\
-        \hline
-            1993 & Chip surface randomization & Total reflection and absorption & No & I, \cite{bergh1973surface}\cite{Schnitzer1993} \\
-        \hline
-            1993 & Thin-film chip architecture & Absorption & No & I, \cite{Schnitzer1993} \\
-        \hline
-            1996 & Patterned sapphire substrate  & Total reflection and absorption & Yes & I, \cite{Tadatomo2001} \\
-        \hline
-            Ongoing & Chip design for high LEE & Total reflection and absorption & No & I, \cite{Haerle2004} \\
-        \hline
-            $\sim$2000 & Silver p-contacts & Absorption & No & I, \cite{kondoh2001nitride} \\
-        \hline
-        \multicolumn{5}{c}{Internal quantum efficiency (IQE) $\eta_{IQ}$} \\
-        \hline
-            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
-        \hline
-            1994 & Double heterostructure & Higher radiative recombination probability & No & \cite{Nakamura1994} \\
-        \hline
-            1996 & Multiple quantum well & Higher radiative recombination probability & No & \cite{Koike1996} \\
-        \hline
-            Ongoing & Active region \newline Doping & Radiative recombination probability & No & I, \cite{schubert2018light} \\
-        \hline
-            Ongoing & Epitaxy \newline Improvements & Radiative recombination probability & No & I \\
-        \hline
-            Ongoing & Chip architecture \newline Improvements & Radiative recombination probability & No & I \\
-        \hline
-        \end{NiceTabularX}
-        \vspace{2mm}
-    \caption*{Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
-'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
-\textit{Spillover} column indicates if the innovation involved technology spillovers.
-\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
-    \vspace{-20mm}
-    \label{tab:innovations_1}
-\end{table}
-
-\begin{table}[H]
-    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note that this table continues \cref{tab:innovations_1}.}
-    \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
-    \hline
-    \multicolumn{5}{c}{Droop $\eta_{droop}$} \\
-        \hline
-            ongoing & Lower current Density & Current density & No & I \\
-        \hline
-            ongoing & Epitaxy \newline Improvements & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
-        \hline
-            ongoing & Chip Architecture \newline Improvements  & $\ominus/\oplus$ Distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
-        \hline
-            $<$ 2017 & Defect getting \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
-        \hline
-        \multicolumn{5}{c}{Light conversion efficiency (CE) $\eta_{C}$} \\
-        \hline
-            \textit{Year} & \textit{Innovation} & \textit{Area of improvement} & \textit{Spillover} & \textit{Source} \\
-        \hline
-            Ongoing & Lower current density & Current density & No & I \\
-        \hline
-            Ongoing & Epitaxy \newline Improvements & Charge distribution \newline Defect density & No & I, \cite{bhardwaj2016progress} \\
-        \hline
-            Ongoing & Chip architecture \newline improvements  & Charge distribution \newline Defect density & No & I, \cite{Wildeson2017} \\
-        \hline
-            $<$ 2017 & Defect gettering \newline underlayer & Defect density & No & I, \cite{haller2017burying} \\
-        \hline
-        \end{NiceTabularX}
-        \vspace{2mm}
-    \caption*{Note: \textit{Year} column indicates the first instance of application of corresponding invention in white LEDs.
-'Ongoing' indicates improvements that are incremental in nature and have been ongoing since the earliest days of LED manufacturing, with no individual breakthroughs identified.
-\textit{Spillover} column indicates if the innovation involved technology spillovers.
-\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
-    \label{tab:innovations_2}
-\end{table}
-
-\clearpage
-\section{Phosphor-Related LED Innovations}
-\label{sec:innovation_phosphor}
-
-This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics.
-These include a detailed description of the history of those innovations, which we use to identify the details of involved technology spillovers.
-In addition, spectral data for the first identified LED products that used those innovations is presented in \cref{fig:phosphor_spectrum}, which we use to calculate consumer experience metrics, as well as LED spectral efficiency.
-
-\textbf{YAG and YGAG Phosphors}
-
-Prior to Nakamura’s invention of highly efficient blue LEDs in the 1990s, Japan’s Nichia Corporation had not sold commercially successful semiconductor products, instead specializing in phosphors for cathode ray tubes (CRT) and fluorescent lamps \cite{nakamura2013blue}  .
-Nevertheless, extensive firm expertise in this area helped Nichia’s Yoshinori Shimizu formulate the principles of using CRT phosphors to convert blue light from Nakamura’s LEDs into white light in 1994 \cite{shimizu1994sheet}\cite{cho2017white}.
-By 1996, Shimizu and his colleagues developed \cite{bando1996}\cite{shimizu1999light} the first practical LED application of a well-known Yttrium Aluminium Garnet (YAG) CRT phosphor activated with cerium \cite{blasse1967new}, enabling the first commercial white LED products manufactured and sold by Nichia since late 1996 \cite{bando1998development}\cite{cho2017white}.
-
-Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 6 in the main article).
-An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG).
-Used in combination with red-emitting sulfide phosphors, by 2003 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}.
-However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased.
-New chemically stable and non-toxic red phosphors were needed.
-
-\textbf{258, SLA and SALON phosphors}
-
-In 1997, Hubert Huppertz and Wolfgang Schnick, working at the University of Bayreuth in Germany, synthesized the first compound in a new class of rare earth nitridosilicate materials \cite{Huppertz1997} later dubbed “258” due to a proportion of elements in its chemical formula.
-The luminescent properties of these materials were identified by the Schnick’s group, by then at Ludwig-Maximilians University of Munich, in 2000 \cite{Hppe2000} after a suggestion made to Schnick at a conference following earlier reports of good luminosity properties of europium-doped compounds \cite{Qiua1998}.
-U.S.-based LED manufacturer Lumileds applied for a patent for the first class of red LED phosphors based on the 258 nitridosilicate chemistry in 2003 \cite{mueller2004phosphor}.
-The first use of the 258 phosphor in a commercial “Luxeon” LED package was then reported in a joint publication co-authored by inventors from Lumileds and researchers from the Schnick’s group in 2005 \cite{MuellerMach2005}.
-
-Further efforts in LED phosphor development were directed towards synthesizing a red narrow-band phosphor.
-Narrow LED emission peak widths yield the highest luminous efficacy of radiation, as in this case less light is emitted in the far-red range of the spectrum in which the human eye is not very sensitive.
-After synthesizing several narrow-band phosphors emitting in yellow \cite{Hppe2004} and cyan \cite{Kechele2009}, the Schnick’s group identified the local cubic cation coordination structure of the cyan phosphor compound as the reason for its narrow band width \cite{lumi2016narrow}.
-A search for a structurally analogous nitride compound with the narrow red instead of the cyan emission was undertaken.
-After several unsuccessful attempts, the sought-after cuboidal nitride compound was found in a 2008 publication led by Francis DiSalvo \cite{Park2008Sr}.
-Based on information provided in this work, Schnick and colleagues synthesized and studied the spectral properties of a new narrow band red SLA phosphor in 2013 \cite{schmidt2013new}\cite{Pust2014}\cite{schmidt2017phosphors}.
-The material was introduced in commercial LED devices by Lumileds in 2015 \cite{lumi2016narrow_whitepaper}.
-
-The most recent red narrow-band phosphor innovation included in Table 1 and Figure 6 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}.
-The first U.S.
-patent application for this phosphor was filed in 2017 \cite{seibald2019phosphor}.
-The SALON phosphor is a derivate of the SLA phosphor.
-Therefore, it is the only innovation related to consumer experience metrics identified in our study that seemingly not involved technology spillovers.
-
-\textbf{PFS Phosphor}
-
-Down conversion with ultra-narrow-band phosphor can achieve the highest spectral efficiency.
-However, few such phosphors have been identified, with even less exhibiting desirable material properties such as thermal stability \cite{Phillips2007}.
-The first commercially successful ultra-narrow-band red phosphor was developed by General Electric (GE).
-It is based on a potassium fluorosilicate (PFS) compound activated with manganese ions.
-Its luminescence was first recorded by Adrian Paulusz at GE in 1972 \cite{paulusz1973efficient}.
-In the early 2000s, while searching for potential new LED phosphor materials for GE’s lighting business at GE Lumination, Emil Radkov rediscovered Paulusz’s findings in the literature.
-Following extensive research on PFS chemical synthesis and material properties conducted in collaboration with the University of Sofia in Bulgaria, Radkov’s Alma Mater, the PFS phosphor had been under development at GE since 2005 \cite{radkov2006red}\cite{radkov2009red}.
-This work, supported by public funding from the U.S.
-Department of Energy (DOE) Solid-State Lighting program \cite{doesslprogram}, resulted in a series of critical improvements in the PFS phosphor properties \cite{Setlur2010}\cite{lyons2012color}, eventually enabling its commercialization under the “TriGain” brand in 2015 \cite{trigain_spectrum}\cite{setlur2015trigain}\cite{Murphy2015}.
-
-\textbf{Quantum Dots for Light Down-Conversion}
-
-Quantum dots (QD) are semiconductor nanocrystals whose quantum size effects make QDs behave as “artificial atoms”.
-Semiconductor quantum dots were first synthesized in the Soviet Union in 1981 \cite{ekimov1981quantum} and at Bell Labs in the U.S.
-in 1983 \cite{Rossetti1983}.
-Luminescent properties of quantum dots were first empirically observed in 1984 \cite{fojtik1984photo} and extensively studied in the early 1990s.
-The key feature of QD luminescence discovered in those studies is that its colour is determined by the QD particle size, making it possible to create pure monochromatic blue, green and red light sources just by tuning the QD size.
-The first application of QDs in LEDs was reported in 1994 in an electroluminescent hybrid QD-polymer LED.
-However, this LED type could not be used in general illumination due to its very low luminous efficacy.
-An alternative application of QDs as a kind of a “phosphor” for light down conversion from an LED light source was proposed in the early 2000s as part of the U.S.
-Department of Energy (DOE)-funded “A Revolution in Lighting“ project at Sandia National Laboratory \cite{simmonsfinal}.
-This concept was successfully demonstrated by Sandia researchers on a commercial LED in 2003 \cite{shea_rohwer_development_2004}\cite{noauthor_sandia_nodate} and was swiftly taken up and advanced further by a group in Taiwan \cite{Chen_2005}\cite{Hsueh_Shih_Chen_2006} The first commercial application of QDs in an LED lamp was brought about by a collaboration between an MIT-born startup QD Vision and the U.S.-based luminaire manufacturer Nexxus Lighting in 2009 \cite{ledprof_nexxusqd}, \cite{bourzac2013quantum}.
-However, rapid advances in the spectral and conversion performance of down-conversion phosphors and high manufacturing cost of quantum dots resulted in the discontinuation of this product.
-After finding market success in display backlighting first demonstrated by Samsung in 2010 \cite{Jang2010} and commercialized by QD Vision in Sony television sets in 2013 \cite{bourzac2013quantum}, QDs returned to the general lighting market in products offered by Lumileds \cite{noauthor_global_2017}\cite{noauthor_quantum_2020} around 2017 and Osram in 2019 \cite{osramqdots} in the form of mid-power LED packages that combined QDs with traditional phosphors for light down conversion.
-
-\section{Spectral Data for Phosphors}
-
-\begin{figure}[H]
-	\centering
-    \includegraphics[width=0.9\textwidth]{./figures/phosphor_spectrum-comparison.pdf}
-	\caption{\textbf{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data.} Corresponding LED products that used phosphor innovations listed in Table 1 in the main article are each indicated by the phosphor label and spectral data publication year.
-Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 6 in the main article.
-The desirable direction of improvements towards higher LER at higher CRI is indicated by a red arrow.
-Metrics were calculated from spectral data shown in the bottom panels using the \texttt{colour-science} package for Python \cite{colour-science_software}.
-Bottom two rows of panels: spectral data for the first identified LED products that used phosphor innovations listed in Table 1 in the main article.
-The luminosity function \cite{cie-term-lumeff}, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel.
-Peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel.
-Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: ?
--  other parts of phosphor mixture not disclosed, G$^*$ - \ch{CaSrS:Eu^{2+}}, Y$^\star$ - $\beta-$SiAlON:Eu$^{2+}$, G$^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, O$^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}.
-Source (top panel): own elaboration based on spectral data.
-Arb.
-Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum.
-Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
-\label{fig:phosphor_spectrum}
-\end{figure}
-
-\clearpage
-\section{Full List of Sources for Figures}
-\label{sec:sources}
-
-\begin{table}[h!]
-\captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for Figure 1.}
-    \begin{NiceTabularX}{\textwidth}{|l|X|}
-    \hline
-    \textit{Technology} & \textit{References} \\
-    \hline
-    LED & Own research, compare \cite{zenodo_weinold_led_history} \\
-    \hline
-    CFL (<1984) & \cite{Bouwknegt1982}\cite{Vrenken1983} \\
-    \hline
-    CFL (1984-2011) & \cite{eger2018origin} \\
-    \hline
-    CFL (>2011) & \cite{Guan2015} \\
-    \hline
-    Fire, Incandescent, HID & \cite{azevedo2009transition} augmented by own calculations based on \cite{benesch1905beleuchtungswesen} \\
-    \hline
-    Max. efficacy & \cite{Murphy2012} \\
-    \hline
-    \end{NiceTabularX}
-    \vspace{2mm}
-    \caption*{Abbreviations: CFL - Compact Fluorescent Lamp, HID - High-Intensity Discharge}
-\end{table}
-
-\begin{table}[h!]
-\captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for Figure 2.}
-    \begin{NiceTabularX}{\textwidth}{|l|X|}
-    \hline
-    \textit{Source} & \textit{References} \\
-    \hline
-    Stiftung Warentest (Germany) & \cite{Warentest2008}\cite{Warentest2009_1}\cite{Warentest2009_2}\cite{Warentest2010_1}\cite{Warentest2010_2}\cite{Warentest2011}\cite{Warentest2012}\cite{Warentest2013}\cite{Warentest2014_1}\cite{Warentest2014_2}\cite{Warentest2015}\cite{Warentest2016_1}\cite{Warentest2016_2}\cite{Warentest2018} \\
-    \hline
-    Konsument (Austria) & \cite{Konsument2010} \\
-    \hline
-    Which (UK) & \cite{Which2020} \\
-    \hline
-    Industry Periodical & \cite{PM2020} \\
-    \hline
-    Government Report (US) & \cite{national2013assessment} \\
-    \hline
-    \end{NiceTabularX}
-\end{table}
-
-\begin{table}[h!]
-\captionsetup{justification=raggedright,singlelinecheck=false}
-\caption{List of sources for Figure 3.}
-    \begin{NiceTabularX}{\textwidth}{|l|X|}
-    \hline
-    \textit{Publication Type} & \textit{References} \\
-    \hline
-    Scientific Publications & \cite{plossl2010wafer}\cite{bierhuizen2007performance}\cite{gencc2019distributed}\cite{chong2014performance} \\
-    \hline
-    Patents & \cite{patent1999uemura}\cite{patent1998takaoka}\cite{patent1999komaki}\cite{patent1999komaki}\cite{ludowise2006resonant}\cite{camras2005iii}\cite{steigerwald2004contacting} \\
-    \hline
-    Other Publications & \cite{craford2015}\cite{sun2016}\cite{yole2013packaging} \\
-    \hline
-    \end{NiceTabularX}
-\end{table}
-
-\begin{table}[h!]
-\captionsetup{justification=raggedright,singlelinecheck=false}
-\caption{List of sources for Figure 6.}
-    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|}
-    \hline
-    \textit{Phosphor/QD} & \textit{Company} & \textit{YOI} & \textit{Reference} \\
-    \hline
-    YAG & Nichia & 1998 & \cite{bando1998development} \\
-    \hline
-    YGAG & Lumileds & 2002 &  \cite{Mueller2002} \\
-    \hline
-    258 & Lumileds & 2005 & \cite{MuellerMach2005} \\
-    \hline
-    SLA & Lumileds & 2014 & \cite{Pust2014} \\
-    \hline
-    PFS & GE & 2015 & \cite{Murphy2015} \\
-    \hline
-    QD & Lumileds & 2017 & \cite{lumileds2016qd} \\
-    \hline
-    Salon & Osram & 2019 & \cite{osram2019qd} \\
-    \hline
-    SALON & Osram & 2019 & \cite{Hoerder2019} \\
-    \hline
-    \end{NiceTabularX}
-\end{table}
-
-\begin{table}[h!]
-\captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for \cref{fig:haitz_new}.}
-    \begin{NiceTabularX}{\textwidth}{|l|X|}
-    \hline
-    \textit{Metric} & \textit{References} \\
-    \hline
-    Cost & \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} \\
-    \hline
-    Flux & \cite{haitz2007seoul}\cite{haitz2010osram}\cite{haitz2010lumi-1}\cite{haitz2010cree}\cite{haitz2011lumi-1}\cite{haitz2014nichia}\cite{haitz2015cree-1}\cite{haitz2015cree-2}\cite{haitz2018cree}\cite{haitz2019osram}\cite{haitz2019lumi-1}\cite{haitz2019lumi-1}\cite{haitz2020lumi-1}\cite{haitz2013cob}\cite{haitz2016cob}\cite{haitz2017cob}\cite{haitz2019cob}\cite{haitz2019lumi-1}\cite{haitz2019lumi-2} \\
-    \hline
-    \end{NiceTabularX}
-\end{table}
-
-\begin{table}[h!]
-\captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for \cref{fgr:subeff}.}
-    \begin{NiceTabularX}{\textwidth}{|l|X|}
-    \hline
-    \textit{Panel (Sub-Eff.)} & \textit{References} \\
-    \hline
-    Panel A1 (FVE) & \cite{nichia2001data}\cite{lumi2003lumi}\cite{gen2005data}\cite{candlepwr2005data}\cite{lumi2006data}\cite{lumi2007data}\cite{nichia2008data}\cite{lumi2008data}\cite{osram2008data}\cite{jeong2011high}\cite{osram2012data}\cite{osram2013data}\cite{osram2014data} \newline \cite{lumi2016data_1}\cite{lumi2016data_2}\cite{epistar2017data}\cite{osram2017data_1}\cite{osram2017data_2}\cite{samsung2017data}\cite{samsung2018data}\cite{osram2018data}\cite{epistar2018data}\cite{lumi2019data} \\
-    \hline
-    Panel A2 (Droop) & Data calculated from luminous intensity curves of respective device datasheets: \cite{datasheet_osram_topled}\cite{osram2008data}\cite{osram2008gdplus}\cite{osram2018csp}\cite{datasheet_lumileds_lux1}\cite{lumi2008data}\cite{lumi2016data_1}\cite{lumi2016data_2}\cite{samsung2018data} \\
-    \hline
-    Panel B1 (IQE) & Data extracted from respective device datasheets and scientific publications (for a complete list, compare \cite{zenodo_weinold_led_history}).
-Additional data was extracted from government reports: \newline
-\cite{doe_ssl_multiyear_2006}\cite{doe_ssl_multiyear_2007}\cite{doe_ssl_multiyear_2008}\cite{doe_ssl_multiyear_2009}\cite{doe_ssl_multiyear_2010}\cite{doe_ssl_multiyear_2011}\cite{doe_ssl_multiyear_2012}\cite{doe_ssl_multiyear_2013}\cite{doe_ssl_multiyear_2014}\cite{doe_ssl_rnd_2015}\cite{doe_ssl_rnd_2016} \\
-    \hline
-    Panel B2 (LEE) & \cite{lee2005analysis}\cite{krames2007status}\cite{Jang2004}\cite{Horng2013}\cite{Liao2010}\cite{HungWenHuang2005}\cite{Leem2007}\cite{Huang2008}\cite{Wang2009}\cite{Huh2003}\cite{Horng2008}\cite{Gao2008}\cite{Chang2003}\cite{Zhou2012} \newline \cite{ChunJuTun2006}\cite{Hua2009}\cite{Matioli2010}
-\cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
-    \hline
-    \end{NiceTabularX}
-\end{table}
-
-\newpage
-
+\addcontentsline{toc}{section}{Supplementary References}
 \printbibliography[title=Supplementary References]

### 2024-03-29 2_SSL_NE/article/article_main.tex
added section to supplement

@@
 \begin{abstract}
+\hl{
 Since their introduction to the market in 1996, white light-emitting diodes (LEDs) have greatly improved in performance, efficiency, and manufacturing cost.
-These advancements have been crucial for reducing global carbon emissions in the lighting sector which accounts for nearly 20\% of global electricity consumption.
-Understanding the extent and sources of the rapid progress in white LED technology can thus provide valuable insights for accelerating innovation in other demand-side clean energy technologies.
+Understanding the extent and mechanisms of the rapid progress in white LED technology can provide valuable insights for accelerating innovation in other demand-side clean energy technologies critical for reducing global carbon emissions.
 Here we show, through cost and performance modeling based on data from literature review, patent analysis, and expert interviews, that the efficiency of top-performing warm white GaN-based LED packages increased from 5.8\% in 2003 to 38.8\% in 2020.
@@
 We find that technology spillovers from other sectors accounted for at least 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
+}
 \end{abstract}

### 2024-03-29 2_SSL_NE/article/article_supplementary.tex
added section to supplement

@@
 \linenumbers
-
-% custom section designations
-% https://tex.stackexchange.com/a/712523
-
-\newcounter{suppnotecounter}
-\newcommand{\suppnote}[1]{%
-  \stepcounter{suppnotecounter}%
-  \subsection*{\large\bfseries\filright Supplementary Note \arabic{suppnotecounter}: \\ #1}%
-  \addcontentsline{toc}{subsection}{#1}%
-  \label{suppnote:#1}
-}
-
-\newcounter{suppdisccounter}
-\newcommand{\suppdisc}[1]{%
-  \stepcounter{suppdisccounter}%
-  \subsection*{\large\bfseries\filright Supplementary Discussion \arabic{suppdisccounter}: \\ #1}%
-  \addcontentsline{toc}{subsection}{#1}%
-}
-
-\newcounter{suppmethcounter}
-\newcommand{\suppmeth}[1]{%
-  \stepcounter{suppmethcounter}%
-  \subsection*{\large\bfseries\filright Supplementary Method \arabic{suppmethcounter}: \\ #1}%
-  \addcontentsline{toc}{subsection}{#1}%
-}
-
-\newcounter{supprefcounter}
-\newcommand{\suppref}[1]{%
-  \stepcounter{supprefcounter}%
-  \subsection*{\large\bfseries\filright Supplementary References \arabic{supprefcounter}: \\ #1}%
-  \addcontentsline{toc}{subsection}{#1}%
-}
 
@@
 
-\addcontentsline{toc}{section}{Supplementary Notes}
-
-\suppnote{Previous Literature on Technological Progress in LEDs}
+\section{Supplementary Notes}
+
+\subsection{Previous Literature on Technological Progress in LEDs}
 \label{sec:prev_lit}
-
-\cref{sec:prev_lit}
 
@@
 
-\suppnote{Manufacturing Process Steps}
+\subsection{Manufacturing Process Steps}
 \label{sec:manufacturing_steps}
@@
 
-\suppnote{Historical Progress in Chip Architectures}
+\subsection{Historical Progress in Chip Architectures}
 \label{sec:chip_architecture}
@@
 Black frames around certain designs indicate chip designs not brought to large-scale production.
+Additional context on this figure is provided in \cref{sec:subeff_progress}.
 Chip architecture abbreviations: TF - Thin-Film; FC - Flip-Chip; CSP - Chip-Scale Package; Material abbreviations: GaN - Gallium Nitride, ITO - Indium Tin Oxide, MQW - Multiple Quantum Well.
@@
 
-We summarize the information collected and analysed on the progress in GaN-based LED chip architectures and manufacturing processes in \cref{fgr:chip_architecture_overview}, where we show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
-We use this information as input into LED performance calculations and manufacturing cost modeling.
-
-\clearpage
-\suppnote{Historical Progress in Sub-Efficiencies}
+\clearpage
+\subsection{Historical Progress in Sub-Efficiencies}
 \label{sec:subeff_progress}
@@
 Panels B1-B2 show data for internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
+Spectral data related to spectral sub-efficiency is shown in \cref{fig:phosphor_spectrum}.
+Additional context for the panels is provided below.
 Source: own synthesis of data from the list of sources provided in \cref{sec:sources}.
@@
 \clearpage
-
-\textbf{Additional Notes on \cref{fgr:subeff}}
-
-Data illustrated in \cref{fgr:subeff} was used to determine the performance of best-in-class devices across all sub-efficiencies. The results are shown in Figures 4-6 in the main text.
 
@@
 
-\textbf{Additional Data on Phosphor Performance}
-
-Note that \cref{fgr:subeff} does not include sub-efficiencies related to spectral performance of LEDs: spectral efficiency $\eta_S$ and conversion efficiencies of different phosphors $\eta_{C_R}$ and $\eta_{C_{Y/G}}$.
-Typically, there is no direct empirical data available for these metrics in the literature, which made their inclusion in the (mostly) empirical \cref{fgr:subeff} not relevant.
-Instead, we calculate the values of these metrics used in Figures 4 and 5 in the main text from spectral and luminous efficacy radiation data presented in \cref{fig:phosphor_spectrum} below (see also Methods section in the main text).
-
-\clearpage
-\suppnote{Non-Phosphor-Related LED Innovations}
+\textbf{Additional Notes on \cref{fgr:chip_architecture_overview} and \cref{fgr:subeff}}
+
+The data shown in \cref{fgr:subeff} and \cref{fig:phosphor_spectrum} was used to determine the performance of best-in-class devices across different sub-efficiencies.
+The results of these efficiency calculations are shown in Figures 3-5 in the main text.
+\cref{fgr:chip_architecture_overview} provides important context by illustrating the associated changes in chip architecture.
+This show the evolution of LEDs from classical chips with lateral current spreading to chip-scale package flip-chip architectures.
+We provide two examples:
+
+Panel B2 of \cref{fgr:subeff} shows how light-extraction efficiency can be improved from $\sim40\%$ to $\sim95\%$ by switching from a planar Indium-Tin Oxide (ITO) surface to a patterned sapphire substrate (PSS).
+This is listed as a spillover in Figure 4 in the main text.
+Panel A in \cref{fgr:chip_architecture_overview} shows where on the chip these changes take place and provides typical examples of the relevant chip types.
+
+Panel A1 of \cref{fgr:subeff} shows forward voltage was reduced (and therefore forward voltage efficiency increased) over time.
+This trend is observable both within the same chip architectures and between chip architectures.
+The most recent data points for 2018-2019 show how classical chips still have the highest forward voltage.
+Panels A-D in \cref{fgr:chip_architecture_overview} provides typical examples of the relevant chip types.
+
+\clearpage
+\subsection{Non-Phosphor-Related LED Innovations}
 
@@
 \clearpage
-\suppnote{Phosphor-Related LED Innovations}
+\subsection{Phosphor-Related LED Innovations}
 \label{sec:innovation_phosphor}
@@
 
-\suppnote{Spectral Data for Phosphors}
+\subsection{Spectral Data for Phosphors}
 
@@
 \clearpage
-\suppref{Full List of Sources for Figures}
+\subsection{Full List of Sources for Figures}
 \label{sec:sources}
@@
 \clearpage
-\addcontentsline{toc}{section}{Supplementary Methods}
-
-\suppmeth{Domain Definition}
-\label{sec:domain}
-
-\textit{Elsevier Scopus}\footnote{\url{https://www.scopus.com/search/}} advanced search query:
-
+\section{Supplementary Methods}
+
+\subsection{Literature Review Process}
+\label{sec:litrev}
+
+At the first stage of the comprehensive literature review described in section 7.2 in the main text, we conducted a search in specialized patent and publication databases for the most highly cited and/or relevant publications and patents containing information about innovations and overall historical progress in white LED lighting technology.
+
+As several key reviews of technological progress in this field show, the dominant design of white LED lighting that emerged as a result of this progress and diffused at global scale in the lighting market is the GaN-based warm white phosphor-converted LED.
+See also \cref{sec:prev_lit} for a review of these publications.
+We therefore decided to focus our analysis exclusively on the evolution of this variation of white LED lighting technology as the only variation that has achieved global market success.
+Accordingly, we restricted our literature search only to publications focused on GaN-based LED lighting by excluding LEDs with other prominent semiconductor material compositions (i.e., GaAsP, AlGaInP, GaAlAs, ZnO, and perovskites), and on applications of LEDs in lighting by excluding other applications related to solar energy, lasers, or emittance in non-visible parts of the spectrum (e.g., ultraviolet LEDs).
+
+The search was conducted through the \textit{Elsevier Scopus} database using the query provided in \cref{tab:scopus}.
+The exclusion filters placed additional restrictions on topical areas to focus the search outcomes only on the scientific and engineering basis of LED technology.
+The top 300 most highly cited papers were analyzed in the outcomes of this search both for their own content, and as an input to the second stage of the literature search described in section 7.2 of the main text that ensured the comprehensiveness of our findings: an iterative ‘snowballing’ search for backward citations in the publications identified at the first stage.
+
+\begin{table}[h!]
+\caption{Elsevier Scopus advanced search query.}
 \begin{pseudocode}
@@
 \end{pseudocode}
-
-\clearpage
-
-\suppmeth{Metrics for Quantifying Technological Progress in LEDs}
+\label{tab:scopus}
+\vspace{-4mm}
+\end{table}
+
+Additional search was conducted in patent databases \textit{The Lens} and \textit{Google Patents}.
+The search was restricted to a relatively broad CPC patent class H01L 33 (‘Semiconductor devices with at least one potential-jump barrier or surface barrier specially adapted for light emission […]’) and its constituent subclasses, focusing on most highly cited granted patents in this patent class and manually excluding those of them that were not relevant to white LEDs.
+In practice, this meant that the first-stage patent search was restricted to patents granted after 1990.
+
+Finally, to complement the search in publications and patents, we also surveyed the archived websites of key LED manufacturers for any materials dedicated to their interpretation of LED history, history of their own products, and major product announcements.
+
+\clearpage
+
+\subsection{Metrics for Quantifying Technological Progress in LEDs}
 
@@
 \clearpage
-\suppmeth{Device Performance Metrics}
+\subsection{Device Performance Metrics}
 \label{sec:device_performance_metrics}
@@
 \clearpage
-\suppmeth{Consumer Experience Metrics}
+\subsection{Consumer Experience Metrics}
 \label{sec:metrics_consumer}
@@
 
-\suppmeth{Background of Interviewed Experts}
+\subsection{Background of Interviewed Experts}
 
@@
 \clearpage
-\suppmeth{Cost Model Structure and Limitations}
+\subsection{Cost Model Structure and Limitations}
 \label{sec:costmodel_structure}
@@
 
-\suppmeth{Yielded Cost}
+\subsection{Yielded Cost}
 \label{sec:cost_computation}
@@
 
-\suppmeth{Contribution of Variables}
+\subsection{Contribution of Variables}
 \label{sec:contribution_variables}
@@
 
-\addcontentsline{toc}{section}{Supplementary Discussion}
-
-\suppdisc{Cost Model Sensitivity Analysis}
+\section{Supplementary Discussion}
+
+\subsection{Cost Model Sensitivity Analysis}
 
@@
 \clearpage
-\suppdisc{Cost Data Comparison with Industry Projections}
+\subsection{Cost Data Comparison with Industry Projections}
 \label{sec:cost_model_calibration}

### 2024-03-30 2_SSL_NE/article/article_main.tex
edits in supplement

@@
 \linenumbers
+
+\newcommand{\hlcolor}[2][yellow]{ {\sethlcolor{#1} \hl{#2}} }
 
@@
 Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details).
-Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend.
-The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
+\hlcolor[green]{Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend.} The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
 \cite{Murphy2012}.
@@
 
-The historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with LED innovations and manufacturing process improvements that have driven this progress, is described in Supplementary Note 4.
-Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020.
+\hlcolor[green]{Using the mixed-methods approach described in Methods, we have collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress.
+The results of our data collection efforts are presented in Supplementary Figure 7 and discussed in Supplementary Notes section 4.}Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020.
 The waterfall diagrams of electric power input losses in \cref{fgr:waterfall} show how improvements in individual sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to 12.7\% in 2010, 32.5\% in 2016 and finally to 38.8\% in 2020.

### 2024-03-30 2_SSL_NE/article/article_supplementary.tex
edits in supplement

@@
 The historical development development of light-emitting diodes from a novelty semiconductor experiment into powerful lighting devices has received a considerable amount of attention following the recognition of pioneering work on blue LED by Japanese researchers Shuji Nakamura, Isamu Akasaki and Hiroshi Amano with the 2014 Nobel Prize in Physics \cite{Nakamura2015}.
-The sources and disaggregated contributions of subsequent improvements in LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006,krames2007led,laubsch2009high,hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
+The sources and disaggregated contributions of subsequent improvements in white LED technology, however, have not been documented in the literature in a systematic fashion, though there were notable publications reporting on the progress in the design of devices \cite{Shchekin2006,krames2007led,laubsch2009high,hahn2014development}, or the technological improvements  underlying the improvement in overall device efficiency for best performers in 2009 \cite{tsao2010solid} and 2016 \cite{pattison2017solid}.
 
@@
 
-Due to these limitations, any recommendations on future improvements in LED technology based on these studies are fragmented at best.
+Due to these limitations, any recommendations on future improvements in white LED technology based on these studies are fragmented at best.
 Furthermore, the effect of individual innovations and technology spillovers, which has been investigated in solar PV \cite{kavlak2018evaluating,kolesnikov2021spillovers,nemet2019solar}, and to some extent in lithium-ion batteries \cite{Stephan2021}, has not yet been studied in the context of lighting.
 
-In our study, we overcome the limitations of previous literature by, first, synthesizing data provided in multiple publications across an ensemble of various cost, performance and consumer experience metrics for tracking progress in LED technology over an extended period of time.
+In our study, we overcome the limitations of previous literature by, first, synthesizing data provided in multiple publications across an ensemble of various cost, performance and consumer experience metrics for tracking progress in white LED technology over an extended period of time.
 Second, our key advantage and contribution compared to these publications is a multi-method approach that we take to fill in the data gaps and augment information obtained from the literature sources by our own performance calculations, a bottom-up cost model and a series of expert interviews.
@@
 \includegraphics[width=15.5cm]{./figures/subefficiencies_progress.pdf}
-\caption{\textbf{Historical improvements in LED sub-efficiencies.} Panels A1-A2 show data for key metrics used to calculate forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
+\caption{\textbf{Historical improvements in white LED sub-efficiencies.} Panels A1-A2 show data for key metrics used to calculate forward voltage efficiency $\eta_{Vf}$ and droop efficiency $\eta_{droop}$.
 Panels B1-B2 show data for internal quantum efficiency $\eta_{IQ}$ and light extraction efficiency $\eta_{LE}$.
@@
 \begin{table}[H]
-    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note this table is continued in \cref{tab:innovations_2}.}
+    \caption{\textbf{White LED innovations and technology improvements affecting key device sub-efficiencies.} Note this table is continued in \cref{tab:innovations_2}.}
     \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
@@
 \begin{table}[H]
-    \caption{\textbf{LED innovations and technology improvements affecting key device sub-efficiencies.} Note that this table continues \cref{tab:innovations_1}.}
+    \caption{\textbf{White LED innovations and technology improvements affecting key device sub-efficiencies.} Note that this table continues \cref{tab:innovations_1}.}
     \begin{NiceTabularX}{\textwidth}{ |l|X|X|l|l| }
@@
 These include a detailed description of the history of those innovations, which we use to identify the details of involved technology spillovers.
-In addition, spectral data for the first identified LED products that used those innovations is presented in \cref{fig:phosphor_spectrum}, which we use to calculate consumer experience metrics, as well as LED spectral efficiency.
+In addition, spectral data for the first identified white LED products that used those innovations is presented in \cref{fig:phosphor_spectrum}, which we use to calculate consumer experience metrics, as well as LED spectral efficiency.
 
@@
 After several unsuccessful attempts, the sought-after cuboidal nitride compound was found in a 2008 publication led by Francis DiSalvo \cite{Park2008Sr}.
-Based on information provided in this work, Schnick and colleagues synthesized and studied the spectral properties of a new narrow band red SLA phosphor in 2013 \cite{schmidt2013new}\cite{Pust2014}\cite{schmidt2017phosphors}.
-The material was introduced in commercial LED devices by Lumileds in 2015 \cite{lumi2016narrow_whitepaper}.
+Based on information provided in this work, Schnick and colleagues synthesized and studied the spectral properties of a new narrow band red SLA phosphor in 2013 \cite{schmidt2013new}\cite{Pust2014}.
+The material was introduced in commercial white LED devices by Lumileds in 2015 \cite{lumi2016narrow_whitepaper}.
 
@@
 
+\clearpage
+
 \begin{table}[h!]
@@
 \captionsetup{justification=raggedright,singlelinecheck=false}
-    \caption{List of sources for \cref{fig:haitz_new}.}
-    \begin{NiceTabularX}{\textwidth}{|l|X|}
-    \hline
-    \textit{Metric} & \textit{References} \\
-    \hline
-    Cost & \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} \\
-    \hline
-    Flux & \cite{haitz2007seoul}\cite{haitz2010osram}\cite{haitz2010lumi-1}\cite{haitz2010cree}\cite{haitz2011lumi-1}\cite{haitz2014nichia}\cite{haitz2015cree-1}\cite{haitz2015cree-2}\cite{haitz2018cree}\cite{haitz2019osram}\cite{haitz2019lumi-1}\cite{haitz2019lumi-1}\cite{haitz2020lumi-1}\cite{haitz2013cob}\cite{haitz2016cob}\cite{haitz2017cob}\cite{haitz2019cob}\cite{haitz2019lumi-1}\cite{haitz2019lumi-2} \\
-    \hline
-    \end{NiceTabularX}
-\end{table}
-
-\begin{table}[h!]
-\captionsetup{justification=raggedright,singlelinecheck=false}
     \caption{List of sources for \cref{fgr:subeff}.}
@@
 \cite{lee2005analysis}\cite{Zhu2015}\cite{Ding2015}\cite{Taki2019}\cite{Shchekin2006}\cite{Hu2016}\cite{Horng2010}\cite{Lin2016}\cite{Yue2018}\cite{Zhao2012}\cite{Zhu2015}\newline \cite{Ding2015}\cite{wierer2001high}\cite{Steigerwald2002}\cite{DaeSeobHan2006}\cite{Wang2006}\cite{Lee2007}\cite{Shen2007}\cite{Huang2006}\cite{Zhmakin2011} \\
+    \hline
+    \end{NiceTabularX}
+\end{table}
+
+\begin{table}[h!]
+\captionsetup{justification=raggedright,singlelinecheck=false}
+    \caption{List of sources for \cref{fig:haitz_new}.}
+    \begin{NiceTabularX}{\textwidth}{|l|X|}
+    \hline
+    \textit{Metric} & \textit{References} \\
+    \hline
+    Cost & \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid} \\
+    \hline
+    Flux & \cite{haitz2007seoul}\cite{haitz2010osram}\cite{haitz2010lumi-1}\cite{haitz2010cree}\cite{haitz2011lumi-1}\cite{haitz2014nichia}\cite{haitz2015cree-1}\cite{haitz2015cree-2}\cite{haitz2018cree}\cite{haitz2019osram}\cite{haitz2019lumi-1}\cite{haitz2019lumi-1}\cite{haitz2020lumi-1}\cite{haitz2013cob}\cite{haitz2016cob}\cite{haitz2017cob}\cite{haitz2019cob}\cite{haitz2019lumi-1}\cite{haitz2019lumi-2} \\
     \hline
@@
 \subsection{Metrics for Quantifying Technological Progress in LEDs}
+\label{sec:metrics}
 
@@
 
-\textbf{Distinction between Metrics}
-
-We use a well-established distinction between metrics.
+\textbf{Distinction between Metrics of Efficiency and Consumer Experience}
+
+We use a well-established distinction between the two types of LED lighting technology performance metrics.
 Those metrics which are related to the spectrum of the light source only ("color quality metrics" \cite{DLC_SSL_Requirements}\cite{dilaura2011lighting}), we group under the term consumer experience metrics.
-Those metrics, which are determined by the number and energy of photons leaving the device (as a function of the number of electrons entering it), we group under the term device performance metrics.
+Those metrics, which are determined by the number and energy of photons leaving the device (as a function of the number of electrons entering it), we group under the term LED efficiency metrics.
 As such, our distinction is physical at the fundamental level.
@@
 
-A highly cited and frequently called-upon metric is \textit{flux/package}, as described in what has become known as Haitz's law.
+A highly cited and frequently called-upon metric is \textit{flux per package}, as described in what has become known as Haitz's law.
 An observation rather than a law of nature, it describes increasing device performance and decreasing manufacturing cost.
@@
 The continuing rise in performance inferred from this chart is disconnected from the actual, much lower, performance of single chips \cite{weinold2021compound}.
-This is shown in \cref{fig:haitz-org}-\cref{fig:haitz}.
+This is shown in \cref{fig:haitz-org}-\cref{fig:haitz_new}.
 
@@
 The original data by Haitz for red LEDs from 2000 is indicated for reference.
-Note the different range on the abscissa compared to \cref{fig:haitz_new}.
+Note the different range on the horixontal axis compared to \cref{fig:haitz_new}.
 Sources: \cite{haitz1999case}\cite{haitz2011solid} and  \cref{fig:haitz_new}.}
@@
     \includegraphics[width=0.85\textwidth]{./figures/haitz_law_white.pdf}
-    \caption{Adaption of \textit{"Haitz's Law"}.
+    \caption{Corrected version of \textit{"Haitz's Law"}.
 Note that the increase in flux per single chip is not increasing as starkly as the flux for multi-chip packages.
@@
 In 2000, the highest performing devices yielded around 10 lm, just below the output of a candle as defined in the unit candela (1 cd=12.57 lm) \cite{haitz2011solid}.
-Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}.
+Progress in this metric was commonly rendered together with the associated reduction in retail prices as “Haitz’s Law”\cite{haitz1999case,haitz2011solid}, which we discuss in \cref{sec:metrics}.
 Today, however, the devices with the highest performance yield in excess of 1600 lm, the equivalent of a 100 W incandescent bulb \cite{cree2020bright}.
@@
 We therefore use a combination of the total device efficiency ("lamp efficiency") and the sub-efficiencies that describe different physical loss channels within the device to describe progress in LED technology.
-In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
+In line with previous publications \cite{schubert2018light,tsao2010solid}, we use the following sub-efficiencies: forward voltage efficiency $\eta_{V_f}$, light extraction efficiency $\eta_{LE}$, internal quantum efficiency $\eta_{IQ}$, droop $\eta_{droop}$, conversion efficiency $\eta_{C}$, spectral efficiency $\eta_{S}$.
 The equations defining these sub-efficiencies are provided below.
@@
 \begin{equation}
-    \eta_{IQ} = \frac{\# \text{of photons created/s}}{\# \text{of electron-hole pairs in/s}}
+    \eta_{IQ} = \frac{\# \text{of photons created/s}}{\# \text{of electron-hole pairs injected/s}}
 \end{equation}
@@
 The third , scattering and absorption processes in the phosphor material.
-These three effects are listed individually in Figure 4.
+These three effects are listed individually in Figure 4 in the main text.
 The overall light conversion efficiency is defined as
@@
 
-We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up LED manufacturing cost model with process-step resolution.
+We address the limitations of data availability for the chip-level LED manufacturing costs and sales prices by developing and applying a bottom-up white LED manufacturing cost model with process-step resolution.
 In our case, it is enabled by the general availability of historical data on prices of relevant raw materials, components, and manufacturing equipment, and the relative cost data on manufacturing processes, which is occasionally published as part of industry press releases \cite{ledinside2013csp}\cite{seoul2015csp}.
 
-We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:costmodel_details}.
+We describe our manufacturing cost model in the Methods section in the main text of the article, and discuss its details, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, explanation of our cost modelling approach based on yielded costs, as well as the model’s limitations, in \cref{sec:costmodel_structure}.
 The outcome of the cost model is the cumulative yielded manufacturing cost per LED package, which includes all costs associated with producing the chip, including running costs of the factory.
@@
 This standard has been widely adopted industry and provides a standardized comparative analysis of the economics of the manufacturing chain \cite{ragona2002cost}\cite{wouters2018industry}.
-An overview of the parameters used in the model, according to \textit{SEMI Standard 35} is provided in \cref{semi_pyramid}.
+An overview of the parameters used in the model, according to \textit{SEMI Standard 35} is provided in \cref{fig:semi_pyramid}.
 Note that the full ensemble of inputs are required only for an industry cost model used by manufacturers for detailed forecasting.
@@
 While generally comprehensive, the model is limited by the availability of performance data for the most recent manufacturing equipment, as well as the confidential nature of cost structures of Asia-based manufacturing lines.
-A complete treatment of the limitations of the model is provided in \cref{sec:costmodel_limitations}.
+A complete treatment of the limitations of the model is provided in \cref{sec:costmodel_structure}.
 
@@
 
-The model is set up up to capture the cost structure of a hypothetical factory employing the most up-to-date equipment in every model year.
-It operates at a fixed-cost basis (labor cost, electricity cost, cleanroom cost, etc.), which was - due to limitations in data availability - taken to be based on a US-based semiconductor manufacturing plant.
+Due to limitations in data availability, the cost model is set up to capture the cost structure of a hypothetical factory employing the most up-to-date equipment in every model year.
+It operates at a fixed-cost basis (labor cost, electricity cost, cleanroom cost, etc.).
 However, this does not correspond to any \textit{real, physical} location.

### 2024-03-30 2_SSL_NE/article/article_main.tex
edited main text

@@
 The full list of identified LED innovations considered in our study is provided in Table SI2 in Section 3 of the Supplementary Information document.
-The list of corresponding technology spillovers is provided in \cref{tab:spillovers} in the Discussion section below.
-Through the index decomposition analysis described in Section 2.5 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8\% can be attributed specifically to technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
+The list of corresponding technology spillovers is provided in \cref{tab:spillovers}.
+Through the index decomposition analysis described in Section 2.5 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8 percentage points can be attributed specifically to innovations driven by technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
@@
 Notably, some sub-efficiencies for the most recent devices considered in our study are now within $\sim10\%$ of their respective physical limits.
-The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements, which is important, given that efforts to improve LED performance across different efficiency loss channels through R\&D continue \cite{cho2017white, Weisbuch2020}.
+The exception is spectral efficiency which, at $\sim17\%$ below the physical limit, shows larger potential for further improvements, which is important, given that efforts to improve white LED performance across different efficiency loss channels through R\&D continue \cite{cho2017white, Weisbuch2020}.
 
@@
  \captionsetup{font=footnotesize}
- \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power LED package for 2003, 2010, 2016 and 2020}.
+ \caption{\textbf{Waterfall diagrams of the loss channels in a generic mid/high-power white LED package for 2003, 2010, 2016 and 2020}.
 Losses are normalized to 1W of electric power input.
@@
 As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.
+All innovations in this list are primarily a result of R\&D efforts.
 
-Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 4, we find that only a single innovation related to LED consumer experience improvements was initially intended exclusively for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
+Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 4, we find that only a single innovation related to LED consumer experience improvements was purposefully developed for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
 All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:spillovers}).
@@
 We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
-Our model shows that for the wafer-level manufacturing steps illustrated in Supplementary Figures 2-6, yield improvements and an increase in  wafer size together are responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.
+Our model shows that for the wafer-level manufacturing steps illustrated in Supplementary Figures 1-5, improved manufacturing yields and increases in the wafer size are jointly responsible for the largest contribution — 95.5\% — to the overall reduction of manufacturing cost per LED chip between 2003 and 2020.
 
-Among the factors contributing to the LED cost reductions over time, improved manufacturing yields and increases in the wafer size rather than particular LED innovations are found to be responsible for the largest contribution to the overall cost reduction.
 In the case of manufacturing yield, the higher it is, the less inputs are wasted on the production of a single LED package.
@@
 
-Surprisingly, LED innovations and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time.
+Surprisingly, R\&D efforts to improve LED technology and associated technology spillovers are not found to have a substantial impact on LED manufacturing cost reductions over time.
 There are several explanations for this finding.
@@
 We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
-Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 98\% reduction of manufacturing cost per LED chip between between 2003 and 2020.
-Both were enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular LED innovations, which was confirmed by interviews with experts in LED manufacturing.
+Second, we find that increases in the wafer size and higher overall manufacturing yield, which are jointly responsible for a 95.5\% manufacturing cost reduction per LED chip, were both enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular innovations in LED technology associated with R\&D efforts, as confirmed by interviews with experts in LED manufacturing.
 
@@
 \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
-\caption{\textbf{Historical changes in LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and Supplementary Note 3) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+\caption{\textbf{Historical changes in white LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and Supplementary Note 3) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
 location.
-Panels A-C: Waterfall diagrams of LED manufacturing cost structure split by manufacturing process steps for years 2003, 2012 and 2020.
+Panels A-C: Waterfall diagrams of white LED manufacturing cost structure split by manufacturing process steps for years 2003, 2012 and 2020.
 Process steps on the horizontal axis are sequenced from left to right in the same order as in the modelled LED manufacturing process.
@@
 Following the frameworks for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021} and Kolesnikov et al.
-\cite{kolesnikov2022technology}, we synthesize information from the interviews, historical records, literature review and citations in patents and publications to identify which LED innovations involved knowledge originating in areas of science and technology beyond LED and SSL.
-We find that at least nine LED innovations, listed in Table 2, clearly involved such ‘external’ knowledge.
+\cite{kolesnikov2022technology}, we synthesize information from the interviews, historical records, literature review and citations in patents and publications to identify which innovations in white LED technology were driven by external knowledge originating in areas of science and technology beyond white LED lighting.
+We find that at least nine LED innovations, listed in Table 2, clearly involved such external knowledge.
 We then analyze the sources, mechanisms and enablers of corresponding technology spillovers, as well as their contributions to progress in white LED technology.
@@
 We find that three spillovers associated with the use of YAG/YGAG phosphors in LEDs played the key role in the development of the first commercial white LED lighting products, essentially enabling the solid-state lighting market and industry of today.
-As noted above, subsequent spillovers then had a significant effect on physical device performance, cumulatively contributing at least 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020, as well as nearly all key improvements in consumer experience metrics.
+As noted above, subsequent spillovers then contributed to white LED technology innovations that were cumulatively responsible for at least 8.5\% of the total LED lamp efficiency improvements between 2003 and 2020, as well as nearly all key improvements in consumer experience metrics.
 
@@
 
-Among important enabling factors for the identified spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
-Finally, we find that, on average, it took 26 years from the initial scientific discovery or invention to the moment of its spillover into the LED domain, with this time varying from 5 to almost 70 years.
-In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED market.
+Among enabling factors that we identified for these spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
+Finally, we find that, on average, it took 26 years from the initial invention or discovery taking place in a field of science and technology different from white LED lighting to the first time it got applied in (‘spilled over to’) the white LED lighting domain, with this time varying from 5 to almost 70 years.
+In contrast, it took much less time – just 6 years on average, varying from just a few months to 19 years – to develop a commercial application for the spillover knowledge in the LED lighting market.
 
@@
     \hline
-        \textbf{Disc.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Enabler} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
+        \textbf{Inv.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Enabler} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
         1926 & 1994 & 1996 & LED phosphors &  Use of phosphors for light down conversion in LEDs & ???
 & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
-        1967 & 1996 & 1996 & YAG phosphor & Use of YAG phosphor in a first white LED product & ???
+        1967 & 1996 & 1996 & YAG:Ce phosphor & Use of YAG:Ce phosphor in a first white LED product & ???
 & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
@@
     \end{NiceTabularX}
-    \caption*{Note: Disc.
-- Year of initial discovery, identified  by the earliest literature source describing the discovery of original idea or invention outside the LED domain.
+    \caption*{Note: Inv.
+- Year of initial invention, identified by the earliest literature source describing the original invention or idea in a field outside white LED lighting that eventually ‘spilled over’ to the latter field.
 S/O - Year of spillover to LED; Comm.
@@
 
-The evolution of LED device architecture and performance, as well as the progress in understanding the underlying physical phenomena, are relatively well covered in scholarly literature and patents (see Supplementary Note 1 for a brief literature review).
+The evolution of white LED device architecture and performance, as well as the progress in understanding the underlying physical phenomena, are relatively well covered in scholarly literature and patents (see Supplementary Note 1 for a brief literature review).
 However, information provided in such sources is insufficient for our goals on at least three accounts: First, existing work focuses only on selected performance parameters or overall device efficiency, rather than on providing a comprehensive coverage of the whole device sub-efficiencies for a particular LED product or design.
@@
 To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a comprehensive review of the primary scientific literature \cite{Haddaway_2014}, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies.
-We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in innovation and technology spillovers.
+We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in individual innovations and technology spillovers.
 We briefly describe our approach to each method below and provide further methodological details in Supplementary Note 3.
@@
 
-We collected data on LED performance and characteristics in a comprehensive literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
-This review was structured around the three main goals: 1) tracking the evolution of LED technology over time as indicated by three groups of progress metrics; 2) identifying individual innovations that contributed to this evolution, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the LED technology domain, or in a field of science or technology outside of SSL, making them involved in technology spillovers.
+We collected data on white LED performance and characteristics in a comprehensive literature review that included scientific publications, patents, conference proceedings from the largest semiconductor and optoelectronics conferences, industry periodicals and roadmaps, as well as company presentations and reports.
+This review was structured around the three main goals: 1) tracking the evolution of white LED technology over time as indicated by three groups of progress metrics; 2) identifying individual innovations that contributed to this evolution, and quantifying their impact on device performance and manufacturing cost; and 3) determining whether these innovations had originated within the white LED lighting technology domain, or in a field of science or technology outside of SSL, making them involved in technology spillovers.
 
@@
 The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in Supplementary Note 1 and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
-We also relied on backward citations in these sources for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the LED technology domain or not.
+We also relied on backward citations in these sources for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the white LED lighting technology domain or not.
 
@@
 The structure of our bottom-up white LED manufacturing cost model with process-step resolution is generally based on the 2012 \textit{LEDCOM} cost model \cite{ledcomv2}, but we expand it significantly both in scope and in its ability to capture historical trends.
-The model captures three historical time periods corresponding to different “eras” in LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
+The model captures three historical time periods corresponding to different “eras” in white LED manufacturing: the early period of the first high-power white LEDs around 2003, the period of accelerating consumer adoption of LED lighting around 2012, and the most recent period around 2020, the year of our main data collection efforts.
 For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the Internet Archive, and expert interviews.
@@
 In addition, two LED life-cycle analyses were used to validate the model structure and extract some of the necessary quantitative model inputs.
-These studies captured a large number of LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
+These studies captured a large number of white LED manufacturing process steps and included the details on the use of metals, chemicals and electricity for each manufacturing step.
 
@@
 Costs associated with research and development, administrative overhead, or other investment costs are not considered.
-We note that the purpose of the cost model is not to provide specific estimates of the LED manufacturing cost for a factory of specific size, geographic location or manufacturing capacity.
+We note that the purpose of the cost model is not to provide specific estimates of the white LED manufacturing cost for a factory of specific size, geographic location or manufacturing capacity.
 It is instead intended to capture the impact of specific improvements in the manufacturing process flow on overall cost.

### 2024-03-30 2_SSL_NE/article/article_main.tex
various updates

@@
 \usepackage{lineno}
-\linenumbers
-
-\newcommand{\hlcolor}[2][yellow]{ {\sethlcolor{#1} \hl{#2}} }
+%\linenumbers
 
@@
 \begin{abstract}
-\hl{
+
 Since their introduction to the market in 1996, white light-emitting diodes (LEDs) have greatly improved in performance, efficiency, and manufacturing cost.
@@
 We find that technology spillovers from other sectors accounted for at least 8.5\% of efficiency improvements and nearly all consumer experience enhancements, playing an important role in widespread LED adoption in lighting.
-}
+
 \end{abstract}
@@
 A rapid reduction of global carbon dioxide emissions is urgently required in order to mitigate the effects of climate change \cite{Forster2019}.
-While the Paris Agreement of 2015 requires signatory states to develop national emissions reduction strategies, these are not legally binding \cite{bodansky2016legal}.
+While the 2015 Paris Agreement requires signatory states to develop national emissions reduction strategies, these are not legally binding \cite{bodansky2016legal}.
 However, some countries have introduced more ambitious legislation, including the European Union, which has set a target of net zero emissions by 2050.
@@
 
-Achieving such ambitious and critically important targets will require both the deployment of new clean energy technologies \cite{iea2020cleanenergy}, and the acceleration of innovation in existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}.
-To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance and consumer experience are needed.
-This requires understanding how these cost reductions and performance improvements can be achieved \cite{Stephan2021,Ziegler2021}.
+Achieving such ambitious and critically important targets requires both developing new clean energy technologies \cite{iea2020cleanenergy} and accelerating the deployment of existing supply-side \cite{sinn2012green} and demand-side energy technologies \cite{rgeVorsatz2009}.
+To ensure rapid adoption of these technologies, significant reductions in their costs and improvements in their performance are needed.
+This requires understanding how these cost reductions and performance improvements can be achieved \cite{nemet2019solar,Stephan2021,kavlak2018evaluating,Ziegler2021}.
 
-There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{WRIGHT_1936, Arrow_1962}, and economies of scope \cite{johansson2012global, national2016power, iea2020perspectives}.
-The role of these factors at different stages of the innovation life cycle, from research and technology development to demonstration, market formation and diffusion \cite{grubler2012policies}, is an area of active research  \cite{kavlak2018evaluating, Ziegler2021}.
-Innovation is also driven by forces of supply and demand \cite{Mowery1979}.
-Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.
+Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning by doing\cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale\cite{johansson2012global, iea2020perspectives}.
+The role of these mechanisms at different stages of technology life cycle, from research and development to demonstration, market formation and diffusion\cite{grubler2012policies}, is an area of active research \cite{kavlak2018evaluating,Ziegler2021}.
+Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Acemoglu_2022}.
+For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Castelnuovo_2005} more often than other mechanisms can result in substantial or even radical improvements in technology performance.
+On the other hand, learning-by-doing is typically associated with continuous incremental improvements in the technology resulting from accumulated experience in technology demonstration, manufacturing, and use.
+Both mechanisms, as well as economies of scale, can also lead to radical (i.e., discontinuous) or incremental reductions in technology costs.
 
-The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies, Anadon2016}.
+Importantly, different mechanisms of innovation in a particular technology play out within a broader innovation system comprising multiple actors, networks, and institutions\cite{grubler2012policies, Anadon2016}.
+As a result, these mechanisms are affected by various drivers and barriers, including forces of supply and demand\cite{Mowery1979}, public policies, and regulation\cite{anadon2009policy}.
+
 Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}.
-While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2021spillovers}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2021spillovers} by considering technology spillovers as the application of external knowledge in a technology where knowledge is considered external if it has been developed for application in other technologies, sectors, or scientific disciplines.
+While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2021spillovers}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2021spillovers}.
+We define technology spillovers as the application of external knowledge in a technology where knowledge is considered external if it has been developed for application in other technologies, sectors, or scientific disciplines.
 There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies.
+For example, spillovers can provide critical external knowledge inputs to the R\&D process \cite{kolesnikov2021spillovers}, enable the re-use of experience from different sectors in learning-by-doing\cite{iea2020cleanenergy}, and lead to cost reductions, e.g., via repurposing the manufacturing equipment from other sectors\cite{Stephan2021}.
 
-Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently   accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}.
-It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements.
+The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diode (LED) lighting.
+Among demand-side technologies, general illumination is a particularly important area for climate change mitigation efforts, as it currently accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}.
+It is also an area of rapid recent technological change: since the introduction of first commercial white LEDs in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements.
 As shown in \cref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}.
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
  \centering
@@
 Luminous efficacy is the measure of how efficiently a light source converts electrical energy into visible light that can be perceived by the human eye, taking into account the wavelength sensitivity of the eye (see Supplementary Note 2 for details).
-\hlcolor[green]{Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend.} The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
+Dashed lines are guides to the eye, based on a 3rd-order polynomial fit to the data trend.
+The physical limit for an ideal light source with a colour rendering index of CRI=90, denoted as $K_{max}^{CRI90}$, is shown as a black horizontal line, as per calculations by Murphy et al.
 \cite{Murphy2012}.
@@
  \label{fgr:history_efficacy}
- \vspace{-7mm}
 \end{figure}
 
-\clearpage
-
-\begin{figure}[H]
+\begin{figure}[h!]
 \centering
@@
 Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period.
+Note that projections are consistently at the lower end of the retail price data.
+This is because retail price data includes more expensive luminaire types, while projections are for traditional light bulbs only.
+Projections also do not include value added tax.
 Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications.
@@
 
-Despite this impressive history, the sources of technological progress in white LEDs have not received as much attention from researchers as innovation in supply-side energy technologies, such as solar photovoltaics \cite{kavlak2018evaluating, nemet2019solar}, wind energy \cite{qiu2012price, jennings2020policy}, or lithium-ion batteries \cite{Ziegler2021, Stephan2021}.
-To the best of our knowledge, no study has comprehensively discussed the sources or extent of historical progress across a range of metrics related to LED cost and performance since the introduction of the first commercial white LED products (see Supplementary Note 1 for a brief review of previous literature on this topic).
+Despite this impressive history, the mechanisms of technological progress in white LEDs have not received as much attention from researchers as in supply-side energy technologies, such as solar photovoltaics \cite{kavlak2018evaluating, nemet2019solar}, wind energy \cite{qiu2012price, jennings2020policy}, or lithium-ion batteries \cite{Ziegler2021, Stephan2021}.
+To the best of our knowledge, no study has comprehensively discussed the mechanisms or extent of historical progress across a range of metrics related to LED cost and performance since the introduction of the first commercial white LED products (see Supplementary Notes section 1 for a brief review of previous literature on this topic).
 This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012, Creutzig2018}.
-Understanding the extent to which individual innovations and technology spillovers contributed to improvements in white LED technology, how their effect compares to other sources of improvements such as economies of scale, and how these innovations and spillovers occurred (i.e., by what mechanisms and actors) will provide valuable lessons for innovation in other demand-side energy technologies and clean energy innovation in general.
+Understanding the extent to which individual innovations and technology spillovers contributed to overall improvements in white LED technology cost and performance, by which mechanism these innovations and spillovers occurred, and what factors affected this process will provide valuable lessons for innovation in other demand-side energy technologies and clean energy innovation in general.
 
-\clearpage
-
-To address these questions, we use three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost.
-The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Note 2.
-We find that  total LED device efficiency increased from 5.8\% to 38.8\% between 2003 and 2020, of which at least 8.5\% can be attributed specifically to technology spillovers.
-Our interviews have also revealed the important role of incremental manufacturing process improvements and learning-by-doing \cite{WRIGHT_1936, Arrow_1962} in this progress in LED efficiency.
-However, further research is needed to quantify the contribution of learning-by-doing to this progress.
-On the other hand, we find that technology spillovers contributed to nearly all consumer experience improvements in LED-based lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
-Finally, our LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid-power classic-chip-architecture GaN-based phosphor-converted warm-white LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020.
-In contrast with LED performance improvements, where progress was driven mostly by innovations in LED technology, the dramatic decline in LED manufacturing cost has been a product of higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.
-Overall we find that technology spillovers are most prevalent in those device loss channels that are well understood at the physical level.
-Improvements in device loss channels that are governed by complex quantum effects at the atomic level, and can only be described heuristically, have come from incremental manufacturing process improvements and learning-by-doing.
-Directing research efforts to better understand fundamental physical or chemical processes related to novel clean energy technologies is therefore likely to increase the possibility of technology spillovers.
+To address these questions, we first identify three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost.
+The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Notes section 2.
+Next, using data from literature review, patent analysis, and expert interviews, we trace the historical improvements in these metrics in warm white GaN-based phosphor-converted LEDs—currently the most widely adopted variant of white LED technology which we thus choose as the focal technological area for our analysis—from their introduction to the lighting market in 2003 to 2020, the year with the most recent data available at the time of data collection.
+We find that total LED device efficiency of such LEDs increased from 5.8\% to 38.8\% between 2003 and 2020.
+A majority of this increase can be attributed to a series of innovations in white LED technology resulting from R\&D efforts, with at least 8.5\% of the cumulative increase attributed specifically to technological innovations driven by technology spillovers.
+Our interviews also revealed a lesser but still important role of incremental manufacturing process improvements most likely resulting from learning-by-doing.
+However, substantial further research is needed to quantify the relative contribution of learning-by-doing vis-a-vis technological innovations to this progress.
+On the other hand, we find that innovations driven by technology spillovers contributed to nearly all consumer experience improvements in white LED lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
+Finally, our bottom-up LED manufacturing cost model shows a 95.5\% decrease in the cost of producing low-to-mid-power classic-chip-architecture GaN-based phosphor-converted warm-white LEDs from 1.11\$ to 0.05\$ (in 2020 USD) between 2003 and 2020.
+In contrast with LED performance improvements, where progress has been driven mostly by R\&D efforts, the dramatic decline in LED manufacturing cost has been a product of learning-by-doing resulting in higher yields across manufacturing steps and economies of scale resulting from increases in the sapphire wafer size.
+We provide the details of our findings for the progress in white LED efficiency, cost, and consumer experience, as well as the role of technology spillovers in this progress, in the remaining sections below.
 
@@
 
-\hlcolor[green]{Using the mixed-methods approach described in Methods, we have collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress.
-The results of our data collection efforts are presented in Supplementary Figure 7 and discussed in Supplementary Notes section 4.}Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020.
+Using the mixed-methods approach described in Methods, we have collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress.
+The results of our data collection efforts are presented in Supplementary Figure 7 and discussed in Supplementary Notes section 4.
+Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020.
 The waterfall diagrams of electric power input losses in \cref{fgr:waterfall} show how improvements in individual sub-efficiencies led to improvements in the overall white LED lamp efficiency from $\eta_L=5.8\%$ in 2003 to 12.7\% in 2010, 32.5\% in 2016 and finally to 38.8\% in 2020.
@@
 In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels.
+The estimates shown here for the contributions of different channels do not show uncertainty ranges because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, and the fact that we focus on a best case scenario.
+Thus, our point estimates should be interpreted as the expected values in a best case scenario.
 We find that there has been consistent progress across all device sub-efficiencies in the recorded period.
@@
     \caption{\textbf{LED innovations related to improvements in consumer experience metrics.}}
-    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|l|}
+    \begin{NiceTabularX}{\textwidth}{|l|l|l|X|X|}
     \hline
-        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} & \textbf{Ref.} \\ \hline
-        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs & SI 3.2.2.1 \\ \hline
-        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs & SI 3.2.2.1 \\ \hline
-        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor & SI 3.2.2.2 \\ \hline
-        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion & SI 3.2.2.4 \\ \hline
-        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor & SI 3.2.2.3 \\ \hline
-        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor & SI 3.2.2.2 \\ \hline
-        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor & SI 3.2.2.2 \\ \hline
+        \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} \\ \hline
+        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs \\ \hline
+        1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs \\ \hline
+        2002 & 258 & $(Ba,Sr)_2 Si_5 N_8:Eu^{2+}$ & Europium-doped nitridosilicate phosphor & First red LED phosphor \\ \hline
+        2003 & QD & N/A & Quantum dot-based phosphor & First use of QD for LED light down conversion \\ \hline
+        2005 & PFS & $K_2 SiF_6: Mn^{4+}$ & Manganese-activated potassium fluorosilicate (PFS) phosphor & First ultra-narrow-band red LED phosphor \\ \hline
+        2013 & SLA & $Sr[Li Al]_3 N_4 ]:Eu^{2+}$ & Europium-doped cuboidal nitridolithoaluminate phosphor & Improved narrow-band red phosphor \\ \hline
+        2016 & SALON & $Sr[Li_2 Al_2 O_2 N_2]:Eu^{2+}$ & Europium-doped oxonitride phosphor & High-performance ultra-narrow-band red phosphor\\ \hline
     \end{NiceTabularX}
@@
 We also previously showed that overall brightness is no longer a target metric in chip development \cite{weinold2021compound}.
-Second, we find that increases in the wafer size and higher overall manufacturing yield, which are jointly responsible for a 95.5\% manufacturing cost reduction per LED chip, were both enabled by advances in manufacturing equipment (e.g., for epitaxy and chip dicing) and incremental manufacturing process improvements from learning-by-doing, rather than particular innovations in LED technology associated with R\&D efforts, as confirmed by interviews with experts in LED manufacturing.
+Second, we find that an increase in wafer size and higher overall yield, taken together, are jointly responsible for a 95.5\% reduction of manufacturing cost per LED chip between 2003 and 2020.
+Both were enabled by advances in manufacturing equipment (e.g., in epitaxy) enabling the economies of scale in manufacturing, as well as incremental manufacturing process improvements from learning-by-doing, rather than LED technology innovations stemming from R\&D, which was confirmed by our interviews with experts in LED manufacturing.
 
@@
 
-Among enabling factors that we identified for these spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
+Among the enabling factors that we identified for these spillovers, we highlight public mission-driven R\&D funding; industry-academia partnerships; firm experience in multiple industries; conferences that brought together researchers from academia and industry; cultural and language proximity; freedom of search in academia; and university alumni networks.
 Finally, we find that, on average, it took 26 years from the initial invention or discovery taking place in a field of science and technology different from white LED lighting to the first time it got applied in (‘spilled over to’) the white LED lighting domain, with this time varying from 5 to almost 70 years.
@@
 
-In this study, we use a multi-method approach to synthesize evidence from multiple sources and, for the first time, reconstruct a comprehensive picture of the rapid technological progress in white LED technology throughout its history since the introduction to the market in 1996 and across an ensemble of metrics related to device performance, cost, and consumer experience.
-Improvements in these metrics are traced to sources in innovation, technology spillovers, and economies of scale in the manufacturing process.
-Improvements in device efficiency are shown in \cref{fgr:waterfall} and \cref{fgr:breakthroughs_efficiency}.
-Improvements in consumer experience are shown in \cref{fgr:consumer_experience}.
-Improvements in manufacturing cost are shown in \cref{fgr:costmodel}.
+In this study, we use a multi-method approach to synthesize evidence from multiple sources and, for the first time, reconstruct a detailed picture of the rapid technological progress in white LED technology throughout its history since the introduction to the market in 1996 and across an ensemble of metrics related to device performance (\cref{fgr:waterfall} and \cref{fgr:breakthroughs_efficiency}), manufacturing cost (\cref{fgr:costmodel}), and consumer experience (\cref{fgr:consumer_experience}).
+Improvements in these metrics are traced (with contributions quantified, where possible) to specific white LED technology innovations resulting from R\&D activities, technology spillovers, learning-by-doing, and economies of scale in the manufacturing process.
 
-Our analysis of the sources, mechanisms and enablers of the identified technology spillovers has important practical implications, highlighting the critical role played by a deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.
-Specifically, deep physical understanding of LED device energy loss channels had enabled important innovations in LEDs that increased several sub-efficiencies in LEDs and will continue to do so, as expected by eminent experts in the field .
-This suggests that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology and its applications both in SSL and technology areas beyond general illumination.
-Corresponding knowledge spillovers can be enabled and deliberately stimulated by various measures such as knowledge exchange events and long-term partnerships between academia and industry, interdisciplinary training and hiring, dedicated mission-driven public R\&D funding, and ensuring certain freedom of search in academia.
+Specifically, we find that progress in LED efficiency and consumer experience has been predominantly driven by innovations resulting from R\&D activities, but the relative contribution of technology spillovers to these innovations has been very different: at least 8.5\% for efficiency improvements and virtually all 100\% for consumer experience metrics.
+In addition, we observe (but so far have not been able to quantify) a likely small but meaningful role of learning-by-doing in efficiency improvements.
+Interestingly, we find that technology spillovers are most prevalent in those device loss channels that are well understood at the physical level (e.g., light extraction and spectral efficiencies, see Fig.
+4 and Table 2).
+We find that improvements in device loss channels that are governed by complex quantum effects at the atomic level (e.g., droop) and at present can only be described heuristically, have, on the other hand, come mostly from incremental manufacturing process improvements resulting from learning-by-doing with little evidence for spillovers.
+
+In contrast with efficiency and consumer experience improvements, we find that dramatic manufacturing cost reductions in white LEDs have been mostly driven by learning-by-doing allowing higher yields across manufacturing steps, and economies of scale resulting from increases in the sapphire wafer size.
+We estimate a combined contribution of these two factors to cost reductions for warm white GaN-based phosphor-converted LEDs to be at least 95.5\%, leaving only a minor residual of 4.5\% for all other mechanisms of innovation, including R\&D.
+
+Interestingly, comparable analyses of historical cost reductions for supply-side energy technologies show a very different pattern of relative contributions of different mechanism to cost dynamics.
+For example, Kavlak et al.\cite{kavlak2018evaluating} showed that for solar photovoltaic modules, increases in wafer size and manufacturing yield (with the latter associated with learning-by-doing) contributed only 11\% and 7\% cost reductions from 1980 to 2012, correspondingly, while innovations resulting from R\&D that increased the efficiency of solar modules reduced the costs by 23\%, with remaining cost declines explained by economies of scale and declines in material input prices.
+Similarly, Ziegler et al.\cite{Ziegler2021} showed that technological innovations resulting from public and private R\&D accounted for as much as 54\% of cost declines in lithium-ion batteries from the late 1990s to early 2010s.
+while economies of scale contributed 30\%, and learning-by-doing associated primarily with improved manufacturing yields reduced the costs only by 2\%.
+Such dramatic differences in relative contributions of the same mechanisms to the cost dynamics between white LED lighting, solar photovoltaics, and lithium-ion batters, as well as our findings on the progress in consumer experience metrics, indicate that LED lighting as a demand-side technology can be more directly affected by end-user needs and demands than technologies focused on energy generation and storage.
+It may also have sufficiently different characteristics affecting its manufacturability and marketability that require different types of policy support and response than supply-side energy technologies.
+Whether our findings for white LED lighting are generalizable to broader demand-side technologies, and what are the reasons behind the observed differences in the patterns of technological progress in different types of energy technologies, remains to be explored in future research.
+
+Our analysis of the sources, mechanisms and enablers of the identified technology spillovers highlights the critical role played by a deep understanding of the physical, chemical and optical phenomena underlying the operation of LEDs, as well as materials science and technology and nanotechnology involved in the production of LEDs, for past and future advances in LED and solid-state lighting technology.
+Specifically, deep physical understanding of LED device energy loss channels had enabled important innovations in LEDs that increased several sub-efficiencies in LEDs and will continue to do so, as expected by eminent experts in the field whom we interviewed.
+A practical implication of this finding is that additional research in these areas and a more deliberate search for relevant external knowledge may accelerate expected future advances in LED technology and its applications both in SSL and technology areas beyond general lighting.
+According to our observations, corresponding knowledge spillovers can be deliberately stimulated, among other factors, by measures such as knowledge exchange events and long-term partnerships between academia and industry, interdisciplinary training and hiring, dedicated mission-driven public R\&D funding, and ensuring certain freedom of search in academia.
 These observations also further reinforce broader arguments made against the dichotomy of basic and applied research \cite{narayanamurti2016cycles, narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
 
-Finally, the lessons learned about the drivers of innovation, technology spillovers, cost reductions and performance improvements in LEDs may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}.
-By comparing these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
+While this study focused on quantifying the impact of spillovers on efficiency and consumer performance metrics (which were essential for adoption) and the enabling mechanisms of spillovers, the large role of the combination of manufacturing scale and learning-by-doing on cost reductions also stands out.
+This role cannot be explained without the demand-pull policies related to regulations and bans around incandescent lightbulbs, which paved the way for scale up and industry investments in manufacturing.
+In this way, the role of demand-pull policies is pivotal when it comes to cost metrics, as it has been with solar PV \cite{nemet2019solar}, although perhaps even greater with white LED lighting.
+
+While additional work is necessary to determine if the patterns we find for innovation in white LEDs apply in other demand side technologies, there are some emerging findings that would have implications.
+First, the fact that efficiency channels with greater physical understanding were more likely to be shaped by spillovers could indicate that investing into research that develops such physical understanding could reduce the cost of entry for new researchers and inventors and lead to spillovers and faster innovation.
+This could be more important for small-scale consumer facing technologies or products (e.g., cars, refrigerators, windows, food) that may face lower barriers to entry in research compared to, for instance, nuclear power or offshore wind.
+Second, we also find that innovation focused on consumer performance was incredibly important for this demand side consumer-facing technology and that the improvement mostly relied on integrating external knowledge, suggesting that a joint focus on cost and consumer experience in R\&D is both necessary but also requires a particularly high level of external knowledge integration, and possibly wider hiring practices.
+This finding also suggests that innovation in demand-side technologies, which have been according to some scholars ‘marginalized’ \cite{Wilson2012}, should be further promoted.
+A third possible hypothesis to test with future work on other demand side technologies is related to potentially larger contribution of demand-pull policies on cost metrics over time through their impact of scale up and learning by doing compared to R\&D.
+If corroborated in other technologies, this finding would add to evidence on the major role of regulations, bans and incentives shaping technology costs and trajectories.”
 
@@
 where $L(F_{t=T}, F_{t=0})$ is the logarithmic mean of $F$ values at times $t=0$ and $t=T$.
-In practice, this mean can be approximated, simplifying calculations.
+As we show in Supplementary Method 8, this mean can be approximated in practice, simplifying calculations.
 These terms contain no residuals, therefore it can be shown that the overall improvement in the device efficiency due to improvements in individual sub-efficiencies is equal to the sum of these improvements in individual sub-efficiencies:

### 2024-03-30 2_SSL_NE/article/article_supplementary.tex
various updates

@@
 Years indicated correspond to the earliest identified patent priority date.
-Black frames around certain designs indicate chip designs not brought to large-scale production.
+Dashed black frames around certain designs indicate chip designs not brought to large-scale production.
 Additional context on this figure is provided in \cref{sec:subeff_progress}.
@@
 
-While generally comprehensive, the model is limited by the availability of performance data for the most recent manufacturing equipment, as well as the confidential nature of cost structures of Asia-based manufacturing lines.
-A complete treatment of the limitations of the model is provided in \cref{sec:costmodel_structure}.
-
 \clearpage
@@
 
+Unfortunately, benchmarking the total cost computed with our model against industry data is not possible.
+While the US Department of Energy has managed to convince companies to share their cost structure (shown in Supplementary Figure 16), they never share \textit{absolute} cost.
+The closest industry publication ever comes to disclosing absolute cost data is in their reporting data on \$/klm at the chip-level.
+Our model (like most other manufacturing cost models we reviewed) computes only the total manufacturing cost, since a combined computation of cost and performance would require a highly complex treatment not only of the cost associated with equipment performance, but also of the associated changes in the bulk semiconductor, phosphor, packaging, etc.
+To the best of our knowledge, this has not been attempted, at least in a publicly available fashion, for solid-state lighting devices.
+
+\clearpage
 \subsection{Yielded Cost}

### 2024-03-30 2_SSL_NE/article/article_main.tex
more edits

@@
 \usepackage{lineno}
-%\linenumbers
+\linenumbers
 
@@
 
-Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning by doing\cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale\cite{johansson2012global, iea2020perspectives}.
-The role of these mechanisms at different stages of technology life cycle, from research and development to demonstration, market formation and diffusion\cite{grubler2012policies}, is an area of active research \cite{kavlak2018evaluating,Ziegler2021}.
+Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning by doing \cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale \cite{johansson2012global, iea2020perspectives}.
+The role of these mechanisms at different stages of the technology life cycle, from research and development to demonstration, market formation and diffusion \cite{grubler2012policies}, is an area of active research \cite{kavlak2018evaluating,Ziegler2021}.
 Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Acemoglu_2022}.
@@
 
-Importantly, different mechanisms of innovation in a particular technology play out within a broader innovation system comprising multiple actors, networks, and institutions\cite{grubler2012policies, Anadon2016}.
-As a result, these mechanisms are affected by various drivers and barriers, including forces of supply and demand\cite{Mowery1979}, public policies, and regulation\cite{anadon2009policy}.
+Importantly, different mechanisms of innovation in a particular technology play out within a broader innovation system comprising multiple actors, networks, and institutions \cite{grubler2012policies, Anadon2016}.
+As a result, these mechanisms are affected by various drivers and barriers, including forces of supply and demand \cite{Mowery1979}, public policies, and regulation \cite{anadon2009policy}.
 
@@
 There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies.
-For example, spillovers can provide critical external knowledge inputs to the R\&D process \cite{kolesnikov2021spillovers}, enable the re-use of experience from different sectors in learning-by-doing\cite{iea2020cleanenergy}, and lead to cost reductions, e.g., via repurposing the manufacturing equipment from other sectors\cite{Stephan2021}.
+For example, spillovers can provide critical external knowledge inputs to the R\&D process \cite{kolesnikov2021spillovers}, enable the re-use of experience from different sectors in learning-by-doing \cite{iea2020cleanenergy}, and lead to cost reductions, e.g., via repurposing the manufacturing equipment from other sectors \cite{Stephan2021}.
 
-The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diode (LED) lighting.
-Among demand-side technologies, general illumination is a particularly important area for climate change mitigation efforts, as it currently accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}.
+The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diodes (LEDs) for general illumination.
+Among demand-side technologies, lighting is a particularly important area for climate change mitigation efforts, as it currently accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}.
 It is also an area of rapid recent technological change: since the introduction of first commercial white LEDs in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements.
@@
 \cite{azevedo2009transition}.
-See Supplementary Note 16 for the full list of sources and references.}
+See Supplementary Note 8 for the full list of sources and references.}
  \label{fgr:history_efficacy}
@@
 Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications.
-See Supplementary Note 16 for the full list of sources and references.}
+See Supplementary Note 8 for the full list of sources and references.}
   \label{fgr:cost_lamp_small}
@@
 
-These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent light bulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}.
+These dramatic improvements in lighting technology, supported by the introduction of efficiency regulations phasing out incandescent light bulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}.
 As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year in the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced in the same year by all photovoltaic installations in these regions.
@@
 To address these questions, we first identify three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost.
-The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Notes section 2.
+The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Method 2.
 Next, using data from literature review, patent analysis, and expert interviews, we trace the historical improvements in these metrics in warm white GaN-based phosphor-converted LEDs—currently the most widely adopted variant of white LED technology which we thus choose as the focal technological area for our analysis—from their introduction to the lighting market in 2003 to 2020, the year with the most recent data available at the time of data collection.
+
 We find that total LED device efficiency of such LEDs increased from 5.8\% to 38.8\% between 2003 and 2020.
@@
 
-Using the mixed-methods approach described in Methods, we have collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress.
-The results of our data collection efforts are presented in Supplementary Figure 7 and discussed in Supplementary Notes section 4.
+Using the mixed-methods approach described in Methods, we collected data on the historical progress in individual sub-efficiencies of warm white phosphor-converted GaN-based LED devices, along with white LED innovations that have driven this progress.
+The results of our data collection efforts are presented in Supplementary Note 4.
 Using this information, we calculate the overall lamp efficiency $\eta_L$ for the best performing mid/high-power phosphor-converted warm white LED devices in four years: 2003, 2010, 2016 and 2020.
@@
 \cref{fgr:breakthroughs_efficiency} shows the overall magnitude of contributions of identified LED innovations and technology spillovers to improvements in LED efficiency over time across different sub-efficiencies.
-The full list of identified LED innovations considered in our study is provided in Table SI2 in Section 3 of the Supplementary Information document.
+The full list of identified LED innovations considered in our study is provided in Supplementary Note 5.
 The list of corresponding technology spillovers is provided in \cref{tab:spillovers}.
@@
 In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels.
-The estimates shown here for the contributions of different channels do not show uncertainty ranges because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, and the fact that we focus on a best case scenario.
+Since we focus on a best case scenario, and because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, these estimates do not show uncertainty ranges.
 Thus, our point estimates should be interpreted as the expected values in a best case scenario.
@@
 When the spectrum has significant energy at the edges of the sensitivity curve, energy is lost.
-Details for each architecture are provided in \cref{fgr:chip_architecture_overview}.
+Details for each architecture are provided in Supplementary Note 3.
 Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4.
@@
 $\eta_{C_R}, \eta_{C_{Y/G}}$ and $\eta_S$ are relevant only to warm white spectrum LEDs, which were not available in 1997.
-Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures $\eta_{LE}$ no longer depends on ITO, see \cref{fgr:chip_architecture_overview}.
+Note: ITO current spreading layer affects different sub-efficiencies in different chip architectures, e.g., in modern flip-chip architectures $\eta_{LE}$ no longer depends on ITO, see Supplementary Note 3.
 Physical limits on sub-efficiencies are indicated by black horizontal lines.
@@
 In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable.
-Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into the white light required for general illumination .
+Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into white light.
 It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs.
-The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor that could generate only cool white light \cite{bando1998development}.
+The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor that resulted in cool white light only \cite{bando1998development}.
 As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.
@@
 
-Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 4, we find that only a single innovation related to LED consumer experience improvements was purposefully developed for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
+Notably, from detailed descriptions of the history of innovations in this list, provided in Supplementary Note 6, we find that only a single innovation related to LED consumer experience improvements was purposefully developed for application in solid-state lighting: the 2016 SALON phosphor compound \cite{Hoerder2019,seibald2019phosphor}.
 All other innovations in the list were either originally developed for non-LED applications or prominently used knowledge from areas of science and technology beyond LED or SSL (see also \cref{tab:spillovers}).
@@
 
-\begin{table}[H]
+\begin{table}[h!]
     \footnotesize
@@
 These differ from the years shown in \cref{fgr:consumer_experience} which correspond to the earliest publication of spectral data for LED products that relied on those innovations.
-Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 4.
+Detailed descriptions of the history of each innovation and relevant spectral data are provided in Supplementary Note 6.
 Desig.=Designation, Ref.=Reference.}
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
  \centering
+ \vspace{-5mm}
  \includegraphics[width=\textwidth]{figures/breakthroughs_consumer-experience.pdf}
@@
 Horizontal red arrow indicates desirable higher values of colour rendering index (CRI).
-See Supplementary Note 16 for the full list of sources and references of all phosphor represented.}
+See Supplementary Note 8 for the full list of sources and references of all phosphor represented.}
  \label{fgr:consumer_experience}
@@
 
+\clearpage
 \section{Manufacturing Cost Improvements}
@@
 In the case of wafer diameter, the larger the wafer, the more LED chips can be produced from a single wafer.
-The wafer diameter commonly used in LED manufacturing has been steadily increasing from 50mm ($\sim2$ inch\footnote{Industry measures wafer size in mm.
-Still, a 50mm=1.9685Inch wafer is frequently referred to as a "2-inch wafer".}) in 2003 to 200mm ($\sim$8 inches, referred to as “8 inch") in 2020.
+The wafer diameter commonly used in LED manufacturing has been steadily increasing from 50mm ("2 inch"\footnote{Industry measures wafer size in mm.
+Still, a 50mm=1.9685Inch wafer is frequently referred to as a "2-inch wafer".}) in 2003 to 200mm (“8 inch") in 2020.
 We capture this effect in the model by calculating the associated number of die per wafer (DPW) for each representative wafer diameter d \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$ (see Supplementary Figure 7 in Supplementary Note 3).
@@
 
-Our findings are further supported by a preliminary sensitivity analysis, presented in Section 5 of the Supplementary Information section, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
+Our findings are further supported by a preliminary sensitivity analysis, presented in Supplementary Discussion 1, where we find that the sensitivity of the cost model to variation in its main parameters decreases over time with the increase of the number of DPW.
 
-In Supplementary Note 6, we also provide a comparison of our model results with past cost report and projections published by the U.S.
+In Supplementary Discussion 2, we also provide a comparison of our model results with past cost report and projections published by the U.S.
 Department of Energy (DOE) on the basis of the \textit{LEDCOM} cost model \cite{ledcomv2} and industry data reported as part of SSL round table meetings.
@@
 \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
-\caption{\textbf{Historical changes in white LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and Supplementary Note 3) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+\caption{\textbf{Historical changes in white LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and Supplementary Method 6 for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
 location.
@@
 Panel D: Cumulative manufacturing yield after each process step for years 2003, 2012 and 2020.
-For a graphic representation of the manufacturing process, see the diagrams in Supplementary Figures 2-6.
+For a graphic representation of the manufacturing process, see the diagrams in Supplementary Note 2.
 Abbreviations: Litho – Lithographic Process, Insp.
@@
 
+\clearpage
 \section{Technology Spillovers}
 
-Following the frameworks for the analysis of technology spillovers proposed by Stephan and colleagues \cite{Stephan2021} and Kolesnikov et al.
+Following the frameworks for the analysis of technology spillovers proposed by Stephan et al.
+\cite{Stephan2021} and Kolesnikov et al.
 \cite{kolesnikov2022technology}, we synthesize information from the interviews, historical records, literature review and citations in patents and publications to identify which innovations in white LED technology were driven by external knowledge originating in areas of science and technology beyond white LED lighting.
-We find that at least nine LED innovations, listed in Table 2, clearly involved such external knowledge.
+We find that at least nine LED innovations, listed in \cref{tab:spillovers}, clearly involved such external knowledge.
 We then analyze the sources, mechanisms and enablers of corresponding technology spillovers, as well as their contributions to progress in white LED technology.
@@
 
+\clearpage
 \begin{table}[h!]
@@
         \textbf{Inv.} & \textbf{S/O} & \textbf{Comm.} & \textbf{LED Innovation} & \textbf{Spillover} & \textbf{Enabler} & \textbf{Origin} & \textbf{Ref.} & \textbf{Area of Improvement} \\ \hline
-        1926 & 1994 & 1996 & LED phosphors &  Use of phosphors for light down conversion in LEDs & ???
-& Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
-        1967 & 1996 & 1996 & YAG:Ce phosphor & Use of YAG:Ce phosphor in a first white LED product & ???
-& Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
-        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & ???
-&  Chemistry (S), Materials science (S) & \cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
-        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & targeted search & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T)  & \cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview})\\ \hline
-        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & targeted search & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare \cref{fgr:chip_architecture_overview}) \\ \hline
-        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & research conference & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
-        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & ???
-& Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo,simmonsfinal,ledprof_nexxusqd,bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
-        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & ???
-& Chemistry (S), Materials science (S) &\cite{paulusz1973efficient,radkov2009red,Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
-        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & ???
-& Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
+        1926 & 1994 & 1996 & LED phosphors &  Use of phosphors for light down conversion in LEDs & Firm history in phosphor manufacturing for CRT applications & Materials science (S), Cathode ray tubes (T) & \cite{bright1972electric,shimizu1994sheet,cho2017white} & Enabled light down conversion in LEDs \\ \hline
+        1967 & 1996 & 1996 & YAG:Ce phosphor & Use of YAG:Ce phosphor in a first white LED product & Firm history in phosphor manufacturing; firm working in multiple sectors & Chemistry (S), Materials science (S), Fluorescent lighting (T), Cathode ray tubes (T) & \cite{blasse1967new,bando1996,bando1998development,shimizu1999light,cho2017white} & Enabled white LED products, $\eta_S$, $\eta_C$ \\ \hline
+        1967 & 1996 & $<$2002 & YGAG phosphor & Use of YGAG phosphor in first warm white LEDs & Firm history in phosphor manufacturing; firm working in multiple sectors &  Chemistry (S), Materials science (S) & \cite{holloway1969optical,bando1998development,shimizu1999light,Mueller2002} & Enabled warm white LEDs, $\eta_S$, $\eta_C$ \\ \hline
+        1982 & 1996 & $<$2010 & Patterned sapphire substrate (PSS) & Use of anti-reflective properties of substrate patterns in LEDs & not identified & Optics and photonics (S), Materials science and technology (S,T), Nature-inspired material design (T) & \cite{moharam1982diffraction,krames1998ordered,feezell2018invention,Narukawa_2010} & $\eta_{LE}$, $\eta_{IQ}$ (depending on the chip architecture, compare Supplementary Note 3)\\ \hline
+        1971 & 1999 & $<$2005 & Indium tin oxide (ITO) current spreading layer & Use of ITO current spreading layer in white LEDs & Public mission-driven R\&D funding; Flexibility of public funding (through DARPA); Industry-academia partnership & Optics and photonics (S), Materials science and technology (S,T), Optoelectronic devices (T) & \cite{vossen1971rf,fraser1972highly,margalith1999indium} & $\eta_{Vf}$, $\eta_{LE}$ (depending on the chip architecture, compare Supplementary Note 3) \\ \hline
+        1997 & 2002 & 2005 & 258 phosphor & Use of luminescent ‘258’ nitridosilicate compound as LED phosphor & Freedom of search; Communication at a conference; Cultural and language proximity; International industry-academia partnership & Chemistry (S), Materials science (S) &\cite{Huppertz1997,mueller2004phosphor,MuellerMach2005} & $\eta_S$, $\eta_C$ \\ \hline
+        1984 & 2003 & 2009 & Quantum dot-based phosphor & Use of quantum dots for light down conversion in LEDs & Public mission-driven R\&D funding; Communication at a conference - integration enabler; Commercial success on a different market - integration enabler & Solid-state physics (S), Photochemistry (S), Nanotechnology (T) &\cite{fojtik1984photo,simmonsfinal,ledprof_nexxusqd,bourzac2013quantum} & $\eta_S$, $\eta_C$ \\ \hline
+        1972 & 2005 & 2015 & PFS phosphor & Use of knowledge in luminescent materials and skills in "wet" chemical synthesis to synthesize PFS compound and optimize it as LED phosphor & Public R\&D funding - integration enabler; International industry-academia partnership; Alumni connections (?) & Chemistry (S), Materials science (S) &\cite{paulusz1973efficient,radkov2009red,Murphy2015} & $\eta_S$, $\eta_C$ \\ \hline
+        2008 & 2013 & 2015 & SLA phosphor & Use of knowledge about existing cuboidal nitride compounds to identify and synthesize structurally similar SLA phosphor & Industry-academia partnership & Structural chemistry (S), Materials science (S), Solid-state physics (S) &\cite{Park2008New,schmidt2013new,Pust2014} & $\eta_S$, $\eta_C$ \\ \hline
     \end{NiceTabularX}
@@
 
+\clearpage
 \section{Discussion}
@@
 In addition, we observe (but so far have not been able to quantify) a likely small but meaningful role of learning-by-doing in efficiency improvements.
-Interestingly, we find that technology spillovers are most prevalent in those device loss channels that are well understood at the physical level (e.g., light extraction and spectral efficiencies, see Fig.
-4 and Table 2).
+Interestingly, we find that technology spillovers are most prevalent in those device loss channels that are well understood at the physical level (e.g., light extraction and spectral efficiencies, see \cref{fgr:breakthroughs_efficiency} and \cref{tab:spillovers}).
 We find that improvements in device loss channels that are governed by complex quantum effects at the atomic level (e.g., droop) and at present can only be described heuristically, have, on the other hand, come mostly from incremental manufacturing process improvements resulting from learning-by-doing with little evidence for spillovers.
@@
 We estimate a combined contribution of these two factors to cost reductions for warm white GaN-based phosphor-converted LEDs to be at least 95.5\%, leaving only a minor residual of 4.5\% for all other mechanisms of innovation, including R\&D.
+This effect of manufacturing scale and learning-by-doing on cost reductions cannot be explained without the demand-pull policies related to regulations and bans around incandescent lightbulbs, which paved the way for scale up and industry investments in manufacturing.
+In this way, the role of demand-pull policies is pivotal when it comes to cost metrics, as it has been with solar PV \cite{nemet2019solar}, although perhaps even greater with white LED lighting.
 
@@
 These observations also further reinforce broader arguments made against the dichotomy of basic and applied research \cite{narayanamurti2016cycles, narayanamurti2021genesis} and the calls for open, inclusive and flexible research cultures \cite{Stephan2021}.
-
-While this study focused on quantifying the impact of spillovers on efficiency and consumer performance metrics (which were essential for adoption) and the enabling mechanisms of spillovers, the large role of the combination of manufacturing scale and learning-by-doing on cost reductions also stands out.
-This role cannot be explained without the demand-pull policies related to regulations and bans around incandescent lightbulbs, which paved the way for scale up and industry investments in manufacturing.
-In this way, the role of demand-pull policies is pivotal when it comes to cost metrics, as it has been with solar PV \cite{nemet2019solar}, although perhaps even greater with white LED lighting.
 
@@
 To overcome the limitations of existing literature and patent analysis, in this study we rely on a multi-method approach to data collection and analysis, augmenting information obtained from a comprehensive review of the primary scientific literature \cite{Haddaway_2014}, device datasheets, relevant patents, and industry publications with information gained from semi-structured interviews with experts from academia and industry, bottom-up manufacturing cost modelling, and own calculations of LED device sub-efficiencies.
-We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Note 2) and identify its sources in individual innovations and technology spillovers.
-We briefly describe our approach to each method below and provide further methodological details in Supplementary Note 3.
+We then synthesize information obtained with multiple methods to track the historical progress in white LED technology over time across the three groups of metrics (see Supplementary Method 2) and identify its sources in individual innovations and technology spillovers.
+We briefly describe our approach to each method below and provide further methodological details in Supplementary Methods.
 
@@
 All interviews were conducted between November 2019 and April 2022 by means of video conferencing and lasted for about one hour.
-An anonymized summary of the background of interviewed experts is provided in Supplementary Table 1.
+An anonymized summary of the background of interviewed experts is provided in Supplementary Method 5.
 Notably, all our interviewed experts came from Europe or USA, with none representing Asia, which potentially may have biased our findings particularly for the earliest and latest periods of white LED history dominated by LED manufacturers in Japan and China, correspondingly.
@@
 Specifically, sub-efficiencies related to the emission spectrum of phosphor down-converted LED devices (i.e., spectral efficiency and light conversion efficiency)  were computed from the spectral data, often reported in LED device specifications, using the \texttt{colour-science} package for Python.
-We also used the same approach to calculate CRI and luminous efficacy of radiation of LED devices (see Supplementary Note 4 and Supplementary Figure 10 for the associated spectral data and calculation results).
+We also used the same approach to calculate CRI and luminous efficacy of radiation of LED devices (see Supplementary Note 7 for the associated spectral data and calculation results).
 This approach allowed us to quantify the improvements related to consumer experience and phosphor development in LEDs.
@@
 For each of these three years, the most prevalent manufacturing equipment was identified through industry periodicals, archived website data from the Internet Archive, and expert interviews.
-Because the architecture of LED chips has changed significantly since the introduction of the first commercial white LED devices in 1996, three different chip architectures are considered in the model: classical chips, flip chips, and chip-scale package flip chips (see Figure 3 and Supplementary Figures 2-6 for the details of each architecture).
+Because the architecture of LED chips has changed significantly since the introduction of the first commercial white LED devices in 1996, three different chip architectures are considered in the model: classical chips, flip chips, and chip-scale package flip chips (see Supplementary Notes 2-3 for the details of each architecture).
 The details of the manufacturing process for each architecture were collected from the scientific literature, textbooks and relevant patents.
@@
 
-The cost model we developed is based on a cumulative approach to yielded cost\cite{becker2001use}.
+The cost model we developed is based on a cumulative approach to yielded cost \cite{becker2001use}, which we describe in detail in Supplementary Method 7.
 In this approach, the yielded cost $C_{Y_i}$ of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$.
@@
 
-Further details on our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, detailed calculations for the yielded costs, as well as the model’s limitations, are provided in Supplementary Note 3.
+Further details on our manufacturing cost model, including its structure and equations, manufacturing process flows for the chip architectures under consideration, input data, detailed calculations for the yielded costs, as well as the model’s limitations, are provided in Supplementary Method 6.
 

### 2024-03-30 2_SSL_NE/article/article_main.tex
various updates

@@
 Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Acemoglu_2022}.
-For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Castelnuovo_2005} more often than other mechanisms can result in substantial or even radical improvements in technology performance.
+For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Castelnuovo_2005} more often than other mechanisms have been associated with substantial or even radical improvements in technology performance.
 On the other hand, learning-by-doing is typically associated with continuous incremental improvements in the technology resulting from accumulated experience in technology demonstration, manufacturing, and use.
@@
 Our interviews also revealed a lesser but still important role of incremental manufacturing process improvements most likely resulting from learning-by-doing.
-However, substantial further research is needed to quantify the relative contribution of learning-by-doing vis-a-vis technological innovations to this progress.
+However, substantial further research is needed to quantify the relative contribution of learning-by-doing vis-a-vis R\&D to the efficiency increase during that time period.
 On the other hand, we find that innovations driven by technology spillovers contributed to nearly all consumer experience improvements in white LED lighting technology, thus playing an important role in widespread adoption of SSL in general illumination applications.
@@
 In \cref{fgr:breakthroughs_efficiency} we also compare, for the first time, efficiency improvements across sub-efficiencies over time, contrasting them with the physical limits of the corresponding loss channels.
-Since we focus on a best case scenario, and because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, these estimates do not show uncertainty ranges.
+Since we focus on highest performing devices, and because we lack reliable information about the distribution of input parameters, which would require us to arbitrarily assign range values for inputs, these estimates do not show uncertainty ranges.
 Thus, our point estimates should be interpreted as the expected values in a best case scenario.
@@
     \caption*{Note: Inv.
-- Year of initial invention, identified by the earliest literature source describing the original invention or idea in a field outside white LED lighting that eventually ‘spilled over’ to the latter field.
+- Year of initial invention, identified by the earliest literature source describing the original invention or idea in a field outside white LED lighting that eventually ‘spilled over’ to white LED lighting.
 S/O - Year of spillover to LED; Comm.
@@
 A third possible hypothesis to test with future work on other demand side technologies is related to potentially larger contribution of demand-pull policies on cost metrics over time through their impact of scale up and learning by doing compared to R\&D.
-If corroborated in other technologies, this finding would add to evidence on the major role of regulations, bans and incentives shaping technology costs and trajectories.”
+If corroborated in other technologies, this finding would add to evidence on the major role of regulations, bans and incentives shaping technology costs and trajectories.
+
+Finally, the lessons learned about the mechanisms and drivers of rapid progress in white LED technology may have broader implications both for other demand-side or energy-efficiency technologies, and for low-carbon energy technologies with similar characteristics such as the granularity of technology, its modularity, and complexity \cite{malhotra2020accelerating, Wilson2012}.
+By comparing the role of these factors at a granular level across different low-carbon and energy efficiency technologies, we can identify and generalize key cross-technology patterns or differences that would help us formulate recommendations for industry and policymakers aimed at accelerating further clean energy innovation for climate change mitigation.
 
@@
 and S.K.
-carried out the main experiments under the supervision of L.D.A.
-The data was analysed by M.P.W.
-All authors contributed to the discussion of the results and commented on the manuscript.
-The draft was written by M.P.W.
+carried out the main empirical work under the supervision of L.D.A.
+Performance simulation software code was written by M.P.W.
+Data was analyzed by M.P.W.
 and S.K.
+All authors contributed to the writing of the manuscript.
 

### 2024-06-20 2_SSL_NE/article/article_main.tex
updated references

@@
 
-Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning by doing \cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale \cite{johansson2012global, iea2020perspectives}.
+Improvements in technology cost and performance, which in this study we generally call “innovations”, can occur via different mechanisms, including (but not limited to) learning-by-doing \cite{WRIGHT_1936, Arrow_1962}, targeted research and development (R\&D) efforts, or economies of scale \cite{johansson2012global, iea2020perspectives}.
 The role of these mechanisms at different stages of the technology life cycle, from research and development to demonstration, market formation and diffusion \cite{grubler2012policies}, is an area of active research \cite{kavlak2018evaluating,Ziegler2021}.
-Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Acemoglu_2022}.
-For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Castelnuovo_2005} more often than other mechanisms have been associated with substantial or even radical improvements in technology performance.
+Different mechanisms can result in different kinds of innovations, characterized by what changes in the technology (e.g., its architecture design, materials, components, or manufacturing process) and the scale of resulting improvements (e.g., radical or incremental) \cite{Garcia_2002}.
+For example, R\&D efforts (sometimes also called “learning-by-researching”) \cite{Cohen_1989} more often than other mechanisms have been associated with substantial or even radical improvements in technology performance.
 On the other hand, learning-by-doing is typically associated with continuous incremental improvements in the technology resulting from accumulated experience in technology demonstration, manufacturing, and use.
@@
 
-The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diodes (LEDs) for general illumination.
+The aim of this paper is to identify and, where possible, quantify the contribution of different mechanisms and technology spillovers to the historical progress in a specific technology – white light-emitting diodes (LED) for general illumination.
 Among demand-side technologies, lighting is a particularly important area for climate change mitigation efforts, as it currently accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}.
@@
 Shown for reference are the average prices for compact fluorescent (CFL) and incandescent light bulbs, with the latter assumed constant based on the average in the covered time period.
-Note that projections are consistently at the lower end of the retail price data.
-This is because retail price data includes more expensive luminaire types, while projections are for traditional light bulbs only.
-Projections also do not include value added tax.
+Note that DOE projections are consistently at the lower end of the retail prices, because retail price data often includes expensive luminaire types, while projections were made for traditional light bulbs only and did not include sales taxes.
 Source: own synthesis of data on sales prices collected from various consumer watchdog databases and publications.
@@
 
-To address these questions, we first identify three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED chip manufacturing cost.
+To address these questions, we first identify three sets of metrics to quantify historical progress in LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, specifically the colour rendering index (CRI) and colour temperature (CT); and 3) LED package manufacturing cost.
 The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Method 2.
@@
 We find that total LED device efficiency of such LEDs increased from 5.8\% to 38.8\% between 2003 and 2020.
-A majority of this increase can be attributed to a series of innovations in white LED technology resulting from R\&D efforts, with at least 8.5\% of the cumulative increase attributed specifically to technological innovations driven by technology spillovers.
+The majority of this increase can be attributed to a series of innovations in white LED technology resulting from R\&D efforts.
+8.5\% of the increase can be attributed specifically to technological innovations driven by technology spillovers, corresponding to an increase in device efficiency of 2.8\%.
 Our interviews also revealed a lesser but still important role of incremental manufacturing process improvements most likely resulting from learning-by-doing.
@@
 The list of corresponding technology spillovers is provided in \cref{tab:spillovers}.
-Through the index decomposition analysis described in Section 2.5 in the Supplementary Information we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8 percentage points can be attributed specifically to innovations driven by technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
+Through the index decomposition analysis described in Supplementary Method 8 we find that out of the overall LED efficiency increase of 32.9\% from 5.8\% to 38.8\% between 2003 and 2020, at least 2.8 percentage points can be attributed specifically to innovations driven by technology spillovers identified in this study, corresponding to 8.5\% of the total LED efficiency improvements between 2003 and 2020.
 
@@
 Details for each architecture are provided in Supplementary Note 3.
-Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 4.
+Details on the phosphors for light down conversion are provided in \cref{tab:phosphors} and Supplementary Note 6.
 Abbreviations: Scat.
@@
 = Nonradiative.
-Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 9 and spectral data in Supplementary Figure 10, adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
+Source: own elaboration based on data on sub-efficiencies presented in Supplementary Figure 7 and spectral data in Supplementary Figure 8, adapted from publications on the respective phosphors: YGAG (2003)\cite{Mueller2002}, 258 (2010)\cite{MuellerMach2005}, PFS (2016)\cite{Murphy2015}, SALON (2019)\cite{Hoerder2019}.}
  \label{fgr:waterfall}
@@
 In general illumination applications, a high colour rendering index (CRI) in combination with a specific, tunable range of possible colour temperatures is desirable.
-Both metrics are determined by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into white light.
-It allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations  associated either with phosphors or quantum dots used for light down conversion in LEDs.
-The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor that resulted in cool white light only \cite{bando1998development}.
+Both metrics are determined soley by LED device emission spectra, which, in turn, depend on the properties of materials used for the conversion of blue light generated by conventional GaN-based LEDs into white light.
+
+This allows us to establish the links between all major improvements in the two consumer experience metrics considered in this study and individual LED innovations associated either with phosphors or quantum dots used for light down conversion in LEDs.
+The first commercial white LED produced by Nichia in 1996 used a YAG (Yttrium Aluminium Garnet) phosphor activated with Cerium that resulted in cool white light only \cite{bando1998development}.
 As shown in \cref{fgr:consumer_experience}, after a series of innovations listed in \cref{tab:phosphors}, LEDs today can be tuned for high CRI performance and a range of desirable colour temperatures.
@@
         \textbf{Year} & \textbf{Desig.} & \textbf{Chemical formula} & \textbf{Description} & \textbf{Significance} \\ \hline
-        1996 & YAG & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs \\ \hline
+        1996 & YAG:Ce & $Y_3 Al_5 O_{12}:Ce$ & Yttrium aluminium garnet (YAG) phosphor activated with cerium & First LED phosphor, enabled white LEDs \\ \hline
         1996 & YGAG & $(Y_{1-x} Gd_x)_3 Al_5 O_{12}:Ce$ & Gadolinium-doped YAG phosphor & First red-shifted phosphor, enabled warm white LEDs \\ \hline
@@
  \caption{\textbf{Historical improvements in consumer experience metrics of phosphor-converted white LEDs.} Data points show the color temperature (CT) and colour rendering index (CRI) performance of the earliest identified representative white LED products with published spectral data that used phosphors listed in \cref{tab:phosphors}, each indicated by the phosphor label and publication year.
-Corresponding spectral data used to calculate CT values is provided in Supplementary Note 4.
+Corresponding spectral data used to calculate CT values is provided in Supplementary Figure 8.
 Three data points representing compact fluorescent lamps (CFLs)\cite{cie_reference}, shown as black triangles, are provided for comparison.
@@
 We find that the manufacturing cost of a single such LED decreased from 1.11\$ (in 2020 USD) in 2003 to 0.11\$ (2020 USD) in 2012 and 0.05\$ in 2020, a 95.5\% overall decrease.
-Our model shows that for the wafer-level manufacturing steps illustrated in Supplementary Figures 1-5, improved manufacturing yields and increases in the wafer size are jointly responsible for the largest contribution — 95.5\% — to the overall reduction of manufacturing cost per LED chip between 2003 and 2020.
+Our model shows that for the wafer-level manufacturing steps illustrated in Supplementary Figures 1-5, improved manufacturing yields and increases in the wafer size are jointly responsible for the largest contribution — 91\% — to the overall reduction of manufacturing cost per LED chip between 2003 and 2020.
 
@@
 Still, a 50mm=1.9685Inch wafer is frequently referred to as a "2-inch wafer".}) in 2003 to 200mm (“8 inch") in 2020.
-We capture this effect in the model by calculating the associated number of die per wafer (DPW) for each representative wafer diameter d \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$ (see Supplementary Figure 7 in Supplementary Note 3).
+We capture this effect in the model by calculating the associated number of die per wafer (DPW) for each representative wafer diameter d \cite{de2005investigation}: $d(2003)\sim 50 mm \rightarrow851 DPW$, $d(2020)\sim200 mm \rightarrow 26,838 DPW$ (see Supplementary Figure 13 in Supplementary Methods 7).
 With more than a thirty-fold increase in the number of die per wafer between 2003 and 2020, the contribution of the whole-of-wafer processing steps to the total cost of manufacturing an individual LED chip and package has dramatically declined over time.
@@
 \includegraphics[width=17.5cm]{figures/costmodel_results_years.pdf}
-\caption{\textbf{Historical changes in white LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and Supplementary Method 6 for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
+\caption{\textbf{Historical changes in white LED manufacturing cost structure.} Presented cost calculations are an outcome of our cost model (see \cref{sec:methods} and Supplementary Method 6) for a single low-to-mid power GaN-based, classic chip, phosphor-converted white LED package, assuming an ideal factory with state-of-the-art equipment at a U.S.
 location.
@@
 Ref.
-column lists literature sources for the represented discoveries, innovations and spillovers.
-Area of Improvement column represents the impact of spillovers on different aspects of white LED technology, e.g., improvements in particular sub-efficiencies.}
+column lists literature sources for the represented inventions, innovations and spillovers.
+Area of Improvement column represents the impact of spillovers on different aspects of white LED technology, e.g., improvements in particular sub-efficiencies.
+See Supplementary Note 6 for a detailed description of phosphor-related innovations and spillovers.}
     \label{tab:spillovers}
@@
 In contrast with efficiency and consumer experience improvements, we find that dramatic manufacturing cost reductions in white LEDs have been mostly driven by learning-by-doing allowing higher yields across manufacturing steps, and economies of scale resulting from increases in the sapphire wafer size.
-We estimate a combined contribution of these two factors to cost reductions for warm white GaN-based phosphor-converted LEDs to be at least 95.5\%, leaving only a minor residual of 4.5\% for all other mechanisms of innovation, including R\&D.
+We estimate a combined contribution of these two factors to cost reductions for warm white GaN-based phosphor-converted LEDs to be at least 91\%, leaving only a minor residual of 9\% for all other mechanisms of innovation, including R\&D.
 This effect of manufacturing scale and learning-by-doing on cost reductions cannot be explained without the demand-pull policies related to regulations and bans around incandescent lightbulbs, which paved the way for scale up and industry investments in manufacturing.
@@
 For example, Kavlak et al.\cite{kavlak2018evaluating} showed that for solar photovoltaic modules, increases in wafer size and manufacturing yield (with the latter associated with learning-by-doing) contributed only 11\% and 7\% cost reductions from 1980 to 2012, correspondingly, while innovations resulting from R\&D that increased the efficiency of solar modules reduced the costs by 23\%, with remaining cost declines explained by economies of scale and declines in material input prices.
-Similarly, Ziegler et al.\cite{Ziegler2021} showed that technological innovations resulting from public and private R\&D accounted for as much as 54\% of cost declines in lithium-ion batteries from the late 1990s to early 2010s.
-while economies of scale contributed 30\%, and learning-by-doing associated primarily with improved manufacturing yields reduced the costs only by 2\%.
-Such dramatic differences in relative contributions of the same mechanisms to the cost dynamics between white LED lighting, solar photovoltaics, and lithium-ion batters, as well as our findings on the progress in consumer experience metrics, indicate that LED lighting as a demand-side technology can be more directly affected by end-user needs and demands than technologies focused on energy generation and storage.
+Similarly, Ziegler et al.\cite{Ziegler2021} showed that technological innovations resulting from public and private R\&D accounted for as much as 54\% of cost declines in lithium-ion batteries from the late 1990s to early 2010s, while economies of scale contributed 30\%, and learning-by-doing associated primarily with improved manufacturing yields reduced the costs only by 2\%.
+Such dramatic differences in relative contributions of the same mechanisms to the cost dynamics between white LED lighting, solar photovoltaics, and lithium-ion batteries, as well as our findings on the progress in consumer experience metrics, indicate that LED lighting as a demand-side technology can be more directly affected by end-user needs and demands than technologies focused on energy generation and storage.
 It may also have sufficiently different characteristics affecting its manufacturability and marketability that require different types of policy support and response than supply-side energy technologies.
@@
 This could be more important for small-scale consumer facing technologies or products (e.g., cars, refrigerators, windows, food) that may face lower barriers to entry in research compared to, for instance, nuclear power or offshore wind.
-Second, we also find that innovation focused on consumer performance was incredibly important for this demand side consumer-facing technology and that the improvement mostly relied on integrating external knowledge, suggesting that a joint focus on cost and consumer experience in R\&D is both necessary but also requires a particularly high level of external knowledge integration, and possibly wider hiring practices.
+Second, we also find that innovation focused on consumer experience was incredibly important for this demand side consumer-facing technology and that the improvements mostly relied on integrating external knowledge, suggesting that a joint focus on cost and consumer experience in R\&D is both necessary but also requires a particularly high level of external knowledge integration, and possibly wider hiring practices.
 This finding also suggests that innovation in demand-side technologies, which have been according to some scholars ‘marginalized’ \cite{Wilson2012}, should be further promoted.
-A third possible hypothesis to test with future work on other demand side technologies is related to potentially larger contribution of demand-pull policies on cost metrics over time through their impact of scale up and learning by doing compared to R\&D.
+A third possible hypothesis to test with future work on other demand side technologies is related to potentially larger contribution of demand-pull policies on cost metrics over time through their impact of scale up and learning-by-doing compared to R\&D.
 If corroborated in other technologies, this finding would add to evidence on the major role of regulations, bans and incentives shaping technology costs and trajectories.
@@
 
-There are several important avenues of future research that are opened up by our analysis.
+There are also several important avenues of future LED research that are opened up by our analysis.
 First, future work could expand the cost model by collecting and including data for a broader set of chip architectures.
@@
 
-Relevant sources for the review were found in an iterative search process that involved two components.
-The first was the search in specialized patent and publication databases as well as company websites.
-The second component was the analysis of backward citations in the identified sources, starting from the reviews mentioned in Supplementary Note 1 and then iteratively repeating it for all newly identified sources, until no further relevant and significant new sources were found.
+Relevant sources for the review were found in an iterative search process that involved two stages.
+The first stage was the search in specialized patent and publication databases as well as company websites, following the search procedure and database search queries described in Supplementary Methods 1.
+The second stage was the analysis of backward citations in the identified sources, starting from the reviews mentioned in Supplementary Note 1 and then iteratively repeating it for all newly identified sources at the first stage, until no further relevant and significant new sources were found.
 We also relied on backward citations in these sources for the identification of technology spillovers, considering cited documents as indicators of knowledge origins of an innovation and analyzing whether those documents belonged to the white LED lighting technology domain or not.
@@
 According to this method, for an overall device efficiency function $F$ that is the product of variables $a, b$ that represent sub-efficiencies, the contribution of the change in a single sub-efficiency variable $a$ between times $t=0$ and $t=T$ can be estimated as \cite{Ang2019}
-%
+
 \begin{align}
@@
 
-The cost model we developed is based on a cumulative approach to yielded cost \cite{becker2001use}, which we describe in detail in Supplementary Method 7.
+The cost model we developed, presented in Supplementary Method 6, is based on a cumulative approach to yielded cost \cite{becker2001use}, which we describe in detail in Supplementary Method 7.
 In this approach, the yielded cost $C_{Y_i}$ of process step $1$ is defined as the ratio between the total cost of step 1 $C_1$ and the yield of step 1 $Y_1$.
@@
 
-The manufacturing cost model and associated input data is available are the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410658.
+The manufacturing cost model and associated input data are available in the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410658.
 The Python scripts that were used to compute the light quality metrics of \cref{fgr:consumer_experience} are available in the Zenodo repository with the identifier DOI: 10.5281/ZENODO.8410789.

### 2024-06-20 2_SSL_NE/article/article_supplementary.tex
various updates
updated references

@@
 Source: adapted and compiled from multiple patents and industry publications.
-See Supplementary Note 16 for the full list of sources and references.}
+See \cref{sec:sources} for the full list of sources and references.}
 \label{fgr:chip_architecture_overview}
@@
 \textit{Spillover} column indicates if the innovation involved technology spillovers.
-\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
+\textit{Source} column indicates the source of information about the innovation, with 'I' indicating expert interviews as such a source.}
     \vspace{-20mm}
@@
     \multicolumn{5}{c}{Droop $\eta_{droop}$} \\
+        \hline
+            \textit{Year} & \textit{Innovation} & \textit{Area of Improvement} & \textit{Spillover} & \textit{Source} \\
         \hline
@@
 \textit{Spillover} column indicates if the innovation involved technology spillovers.
-\textit{Source} column indicates the source of information about the innovation or improvement, with 'I' indicating expert interviews as such a source.}
+\textit{Source} column indicates the source of information about the innovation, with 'I' indicating expert interviews as such a source.}
     \label{tab:innovations_2}
@@
 
-\subsection{Literature Review Process}
+\subsection{Literature Search Process}
 \label{sec:litrev}
@@
 This makes attribution of changes in this metric to individual changes in device design or manufacturing difficult.
-For this reason, we do not use this metric only to show historical progress across different lighting technologies in Figure 1 in the main text of the article.
+For this reason, we use this metric only to show historical progress across different lighting technologies in Figure 1 in the main text of the article.
 
